### 2015-12-14 source/10/Paper proposal for EAGLE final.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Abigail S Graham}
\date{2015-11-16}
\begin{document}
\begin{enumerate}
\item FACE VALUE: THE POWER OF IMAGES AT APHRODISIAS
\end{enumerate}
How Digital Resources Can Tranform our use of Palaeography in understanding Inscriptions.

\bigskip

\ Dr. Abigail Graham

\bigskip

The University of Warwick (Lecturer)

\bigskip

Abigail.graham@warwick.ac.uk

\bigskip

Possibly for subject panel: Dating criteria for epigraphic material: possible approaches.

\bigskip

Abstract:

\bigskip

\ This paper will consider how the publication of a large digital corpus (The Inscriptions of Aphrodisias [Reynolds,
Roueche \& Bodard, 2007] has shaped the assessment of inscriptions, particularly regarding palaeography and the dating
of inscriptions. A case study of dedicatory inscriptions from the Temple of Aphrodite at Aphrodisias will explore how
our approach to palaeography and dating has evolved with digital resources, identifying areas where challenges remain
and considering how improvements could be made in both our approach to and in publication of epigraphic materials.

\bigskip

Keywords

\bigskip

Dating, Palaeography, Ordinatio, Letter forms, Aphrodisias, Context, Recarving.

\bigskip

\bigskip

1.1 Introduction

\bigskip

\begin{enumerate}
\item \begin{enumerate}
\item \begin{enumerate}
\item Caveat Lector: \ Defining Palaeography and traditional approaches
\end{enumerate}
\end{enumerate}
\end{enumerate}

\bigskip

As the co-ordinator and lecturer on three graduate level courses of Roman Epigraphy, I am invariably asked the same
question: how do you use letterforms to date inscriptions? My answer is always the same: “very carefully”. \ Studying
palaeography within the discipline of ancient epigraphy can be a journey into thorny hedge where one can easily “fall
into that category of human endeavor known as stylistic attribution and inevitably involve subjectivity”.\footnote{
Tracy, 1995, 3-4. } Stephen Tracy advises that palaeographic surveys should be carried out with caution: “Caveat Lector
must needs be our motto”. \ The study of palaeography is problematic on a number of levels, both in the way it is
defined and the ways in which it is employed. Many scholars dismiss lettering as a means of dating and they make an
important point: dating by a single criterion, especially a stylistic one, is somewhat precarious. When studied in
isolation, letterforms present stylistic variations that may be characteristic of a specific individual, workshop or an
urban area. While observations about carving techniques can be helpful in specific case studies, they are more
problematic when applied generally on a broader scale (e.g. to larger geographical areas or time trame) where
archaisms, local styles, and variations can create distortions.\footnote{ Susini 1973; Petrucci 1993; Di Stefano
Manzella 1995 163-181, Cooley 2012, 433; Edmundson 2015 122-125.} Similar caution would be applied in dating a
sculpture on the basis of a hair fragment. Analysis of statues considers a number of factors, material, hairstyle,
drapery and/or context. Inscriptions are doubly difficult, as they fall into categories of both text and an object.
While there is potentially more information, there is also a greater chance that it will be contradictory. Thus the use
of lettering, an “imprecise science”, is better used in combination with a number of different factors.\footnote{
Harris 1989, 26-9; Bodel 2001 3-5; Di Stefano Manzella 2007; 393-418 and Cooley 2012, 432-435. } Despite the
aforementioned limitations, a number of informative palaeographic studies have been produced. The success of these is
based on detailed commentaries on a specific corpus of material, a transparent methodology and the incorporation of
numerous high quality images.\footnote{ For example: Gordon 1957 offers more than 50 plates and figures, Di Stefano
Manzella 1987 provides 218 illustrations, Tracy 1990, 1995 and 2003 uses over 60 plates and figures in each volume. \ }

\bigskip

A second issue in the study of palaeography involves the access and publication of epigraphic materials. \ Access to
large corpora of inscriptions has traditionally been limited to a small audience of scholars and site visitors. Apart
from a few museum collections, which happen to have inscriptions arranged in roughly chronological order,\footnote{
Perhaps the finest example of a museum in which one can gain and understanding of palaeography (and inscriptions as
whole) is the Museo Epigrafico Nazionale, Roma. } there are few places where one can visually experience the
development of carving styles over time. For the lucky few who attain access (and permission) to study a large corpus
of inscriptions, publishing these texts with supporting images can be challenging and expensive. Studies of
palaeography in ancient inscriptions have often been, by necessity, selective with images making it difficult for both
the author and readers to develop a detailed understanding of carving trends and practices. In this traditional format,
inscriptions were also separated from their archaeological context and the accompanying artwork while the reader,
viewing only the lettering, was often removed from the visual elements of the inscription (e.g. the type of stone, use
of spacing and decoration, letter size). \ This is a suboptimal way of assessing epigraphic evidence.

\bigskip

The advent of online corpora have increased both the access and the development of discussions regarding epigraphic
monumentality, including new methodologies, approaches as well as attempts to redefine genre classifications and
terminology.\footnote{ Woolf 1996, 22-39, Eck 2010, Panciera 2012, 1-10, Graham 2013, 1-17. \ For a recent collection
and analysis of epigraphic databases, see Tom Elliot’s work \ (chapter 5) in Bruun and Edmundson 2015, 78-85. \ } \ A
number of recent studies have used changes in the appearance of inscriptions over time, such as the use of different
media, decorative and paratextual elements\footnote{ \ One marks in the text see Susini 1973, 26 and Cooley in Jansen
2014, 143-155.} (e.g. ivy leaves as interpuncts (hederae distinguentes),\footnote{ \ Hommel 1970, 293-303. }
abbreviations\footnote{ Gordon, “Supralineate Abbreviations” 1957, 111.} and spacing between words\footnote{ On using
space to reconstruct inscriptions see Alföldy 1995, 195-226, Grasby 2002, 151-176.} alongside lettering, as dating
criteria.\footnote{ For an extensive list of dating conventions cf..Di Stefano Manzella 1987, Chapter 20. }
\ Panciera’s recent article, in particular, advocates the significance of the public context and visibility of
inscriptions.\footnote{ “I would propose to regard as an ‘inscription’ any particular type of written human
communication of the sort that we would today call unidirectional…not being addressed to a person or to a group but to
a collectivity, and for this reason is made with the location, writing technique, graphic form and impagination, mode
and register of expression chosen because they are most suitable to the attainment of its intended goal.” (Panciera
2012: 8).} \ In this vein, it is worth considering how the physical characteristics of an inscription belong within a
broader assessment of a culture of writing. Is palaeography the study of letterforms alone? Was a focus on letterforms
a deliberate choice or a product of the traditional constraints in accessibility and publication of epigraphic
materials? The following case study from Aphrodisias will explore these questions further.

\bigskip

1.1.3 Outline

\bigskip

The focus of this paper is a series of rather unimpressive column dedications from the temple of Aphrodite at
Aphrodisias. \ While the information recorded in the texts is unremarkable, the journey of these inscriptions from
their publication as part of an epigraphic corpus to a recent digital publication reveals an epic transformation in the
format and approach to these materials. \ This study will begin with brief overview of previous published editions
(e.g. MAMA volume VIII, 1962) followed by an assessment of the information available on the current Inscriptions of
Aphrodisias (2007) website. Through careful assessment of text (including formula, vocabulary and spelling) and its
presentation (the arrangement of the text, use of decoration and spaces, as well as lettering) in the images provided,
this survey will demonstrate how the availability of published images and inclusion of dating criteria have increased
the amount of information available whilst also adding clarity to the process of dating an inscription. By examining
how we define and use palaeography to evaluate inscriptions on the Inscriptions of Aphrodisias website (IAph 2007), we
can observe how the discipline has evolved and what changes may be possible in approaches and the publication of
epigraphic material.

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

[Warning: Draw object ignored]

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\begin{enumerate}
\item \begin{enumerate}
\item Publishing inscriptions: A Case Study of Column Dedications from The Temple Of Aphrodite at Aphrodisias 
\end{enumerate}
\end{enumerate}

\bigskip

1.2.1 A brief introduction to the inscriptions IAph 1.4-1.6.

\bigskip

Three inscriptions, each of which record the dedication of a column at the Temple of Aphrodite, will be the subject of
this survey. \ The first two texts were noted as early as the 18th century and copied in a notebook by the British
Architect Deering in the early 19th century. A third version of the text, was \ uncovered during excavations at the
temple site by the French Engineer Gaudin in the early 20th century.\footnote{ A comprehensive history of the history
and bibliography, as well as a description of the resources can be found on the IAph website:
http://insaph.kcl.ac.uk/iaph2007/iAph010006.html. } The inscriptions are recorded on tabella ansata (0.745 x 0.465m) as
part of fluted marble columns of the peristasis, some of these have been reconstructed in modern restorations on the
site.

\bigskip

These texts can be dated through a number of different criteria. \ Contextual association with construction of the
temple, dates between the end of the 1st century BC and the early 1st century AD, though the land was clearly in use
well before this time. \footnote{ Smith 1996, 41-42 and Reynolds 1990, 37. References to Sulla’s dedications are in
Appian BC 1.97. \ Caesar and Augustus’ acknowledgements of the sacred space are evident in IAph 8.27, 8.31 and 1.1. }
\ This is corroborated by coins depicting the temple, which date from 2 BC- 14 AD.\footnote{ Coin type 41 depicts
Augustus (OBV) and the temple of Aphrodite (REV) in MacDonald 1992, Plate V R131. } Prosopography is also informative:
Gaius Julius Zoilos (freed by Caesar or Octavian), who dedicated the theatre at Aphrodisias ca. 28 BC, also dedicated
the naos of the temple (IAph 1.2), perhaps posthumously. \footnote{ The titles on the architrave inscription, soter and
euergetes, are absent from Zoilos’ other inscriptions and may imply a posthumous dedication, perhaps by the boule and
demos (Reynolds 1990, 38).} The benefactor of the columns in this case study, Eumachos Diogenes, also comes from an
established family that flourished well into the 2nd century AD.\footnote{18 The success of Eumachus Diogenes’ family,
which included the first known Aphrodisian to hold a procuratorship in the 2nd century AD, reveals an enduring
significance for his family monuments in the city (Reynolds 1999b, 327-334).} \ Finally, the formula of the text the
inscription and the vocabulary, particularly the term philokaisar, suggest a late Republican or Augustan
date.\footnote{ The formula of the dedication, which lists the benefactor first is evident throughout the Hellenistic
period in Aphrodisias (and Asia Minor) until the early Imperial Period, after which a new formula (beginning with
recipients (e.g. Aphrodite and Imperial recipients) is predominant (cf. Graham 2013, 4-7). \ The Augustan and early
Tiberian uses of the philokaisar are known from a dedication at Ioulis dated ca. 27 BC - AD 14 (SEG XLVIII (1998) no.
1129) and in a monument to Ti. Cl. Drusus in Patara (SEG XLIV (1994) no. 1205). For significance and date of the title
see Buraselis 2001, 101-105. } \ Before encountering the inscriptions face to face, the dating, function and meaning of
these inscriptions appears to be quite straightforward. \ So let us examine the experience of viewing these
inscriptions various published formats.

\bigskip

1.2.2. Publishing IAPH 1.4 and 1.5 in MAMA VIII (nos. 347 \& 348): A series of copies?

\bigskip

Cormack published the first comprehensive catalogue of inscriptions from Aphrodisias as part of his Monumenta Asiae
Minoris Antiqua volume VIII in 1962. \ This was a great undertaking that included texts as well as images. Two of the
three texts \ (IAph 1.4 \& 1.5) were published here as MAMA 437 and MAMA 438 respectively, while the third text was
merely mentioned as a further copy.

\bigskip

\bigskip

\bigskip

\bigskip

[Warning: Draw object ignored][Warning: Draw object ignored]

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

[Warning: Draw object ignored]

\bigskip

The two published texts reveal similar inscriptions with a few variations.\footnote{ MAMA 437 and 438 are represented in
their digital format as published the Packhum website \ (Aphrodisias 108 and 109 (respectively)).
http://epigraphy.packhum.org/inscriptions/main?url=oi\%3Fikey\%3D256987} \ IAph 1.4 (MAMA 437) has a slightly less
impressive ordinatio, particularly in line 4 with an odd line break, a misspelling of Ammias (as Amias) and Adrastus
(line 5) with a sigma omitted. It is difficult to determine how or why these errors occurred without indications of
spacing or an examination of the stone. \ One could argue that spelling and arrangement of the text were not important
to a broader audience at Aphrodisias, however, this is negated by a second copy of the dedication (IAph 1.5, MAMA 438),
which portrays a more skillful execution and arrangement of the text. Spelling errors are rectified and the arrangement
of the text creates a distinction in Line 5 between Eumachus and Ammias, and for the demos, which is isolated on the
last line. These subtle variations only emerge with a close reading of the text, and it would be easy to overlook them.
The published format of the texts also makes them appear similar in lettering and arrangement (e.g. left indentation).
\

\bigskip

The small (thumbnail) and low quality image provided in MAMA 437 provides a basic understanding of the appearance of the
inscription, though it does not allow for a critical assessment of the lettering or arrangement of the text. The image
reflects lettering that is recognizable (to the trained eye) as late Republican/ early Augustan at Aphrodisias with
deep incisions, small serifs, as well as varying letter heights, but it is not sufficient for a detailed study. MAMA
438 is not published with a photograph, so one must rely upon a comparison of the texts alone.

\bigskip

\ The third text version of this text is mentioned in the commentary and based on the information provided (two similar
texts and a reference to a “copy”) one would expect that it was very similar. The same editorial practice is employed
in the parallel column dedications from a different benefactor (I.Aph 7 \& I.Aph 8), where 2nd version of the text is
noted (IAph 1.8) but only one (I.Aph 1.7, MAMA 450) was published. The apparent verisimilitude of the column
dedications is corroborated by the published images, which depict two inscriptions (MAMA 437 and MAMA 450) with similar
late 1st c. BC/early 1st c. AD lettering. When publishing a large corpus of inscriptions, referring to copies is
understandable. However, treating texts as “copies” as opposed to individual monuments can, and in this case will,
prove problematic.\footnote{ For another study of “copy” inscriptions at Aphrodisias see Graham 2015 (forthcoming).}
While MAMA VIII offered a broader audience access to the inscriptions at Aphrodisias, the approach to the inscriptions
and the quality of the images also imposed limitations that made it difficult to analyse the physical elements of the
inscription or to question the proposed texts and restorations. \

1.2.3 \ Living in a Digital World: The Publication of Column Dedications in IAph 2007 (IAph 1.4.)

\bigskip

The Inscriptions of Aphrodisias project was a groundbreaking endeavour both for publishing epigraphic materials and
employing Epi Doc conventions with XML for marking up ancient documents. Its success has inspired a number of
significantly larger projects with international scope and collaboration. The approach to the materials is detailed
with categories covering physical characteristics, history of discovery and bibliography (which is important given the
wealth of superlative scholarship).\footnote{ It is worth noting that websites, however they develop, will remain
supplementary to published scholarship.} \ In addition to search functions, the page is more interactive, the texts
have been revisited, translations added, as well as commentary with weblinks to parallel texts. The organization of the
inscriptions by context allows the reader to view the inscriptions in context and to gain a better understanding of the
area’s “epigraphic habit” (Figure 1: Screenshot of \ IAph 4 webpage).

\bigskip

 \includegraphics[width=14.605cm,height=8.726cm]{PaperproposalforEAGLEfinal-img001.png}

\bigskip

\bigskip

In addition to the format of the webpage, a number of images have been published along with a drawing from Deering’s
notebook from the early 19th century. \ These resources provide an opportunity not only to view the inscription but to
see how it has been studied over time. The dating of the inscription includes a list of the criteria upon which it was
based: lettering, context, prosopography, adding a degree of transparency to the dating process. While the lettering is
not subject to further description, the inclusion of weblinks to parallel texts allows the viewer to compare and
contrast different inscriptions, developing his/her understanding this element.

\bigskip

The published texts of IAph 1.4 and MAMA 437 present one significant variation: the inclusion of space indicators within
the text. This practice allows the reader to see how the use of space relates with the text of the inscription,
particularly in the case of line 4, where challenges in the text (misspelling of Ammias and carrying over of a single
letter from the previous line), can be observed in the inclusion of a vacat at the end of the line.\footnote{ Vacats,
more sparingly employed in late Republican and Augustan inscriptions at Aphrodisias, usually serve a grammatical and/or
decorative function (e.g. giving distinction to a name or key elements/individuals in the text (cf. building
dedications at the Sebasteion IAph 9.1, 9.25, 9.112). This vacat serves little grammatical function, and is likely to
be a result of the arrangement of the text/ omission in Ammias’ name. } \ The indentation on the last line also reveals
a left orientation of the line which is more common in Late Republican/Augustan texts at Aphrodisias (e.g. IAph 1.1,
1.7, 1.8 and 1.38), as opposed to the “justified” approach (indentations on both sides of the text to accentuate a name
or word), which is more common in later Imperial inscriptions at Aphrodisias.

\bigskip

\bigskip

\bigskip

[Warning: Draw object ignored]

[Warning: Draw object ignored]

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

High quality images on the website facilitate connections between the published text and the inscription. In the image
of IAph 1.4 (Figure 2. Photograph of IAph 1.4), one can to observe how descriptive elements manifest themselves in an
inscription: how varying letter sizes (2-3 cm) represent a lack of uniformity and crowding as space becomes more
cramped, and how, after the beautifully spaced upper lines (lines 1-3) the carver struggled (line 4) to fit the letters
in, possibly noticing his spelling error only when there was extra space at the end of the line. \

\bigskip

\bigskip

 \includegraphics[width=8.758cm,height=7.668cm]{PaperproposalforEAGLEfinal-img002.jpg}

\bigskip

Alongside spelling and the arrangement of the inscription, Letterforms are more readily observed. Overall, the lettering
is less uniform than later Imperial dedications, serifs are small and the strokes are both thick and deeply incised.
Letter height varies significantly as does the size of omicrons (though this can be seen in a poorly rendered text
throughout the Roman period). Angular forms such as alpha, lambda, and mu all intersect at the top meeting at sharper
angles than the more square versions of these letters that predominate in Imperial (post Augustan)
inscriptions.\footnote{ Compare with Hadrianic Lettering in the next section of the paper.} \ Letters also bear
stylistic elements of the time: epsilons have a connected middle bar (this is often disconnected in Imperial versions
of this letter), the rhos have small legs, making them appear more like their Latin counterpart, and the omicron has an
oval shape with serifs only at one end of the lower bars.\footnote{ “Legged” rhos are observed primarily in Augustan
and Julio-Claudian dedications and also on coins (cf. note 16). This style letterform is increasingly less common after
Flavian period at Aphrodisias. } Comparing these letterforms with parallel dedications at the temple (IAph 1.1, 1.2
(Zoilos’ dedications), 1.7 and 1.8) reveals many similarities. Letterforms alone are not diagnostic but when combined
with other elements, they can add to our understanding of a date.

\bigskip

\bigskip

1.2.4. Looks can be deceiving. When Inscriptions defy our expectations: IAph 1.5 and IAph 1.6.

\bigskip

Based on the publication of these inscriptions in MAMA VIII, one could easily come to the conclusion that all the column
dedications at the Temple of Aphrodite looked quite similar. However, upon visiting the site of Aphrodisias in the
summer of 2004, I was (yet again) to be denied a simplistic interpretation of an inscription. Courtesy of the IAph2007
website, a broader audience can now bear witness to complex nature of these inscriptions. \ The first disparities
between the two inscriptions (IAph 1.4 and IAph 1.5) can be observed in a comparison of letter sizes. IAph 1.4 records
letters between 2-3 cm (with a variation of 1 cm), while IAph 1.5 varies only between 2.5- 2.75 cm (a variation of
.25cm). The reductions in variation of letter size are part of a general standardization in letterforms during the
Julio-Claudian period at Aphrodisias.\footnote{ These Julio-Claudian texts record little if any variation in the size
of letter forms. } While significant variations in letter size can be observed in poorly executed inscriptions
throughout Aphrodisias’ history, one may not expect such a significant variation between two “copies”.\footnote{
A.Graham 2015 (forthcoming) illustrates a further case study of “copy” inscriptions at Aphrodisias.} \ The published
text of IAph 1.5 also indicates indentations on both sides of the word demos in the bottom line. While double
indentations are not unknown in Augustan inscriptions at Aphrodisias, they are more common in later Imperial
inscriptions.\footnote{ Double indentations in an Augustan text are attested (IAph 12.301, dating to 23-25 BC), but the
practice is more common in Julio-Claudian dedications (Sebasteion: 9.34, 9.36, 9.37, 9.38, 9.39, The city wall 12.515
(Claudian)).} These discrepancies, which suggest that the two inscriptions may have been less similar in appearance,
illustrate the importance of reading the descriptive elements of an inscription carefully.

\bigskip

[Warning: Draw object ignored][Warning: Draw object ignored]

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

An image \ (Figure 3 Photograph of IAph 1.5) illustrates further differences between the appearance of this inscription
and the comparative materials from the site (IAph 1.1, 1.2, 1.4, 1.7 and 1.8).

\bigskip

\bigskip

 \includegraphics[width=9.036cm,height=6.985cm]{PaperproposalforEAGLEfinal-img003.jpg}

The high quality image reveals square and regular letterforms (e.g. omicrons are less oval and more precisely rendered
throughout) with a clear contrast between slender strokes and large deep triangulate serifs. There are no legged rhos,
the middle bars of the epsilons are separated from the stem, and the omegas are circular with two large bars that are
heavily serifed on both ends (e.g. compare the omega on the bottom line of IAph 1.4 with its counterpart of IAph.1.5).
These traits along with others (e.g. letter size and use of spacing noted above) are commonly attributed to late 1st
/2nd c. AD inscriptions at Aphrodisias, and are absent from comparative Augustan materials on the site (e.g. IAph 1.1,
1.2, 1.4, 1.7 and 1.8, including a Julio-Claudian dedication 1.102). Although one must be careful with stylistic
factors, especially when archaisms can be used, it is worth noting when a number of features that are attributed to a
later period, seem to suddenly emerge nearly a century early.\footnote{ Archaising and imitation of earlier lettering
is evident in at least one inscription (IAPH 13.116) Reynolds 1982, 155-166). } Any one of these features, letter
variations, stylistic letterforms, use of serifs, indentations in the text, correction of errors in a previous text,
would not stand well alone, but taken together, a more compelling case can be made for the reassessment of this
inscription’s date. \

\bigskip

The website records the date of this inscription just like IAph 1.4: citing “context, lettering and prosopography.” For
those trying to gain an understanding of letterforms at Aphrodisias, this is somewhat confusing and it is not the only
example in which the lettering and organization of the inscription do not match date provided.\footnote{ Honours for
Zoilos (IAph 8.203) is dated by “prosopography” to the 1st c. BC but the letters are square Imperial forms and do not
match Zoilos’ other dedications (8.1, 1.1,1.2). Honours for P. and M. Vinicius (IAph 3.101) are dated as “Augustan” and
“Tiberian” by “lettering” though prosopography is known (there is some controversy cf. Reynolds 1982,175). Stylistic
elements (arrangement, spacing and decoration), particularly in M. Vinicius’ base, reflect qualities of
Claudian-Flavian period inscriptions (which is not excluded by prosopographical dates). \ In both cases, conflicting
dating criteria are omitted. \ \ \ } The date of the text is not incorrect, insofar as it was originally inscribed at
this time and the prosopography as well as the formula of the text support this. \ However, one struggles to see how
the lettering and/or the arrangement of the text corroborate this date. An answer may be found in further examination
of the inscription’s context and prosopography. In addition to a number of earthquakes, new temenos of the temple was
added under the emperor Hadrian, a time during which Athenagoras’ descendants were alive and prospering in
Aphrodisias.\footnote{ Smith and Ratté, 1995, 43. For an analysis of the temenos plan, see Doruk in Roueche and Erim
1990, 66-74. \par For Athenagoras descendants, cf. note 18 and Reynolds 1999b, 327-334} \

\bigskip

One benefit of the website is that it facilitates searches for parallel texts in this context and this period. A
Hadrianic building dedication from the temple, IAph 1.174 (Figure 4: Photograph of IAph 1.174) reveals similarly
rendered letterforms with little variation in size and a number of stylistic similarities: thin strokes with deep
triangulate serifs, the omicron with double serif bars, an epsilon with a separated crossbar. Similar observations can
be made on a number of Trajanic and Hadrianic texts at Aphrodisias (IAph 4.308, 5.9, 5.208 (Hadrianic Baths)). A theory
of later recarving would reconcile a number of disparities in this inscription.

\bigskip

\bigskip

 \includegraphics[width=10.836cm,height=4.092cm]{PaperproposalforEAGLEfinal-img004.jpg}

\bigskip

This is not to say that IAph 1.5 was definitely a Hadrianic recarving, but to observe that, contrary to what is recorded
in the IAph2007 dating criteria of this inscription, the lettering and arrangement of text on the stone do not reflect
an inscription that is an obvious contemporary with the other column dedications at the temple. The context and
prosopography of the inscription do not rule out a later date and the text, which rectifies issues in both organization
and orthography of IAPh 1.4, reflects an inscription that may have responded to an earlier monument.

The entry for IAph.1.6 represents a different response to a similar problem. \ It demonstrates that scholars are willing
to use letterforms to support a recarving when the case is sufficiently extreme. The text, only mentioned in MAMA VIII
as a “copy”, presents a number of irregularities, repeating some errors, correcting others, and making quite a few new
ones. Whilst maintaining the first 3 lines of the arrangement in IAph 1.4, this inscription compounds the error on line
4 by reduplicating two letters in the name Diogenes. Line 5 repeats the misspelling of Ammias, corrects the misspelling
of Adrastos (line 6), but creates two new errors in line 7, missing out the nu on Molon and Olynpias, then repeating
the column phrase in lines 9- 10.

\bigskip

[Warning: Draw object ignored]

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

Reading this inscription is even more of challenge (Figure 5. Photograph of I.Aph 1.6), one can suffer vertigo as the
lines run up and down and the letters run into the margins. There is no use of spaces or decorations to clarify or
distinguish sections of the text and there are quite a few inadvertent errors to make it difficult, even to the trained
eye. Although the prosopography and context support an earlier date, the entry suggests that this “inelegant” text was
“recarved ?” The entry does not, however, propose a date or explain why, in this instance, recarving is a viable
conclusion. \ While a methodology is evident in the IAph2007 dating format, it is not employed consistently or
transparently in this case. \

\bigskip

\bigskip

\bigskip

 \includegraphics[width=7.23cm,height=5.549cm]{PaperproposalforEAGLEfinal-img005.jpg} 
\includegraphics[width=7.396cm,height=5.456cm]{PaperproposalforEAGLEfinal-img006.jpg}

\bigskip

\bigskip

The inclusion of numerous images proves crucial here, where moss now covers parts of the text that were legible in the
1970’s. The lunate omega on the bottom line (barely visibly in the recent photo) along with the lunate sigmas indicate
that the inscription is from the late Antique period at Aphrodisias, possibly after an earthquake in 359 BC (for
parallels see ALA 29 and 30), before the conversion of the temple into a Christian church (after 450 AD).\footnote{
R.R.R. Smith 1996,41-42.} Acknowledging that the text was reinscribed centuries later affords further insight into both
the inscription and act of recarving as a process that often changed an inscription but did not necessarily improve it.

\bigskip

The three inscriptions betray fundamental differences, which reveal a rich and complicated tale of a column’s life at
Aphrodisias. While minor differences in the text do not change our translation of the words, a close analysis of the
resulting inscription informs our understanding of the arrangement of the text, the potential dating of the
inscription, as well as the relationship between text and monument. Recarving was not a highly unusual phenomenon in
the ancient world but by minimizing the conflicting elements of these inscriptions, one potentially overlooks this
aspect of an inscription.\footnote{ Thomas and Witschel 1995, 135-177. }

\bigskip

1.2.5 \ Conclusions on the case study

\bigskip

This survey of column dedications has demonstrate how a series of inscriptions, which were represented as a series of
similar texts in MAMA VIII can be seen in a fundamentally different way in IAPh 2007. The digital publication has a
number of advantages: it applies a more rigorous approach to the text in XML thus illustrating the arrangement of the
inscription more clearly, the date is given some transparency through a list of applied criteria, the descriptors, such
as the lettering size of the inscriptions, provide more information. The images of the inscription in drawing and its
context provide an invaluable resource that allow the reader to better understand the relationship between text,
inscription and context, including references and better accessibility to parallel inscriptions. In terms experiencing
an inscription in a digital world, this is probably as close a person can come to assessing the face value of an
inscription. \ It is a tremendous step forward in addressing the longstanding limitations inherent in the publication
of epigraphic evidence, though some challenges remain. \ The final section will consider how we might further use this
information, in terms of adding clarity and transparency to the dating process, as well as in our approach to these
materials. \

\bigskip

\ 1.3. \ Palaeography in a Digital World: Monumental Problems \& Solutions

\bigskip

\begin{enumerate}
\item \begin{enumerate}
\item \begin{enumerate}
\item Digital Epigraphy. New Method: New Methodology?
\end{enumerate}
\end{enumerate}
\end{enumerate}

\bigskip

For students and academics alike, dating by letter forms is an almost “mystic” practice: something upon which experts
often comment but more rarely explain or demonstrate in practice. \ Recent epigraphic handbooks keep the term
palaeography at arms length, using it interchangeably with discussions of letterforms.\footnote{ \ In the index (Bruun
and Edmundson et al., 2015, p. 880) \ “palaeography” is cross-referenced with “Letter forms” and though used by O.
Salomies (chapter 9), the word is not defined in the preceding chapters (Bruun and Edmundson et al., 2015, 155).
\ Letterforms are described generally with charts of figure numbers rather than images to directly illustrate changing
styles (Edmundson 2015, pp. 123-4). Cooley’s manual offers a more detailed assessment of lettering with case studies to
illustrate the limitations of using lettering (Cooley 2012, 423-33)} \ The caution of these scholars is justified and
understandable: conclusions based on a single element of an inscription are precarious and, perhaps more importantly,
they represent a mode of scholarship that views inscriptions in a fundamentally different way than they were viewed in
antiquity. Writing, both ancient and modern, is a product of a number of factors, all of which functioned together in
the image of writing (margins, lettering, indentations punctuation).\footnote{ \ Woolf 1996, 25-27, Bodel 2001, 3-5,
\ Cooley 2012, 433-437; Panciera 2012, 1-10. \ } As few would look at a document today and say “that’s a fine Helvetica
10 point!,” we should be cautious in an assessment of palaeography that excludes the visual elements which were
inextricably linked an inscription’s appearance (e.g. alignment, margins, spacing, punctuation). Modern definitions and
studies of manuscripts suggest a broader scope of inquiry, which includes handwriting together with decorations and
spatial organization as well as subsequent comments in the margins.\footnote{ \ Buonocore 2015, 21-37. } These
practices suggest a discipline with an interest not only in the evolution of lettering but in the appearance and
development of a culture of writing.

\bigskip

While traditional text-based modes of publication such as MAMA VIII have, at times, constrained the study of hands to an
analysis of lettering, digital corpora offer new opportunities to view, analyse and incorporate broader scope of visual
elements in the interpretation of an inscription. \ This is not to say that the traditional methods or charts of
letterforms development should be abandoned, merely that the methodology could be expanded, as it has been in this case
study, to include aspects of textual organization (spacing, use of decoration and punctuation, letter size). The
question arises: how do we achieve this through the website materials?

\bigskip

\bigskip

1.3.2 Employing change: How we might improve presentation of materials on the website.

\bigskip

One difficulty of the current site, observed in the assessment of IAph 1.5 is that despite the clear criteria for
dating, the terms do not reflect a consistent methodology in dating, or the contradictions in the process. This is
dangerous for those who simply accept the dates provided, and confusing for those who try to apply or develop a sense
of letterform development. When a recarving is suggested IAph 1.6, there is little to explain how we know it was
recarved (letterforms) or when the recarving took place (use of parallel texts), though both resources are available on
the website. The wealth of visual, textual and contextual information in IAph 2007, offers an opportunity, not only
clarify, but to add greater transparency to the process of dating and how lettering is used in an inscription. This
could be achieved by adding a few pages to explain and illustrate elements dating criteria (each with a significant
Caveat Lector on the use of these elements). Firstly, one could add a page outlining the dating criteria, providing a
brief description of each (e.g. context, lettering and prosopography) together with a sample case study (or two) of how
these factors are applied in an inscription.

\bigskip

\ Further improvements, could be made with additional pages on lettering and context. \ While lettering is undoubtedly
employed in dating inscriptions at Aphrodisias, one must also consider how this dating is represented in the published
materials. Epigraphers, who have described monumental lettering as ‘Augustan’, ‘distinctively Julio-Claudian’ or
‘Domitianic’ have already acknowledged that such distinctions exist at Aphrodisias, but the classifications remain
undefined.\footnote{ Studies of inscriptions describe lettering as “distinctively Julio-Claudian” (Reynolds 1981 (nos.
2\&3) 317-318), “Triumviral or Augustan” (Reynolds 1982 docs. 35-37, 159-163), and ‘Domitianic’ \ (Chaniotis 2004, no.
14), suggest that such distinctions are evident at Aphrodisias. \ } While such definitions are often problematic, so is
a situation in which a broader audience accepts and uses a series of dates, without understanding how certain criteria
have been used in formulating a date. Undertaking a detailed palaeography for a site is a massive endeavor, and one
that can prove contentious on a number of levels. However, a page about lettering (together with aforementioned element
of textual organization) could be added with brief descriptions of trends (periods of ca. 60-70 years) and references
to a few inscriptions as illustrations. This could be supported by a single case study, such as this one, to show how
these elements are used as well as how they can be problematic.

\bigskip

\ \ Finally, while inscriptions are given a good deal of context in IAph 2007, the epigraphic information remains
separated from the archaeological studies. Both of these are factors in the dating process and, as we observed in IAph
1.5 and 1.6 can often facilitate the study of an inscription, both in reconciling discrepancies and searching for
parallel inscriptions. \ With inscriptions grouped by context and an excavation history that is referenced (but not
explained) it might be helpful to have a brief building history for each context. In the case of Aphrodisias, these
materials and further bibliographies are \ available on NYU excavation websites and could be easily connected with
weblinks.\footnote{ NYU website. \ http://www.nyu.edu/projects/aphrodisias/home.ti.htm. \ Recent excavation reports are
reference here as well: \ http://www.nyu.edu/gsas/dept/fineart/academics/aphrodisias/aphrodisias.htm.}

\bigskip

The benefits and the challenges observed in the IAph 2007 digital publication are applicable to a number of digital
resources. \ As we attempt to bring corpora of unprecedented size to a digital realm, we must consider not only how we
represent this information but how we engage a wider readership in epigraphic materials. It would not take a great deal
of work to augment the scope of IAph 2007 from an academic resource into one also achieves a didactic aim of
illustrating how we use epigraphic evidence to date inscriptions. We have the potential now, to present inscriptions at
face value, not only as texts but as contradictory objects whose stories, whether conveyed on stone or a computer app,
are subject to the same conventions, complexities and imperfections as the humans who created them.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/13/EAGLE2016_Full_Paper_Verreth.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Herbert Verreth}
\date{2015-11-15}
\begin{document}
[Warning: Draw object ignored]Abstract

The Trismegistos database has recently created a geographical index for all Latin inscriptions. For the moment we have
67.820 geographical references attested in Latin documentary texts, but this rough starting material still has to be
refined. This paper describes how we undertook this task, which problems we encountered while doing so, and the choices
we made for the presentation of the material.

Keywords

Imperium Romanum, Latin, epigraphy, topography, geography, Trismegistos.

\bigskip

\ \ The Trismegistos (TM) database (http://www.trismegistos.org) of the Katholieke Universiteit Leuven in Belgium
gathers the metadata for all documentary and literary texts from Egypt and the Ancient World in general written in
whatever language or script between 800 BC and 800 AD. In this respect we try to collaborate as much as possible with
other scientific databases all over the world. We started some ten years ago with the papyrological material from Egypt
(both Greek and Egyptian), and the last few years - through the collaboration with EAGLE - we have also been
incorporating Latin epigraphical texts from the whole Roman word. For the moment TM shows the metadata for 479.629
texts, 306.925 of them written in Latin or containing Latin passages, but the number keeps on growing. Our direct EAGLE
partners Epigraphic Database Heidelberg (EDH), Epigraphic Database Roma (EDR), Hispania Epigraphica Online Database
(HEp) and Epigraphic Database Bari (EDB) mainly focus on inscriptions on stone and other important texts from the
regions they cover, but they usually omit the inscriptions and stamps on instrumentum domesticum or other minor
materials. Since TM also wanted to incorporate these texts, we were glad to find them in the Epigraphik-Datenbank
Clauss-Slaby (EDCS), which contains virtually all published Latin inscriptions, to fill in the missing gaps, a process
which is still going on.

\ \ Trismegistos is a relational database created within the computer program Filemaker (with as latest version
Filemaker Pro 14). Its contents are uploaded weekly to an online MySQL / PHP environment. Attached to the main TM Text
file are numerous other files, such as Collections, Archives, People, Names and Places. These files automatically copy
the relevant information for every card from the main file, so that no double work is needed. With the help of the
Papyrological Navigator (PN) (http://www.papyri.info) we developed a PHP environment with the assistance of a
programmer (Jeroen Clarysse), to tag all words starting with a capital occurring in the published papyrological texts,
which yielded in the end a full index of all personal names and toponyms in every Greek and Latin papyrus. This process
is nowadays called Named-Entity Recognition [NER]\footnote{ For more information on NER, cf. Broux, Yanne - Depauw,
Mark, Developing onomastic gazetteers and prosopographies for the ancient world through named entity recognition and
graph visualization: some examples from Trismegistos people, in 2nd International workshop on computational history. 2.
Barcelona, 11 November 2014 (Lecture notes in computer science, 8852), Berlin - Heidelberg, 2015, p. 304-313.}. These
TM files within People and Places are freely accessible online and can be looked up, questioned and investigated in a
number of ways. But that is a different story. Here we want to focus on our project to do the same for all published
Latin inscriptions.

\ \ The main credits for the whole set-up of this new project go to Mark Depauw, the Trismegistos director. The general
idea remained the tagging of all words starting with a capital, but this time Depauw tried a new approach, which was
completely imbedded within Filemaker. Since the Roman 'tria nomina' hardly occur in the Greek papyri, also a new
onomastical structure had to be devised for the automatic recognition. As a test case we choose the full text corpus of
the EDCS. All these texts were 'cut up' in capital clusters, i.e. strings of consecutive words all starting with a
capital. Words as filius, nepos and libertus were added to the string so that in most cases the full identification of
a person could be grouped together, e.g. Quintus Caecilius Quinti filius Quirina Mustacus (TM 332260), Caio Annioleno
Cai filio Arnensi Karthaginiensi Galliano (TM 349961) or Maesiaes Cai libertae Chrysidis (TM 244384). A minor
disadvantage of this corpus was the capitalisation of the first word of every inscription (which is not done in the
PN), which yields quite a significant number of mere nouns in the group of expected personal names and toponyms. Also
other words starting with a capital were not our prime goal: names of gods, religious festivals, months, army units,
ships, animals, mythological persons and Roman numbers. Excluded (for the moment) are also the names of the emperors
and the members of the imperial family, often occurring in a complicated titulature which is not easy to standardise.
In the end this yielded 898.134 capital cluster cards. From our previous projects we already had a fairly elaborated
reference corpus of Roman personal names, which was now expanded and used by Depauw to match every word in the capital
clusters with the names in the reference corpus. If there was a match, the case of the ending was added, e.g.

\ \ Caio Annioleno Cai filio Arnensi Karthaginiensi Galliano

\ \ dat\ \ dat \ \ \ \ \ gen \ filius\ \ tribus\ \  \ \ \ \ origo\ \ \ \ dat

\bigskip

 \includegraphics[width=10.881cm,height=7.505cm]{EAGLE2016FullPaperVerreth-img001.png}

\bigskip

The technical details of this complicated process are better discussed by Depauw himself at some other occasion. On the
basis of this matching all capital cluster cards were split up in two groups: (1) 'yes', this card contains a personal
name [454.183], and (2) 'no', this card does not contain a personal name [443.951]. Within the second group also other
labels have been added, e.g. army [23.201], god [15.663], emperor [41.944], which will be useful for future research.

\ \ This is also the phase where the toponyms come in (which can occur both in the first and in the second group).
Within the EAGLE project we created already a fairly large reference corpus for toponyms from the Roman empire, but
this corpus was enlarged by entering the toponyms occurring in the Itinerarium provinciarum Antonini Augusti [3.434]
and the Tabula Peutingeriana [3.287]. The TM Geo file now contains 46.277 toponyms from all over Egypt and the Roman
empire, both ancient and modern, which cover most of the places where ancient texts have been found, and most of the
toponyms mentioned in Egyptian papyrological and Latin epigraphical sources.\footnote{ For more information on TM Geo,
cf. Verreth, Herbert, Topography of Egypt online, in Proceedings of the 27th International Congress of Papyrology in
Warsaw, July-August 2013 (forthcoming).} We played with the idea of automatic matching, like we did for the personal
names, but except for the case of the relatively straightforward tribus names, this yielded no satisfactory results. A
lot of toponyms resemble personal names (e.g. Florentia, Venusia, (Fundus) Bassianus), and the automated identification
of strings such as Colonia Ulpia Traiana Augusta Fidelis Lepcis Magna (TM 198383) or Municipium Augustum Hipponiensium
Regiorum (TM 200133) seemed too cumbersome. In the end, we settled for plan B, which shows that in 'Digital
Humanities', the human component is still essential: we had to go through all 900.000 capital cluster cards manually,
identifying the toponyms in every cluster, adding the corresponding TM Geo number and - whenever necessary - correcting
the indications for yes/no and the automatic identification of the cases of the personal names. It was six months of
tedious work, resulting in 61.139 capital clusters with at least one toponym. No doubt some toponyms escaped our
attention, but we do hope to have identified the majority of the names involved. In this process, however, we also
encountered some set backs, especially in the longer texts and in the more complicated wooden or metal tablets: in the
beta version on which we worked, the creation of the capital cluster strings was not always so perfect as we had hoped
for, and also the line numbers automatically assigned to each cluster string have sometimes gone astray. Mark Depauw is
developing a new and improved version, especially in preparation for the much larger batch of personal names, where it
is virtually impossible to manually correct everything that has gone wrong. Due to these problems I guess that we now
have about only 80 to 90 \% of the toponyms occurring in all Latin inscriptions, but on the whole we are quite pleased
with the result and in due time the remaining toponyms no doubt will find their way into the database also.

\ \ Phase 1, the identification of toponyms in the capital cluster strings, was finished in the beginning of July 2015,
and we have now entered phase 2, the incorporation of the capital cluster file into the 'real' TM Georef file. For
every place listed in TM Geo we try to list all the ancient text references where that place is attested. The file of
these geographical references (TM Georef) is directly linked with the main TM Text file, so that every reference
automatically receives a chronological and a geographical context. Every toponym found in the capital cluster file is
exported to a separate Georef card. When a toponym exists out of several consecutive elements, like the colonia and
municipium examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as 'Bythina et
Ponthus', which cannot belong to the same capital cluster string because of the intermediary 'et', are exported double
and then afterwards joined manually. For the moment TM Georef lists 67.820 geographical references attested in Latin
documentary texts and 10.474 references attested in Latin literary texts (but except for Egypt the latter have not yet
systematically been entered). The Latin references make out almost 40 \% of the total of 196.794 Georef cards.

\bigskip

 \includegraphics[width=10.881cm,height=5.008cm]{EAGLE2016FullPaperVerreth-img002.png}

\bigskip

\ \ In this phase we also start comparing the reading of the toponym in EDCS with the readings of the same passage in
EDH, EDR, HEp and EDB, which can all be shown simultaneously in the Filemaker database. In theory it is possible to
automatically look for differences in readings among all these databases, but because each of them has its own
approach, there will be so many small differences in line numbering, punctuation, the use of uncertainty dots and the
way unconventional spelling in an inscription is indicated, that we doubt that there will be many exact matches. We
therefore think that it will not be worth the enormous amount of work that it would involve. Human observation is again
the answer and we do hope that our partners and the users of the geographical index will point out to us any mistakes
we have made or obsolete readings we have kept. For the online version we have to talk with our partners whether they
want to have their texts also shown on the TM page (like TM does for the texts from PN) or not. For the Open Access
CC-0 texts in the Europeana EAGLE portal, this will in any case be implemented in the future. Anyway it is always
possible to put a direct link on the Georef page to every partner that has the text in its corpus.

\ \ Since the users of TM must be able to look for specific spelling variants of each toponyms, all these references are
presented the way they are on the stone, with as little additions or emendations as possible, except of course for any
abbreviations at the end of the word; e.g. T(h)ra[c(um)] becomes Tra[c(um)], and Rom(a)e becomes Rome. In another field
the standardized nominative case is given, without brackets or uncertainty dots; e.g. Trax and Roma.

\bigskip

 \includegraphics[width=10.881cm,height=11.984cm]{EAGLE2016FullPaperVerreth-img003.png}

[Fig. 3]

\ \ For every text in TM we try to give a reference to the most authoritative edition, where the user can find the best
and most up to date reading and interpretation of that text. As authoritative editions we preferably use CIL, Année
épigraphique and more recent major editions such as RIB, ILAlg, ICUR or I. Alex. Imp. Any corrections to the reading of
the toponym with regard to that edition are to be listed in the field Bibliography, while the obsolete reading is
recorded in the field Note. If the correction comes from one of the online full text databases, we add a reference to
the number of the text in that database.

\ \ A major problem is the dating of the texts. Unfortunately not every edition provides a date for each inscription.
Even if the scholar who publishes the text, has a fairly good idea of the century or range of time to which the text
might belong, it is not always mentioned explicitly in the edition. For every Latin text for which TM did not receive a
date from its partners, we added the broad range of 199 BC till AD 799, hoping that this dating will become more
refined in the future.

\ \ The third and final phase involves the context of the toponym. In the field Detail we give a plain translation of
the immediate phrase to which the toponym belongs. The translation should be a standardised as possible, with termini
technici preferably added in Latin, so that the users can easily search for them in the database; e.g. 'Tiberius Iulius
Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris', 'praefectus
Aegypti', 'cohors I Tungrorum'. If the place is explicitly ascribed to a provincia or a region, this provincia is
listed in the field 'Administrative situation'. If a town is explicitly called an oppidum, a vicus or a civitas, this
information is listed in the field Status. By adding this information in searchable fields we hope that the user can
start asking quite specific questions; e.g. the first and last attestation of a place in the sources; the periods in
which a town was called a colonia or a municipium; the places in which the 'ala I Thracum Mauretana' has been attested.

\ \ TM is a relational database, which implies that is possible to get to the information from different angles. If
someone is studying a certain text, he can get a list of all toponyms mentioned in that text. On the other hand, if
someone is examining a certain place, he can find the list of all attestations for that place, in the order that he
wants. In some cases a scholar can have very specific questions, which are difficult to search through the online TM
search interface; it is quite well possible, however, that these questions can be easily answered in the more complex
Filemaker structure we have at our disposal; just contact us and we will try to solve the problem for you.

\ \ We are aware that this is a very succinct presentation of the new and exciting developments in Trismegistos Places,
but everybody interested is always more than welcome to ask for more information. Please feel free to provide us with
any addenda or corrigenda to the database you might have.
\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/17/Zuddas Epigraphy in Italian High Schools (final).tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Utente}
\date{2015-11-16}
\begin{document}
Abstract

\begin{figure}
\centering
\begin{minipage}{10.901cm}
1.\ \ EPIGRAPHY IN ITALIAN HIGH SCHOOLS\newline

Enrico Zuddas

University of Perugia

enricozudd@yahoo.com
\end{minipage}
\end{figure}
The paper focuses on the possible uses of Epigraphy when teaching Latin to Italian High School students. Inscriptions
(especially from the school’s territory) make Latin “come to life” because even the simplest of texts is real.
Inscriptions are an excellent tool to internalize declensions, but they can also broaden the horizons offered by the
literary sources. The Eagle databases offer ready to use materials to teachers who are less familiar with the
traditional supports.

Keywords

High School Education, Latin Grammar, Lexicon, Translations, EDR Database, Roman Asisium.

\bigskip

1.1. Introduction

High school students often feel that Latin is not a “real” language. Every attempt to use Latin as a modern language, by
the creation of words from the contemporary world or by using it in a conversation, makes Latin seem even more
artificial.

In this framework, for a Latin teacher the first two years of high school are the most challenging. This is when
students have to acquire the grammar, but hardly see the point of learning so many rules by heart (even if the
prescriptive approaches have now given way to descriptive linguistics). However, when students start studying
Literature and reading literary texts, they generally realize that it was worth making the effort. Hence, one can
imagine the frustration of Liceo Linguistico (Foreign Languages High School) pupils, who only do Latin for the first
two years, and will never see the results of such hard work.

One could object that in the first two years teachers should mix grammar and culture: but with only two lessons per
week, and classes with a high number of students, there would simply be not enough time. Moreover, the approach to
unedited literary texts is arduous because they contain too many unknown structures and often need to be at least
partially translated.

Inscriptions can be an effective solution to this problem: they make Latin authentic. They are often short and therefore
quite easily readable. They exist in reality, they can be seen and even touched – which should never be forgotten in a
society where visual communication is so important. Through epigraphy the Classics become, in a way, “multimedia”: this
is what scholars mean by the expression “words on stone”. Even if generally, when visiting a museum, inscriptions are
not of such interest as the statues or mosaics, they do strike students as something concrete and, in a certain sense,
“alive”. That is why they are a good tool to improve the ability to read, understand and translate Latin.

The use of local materials, which can be checked out personally, may more easily arouse the learners’ attention. That is
why in this paper I will mainly refer to inscriptions from modern Umbria (regiones VI and VII, from Perusia) and
especially from Asisium, used at school during my experimental lessons. The local museum, best known as the “Foro
romano”, hosts a huge collection of inscriptions that were catalogued in 2008 by pupils of the Liceo Properzio, under
my supervision, with details of the type of inscription, material, place of origin and chronology. Previously a booklet
about Roman Assisi had been
written\footnote{http://www.liceoassisi.it/index.php?option=com\_content\&view=article\&id=47:assisi-romana\&catid=6:i-nostri-risultati\&Itemid=34.},
both in Italian and English, comprising a page about the famous tetrastyle with a translation of the inscription on the
base (CIL 11, 05372 = EDR025323, later inserted in the MediaWiki page), which recalls the official inauguration of the
aedicule: at the time, as was the Roman custom, money was given to decurions, seviri Augustales and the common people.
Thanks to these few lines, the class is taken on a journey to ancient Asisium, and gets a lot of feedback about
religion, architecture, social life and economy in the first century A.D.

1.2. How to use inscriptions

Inscriptions in high school can be used in a variety of ways, but particularly in the following fields:

\begin{itemize}
\item elements of Latin grammar, syntax and linguistics\footnote{ Hartnett 2012; McCarthy 1992.};
\item culture and history;
\item and also, to a lesser extent, lexicon. 
\end{itemize}

\bigskip

1.2.1. Lexicon in context

Although the acquisition of lexicon is essential in order to access the conceptual categories of a culture, it is
extremely difficult to learn words without using them actively, and it is a terrible mistake, often made by school
textbooks, to provide long lists of terms, especially if de-contextualized and only based on the frequency with which
they are used.

Many manuals, influenced by modern languages, offer lexicon in concrete fields (such as food, clothing or education),
but this does not necessarily entail that the students will be more attracted by these topics. There is also the risk
of reducing culture to anecdotes. Experience shows that with a limited amount of time and taking into account the
selective memory of teenagers, you have to choose what your priority is. The main reason why a school-level student
should do Latin is because of the deep impact of Roman cultural and linguistic heritage on our world: therefore, in my
opinion, the words that students need to learn the most are:

\begin{itemize}
\item the ones that are relevant for their Italian derivates (e.g. os, oris);
\item the ones that belong to the most significant semantic fields (always keeping in mind that some words are more
important than others: for instance, knowing the difference between bellum and pugna helps to develop the comprehension
of the two different categories; on the contrary, a non-specialized student does not need to know who a primus pilus
was);
\item the abstract ones that are fundamental to understanding the Roman way of thinking (e.g. imperium, virtus, fas).
\end{itemize}
Accordingly, a teacher should concentrate on those aspects of Roman society that are still pre-eminent for us:
archaeology (Roman buildings, domus, theatres, roads), myths and religion, politics (Empire, war and globalization).
Inscriptions certainly offer a less varied lexical repertoire than a literary text, but can still help to give
substance to these contexts.

In an inscription such as CIL 11, 05400 = EDR025350:

P. Decimius P. l. Eros / Merula, medicus / clinicus, chirurgus, / ocularius, VIvir. / Hic pro libertate dedit
((sestertium)) (quinquaginta milia). / Hic pro seviratu in rem p(ublicam) / dedit ((sestertium)) (duo milia). / Hic in
statuas ponendas in / aedem Herculis dedit ((sestertium)) (triginta milia). / Hic in vias sternendas in / publicum
dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni ((sestertium))...

not only does a student make contact with the technical lexicon of medicine and words related to building activities
(statuas ponere, vias sternere), but he can also perceive the “evergetic spirit” of an ancient society: a physician was
usually a freedman coming from the East, who could be very rich and spend his money for public utility; whenever a
person obtained a priesthood, he used to pay a summa honoraria. The text can also be used for linguistic purposes
(prepositions like pro; partitive genitive; deponent verbs).

This does not mean that inscriptions are not useful for seeing specific abstract words in their context: when talking
about Roman virtues and mos maiorum, the famous Clipeus from Arles (\emph{AE} 1952, 0165 = \emph{AE} 1994, 0227) is a
perfect complement to chapter 34 of the Res Gestae (formally an inscription, too): \emph{Senatus / populusque Romanus /
imp(eratori) Caesari / Divi f(ilio) / Augusto / co(n)s(uli) VIII dedit clupeum / virtutis, clementiae, / iustitiae,
pietatis erga / deos patriamque.}

\bigskip

1.2.2. Grammar

One of the biggest problems in teaching Latin to young students is that they find it really difficult to understand
declensions. In the preliminary lessons the sentences used are very simple and can be understood and even translated
without really acquiring the syntactic function of the words.

For instance, if I say Iulia rosas amat, even an Italian who knows no Latin can get the meaning of the sentence.
Problems are encountered when the sentences and texts to translate become more complex: only when it is too late does
an ill-prepared teacher realize that the class has not internalized the language system and the patterns.

In funerary inscriptions the value of cases is essential. The texts can be simple, but if you want to understand who is
dead, who is the dedicant and what the relationship between them is, you have to distinguish dative and nominative
cases. Moreover, the concordance of the different elements of the onomastics (especially praenomina and filiation) is
good way to encourage (and, for a teacher, to test) the learning of declensions not in a merely mnemonic, but also in
an active way.

Examples:

\textwide{\FilledBigTriangleRight} CIL 11, 05501 = EDR025449: Noniae Privatae, C. Propertius -{}-{}-{}-{}-{}-

You can use an inscription as easy as this one to revise first and second declensions, and at the same time to teach the
structure of a Roman name.

\textwide{\FilledBigTriangleRight} CIL 11, 05461 = EDR025411: A(n)noru(m) XIX. Calventia C(ai) f(ilia) Polla, L(ucius)
Vistinius vir, Gavia mater posuer(unt).

This inscription offers the chance to get feedback about mistakes in the use of language, or to reflect on mortality and
marriage.

\textwide{\FilledBigTriangleRight} CIL 11, 05399 = EDR025349: P(ublius) Decimius P(ubli) l(ibertus) Eros Merula VIvir
viam a cisterna ad domum L(uci) Muti stravit ea pecunia -{}-{}-{}-{}-{}-

With an inscription like this you can address different topics, both historical (the importance of freedmen, the
imperial cult) and linguistic ones (indirect complements, uses of is ea id as an adjective and to introduce a relative
clause that, in this case, has evidently been lost).

\bigskip

At the beginner’s level, the sentences used for examples, exercises and translations are fictitious and often banal.
They give a false image of antiquity: for instance, when you start with the first declension you get the wrong
impression of a “female” world simply because the male nouns are rare. Students are unimpressed; these sentences have
no significance for them. Those provided by inscriptions are equally easy and short, but they are not banal because
they are a mirror of a society, they offer a historical perspective; there is always a story behind them. Even the
simplest ones can give us precious information.

Let us consider a famous inscription from the Cathedral in Assisi (CIL 11, 05390 = EDR025340):

Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(- - -),
Ner(o) Babrius T(iti) f(ilius), C(aius) Capidas T(iti) f(ilius) C(ai) n(epos), V(ibius) Voisienus T(iti) f(ilius)
marones murum ab fornice ad circum et fornicem cisternamq(ue) d(e) s(enatus) s(ententia) faciundum coiravere.

The comprehension of the text is easy, especially with the abbreviations solved. Even so, many considerations can be
made:

\begin{itemize}
\item for grammar: words of third declensions such as maro and fornix; complements of direction; gerundive to express
purpose; use of the form {}-ere in the perfect tense personal endings;
\item for history: the use of the Latin language prior to the Social War, as proof of the intense Romanization of the
area at the end of the II century B.C.\footnote{ Coarelli 1991.}; the presence of elements in the names that are not
Roman but of Umbrian origin;
\item for archaeology: the building of terraces to create public spaces in a town like Assisi established on a hill; the
identification of the area around San Rufino as the “acropolis”; the incorporation of the Roman wall into the left nave
of the church (the inscription being still in situ).
\end{itemize}

\bigskip

Schoolbooks rarely present these opportunities offered by inscriptions, so the main problem for a teacher is to access
appropriate material\footnote{ See the observations made by Carpenter 2006.}. Holding a PhD in Late Antiquity and
Umbrian inscriptions, I am fortunate enough to already know many sources. However, a graduate may not have enough
knowledge in epigraphy, may not be familiar with the Corpus Inscriptionum Latinarum; that is why the EAGLE project can
also offer tools for further insights.

Here are some examples taken from manuals where you can see the different attitude of authors towards the epigraphic
material:

\textwide{\FilledBigTriangleRight} Barbieri 2015, 39: the inscription of the architect C. Vettius Gratus (CIL X 3392) is
merely a decorative element on the page, without any relation to the topic (phonetic changes from Latin to Italian);

\textwide{\FilledBigTriangleRight} Domenici 2012, 40: programmata from Pompeii are introduced to explain the Roman
naming system;

\textwide{\FilledBigTriangleRight} Gambis Meini, Roffi, Guidotti Bacci 2013, 269-270: tabellae defixionum are used for
different purposes (demonstrative adjectives, functions of subjunctive); the subject (magic in the ancient world) may
not be relevant in a school context (the possession of such knowledge is not required), but sounds very intriguing to
students.

\bigskip

\bigskip

\bigskip

1.3. Translations

Translating inscriptions is a difficult task for everyone and especially for high school students.

First of all, they will not find any help elsewhere: every other Latin text can easily be found – even if not always
correctly translated – on student internet sites and in blogs (such as Splash Latino). But we all know very well that
the Internet still lacks many Italian translations of inscriptions.

Secondly, the style and the structures are different. Even when all abbreviations and integrations are explained (a
teacher should at least show the meaning of round and square brackets but should not ask a pupil to solve an
abbreviation, except the easy ones), the word order cannot immediately be reconstructed, especially in decrees and
carmina epigraphica. School dictionaries are not intended for interpreting epigraphical lexicon: some words may not be
present, their meaning may not always be explained. A lot of institutions and formulae which are clear to a specialist
(e.g. quattuorvir iure dicundo) may be hard to understand or to translate. It is, however, also the case that a teacher
cannot dedicate too much time to introducing these words to the class because his ultimate aim is different; so it is
better to focus on materials that do not contain much specific lexicon.

Examples:

\textwide{\FilledBigTriangleRight} CIL 11, 04431 = EDR025160: [Inf]austo, levis umbra, tuo mihi flebilis hora / sorte
tua certe tempus in omne fuit.

This funerary inscription from Ameria contains two verses, but the position of the words is tricky; infausto may be
taken as an adjective (as it is most commonly) and its meaning is basically the same as sorte tua; students who are not
used to poetry may not recognize the anastrophe tempus in omne.

\textwide{\FilledBigTriangleRight} CIL 11, 04391 = EDR025123: Iuliae M(arci) f(iliae) Felicitati, / uxori C(ai) Curiati
Eutychetis / IIIIvir(i), magistrae Fortu/nae Mel(ioris), coll(egium) centonarior(um) / ob merita eius. Quo honore /
contenta sumptum omnem / remisit et ob dedic(ationem) ded(it) sin/gulis ((sestertios)) XX n(ummos) et hoc amplius /
arkae eorum intul(it) ((sestertium)) V m(ilia) n(ummum) / ut die natalis sui (ante diem) V Id(us) Mai(as) / ex usuris
eius summae epu/lantes imperpetuum divider(ent), / quod si divisio die s(upra) s(cripta) celebrata non / fuerit tunc
pertineb(it) omn(is) summa / ad familiam publicam.

Specialists are well used to an inscription like this. A high school student may find unexpected difficulties in
understanding the meaning of magistra (not teacher but priestess), but also centonarius or even hoc amplius; not to
mention the Roman calendar system, which always takes too much time to explain!

Nevertheless, this kind of challenge is exactly what makes inscriptions the perfect tool to fully appreciate what the
art of translation is.

The article published by F. Bigi in the Proceedings of the First EAGLE International Conference\footnote{ Bigi 2014.}
includes many observations regarding the problems that may be encountered, for instance when translating names and
titles. In particular, I find the suggestion that round brackets should be used in the translations to provide further
explanations about specific offices rendered with the technical derivative word, or for concepts omitted in the
original Latin text, very useful. Examples of this could be words such as centonarii or marones seen in inscriptions
mentioned above, but also the following ones:

\textwide{\FilledBigTriangleRight} CIL 11, 04213 = EDR130908: Interamna Nahars should be further qualified as “Terni” to
help those readers who are not familiar with Umbrian cities;

\textwide{\FilledBigTriangleRight} CIL 11, 01925 = EDR142701: the names of the emperors M. Aurelius Antoninus and M.
Antoninus Pius Germaticus Sarmaticus need to be explained (Caracalla and Marcus Aurelius) to avoid confusion. It should
be noted that in my classes proper names, and not only those of emperors but also of other people, were generally
translated into Italian, even if it is advisable to transcribe them in the nominative case\footnote{
http://www.eagle-network.eu/wiki/index.php/Guidelines\_for\_Translators.}; the same thing was done with cognomina ex
virtute, considering that they are intuitively interpretable for an Italian.

\bigskip

This type of activity works better with Liceo Classico (Grammar School) pupils, who do translations from Latin and Greek
almost every day and who are more at ease with the use of dictionaries. However, there are still hurdles to overcome.
The ministerial syllabus set out for the course focuses on Literature and culminates in a specific exam requirement,
the translation of a piece of literary prose. Is this “epigraphical” activity helpful? Does it take up too much of the
time which should be employed in translating the Classics? The answer to both these questions is “yes”. On the one
hand, as I said, students have to “jump into translating” without a net (the Net, in fact). On the other hand, if at
the end of the final year pupils are required to translate a passage from certain authors, then clearly it would be
more appropriate for them to concentrate on this activity as much as possible during the months prior to the exam. For
this reason, epigraphy can only be a supplement to traditional assignments; the Italian national curricula are
apparently very free, but at the same time they are very rigid. Yet, a few forays into epigraphy can be stimulating,
because the class perceives them as an intriguing novelty, especially if not subject to assessment. After all, it would
probably be too difficult to prepare a test with grades and scores on this subject and could deprive this activity of
its extemporaneous and enjoyable aspect.

1.4. Inscriptions and Literature: a few samples

On a more advanced level, inscriptions may also integrate certain aspects related to the study of Latin Literature. The
most typical example could be a comparison between the Tabula Claudiana (CIL 13, 01668) and Tacitus’ account (Annales
XI, 23-24)\footnote{ On which see Jahn 1993.}: reading the original document is a privileged occasion to determine how
reliable the historian is when using his sources.

Highly original suggestions have been provided by Mauro Reali, who is also the author of different school manuals, in a
paper published online\footnote{http://mediaclassica.loescher.it/nuove-e-\%93vecchie\%94-forme-di-multimedialita.n2799
; see also Reali, Turazza 2015.}. Being an expert on the subject, he offers a comparison between the “noble” form of
the political-philosophical amicitia presented in Cicero’s Laelius and the term \emph{amicus} mentioned in inscriptions
from the lower levels of society, such as CIL 05, 05300 from \emph{Comum} (a funerary stela made by a freedman for
Pliny the Younger) or CIL 05, 05923 = EDR124245 regarding a strange case of “friend deletion” long before the Facebook
era\footnote{ http://www.laricerca.loescher.it/lingue-classiche/327-un-amico-o-amicus-e-per-sempre.html.}.

An interesting example from Umbria can be found in AE 1992, 0560-0561 = EDR150769 and EDR150784. The first gravestone
recalls the acquisition of a tomb - which had previously been despoiled - by an heir of the founder, who then installed
another cippus for 40 friends (amicis meis, i.e. freedmen probably belonging to the same association):

Viator, resiste et rogo / te et lege. Post annos XXVII ven[i] / Hispellum, in patriam meam. Scio / me oportere colere
hunc locum / ubi ossa meorum requiescunt et mea / et amicorum meorum. Ex hoc sepulch[ro] / cippi perierunt duo et
frontes duae sciun[t] / qui surupuit et acturi simus et legimus, / satis est testium etqs.

As for the Augustan age, Reali suggests showing some monumental inscriptions of the princeps (he impressively goes so
far as to compare the qualifications \emph{Imperator Caesar Augustus} to a modern logo or even a \emph{hashtag}). On
this subject, the altars Augusto sacrum put up by Perusia restituta (CIL 11, 01923 = EDR142666, EDR142667, EDR142668,
EDR142669), even with a simple text, offer the opportunity to deal with an aspect as crucial as the imperial cult.
Traditionally, we read that Augustus was worshipped directly only in the Eastern provinces, but not in Rome and Italy.
The inscriptions from Perusia testify that things are effectively more complex; these documents also offer remarkable
information about the restitutio of the town, destroyed at the end of the bellum in 40 B.C. A few years later Perusia
would become Augusta, as you can read on the city gates, especially on the newly restored Etruscan Arch (CIL 11, 01929
= EDR142706).

Teaching, as I do, in a school named after Propertius, I always stress the importance of reconstructing the origin of
the poet through the epigraphical data of the gens Propertia: the greatest number of written documents of the family
having been found in Assisi. For more than two hundred years, beginning with the Vois(ienus) Ner. (filius) Propertius
mentioned among the Umbrian magistrates of the late II century B.C., this gens stands out in the town for its social
influence and wealth\footnote{ Forni 1986; Zuddas 2006.}. It is always very exciting to combine the information on
Passennus Paullus Propertius Blaesus given by CIL 11, 05405 = EDR025355 and that contained in two Letters by Pliny (VI,
15 and IX, 22). Pliny, showing great concern for his friend’s illness but also great esteem for him as an elegiac poet,
asserts that he is a descendant and fellow citizen of Propertius; the inscription (the front of an honorific base),
providing the full name, with the tribe Sergia typical of the inhabitants of Asisium, is indirect, but clear, evidence
that the Augustan poet was born there. The information provided by the inscription and the literary text integrates
perfectly. Students find it fascinating to look for traces of the poet inside the town, especially when they read the
following graffito on an interior wall of a Roman house underneath the church of Santa Maria Maggiore (EDR028769):

[{}- - - I]o[323?]vino consulibb(us) (ante diem) VIII Kal(endas) Martias domum oscilavi Musae.

This house was still visited in the fourth century A.D., which really supports the theory that it used to be the poet’s
residence, and remained an object of reverence for centuries\footnote{ Boldrighini 2014, 244-246.}. Through the stories
of Passennus Paullus and Sextus Propertius macro-history and local history meet to make the past come alive in every
corner of the modern town.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/18/EAGLE2016_Full_Paper_Broux.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Yanne Broux}
\date{2015-09-16}
\begin{document}
[Warning: Draw object ignored]Abstract

Facebooking the past. The idea grew while developing a database of all people mentioned in texts from Greco-Roman Egypt
(350 BC–AD 800). Thanks to Trismegistos’ role in EAGLE, Named Entity Recognition can now be applied to almost 500,000
Latin inscriptions from the Roman Empire, and some 400,000 clusters containing personal names can be extracted. This
collection of names will lead to a large-scale study of naming practices in the ancient world, and how these reflect
changes in society at large.

\bigskip

Keywords

Latin epigraphy, Trismegistos, Named Entity Recognition, Social Network Analysis, Onomastics, Prosopography

\bigskip

1.1. Trismegistos: the early years

Facebooking the past. The idea grew a couple of years ago, while developing a database of all people mentioned in texts
from Greco-Roman Egypt. While probably not exactly considered Big Data by those who actually work with BIG data, the
500,000 or so attestations of individuals in Trismegistos open up some prospects for quantitative analysis, something
historians still tend to shy away from. One of the approaches I have been exploring is Social Network Analysis [SNA].
SNA was developed in the 1960s in mathematics, anthropology and sociology and measures structural forms of relations
between individuals, places and/or events. Over the past couple of decades, it has found its way to numerous other
fields, such as physics, neuroscience, and recently also (modern) history. Within ancient history, however, SNA still
needs to obtain a firm footing.

To reconstruct proper social networks, a decent prosopography is indispensable. In a traditional scholarly setting, this
implies time-consuming and painstaking manual labor. Fortunately, in a digital environment, there are other ways, and
Trismegistos (www.trismegistos.org) forms an ideal starting point. Trismegistos grew out of a long tradition of
databases and prosopographies (structured lists of people), as well as various other Ancient History projects at KU
Leuven. The original idea of Trismegistos was to foster interdisciplinarity in the study of Ancient Egyptian society by
creating a central database with metadata about published papyrological texts from Greco-Roman Egypt, in a first
instance written in Greek, Latin and Egyptian (including hieroglyphic, hieratic and demotic). The inclusion of Egyptian
soon dissolved the disciplinary boundary with epigraphy, broadened the chronological window, which was eventually set
to 800 BC – AD 800, and led to the inclusion of further languages such as Coptic, Aramaic and Arabic.

Apart from texts, since 2008 Trismegistos also intensively deals with places and people. Building on open access to the
full text in repositories, Named Entity Recognition procedures are used to create lists of toponyms and anthroponyms
that occur in the ancient sources. This was first applied during the socio-onomastic project ‘Creating Identities in
Graeco-Roman Egypt’ (KU Leuven OT project 2008-2012), on a corpus of about 50,000 papyri and ostraca in the Duke
Databank of Documentary Papyri (papyri.info). With the additional support of a Hercules Grant, (‘An Interdisciplinary
Database of Proper Names in Late Pharaonic, Graeco-Roman and Byzantine Egypt (ca. 800 BC – AD 640)’; 2008-2014) the
work on the core data could be finished in just over two years, resulting in a database with almost half a million
references to people (Trismegistos People) and an additional hundred-thousand or so place names (Trismegistos Places)
mentioned in texts from Egypt.

On the basis of Trismegistos People, several studies on naming practices, modes of identification and identity issues in
Greco-Roman Egypt have been published. Two PhDs focused on the longstanding tradition of double names: the first dealt
with the Ptolemaic period, when this form of polyonymy served to cross ethnic borders [Coussement forthcoming], the
other with the Roman period, when the practice was adapted by the local elite to distinguish themselves from the hoi
polloi and to resemble Roman nomenclature [Broux 2015a]. The data of the fourth century AD provided new insights into
the spread of Christianity in Egypt [Depauw and Clarysse 2013]. The development of a standardized identification
cluster, consisting of a person’s name, patronymic and metronymic [Broux and Depauw 2015a], as well as the use of fixed
expressions to denote illegitimacy [Broux 2015b], name change [Broux 2013a] and official identification [Broux,
Coussement and Depauw 2010], were all related to the legal reorganization of the population under the Romans and the
ensuing tax reform [Broux 2013b]. Most recent studies focus on the influence of Roman socio-linguistic practices on
Greek and Egyptian conventions [Depauw forthcoming] and how network analysis can provide us with new insights regarding
onomastic habits and what they say about cultural identity and social status [Broux 2015c].

1.2. Going global: an encompassing source guide for the ancient world

1.2.1. Expanding Trismegistos Texts

Like Facebook, however, Trismegistos wants to grow, and get the entire ancient world on board. To achieve this goal,
Trismegistos’ core database, the text database, must first be expanded by broadening its chronological and geographic
horizon. Since 2013 the team has been actively working toward the inclusion of all texts from antiquity in
Trismegistos. This implies including Latin and Greek inscriptions, an estimated 700,000 texts. Contacts with the Latin
epigraphic database in Heidelberg resulted in Trismegistos’ participation in the Europeana EAGLE project, coordinating
the disambiguation across all partners. This added about 150,000 new texts to the Trismegistos database. The remaining
300,000 will be integrated from another source (Clauss-Slaby: www.manfredclauss.de), so that the coverage of Latin will
soon be exhaustive. For Greek, good contacts have been established with the main players in the field (PHI:
epigraphy.packhum.org/inscriptions/; DC3 [i.e. Duke, SEG, Claros]: blogs.library.duke.edu/dcthree/), and the aim is to
become partners in a project dealing with the 250,000 or so inscriptions during the next year. At the same time
Trismegistos is also working towards a complete coverage of indigenous languages and scripts, which are often separate
fields, isolated from the ‘classical’ world. New partnerships for Etruscan have been set up, and soon also for Punic
(CIP: cip.cchs.csic.es) and South-Arabian (DASI: dasi.humnet.unipi.it), while for the Italic languages, for Gallic,
Lepontic, Venetic, and Messapian information has already been digitized on the basis of existing corpora.

\bigskip

1.2.2. Expanding Trismegistos People

With the experience obtained during the extraction of names from the Greek papyri, Named Entity Recognition will be
applied to the Latin inscriptions incorporated in Trismegistos so far. This will result in the addition of an estimated
500,000 or more new references to names/individuals to Trismegistos People. Network analysis will be used to help with
the disambiguation of individuals.

\bigskip

1.2.2.1. Named Entity Recognition

The collection of references to people and their names in the inscriptions will be carried out by applying Named Entity
Recognition [NER] to the clusters of capitalized words extracted from the full text repositories of Greek and Latin
inscriptions.

NER was originally developed by computational linguists in the 1990s to detect and classify pre-defined elements in
texts, but quickly spread to other fields, such as biology and genetics, and is now gaining momentum in the Digital
Humanities [van Hooland et al. forthcoming]. The problem with NER-systems, however, is that techniques designed for one
genre or field do not necessarily work for others, due to specific text properties (some follow strict writing
constraints, e.g. scientific or news articles, while others, such as email or tweets, are more informal), or due to
language-related grammatical and syntactical formats.

For ancient Greek and Latin, the languages of the inscriptions under scrutiny here, there were no ready-to-use NER
solutions. In a case language, proper names are more variable; the different accents on vowels make letters less easily
recognizable; and in particular the onomastic system and the way people are identified is completely different from
modern languages. Therefore a strategy needed to be developed to cope with the multilingualism of the sources and the
declensions of the inflected languages.

Trismegistos opted for a combination of a gazetteer and a rule-based approach. For the gazetteer a three-tiered
onomastic database structure was developed in Trismegistos People, dealing with names, name variants, and declined name
variants respectively. The first database, NAM, currently has 34,106 entries, e.g. the name Isidoros. Each name is
connected to a set of variants in all possible languages. As a rule, only very minor dialectal or orthographical
variation is allowed in the ‘native’ language, in this case Greek (e.g. \textgreek{E>is'idwroc }and
\textgreek{>Is'idoroc). }Many of the variants are actually created by renderings of a name in other languages, e.g.
[1EC8?]sytrs, [21D?]ysydwrs or [2C93?][2CA5?][2C93?][2C87?][2CB1?][2CA3?][2C89?] in Egyptian and Coptic. In all there
are 177,290 variants in the NAMVAR database. Finally, for each of the variants, the declined forms were created, e.g.
\textgreek{>Is'idwrou (}genitive) or \textgreek{>Is'idwrwi (}dative). This NAMVARCASE database is the largest with
667,677 entries, and this set is used as a gazetteer for NER. It was developed on the basis of the set of names of some
40,000 individuals with titles listed in the Prosopographia Ptolemaica (KU Leuven) and was supplemented with new names
collected during the ‘Creating Identities’ project [Depauw and Van Beek 2009]. Through cooperation with the Lexicon of
Greek Personal Names (Oxford University) the names of some 300,000 individuals can now also be integrated.

Apart from this gazetteer, rules were developed to cope with the combination of names, or more correctly declined name
variants, as well as with the combination of names with non-onomastic introductory terms (e.g. ‘son of’) in the
identification of individuals. For Greek texts, this is pretty straightforward, as people were identified by their name
(generally a single name, in some cases a double name) followed by genealogical identifiers (in the genitive form)
only. The onomastic habits encountered in Latin inscriptions differ significantly, however. Here the majority of the
individuals follow the Roman naming system. This consists of a fossilized praenomen, a nomen gentilicium (the
equivalent of our family name), and a personal cognomen. Often the patronymic was inserted between the gentilicium and
cognomen; sometimes even the papponymic and the voting tribe, e.g. Marcus Tullius Marci filius Marci nepos Cornelia
Cicero (‘Marcus Tullius Cicero, son of Marcus, grandson of Marcus, [of the] Cornelia [tribe]’). A completely new set of
rules is therefore being developed to apply to the Latin inscriptions.

\bigskip

1.2.2.2. Human quality control and prosopographical identifications

Once the clusters of capitalized words have been extracted and have been matched to the onomastic gazetteer and rules
for name combinations, a human check will be performed. This includes tasks which are not easily automated:
interpreting declined name variants as attestations of a specific case where the mere form is ambiguous; deciding
whether some ambiguous entries are toponyms or anthroponyms; and reviewing the results of the cluster interpretation
rules and adding relevant information where necessary.

All this could be labeled ‘quality control’, but we will also rely exclusively on humans for the logical next step when
developing a prosopography: the identification of namesakes as attestations of the same person. Since the systematic
review will be performed text per text, only intratextual identifications can be implemented at this stage.

\bigskip

1.2.2.3. Social Network Analysis

Social Network Analysis [SNA] was developed in the 1960s in mathematics, anthropology and sociology and measures
structural forms of relations between individuals. It has huge potential for future historical research, not only by
applying ‘traditional’ network analysis for the study of social interaction, but also by developing new,
non-conventional techniques.

In the wake of the automatic extraction of individuals from Greek papyri during the ‘Creating Identities’ project,
people appearing in more than one text could not be identified and were therefore entered in Trismegistos People under
multiple records. The identification of these doubles is a difficult and time-consuming process, not in the least
because of the high degree of homonymy: in village communities, similar names were common, and in families, names were
often passed down every other generation. This makes it difficult to distinguish between one person and another. Broux
and Vanbeselaere therefore developed a new procedure to identify individuals with the help of graph visualization and
SNA [Broux and Vanbeselaere forthcoming]. By linking people based on co-occurrence in the same texts, an overview of
all the data is presented in a single network, where one can “zoom in” on specific individuals and compare their
“surroundings”. When specific clusters of names reappear, these are likely the same individuals mentioned in different
texts (see Fig. 1: each circle [or ‘node’] represents an individual; some of these need to be merged, since they
actually refer to the same person).

\bigskip

 \includegraphics[width=9.984cm,height=9.454cm]{EAGLE2016FullPaperBroux-img001.jpg}

Fig. 1.1. Visualizing individuals co-occurring in texts.

\bigskip

Additionally, the developed identification method enables us to discern family components. Mapping genealogical
relationships is often problematical, especially when individuals are attested in different capacities. Someone who is
mentioned as the father of an athlete in a victory list is not easily recognized as the state official in a petition.
In other words, social and professional links do not always overlap with family ties, and these need to be synced to
provide a complete picture and to prepare the data for the next step: analyzing networks of names.

The application of SNA to onomastics is another avenue that has not yet been explored. By linking names on the basis of
genealogical relations (since names are the result of conscious choices made by parents), the co-occurrence of names in
communities can be mapped, which opens up new possibilities for quantification and interpretation [Broux 2015c]. A
name’s popularity can be calculated by means of its in-degree (i.e. how many other names point to it), while the
density of the network, the number of reciprocal links and the weight of these links can tell us something about the
social motives behind these naming patterns. For the local elite of Roman Egypt, for example, descent was of prime
importance, since membership was strictly hereditary, and by limiting themselves to a specific collection of names,
they could express family and community ties. Moreover, networks like these can help us evaluate the perception of
names in antiquity, as well as determine the linguistic origins of undefined names, on the basis of their location in
the graph.

1.3. Goals

1.3.1. Study of naming practices

The collection of names from across the entire Mediterranean will lead to a large-scale study of naming practices in the
ancient world, and how these reflect changes in society at large. A major transition point is of course the steady
integration of regions and states across the Mediterranean and Western Europe into the Roman Empire. The focal point
will therefore be the impact of the Roman occupation on traditional naming and identification conventions in different
provinces. Regions where pre-Roman material is also available (e.g. Gaul, Magna Graecia, Asia Minor) are especially
significant when mapping aspects of continuity and change chronologically. Moreover, results from both eastern and
western provinces will be compared to study uniformization, whether imposed from above or spread out from below.

\bigskip

1.3.2. Towards a Facebook of the ancient world

Eventually, the goal of Trismegistos is to recreate a prosopography of the Greco-Roman world. Reconstructing social
networks of the past will help us gain a better understanding of the mechanisms of interaction in the ancient
Mediterranean, not only on the micro level (individuals), but also on the mesa (communities) and even macro (regions,
empires) levels. At the same time connections and communication across these different levels can be analyzed: how
individuals, as members of local communities, were integrated into larger political structures (top-down approach), and
how these communities responded to impositions from above (bottom-up approach). Social models, such as the six degrees
of separation theory, can be tested, to check whether our ‘small world’ perception is indeed the result of present-day
technology and mass-communication, or if similar structures of interconnectivity existed, and, if so, what the
conditions for this ancient globalization were back then. Mark Zuckerberg out. Enter Trismegistos.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/20/EAGLE2016_BONACINI - Pilot project at Paolo Orsi Museum.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%

\title{}
\author{}
\date{2015-12-11}
\begin{document}
Digital enhancement of the “Paolo Orsi” museum: a Google Street View 360° pilot project tour \newline

Elisa Bonacini

Humanities Department, University of Catania; IEMEST (Istituto Euro Mediterraneo di Scienze e Tecnologie), Palermo,
Italy

Email: e\_bonacini@hotmail.com; elisa.bonacini@unict.it

\newline
Abstract

\ \ \ \ \ \ The aim of this paper is to present the pilot project in progress at the “Paolo Orsi” Archaeological Museum
(Syracuse, Italy). Thanks to a free partnership with Google Business Photos, we have managed to map the entire museum
for a online 360° tour on Google Street View. A dozen archaeological finds have been selected for 360° virtual tours,
provided with descriptive sheets. Among them the beautiful ‘inscription of Nassiane’, from the Catacombs of San
Giovanni in Syracuse, has been selected.

\bigskip

Keywords

Google, Sicily, Sicilian Cultural Heritage, Virtual Museum, Virtual tour, Digital Heritage.

\bigskip

\bigskip

1.1. The “Paolo Orsi” Archaeological Museum

The “Paolo Orsi” Regional Archaeological Museum of Syracuse, together with the “Antonino Salinas” Regional
Archaeological Museum in Palermo, is the most important Sicilian archaeological museum and it is one of the most
important and richest archaeological museums in Italy.

The National Archaeological Museum of Syracuse was born by a royal decree in 1878, known as “national” for its
collections’ importance and size. Well-placed inside the historical palace in the Cathedral Square on Ortigia island,
it was directed by Paolo Orsi from 1895 to 1934.

The archaeological collection has been enlarged by over 70 years of archaeological research making it necessary to move
the collection to a new museum space. Designed by the architect Franco Minissi, the new museum was built in the Villa
Landolina garden between 1967 and 1986 and inaugurated in January 1988. The collection consists of artefacts from the
prehistoric, Greek, Roman and Christian periods found in archaeological excavations in Syracuse and in other Sicilian
sites. The museum space is divided in three levels (floor 1, 2 and basement), distributed around a central space which
is dedicated to the history of the museum and temporary exhibitions. First level is divided in three sectors (A – C)
and testifies the history of central-eastern Sicily from prehistoric ages to the Greek one. On the upper floor, sectors
D and F were inaugurated in 2006 and contain finds from the Hellenistic-Roman and Christian periods. Section E will
open next year with findings from sites in central-eastern Sicily (as Centuripe, Morgantina, Tindari and so on).
Moreover, a precious and unique collection of coins and medals from archaic to the medieval age is located in the
basement, opened in 2010.

\bigskip

1.2. The project: reasons and birth

This project born from the desire to fill the deep gap in the promotion and enhancement of Sicilian cultural heritage.

Sicily has the highest number of UNESCO heritage sites (7/51 in total\footnote{1997: Valley of Temples in Agrigento;
1997: Villa del Casale; 2000: Eolian Islands; 2002: Late Baroque Towns of the Val di Noto (South-Eastern Sicily); 2005:
Syracuse and the rock necropolisof Pantalica; 2013: Mount Etna; 2015: Arab-Norman sites, Palermo and the Cathedral
Churches of Cefalù and Monreale.}) and of UNESCO intangible cultural heritage (3/6\footnote{2008: Opera dei Pupi,
Sicilian puppet theatre; 2013: Mediterranean diet (transnational); 2014: Traditional agricultural practice of
cultivating the ‘vite ad alberello’ (head-trained bush vines) of the community of Pantelleria.}) in Italy and in the
world. Infact, Sicily has some of the largest and most important archaeological sites in the world: the temple of
Concordia in Agrigento, for its exceptional state of preservation has become the symbol of UNESCO itself.

Despite of this, Sicilian cultural heritage struggles to be present on Google’s platforms such as Google Art Project and
others, as it should, compared to other Italian cultural sites (Bonacini 2013; Bonacini 2014).

It is not so difficult to explain where the difference between the island and its mother country comes from. Sicily has
the status of independent region, therefore it has an exclusive competence in the field of regional cultural heritage.
Sicilian heritage is released from any convention that the Italian Ministry of Cultural Heritage and Tourism has signed
since 2009 with Google. The Regional Department of Culture and Sicilian Identity has never bothered to solve this
really huge gap regarding its cultural heritage and landscapes.

In Street View Gallery\footnote{https://www.google.com/maps/views/streetview?gl=us.}, which now has contributed a great
number of users from all over the world, thousands are the spherical and geo located photos of Italian places. However,
tightening the selection to “Landmarks of Italy”
\footnote{https://www.google.com/maps/views/streetview/italy-highlights?gl=us.}, Sicily has only 9 spherical photos,
showing the beaches of the Aeolian Islands (7), the islands of Favignana (1) and Marettimo (1).

Among Art Project’s 605 museum collections\footnote{https://www.google.com/maps/views/streetview/art-project?gl=it.}, 47
are Italian, especially relevant to Rome, Turin, Venice and Milan. One of these, not the best artistic production, is a
Sicilian contemporary collection, relevant to the International Festival of Street Artists in Giardini Naxos (Me). No
other Sicilian museum, collection or archaeological site has been included in Art Project.

Browsing on the Street View Gallery, 21 sites are all over inscribed on the World Wonders
Project\footnote{https://www.google.com/maps/views/streetview/world-wonders-project?gl=it.}: for Italy, only Pompeii
and the historic center of Florence are inscribed in this still restricted list. Many more sites can be visited
virtually from the same project linked on Google Cultural Institute’s
website\footnote{www.google.com/intl/it/culturalinstitute/worldwonders/. }: 172 sites in the world, 22 in Italy and,
finally, 2 of them in Sicily. Until last year the Unesco site of the baroque town Val di Noto was the only one in
Sicily; currently Mount Etna, has been added.

Google organized its Google Camp 2014 and 2015 editions by selecting as exceptional locations two of the most evocative
archaeological sites in the world, Selinunte and Agrigento. Nevertheless the beauty and heritage of Sicily, ironically,
is not on Google’s platforms.

This project was born in collaboration with Mr. Gianfranco Guccione, a certified Google Business Photo photographer,
while he was working as a freelance consultant at the General Direction of the Regional Department for Cultural
Heritage and Sicilian Identity in 2014. He proposed to realize the Street View mapping of a museum and an
archaeological site in Sicily, considering the possibility of creating “augmented” virtual tours (3D virtual tours of
objects displayed in museum’s windows and virtual aerial tours, with the addition of text and audio descriptions).

The profound reason of this project consists in an effort to bridge this gap, “increasing” fruition and enhancement of
Sicilian cultural heritage.

With the agreement between the General Direction of the Regional Department for Cultural Heritage and Sicilian Identity
and the European Coordination of Google Business Photo, it was decided to choose as a sample of this project two
cultural regional institutions, the “Paolo Orsi” Museum in Syracuse and the Valley of the Temples in Agrigento, both
UNESCO sites since 2005 and 1997. The project was then structured as a research fellow project at the University of
Catania and carried out by the present writer in close collaboration with Mr. Guccione.

The first part of the project at the “Paolo Orsi” Museum, which we will discuss here, is about to be completed. The
second part at Valley of the Temples in Agrigento is going to start.

\bigskip

1.2.1. The 360° tour on Google Maps: some technical data

A large photographic survey began in August 2014, with the aim of mapping all areas of the first and second level open
to visitors (only the collection of coins and medals, for security reason, has not yet been photographed).

A total of 3.924 shots to get about 327 360° virtual tours have been made using a mobile station made up by a reflex
camera with fisheye type camera lens, tripods with panoramic head. Because of the peculiarity of the light in the
different museum’s sectors, it was necessary to adjust the brightness each time. The windows in the winding path often
reflect one another and precautions were taken to avoid, as far as possible, those refractions.

Once loaded on Google’s software Business Photos, the pack of images from the 327 virtual tours were geo located in
Google Maps Street View, mounted avoiding defects of sight between the images.

The 360° virtual tour makes it possible browsing the entire museum and its collection, between levels connected by
arrows, placed online at the link https://goo.gl/maps/oagnd8urP1H2 (for the first level) and
https://goo.gl/maps/vrpDfuPPgwM2 (for the second level). As you can see in fig. 1.1., by clicking on the first link,
the remote user can enter the museum through the Street view panel: the 360° virtual tour begins at the entrance and
the user can browse moving with the directional arrows, between sectors and levels.

The project provided the opportunity to carry out 360° virtual tours of some exposed archaeological finds, like an
“augmented” virtual tour, certainly innovative compared to what Google’s platforms provide.

Art Project, for example, allows you to view points of interest along the path in a museum, but captions are short and
photos are static. A pilot initiative has been recently launched featuring the possibility to in-browse almost 300 3D
photos of objects from the collections of six cultural institutions in the world. However 242 of these consist of scans
of animal skulls (pieces of nature, not works of art) from the California Academy of Science collection; 22 objects of
art come from Museo d’Arte Orientale (the only Italian cultural institution to join the project until now). These 3D
objects, described by short captions, are available to users to be rotated and zoomed. The aim of this new Google’s
project is to begin building the most important and largest database of 3D scan art of works worldwide.

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{m{10.872cm}}
{\centering  \includegraphics[width=8.911cm,height=4.632cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img001.jpg}
\par}

~

Fig. 1.1. The virtual entrance to Archaeological Museum “Paolo Orsi” on Google Street View.

~
\\
\end{supertabular}
\end{flushleft}
As for the displaying on Google Maps of the 360° virtual tour of the objects, we must specify that Maps so far does not
support the integration of menus, captions, photos, video, info inside the Street View virtual tour technology.
Infact,it \ is only allowed navigation in 360°, ie a virtual walk.

We tried to find an answer in order to show on Maps all the information, captions, maps, levels and 360° virtual tour of
objects. Thanks to customized i-frame we made these virtual tours possible with their captions and the maps with
different levels and clickable points of interest, adding them to the existing virtual tour of the Museum, already on
Maps, through a link containing the Google mapping of the Museum. In this way the virtual tour of the Museum - made
with Google standards - is “augmented” by another virtual walk much more exhaustive, located via link on the Google
Maps board of the Museum, where you can view all these additional items.

Both the virtual tour of the museum and the virtual tour of the objects will also be placed on the Museum's website, on
the Regional Department for Cultural Heritage and Sicilian Identity portal
(http://www.regione.sicilia.it/beniculturali/museopaoloorsi/). Because of the profound principle that this non-profit
project would improve the visibility and promotion of Sicilia Cultural Heritage, the Museum’s virtual tour and its
reproductions belong to Google; Museum’s objects virtual tours, instead, belong to Paolo Orsi Museum, which is free to
reuse them.

\bigskip

1.2.2. The “Paolo Orsi” Museum 360° tour

Coming from Sector A, after turning around a couple of casts of dwarf elephants from Spinagallo cave (Syracuse), a
remote user can see the displayed artifacts, starting with Neolithic phase of Stentinello (VI millennium B.C.) to reach
the great exhibition space dedicated to Bronze age: Ancient Bronze age (facies of Castelluccio), Middle Bronze age
(facies of Thapsos), Late Bronze age (facies of Pantalica) and Final Bronze age (facies of Finocchito), where \ the
large containers from Thapsos and Pantalica necropolis.

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{m{10.681cm}}
{\centering  \includegraphics[width=8.98cm,height=4.697cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img002.jpg}
\par}

~

Fig. 1.2. 360° tour of Sector A: a couple of casts of dwarf elephants from Spinagallo cave. \\
\end{supertabular}
\end{flushleft}

\bigskip

In the Sector B1 remote users can admire findings from the first colonies founded by the Greeks in eastern Sicily
(Naxos, Zancle, Leontinoi, Katane, Megara Hyblaea) with some of the most important Greek masterpieces, as the naked
sculptures of young men from Leontinoi and Megara Hyblaea.

The B2 Sector introduces the visitor to the archaeological finds from the city of Syracuse, from its foundation to
classical age. Here the most important spaces are those dedicated to the architecture of archaic and classical temples
(Athenaion, Ionic temple, Olympeion and Apollonion), to the statuary and the terracotta findings from urban excavations
during the last decades (Piazza Duomo, Ortigia, Piazza della Vittoria) and to the urban and ex-traurban necropolis
(Fusco, Giardino Spagna) with rich funerary kits. \

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{m{10.681cm}}
{\centering  \includegraphics[width=8.89cm,height=4.641cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img003.jpg}
\par}

~

Fig. 1.3. 360° tour of Sector B2: the findings from Athenaion and Ionic temple in Syracuse. \\
\end{supertabular}
\end{flushleft}

\bigskip

Sector C is dedicated to the colonies founded by Syracuse - Eloro (670 B.C.), Akrai (664 B.C.), Kasmenai (644 B.C.) and
Camarina (598 B.C.) -, to Gela (689 B.C.) and Agrigento (580 B.C.), the largest colonies of south-eastern Sicily with
their ceramics, architectural remains of temples, findings from sanctuaries and necropolis, as well as finds from other
indigenous hellenized centres.

Sector D on the second level contains finds from the Hellenistic age to the Roman period, including statuary, beautiful
portraits from the Roman age, architectural pieces, ceramics, mosaics, cinerary urns and various handcrafts.

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{m{10.681cm}}
{\centering  \includegraphics[width=8.708cm,height=4.521cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img004.jpg}
\par}

~

Fig. 1.4. 360° tour of Sector D: the Venus Landolina. \\
\end{supertabular}
\end{flushleft}
They document the multiple aspects of life in Syracuse and come from urban necropolis from III-II century B.C. Here are
masterpieces like the wonderful statue of the Venus Anadiomene, called Venus Landolina, here in Fig. 1.4.

In Sector F finds from the various catacombs in the city are on show, documenting life in the Christian era. Here is the
Sarcophagus of Adelphia, a Christian marble sarcophagus found in the Rotunda of Adelphia inside the Catacombs of San
Giovanni, just near the museum.

\bigskip

1.2.1.1. 360° virtual tours of archaeological selected objects

The Museum staff has selected 12 objects along the path between the two levels, which can be linked as points of
interest placed on the Museum map - where the remote user can identify sectors, plans and POI - with virtual 360° tours
equipped with labels with captions and descriptive texts.

Here are the selected objects:

[Warning: Draw object ignored]

Fixed on tripods the reflex camera with a quadrangolar lens, the objects have been photographed flipping them on a
graduated portable rolling disk. Each 360° object virtual tour took a number of about 88 shots, for a total of 1.062
shots, mounted with specific software. In this way the remote user, clicking on the preview pictures, can admire the
selected object in all its sides, by moving the mouse on the right and left and zoom in-out.

\bigskip

1.2.1.3. 360° virtual tours of \ Nassiane inscription

The last selected object we present here is a disk from Sector F: it is a marble disk, discovered by Paolo Orsi in the
Catacombs of San Giovanni in 1894 (Orsi 1895: 509-510), decorated by a wreath of laurel leaves on one side; the other
side was reused in the 4th century A.D. for the funerary inscription of Nassiane, woman who died at age 32 in God
faith.

This object, which the remote user can observe with a 360° virtual tour, has been selected both for the peculiar reuse
and because it documents the religious syncretism of the early centuries of Christianity. Re-using of architectural is
well documented in the catacombs of Syracuse (Sgarlata 2013). It was found shattered in front of a burial, distinct
from others, known as “the Saint's Tomb” (Sgarlata 2004: 40-44).

\bigskip

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{m{10.681cm}}
{\centering  \includegraphics[width=10.029cm,height=5.225cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img005.jpg}
\par}

~

Fig. 1.5. 360° tour of the Nassiane inscription. \\
\end{supertabular}
\end{flushleft}
\ This is the Greek text:

\bigskip

chrismon \textgreek{c qrist{\textless}i{\textgreater}an~hc semn~hc >aganófronoc [>hd]`e fil'androu
}N\textgreek{assian~hc t'umbon e>isor~a|c, f'ile, k'imenon [<`wde <'h]tic semnos'unh|sin [>erize]to Phnelop'ih| }palma
chrismon \textgreek{>enj'ade k~ite [}N\textgreek{ass]ian'h, z'hsasa [>en J(e)~w|] kal~wc ka`i >am'emptwc >`eth
lb}\textgreek{', }\textgreek{m}\textgreek{~hnac i'.}

\bigskip

The inscription says: “Oh friend, you see here the tomb of Nassiane, Christian, ripe, sweet, fond of her husband, who
competed for virtue with Penelope. Here lies Nassiane, she lived well and blamelessly in God (?), 32 years and 10
months”. This inscription documents such as epigraphic formulas affected by religious syncretism (syncretism generally
means that a complex of phenomena and concepts derive from the meeting and fusion of different religious forms) between
Pagans and Christians, for which a Christian like Nassiane could have competed in life with Penelope, the hero
Odysseus’ wife and symbol of marital devotion. Its circular shape suggested that it was a table for the ritual of
refrigerium, the funeral feast that symbolically was consumed with the dead, even this syncretic practice \ dates back
to the pagan world (Sgarlata 2013, Scandurra 2014).

\bigskip

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{m{10.681cm}}
{\centering  \includegraphics[width=6.205cm,height=6.174cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img006.jpg}
\par}

~

Fig. 1.6. The Nassiane inscription, recto (courtesy by the “Paolo Orsi” Museum). \\
\end{supertabular}
\end{flushleft}

\bigskip

1.3. Conclusions

As discussed elsewhere (Bonacini 2013 and 2014), Google is undoubtedly the most active entity in the world committed to
preservation, dissemination and promotion of cultural heritage, well above any public institution, through an
unparalleled campaign of digitization open to users’ collaboration. This could happen because Google itself has an
incomparable capacity of economic investment. Even large international projects of digitization (such as Europeana
itself) \ are not able to compete with Google.

Therefore, after initial hesitation towards these Google’s initiatives, now most of museums and cultural institutes in
the world have seen in Google a partner that enables them to progress in the online visibility and in the process of
heritage digitization.

Regarding the project here presented, which concerns one of the two selected sites, we can rightfully say that it is the
first archaeological museum in the world - needless to say, the first museum in Sicily - entirely browsable on Google
Maps platforms with a virtual tour in all exhibition halls and 360° virtual tours with integration of captions and full
description of artworks.

In the near future we hope to allow 360° visualization of a greater number of objects, with their accompanying captions
translated at least in English and in audio version.

Thanks to this project we hope that Google itself could realize how the time has come to “rejuvenate” the Google Maps
Street View system, allowing enabled users to apply additional content on the maps.

However, the wide interoperability between Google software the development of new solutions and the integration of
geo-referenced results in the page results on the search engine continues unabated: the “Paolo Orsi” Museum - and with
the museum, the city of Syracuse and the whole of Sicily – will surely take advantage of this new tool for its
visibility.

\bigskip

[Warning: Draw object ignored]

\bigskip

Acknowledgement

To Mr. Gianfranco Guccione (http://www.airworks.it/) goes my heartfelt and deep thanks for having proposed to me this
project and having concluded it together, with the only aim of enhancing the Cultural Heritage in Sicily.

I would like to thank some people, because without them it would not have been easy to achieve this result: Professor
Mariarita Sgarlata, formerly Regional Minister of Cultural Heritage and Sicilian Identity; Dr. Sergio Gelardi, formerly
General Director of Cultural Heritage, Dr. Gaetano Pennino, current General Director of Cultural Heritage; Dr. Enrico
Carapezza, Area Manager for General Affairs of the Department of Cultural Heritage and Dr. Maria Pia Bottino, formerly
member of the cabinet of Regional Minister of Cultural Heritage and Sicilian Identity.

I warmly thank all staff of the “Paolo Orsi” Regional Archaeological Museum in Syracuse for the great availability and
friendship shown to me and Mr. Guccione: Dr. Gioconda Lamagna, executive head of the Museum, Dr. Angela Maria Manenti,
head of relations with the public, all the archaeologists Dr. Anita Crispino, Dr. Agostina Musumeci and especially Dr.
Giuseppina Monterosso, tireless in supporting the project at any time and formerly member of the cabinet of Regional
Minister of Cultural Heritage and Sicilian Identity.

Many thanks to Dr. Peter Shishkov, Google Business Photos/Street View Indoor European Coordinator, for supporting Mr.
Guccione’s initiative.

To the University of Catania, to the Chancellor Prof. Giacomo Pignataro and to the Humanities Department Director Prof.
Giancarlo Magnano San Lio deserve my thanks for allowing me to work on this project.

My deepest gratitude goes to my friend Prof. Annalisa Bonica for the patience she \ has showed in correcting this
article.

This publication has been made on concession of the Regional Minister of Cultural Heritage and Sicilian Identity;
archaeological findings and plans presented here are exclusive property of the “Paolo Orsi” Regional Archaeological
Museum in Syracuse and it not allowed to be copied or duplicated \ by any means.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/21/EAGLE2016_Full_Paper_revised.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{}
\date{}
\begin{document}
AXON. A Selection of Greek Historical Inscriptions

A Database for Research and Teaching

\bigskip

Abstract

The AXON Project has developed a database of Greek historical inscriptions, from the birth of the~polis~in the Archaic
Age to 31 BC. Each entry is provided with the object’s description, a complete lemma, Greek text with critical
apparatus, Italian translation and commentary with keywords and indexes, and updated bibliography. New insights for
data-inclusion have been developed. The database supports enlargement and offers a high degree of searchability. Our
aim is to illustrate the structure, the contents and the solutions we have come up with in the development of the AXON
Project. We will also offer some suggestions for teaching and academic research purposes.

Keywords

Online epigraphic editions; interoperability of digital editions of Greek historical inscriptions; images of Greek
historical inscriptions; digital epigraphy in teaching and research

\bigskip

1. The AXON Project

AXON. A Selection of Greek Historical Inscriptions is a project conceived within the Greek Epigraphy Laboratory
(Director, Prof. Claudia Antonetti), and has been brought into existence with the financial support of the University
Ca’ Foscari of Venice (University Project 2013, Project Coordinator, Prof. Stefania De Vido).

Since October 2014 the members of the AXON Project have been developing a database which includes a great variety of
Greek inscriptions of different chronology, typology, and territory of origin. The most recent advances of traditional
epigraphy as well as the scientific acquisitions in the Digital Humanities have been taken into account. The selection
of texts has been made according to a broader notion of ‘historical’ inscription, including not only significant
military, political, and institutional texts, but also those inscriptions which are essential for the social and
cultural understanding of the Greek world.

AXON includes texts from the birth of the Greek polis in the Archaic Age\footnote{ See the Introduction in Hansen \&
Nielsen 2004; Hansen 2006. } to 31 BC, a chronological frame traditionally related to Greek History (though a future
extension of this chronological limit is not excluded). The epigraphic entries have been prearranged in order to allow
a wide and well-structured description of each document. At the same time, a common and coherent lexicon has been
produced, which will permit an easier indexing of significant words and will make future searches much quicker.

\bigskip

2. A unique model-entry for a great diversity of inscriptions: taxonomy and categorisation

\bigskip

2.1. Entry description

\bigskip

2.1.1. Object’s description

The model-entry has been created with an eye on the object’s thorough description. Here is the object’s categorisation:

\begin{enumerate}
\item Object type
\item Material
\item Object’s dimensions
\item State of preservation
\item Further descriptive elements
\item Date and context of finding
\item Finding site (modern nation, ancient region, ancient and modern name of the city, if known)
\item Actual location (modern nation, city, museum/archaelogical context, inventory number)
\end{enumerate}
The great majority of these categories can be selected from a given number of options from a pull-down menu. Some
categories – such as Object type, Material, or State of preservation – are directly linked with the corresponding
sections in the EAGLE Vocabularies (see http://www.eagle-network.eu/resources/vocabularies/). Furthermore, a hyperlink
has been created between the AXON-entries and Pleiades website (http://pleiades.stoa.org): where the finding site is
known, each entry offers the geographic coordinates and a Googlemaps visualisation. This gives the possibility of
rapidly gathering the information for any single ancient location and allows for searches directly from an interactive
map.

\bigskip

2.1.2. Chronology

The chronological delimitation of each text is supported by many options, as you can see in fig. 1:

\begin{figure}
\centering
\includegraphics[width=10.88cm,height=4.156cm]{EAGLE2016FullPaperrevised-img001.png}
\end{figure}
Fig. 1. Window for the input of data, section Text/Chronology

\bigskip

2.1.3. Alphabet \& language

\bigskip

Each entry provides all the necessary information on the alphabet and language of each inscription:

\begin{enumerate}
\item Type of Inscription (with link to EAGLE’s Vocabularies:
http://www.eagle-network.eu/resources/vocabularies/typeins/. The categories are those used by Guarducci 1967-1978)
\item Text’s structure
\item Writing (Execution technique\footnote{ With link to http://www.eagle-network.eu/resources/vocabularies/writing/.};
different types of epichoric alphabets according to Kirchhoff’s colour-coded map; Local script\footnote{ Following the
categorisation in Jeffery 1990.}; Palaeographic features and letters’ form\footnote{ The letter-form and glyph-form are
based on the symbols of the font Cardo (http://scholarsfonts.net/cardofnt.html), but many have been developed by the
AXON Team on the examples of letter-form given in Jeffrey 1990 (see also
http://poinikastas.csad.ox.ac.uk/browseGlyphs.shtml).}; letters’ heights, description and layout of the text field;
Direction of Text)
\item Language (with an option for any dialect’s peculiarities)
\end{enumerate}

\bigskip

\bigskip

2.1.4. Genetic lemma \& apparatus criticus

The text of each inscription is preceded by a hierarchically arranged lemma (the so called genetic lemma, according to
Louis Robert’s definition\footnote{ Robert \& Robert 1954.}) and is followed by the apparatus criticus.

\bigskip

2.1.5. Italian translation \& commentary

Each entry corresponds to an Italian translation and commentary (in .pdf).

\bigskip

2.1.6. Abstract

The Abstract – with a WYSIWYG interface – includes all the keywords for indexing and lemmatisation:

\begin{figure}
\centering
\includegraphics[width=10.88cm,height=5.449cm]{EAGLE2016FullPaperrevised-img002.png}
\end{figure}
Fig. 2. Example of Abstract in AXON. The different colours allow a selection of words from a drop-down menu.

\bigskip

The Keywords are divided into the following categories (these categories are based on EpiDoc Community Guidelines as
well as on the indexes of the Supplementum Epigraphicum Graecum, SEG):

\begin{enumerate}
\item Persons and names (mainly for ‘historical’ characters)
\item Gods and heroes
\item Place names
\item Geographical names
\item Significant words regarding the history of politics and institutions
\item Other relevant keywords
\item Ancient sources
\end{enumerate}

\bigskip

2.1.7. Bibliography

Finally, an updated bibliography highlights any previous edition for each entry, as well as all the appropriate
secondary sources. SEG abbreviations have been used for epigraphic corpora and other publications (the section
“materiali” on the website gives access to a list of all abbreviations, useful for students, as well\footnote{
http://virgo.unive.it/venicepigraphy/axon/public/axon/pagine/materiali (still being processed).}).

\bigskip

2.2. Internal \& external interoperability of the AXON database

\bigskip

Each entry is related – whenever it seems appropriate – to other entries in the AXON database. A hyperlink connects the
entry with other digital editions of the same text (if available), or with other useful websites, possibly containing
images. Wherever possible, images and/or apographs and/or squeezes of inscriptions have been included. The creation of
a digital archive of images as part of the AXON website is also desirable.

\bigskip

2.3. A simple website interface for the input of data

\bigskip

Since the contributors to the project (i.e. the authors of the entries) are experts from different Italian and European
universities (and not all of them are familiar with the Digital Humanities), and given the great number of entries
planned in the near future, the necessity of a simple and easily understandable interface for the input of data was an
essential issue to the project from the very beginning. Guidelines to EpiDoc have been taken into account in order to
produce a clear structure for the input of data.

Our aim is to establish a growing community of experts, students, and enthusiasts to increase the number of contributors
through lists of inscriptions which have not yet been assigned. At the same time it will be the possible to suggest
other texts which are not included in the lists. To achieve these aims, the project follows an EpiDoc-friendly
structure and is compatible with Europeana EAGLE Project, especially in the use of a common terminology.

\bigskip

\bigskip

3. Searchability

\bigskip

The website is designed to allow for many search options. Beyond the “full text” search and another based on the number,
title and author of the entry (see fig. 3), three other search-possibilities will also be available:

\begin{enumerate}
\item browse all the entries according to the inscriptions’ a) typology, b) chronology, and c) area of origin;
\item access the entries through an interactive map;
\end{enumerate}

\begin{figure}
\centering
\includegraphics[width=10.88cm,height=7.354cm]{EAGLE2016FullPaperrevised-img003.png}
\end{figure}
Fig. 3. Search based on “Full text”, number, title, and author of the entry.

\bigskip

\begin{enumerate}
\item perform an advanced search based on different categories: 
\end{enumerate}
\begin{enumerate}
\item \begin{enumerate}
\item bibliography
\item keywords
\item object’s description and preservation (see fig. 4)
\item chronology
\item text (single words or phrases, typology, dialect, alphabet, letter-form, etc.) (see fig. 5).
\end{enumerate}
\end{enumerate}

\bigskip

\bigskip

\begin{figure}
\centering
\includegraphics[width=10.88cm,height=9.518cm]{EAGLE2016FullPaperrevised-img004.png}
\end{figure}
Fig. 4. Different searching categories.

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

Fig. 5. Advanced text search.

\bigskip

\begin{figure}
\centering
\includegraphics[width=10.88cm,height=9.888cm]{EAGLE2016FullPaperrevised-img005.png}
\end{figure}
Other filters will be employed for each search result. A section entitled “tools” will also be available, and will
include information on the entries’ structure, tables with the contents of different categories, links to Vocabularies
and websites, etc.

4. The AXON Project for teaching and academic research

\bigskip

4.1. Teaching

\bigskip

The AXON Project, as an example of a digital edition of inscriptions (see esp. genetic lemma and apparatus) with a high
degree of clarity for contributors and users, is a useful tool for teaching Greek epigraphy as well as ancient history.
Many contributors are university lecturers / professors of Greek Epigraphy, and the scientific committee includes high
school teachers and instructors in classical languages, making AXON especially well-suited for educational purposes and
for use by students: for engaging them, for example, in the composition of entries. The interoperability of the AXON
website and the cross-references to other Digital Humanities projects are essential elements in the development of this
discipline.

\bigskip

\bigskip

4.2. Significance for the academic community

\bigskip

Each entry is created by an expert contributor and is subject to double-blind peer review, thus assuring an important
contribution to the scholarly community. At the same time, the hyperlinks to other websites and digital editions will
make it easier for the user to check immediately all similar projects. Finally, the indexing allows for the easy
discovery and use of specific information, and will be of fundamental importance for gather together groups of
documents according to particular research needs.

In conclusion, the AXON Project aims at a collaboration of expert scholars from different fields: epigraphy, ancient
history, dialectology, archaeology, digital humanities. It can produce valuable results in the domain of the digital
editing of inscriptions and, more generally, contribute to the advancement of classical studies, opening them up to a
broader audience through the world-wide web.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%\title{}\author{MARTA}
\date{2015-09-18}
\begin{document}
The Digital Edition of the Archaic Latin Inscriptions

(7th{}-5th century B.C.)

Giovanna Rocca – Giulia Sarullo – Marta Muscariello\footnote{\par \ \ Giovanna Rocca (GR) is responsible for § 1 and 2;
Giulia Sarullo (GS) is responsible for § 4 and Marta Muscariello (MM) for § 3.}

Abstract

This project consists in the digital edition of the archaic Latin inscriptions (7th – 5th century B.C.) according to the
EpiDoc Guidelines. The edition is the result of an autoptical examination of the epigraphic documents and of the
text-bearing objects, together with the analysis of previous studies. In the particular case of the Forum inscription,
this led to new discoveries and confirmed old hypotheses. Each text will be presented in an epigraphic chart, enriched
by photos and illustrations.

\bigskip

Keywords

Archaic Latin Inscriptions, Latin Epigraphy, EpiDoc, Digital Humanities, Epigraphic Edition, Forum inscription.

\bigskip

\bigskip

1. Generalia [GR]

The project Iscrizioni Latine Arcaiche (ILA) consists in the digital edition of the inscriptions found in old
Latium\footnote{\par \ \  With the two exceptions of the Vendia’s Urn, found in Cerveteri, and the Garigliano bowl.}
dating back to the period between the 7th and the 5th century B.C. Between the end of the 19th century and the first
decades of the 20th century this corpus consisted of only four inscriptions of a certain length, that is the Duenos
vase (1880), the Fibula Praenestina (1887), the Forum inscription (1899) and the Tibur pedestal inscription (1926),
besides other shorter but still very interesting texts that offered important cultural information (such as the
inscription from the Regia, REX, CIL I2 479). In the second half of the 20th century, the corpus grew significantly and
reached the total of about eighty documents that have not been gathered in a comprehensive edition yet. The updated
corpus includes new and crucial discoveries, some of which have already been published in a traditional way (in print),
such as the inscriptions from Satricum and the fragments found at the Palatinum, whereas others are still unpublished,
as in the case of the recent findings from the Regia. The quantitative and qualitative enrichment provided by recent
new readings of the texts and the new data emerged justify and require a new publication.

Most of the inscriptions in the collection are frustuli or single letters: these texts, though not really relevant from
the linguistic point of view, are important testimony of the use of writing in Latium since the 7th century B.C.

The website is then an absolute novelty in the field of digital epigraphy, since at the moment no online epigraphic
collection specifically dedicated to these documents exists. It is well known that, for every kind of publication, a
digital edition provides several advantages, for example: the possibility to update both the textual corpus and the
bibliography continuously; the hypertextual structure, that allows the user to utilize the edition in different ways;
the opening to a public that is heterogeneous and wider than the one a work addressed to specialists can reach. In
which other ways can the web respond to the specific requirements of a peculiar corpus such as ours with distinguishing
features, different from the later Latin epigraphy optimally reproduced in the Epigraphic Database Rome (EDR)?

First of all, by using the EpiDoc encoding standard, which is compatible with other encoding systems, it will be
possible to transfer our data to EDR and to the EAGLE-Europeana Network (see § 4).

The hypertextual structure of the edition is surely one of its assets, in that it enables the user to find information
about each text (images, bibliographical references, etc.) immediately and consider the inscription in the context of
its place of finding; at the same time, it encourages a direct [201B?]in real time’ comparison between texts. Moreover,
this kind of structure allows us to obviate the inconvenience of an edition that presents only a specifically
[201B?]historical’, [201B?]archaeological’ or [201B?]epigraphic’ point of view, as sometimes happens in \ traditional
editions. This is achieved thanks to cross-references to topics analyzed in depth by specialists, thus making our
edition a useful research tool for various branches of learning.

Besides what has been illustrated so far, the presence of complex indexes, through which it will be possible to locate
the inscription from different starting points, will allow the user to search the texts according to various
parameters: dating, place of finding, object type, and textual type (see § 3).

The choice of the method to adopt as regards the interpretative transcription is problematic: it is clear that it is
absolutely impossible to be completely neutral or objective. Unlike the diplomatic transcription, for which a critical
apparatus can be constructed and be extremely useful, the same is not possible for the interpretative transcription: an
exhaustive critical apparatus would imply a superabundance of information that could negatively influence the
scientific nature and the usability of the edition, especially in the case of inscriptions that have been variously
interpreted since their discovery. For longer inscriptions (e.g. Forum Inscription, Duenos vase, Tivoli inscription,
Garigliano bowl), the numerous readings that have been proposed so far by scholars have been compared and verified in
the light of linguistic criteria and of the new data in order to obtain an edition of the text that, although it cannot
be considered the definitive interpretation and does not solve all the pending issues, poses itself as a new starting
point for future research. A cross-reference to all the other interpretations will offer a complete source of
information and a tool that intends to be useful and exhaustive.

The archaic Latin inscriptions play a fundamental role in the study of the first stages of the language, since they
present particular features that allow us to investigate the various steps that led to “standard Latin”. The language
attested in our inscriptions can be considered a Restsprache insofar as it is not [201B?]readable’ through later Latin
but it can be [201B?]interpreted’. The linguistic commentary will be carried out, in votis, in the second phase of this
project. Here the research focuses on the epigraphic features, that show a plurality of forms and of alphabets in such
a limited corpus. The chronological and geographical distribution of the signs and of the variants in use in the
inscriptions have been analyzed in order to offer valid elements to the study of the evolution of the alphabetic model
between the 7th and the 5th century B.C.

2. Epigraphic News about the Forum Inscription [GR]

Thanks to the agreement and collaboration with public authorities, we were invited to take part in an extraordinary
event (July 3rd 2015), that is the 3D laser scanning of the Forum Cippus (CIL I2, 1)\footnote{Cippus of tuff found
under the Lapis Niger by Giacomo Boni in 1899. The inscription is cut boustrophedon on the four faces of the pillar and
on the edge sliced between face D and face A. The stone (late 6th B.C.) is badly damaged in the upper part so that the
beginnings and the endings of the lines are lost. The text has been interpreted as a sacred law.}. The autopsy, carried
out with the help of a strong source of light, and the observation of the scanning in real time clarified, hopefully in
a definite way, some controversial issues about the presence of dividing signs on face A and, at the same time, opened
a new perspective on the reading of line 16 (face E).

One of the epigraphic problems concerning this inscription consisted in the absence of punctuation on face A, in
comparison with the other faces in which three vertical dots divide the syntactical units. This lack was particularly
suspicious in a sequence in which there are no exegetical alternatives (SAKROSESED = sakros esed). As a consequence,
scholars tried to find an explanation for this absence: the inscription was carved by different hands; the inscription
was made up of different texts, each copied from different drafts; the antigraph was in scriptio continua and the
inscriber was not familiar with this procedure; the punctuation was not accurately assigned.

As a matter of fact, face A conforms to the others, showing three vertical dots after sakros and also, although less
deep and more worn-out, after esed\footnote{The details of significant portions of the inscriptions will be shown on
the website with several photographs. } (Fig. 1.1). This fact was highlighted by Gamurrini (1899) and the three dots
appear in the apograph in Hülsen 1899, col. 1003 and they could also be guessed in the photograph Anderson 3192
(Archivio Alinari). Nevertheless, the post Goidanich (1943) vulgata and the difficulties in checking in person the
stone, because of its almost unreachable and scarcely illuminated position that led to the publication of studies not
supported by an autoptical check, perpetuated a reading influenced by the uncertainty in distinguishing between natural
cavities or notches due to the nature of the stone and [201B?]significant’ holes caused by the tool used for writing.

Fig. 1.1. Forum Cippus, Face A and Face E (by Marta Muscariello).

\begin{figure}
\centering
\includegraphics[width=4.837cm,height=7.225cm]{IscrizioniLatineArcaicheEAGLE2016-img001.png}
\end{figure}

\bigskip

The most relevant news arrives from the analysis of line 16 (face E). This has been read up to now as loiuquod()qo
(Wachter 1987), loi\{u\}quiod (Vine 1993), LOIUQUIOD[QO] (Baldi 1999), LOIUQUIOD,QO/// (Hartmann 2005)\footnote{The
final QO is based only on Goidanich 1943.}, in order to explain the [201B?]unusual’ shape of V (letter no. 4) that has
been recently read as F (Prosdocimi 2010). As the scanning showed, the first vertical stroke, read as an I, is much
closer to the sign that looks like a V than it appears on the apographs and the photographs published so far;
especially shots taken from an oblique and not a frontal point of view can be misleading.

This point can result fundamental in understanding how the inscriber worked on the stone and how he corrected the
sign\footnote{Other corrections can be found on the last line of face C (kapia on kapa) and on the second line of face
D where a V was corrected into a koppa. }. Hypothesis no. 1 (which is also the simplest one): the sequence to write was
LOIVQVIOD but, after he cut a first vertical stroke, the inscriber mistakenly started to cut the oblique stroke of a V
instead of a second vertical stroke that would begin the V; he recognized the error and cut a vertical stroke next to
the first one and finally another oblique stroke that reached the bottom of the second vertical stroke, completing the
V. Hypothesis no. 2 presumes the same order in cutting the strokes, but for a different reason: the sequence to write
was LOVQVIOD, the inscriber started cutting the first stroke but he found an obstacle (i.e. a hole in the stone), so he
continued with the oblique stroke on the right (thinner than the others) but he changed his mind and cut a second
vertical stroke close to the first one and joined it with a new oblique stroke. The short inner stroke has been
considered as a correction, i.e. a deletion, and caused the expunction of the whole letter or its reading as an F; as a
matter of fact, it is nothing else but the result of a first try to cut the oblique stroke. Of course, we could be more
precise after we receive the outcomes of the scanning, that will be ready soon. The following sign (nr. 4 or 5,
depending on the reading), instead, is surely a koppa and not an O.

Our reading proposal has the advantage of illustrating the sequence of the inscribing but does not solve the
interpretative problems: a louquiod instead of a loiuquiod / loiquiod (for which lucus, eloquium, licium and liquidus
have been proposed) still needs to be explained and both still await a solution.

3. The Epigraphic Chart [MM]

In the website, each text of the ancient Latin corpus is presented by an ad hoc designed chart in order to meet the
peculiar requirements of this kind of corpus. The chart is organized in items that concern every important aspect
regarding the material and cultural contextualization of the inscription, from the archaeological support to the
epigraphic features. Such a detailed scheme contributes to making our project a research tool both as a complete source
for information retrieval and as an updated starting point for the study of the texts and of the language.

In this initial phase of \ Latin literacy, the linguistic data are not sufficient to establish the linguistic features
of this language – the corpus chiefly consists of hapax. As a consequence, the contextual data of the inscribed object
results being of great help to comprehend the text. For example, the new archaeological data found during the recent
excavation campaign in the Comitium carried on under the direction of P. Fortini are providing helpful information for
the study of the Forum Inscription; in the past, the collection of all the fragmentary instrumental Latin inscriptions
up to the 4th century B.C. published by G. Colonna in 1980\footnote{In Stibbe et al. 1980.} supplied, at least
partially, the extent of the alphabetization developing in Latium Vetus, subtracting the major inscriptions from a sort
of “documental isolation”; moreover, the data concerning the interaction with other inscriptions of ancient Italy are
fundamental, since these are different in languages and alphabets: besides the Etruscan examples in Rome, we can
remember the case of Satricum, where both Latin and Volscan are attested\footnote{See Rocca 1997, 189-198.}, or the
Garigliano bowl which bears, together with the Latin inscription engraved inside, an inscription in Italic alphabet and
language on the external body of the vase\footnote{On the relationship between the two inscriptions, see Antonini
2012.}, that has also given a hint for a particular interpretation of the Latin text.

The first item of the chart contains the ID tag assigned to the inscription in the ILA project, that identifies the text
with the find-spot (using the ancient place name whenever possible) followed by a progressive number: for example, the
Tibur pedestal inscription (CIL I2 2658) is denominated “Tibur 1”, the inscription on the Garigliano bowl is called
“Garigliano 1”. Given the small number of find-spots, we decided to use the full name of the place, instead of
abbreviations; possible new findings that could emerge in the future from the same site will be easily added to the
corpus simply by increasing the number. Beside the ID tag attributed to the inscription we quote the most common names
attributed to the object, stratified in time in literature and well known to the scholars (for example “Duenos vase” /
“Vase of the Quirinal”). We then indicate here, in the item [201B?]Present collocation’, where the inscription is
preserved with, when possible, the inventory number.

The chart continues with a group of three items that form the section dedicated to the [201B?]Archaeological data’: the
‘Description of the object’ (with the general type, the possible peculiar features, the function and the dimensions of
the inscribed object); the ‘Provenance’ (that is the place where the object was found and its exact archaeological
context\footnote{In the case of mobile objects a different place of fabrication can be presumed, as in the case of the
Vendia’s Urn, found in Cerveteri, but considered by some to have been fabricated somewhere else in Latium.}); the
‘Date’, which quotes the hypotheses given by scholars about the chronological coordinates of the inscription.
Concerning this item, we must keep in mind that the dating of the antiquissimae is often approximate, and it is based
on different criteria (at times archaeological, at others epigraphic or linguistic or with convergences of two or of
all these factors); in some cases, the gap is so wide to almost seem fluctuant depending on which criterion is
considered. Without doubt this long-standing problem must be held in consideration, also remembering that for some
inscriptions a chronological lowering to the 4th or even the 3rd century B.C. has been proposed. The difficulty in
dating the objects and the rarity of the findings is surely connected to the difficulty in defining the specific
features of the language testified by these inscriptions\footnote{On the periodization of Latin, P. Cuzzolin and G.
Haverling state: “The division of the history of a language into different periods implies that we have a rather clear
picture of what language we have dealing with. At two points in the history of Latin we are not quite sure of this: the
exact moments in which Latin is born and in which it is transformed into Romance are not easily determined. The problem
is to determine what is Latin and what is not: unfortunately there is no overall agreement on whether all of the early
inscriptions considered to provide us with early examples of Latin actually do that.” (Cuzzolin, \ \ Haverling 2009,
20).}.

The charts present several photographs of the inscriptions, taken during the photographic campaign carried out by the
project’s team. Enlargements of some useful or problematic details of the inscription have been added: the richness in
images is related to the participation of the project in the Europeana network, a database of images of the European
cultural heritage. Although the images illustrate the inscription in an optimal way, fac-similes of each text will also
be provided, clearly related with the transcription provided by the editor.

We offer two kinds of transcription, the ‘Diplomatic transcription’ and the ‘Interpretative transcription’, which
provides the edition of the text. The need for a diplomatic transcription is due to the problematic nature of archaic
texts: in the case of the Duenos vase, as it is well known, the second section of the text is almost always given by
scholars in diplomatic transcription because of the difficulties in segmenting the phrases into words (but with
attempts of interpretations of small portions).

In the item [201B?]Textual typology’ inscriptions are classified according to the nature of the text, taking into due
consideration the peculiarities of this corpus. In the chronological span between the 7th and the 5th century B.C. the
codification of formularies both of possession and of gift/dedication is still in fieri in the various linguistic
branches spoken in ancient Italy (with the exception of Greek); for this reason, from a classificatory point of view an
inscription can be ‘anomalous’ in two ways: on the one hand, it can lack the typical elements of a formulaic scheme
that will be fixed later on, thus requiring a further interpretative effort in order to assign it to a specific textual
category; or, on the other hand, it can result more complex than the standard formula and present elements that can be
related to more than one textual typology: in this case, the object type and the archaeological context are determinant
for the overall classification\footnote{On this subject, see two recent publication, Poccetti 2009 and Maras 2015.}.

A further group of items composes the ‘Epigraphic data’ section: the analysis begins from the ‘Position of the
inscription’ on the object, which indicates the relationship between the text and the inscribed object, an aspect that
has important consequences on the function and the fruition of the inscription\footnote{An important methodological
point was established by Susini 1982.}. We then have ‘Scriptura’, where the execution technique is described;
‘Direction of writing’, i.e. right-to-left, left-to-right, boustrophedon, etc.; ‘Dividing signs’, in which the presence
of punctuation and its eventual function is signaled; ‘Dimensions of the letters’, which is important as regards the
visibility of the inscription in relation with the object and the observer. This section ends with the ‘Epigraphic
commentary’, containing the description and analysis of the letters one by one both from the formal (shape-model of the
letter) and the factual point of view (possible particular features in the execution of the inscription) and some
general observations.

The ‘Notes and issues’ item gathers historical-bibliographical notes, observations and discussions on the most
problematic points of each inscription: the insertion of the discussion at the end of the chart offers the advantage of
having all the basic information on the object and on the inscription immediately available, while the study in depth
of the issues that deserve a thorough analysis is treated in a separate section.

The chart is closed by the ‘Bibliography’ section. The ‘Editio princeps’ and the possible ‘First notice’ (if the
inscribed object had been mentioned in a publication preceding the first edition of the text) are indicated in two
separate items. Then the complete bibliography of the inscription follows in chronological order, from the most dated
to the most recent publications; the chronological order, in comparison with the alphabetical one, allows us to find
more easily the latest works on the inscriptions or those published in a certain period in the history of the studies.

4. Technical notes [GS]

An epigraphic corpus can be digitalized in different ways, according to the specific issues that each project intends to
tackle. Unlike EDR that, as the other projects constituting the Electronic Archive of Greek and Latin Epigraphy
(EAGLE)\footnote{http://www.eagle-eagle.it/. The other databases related to EAGLE are the Epigraphic Database Bari
(EDB), http://www.edb.uniba.it/, the Epigraphische Datenbank Heidelberg (EDH),
http://www.uni-heidelberg.de/institute/sonst/adw/edh/index.html and Hispania Epigraphica (HE),
http://www.eda-bea.es/.}, is a database, the archaic Latin inscriptions have been digitalized according to the EpiDoc
Guidelines\footnote{For further information on EpiDoc and its history, see Elliott, Bodard, Cayless et al. 2006-2013.
The guidelines are available at http://www.stoa.org/epidoc/gl/latest/.}. This is a set of specifications and encoding
tools in XML (eXtensible Markup Language) for the scientific edition of ancient texts based on the Text Encoding
Initiative (TEI), a set of XML specifications designed for the digital publication of texts and manuscripts for
research purposes\footnote{http://www.tei-c.org/index.xml. See also Burnard, Baumann 2014 (About these Guidelines):
“The TEI encoding scheme is of particular usefulness in facilitating the loss-free interchange of data amongst
individuals and research groups using different programs, computer systems, or application software”.}. EpiDoc is
becoming more and more a point of reference for digital epigraphic projects\footnote{On the relationship between EAGLE
projects and EpiDoc see Felle 2012.} and it is also the standard chosen for the aggregation of the archives’ data in
the recently constituted network, again called EAGLE (Europeana Network of Ancient Greek and Latin
Epigraphy)\footnote{http://www.eagle-network.eu. About the new Best Practice Network, co-founded by the European
Commission, see http://www.europeana.eu.}, in which, besides the data of the EAGLE archives, the archaic Latin
inscriptions corpus will also converge. As a matter of fact, inscriptions encoded with EpiDoc are not only compatible
with other projects created according to these Guidelines, but they can also be transferred from a system into another
without losing any information; actually, since XML consists in a semantic markup, that is related to the content of
the information and not to its appearance, it is possible to modify the look of the final result by simply operating on
the style sheet\footnote{In XML, all information related to the formatting of the text are registered on a separate
file called style sheet, see Bodard 2010, 104, 110-111.}, not having to revise the single files. This will facilitate
the integration of the archaic Latin inscriptions into wider digital collections such as EDR and EAGLE-Europeana.
Moreover, since the file thus encoded can also be translated into another markup language, their survival despite any
future technical evolution is guaranteed\footnote{Tissoni 2008, 37-38 and Bodard 2010, 104-105.}.

Furthermore, the XML edition of an inscription (or of an entire corpus) created according to the EpiDoc Guidelines will
produce a digital edition of the text that complies to Leiden Conventions\footnote{On Leiden Conventions, the standard
used to annotate epigraphic documents and papyri in printed editions, see Krummrey, Panciera 1980, Panciera 2006a and
2006b. About their use in EpiDoc’s files see Elliot 2007, Mahoney 2006, 229 and Bodard 2010, 105.} that will show the
same typographical marks a printed edition following the Leiden system would have, thus being immediately
comprehensible to any epigrapher.

The archaic Latin inscriptions pose various epigraphic problems, related to their antiquity, that require specific
solutions also with regard to the markup. Since the EpiDoc Guidelines were originally conceived to encode in XML later
epigraphic documents, it has been necessary to adapt these Guidelines to respond to the peculiar issues of this
corpus\footnote{See Sarullo 2011, 162-167, where a few examples of markup are quoted. }. For this reason, some elements
have been adapted and others have been specially designed, and this was possible thanks to the fact that XML is an
extensible system.

The major encoding issues concern the rendering of the direction of writing and of reversed and upside-down letters.
Unlike later texts, predominantly left-to-right, the inscriptions of the ILA corpus show a certain degree of
fluctuation in the direction of writing. Besides left-to-right, right-to-left and boustrophedic inscriptions, there are
also some particular cases, such as the lamina from Lavinium (CIL I2 2833) and the Tibur pedestal inscription (CIL I2
2658), that requires a specific treatment. For these texts, it was necessary to adopt the {\textless}lb
rend={\textquotedbl}up-to-down{\textquotedbl}{\textgreater} and {\textless}lb
rend={\textquotedbl}down-to-up{\textquotedbl}{\textgreater} elements in order to render the peculiar directions in
which the text was cut\footnote{The issue of the direction of writing was the subject of much debate at the 6th EAGLE
International Event Off the beaten track. Epigraphy at the borders (Bari, September 24th{}-25th, 2015). The discussion
highlighted how this is a matter of great relevance both for the archaic inscriptions and the testimonies from late
antiquity and the necessity to establish a common standard to encode the instances of “non-standard” directions of
writing emerged. For this reason, this issue will surely be the object of further consideration in the future.}.
Reversed and upside-down letters are usually left unmarked in traditional epigraphic editions and we decided to comply
to this practice. Nevertheless, the {\textless}hi{\textgreater} element has been used to mark up these letters, with
two different values of @rend. For reversed letters, the {\textless}hi
rend={\textquotedbl}reversed{\textquotedbl}{\textgreater} was used, a tag that in the EpiDoc Guidelines is used to
encode the litterae inversae, enclosed in double round parentheses, such as in ((C)) for mulier\footnote{See Panciera
2006b, 1722.}. For upside-down letters, a new value was provided, {\textless}hi
rend={\textquotedbl}upside-down{\textquotedbl}{\textgreater}, since none of the allowed values of @rend for the
{\textless}hi{\textgreater} element is suitable for this issue.

XML also allows us to encode the semantic structure of the texts. This kind of markup does not influence how the text is
displayed but it is essential to generate the Index verborum and to allow a word-based search within the corpus. The
antiquity of the documents compelled to index the words as they appear in the inscription, because in most of the cases
a lemmatization would be forcing; for the same reason, some sequences that remain difficult to interpret were not
segmented and the search for a portion of text will be possible.

Finally, the EpiDoc file contains all information about the text-bearing object (description, dating) and the text
(critical apparatus, commentary); the encoding of these data allows us to generate the indexes that, together with the
bibliographical references and the images, enrich the digital edition and make the utilization of the text more
complete.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Noemí Moncunill Martí}
\date{2015-11-15}
\begin{document}
Abstract

\begin{figure}
\centering
\begin{minipage}{10.901cm}
1.\ \ Epigraphy and onomastics in the Hesperia databank\newline
\newline

Noemí Moncunill1, Javier Velaza2

Université de Paris-Sorbonne (Paris IV)1, University of Barcelona2

Email(s): velaza@ub.edu

nmoncunill@gmail.com

velaza@ub.edu

\bigskip
\end{minipage}
\end{figure}
The first part of this work provides a general overview of the features and advantages of the digital epigraphic corpora
on the basis of the experience gained in the last years within the Hesperia project. The second part of the paper
provides a detailed presentation of the new sections available in the Hesperia databank devoted to indigenous personal
names and divinity names.

\bigskip

Keywords

Hispania, epigraphy, onomastics, digital corpora, Palaeohispanic languages and writings

\bigskip

\bigskip

\bigskip

1. Introduction.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
(2014SGR63), on the one hand, and 655938 — ZEPHYROS — H2020-MSCA-IF-2014 project, on the other.}

The aim of this work is to do a general overview of the computer-based epigraphic corpora, to think about their features
and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.

The first thing that necessarily needs to be highlighted is that traditional epigraphic corpora have certain limits that
are imposed by their morphology and format. If summarizing, there are three main and most evident limitations:

a) One-dimensional format. The monument's description, the textual edition, the critical apparatus and remarks are
displayed one after another, with no wider possibility to link one to another than through some indices, which by the
way are rarely detailed enough. Not even in the most extended indices, as those in CIL, the connexions between
inscriptions through such relevant data as their date, palaeography or onomastics are set clear. Obviously, those
connexions can be set by the users in their mind, but when the corpus displays thousands or tens of thousands of
inscriptions, it looks too optimistic to take for granted that most readers will have the aptitude such work takes.

b) Ephemeral usefulness. It looks also obvious that a traditional corpus will necessarily be always ephemeral for many
reasons: new inscriptions cannot be added, mistakes cannot be corrected, text editions cannot be improved and
modifications based on new evidence or knowledge cannot be introduced, among other limitations. Of course some of these
problems have hardly been solved by publishing some supplements, but it is widely known that these are partial
solutions only.

c) Authoritative work. In a traditional corpus, editors impose their knowledge in a large range of features, and thus
they interfere in the utility of other views, to the extend of depreciating them too much. When it comes to an
epigraphic corpus that is written in a wide-known language, such as Latin, this auctoritas becomes particularly
powerful regarding the reading of the epigraphic text, its checking being not possible all the time for users,
particularly for those who are not experts on the subject. But when it comes to fragmentary languages, with very low or
even no language deciphering at all, and with many doubts on decoding, the truth is that the influence of the editor is
extremely high and involves some risks that they do not always come to terms with successfully.

In our opinion, these limitations can be solved or, at least, diminished by a proper and efficient use of the tools IT
puts to our disposal. Although this is no longer anything of new, any attempt to use them so far has brought to an
output too close to the traditional, non-innovative epigraphic corpora, considering the features of such new tools. To
get over one-dimensional formats, an open, adaptable digital structure must be created that shall take the most out of
every single datum in the database. Besides, getting over an ephemeral usefulness requires the possibility not just to
keep adding new data, but also to connect the database with any other tool with a similar structure and to extend the
platform with any new or advisable features. Finally, the authoritative temptation must be also avoided, particularly
through systems that shall bring to users the possibility to take their own decisions in certain issues of an open
debate and, at the end of the day, to customize their own corpus.

Along with the above-said, some other advantages any digital edition shall bring must be taken into account, although it
looks unnecessary to discuss them: for instance, an online building of the corpus shall allow cooperation and
simultaneous work of interdisciplinary teams that will not only contribute to speed up the work and check the
achievements, but also to enrich the views in such studies and the analysis of the monuments and their texts. The
open-access format of the corpus means also a big leap for research in this field, along with an important service of
transferring knowledge to society. However, it must be recalled that every corpus shall be adapted to the specific
features of the documents and their epigraphic culture (and other sources, eventually) has left to us. And such terms
require a constant dialogue between epigraphists and technicians since the very beginning of the project, in spite of
the fact that they do not always come to understand each other. Those projects that had any of these groups prevailing
at the start of the project have proved to have insuperable lacks through their stages.

2. The Hesperia project

All above-mentioned principles have been taken into account in Hesperia project, its main goal being to gather all
linguistic evidences from Palaeohispanic languages, that is, pre-Roman languages in the Iberian Peninsula\footnote{
General presentations of the project can be found in Luján - Orduña (forthcoming), Orduña - Luján (forthcoming) and
Velaza 2014.}. First of all, we need to recall these are very different materials as for their quantity, quality and
reliability. Ancient Hispania left us epigraphic texts in at least four languages: Tartessian, Iberian, Celtiberian and
Lusitanian -{}-the possibility there are also some texts in Palaeobasque language is still sub iudice\footnote{
Concerning this subject, see Velaza 2009.}{}-{}-, mainly written in four epichoric scripts -{}-Tartessian,
southeastern-Iberian, northeastern-Iberian and Greek-Iberian script.

 \includegraphics[width=10.987cm,height=7.835cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img001.jpg}

Fig. 1: Map of the Palaeohispanic inscriptions attested in the Iberian Peninsula.

The current knowledge of each of these scripts is in very different stages, just as when it comes to their languages
deciphering. Along with the epigraphic texts there are also other evidences, as those from onomastics -{}-anthroponymy
and theonymy preserved in Roman inscriptions found in these territories, and toponymy from Classic sources-{}- and some
notes transmitted by several authors.

Hesperia was born as a natural successor in the digital era of the main corpus for Palaeohispanic languages, that is
Monumenta linguarum Hispanicarum, published between 1975 and 2001 by Jürgen Untermann. The project conception is due to
Javier de Hoz and its digital platform has mainly been developed by Eduardo Orduña. Nowadays it is coordinated by an
interdisciplinary and interuniversity team with researchers from the Universidad Complutense de Madrid, the Universidad
del País Vasco, the Universidad de Zaragoza and the Universidad de Barcelona. There are currently several sections open
for searching (http://hesperia.ucm.es/).

 \includegraphics[width=10.896cm,height=7.669cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img002.png}

Fig. 2. Dispersal areas of Palaeohispanic inscriptions and sections available in Hesperia databank

i) Partial epigraphic corpora from B zone (Iberian inscriptions from the south of France), K zone (Celtiberian
inscriptions) and L zone (Lusitanian inscriptions). All of them follow the same data form made of six tabs
corresponding to {\textquotedbl}general features{\textquotedbl}, {\textquotedbl}text{\textquotedbl},
{\textquotedbl}epigraphy and paleography{\textquotedbl}, {\textquotedbl}pictures{\textquotedbl},
{\textquotedbl}arqueological context{\textquotedbl} and {\textquotedbl}bibliography{\textquotedbl}. The databases are
linked between them through a powerful engine for simple and complex searches, and pdf documents as well as maps can be
created with the resulting outcomes. Work is currently being focused on C and D zones (Iberian inscriptions in
Catalonia), which is scheduled to be open for search by the end of 2015.

ii) A numismatic database with its specific four-tab structure: {\textquotedbl}general features{\textquotedbl},
{\textquotedbl}inscription{\textquotedbl}, {\textquotedbl}language and writing{\textquotedbl} and
{\textquotedbl}bibliography{\textquotedbl}, where all Paleohispanic mints are gathered, no matter their scripts or
languages.

iii) An onomastic database, with a specific search engine allowing combined searches with any of the elements typical
from the identified onomastic formulae.

Beyond finishing, checking and updating the already-open sections, the team currently works on transversal fields, such
as lexicon, as well as on creating several kinds of tools, as the possibility of displaying the corpus according to
alternative readings -{}-which turns out to be essential in epigraphic contexts with a lower level of decoding, as in
Tartessian, although its applicability to other fields is still being tested-{}-.

Besides, the team has also decided to consider the potential applicability of this corpus structure to other cultures of
the western Mediterranean. With such aim, prof. Francisco Beltrán leads the AELAW project (Ancient European Languages
and Writings), with the participation of researchers from ten different countries, and that has been granted a COST
program in order to build in future a single corpus with the evidences from all fragmentary languages documented in the
western Mediterranean for the Ancient times.

3. The new sections on onomastics.

The Hesperia databank has been recently extended with a new resource on indigenous onomastics. More specifically, the
new sections now available are devoted to the anthroponymy and theonymy, leaving for the near future the data referring
to toponymy and ethnonymy, which are not yet available in open access.

In this new section we have compiled all the pre-Roman personal names and divinity names attested both by direct and
indirect tradition. This represents, so far, a total of nearly 6.000 records. Names attested in direct sources are
those that can be identified in epicoric epigraphies; the second group, in its turn, contains the names that can be
identified in the so-called colonial epigraphy, as well as in literary sources. Regarding the chronology of the
epigraphic material, the oldest texts are the Iberian, which can be dated back to the 5th century BC, whereas the most
recent ones are the Latin inscriptions from the high-imperial age. Thus, the processed information can be classified
into one of these groups:

\begin{itemize}
\item Indigenous names in Iberian inscriptions
\item Indigenous names in Celtiberian inscriptions 
\item Indigenous names in Lusitanian inscriptions
\item Indigenous names in Greek inscriptions
\item Indigenous names in Latin inscriptions
\item Indigenous names in ancient authors
\end{itemize}
The Greek and Latin epigraphy group mainly contains inscriptions from the Iberian Peninsula; nevertheless there are also
a few exemplars coming from outside this territory, but which clearly refers to individuals of Palaeohispanic origin.
The most significant document that fits into this category is the Ascoli bronze, which displays a long list of Iberian
equites to which the Roman citizenship was given. As a matter of fact, this artefact has become the most important
document for the study of Iberian personal names.

As said above, the database contains about 6.000 records so far; each of them reports a divinity name and/or the whole
onomastic formula of individuals whose complete name presents at least one indigenous element. This means that each
item is devoted to a concrete person, whose onomastic formula might be composed by several indigenous names. The
information of each record is completed with a bibliographical apparatus, and the find-spot coordinates, together with
a map. Moreover, the database allows using this geographic information to create new linguistic maps, which represents
one of the main strengths offered by this new tool of the Hesperia databank.

The following could be a good example of one of the above-described records. This Latin inscription from southern Spain
mentions a woman with a Latin nomen, Aelia, followed by an Iberian cognomen, Belesiar, as well as the name of an
indigenous divinity, Betatun. Thus, a single table records an indigenous personal name together with an indigenous god.
The lower part is reserved for the bibliographical references, the geographic information and, last but not least, the
map.

 \includegraphics[width=11.749cm,height=6.712cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003.png}

Fig. 3. Personal-names and divinity-names record in Hesperia databank.

A slightly different kind of record is conceived to compile the personal names attested in literary sources. As shown
below, in these cases the bibliographical field is merely used to mention the ancient passages in which the name is
attested; there is no geographical information, and the “observations” field is often used just to report the different
graphical variants of the name.

 \includegraphics[width=11.883cm,height=5.66cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img004.png}

Fig. 4. Personal-names and divinity-names record in Hesperia databank.

The goal of Hesperia is to provide, in the first place, an exhaustive repertory of all the Palaeohispanic names. Thus,
with the information available a map can be easily generated with all the spots where at least an indigenous name is
attested (in green); or a divinity name (in blue); or, finally, combine those maps in a single interactive map (in
green and blue), where each point is directly linked with the corresponding records.

 \includegraphics[width=7.031cm,height=5.359cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img005.png}

Fig. 5: Dispersal area of indigenous personal names.

 \includegraphics[width=6.98cm,height=5.262cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img006.png}

Fig. 6: Dispersal area of indigenous divinity names

 \includegraphics[width=6.32cm,height=6.045cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img007.png}

Fig. 7: Disperal areas of indigenous divinity names and personal names

Thanks to the search engine, all these points can then be redistributed into smaller groups to draw linguistic areas or
isoglosses, on the basis, for instance, of the attestation of a well-known anthroponymic element or the dispersal area
of significant phonetic features. In the following maps it is displayed, in the first place, the attestation area of
names containing the element biur, which clearly corresponds to the Iberian area, that is, the non indo-European half
of the Iberian peninsula; and, in the second place, it is displayed the dispersal area of the name Tancinus, which
clearly corresponds to the indo-European part.

 \includegraphics[width=6.491cm,height=4.833cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img008.jpg}

Fig. 8: Dispersal area of the anthroponymic element biur.

 \includegraphics[width=6.092cm,height=4.583cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img009.jpg}

Fig. 9: Dispersal area of the name Tancinus.

It must be pointed out that the Hesperia project is, in general terms, following the main standards for epigraphic
databanks and text editing, and we are now exploring ways to interact with the linked open data ecosystem (tagging
persons for compatibility with SNAP, places for compatibility with Pelagios/Pleiades, text references for compatibility
with IDEs, citations with CTS, object metadata with EAGLE, etc.).\footnote{ We thank Gabriel Bodard for his
observations at this regard.}

The example below could provide an easy way to show to what extent the interrelationship between different open-access
databanks might be useful even in the present stage of the Hesperia project, where prosopography and toponymy have not
yet been fully developed.

The following file contains information on a person’s or god’s name observed in an Iberian rock inscription in the
Pyrenees. Nevertheless, the structure of the word as well as its phonetic features show no possible interpretation in
Iberian. In consequence, other possibilities have to be considered, such as it could actually be an adaptation of a
Greek name. As a matter of fact, the same name is attested as a mythological character in literary sources (Parthenius
of Nicaea XXX 1; Ovid, Ibis 434), as well as record as a personal name in open-access databanks such as LGPN
(V5a-20577) or Trismegistos (Per 222620). Therefore, in this particular case, the linking between open-access data
could be helpful for the linguistic analysis itself.

 \includegraphics[width=10.862cm,height=6.325cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img010.png}

Fig. 10: record of a possible Greek name attested in an Iberian inscription.

A similar consideration might also apply, for instance, to another important document in the Palaeohispanic epigraphic
landscape, namely the III Botorrita Bronze plaque, written in Celtiberian language and script but containing a
considerable amount of foreign names, such as Latin or Greek.\footnote{ Vid. Unterman’s proposal in MLH IV, K.1.3. }
This could be the case for markos (Marcus), bolora (Flora), kinbiria (Cimbria), antiokos (Antiochus), bilinos
(Philinus), bilonikos (Philonicus), tais (Thais), tiokenes (Diogenes), and so on. We could finally mention, in this
same regard, the Gaullish names observed in some Iberian inscriptions from southern France: aśetile (Adsedilus, CIL III
5373), tesile (Tessillus, CIL III 14368.28), uaśile (Vassil(l)us, CIL XII 2286), katubaŕe (Catumarus, CIL 3, 4263),
etc.

To sum up, the new dynamic sections of Hesperia provide an essential boost to a field of study, the detection of
anthroponymic areas, which, thanks to the previous work by M. Gómez Moreno and J. Untermann, has been proved to be very
productive for the comprehension of the linguistic diversity of the ancient Iberia. The study of the indigenous
languages on the basis of the distribution of their personal names is actually essential, sometimes even the only
available means, for the definition of these areas that remained anepigraphic in pre-Roman times (see the map in fig.
2, with the dispersal areas of Palaeohispanic inscriptions). Obviously, one of the main advantages of the digital
format is that it offers the possibility to regularly update the corpus with new data. However, the most remarkable
difference from a traditional corpus is that it allows the user to connect and freely cross the information, and to
project the results automatically on a map, which makes this a powerful resource for new research and new results.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/29/EAGLE2016_camera_ready.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Angelos}
\date{2015-11-21}
\begin{document}
Abstract

\begin{figure}
\centering
\begin{minipage}{10.901cm}
1.\ \ AUGMENTING THE WORKSPACE OF EPIGRAPHISTS\newline
An interaction design study\newline

Angelos Barmpoutis1, Eleni Bozia2

Digital Worlds Institute, University of Florida1

Department of Classics, University of Florida2

Email(s): angelos@digitalworlds.ufl.edu, bozia@ufl.edu
\end{minipage}
\end{figure}
\ \ \ \ \ \ This paper presents the results of an interaction design study that focuses on the use of natural user
interfaces for professionals in the fields of epigraphy and archaeology. This study proposes solutions for utilizing
the sensors that can be found in popular handheld devices, such as tablets and smart phones, in order to naturally
perform common tasks from the typical workflow of epigraphists. The developed interface allows the users to naturally
hold digitized inscriptions, interact with them in order to relight or manipulate them as if they were real physical
objects, and interact with metadata or other multi-modal data, such as text and images.

Keywords

Mobile Applications, Interaction Design, Natural User Interfaces, 3D models, Archaeology

1.1. Introduction

The technological advances in the last decade have equipped the general public with several handheld electronic
commodities that changed significantly daily routine in a personal and professional level and contributed to the user's
quality of life not only in developed countries but also in developing economies in Africa and Asia [Osman 2011].
Handheld devices, such as tablets and smart phones, not only connect the users with tremendous amount of information
through the internet, but also offer interfaces for natural user interaction that enable non-technology-oriented
populations to use computers intuitively.

In the fields of epigraphy and archaeology, the areas of digital epigraphy and computational archaeology have benefit
from the use of several of the sensors available in hand-held devices. Crowd-sourcing of photographic data and
geo-spatial information, augmented-reality navigation in archaeological spaces and museums, and 3D scanning of
historical artifacts, using smart phones and tablet computers,, are few of the exemplar applications of hand-held
sensors in epigraphy and archaeology. One common component in all the aforementioned applications is the ability to
record tridimensional data either in the form of geo-spatial coordinates, or in the form of local 3D point coordinates
needed for augmented-reality interaction, or for the construction of triangular meshes of 3D models. \

\ \ There are several examples in literature that present 3D digitization projects that have been undertaken by museums
including the Epigraphical Museum of Athens [Sullivan 2011, Papadaki et al. 2015], Museo Arqueológico Nacional de
Madrid [Ramírez-Sánchez et al. 2014], Museo Nazionale Romano di Palazzo Altemps [Barmpoutis et al. 2015], Museo
Geologico Giovanni Capellini di Bologna [Abate \& Fanti 2014], National Museums Liverpool [Cooper et al. 2007],
Smithsonian Institution [Wachowiak and Karas 2009], and several other museums and institutes [Gonizzi Barsanti and
Guidi 2013, Landon and Seales 2006, Levoy et al. 2000].

Several novel methods for scanning, processing, and analyzing 3D models of inscriptions have been developed, including
methods for text extraction from inscriptions [Aswatha et al. 2014, Sullivan 2011], accurate 3D scanning of
inscriptions [Papadaki et al. 2015], visualization of inscriptions [Bozia et al. 2014], as well as 3D applications for
other archaeological artifacts [Babeu 2011, Pollefeys et al. 2001, Malzbender et al. 2001, Esteban \& Schmitt 2004].
Comparative studies of 3D scanning methods for cultural herirage can be found in [Pavlidis et al. 2007] and [Böhler \&
Marbs 2004].

\ \ The aforementioned examples show that the use of 3D technologies in epigraphy and archaeology has been a
well-studied topic over the past two decades. However, there is a notable disconnect between the research on these
technologies and the actual use in the professional epigraphic and archaeological practice as it has been hard for
non-technology-oriented audiences to handle and manipulate tridimensional data, using conventional computer equipment.
Furthermore, without mechanisms for proper user interaction, a 3D model that is projected on a 2D screen is not
significantly advantageous compared to a set of 2D photographs.

The recent advances on Natural User Interfaces (NUI) along with their marketing as low-cost general-purpose devices
(smart phones and tablets) have created a nurturing environment for integrating them in cultural heritage applications.
Popular low-cost NUIs, such as touch screens, marker-less position trackers, motion sensors, and head-mounted displays
have been recently studied and employed by museums as mechanisms for multi-sensory virtual experiences [Ujitok \&
Hirota 2015, Soile et al. 2013, Ikei et al. 2015].

This paper tries to fill the gap between the 3D technologies and their actual professional application in the field of
epigraphy by proposing innovative uses of NUIs specially designed to serve epigraphists. This is, to the best of our
knowledge, the first systematic interaction design study in the field of epigraphy. This study proposes solutions for
utilizing the sensors that can be found in popular handheld devices to naturally perform common tasks from the typical
workflow of epigraphists. The developed interface allows the users to naturally hold digitized inscriptions and
interact with them in order to relight or manipulate them, as if they were real physical objects, and also interact
with metadata or multi-modal data, such as text and images.

1.2. Understanding the workflow of epigraphists

Understanding the users is one of the integral steps of interaction design, which is an iterative process during which
representative users interact with preliminary designs and provide useful feedback [Preece et al. 2015]. For the
purposes of this study, our team interacted with early adopters of our prototype system, who were epigraphists and
conservation specialists from Cornell University, the University of California, Berkeley, the University of Lyon 2, the
Berlin-Bradenburg Academy of Sciences and Humanities, the U.K. National Archives, and the University of Florida. The
goals of our interaction were twofold: a) to study the various forms of physical interaction that epigraphists have
with an inscription as a real physical object \ and b) to expose epigraphists to a digital interface that imitates
their interaction routine, using digital replicas of physical objects.

The first part of our study revealed 3 common types of interaction with the inscriptions as physical objects:

\begin{enumerate}
\item Change of point of view: Observation of the inscription from different viewing angles assists epigraphists
understand better the shape of the inscribed letterforms. \ 
\item Change of lighting conditions: Relighting the inscription by introducing artificial shadows or additional light
sources from different angles may reveal details that were not legible in the original lighting conditions.
\item Magnification of inscribed details: Close observation of an inscribed region of interest, with or without
artificial magnification, may assist epigraphists in assessing weathered fragments and make a better informed decision
regarding the deciphering of the original text.
\end{enumerate}
It should be noted that in addition to the above 3 types of interaction, there are two additional interactions that are
special cases of I and II. More specifically, the physical object can be either portable (such as a small fragment of
stone or other material) or not (when the inscription is on an inscription bearer). \ In the case of a handheld object,
interactions I and II involve manual movement of the inscription with respect to the fixed observer (case I) or the
fixed light source (case II), while in the case of large rigid objects the observer and the light source move with
respect to the fixed inscription. \

According to the above analysis, in the case of digitized inscriptions a NUI should provide the means for an epigraphist
to {\textquotedbl}hold{\textquotedbl} the virtual object, {\textquotedbl}move{\textquotedbl} the point of view with
respect to the virtual object, {\textquotedbl}manipulate{\textquotedbl} the virtual object with respect to the virtual
light source, and {\textquotedbl}focus{\textquotedbl} on details of interest. The next section presents a NUI-based
interaction design that proposes natural solutions to the aforementioned forms of interaction that seamlessly imitate
the typical workflow of epigraphists. \ \ \

1.3. Natural User Interface design for epigraphy

Natural User Interfaces consist of sensors that track the natural behavior of users and provide a natural form of
interactivity with computers and other electronic devices. The common forms of NUI sensors are: pressure sensors for
sensing touch gestures (e.g. touch screens and touch pads), motion sensors for sensing user-initiated changes in the
orientation and acceleration of the device (e.g. accelerometer, gyroscope, and compass), and position sensors for
tracking changes in the relative position of the user with respect to the device, such as body motions (e.g.
Microsoft's Kinect), finger motions (e.g. Occipital's Leap Motion), eye movements, and others.

\ An optimal interaction design solution should be intuitive, minimalistic, and non intrusive [Preece et al. 2015].
Therefore, in order to design interaction for epigraphy one should choose devices that are easily accessible by
epigraphists and do not interfere with their workspace (e.g. avoid introducing new devices or external sensors). All
forms of interaction described in Sec. 1.2 can be implemented, using motion and pressure sensors, which can be easily
found in tablet computers or smart phones. In both types of handheld devices the virtual object can also be assumed
handheld, without loss of generality, in order to generate a multi-sensory experience for the user (i.e. holding the
device = holding the digital inscription). \ \ Hence, NUI design is possible by utilizing accessible devices and
without the use of external sensors as it is described in details in the following sections.

\bigskip

1.3.1. Natural interactive relighting of 3D models

In order to achieve natural interactive relighting of an inscription, the system should imitate the process of
relighting a handheld physical object (such as a paper cast of inscription) by reorienting the object with respect to
the light of the environment. Without loss of generality, we can assume that the default virtual lighting source is
located on the ceiling, right above the device, which is also very intuitive choice as it is the most probable
real-world lighting condition. Under this assumption, a gyroscope, a sensor that tracks the orientation of the device
with respect to the gravitational vector, is enough to track the slope of the device with respect to the virtual light.
The top row of Fig. 1.1 shows the approximated real-world orientation of a tablet computer as it was estimated using
the gyroscope of the device. The orientation is updated in real-time as the user moves the device.

 \includegraphics[width=10.881cm,height=2.72cm]{EAGLE2016cameraready-img001.png} 
\includegraphics[width=10.881cm,height=2.72cm]{EAGLE2016cameraready-img002.png}

Fig. 1.1. Top row: Illustration of interactive manipulation of the virtual lighting by moving the device. The figures
show the corresponding field of normal vectors in the 3D space as computed using the gyroscope of the device. Bottom
row: Demonstration of interactive relighting of a 3D digitized inscription. Different virtual lighting angles reveal
different inscribed details. \ \ \

The estimated orientation of the device can be used in order to relight the depicted 3D model of inscription using the
angle of the device with the direction of the virtual light source. The bottom row of Fig. 1.1 demonstrates interactive
relighting of a digitized paper cast. Epigraphists can relight an inscription by reorienting the tablet as if it were a
real physical object. This process matches perfectly with the physical interaction of epigraphists with real
inscriptions and can be intuitively extrapolated to the case of 3D models of large inscriptions that were not handheld
in the real world (see Fig. 1.3).

\bigskip

1.3.2. Natural interactive manipulation of 3D models

Another important form of interaction in the epigraphic routine is the change of the point of view in order to
understand better the structural details of the inscribed letterforms. Assuming that the model of the inscription is
parallel to the screen of the device, the change of point of view involves only change of the perspective projection of
the digital object without any virtual rotation. In such case the rotation of the object is equivalent to the physical
rotation of the device without any virtual rotation of the object.

Figure 1.2 shows 15 different projections of the same virtual cube that correspond to the change of perspective caused
by moving the observer's head parallel to the screen. Note that all cubes are parallel to each other.

\bigskip

 \includegraphics[width=10.881cm,height=5.611cm]{EAGLE2016cameraready-img003.png}

Fig. 1.2. Visualization of same-size boxes using 15 different perspective projections with the same FOV angle and
different cropping parameters. None of the boxes is rotated in the space. \ \

\bigskip

In the case of a tablet computer, the change of perspective can be implemented using the accelerometer of the device,
which senses non-gravitational accelerations in the 3D space. With the logical assumptions that: a) the tablet is
initially facing the user, and b) the user's eyes remain in a relatively fixed position in the 3D space (otherwise an
eye-tracker should be required), the change of perspective can be realistically achieved by naturally reorienting the
tablet as shown in the top row of Fig. 1.3. The superimposed boxes in this figure were estimated, using the
accelerometer reading of the device for three different orientations of the tablet.

Interactive manipulation of a 3D digitized inscription bearer is shown in the bottom row of Fig. 1.3. Different sides of
the bearer can be observed by reorienting the tablet naturally. In this example a large inscription model with its
bearer was chosen in order to demonstrate that the proposed interaction design remains intuitive independently of the
scale. It should also be noted that the interactive manipulation of the perspective can be optionally performed
simultaneously with the interactive relighting as shown in Fig. 1.1 (bottom) in order to perform a more realistic
interaction that causes relighting and change of point of view at the same time. \ \ \

\bigskip

 \includegraphics[width=10.881cm,height=2.72cm]{EAGLE2016cameraready-img004.png} 
\includegraphics[width=10.881cm,height=2.72cm]{EAGLE2016cameraready-img005.png}

Fig. 1.3. Top row: Illustration of interactive manipulation of the perspective projection by moving the device and using
the {\textquotedbl}fixed eye{\textquotedbl} assumption. Bottom row: Demonstration of interactive visual inspection of a
3D digitized inscription along with the inscription bearer. The user can view the object from different perspectives,
using natural motions. \

\bigskip

1.3.3. {\textquotedbl}Touching{\textquotedbl} the metadata: Interacting with multi-modal data

The forms of interactions presented in 1.3.1 and 1.3.2 involved only the motion sensors of a tabled computer without
utilizing the pressure sensors of the screen of the device. The commonly used touch gestures (such as 2-finger twist to
rotate, 2-finger pinch to zoom, and 2-finger translate to move) can be employed in order to enhance the proposed NUI
design. In addition to the aforementioned gestures, a tap gesture could activate regions of interest with additional
modalities of information such as text, images, and metadata. The user can interactively browse the different forms of
data by using intuitive touch gestures as shown in Fig. 1.4. This set of 2D interactions along with the 3D NUI design
presented earlier can compose an intuitive yet powerful workspace for an epigraphist who can now perform digitally
several parts of the epigraphic workflow. \

\bigskip

 \includegraphics[width=10.881cm,height=5.754cm]{EAGLE2016cameraready-img006.png}

Fig. 1.4. Screenshot of our interactive environment. The user interacts with the 3D object, using touch gestures and
selects one of the regions of interests. This action initiates other data tools such as the image viewer or the edge
filter as shown in this example.

1.3. Conclusions and future directions

In this pilot study, a complete set of natural user interactions was designed based on the physical interactions of
epigraphists with real inscriptions. The proposed interactions utilize the existing sensors in a typical tablet
computer or smart phone in order to interactively relight a digitized inscription and manipulate the user's
perspective, using a set of intuitive gestures that imitate the natural interaction with a physical object. In the
proposed design, the epigraphist can {\textquotedbl}hold{\textquotedbl} a digital inscription, relight it by
reorienting it as a tangible object, observe it from different perspectives, and finally interact with other modalities
by following a set of 2D touch gestures. The prototype system was developed as part of the Digital Epigraphy and
Archaeology (DEA) project using the open-source library VisiNeat for 3D visualization and interaction and is compatible
with iOS, Android, and Microsoft RT tablet and smart phone devices. The interface is available through the web-site of
the project: http://www.digitalepigraphy.org

In the future we plan to quantitatively evaluate the designed interface by tracking the user activities and analyze
their motion patterns in the 3D space while they are interacting with their handheld device.

\bigskip

\begin{figure}
\centering
\begin{minipage}{10.901cm}
Bibliography
\end{minipage}
\end{figure}

\bigskip

\begin{figure}
\centering
\begin{minipage}{10.901cm}
Acknowledgement

\bigskip
\end{minipage}
\end{figure}
We would like to acknowledge our collaborators for their invaluable interaction and feedback during the past three
years. We would also like to thank the UF College of the Arts and the UF Center for Greek Studies for providing
continuous funding support for this project.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/3/EAGLE2016_Full_Paper_Template.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage{supertabular}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%

\title{Romans 1 by 1. Documenting a population database for the Roman world}

\author[VARGA]{Rada Varga}

\address{Babeș-Bolyai University, Cluj-Napoca}

\cortext[first]{Corresponding author. Email: radavarga@gmail.com}

\begin{document}

\maketitle

\begin{abstract}
 The present article briefly documents the epigraphic division of a developing online population database for the Roman
Empire, accessible at \url{www.romans1by1.com}. The paper presents the motivation of constructing the database and its
envisioned architecture, in relation with the various sources, while emphasizing on the steps and procedures required
in order to transpose epigraphical information into an ancient population database. 
\end{abstract}
\keywords{Epigraphy, middle classes, provincial society, occupational inscriptions, prosopography}

\noindent Romans 1 by 1 database and afferent website were created for filling in an existing gap in the study of Roman-era
population. The database tries to begin answering to the need of properly cataloguing all attested inhabitants of the
Roman Empire. Of course, this is a tremendous task and \url{www.romans1by1.com} is only a first step, circumscribed for now
to a very specific category of sources.

\section{Motivation}

\noindent But why would an ancient population database be essential? Because a digital resource focused on individuals would
reveal linkage possibilities that otherwise elude us, it would finally give us the complete and accurate image of the
Roman attested population and, through codifications, it would open the way towards computer-assisted in-depth analyses
on all relevant aspects imagined (epigraphic patterns, religiosity, migrations, onomastics, occupations, family data,
etc.). A complete, aggregate database will allow a longitudinal (diachronic) view on the attested Roman population from
a certain area and ideally from the whole Empire, while also opening transversal (a section in time) perspectives.

The database in itself would have three components (see below Fig.~\ref{fig:schemadatabase}), following the best practices in the field
(\citet{doi:10.3200/HMTS.37.1.34-38}): the sources database (with ``facsimile'' transcription of the sources’ text), the central
database (the complete, corrected, integrated, standardized and coded form of the sources database) and the data
releases (destined for on-line usage, allowing easy extraction of data in view of analyses).

Alongside these components, which actually represent various steps in the data preparation process, a population
database for the Roman Empire should be built on three pillars, which we might call units or divisions of the database:
epigraphic, literary and archaeological, each of them requiring different expertise, different approaches and different
standards for the individual recording forms. In the end, of course, all three types of individual records will have to
be integrated in a central standardized component, whose structure is to be developed by merging the three
aforementioned divisions; consequently, its configuration will ideally take shape only after all three units enjoy a
stable architecture at least in the sources component of the database.

The construction of the database will follow a series of steps imposed by best practices: creating a repository of
sources; introducing, integrating, standardizing, coding and storing the information; enriching and disseminating the
information. The codifications hold an essential part, not only in the individual linkage procedures, but in the
analysing process as well. At the point when the database will comprise enough data, properly recorded and with all
codifications undertaken, the usage of statistical software in order to identify trends and run comparisons over large
scale geographical and administrative units might result in a better than ever understanding of Rome’s
social history.

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.50]{1.png}
\caption{Schema of the database}
\label{fig:schemadatabase}
\end{figure}

\section{The sources}

\noindent Within this theoretical frame, whose amplitude implies a gradual and long-term approach, we have chosen to start by
building the epigraphical division of the database. Thus, we began developing a project on the middle classes of the
Low Danube provinces and their epigraphic attestations. The database created in this context together with the platform
www.romans1by1.com represent the skeleton of the future population database of the Roman Empire.

For now, we are solely focusing on characters attested epigraphically – thus on inscriptions as sources. As constructing
a metadata suitable for all social and professional categories of the provincial world is very complex, our database is
created for accommodating all information epigraphically provided on members of the provincial middle classes.
Terminologically, we consider all those who manifest themselves epigraphically, without being members of the imperial
or provincial elite, as member of the middle socio-economical layers. Likewise, we have excluded active militaries (but
not their families), as their appurtenance to the army creates their social status.

Regarding the data entry, we are trying to remain faithful to the source and to record, during the first phase, only the
minimum of deduced information (gender, juridical status, ethnicity of the name). At the same time, we operate some
conventional onomastic transcriptions (AEL will be Aelius from the start, etc.). All these basic normalization
procedures are being thoroughly explained and documented in the data entry manual, and the reason why we chose to apply
them is to speed up data standardization, run analyses and publish results on small samples, whose standardization
procedures would otherwise be overwhelmingly time-consuming in relation with the results.

\section{Database architecture}

\noindent The core of the database’s epigraphical division is formed by a table used for recording personal
individual data (labelled Personal data), around which the entire network of components needed to ensure proper
information recording is built. However, when starting data entry, one must begin by recording information about the
source in use.

The first table to be filled in is the file of the source – Inscriptions. First of all, each inscription gets a source
code, formed of 5 digits and a symbol/acronym of the province’s name (D for Dacia, MS for Moesia Superior, DAL for
Dalmatia, etc.) – so we have, for example 00001MS. The Inscriptions table has text fields, as well as value lists.
Then, we have text strings for: Relevant expressions, Stylistic details, Atypical features, Observations, Place of
discovery, Place of provenience, Ancient name provenience, Timestamp/Timeframe and External links. Although we are
aware that some of the data (Timestamp/Timeframe) could have benefitted from a standardized form, at this point we
opted for vaster possibilities of expression and adaptation. Other information will be filled using value lists, as
standardization is more suited: Type of inscription, Language, Material. For these we will use the Eagle
vocabularies,\footnote{\url{http://www.eagle-network.eu/resources/vocabularies}} because they offer an already
standardized language.

The Coordinates table links each inscription with the latitude and longitude of its place of provenience, in order to
place it on a map.

The Inscription bibliography was conceived so that extracting complete or selective bibliographical lists would be
possible. Thus, a normalization table includes all bibliographical titles referred to and from it; through a value
list, one can select the Bibliography abbreviation. The exact reference is presented and detailed in the Details and
Comments text strings. Of course, all data are linked to the Inscription code, selected as well from a value list.

Only after properly documenting the source one can access the main and most complex table, Personal data, where
basically the individual file of each person is created. At this point, each new entry represents a singular epigraphic
attestation of an individual, and a unique ID is generated, which will help link the respective character throughout
the various components of the database and with other database entries. The person is also manually linked to the
source using a value list of the inscriptions’ codes. In the event of one person being attested in
multiple epigraphs, each attestation will represent a new entry, and it will be assigned a new unique ID, which will be
doubled, during linkage procedures, by a common ID for all instances of the same person.

The Personal data table was built to host a large variety of information offered by epigraphs, and its structure has
proved, up to this day, rather stable. However, if need be, it can always be extended in order to accommodate any new
kind of information. A first set of its variables are name-connected: Praenomen, Nomen, Cognomen/Personal name,
Father/Master name, Agnomen, Signum, where each also has a drop-down associated for Ethnicity and the Agnomen and
Signum an Observations string. While we believe the possibilities of actually identifying signa for members of the
middle classes, during the Principate period, is rather reduced, we opted for facilitating their correct registration,
in case they are discovered. Other data regard Natione, Ethnicity, Origo and Domus – if and how they are mentioned in
the inscription. As acknowledged above, some information will be recorded, even if they are deductive and not literally
written in the source: Gender and Juridical status (though the servile one often is literally recorded). For the
latter, we have opted for a check box, which, if checked, opens all the available possibilities. The rest of the fields
accommodate supplementary information, when and how there is the case: Occupation, Collegium, Deities, Age (at death),
Details of life/death and Observations.

The Occupation field requires some special attention, as it has the Occupation code associated with it; as we are trying
to propose a codification of Roman occupations/professions, based on and adapting the HISCO classification
model\footnote{\url{http://historyofwork.iisg.nl/} }. While a raw classification and codification based on HISCO might only
be a slight challenge, finding a theoretical model might prove to be a more elaborated task. What we aim at is
constructing the “metadata” for the codification of Roman professions attested on stone and analyse how deep the
classification can realistically go. Much alike other normalization fields present in the sources component of the
database, Occupation code was implemented because it helps dealing faster with small and medium size samples (up to
hundreds of people) in view of publishing preliminary results which are vital for dissemination and further financial
support of the project.

Two important variables, for statistics and working with the data, are Dedicated by and Dedicated for, which state the
relation of the recorded individual with the epigraph and with other persons mentioned by it. Both appear in the simple
form of check boxes.

Another particular check box, in need of supplementary explanations, is Later. The database was conceived for attested
civilians from the middle classes, but sometimes they are associated in inscriptions with militaries or representatives
of the elites. In this case, we have to register a minimum of data on the later as well, in order to build a wider
image of the characters that we focus on. When checking Later, it opens Status, which at its turn opens the following
options in the form of check boxes: Senator, Knight, Local magistrate, Decurion, Imperial priest, Imperial slave,
Imperial freedman, Military personnel. If checked, each of these boxes opens a Details text box; additionally, Local
magistrate and Decurion open a City/Town value list, while Military personnel, opens Rank and Unit value lists. While
elite members and military personnel will at some point be added to the database, for the currently running project
purposes, their social and professional status, together with a minimum of relevant details and relation with the
recorded individuals represent enough information.

\begin{center}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{3.711cm}|m{2.941cm}|m{4.046cm}|}
\hline
Field label &
Field id &
Data type\\\hline
Praenomen &
rperson\_praenomen &
Text\\\hline
Ethnicity praenomen &
rperson\_eth\_praen &
Value list\\\hline
Nomen &
rperson\_nomen &
Text\\\hline
Ethnicity nomen &
rperson\_eth\_nom &
Value list\\\hline
Cognomen/Personal name &
rperson\_cognomen &
Text\\\hline
Ethnicity cognomen/personal name &
rperson\_eth\_cogn &
Value list\\\hline
Father/Master name &
rperson\_father &
Text\\\hline
Ethnicity father/master name &
rperson\_eth\_fth &
Value list\\\hline
Agnomen &
rperson\_agnomen &
Text\\\hline
Ethnicity agnomen &
rperson\_eth\_agn &
Value list\\\hline
Observations agnomen &
rperson\_remark\_agn &
Text\\\hline
Signum &
rperson\_signum &
Text\\\hline
Ethnicity signum &
rperson\_eth\_sign &
Value list\\\hline
Observations signum &
rperson\_remark\_sign &
Text\\\hline
Natione &
rperson\_nation &
Text\\\hline
Ethnicity &
rperson\_nat\_ethnicity &
Text\\\hline
Gender &
rperson\_rgender\_id &
Value list\\\hline
Juridical status &
rperson\_jstatus &
Checkbox. \ Opens checkboxes:

Citizen/Libertus/Peregrine/Slave/

Veteranus (opens Unit/Rank - valuelists)\\\hline
Tribus &
rperson\_rtribus\_id &
Text\\\hline
Origo &
rperson\_origo &
Text\\\hline
Domus &
rperson\_domus &
Text\\\hline
Collegium &
rperson\_collegium &
Text\\\hline
Occupation &
rperson\_occupation &
Text\\\hline
Occupation code &
rperson\_occ\_code &
Text\\\hline
Deities &
rperson\_deity &
Text\\\hline
Age &
rperson\_age &
Text\\\hline
Details of life/death &
rperson\_death\_details &
Text\\\hline
Dedicated for &
rperson\_dedicated\_for &
Checkbox\\\hline
Dedicated by &
rperson\_dedicated\_by &
Checkbox\\\hline
Inscription code &
rperson\_rinscription\_ids &
Value list\\\hline
Later &
rperson\_now\_stat &
Checkbox. Opens checkbox: Status\\\hline
Status &
rperson\_status &
Checkbox. Opens checkboxes:

Senator (opens Details – text)/

Knight (opens Details – text)/

Local Magistrate (opens Details – text, City – value list)/

Decurion (opens Details – text, City – value list)/

Imperial priest (opens Details – text)/

Imperial slave (opens Details – text)/

Imperial freedman (opens Details – text)/

Military personnel (opens Details – text, Unit/Rank – value lists)\\\hline
Observations &
rperson\_observation &
Text\\\hline
%\caption{Metadata of the Personal data table}
\end{supertabular}
\label{tab:personaldata}
\end{center}

Based on the personal ID given to each individual, the Relationship table will solely name the relationship between
individuals (A to B and B to A), choosing from a drop-down menu. The relationships have been encoded from the start, in
order to make processing quicker; thus first-degree relationships have 10-codes (101-Husband, 102-Wife), second and
third degree relations 20- and respectively 30-codes, non-family relations were given 40- codes and 50-s for the
unspecified/unreadable relations. Male relations were given odd numbers and female ones – even numbers.

\section{Conclusions}

\noindent Romans 1 by 1 is a first step towards a comprehensive and exhaustive electronic resource for the attested population of
the Roman Empire. The following normal steps for expanding and enriching the database is elaborating the fitted
metadata for provincial elites and military personnel epigraphically attested.

\section*{Acknowledgement}
This work began being developed during the period of a residential scholarship at the Hardt Fondation, Vandoeuvre. Its
expansion to the area of the whole Latin language Empire is being undertaken with the support of a postdoctoral
fellowship of the Fritz Thyssen Stiftung. Also important was the financial support of a POSDRU scholarship, granted by
the Babeș-Bolyai University.

This work began being developed during the period of a residential scholarship at the Hardt Fondation, Vandoeuvre. Its
expansion to the area of the whole Latin language Empire is being undertaken with the support of a postdoctoral
fellowship of the Fritz Thyssen Stiftung.

\bibliographystyle{../../sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/31/EAGLE2016_I Sicily_final copy.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\title{}
\author{a}
\date{2015-11-18}
\begin{document}
[Warning: Draw object ignored]Abstract

This paper introduces the EpiDoc corpus of inscriptions on stone for ancient Sicily, I.Sicily. The project is one of the
first attempts to generate a substantial regional corpus in EpiDoc. The project is confronting a number of challenges
that may be of wider interest to the digital epigraphy community, including those of unique identifiers, linked data,
museum collections, mapping, and data conversion and integration, and these are briefly outlined in the paper.

Keywords

Sicily, Epigraphy, Epidoc, Greek, Latin, Pleiades, multilingualism

\bigskip

1.Introduction: What is I.Sicily

I.Sicily is an online, open access, digital corpus of the inscriptions on stone from ancient Sicily.\footnote{ The
corpus will be mounted at www.csad.ox.ac.uk/sicily/isicily, but is presently on a development server. A public face is
currently maintained via a blog at http://isicily.wordpress.com/, as well as on Facebook at www.facebook.com/ISicily
and on Twitter via I.Sicily@Sicilyepigraphy.} The corpus aims to include all texts inscribed on stone, in any language,
between approximately the seventh century BC and the seventh century AD. The corpus currently contains records for over
2,500 texts, and when complete is likely to contain c.4,000. The corpus is built upon a conversion from a legacy
dataset of metadata in MS Access to EpiDoc TEI XML.\footnote{ http://sourceforge.net/p/epidoc/wiki/Home/ [accessed
26.09.2015].} \ The XML records are held in an eXist database for xQuery access, and additionally indexed for full-text
search using SOLR/Lucene. The corpus and related information (museum list, bibliography) are published as Linked Data,
and are manipulated through a RESTful API. The records are queried and viewed through a web interface built with
AngularJS and jQuery javascript components. Mapping is provided in the browser by the Google Maps API, and ZPR (Zoom,
Pan, Rotate) image-viewing is provided by the IIIP image server and the OpenSeadragon javascript library.

At the time of writing (September 2015), the main conversion routine is being refined and the epigraphic texts are being
collated for incorporation into the records. An ancillary database of museum collections in Sicily has been constructed
and bibliography is held in a Zotero library. Extensive search facilities will be provided, including map-based and
bibliographic searching. Individual inscriptions and individual museums will both be provided with URIs, as will
personal names and individuals; places will be referenced using Pleiades; epigraphic types, materials, and supports
using the EAGLE vocabularies.

2. The motivations for and origins of I.Sicily

The existing epigraphic landscape in Sicily is extremely diverse in two primary regards: on the one hand, the island has
a very mixed cultural and linguistic make-up, meaning that the epigraphic material is itself extremely varied, with
extensive use throughout antiquity of both Greek and Latin, as well as Oscan, Punic, Sikel, and Hebrew;\footnote{
Recent overview of much of the linguistic tradition in Tribulato 2012; and of the epigraphic material in Gulletta
1999.} on the other hand, the publication of this material has a very uneven record and despite an excellent
pre-twentieth century tradition, the existing copora are far from complete and the ability of key journals such as SEG
or AE to keep pace with local publication has been limited.\footnote{ For an overview of the corpora tradition up to
the twentieth century, see De Vido 1999.} A limited number of museum-based corpora have been published in recent
decades (for Catania, Palermo, Messina, and Termini Imerese, as well as the material from Lipari), but this has not
greatly improved the overall situation.

The combination of these two factors already means that locating, identifying, or working with a Sicilian inscription,
or its publication record, is extremely challenging for anyone without extensive experience of the material. The
situation is compounded by the universal and familiar challenges of the recording and accessibility of archaeological
collections, whether held in museums, in archaeological stores, or elsewhere, and the lack of consistency in the
publication of \ new material.

As noted in the introduction, some of the impetus for I.Sicily comes from a desire to exploit a substantial legacy
dataset in MS Access. This consists of a single table originally constructed in MS Access 2000, and maintained
erratically from the year 2000 onwards. The original purpose of this table was to gather data to assess the ‘epigraphic
habit’ of ancient Sicily, and consequently the texts themselves were not the primary focus. However, the extent to
which the dataset facilitated further study made increasingly clear its potential value for the study of Sicilian
epigraphy.\footnote{ The principal results were published in Prag 2002, revised in Prag 2004: 159-188; cf. Prag 2003,
2007, 2008, 2010.}

In its final form the table holds data across 39 different fields, for 2575 records. 17 of these fields detail
publication history (corpora references and other bibliography); the other fields record information on the language,
date, provenance, current location, epigraphic type, form and material of the inscriptions, together with a free-text
field recording further information about the inscription and fields to record any autopsy undertaken. Almost all of
this data is derived from existing publications.

The conversion from the original MS Access dataset was developed through a pipeline of known conversions going from MS
Access to CSV to TEI P5 XML. The XSLT transformation of the table of data from TEI P5 XML to EpiDoc XML is the point in
the process where further up-conversions of the data were made. These include the creation of the hierarchical EpiDoc
XML as well as normalisation of dating and bibliographic records. This conversion is not meant to be repeated as the
dataset, once converted to EpiDoc XML, will be edited in the I.Sicily website. While the conversion preserves the data
from the MS Access dataset, it restructures and where possible improves or normalises it.

By virtue of the fact that I.Sicily begins from such rich metadata, to which texts, images, and further data will be
added over time, and because this is in turn being supplemented by an on-going programme of autopsy, the form and
content of I.Sicily is intended to be more akin to that of a true corpus than simply a text-database, seeking to
combine a full record of past publication and study with a fully revised edition, and potential for multiple
individuals to contribute to a process of on-going revision (see Fig. 1 for a draft edition of one inscription (AE
1962.314 = I.Sicily 820)).

\bigskip

3. The aims of I.Sicily

We outline briefly five areas in which I.Sicily aims to develop, facilitate and improve the study of epigraphic material
from ancient Sicily.

\bigskip

3.1. Multilingualism

Sicily is traditionally described as a ‘melting pot’, the ‘crossroads of the Mediterranean’. The negative consequences
of the separation of epigraphic material according to linguistic traditions have recently been highlighted and directly
confronted by the Corpus Inscriptionum Iudaeae/Palaestinae (CIIP), edited by H. Cotton et al.\footnote{ Original notice
in Cotton et al. 1999; presentation in Cotton and Price 2007.} I.Sicily sets out to follow in that mould, since the
different linguistic traditions of Sicily not only exist side-by-side but interact constantly throughout the island’s
history, and no study of the epigraphic material can afford to ignore contemporary and parallel material in the other
languages.\footnote{ See e.g. Manganaro 1993, Prag 2002, Salmeri 2004, Korhonen 2011, Tribulato 2012.} The situation
created by basic technologies such as Unicode and EpiDoc XML mean that there is now no reason not to be language
agnostic in the inclusion of material (the point may be obvious, but the tendency towards language-specific corpora is
still marked). The opportunities and possibilities offered by these technologies are considerable, even at the most
basic level, since, for example, searching can be made language specific or language neutral. One obvious area where
Sicilian studies are currently hampered by this partitioning is in the study of onomastics. The Lexicon of Greek
Personal Names records most instances of Greek names for the island, but Sicily is no less rich in non-Greek names
(Latin and others), and at present there is no onomasticon for the island.\footnote{ www.lgpn.ox.ac.uk/ [accessed
26.09.2015].} Simply by the marking-up and indexing of all names in the island’s inscriptions, I.Sicily will generate a
powerful tool for future study. Although I.Sicily in its first phase is not undertaking morphological or syntactical
mark-up, the encoding of all these text in XML constitutes the necessary first stage in such a development, and we see
this as a highly desirable future project, and the possibilities for the field of historical linguistics are
considerable. The incorporation of a full range of metadata on the epigraphic support, geographical location,
chronology, etc. will likewise allow detailed analysis of cultural patterns and their relationship to language-use over
time.\footnote{ See Prag 2002 for a first effort in this direction.}

\bigskip

3.2. Identification and Bibliography

Sicily presents a particularly extreme version of the common problem of identifying a text and its publication record.
No existing corpus in either Greek or Latin comes close to full coverage (CIL X and IG XIV are the largest individual
traditional corpora for the region, but both are over 125 years old and cover less than 30\% of the material now
known).\footnote{ Mommsen 1883; Kaibel 1890.} Existing online databases improve on this situation, but the results
obtainable are of very varied value. The most comprehensive, in terms of the range of data recorded, is EDR (with which
I.Sicily is collaborating), which currently reports 1,906 records for ‘Sicilia’; but this reduces to 833 when limited
to texts on stone (‘lapis’ or ‘marmor’); contrast I.Sicily, with 2,563 records at the time of writing.\footnote{
www.edr-edr.it/ [accessed 26.09.2015]} Clauss Slaby reports 4,374 records for ‘Sicilia’ (including Christian
inscriptions, excluding ‘sigilla impressa’), but the return is inclusive of all kinds of epigraphic material, without
indication or discrimination, contains some duplication, is much harder to reconcile to existing records, and records
only text.\footnote{ www.manfredclauss.de/ [accessed 26.09.2015].} The PHI database of Greek inscriptions has a rich
record of published Greek texts, but is text only and limited in outputs.\footnote{
http://noapplet.epigraphy.packhum.org/regions/1156 [accessed 26.09.2015].} SEG references are available for 733
inscriptions on stone and AE references for 328 (data taken from the I.Sicily database and based upon comprehensive
manual trawls of SEG and AE).

One major aim of I.Sicily, therefore, is to generate unique identifiers for each inscription - the I.Sicily number, in
the form ISic 1234. These will be maintained as URIs, of the form:

\ \ http://csad.ox.ac.uk/sicily/inscription/ISic1234

I.Sicily is well placed to do this since its initial dataset is primarily a bibliographic concordance of the lapidary
inscriptions of Sicily. One of the associated outputs of the project will therefore be an online bibliography for
Sicilian epigraphy, and an online Zotero library has already been created with over 700 records which are referenced in
the EpiDoc files.\footnote{ www.zotero.org/isicily (private at the time of writing, but to be made public when the main
site is launched) [accessed 26.09.2015].} A locally cached version of the bibliography will be presented at the
I.Sicily site to facilitate detailed bibliographic searching (including the identification of inscriptions by
publication) and to allow the generation of customised concordances.

A further element of bibliographic information which I.Sicily will include is the cross-referencing and linking to
online editions of major antiquarian corpora of Sicilian inscriptions. A growing number of these are already available
in digital format and several are already mounted in the Arachne archive, making direct page-citation
possible.\footnote{ E.g. Castelli 1784, at http://arachne.uni-koeln.de/books/Castelli1784 [accessed 26.09.2015].}

The richness of I.Sicily’s records in this area mean that I.Sicily is currently collaborating with both Trismegistos and
IDES (‘Integrating Digital Epigraphies’).\footnote{ www.trismegistos.org/ [accessed 26.09.2015] and http://ides.io/
[accessed 26.09.2105].} The former aims to generate TM numbers for all the Sicilian material (which I.Sicily will
include); the latter is to assist IDES in the refining of links between, e.g., PHI and SEG records, and to improve
I.Sicily’s own recording of PHI numbers.

\bigskip

3.3. Identification and Collections

The traditional focus of epigraphic study upon the text, rather than the epigraphic support, means that epigraphic
publication in the past has frequently been relatively limited in the information which it has recorded about the
object on which the inscription is inscribed. This is a familiar complaint, and one which I.Sicily will address
wherever possible through full object description and a rich photographic record. However, a corollary of this general
problem is a very low level of information regarding current location and in particular the infrequent recording of
museum inventory numbers or similar information. This situation is inevitably exacerbated by the substantial (and very
positive) reorganisation and redevelopment of museum collections in Sicily recent decades – including a significant
increase in the number of museums and public collections.

I.Sicily is making use of the TEI {\textless}msIdentifer{\textgreater} element, with its associated sub-elements in
order to record details of institutional collections and inventory numbers wherever possible.\footnote{
www.tei-c.org/release/doc/tei-p5-doc/en/html/ref-msIdentifier.html [accessed 26.09.2015].} In order to maximise the
value of this, we have adopted two further courses of action. In the first place, as part of the larger ambition of
undertaking autopsy of every stone contained within the corpus, we are working in close collaboration with museums on
the island to improve our records of individual museum holdings. Where possible we aim to include associated archival
information, such as copies of inventory records. This work currently includes a major sub-project to catalogue the
epigraphic collection of the Museo Archeologico Regionale Paolo Orsi at Siracusa, and we are also currently working
with collections at Adrano, Halaesa (Tusa, ME), and Catania.\footnote{ We gratefully acknowledge the ongoing support of
dott.ssa G. Lamagna and dott.ssa A.M. Manenti at Siracusa, as well as previous directors of the Museo Archeologico at
Siracusa, dott.ssa C. Ciurcina and dott.ssa B. Basile; of dott.ssa A. Merendino at Adrano; of dott.ssa G. Tigano and
dott. R. Burgio at Messina; and of dott.ssa M.G. Branciforte at Catania.} It is hoped that this work will be of
considerable value to the museums themselves, since access to the I.Sicily records should facilitate the curation,
display and accessibility of the inscriptions (see below also on translations), and we welcome future collaboration
with other museums on the island.

Secondly, in collaboration with Dr Michael Metcalfe, I.Sicily has developed a database of Sicilian archaeological
collections (130 at the time of writing). This database is mounted online alongside the epigraphic corpus, in a
searchable format, including map-based searching. In order to facilitate the generation of linked data, the individual
museum records will be maintained with URIs, of the form:

\ http://csad.ox.ac.uk/sicily/museum/SicMus123

The linking of the epigraphic and museum databases will enable the searching and reporting of inscriptions by museum
collection as well as the easy locating of the appropriate collection.

\bigskip

3.4. Location, location, location

I.Sicily is actively generating rich geo-data for the individual inscriptions, both for the original findspot/provenance
and the current location (whether museum-based, on-site, or elsewhere), and we aim to provide map-based searching for
inscriptions, as well as text-based searching by ancient and modern place-names. In addition to full listing wherever
possible of both ancient and modern place names for epigraphic provenance, we are working to provide detailed location
information for each find-spot and the inscription’s current location, through a combination of library and map-based
research and the use of autopsy and GIS recording. At present geo-data is being recorded in two forms, both through the
use of explicit geographical locations in the form of longitude and latitude records in decimal degree form (using
{\textless}geo{\textgreater} elements), and through the use of Pleiades URI references wherever possible.\footnote{
http://pleiades.stoa.org/ [accessed 26.09.2015].} We are committed to the long-term use of Pleiades as our primary
reference for ancient places, and to that end we aim to update and improve the Pleiades data for Sicilian locations, in
particular name data and sub-locations, in conjunction with the editing of the I.Sicily records.\footnote{ See e.g.
Wilson et al. \ 2015. Valeria Vitale (KCL) is currently undertaking a significant programme of data improvement in
Pleiades on behalf of I.Sicily; we are grateful to Tom Elliott and Jeffrey Becker for their support.}

\bigskip

3.5 Searching

In order to support the aims outlined above, I.Sicily has taken a different approach to search and browse. Although
standard form-based search with paged results, like that of Google, makes sense for very large result sets, the
comparatively small number of records in I.Sicily (thousands versus the estimated 30 trillion web pages indexed in
Google) lends itself to a more direct and interactive approach - a spreadsheet/grid model (similar to Microsoft Excel)
that runs directly in the browser. Although it is tempting to repeat the standard web-form model, following the
argument that that’s what users expect, the spreadsheet approach will be much easier to use, narrowing quickly and
accurately to more easily interpreted results. Further, any subset of the spreadsheet, generated from interactive
filtering, can, with a single button push, be exported to CSV (comma separated values) for use outside I.Sicily. The
spreadsheet interacts particularly well with maps: all findspots or museums in a filtered subset of the grid can be
simultaneously shown on the map (see Fig. 2). The spreadsheet model also provides a very quick and intuitive (since so
many people are familiar with spreadsheets) means for editing records (in this case, inscriptions and museums) online.
This web-based spreadsheet model has only recently become feasible for the web, as web browsers have added more
functionality and new javascript libraries have been developed.

\bigskip

3.6. Translations

As was extensively discussed at the first EAGLE conference (Paris 2014), the creation and availability of translations
is a major goal of the EAGLE project and its collaborators, and I.Sicily is no less committed to that
ambition.\footnote{ See Orlandi et al. 2014: Part II.} Translations are rarely available for any of the published
Sicilian inscriptions.\footnote{ French translations appear in Dubois 1989 and Dubois 2008.} It is obvious that the
inclusion of translations will make the material much more accessible to a wider audience both of students and the
general public. Equally, provision of translations will add to the value of the database as a resource for museums and
others curating the inscriptions recorded in the database. To that end, a long-term ambition of I.Sicily is to include
translations wherever possible in both English and Italian. We see this as one obvious area where public contribution
(‘crowd-sourcing’) will be invaluable (see below).

4. Limitations and future ambitions

The scale of the enterprise, and the available resources, mean that in its current form the project has limited itself
to inscriptions engraved on stone (the coverage of rupestral inscriptions/graffiti and of inscriptions painted on
stone/plaster is regrettably uneven). However, there is no reason in principle not to extend coverage in future to
include inscriptions on other materials. Similarly, although as noted above the current project does not include a
programme to mark up linguistic features of the texts, the commitment to the long-term maintenance of the corpus and
the open availability of the underlying XML records means that such a project would be entirely possible in the future.

A core principal of the project is that wherever possible an inscription record should be supported by recent autopsy
and not simply derived from the existing literature. Necessarily, this process is a slow one, and the majority of
records at this stage consist of information derived from secondary sources (earlier editions and other publications).
Individual inscription records will contain a clear indication of the editorial state of the record (from unchecked
through to fully edited) and additionally whether the record is underpinned by autopsy. In both cases, clear records
will be kept of editorial responsibility, autopsy and authorship as appropriate. In order to speed up the development
of the corpus, and to encourage those working on the material to take ownership of it for themselves, we aim to enable
individuals to submit new records and emendations or additions to existing records (such as translations, images,
location information), both in the Epigraphic database and the Museums database. To this end, we welcome collaboration
with those undertaking epigraphic projects in Sicily, and aim to offer the ability for other projects to publish their
editions through I.Sicily.\footnote{ We are currently establishing just such a collaboration with Prof.ssa M. Sgarlata
and Drs Lo Faro and Gradante, in support of their project to produce the volume of Inscriptiones Christianae Italiae
for the San Giovanni Catacombs in Siracusa.} We are also exploring the potential of the corpus as a teaching resource
both for epigraphy in general and for the teaching of EpiDoc. This latter aspect has already been initiated through a
Teaching Project Award (2014-2015) from the Humanities Division of the University of Oxford, and we aim to develop this
further in the coming year, as part of the work of incorporation and conversion of texts into the existing dataset.

It is our long-term ambition that I.Sicily might become the default location for the publication and dissemination of
Sicilian inscriptions; in the shorter term, we hope that it will serve as \ valuable portal in the world of Sicilian
epigraphy and of ancient world open linked data, greatly improving the accessibility of Sicilian epigraphy and so
enriching the study of the ‘crossroads of the Mediterranean’.

\bigskip

\bigskip

 \includegraphics[width=7.197cm,height=18.653cm]{EAGLE2016ISicilyfinalcopy-img001.jpg} Fig. 1: sample edition.

 \includegraphics[width=10.881cm,height=7.966cm]{EAGLE2016ISicilyfinalcopy-img002.png}

\bigskip

Fig. 2. Screenshot of map-based searching (museum locations) and of part of the spreadsheet/grid search interface
employed in I.Sicily.

\bigskip

Bibliography

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

[Warning: Draw object ignored]I.Sicily gratefully acknowledges the financial support of the John Fell Fund of the
University of Oxford, and of the Warden and Scholars of the House or College of Scholars of Merton in the University of
Oxford.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/33/TIGLIO.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{TIGLIO. Translations and Images of Greek and Latin Inscriptions Online}

%%%%%%%%
%AUTHORS:%
%%%%%%%%
\author[PERSEIDS]{Almas Bridget}
\author[DC3]{Baumann Ryan}
\author[PERSEIDS]{Beaulieu Marie-Claire}
\author[DC3]{Cayless Hugh}
\author[UHEI]{Cowey James}
\author[UHEI]{Liuzzo Pietro\corref{first}}
%\author[AIO]{Lambert Stephen}
\author[AIO]{McCourt Finlay}
\author[DC3]{Sosin Joshua}

\cortext[first]{Corresponding author. Email: pietro.liuzzo@zaw.uni-heidelberg.de}

\address[UHEI]{Ruprecht-Karls-Universität Heidelberg - Marstallstraße 6, 69117, Heidelberg (DE)}
\address[AIO]{Glasgow University - Attic Inscriptions Online (UK)}
\address[PERSEIDS]{Tufts University (USA)}
\address[DC3]{Duke Collaboratory for Classics Computing (USA)}

\begin{document}

\maketitle

\begin{abstract}
This paper describes the aims of a project funded by The Andrew W. Mellon Foundation to speculate on the best ways to deal with two forgotten types of content in the realm of ancient epigraphy: translations and images. Translations available online are less then 3\% of the total available transcriptions online; images are often subject to policies which make extremely difficult their use for research, publication, even simple viewing. There has been no thought given to these before in an articulated manner although these are the types of content which can bring a much larger group of users to epigraphy.   
\end{abstract}

\keywords{Translations, Images, Greek and Latin, Epigraphy, Online}

\section{Introduction}\label{sec:intro}

Nearly all Greek and Latin epigraphic texts are available, sometimes in multiple versions, after three decades of continuous digitization and online publication. Without counting repeated inscriptions and \textit{instrumenta}, there are about 300.000 Latin inscriptions\footnote{EAGLE disambiguated total: 235.626 ; EDCS not disambiguated total without \textit{instrumenta} and \textit{inscriptiones christianae}: 308.581. On the definition of Roman Epigraphy, see \citet{Panciera2012}.} online and about 200.000 Greek Inscriptions\footnote{Data from the latest Integrating Digital Epigraphies's (\url{https://youtu.be/OPfDj_hjeok}) harvest from the Packard Humanities Institute, Searchable Greek Inscriptions (\url{http://epigraphy.packhum.org/}), with some duplicates, 207.964.}. In some databases there is also abundant metadata and  a structured bibliography. The situation for translations and images of these inscriptions (text and support) available in the digital space is nevertheless quite different. The ratio between images of inscriptions and text is of one image every two inscriptions,\footnote{At the time in which this paper is being written (\today) the Epigraphic Database Heidelberg has ca. 35.000 photos and ca. 71.000 texts (0,5); The Epigraphic Database Roma has a slightly better ratio with 45.000 photos for 71.000 texts (0,6); The Epigraphic Database Bari has 34538 texts and 10341 images (0,3). In the EAGLE aggregator, the total ratio (excluding the related content of Arachne), is of 0,79 images for each text (235626 documental entities per 185999 visual entities) because smaller corpora tend to have a better photographic documentation.} but those inscriptions which have photographic documentation usually have many photos.\footnote{The Epigraphic Database Heidelberg has to date ca. 14.000 records with a photo or a drawing attached, bringing the average number of images per inscription to 2,5.} Translations are present only in smaller corpora edited online in most cases and the Attic Inscriptions Online project\footnote{\url{https://www.atticinscriptions.com}} is an \textit{uniquum} in its intent to provide mainly translations of inscriptions.\footnote{\citet{Lambert2014}} There are many publications offering translations in print, but these are not published online. An estimate calculation, based on the 11.000 translations present in the EAGLE Media Wiki, and known collections of printed translations of inscriptions, compared to a total of texts usefully translatable of around 300,000 texts, brings to an alarming 10\% of translated texts, of which only a third (slightly more than 3\%) is online. Translations are perhaps not a priority for researchers who know Greek and Latin, but are a way to clarify the interpretation of a text and an invaluable tool for didactical purposes and teaching: they are the only way in which an inscription can reach a wider public in a significant way as part of cultural heritage. The same could be said for images, even more obviously, since researchers also need them because: 1) they cannot always reach the place where an inscription is stored to study it  (given that the inscription is still there); 2) there are cases  in which a photo might be all we are left with and these are quickly increasing as monuments get lost or are destroyed. The imbalance in the documentation is thus pressing, since translations and images are our two best controls on the constitution and interpretation of ancient documentary texts. To an extent, digital epigraphy today is the direct descendant of epigraphy’s 19th century analog self: many texts, few translations, few images. This project aims to take initial steps to redress that imbalance, building resources that allow epigraphists and ancient historians to bring translations and images more closely into the suite of existing digital epigraphy resources.

\section{Problems}
Let's look at the data we have. The EAGLE project has gathered some insights on small collections of images of inscriptions openly published online, on Wikimedia Commons,\footnote{\url{https://commons.wikimedia.org/wiki/Category:Media_contributed_by_EAGLE}} and on a set of translations of inscriptions, collected in the EAGLE Media Wiki.\footnote{\citet{Liuzzo2014} \url{http://www.eagle-network.eu/wiki/index.php/Main_Page}} We shall compare their impact and reach on the wider public to that of the texts of inscriptions, to underline the urgency for these materials to be produced also in order to bring epigraphy outside its restricted circles.
Let's look at the visits to the Epigraphic Database Clauss-Slaby, the largest collections of texts (with minimal metadata and no directly stored image): EDCS has an average 3000 requests per day.\footnote{This can be easily monitored looking at the counter on the website at the end of each day. No better statistics are available.} The result page is always one, containing all the results from the database, which has a total of 491353 texts. We have no means to provide better data unfortunately but for the comparison these will be enough. The images collected under the category ``Media Contributed by EAGLE'' on Wikimedia Commons contains instead around 8000 photos of inscriptions and we have some good insights on this data\footnote{Thanks to user:Fae and the authors of the BaGLAMa 2 tool. See \url{https://commons.wikimedia.org/wiki/Category_talk:Media_contributed_by_EAGLE/reports} and \url{https://tools.wmflabs.org/glamtools/baglama2/\#gid=148\&month=201508}}. These photos have been viewed in 19 months 22,236,085 times. Another interesting information is the number of people who have worked on them, by no means only members of the EAGLE project: 7 users have made more than 1000 edits, which could be anything above the figure; 20 have made between 100 and 1000 edits; 71 have made between 10 and 100, and even more interestingly 581 have made between 1 and 10 edits. This is a critical mass of active users, uploading, editing, curating, using data they are interested in. 
The same situation can be noted for the 11.000 translations in the EAGLE Media Wiki, which were viewed in 18 months 1.380.000 times and have seen 280 active users, who have at least made 1 edit. The tool is not well known outside the EAGLE consortium and is a very small prototype, but the fact that it has already attracted such a mass of views is significant. What would happen if we gave 100.000 translations in the way we have given them to the people in the Mediawiki, completely openly? What would happen if those people contributing to images and translations were empowered to operate easily and intuitively to enter more and more data? Inscriptions will never get as many fans as we would like to, but perhaps their content and related resources would be a bit more accessible to non-initiated. 
This comparison confirms also that the usability of resources is measured at a different level when they are made open, and that images and translations have an undeniable higher relevance as an online resource, thus attracting interest also to the transcriptions, while this does not happen the other way around and only those who know what they are looking for will stumble upon an ancient inscription published online. Nothing new in these observations, but this obvious observation is in contrast with the actual situation in which photos are few and translation even fewer. Why so, then? The possible reasons are:
\begin{enumerate}
\item the lack of an entry point which is easy to access and use
\item people get easily worried by copyright due diligence and find diffucult sometime to trak back who is the author or the copyright owner of a photo
\item a lack of coordinated effort, planning and management of the storage of both these types of content
\item researchers working on inscriptions identify their intended audience in a very specific academic community which does not need translations and instead needs edited texts (transcriptions and metadata). 
\item publication of content with (sometimes unnecessary) restrictions
\item lack of time for this effort, unrecognized in academic settings as a contribution to the progress of knowledge, as, sadly, most other digital efforts
\end{enumerate}

We shall point out what has been done to solve these problems and cater for an improvement in this part of documentation and production of online content in the future.

The international group of partners, which includes University of Heidelberg, University of Cardiff, Duke University and Tufts University is holding regular workshop meetings to design and  develop a suite of resources that support generation of epigraphic translations, with peer-review and publication workflows supported by Perseids’ extension of the Son of the Suda On Line code (SoSOL), with publication supported by the EAGLE Mediawiki, and image management, reference ontology, geo- and other services, supported by Integrating Digital Epigraphies, and with Attic Inscription Online translations as the key content stream for development and testing.

\section{Translations}
To face these challenges with regard to translations, it is the opinion of the project team that tying together existing resources is a better way to tackle the issues rather than trying to superimpose a new tool or system. The available building blocks for such systematization of existing resources are the following:

\begin{itemize}
\item the existing local data entry point of Attic Inscriptions Online, in the process of moving to TEI-EpiDoc markup for the underlying data
\item the EAGLE Mediawiki, with the Wikibase Extension, which collects translations from several users as a part of the work of the EAGLE consortium to bring epigraphy to a wider public 
\item the Perseids peer review system,\footnote{\url{http://sites.tufts.edu/perseids/}} which uses the Son of the Suda Online\footnote{\citet{Baumann2013}}
\item Leiden+,\footnote{\url{http://papyri.info/docs/about} and J. Sosin presentation at \url{http://www.stoa.org/archives/1263}} a simplified markup which allows the use of normal diacritics instead of tags to enter XML markup. 
\item identification and disambiguation done by content providers (the epigraphic databases) and by projects such as Trismegistos\footnote{\url{http://www.trismegistos.org/}} for the members of the EAGLE consortium and IDEs\footnote{\url{http://blogs.library.duke.edu/dcthree/projects/}} for Greek Epigraphy projects.
\item referencing and resolution services provided by IDEs which do not just align content relating to one resource but describe the relation among them
\end{itemize}

These consider two kinds of users: 
\begin{itemize}
\item users involved in a project with access to a data entry point in a database (using XML)
\item independent contributors 
\end{itemize}

With the available building blocks what can be done currently for translations is:
\begin{itemize}
\item a standalone javascript library to enter translations using Leiden+ (to be implemented and tested in AIO, as the best candidate for its focus on translations) which implies:
	\begin{itemize}
	\item an enlargement of the encoding guidance and conversion capabilities of Leiden+ to EpiDoc for translations.\footnote{\url{https://github.com/TIGLIOPROJECT/documentation/wiki}}
	\item recommendation on how to mark up translations for the EpiDoc guidelines
	\end{itemize}
	These developments will hopefully be beneficial to the EpiDoc users community as well, which has in the past asked for more guidance on how to encode translations. 
\item An export of AIO in EpiDoc to the Perseids platform in which translations will be peer-reviewed for ingestion in the EAGLE wiki, bypassing the harvest process for EAGLE. This might be useful as a use case for future project willing to publish their translations with the others collected into the EAGLE Media Wiki.
\item facilitate flow between existing tools and services 
	\begin{itemize}
	\item EAGLE and Perseids worked together in the past years in order to integrate the two services offered, but this had a number of limitations, e.g. the requirement for a translation to be already present in the wiki in order to be able to publish another one via Perseids. This reduced the number of items for which the integration could be used to a minimum, forcing the use of workarounds as mock text and placeholders. The integration of the EAGLE wiki with Perseids will now enable users to enter translations for any inscriptions and even a new translation from scratch with a specific new language, thus covering all possibilities for the EAGLE Mediwiki user. This requires nevertheless:
	\begin{itemize}
	\item unique identifiers for Latin and Greek texts, which are currently provided for the first by Trismegistos, and for the second group of texts by IDEs.
	\item a citation URN structure which is agreed upon and otherwise usable. This will be based on the scheme already in use for EAGLE built on CTS URN syntax as in the example: 
	\begin{quotation}
\footnotesize{urn:cts:pdlepi:eagle.tm12345.perseids-translation-1}
\end{quotation}
where the structure is 
\begin{quotation}
\footnotesize{urn:cts:namespace:textgroup.work.version}
\end{quotation}
	\end{itemize}
	\item The complete workflow from data entry to publication in a website and to the common EAGLE resource via Perseids will then be a complete and replicable workflow, scalable for use from larger projects and documented to guarantee easy and sensible connection of the resources online. 
	 \end{itemize}
\end{itemize}

The workflow for the connection of new translations and images to existing online content will be then facilitated in this way. 
A project with its own data entry interface should be able to use the javascript library to enter translations using Leiden+ and following the conventions already extended and public in papyri.info. They should then be able to identify with a TM or IDEs id these translations and infer a URN to push these directly to the Perseids system. Here the translations will be peer-reviewed and then sent both back to the source with an approved status and to the EAGLE Media Wiki. If this database is partner of the  EAGLE project the texts will be harvested separately and the translations linked back from the Media Wiki. From the Media Wiki API they will also be available as such to external users. 
An independent contributor instead will be able to look up the TM text id or IDest of the inscriptions he wants to translate and enter it to the EAGLE Media Wiki. From here this will be sent to the Perseids system and returned as described above.

The short term goal is therefore to stitch together existing resources already in development, end especially to provide ids and a clear citation syntax for all available inscriptions, which will have counter benefits also for any other digital project with these requirements.

\section{Images}
Most of the existing images of inscriptions are currently safely stored in private computers. Large collections of images of inscriptions are available at the major databases, and can count up to thousands of photos of inscriptions or drawings. Large collections of images are also on Flickr (e.g. Visible Words) and Wikimedia Commons (CIL and AE categories provide a good overview of what is available).

The major problem preventing publication is that people do not know if they can share the images which they have. The copyright regulations are too complicated and possible contributors opt for doing nothing instead of taking any unknown and unwanted risk. 
Storage of images of inscriptions in the major databases happens under very strict conditions of reuse and publication and while it is the best possible way to operate for these projects, there can be no mirroring or distribution of the resources available so that the life expectancies of content curated for decades is tied to  the lifetime of these databases. A test has been done with the images of non identified items in the Epigraphic Database Heidelberg, uploaded to Wikimedia Commons and users have contributed to identify a number of those. 
Uploading photos on Wikimedia Commons requires an open license, which cannot always be guaranteed, whereas on Flickr it is possible to retain rights whilst publishing the photos, so that it makes a nicer tool for this kind of content, although it does not allow community editing and batch upload, which is instead possible via tools developed by Europeana and the Wikimedia Foundation for Commons.\footnote{The GLAM Wiki toolset, \url{https://commons.wikimedia.org/wiki/Commons:GLAMwiki_Toolset_Project}.}
A unique repository or a connection hub for all these photos would be a solution to keep the content under sight but this would still require control over resources daily published in Flickr and Commons, an involvement into the communities of users online, together with the extension of citation structures to these groups. There is no easy solution for the copyright side of the problem, but a continued encouragement to share will build towards the critical mass needed for a change in perspective on this issues in the coming few years. During this project we will try to list requirements for a tool to ease out the upload of images online, which will match images and metadata provided in various formats, suggesting ids and adding them in a format compliant to the citation scheme agreed. It is in fact true that the amount of time required to use tools which ask for a one to one upload of images is another important factor which reduces the amount of content shared even from those willing to do so.

\section{User involvement and expert sourcing}
The problems of all digital projects looking at putting together materials from various sources and contributors are on one side to get people involved, on the other to overlook the work done and take care of the administration.  
The possible user scenarios are eventually many more. Two especially deserve to be mentioned here. Most input has been provided in term of new translations in the EAGLE Media Wiki during Workshops and Secondary Schools class work. Some teachers with a background in epigraphy have been contacted by the Epigraphic Database Rome to start and experiment with a didactic model which would include translating inscriptions. The students worked on a specific corpus of inscriptions, studied the text and the support and produced a translation which they entered in the EAGLE Media Wiki with the supervision of their teacher. This experience has proved successful for the students which have seen their contribution directly where it should be, together with other scholarly content. On the other side, every new translation matters in such a state as the one described above. 
The second example is the work done in two consecutive workshops, held in Ercolano and at the Centre for Hellenic Studies in Washington DC of the Ancient Graffiti Project,\footnote{\url{http://ancientgraffiti.wlu.edu/}} which has published multiple translations for all the graffiti of Herculaneum. In this context the usability of the Wikibase software has been tested and it proved to be an extremely intuitive and powerful tool. It takes very little explanation, but there are caveats for this simplicity and namely that it is extremely easy to do things in slightly creative ways, as entering statements as source information or typing an id in a slightly different way, which then need to be monitored and fixed.

\section{Conclusions}
Some of the tasks above have been already carried out, some are under way, but there are some general conclusions which can be summarized. There is a need to unify the resources, agree on standards for reference and citation and provide stable identifiers and citation structures, providing a comprehensive list of epigraphic publications with the relevant abbreviation in use. Although some work has been done, the amount of data makes this task continuously needed together with that of disambiguation. Other efforts need to go in the direction of flexible but harmonized standards for encoding and working on data entry giving priorities probably in a  slightly different way as before, updating the tools to be able to cope. More generally, while tools are abundant and so are guidelines and cookbooks, an agreed venue for coordination of the efforts is still a desideratum, and should include not only scholarly project but also community based efforts such as those of the Wikimedia Commons users and of the Flickr user's groups. These people could be also part of the peer review process, thus contrasting the side effects of an inactive board.\footnote{The last meeting of the project, dealing with images and CTS URN structure still had to take place at the date of submission of the present contribution.}

\section*{Aknowledgements}
The Andrew W. Mellon Foundation has funded the meetings of the project TIGLIO and the work which will be carried out for Attic Inscriptions Online. We would like to thank also all the Wikimedia users who have contributed to parts of this work in various ways and especially Aubrey, CristianCantoro, Fæ, Laurentius, Magnus Manske, Sannita and Wittylama. Another special aknowledgement goes to all users of the EAGLE Mediawiki and to all the volunteers working on photos of inscriptions online.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/36/DatabasesToEpiDoc.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Mapping Epigraphic Databases to EpiDoc}

%%%%%%%%
%AUTHORS:%
%%%%%%%%
\author[UHEI]{Liuzzo Pietro\corref{first}}

\cortext[first]{Corresponding author. Email: pietro.liuzzo@zaw.uni-heidelberg.de}

\address[UHEI]{Ruprecht-Karls-Universität Heidelberg - Marstallstraße 6, 69117, Heidelberg (DE)}

\begin{document}
\maketitle

\begin{abstract}
The Europeana network of Ancient Greek and Latin Epigraphy brings
together most repositories of ancient epigraphic material and aims to
provide historians not just with a useful research tool, but a
curated online edition which has high quality content as well as high
quality data.  In this paper some of the up-conversion, alignment and enrichment tasks are presented.

\end{abstract}

\keywords{EpiDoc, XML, Vocabularies, Linked Data, Data harmonization}

\section{Introduction}\label{sec:intro}

The Europeana network of Ancient Greek and Latin Epigraphy\footnote{\citet{OrlandiCasarosa2014} \citet{Orlandi2014} \citet{liuzzo2014} \url{http://www.eagle-network.eu/}} brings
together most repositories of ancient epigraphic material and aims to
provide historians not just with a useful research tool, but a
curated online edition which has high quality contents as well as high
quality data. Towards this end many steps are needed as the databases on which many institutions have worked for decades cannot be just discarded. Epigraphy enjoys a special position among digital resources as most transcriptions of existing classical inscriptions have been digitized. Many duplicates exist and a lot of the content lacks metadata almost entirely, but the work continues with dedication so that among the 10 \% of digitised European Cultural Heritage, Ancient Epigraphy can certainly claim to play a strong role.\footnote{The data is taken from the Europeana 2020 strategy report available at \url{http://strategy2020.europeana.eu/}.} While the long term aim for epigraphy is on the one hand to keep up with new finds and studies and on the other to have a common and flexible entry point and backend on which to work together, this is not easy to accomplish without a careful process which involves a gradual transition from independent databases (the best option at the time when the EAGLE databases were created) to a common online data entry and editing system working on XML files in EpiDoc, as it has now become feasible.   
In this paper some of the mapping and content harmonization efforts undertaken by the EAGLE BPN to achieve this step towards a common epigraphic resource, will be presented as a case in which this apparently technical task involved deep revision of contents related to the discipline and required discussions and collaborative efforts by the working
groups of digital epigraphists, encoders, epigraphists and developers, coming from several different background and experiences. The result of this work are the automated conversion to TEI EpiDoc XML of 90\% of the text features in the EAGLE databases and the EAGLE Vocabularies, the largest existing controlled and aligned vocabularies for classical epigraphy.\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}}

\section{Why XML}
Choosing XML for an online publication is almost obvious given the current stage of development of the editorial methodology in epigraphy. However there is room for a brief explanation of why this is preferable on a large scale to a relational database. I would like to suggest a comparison here. Coffee is produced by filtering water through a powder more or less thick. This can be accomplished by pushing water through the coffee (as in most systems, from a mocha to an espresso machine) or coffee through the water (as in a french press, where the coffee floats and is pushed down). There is pretty much the same difference when we compare the action of entering data into a database, i.e. a rigid structure of information, and marking up a text, adding tags to the text itself. The first process forces data through a grid however well thought off, while in the second case structure is added on the contents without altering them, but just enhancing their potential so that the richness and values of both contents and structure empower each other. There is just more taste and much more freedom of encoding to the level desired and to the complexity one wants. An additional advantage is the level of interoperability. An XML file is virtually software independent, it is not an Access, FileMaker or whatever database format. A third reason is that an XML file can host a complete textual description with semantic markup and still have the structure of a normal text edition, while the same source can be used to produce a database or an index for interrogation purposes. For example the same markup of an abbreviation \texttt{<expan><abbr>C</abbr><ex>aio</ex></expan>} can be used to produce a database of abbreviation marks and expansions, to output in a diplomatic edition only the character actually on the stone or to print the expanded abbreviation in a critical edition.\footnote{\url{http://www.stoa.org/epidoc/gl/latest/trans-abbrevmark.html}} To print a database to a book would be at least more cumbersome. XML provides also a better service as an archival format being explicit on the values it uses to describe contents, moreover if these follow a schema of agreed value of such elements. This said producing and sharing XML files means that more quality editions can be produced, printed editions can be produced and all sort of outputs can be supported with flexibility; it means that data is safe from software developments and support; it means you don't have to put information into a rigid table which is then hard to restructure, but that you can annotate your text as much as you want thus adding layers and layers of ``possible databases'' of relevant information to be extracted and used.   
The text is what one uses to describe situations and complexities, the markup allows also the machine to know and retrieve that information as in a database: you don't need a field or an element to state what the orientation of your text is, you can simply state and describe as you would do in a normal publication and if you want to reuse information about the text orientation then you can add elements and attributes to do so. XML allows not to worry too much on ``what goes where'' or about ``where do I put that info" because it is not a database, so that the editor can instead focus on ``what is that'' and on accurately describing it. The TEI specification which is EpiDoc allows the specific epigraphic content to be described fully for what concerns the text of inscriptions, drawing on more than a decade long experimenting and development of this standard thanks to a vast and active international community. The simple fact of using the same markup to describe our texts allows us to study them together and to produce and run software for more then just one corpus of inscriptions.

\section{EAGLE Mappings and Metadata modeling}
The EAGLE BPN has taken 18 different content providers and has come to
the decision to take the first steps to bring hundreds of thousands of
ancient Greek and Roman inscriptions to TEI. This required 14 mappings
of local database metadata models to TEI/EpiDoc, as well as the elaboration of
XSL Transformations\footnote{\url{https://github.com/EAGLE-BPN}} for up-conversion\footnote{This is the terminology used by \citet[906]{Kay} to describe transforming without explicit structure in data which has it. In this case it is a transformation from a less explicit database structure to a fully explicit and interoperable one.} and the alignment of the text
markup. Although this was thought to be a trivial task it turned out
that whereas databases claimed to use the same conventions, a considerable amount of differentiation took place due to the data structure and to data entry procedures as well as to policies and internal decisions.

The main harmonization task undertaken has been to align the XML format
of data provided for aggregation and ingestion, to the TEI
specification EpiDoc.\footnote{\url{http://www.stoa.org/epidoc/gl/latest/}} This well established TEI schema, broadly used in
many high quality projects, allows for a very easy alignment, for the
production of an XML file compliant with international standards and
for high flexibility for integration of the vocabularies and places
gazetteer in use.

Moving from a database to a marked up text is a partially mechanical operation which involves a theoretical jump. As coffee is better when water goes through coffee powder rather than when coffee goes through the water (as in a french press), a database has fields to be filled, while marking up a text is a descriptive activity which is attached to information whichever textual form it takes.\footnote{See above.} Mapping from a database to XML forces into the XML a structure and a logic which is that of a database, whereas the freedom and flexibility achieved with XML are yet to be actually realized and exploited.

In order to offer a complete and critically structured endpoint to the user on the side of data, to describe inscriptions and their representations, EAGLE considered beside TEI also CIDOC CRM, studying and providing a full EpiDoc to CIDOC-CRM mapping.\footnote{Still under development, within the ARIADNE project. \url{http://www.ariadne-infrastructure.eu/}. The need for a working group and a variant declension of CIDOC for epigraphy have been highlighted during the Nicosia EAGLE meeting in March 2015.} This would have enabled a further full description in a different logic, that of the web of data, but the attempts made proved the need for more specific efforts.

There are several mappings involved in the aggregation work of EAGLE. Content providers need to map to a common Eagle Metadata Format, and the data produced will then be mapped to the Europeana Data Model\footnote{\citet{D3.1}} in order to be aggregated in Europeana, for a wider dissemination with a special eye for the general user.

\section{EAGLE Metadata Model and Harmonization}

The occasion of a mapping work allows also for other tasks of curation to be performed.
EAGLE Members had in some case the text of a same inscription with three parallel different types of encoding conventions applied.

\begin{figure}[htbp] \includegraphics[width=\columnwidth]{Contentcuration.png} \caption[]{Different Texts before and after harmonization though up-conversion} \end{figure}

In the context of the mapping to EpiDoc of the metadata, the text also underwent transformation with tools which have been developed to support the alignment and harmonization of data from content providers to international standards for what concerns digital editions of inscriptions.

Given the template described in Part III, and ANNEX II of the EAGLE Metadata Schema (\citet{D3.1}), an XSL Transformation converts from string epigraphic texts in marked up TEI-EPIDOC XML, following the \href{http://www.stoa.org/epidoc/gl/dev/}{EpiDoc guidelines} (\citet{Elliott2007}).

These XSLT:\footnote{Based on Chetc.txt (by Hugh Cayless, Elli Mylonas, Gabriel Bodard and Tom Elliott) and further support from the Epidoc Collaborative (especially from Gabriel Bodard)}

\begin{enumerate}
\item allow the conversion of epigraphic texts with various encodings and conventions from string to Epidoc markup and valid against the The EpiDoc RelaxNG schema.
\item Populate appropriate elements with available common URI from the EAGLE vocabularies\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}}

\end{enumerate}

This export set up will also guarantee that contents are kept aligned to the EpiDoc guidelines at all stages guaranteeing an effort free alignment to these international conventions for partners who can continue to apply local conventions for editing. But I would like to give more details on the steps of this process as an example of how it was possible to extract semantics, entities, and patterns from these text while aligning metadata format to an internationally recognized standard.
\bigskip

\subsection{Step 1}
Each project uses different conventions and therefore the regular expressions used to match particular situations are different.
The process of mark up of the string text in div[@type="edition"] is accomplished in several steps to guarantee consistency and precision.

The \texttt{textstructure.xsl} looks for marker of different sections and tokenize them to apply the same XSLT to each section of the text which needs to be contained by an <ab> element. If there is only one part it applies following instructions to that only.

Each section of text is then processed by the \texttt{brackets.xsl}. Normalizing brackets is important for the following steps and splits individual semantic values. The notation [ort 3], which would mean that a supplied text is followed by a gap of three letters, is divided into [ort][3].

The normalized string which results from this process is then passed to the 
\texttt{up-conversion.xsl} which works using a specific operation to search for regular expressions patterns (\texttt{xsl:analyze-string}) and substitute them with correct xml elements.

Running the transformation on the pattern {\textless}E=F{\textgreater} will return the following result:

\begin{verbatim}

<choice><corr>E</corr><sic>F</sic></choice>

\end{verbatim}

The result of this template is then passed to a further template which gives consistent numbers (\texttt{insertnumbers.xsl}). Empty lines do not need to have numbers, so Xpath is used to evaluate where to put a 0 as value of the @n in the {\textless}lb{\textgreater} element. Starting from this

\begin{quotation}
{}-{}-{}-{}-{}-{}-] / e[t?] Q(-{}-{}-) Bl(a)e[sus?] / contub/ernalis / eius / d(e) s(uo) l(ibens) l(aetus) d(edit)
\end{quotation}

The result of this processes is then the following\footnote{The amount of feature supported is much higher and some example of complex text successfully converted can be seen in this presentation online}

\begin{verbatim}

<ab>
<lb n="0"/><gap reason="lost" extent="unknown" unit="line"/>
<lb n="1"/>e<supplied reason="lost" cert="low">t</supplied> 
<abbr>Q</abbr> 
Bl<expan><ex>a</ex><abbr>e</abbr></expan><supplied
reason="lost" cert="low">sus</supplied>
<lb n="2"/>contub
<lb break="no" n="3"/>ernalis
<lb n="4"/>eius
<lb n="5"/><expan><abbr>d</abbr><ex>e</ex></expan> 
<expan><abbr>s</abbr><ex>uo</ex></expan> 	
<expan><abbr>l</abbr><ex>ibens</ex></expan> 
<expan><abbr>l</abbr><ex>aetus</ex></expan> 	
<expan><abbr>d</abbr><ex>edit</ex></expan>
</ab>
\end{verbatim}

\subsection{Step 2}
On the elements which contain information such as Object Type, Material, Execution Technique, etc. which are typically handled with a controlled vocabulary a series of XSL transformations is passed, one for each vocabulary related to that element  to match the content of the element with the vocabulary entry into the SKOS vocabulary stored in git, regularly updated and published as a self standing resource on the EAGLE portal.\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}} What follows is a brief description of such vocabularies and their development.

\begin{figure}[htbp] \includegraphics[width=\columnwidth]{upconversion.png} \caption[]{Stages of transformation of a text: database, template, up-conversion, checking and display with EpiDoc XSLT.} \end{figure}

\section{Classification problems: the EAGLE Vocabularies}
As Di Stefano Manzella \footnote{\citet[109]{Manzella1987}} clearly explained classification is no easy issue in any field: epigraphy is no exception to this rule. Traditionally the CIL VI (Rome) classification has been used as a reference, as this typology has served as a model for all epigraphic production in the Roman Empire. There are nevertheless new glossaries and classification curated by \href{http://cil.bbaw.de/cil_en/dateien/glossar.php#auswahlglossar}{CIL}, which retain the scientifically selective constraints of a formal classification, together with the benefits of this.

Problems are various, and include also the use of terms across vocabularies and the doubts which might be generated by archaeological chance.\footnote{ See \citet[XI-XII]{Piso}, following a current of studies which has its main in G. Susini and J.-N. Bonneville.}

What looks like an \href{http://www.eagle-network.eu/voc/objtyp/lod/29}{altar} could be also the \href{http://www.eagle-network.eu/voc/objtyp/lod/57}{base of a statue}, for example. Or perhaps it could have served two different function has an object of which we might or not have any archeological, contextual or textual trace.

\begin{figure}[htbp] \includegraphics[width=\columnwidth]{statuebase.png} \caption[]{Statue Base} \end{figure}

This is also the case when we deal with techniques of execution of an inscription. Those can also be multiple and an inscription can be for example both a graffito and painted.
And this is just not to mention the extreme complications of having a lot of independent vocabularies, each one for its own sake and small scope purpose. This is often the best option for correct attribution as local habits do vary. But within the framework of harmonization activities, it is important to refer to connected and interrelated definition of terms (not univocal!) and to allow for multiple values to coexist. This is one best practice in the definition of vocabularies which is very important for the nature of content in question. 
This is also the reason why the EAGLE decided to provide not just definitions of all main terms but also examples from different areas and in different state of preservation, and as far as possible also bibliographic reference to authoritative sources.

The EAGLE Vocabularies are the following

\begin{itemize}
\item \href{http://www.eagle-network.eu/voc/typeins.html}{Type of Inscription}
\item \href{http://www.eagle-network.eu/voc/objtyp.html}{Object Type}
\item \href{http://www.eagle-network.eu/voc/material.html}{Material}
\item \href{http://www.eagle-network.eu/voc/writing.html}{Writing and Execution} 
\item \href{http://www.eagle-network.eu/voc/decor.html}{Decoration} 
\item \href{http://www.eagle-network.eu/voc/statepreserv.html}{State of Preservation} 
\item \href{http://www.eagle-network.eu/voc/dates.html}{Dating Criteria} 
\end{itemize}

\begin{figure}[htbp] \includegraphics[width=\columnwidth]{edb.png} \caption[]{EAGLE Vocabularies used by the Epigraphic Database Bari} \end{figure}

The EAGLE vocabularies are maintained in a git repository where they are manually updated and transformed to a readable versions (with another XSLT transformation) published on the EAGLE website as a self standing resource linked to several others. A user is thus able to search for a type of object and read a definition as well as searching for more information on Wikipedia, Wikidata, Wikisource, etc. including the partners' websites. The Vocabularies perform thus both a technical function and are at the same time a living resource which can be read and used independently.

\section{Linked Open Data}
The Europeana LOD practice for metadata recommends the adoption of machine readable vocabularies. Within the EAGLE BPN Linked Open Data practices and approaches have been taken on board for a process which will bring current material some steps closer to that practice.\footnote{\citet{Bizer2009}} Addressing existing problems in classification and the publication of a machine readable vocabulary of values (controlled vocabulary) is one of these steps:\footnote{\citet[4-5]{Harper2012} for the definition. See also \citet{Isaac2012}} it involves identifying and matching values as well as giving them stable identifiers and making of them self standing resources where possible.

In facts the many problems and complexities which \citet{Manzella1987} underlined and exposed can find a solution using the LOD approach.\footnote{\citet{Bizer2009}} Both the limits of choices and hierarchical organization can be bypassed using unique identifiers and relations among those. Polyhierarchic structures as well as the presence of many possible denomination with specific and self standing definition allows for both precision, consistency and sustainability over time of this approach.\footnote{\citet{Harpring2010}}

For example, an inscription can easily be classified both as \emph{magistrati populi romani} and as \emph{decuriones municipales}; a document can be classed both under the \emph{fasti} and under the individual mentioned in the text. These major advantages can be exploited when the effort of publishing open machine readable material is undertaken and a document can be then classified chronologically and alphabetically without the need of an authorial choice.\footnote{I would like to thank prof. Piso for input on this point, given during the work of the Working Group.}

Nevertheless, the major possible advantage is that there is no evident need, with a LOD approach, to distinguish among Greek and Latin inscriptions: they can find their place together in a Linked Data  edition and benefit of other efforts in this direction.

\section{Crosswalking EpiDoc to EDH with XSLT}
The proof of concept of usability of the XML source file is given by the reverse function being possible, i.e. to transform the XML files into a database data structure. A team of people from the EAGLE BPN\footnote{Gabriel Bodard, Pietro Liuzzo, James Cowey, Frank Grieshaber, Brigitte Gräf and Francisca Feraudi Gruénais.} and Scott Vanderbilt, curator of the online edition of the Roman Inscriptions of Britain,\footnote{\url{http://romaninscriptionsofbritain.org/}} worked on mapping the EpiDoc contents of that project to the EDH database model,\footnote{The work was largely based on the previous exercise of this kind, made by Gabriel Bodard and James Cowey for the Inscriptions of Roman Tripolitania.} facing a series of problems and challenges, which brought an interesting discussion, also on principles of ``up-conversion'' and cross-walking, which should be as often as possible multi-directional.\footnote{The files are available here: \url{https://raw.githubusercontent.com/EAGLE-BPN/Epidoc2EDH/master/rib_to_edh-2.xsl}}

Challenges included
\begin{itemize}
\item alignment to EDH internal vocabularies
\item alignment to EDH conventions for bibliographic references
\item use of internal references and retrieval of key based information
\item matching of existing items
\item formatting to EDH conventions differentiated based on the type of content extracted (text or names) 
\end{itemize}

The result was successful and proves once more if needed that an XML source format is a preferable option whatever the variety of desired output are.

\section{Conclusion}
I have presented in this paper the effort of a large consortium of people and institutions belonging to the same field of interest, that of Ancient Roman and Greek Epigraphy.
Mappings and harmonization of data have been for the EAGLE project not just a way to allow machines to do more, but an intellectual effort of revision of contents and establishment of best practices. This work prepares further steps which hopefully will take place soon, in which one repository of inscriptions will be accessible to all to edit, contribute and download data from several websites and interfaces. In such environments related contents based on descriptive URIs, as well as places, personal names and other semantic information will be possible and meaningful and perhaps in a not too far future classicists will have a complete and functioning network of Linked Ancient World Data, including texts, manuscripts, papyri, coins and inscriptions.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}%was eagle2
\end{document}

### 2015-12-14 source/38/EAGLE2016_Final_Paper Modolo.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Mirco Modolo}
\date{2015-11-20}
\begin{document}
Abstract

\begin{figure}
\centering
\begin{minipage}{10.901cm}
FREE REPRODUCTION OF CULTURAL HERITAGE FOR RESEARCH PURPOSES: REGULATORY ASPECTS AND NEW PROSPECTS FOR PROJECT
‘EAGLE’\newline

Mirco Modolo

Free lance archaeologist and archivist

Email(s): mircomodolo@gmail.com
\end{minipage}
\end{figure}
This paper traces the genesis and activity of the “Fotografie libere per i Beni Culturali” movement promoting free
remote reproduction, and reuse for scientific purposes, of manuscripts and antique volumes conserved in Italian
archives and libraries. Simultaneously, evidence highlights the advantages of deregulating cultural heritage images for
historical research, in particular for the study of antique epigraphy developed by the EAGLE project. \

\bigskip

Keywords

Archivi, Biblioteche, Manoscritti, Riproduzioni, Creative Commons

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\begin{enumerate}
\item Free reproduction of cultural heritage outlined in the “Art Bonus” decree
\end{enumerate}
\ L’entrata in vigore, il primo giugno 2014, del decreto “Art Bonus”, che sanciva la libera riproduzione per finalità
scientifiche, purché effettuata a distanza, di tutti i beni culturali rappresentò una novità assoluta per la ricerca
storica: numerosi archivi e biblioteche resero infatti libera e gratuita la riproduzione con la fotocamera dell’utente
di manoscritti e volumi storici, con evidente vantaggio per la ricerca storica e in particolare per lo studio
dell’epigrafia nei manoscritti antiquari.

Appena un mese dopo un emendamento restrittivo, poi divenuto legge, intervenne ad escludere esplicitamente dalla
liberalizzazione i beni bibliografici e archivistici. Alla sorpresa subentrò allora l’irritazione di numerosi studiosi,
alimentata dal paradosso che mentre rimaneva libera la fotografia per il turista nel museo, lo studioso impegnato in
attività di ricerca in archivi e biblioteche non avrebbe invece più potuto beneficiare dei vantaggi della
liberalizzazione.

On the first of June 2014, the ‘Art Bonus’ decree came into effect, sanctioning the free reproduction of all cultural
heritage for scientific purposes (even reproduction carried out at a distance). It represented a new first for
historical research\footnote{\ Decreto-legge 31 maggio 2014 n. 83, art. 12.}: numerous archives and libraries opened
and made free of charge the reproduction of manuscripts and historical volumes with a user’s camera. This provided a
clear advantage for historical research, particularly epigraphy, which considers antique manuscripts from the 15th and
19th centuries an important source for the study of Latin and Greek inscriptions.

As a direct consequence of the Art Bonus decree, some institutions, including the State Archives of Florence (Archivio
di Stato di Firenze), allowed users to make digital copies of records in consultation. Unfortunately, just one month
later, a restrictive amendment modified the original text of the law, explicitly excluding printed books, manuscripts
and archival documents from the liberalisation\footnote{\ Legge 29 luglio 2014, n. 106}. Many scholars’ surprise was
overtaken by irritation, driven by an apparent paradox: museum photography remained free for the tourist, yet the
scholar engaged in research activity in archives and libraries would not reap the benefits of
liberalisation\footnote{\ MODOLO, M., Il sogno infranto delle libere riproduzioni, in Il Giornale dell’Arte, p. 10
(settembre 2014); Brugnoli, A., Ancora sulla riproduzione dei beni culturali, in ROARS (05/10/2014).\par }.

\bigskip

\begin{enumerate}
\item Current state of the problem
\end{enumerate}
Excluding bibliographic and archival records from liberalisation has effectively restored the previous regime, still in
force today: taxing images taken with a personal digital device (in the institutions still allowing this freedom), or
an outright ban on use of a personal device – requiring the paid use of a photographic service. Rates for using a
personal device are the most varied (on average 3 Euro per unit), while some archives require a rental fee for making
reproductions on a personal device in a specialised room. Thus an extreme variety of tariffs emerges, commoditising
research to the detriment of those who contribute to enhancing documentary heritage through their own study.
Furthermore this is in direct contrast with the Italian Constitution which, according to articles 33 and 9, requires
the Republic not only to guarantee free research, but also to actively promote it. An effective “tax” on reproduction
with a personal device becomes a research obstacle; it is absurd to expect a fee for a service provided autonomously by
the researcher himself.

Regarding ‘reuse’ of the image, publication for scientific purposes remains subject to a precise and formal request to
have ‘publication authorised’ granted on stamped paper, hugely time consuming for administrator as well as applicant.
The payment of fees for use (‘canoni di utilizzo’ or royalties) is excluded instead for the publication of images of
cultural heritage (such as archival documents, ancient printed books or ancient inscriptions) in scholarly books or
papers with a cover price of less than 77 Euro and up to 2000 printed copies (as per circular 21/2005 of the ‘Ministero
dei Beni e delle Attività Culturali e del Turismo’).

\bigskip

\begin{enumerate}
\item \begin{enumerate}
\item \ \ Traditional objections to liberalisation of reproductions
\end{enumerate}
\end{enumerate}
The principal objections opposing the liberalisation of reproductions of bibliographic and archival records, and which
must be addressed here, regard both ‘conservationist’ as well as derivational ‘improper use’ risks for the images.

The protection of archival material is one of the principal arguments against free reproduction. Indeed archivists at
the State Archives of Florence (Archivio di Stato di Firenze) directly noted the advantages of free reproduction in the
month “Art Bonus” came into force (prior to the approval of the restrictive amendment): one could observe not only that
the images did not damage the archival material, but rather, being subject to less movement, actually contributed to
its conservation\footnote{\ https://fotoliberebbcc.wordpress.com/2015/06/06/4724/}.

If the remainder of the problem is indeed conservation, it is difficult to understand why this might be considered a
problem only in Italy: the policies of the National Archives of the United Kingdom and the French Archives Nationales
already foresee free reproduction with a user’s personal device. In recent months the British Library has spread across
the web a short video demonstration in order to instruct the users in the proper handling of the codex manuscript.
Specifically, the video prompts the user to employ simple lead cordons to hold open the pages of a codex without
compressing the back of the volume and risking damaging the
binding\footnote{\ http://www.bl.uk/reshelp/inrrooms/stp/copy/selfsrvcopy/book\_photography\_video.mp4.}.

In addition, digital replaces contact photocopy (which still exists in some archives), primarily by reducing the wear
and tear in handling a manuscript over the course of extended periods. As such, reproduction at a distance not only
does not damage media more than normal handling during consultation, but rather may be a powerful ally to its
conservation. For this reason it should be incentivised and not limited or in fact denied as happens today.

To those fearing ‘abuse’ relating to improper use of images (different from those permitted by authorisation) it is easy
to respond that prior authorisation had already been eliminated from “Art Bonus” for photographs of all cultural
heritage other than bibliographical or archival records. In other words, ‘ex post’ control instead of ‘ex ante’
control, characterised as the only genuinely effective measure. It is hard to imagine now, in the digital era, strict
ironclad control over all who request daily permission to reproduce mass of images which are each day authorised only
for “educational purposes”! This is a problem which doesn’t exist in France: the Archives Nationales does not in fact
provide any specific request for autonomous reproduction, while the Bibliothéque National de France allows anyone to
freely download PDF documents or antique volumes in good resolution without the need to include distinctive watermarks.

The application of tariffs for reproduction and reuse of cultural heritage images appears, even more than a means of
generating profits, a pretext designed to discourage reproduction of manuscripts and ancient volumes and to minimise
the risk of abuse associated with using the images for profit-making activity, in a way which might ‘damage’ the public
treasury\footnote{\ http://www.bianchibandinelli.it/2015/05/25/fotografie-libere-un-comunicato-dellassociazione-bianchi-bandinelli/}.

\bigskip

\bigskip

\begin{enumerate}
\item The proposal of \ “Fotografie libere per i Beni Culturali”
\end{enumerate}
Sulla scia del decreto “Art Bonus”, per tentare allora di sanare una situazione di disagio diffusa tra gli studiosi nel
mese di settembre 2014 si è costituito spontaneamente tra i ricercatori un movimento di idee, “Fotografie libere per i
Beni Culturali”, che ha avviato una petizione di massa al ministro dei Beni Culturali Dario Franceschini finalizzata a
promuovere la libera riproduzione in archivi e biblioteche, ripristinando lo spirito originario dell’”Art Bonus” Una
iniziativa senza precedenti non solo per il numero di firme, più di 4200, ma anche per la qualità dei sottoscrittori:
sono presenti, oltre ai massimi rappresentanti in Italia di tutte le discipline storico-umanistiche impegnati nella
ricerca storica e nella tutela dei beni archivistici e bibliografici, anche funzionari della pubblica amministrazione,
studenti, ricercatori e semplici appassionati di storia locale.

Il movimento ha prodotto inoltre una proposta di modifica dell’art. 108 del Codice dei Beni Culturali disciplinante le
riproduzioni, visualizzabile in una pagina del sito web del
movimento\footnote{\ https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/}, con la quale si chiede al
legislatore di limitare i casi di esclusione della liberalizzazione ai beni bibliografici e archivistici
rispettivamente ai volumi a stampa già tutelati da norme sul diritto di autore e ai documenti di archivio contenenti
dati sensibili (privacy).

La libera riproduzione dunque è uno strumento fondamentale che, venendo incontro alle esigenze prioritarie della
conservazione, agevola fortemente l’attività di trascrizione dei documenti e dunque la ricerca storica, consentendo un
significativo risparmio in termini di tempo e di denaro per lo studioso, soprattutto per chi è costretto a spostarsi in
archivi distanti dal proprio luogo di residenza.

La liberalizzazione, intesa come gratuità dello scatto con mezzo proprio che si accompagna all’esenzione dall’obbligo di
autorizzazione preventiva, si potrà estendere, secondo il testo della proposta, non solo alle copie effettuate per
“studio personale”, ma anche per al riuso ‘non commerciale’ delle immagini, prevedendo una eccezione ‘culturale’ per
agevolare le pubblicazioni a stampa scientifiche entro precisi termini. “Fotografie libere per i Beni Culturali”
propone infatti di rendere libera la pubblicazione di immagini di beni culturali in testi scientifici “minori” con
l’obbligo tuttavia di specificare sempre il nome dell’istituto di provenienza e di consegna di un esemplare della
pubblicazione. In questi casi la richiesta formale di ‘concessione’ potrà essere sostituita da una semplice
comunicazione online all’istituto dell’intenzione di pubblicare. Per le pubblicazioni “maggiori”, oltre cioè le soglie
indicate, ed in generale per attività lucrative, rimarrà in vita il regime di concessione, che prevede sia la richiesta
formale di autorizzazione sia il pagamento di canoni d’uso (royalties).

Infine si richiede di creare le condizioni normative per facilitare la condivisione sul web delle riproduzioni di
documenti di archivio, manoscritti e volumi antichi eseguite in sala dagli studiosi.

To meet scholars’ demands, in September 2014 researchers spontaneously formed a movement of ideas, “Fotografie libere
per i Beni Culturali”. The group launched a petition asking Minister of Culture on. Dario Franceschini for free
reproduction in archives and libraries, renewing the original spirit of the “Art Bonus”
decree\footnote{\ https://fotoliberebbcc.wordpress.com/category/adesioni-e-contatti/}. It was an unprecedented
initiative given not only the number of signatures (more than 4200), but also the quality of subscribers: a chorus
comprising the highest representatives from each of the historical-humanitarian disciplines from around Europe:
associations, university docents, researchers, students, and simple fans of local history, along with directors of
state archives and functionaries of the ministry
itself\footnote{\ http://www.ilgiornaledellarte.com/articoli/2015/4/123892.html.}.

Moreover, the movement produced a proposal to amend article 108 of the Codice dei Beni Culturali regulating the
reproduction of cultural heritage (https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/), to eliminate the
exclusion of printed books and documents from the liberalisation archive, hence complying with copyright and personal
privacy for archival documents.

Free photography is therefore a fundamental tool which, without contradicting the demands of conservation, greatly
facilitates the task of transcribing documents and historical research – allowing significant time and money savings
for the scholar, especially those forced to move to archives distant from their own place of residence.

Free photography means the ability to use one’s own smartphone or camera in archives and libraries during consultation,
without the need to request prior written approval (for purposes other than profit). A ‘cultural’ exception should be
provided for a particular form of profit, scientific publishing, given the belief that research is meaningful only if
it can be disseminated in the broadest way\footnote{\ Circolare Mibact 21/2005.}. “Fotografie libere per i Beni
Culturali” in fact proposes to make free the publication of images of cultural heritage in scientific texts in
circulation and of a limited cover price (below 77 Euro and 2000 copies) with the sole duty to always specify the name
of the library/archive of provenance, and to provide a copy of the publication. In such cases the official request for
authorisation by ordinary post may be substituted by a simple communication online at the institute with the intent to
publish. For the “greater” publications, other than the usual indicated above, and in general for moneymaking
activities, the concession regime will remain in force, requiring both a formal authorisation request as well as a
payment for the canon of use (royalties).

\bigskip

\bigskip

\begin{enumerate}
\item The free reuse cultural heritage images
\end{enumerate}
Il tema del riuso delle immagini di beni culturali, cui si è brevemente accennato nell’ambito della proposta di
“Fotografie libere per i Beni Culturali”, si incrocia strettamente con l’esigenza di EAGLE di pubblicare online
iscrizioni latine e greche. A questo scopo EAGLE ha stabilito una convenzione con la Direzione Generale per
l’innovazione tecnologica e la promozione del Ministero dei Beni Culturali in data 21 novembre 2005 per pubblicare sul
web le riproduzioni di iscrizioni latine e greche già edite anteriori al VII sec., mentre per le iscrizioni inedite si
autorizzano nuove riproduzioni “fatti salvi eventuali diritti scientifici” degli studiosi che hanno rinvenuto le
iscrizioni e solo tramite il “consenso” delle soprintendenze interessate. Una seconda convenzione, stipulata nel 2012
con l’Archivio di Stato di Roma, autorizza EAGLE invece a pubblicare di immagini di documenti di archivio che riportano
informazioni su iscrizioni.

\ Prendendo spunto dalla proposta di “Fotografie libere per i Beni Culturali” è utile riflettere sulle modalità per
rendere effettiva la libera divulgazione con ogni mezzo delle immagini di beni culturali per finalità non lucrative già
prevista dall’art. 108 comma 3 del Codice dei Beni Culturali, così come modificato dall’”Art Bonus” (è libera “la
divulgazione con qualsiasi mezzo delle immagini di beni culturali, legittimamente acquisite, in modo da non poter
essere ulteriormente riprodotte a scopo di lucro”). In altre parole sarebbe importante quindi esplicitare a livello
normativo che la divulgazione delle riproduzioni per fini non lucrativi (e dunque anche la pubblicazione online a scopo
scientifico), debba rimanere gratuita e libera dalla procedura burocratica vigente che deve oggi accompagnare qualsiasi
richiesta di concessione alla pubblicazione delle immagini (sostituendola al più, come nel caso delle pubblicazioni a
stampa, con una comunicazione che informi l’istituto di appartenenza sull’uso culturale dell’immagine).

La British Library ad esempio, che proprio nei mesi scorsi ha liberalizzato le riproduzioni per assecondare le richieste
dell’utenza, promuove esplicitamente la condivisione sui social networks delle immagini di manoscritti, riconoscendo il
‘great benefit in sharing images’.

Il riuso delle immagini di beni culturali per finalità educative e scientifiche potrebbe essere regolato attraverso il
ricorso alle licenze Creative Commons. Il Ministero dei Beni Culturali potrebbe chiarire l’ambito applicativo dell’art.
108 comma 3, prevedendo il ricorso a licenze CC-BY-NC-SA by default sulle immagini di beni culturali. In questo modo
verrebbe sancita la libera diffusione delle immagini per finalità diverse dal lucro purché sia sempre specificata la
provenienza del materiale originale e a condizione che nei riusi successivi vengano attribuite le stesse licenze di
origine.

\bigskip

The reuse of cultural heritage images, briefly mentioned in the proposal scope for “Fotografie libere per i Beni
Culturali”, intersects closely with EAGLE’s need to publish Latin and Greek inscriptions online for cultural purposes.
To this end, EAGLE established an agreement with the Italian Ministry of Culture on the 21st November 2005, to publish
on the web low-resolution reproductions of previously published Latin and Greek inscriptions dated before the 7th
century. A second agreement, signed in 2012 with the Archivio di Stato di Roma (State Archives of Rome), authorizes
EAGLE to web-publish images of archival documents, including dates and information of interest for the study of antique
epigraphy. Despite these formal agreements, those working for EAGLE in Italy and across Europe, understand how
difficult it is to publish images of inscription – mostly due to the mistrust of those who still view diffusion of
digital images as a form of “expropriation” and not an occasion of cultural enrichment.

The added-value of digital resides in its ability to disseminate knowledge, and not in static storage, as clearly
demonstrated in numerous international examples which head in the direction of an ever-greater open majority toward
free-reuse: the British Library for instance, which incidentally in the past months has liberalised reproductions to
satisfy the demands of users, explicitly promotes the sharing of images and manuscripts on social networks, recognising
the ‘great benefit in sharing images’. An interesting case is the Metropolitan Museum of New York, which has made
available hundreds of thousands of high-resolution photographs of their own works, even allowing their reuse in
scientific publications, in free format but also free from prior authorisation; the Walters Art Museum in Baltimore
proposes an even more extreme model in giving anyone the possibility to reuse excellent resolution images for any
purpose, including commercial (!) with a “CC0” license.

Free reuse of the images, at least for cultural purposes, has obvious positive effects on the liberal circulation of
scientific content, and also represents an important way to improve the visibility of institutions and their
collections. Museums, archives and libraries should therefore be seen as active and dynamic centres of cultural
promotion, more so than bureaucratic offices for the static conservation of cultural heritage\footnote{\ Cfr. on the
Creative Commons licences in archaeology: Brugnoli, A., Gardini, S., Fotografia digitale, beni archivistici e utenti:
l’impiego e la diffusione di una nuova tecnologia nella normativa e nelle iniziative dell’amministrazione archivistica,
in Archivi e Computer, marzo 2013, pp. 213-256; Gualandi, M. L., L’archeologia italiana di fronte alla sfida dell'Open
Data. Il MOD - Mappa Open Data Archive, in Atti delle Giornate Gregoriane (VII edizione), Bari 2014, pp. 69-75.}.

\bigskip

\begin{enumerate}
\item Encouraging first results and European perspectives for project EAGLE
\end{enumerate}
The effort underlying the collection of thousands of signatures has already led to encouraging openings. A recent
parliamentary question to the minister on whether to restore the original sprit of the “Art Bonus”
decree\footnote{\ http://www.senato.it/japp/bgt/showdoc/showText?tipodoc=Sindisp\&leg=17\&id=914146}, and the
comparable availability shown by the Ministry of Culture, are in fact events which inspire cautious optimism –
primarily because they mark a return to this topic in parliament after one year. It begins a course that will hopefully
soon arrive at the modification of article 108 of the “Codice dei Beni Culturali” regulating reproduction and the
release of an appropriate circular which should affirm, without ambiguity:

\bigskip

1)\ \ the free reproduction in archives and libraries for research purposes and in accordance with complete respect for
privacy and copyright law;

2)\ \ the free reuse of images of cultural heritage for purposes outside of financial gain, or free publication online
and in print within the limits mentioned above\footnote{\ Cfr. n. 10.}.

It is important that these principles remain valid in archives and libraries which pertain not only to the State, but
also to the local administrations and – why not – to the private owners of cultural heritage which, although private,
constitute the patrimony of public interest. If in Italy one might achieve a similar result, as is hoped, the immediate
next objective would be to export this model to other countries in the EU to encourage the widest possible circulation
of information in the field of historical research at the community level.

Photography, as noted here, is an indispensable method for studying not only epigraphic text but also the supportive
material which contains it. The advantage for EAGLE will then be two-fold: the liberalisation of photography in
archives and libraries can facilitate the study of numerous epigraphic repertoires which can be found in manuscripts
and antiquarian sources, while the free reuse of cultural images in museums across Europe may allow the broadest
sharing of inscription reproductions on the web.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{A\&D\&F\&L}
\date{2015-11-15}
\begin{document}
Abstract

\begin{figure}
\centering
\begin{minipage}{10.901cm}
VISUAL FEATURES OF INSCRIPTIONS.

An issue for EDB (and EAGLE)\newline

Antonio, Enrico FELLE

Università degli Studi di Bari {\textquotedbl}Aldo Moro{\textquotedbl} -Italy

Email: antonio.felle@uniba.it
\end{minipage}
\end{figure}
In these last years the amount of digital images of inscriptions increased very quickly: we do not need accurate textual
descriptions of the so-called anaglypha, because we can directly see them. But we have to build a search-by-image,
using photos and drawings but also tagging them with standardized - and shared - labels.

The issue of the {\textquotedbl}illustrated inscriptions{\textquotedbl} brings us to consider more broadly all the
visual features of inscriptions, that were conceived as objects to see, not only to read.

\bigskip

\bigskip

Keywords

Early Christian Epigraphy, Byzantine Epigraphy, Middle Ages Epigraphy, Images, Symbols, Signs, Paleography,
Stonecutters’ workshops

\bigskip

1. {\textquotedbl}Illustrated inscriptions{\textquotedbl} by the Christians of Rome.

\bigskip

In 2012, during a conference in Rome about Late Antique plates decorated with engravings, I presented a paper about the
potentially very useful contribute that the Epigraphic Database Bari (EDB) could offer \ to study and to interpret the
notion and the use of images (signs, symbols, figures and so on) by the Christians of Rome in Late Antiquity, by
analyzing the inscriptions stored in the database [Felle 2013].

Then, the first datum was that only a quarter of these epigraphs displays images or generical non-alphabetical signs
[Felle 2013, p. 101] \ (fig. 1).

 \includegraphics[width=10.119cm,height=5.74cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img001.pdf}

Fig. 1. Percentage in EDB of the {\textquotedbl}illustrated inscriptions{\textquotedbl} \ (chart by A. E. Felle).

\bigskip

After storing in EDB other 10000 inscriptions, since 2012 to the present day, the percentage of figured inscriptions is
still the same: then, I think that we are able to consider this datum enough sure; so we are able to partially correct
the common idea that using images in written monuments is a recurrent, typical and characteristic feature of almost all
the Early Christian inscriptions.

The second datum defined by the 2012 survey was the proportional decreasing of the use of different kind of images from
the first decades of the IV century (the age of Constantine), when a huge and pervasive use of the so-called signa
Christi - first of all the Chi-Rho monogram, with all its variations - prevails on all the other signs and figures
[Felle 2013, 101-102].

Carlo Carletti explained this phenomenon as the result of a will to display explicit signs of a religious identity, such
the Chi-Rho monogram is [Carletti 2008, 68-72; Felle 2007, 365-366]. But we have to underline that the phenomenon is
not exclusive of Christian patrons. We observe more and more recurrent similar {\textquotedbl}signs of
identity{\textquotedbl} also among inscriptions commissioned by Jews, not only in Rome but also in other contexts in
Late Antiquity world where they were [Felle 2007, passim; Felle, in press]. \ \

Going back to the inscriptions by Christians, the use of signa Christi \ in form of monograms strongly reduces the use
of \ other christological signs or figures, as for instance the anchor: this one, very recurrent during all III
century, disappears completely and very quickly, already in the very first decades of IV century [Felle 2012, 103]
(fig. 2).

\bigskip

 [Warning: Image not found]

Fig. 2. Use of different images in inscriptions by Christians of Rome between III and IV cent. (chart by A. E. Felle)

\bigskip

Since its conception, EDB recorded the presence and the different kinds of various signa Christi, both by a checkbox and
in the text field, with standardized descriptions, with the aim to easily retrieve them in the database and to get
valuable results from structured queries about their recurrence.

The results of the 2012 study were mainly of quantitative nature; today, by the existing large repositories of images in
EDB – as like in general in EAGLE and in other similar epigraphic projects – we can improve the qualitative analysis of
this and other features of Late Antique inscriptions.

First of all, \ we have to say that in EDB we faced huge difficulties about the treatment of images other than signa
Christi, or generally other non-alphabetical signs, or also captions related to figures on the slabs, and so on.
Indeed, at the moment we are still not able to automatically obtain by EDB a structured index of the repertoire of the
images. As in other epigraphic databases, in EDB they are recorded directly reporting their descriptions as like they
are in printed editions (sometimes very old, as the first volumes of ICVR, for instance); there, in absence of
pictures, the so-called anaglypha are concisely described by simplified and repetitive clichées to suggest the depicted
subjects to the readers (fig. 3) or by brief descriptions in Latin in transcriptions or also in the commentaries (fig.
4).

\bigskip

\bigskip

 \includegraphics[width=10.59cm,height=5.159cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}

\bigskip

 \includegraphics[width=10.268cm,height=2.639cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}

Fig. 3. Rome, coemeterium Hippolyti, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.12.14) \ and
edition \ in ICVR, VII 19820.

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

 \includegraphics[width=8.202cm,height=2.514cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img004.png}

 \includegraphics[width=9.622cm,height=2.244cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img005.png}

Fig. 4. Rome, coemeterium Maius, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.8.3) and edition in
ICVR, VIII 22407.

\bigskip

Often, these descriptions are different although indicating the same subjects: in the ICVR, the reason of this
disomogeneity is not only the longue durée of the realization of the corpus (seventy years, since 1922 to 1992, when
the last published volume, the tenth, appeared) but also a refined textual variatio. Surely it can be appreciated in
printed editions but, for the aim of our digital archives, produces real difficulties.

Some examples: in EDB different verbal descriptions about the some subject are recorded, such as avis uvam pascitur
(e.g. ICVR, III 8114a, b, c, e), or avis uvas pascitur (ICVR, III 8004b); or \ also avis racemum carpens (ICVR, IX
24020), avis racemum carpit (ICVR V, 14157), avis racemum pascitur (ICVR, V 15194), avis racemum rostro carpit (ICVR,
IV 10934): it is not easy to perceive some difference). However, this variatio prevents right results in retrieving
data in our database.

Moreover, recording in EDB descriptions with different words for the same illustrated subject, \ such as
{\textquotedbl}avis, racemus{\textquotedbl} \ (EDB 19827: ICVR, III 9311, see fig. 5a) or \ {\textquotedbl}avis cum
racemo{\textquotedbl} (EDB 24933: ICVR, III 8044, see fig. 5b) we are not able to retrieve all the occurrences of this
same subject, because they are recorded (both n ICVR and in EDB) in different ways.

\bigskip

 \includegraphics[width=5.076cm,height=2.21cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img006.png}  a
\ \ \ \ \ \ b 
\includegraphics[width=4.207cm,height=2.078cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img007.png}

Fig. 5. Rome, catacomb of Domitilla. Edition of \ ICVR, III 9311 (a) and \ ICVR, \ III 8044 (b). \ \ \ \ \ \

\bigskip

\ This ambiguity prevents to retrieve all the occurrences of the same illustrated subjects and then they adulterate the
result of our queries: I think that we have to correct as soon as possible this ambiguity, in order to establish an
unique way to describe the anaglypha.

One can say that the present ease to obtain and to use digital pictures of the inscriptions overtakes this issue: surely
that's true. But, I do not entirely agree with this point of view.

In these last years - also with the kind help by the Photographic Archive of Papal Commission of Sacred Archaeology
(http://www.archeologiasacra.net/pcas-web/) - \ the amount of images available in EDB increased very quickly: this is
surely an advantage in respect to the situation of only three years ago. At the present day and more over in the
future, we no longer need accurate textual descriptions of the anaglypha: we can directly see them. But, the
possibility to easily view a photo or a drawing of an inscription does not solve the issues related to search and to
retrieve inscriptions bearing given kinds of image, or specific signs, and so on. \

The relative high occurrence of images in Christian inscriptions drives EDB team to try to build a search-by-image,
using photos and drawings, but also tagging them with standardized labels. A “high definition” analysis of non-verbal
language of the inscriptions by Christians of Rome in Late Antiquity surely needs photos, drawings, and so on, but
mostly needs a logical, structured, hierarchically ordered taxonomy of all non textual elements defined by controlled,
firm and shared vocabulary: a thesaurus imaginum\footnote{ Surely the experiences of other projects can be useful to
this aim: I think for instance to the solutions presented during the VIth EAGLE International Event in Bari by the
lecture offered by Rebecca Benefiel and Holly Sypniewski about the Ancient Graffiti of Herculaneum project.}, that can
be a suitable tool also to trail the activity of different stonecutters’ workshops that served – in the case of EDB
inscriptions - the various users of the Roman Christian catacombs.

\bigskip

2. Visual features of the 'written monuments'. \

The issue of the {\textquotedbl}illustrated inscriptions{\textquotedbl} brings us to consider more broadly the visual
features of inscriptions overall. The ancient epigraphs are conceived not only as texts to be read (very few people
were able to do it) but also - and, maybe, firstly - as objects to be seen. I think that we have to realize this
perspective - common in Christian epigraphy studies after all - to understand the communication power of this
communicative medium. Indeed, the topic has been assumed as main theme of the last International Congress of Greek and
Latin Epigraphy in Berlin in 2012 [Öffentlichkeit 2014]; very recently, a just published volume collects various essays
about this same topic just about the Late Antique, Medieval (both Christian and Islamic) world, where this notion of
the inscriptions as 'written monuments' stands out with strong evidence [see Eastmond 2015].

In our projects - we have to admit it - \ the notion that inscriptions are essentially texts is still largely
prevailing: but now we can - consequently, we must - to increase our capability to represent, to record and so to
investigate \ also other, visual features of inscriptions.

\bigskip

2.1. Positioning

First of all, I think to the positioning of the epigraphs in the contexts for which they were created. About EDB, we
already presented in the EAGLE Conference in Paris in 2014 a first attempt to record and to describe the exact
positioning of the inscriptions still found in their original spot, by sharing data with the Domitilla-Projekt (by the
Österreichische Akademie der Wissenschaften and the Deutsches Archäologisches Institut) focused to the frescoes of the
catacomb of Domitilla along via Ardeatina, the largest one in Rome suburb [Felle - Zimmermann 2014]. The positive
collaboration between our two projects \ continued: now we are able to offer to EDB users to view on the updated map of
the catacomb of Domitilla the distribution of the inscriptions still in situ in the four levels of the subterranean
cemetery: one can use them as reliable documents to (re-)consider the history of the complex, to confirm or to deny the
ideas about chronology of its excavation and of its frescoes and about the using of the different zones of the
catacomb.

\bigskip

2.2. Materials and shape

Materials and shape of the written objects communicate im-mediately, to all, before the inscribed texts. Because now we
can do it, we have to display to the users of our projects the real communicative power of the inscriptions, \ that
assume much more meaningfulness when we can see them than when we can read their only texts in a library. By pictures
and 3D rendering of the places where the epigraphs were arranged we should be able also to provide virtual images of
inscriptions \ {}- also lost or incomplete - conceived to be seen and read exactly there. An effective example can be
the dedication by pope Damasus (366-384) to the martyr Ianuarius in the catacomb of Praetextatus on the via Appia (fig.
6).

 \includegraphics[width=8.692cm,height=4.21cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img008.jpg}

 \includegraphics[width=10.141cm,height=9.059cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img009.jpg}

Fig. 6. Rome, catacomb of Praetextatus. Damasus’ dedicatory inscription (ICVR, V 13871) for the martyr Ianuarius (Photo:
PCAS; 3D rendering: G. De Felice).

\bigskip

The bishop of Rome reaches his aim to capture the gaze also of illiterate faithfuls by placing a very large slab of
white marble over the tomb of the martyr, well-lighted by a skylight made on purpose, in strong contrast to the
darkness of the neighbouring galleries of the subterranean cemetery. \

Also the text is aimed at the same goal: the dedication is brief and simple, inscribed using a special writing, very
carefully carved by using a font specially elaborated for Damasus' inscriptions by the fashionable designer Furius
Dionysius Philocalus. The contrast with the common inscriptions in the catacombs, often made by reused marble pieces or
bricks, and very often (not always!) written with rough letters, is impressive. The bishop's intervention is more
meaningfully revealed to all by the visual features of his inscriptions than by their only (metrical or not) texts.

This notion about written monuments is more and more diffused in Western Early Middle Ages and also in Byzantium: one
has no need to read, to perceive the actual and effective messages displayed by inscriptions placed within the fabric
of the walls of Byzantine cities, such as Constantinople or Thessaloniki (fig. 7).

 \includegraphics[width=11.1cm,height=5.673cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}

Fig. 7. Thessaloniki, city walls. Inscriptions in the masonry of a tower near Eptapyrgion (photo: A. E. Felle).

\bigskip

\bigskip

There, the inscriptions are not carved in marble slabs or stone blocks, but they are realized with the same materials of
the walls: simple bricks, but disposed to obtain letters and signs and symbols, visible also from afar. The
inscriptions explain the walls; the walls speak its raisons d'être by the inscriptions, that are in different cases
rich of abbreviations, closed to the reading but open to the sight: writing appears intrinsically significant. In my
opinion, this notion is clearly demonstrated by the unnecessary captions in the icons and in the images of martyrs
(fig. 8), where \ {}- on closer view - the inscriptions are completely useless, if we continue to consider them only as
texts to be read.

\bigskip

 \includegraphics[width=10.749cm,height=9.208cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img011.jpg}

\bigskip

Fig. 8. Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
“useless” caption \ s(an)c(t)a Agnes (photo: A. E. Felle).

\bigskip

\bigskip

\bigskip

\bigskip

2.3. Relationship with the context

Indeed - more over in Late Antiquity and Middle Ages - readability of the texts is not the main property of
inscriptions: rather, the main condition appears their relationship with their contexts. An incisive example can be
offered by the Christian inscriptions bearing biblical quotations [cfr. Felle 2006]: in the middle of the bronze
plating of the marble lintel over the Great Door of the Royal Gates of the Haghia Sophia in Constantinople, an empty
throne is occupied by an open codex, according to the words by Kähler, {\textquotedbl}the only extant plastic
composition dating from the founding period of the church{\textquotedbl} [Kähler 1967, pp. 29-30; 32 taff. 22; 62]
(fig. 9).

\ Fig

\bigskip

On the open codex is inscribed a focused - but barely readable - quotation from John 10, verses 7 and 9, where Jesus
indicates himself as the gate:

\bigskip

John 10.7: \textgreek{E>~ipen} \textgreek{o>~un p'alin <o >Ihso~uc, >Am`hn >am`hn l'egw <um~in <'oti }\textgreek{>eg'w
e>imi <h j'ura t~wn prob'atwn}. (Therefore Jesus said again,~‘Very truly I tell you, I am the gate for the sheep);

[Warning: Draw object ignored]John 10, 9: \textgreek{>eg'w e>imi <h j'ura· }\textgreek{di' >emo~u >e'an tic e>is'eljh|}
\textgreek{swj'hsetai ka`i }\textgreek{e>isele'usetai ka`i >exele'usetai ka`i nom`hn e<ur'hsei} \ ( 9I am the gate;
whoever enters through me will be saved.[a]~They will come in and go out, and find pasture).

\begin{figure}
\centering
\includegraphics[width=4.419cm,height=7.329cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img012.jpg}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=6.655cm,height=6.509cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img013.jpg}
\end{figure}
\begin{figure}
\centering
\begin{minipage}{0.688cm}

\bigskip
\end{minipage}
\end{figure}

\bigskip

\bigskip

Fig. 9. Istanbul, Hagia Sophia. On the left, the Royal Gates in the narthex. On the right, the particular of the image
of the throne with open inscribed codex above the lintel of the central gate (from Kähler 1967).

\bigskip

This the text of the inscription [Felle 2006, n. 505]:

((crux)) \textgreek{e>~ipen <o k('urio)c {\textbar} >eg'w e>imi {\textbar} <h j'ura t~wn {\textbar} prob'atwn;
{\textbar} di> >emo~u {\textbar}{\textbar} >e'an tic {\textbar} e>is'eljh| {\textbar} e>isele'uset(ai) {\textbar}
k(a`i) >exele'uset(ai) {\textbar} k(a`i) nom`hn {\textbar} e<ur'hsei.}

\bigskip

The archeological context and the mirate positioning make tangible, concrete, the sacred words; and the real presence
(not necessarily the readability) of the sacred words give proper and strong sense to their material support and to
entire context, the Royal Gates of the Great Church of Constantinoples [Felle 2015, p. 320 and passim].

\bigskip

2.4. Writing

Scarce or null readability does not imply low quality of the appearance of writing: rather, the writing appears
intrinsecally meaningful such as visual element of the equipment of a simple or rich funerary monument or of a cultual
building. Then, we have to face the issue of the description of the writing not only \ from the \ necessary point of
view of paleography ( we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
also in order to perceive and to understand its non-verbal significance: by its disposition, direction, shape. The
notion of the inscriptions in the Islamic world, \ where often the letters are also - and maybe firstly - images (they
are used as decorative friezes, architectural ornaments, figures) and their clearity and readability are not considered
as necessary, can help us to evocate this feature of the writing (fig. 10).

\bigskip

 \includegraphics[width=11.044cm,height=8.281cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img014.jpg}

Fig. 10. Granada, Alhambra. An example of the writing as decorative frieze and architectural ornaments (Photo: A. E.
Felle)

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

3. Conclusions

The ancient inscriptions actually belong to civilizations where the literacy - with very few exception - was very far
from our standard: to see an inscription with the same point of view of the most part of the citizens in Roman Empire -
and mostly in Late Antiquity - \ we have to become, in some way, illiterate.

In conclusion: we have to consider in digital descriptions of the inscriptions some their {\textquotedbl}visual
features{\textquotedbl} that in our projects - first of all in EDB, of course - are not too considered, although they
are very significant. We need, about encoding these non-verbal features, the same positive results that by Epidoc we
reached \ in encoding the texts: a hard challenge.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/42/Paper_licciardello.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%\title{}
\title{}
\author{Flavia Licciardello}
\date{2015-11-20}
\begin{document}
[Warning: Draw object ignored]Abstract

\ \ \ \ \ \ This paper aims to present an example of a database designed to combine epigraphic, linguistic and
philological data. This database is part of my project on a study of deictic expressions in dedicatory inscribed and
literary epigrams. It includes the results of the analysis of around 600 dedicatory epigrams and it will be used to
extract information on trends and recurrent patterns in the genre.

\bigskip

Keywords

\ Greek epigram, dedication, database, deixis, Hellenistic age.

\bigskip

1.1 Introduction

For a long time, the study of the Greek epigrams has maintained a clear distinction between ‘epigraphic’ and ‘literary’
epigrams. This distinction, however, fails to understand the complexity and the development of this poetic genre. Only
in very recent years have scholars started to highlight properly the important relationship between Hellenistic
‘literary’ epigrams and epigraphic models.\footnote{ The shift in the attitude towards the history of the epigrammatic
genre is well resumed by [Garulli 2012, 5-34]. For other work, which consider the importance of the dialogue between
‘epigraphic’ and ‘literary’ epigrams, see e.g. [Meyer 2005], [Tueller 2008]. On archaic and classical epigrams, see
[Baumbach-Petrovic-Petrovic 2010]. Further bibliography in [Baumbach-Petrovic-Petrovic 2010, 2].}

My PhD project follows this new exegetic line and contributes to our understanding of the inscriptional component of the
Hellenistic epigram.\footnote{ The project is developed within the frame of the C-1 group (‘Deixis and frame of
reference) of the Excellence Cluster TOPOI.} In particular, I focus on the use of deictic expressions in dedicatory
epigrams, i.e. on the use of all those linguistic elements whose meaning and interpretation depend on the spatial and
temporal context where they are uttered.\footnote{ For an overview of the concept of deixis see [Diessel 2012] (with
further bibliography). On deixis in Ancient Greek see e.g. [Felson 2004], [Edmunds 2008] and [Bonifazi 2014].} The
corpus analysed includes inscribed epigrams from the archaic epoch until the end of the 4th century BC, and the
Hellenistic epigrams transmitted in the Greek Anthology or on papyrus. The reference editions used are [Hansen CEG 1-2]
for inscribed epigrams and [Gow-Page 1965] and [Austin-Bastianini 2002] for ‘literary epigrams’.\footnote{ For
convenience, I will later on refer to this second group as ‘literary epigrams’. The definition simply identifies those
epigrams that are transmitted in the Greek Anthology or on papyrus and does not imply any value judgment.}

\ \ In order to deal with this heterogeneous material, I developed a database to register and organise the data obtained
from the analysis of the texts. The specific aim of this database is the registration of all relevant linguistic
features related to spatial, personal and temporal deixis. In addition to this, I put on record for each epigram other
more generic elements, which are related to the dedicatory context, to the linguistic facies and, especially in the
case of inscribed epigrams, to the historical and archaeological context. In this way, on the one hand, I can obtain a
global picture of the whole corpus, which includes data found both in inscribed and literary epigrams. At the same
time, I can keep and easily retrieve all the peculiar features related to each specific text in order to be able to
deal with the material properly, without losing sight of their specificity. The possibility of managing at the same
time these two levels - the whole corpus and the single text – is extremely valuable. In this regard, the database
provides an essential support, since it helps work with complex material on those different levels that must be
considered together, but are usually difficult to keep in focus at the same time.

\bigskip

1.2 Greek dedicatory epigrams and the role of deixis

In the Ancient Greek world it was customary to accompany a dedication to the gods with a – usually – short inscription.
This was normally chiselled on the dedicated object (or on its support) and it recorded the main information related to
the offering. The elements recurring in such inscriptions were the name of the dedicator, the verb of dedication and
the name of the god receiving the dedication.\footnote{ On archaic dedications, see [Day 2010], in part. [p. 1-14] for
an introduction to the genre.} The most common and widespread formula for dedications contained exactly these three
elements: \textgreek{<o de~ina >an'ejhke t~w| je~w|. }A frequent variation was \textgreek{<o de~in'a me >an'ejhke t~w|
je~w|,}\footnote{ The formula here presented (with its two variants) was \ identified by Maria Letizia Lazzarini, in
her study on the formulas of archaic dedications [Lazzarini 1976, 58-40]. Her analysis is based on both verse and prose
inscriptions from the archaic age. However, such basic scheme, with these elements, was normal in epigrams as well and
it remained substantially unchanged in classical and post-classical time.} \ where the speaker is the object itself, as
is clear from the employment of the personal pronoun \textgreek{me. }It is interesting to note that since their first
examples the speaking object was a recurring feature in dedicatory epigrams and it is still frequent in the 4th century
BC.\footnote{ On the topos of the speaking object, see [Burzachechi 1962], [Svenbro 1988, 36-52], [Tueller 2008,
16-27], [Furley 2010, 151-166], [Wachter 2010, 250-260], [Christian 2015, 29-107]. \ } Clearly, the basic scheme here
presented could be varied by omitting some elements (as the name of the dedicator) or adding others (as the generic
name of the object, like the recurrent \textgreek{>'agalma).}

\ \ A crucial moment in the evolution of the epigrammatic genre is the beginning of the Hellenistic era, when epigrams
began to be considered full literary creations and started to be circulated autonomously, no longer limited to one
inscription alone.\footnote{ This trend may be linked to the custom prevalent in the 5th and 4th century BC to copy
epigrams on monuments and then quote them in various texts. See [Gutzwiller 1998, 47ff.]} This development led to the
composition of literary epigrams not intended for inscription, but which in some way maintained the illusion of a
material inscription on a stone. The Hellenistic epigrammatists who worked with inscriptional type of epigrams (and
among these dedicatory epigrams) retained the structure, style and traditions of the epigraphic models, yet the
translation of these into a book context inevitably means that the communicative strategies employed until then had to
be reinvented. The primary reason for such changes is self-evident: the monument intended for inscription and the space
surrounding it do not exist anymore – they have to be imagined into reader’s mind.\footnote{ This involvement of the
reader can be put in relation with what [Bing 1995] defines Ergänzungsspiel, i.e. with the process of supplementation
of the text deliberately incited by Hellenistic epigrammatists.}

This process of reinvention is particularly evident in dedicatory epigrams, which were traditionally chiselled on the
dedicated object itself. In this case, the loss of the original context forces the author to re-elaborate traditional
models, by adding some elements (e.g. the specific name of the dedicated object) which cannot be retrieved from the
surrounding setting.

From a linguistic point of view, this re-elaboration has an impact in the texts on deictic expressions, which were
typically used in inscriptions to lead the gaze of the reader towards the dedicated object. In this epigraphic context
the deictic expressions point to something in front of reader’s eyes (deixis ad oculos), whereas in a literary context
the readers will have to imagine in their mind the invisible referents of deictic expressions (deixis am
Phantasma).\footnote{ The difference between deixis ad oculos and deixis am Phantasma was highlighted and described for
the first time by Bühler in 1934, see [Bühler 1982, 121-126, 133-135].} If on the one hand this change requires a
particular attention to the verbal reconstruction of the setting, on the other hand, the loss of the material context
allows the poet to play with different and new points of view and frames of reference.

The analysis of such deictic expressions can help determine the deictic centre in dedicatory epigrams. The deictic
centre, which normally coincides with the origin of the utterance, works as a point of reference for all deictic
markers and expression.\footnote{ See [Bühler 1982, 102f.] and [Levinson 1983, 63f.].} In the case of dedicatory
epigrams on stone, the deictic centre in spatial terms is generally understood to be in the place of the inscription
itself, which in most cases is the dedicated object. In other words, for all deictic expressions that point to
something close (such as the proximal \textgreek{<'ode ‘}this here’\footnote{ ‘Proximal’ deictic elements are all those
elements which refers to the deictic centre: among these demonstrative pronouns or adverbs such as ‘this’ or ‘here’ and
temporal adverbs such as ‘now’. See [Diessel 2012, 2408f.].}) the occasional reader will look for the referents in the
space close to the inscription. In the evolution of the epigrammatic genre, the loss of the material context produces
an important change, since the strong relation that connects the text with its object starts to fade. This means that
the deictic centre is somehow released from its traditional location on the object and the poet is free to consider and
bring to the text new, different points of view.

\ \ This progressive detachment of the text from its original physical location played a crucial role in the process
that led, out of trivial verse inscriptions, to the emergence of the epigram as a full literary genre. Deictic
expressions were traditionally employed in epigraphic contexts to strictly and clearly bind the text to its material
support. Later on, only the possibility of the separation from a unique location allows the epigram to be circulated
autonomously and reach a wider audience. In this passage, the poet explores new uses of deictic markers in order to
widen the possible references of his text.

My research will try to detect different trends in the use of deictic expressions in order to highlight the similarities
and parallelisms between inscribed and literary epigrams and to elucidate the different deictic strategies employed in
different contexts.

1.3 The database

In the first step of my research, the analysis of the texts is followed by the registration into a database of all the
relevant data related to spatial, personal and temporal deixis. This database is conceived first of all as a tool to
organise the data and make them available for the next steps of the research.

The corpus of epigrams analysed includes dedicatory verse-inscriptions collected in Hansen’s Carmina Epigraphica Graeca
[CEG 1-2], as for epigrams on stone. In this case, I selected those epigrams included by Hansen in the section “Tituli
dedicatorii”. For literary epigrams, I followed Gow’s and Page’s Hellenistic epigrams [Gow-Page 1965], to which I added
the epigrams attributed to Posidippus and edited by Austin and Bastianini [Austin-Bastianini 2002]. In this case, the
selection was made according to the contents of the epigrams and I selected all those epigrams that have a clear
dedicatory frame.

The structure of the database was designed to include all important data related to the text examined, in order to have
for each epigram a complete profile, which includes linguistic, literary, historical and archaeological
aspects.\footnote{ The software used is FileMaker Pro 13.0v4.} The data were organised in three main tables, connected
to each other through a central one (See Fig. 1.1).

[Warning: Draw object ignored]

Fig. \stepcounter{Figura}{\theFigura}.1

Alongside bibliographical and historical-archaeological information, the database records the data obtained by a first
analysis of the form and of linguistic features of the epigram, such as verbal tenses, occurrences of demonstrative and
personal pronouns (deixis); speaking subject and addressee (communicative\_context); verb of dedication, standard name
of the dedicated object (dedicatory\_context).

Such structure offers a balance between the need to record a detailed picture for each epigram analysed and the
possibility of conducting research on single features within the whole corpus. Since each table contains a restricted
amount of data (divided into coherent sections), it is easy to get a simple, clear picture of the recurrence of one
specific feature (and its interconnection with other related aspects) within the corpus and to leave out information
that is not immediately consistent with the research done. When necessary, the tool ‘portal’ helps to find the data
rapidly and to combine the data contained in different tables. In this way, for example, it is possible to look for the
recurrence of a specific tense for the verb of dedication (in the table ‘deixis’) and to see if this is associated with
the use of a specific verb of dedication (in the table ‘dedicatory\_context’) or with a particular epoch (in the table
‘archeological\_context’).

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

1.4 An example: the present tense in dedicatory epigrams

The verbal tenses can work as temporal deictic markers and can play an important role in the definition of the temporal
frame.\footnote{ Though the grammatical category of tense does not always coincide with the semantic category of time,
it is still possible to retrace in the use of a peculiar verbal form a reference to time. However, such analysis must
always consider other linguistic elements, which contribute determine the temporal frame of the text. On temporal
deixis and verbal tense see [Lyons 1977, 677-690]; [Levinson 1983, 73-79]; [Klein 1994, 14-26 and 120-130]. For the
analysis of temporal deixis in Ancient Greek texts see e.g. [D’Alessio 2005] and [Edmunds 2008, 8f.] (with further
bibliography).} In particular, the present tense usually decodes the present time, which is the time of the deictic
centre (‘now’). Since the deictic centre operates as a point of reference for the spatial and temporal orientation,
individuating the deictic centre helps determine the frame of reference of the epigram. This could lay, for example, in
the act of dedication celebrated, in the moment of the composition by the poet or in the moment of reading.\footnote{
On the relation between the temporal frame of the utterance and the deictic centre, see [Levinson 1983, 79f.] and
[D’Alessio 2004].} \

In the definition of the deictic centre in dedicatory epigrams, the analysis of the verb of dedication is particularly
relevant, since this makes clear the relation of the text to the dedicatory act, which is the main piece of information
of the epigram. The most common dedicatory formula presents the verb in the indicative aorist (with augment). Leaving
aside the formulaic aorist and those cases where the verb is not expressed or lost, we find in the corpus analysed a
significant number of cases where the verb that expresses the dedication is in the indicative present.\footnote{ More
specifically, out of a corpus of 598 epigrams analysed, for the verb of dedication the aorist appears 379 times, the
present 50 times. In 172 epigrams the verb is lost or absent. Moreover, some epigrams contains more then one verb of
dedication and these could be expressed in different aspects and moods. }

As for the 41 occurrences of verbs of dedication in the indicative present, a clear distinction can be observed between
epigrams on stone and epigrams with a literary tradition. In the first group, the present form is much more sporadic.
Out of 422 epigrams analysed, we find only 5 clear examples, from different epochs and geographical areas: [CEG 192i]
(Athens, ca. 520? BC \textgreek{>anakeima[i]), [}CEG 302] (Attica, found in Ptoion, ca. 540? BC v. 1
\textgreek{e>im>)}\footnote{ For the verb \textgreek{e>im'i }in dedicatory formulas, see [Lazzarini 1976, 59f.].} [CEG
251] (Athens, ca. 500-480? BC v.1 \textgreek{e>im`i}), [CEG 390] (Apollonia Illyrica, found in Olympia, ca. 450-440? BC
v.1 \textgreek{>a\textlatin{[323?]}}[E32D?]\textgreek{nake'imeja}), [CEG 822] (Geronthrai, 4th cent.? BC v.1
\textgreek{>an'akeitai).}\footnote{ In addition to these, in [CEG 347] and [CEG 775i], the verbs
\textgreek{>anake'im]eja }and \textgreek{kos]mo~umen }respectively are supplied. In [CEG 830ii] the verb
\textgreek{>an'akeitai }refers to another dedication, not to the one celebrated by the epigram.} In all these cases,
the grammatical subject of the verb is the dedicated object. The use of the present form refers therefore to the
present of the object, which, since the dedication, is for the time being in the place of the dedication. As is also
clear from the fact that in most of these cases the object is the speaker, the use of the present verb indicates that
the deictic centre is the dedicated object. It is also interesting to notice that most of the aforementioned epigrams
contain a second verb of dedication, in the formulaic aorist form.\footnote{ The exceptions are [CEG 822] and the
fragmentary [CEG 192i].}

For Hellenistic epigrams of literary tradition, the picture is different. First, the occurrence of dedicatory verbs in
indicative present is less sporadic. Out of 176 epigrams analysed, the present form appears in 33 epigrams. In some of
these cases, the situation is similar to that found in dedications on stone: the subject of the present verb is the
dedicated object, which lays in the temporal deictic centre.\footnote{ However, as opposed to the examples found in
dedications on stone, the object is rarely the speaker. It is also interesting to note that the formulaic
\textgreek{>an'akeimai }is frequently substituted by the simple form \textgreek{ke'imai. }Similarly, in literary
epigrams the simple \textgreek{t'ijhmi }is preferred to \textgreek{>anat'ijhmi, }which is traditional in epigraphic
examples.} An important innovation is the fact that in some epigrams the subject of the verb of dedication in present
form is the dedicator. The present verb refers therefore to the present time of the act of dedication accomplished by
the dedicator. This means that the deictic centre is anchored now to the moment of the dedication, when the dedicator
is obviously present. More evidently, in [Leon. AP VI 288 (HE 2213-2222)] and [Phan. AP VI 299 (HE 2994-3001)], the
dedicator is the speaker and consequently the very deictic centre. Another interesting case is found in [Call. AP XIII
7 (HE 1129-1134)] and [Diosc. AP VI 220 (HE 1539-1554)]. In these two epigrams the dedicator, who speaks in the first
person, pronounces the dedicatory formula, and this is reported by the epigram as a direct discourse. Such examples
indicate that in the development of the Hellenistic epigram the frames of reference multiply. The authors explore new
point of views, not anymore tied only to the dedicated object.

These figures, which obviously requires more in-depth analysis, are an interesting sign of the transformation and
development of the epigrammatic genre in the Hellenistic epoch. Though the poets still move along the path of the
epigraphic tradition, they include new elements in their celebration of the dedicatory act. Elements already employed
from the beginning of the epigrammatic genre are retrieved and renewed, by combining them with new perspectives.

1.5 Conclusion

The database presented here is designed to be a helpful tool in the study of heterogeneous material. This tool will
allow us to apply an interdisciplinary approach, which combines epigraphic, linguistic and philological strategies. It
not only helps to manage a large quantity of data, it also organises the results of the analysis on the texts in a
practical way.

The coherent organisation of the data into a database has two clear advantages. On the one hand, the database provides
an overview of the whole corpus. This could be used, for example, to detect easily specific trends and recurring
elements in the corpus, or to consult and combine the data from different points of view. On the other hand, it is
possible to retrieve rapidly the data connected to each specific text. This comprehensive look, on the whole as well as
in the particular, would be difficult to obtain otherwise and it makes the database a fundamental tool for dealing with
complex materials and applying different levels of analysis.

\ \ It is important to stress that this database aims to enable many possible, different research. The structure is not
based on a single, pre-fixed hypothesis of investigation, but is built so as to let the user interrogate the data in
different ways, without defining a priori the direction of the research.

\ \ Moreover, the database is a dynamic tool and suitable for further and continual additions. This feature is
particularly valuable in the field of the epigrammatic poetry, where continual discoveries of new material (on stone,
but also on papyrus) requires us to enlarge and re-work the corpora constantly. In this regard, the database presented
here is not only a response to the recent need, in the specific field of epigrammatic studies, to create corpora that
combine epigraphic and ‘literary’ materials, it is also open to future additions and new research.

Aknowledgments

\bigskip

The project is funded by the Excellence Cluster Topoi. I am grateful to Dominik Lukas, who helped me develop the
structure of the database. I also thank Dr. Valentina Garulli for her helpful comments.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/43/EAGLE2016_Benefiel_Sypniewski.tex
all files movede from drive

% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%\title{}
\title{}
\author{Rebecca Benefiel}
\date{2015-11-19}
\begin{document}
1.1. Introduction

Our project is working with informal, handwritten wall-inscriptions, or ancient graffiti, which were scratched into the
wall-plaster of ancient towns. \ Several hundred of these handwritten inscriptions have been documented at Herculaneum
and more than 6000 are known from Pompeii. We are contributing these inscriptions to the Epigraphic Database Roma
(www.edr-edr.it), and are creating a linked resource, the Ancient Graffiti Project (ancientgraffiti.wlu.edu), that will
allow users to conduct location-specific searches for graffiti.

Among the many texts written on the walls of these two cities, there sometimes also appear graffiti drawings, or figural
graffiti [Fig. 1].

 \includegraphics[width=8.985cm,height=4.413cm]{EAGLE2016BenefielSypniewski-img001.jpg} \newline
Fig. 1 Graffito from Pompeii [CIL 4, 5215]\newline

This graffito depicts a pair of gladiators, where the two figures are identified with their names and the number of
their victories. The inscription therefore includes both text and image. It was much more common in Pompeii for someone
to write a message on a wall than to sketch a drawing: people wrote their names, greetings to friends, quotations of
literature, and other types of messages. However, we do find a smaller, but not insignificant number of drawings also
inscribed on the wall plaster throughout the town. It is very rare to find a large scene, like the illustration of a
gladiatorial contest with athletes, musicians, and perhaps magistrates, sketched by hand on a funerary monument just
outside the Porta Nocera of Pompeii (CIL 4, 10237; D31 in Cooley and Cooley 2014; cf. also CIL 4, 10236 and 10238 drawn
nearby). \ More commonly, people made small sketches on the walls around them choosing from roughly a handful of
popular designs: \ heads in profile, boats, gladiators, birds, and geometric designs (Langner 2001).

\bigskip

Figural graffiti have provided us with several challenges as we digitize them for the Epigraphic Database Roma and as we
design a way to search for and retrieve such drawings via the Ancient Graffiti Project search engine. In this paper we
will discuss the challenges we face and some of the strategies we have developed in response. \

\bigskip

2. Our Material

First, a little background on figural graffiti and our sources for this data. \ In Herculaneum, we are fortunate that a
significant number of graffiti are still extant and in situ, as roofing has been reconstructed for many buildings to
protect them from the elements. \ Due to the fragile nature of wall plaster, however, especially in Pompeii, many
graffiti that were recorded previously and published in CIL 4, have now been lost. Much of our data, therefore, comes
from verbal descriptions of graffiti that have since disappeared. Furthermore, the different editors of CIL 4 and its
supplements used different methods to denote that a drawing was present, and their practices changed over time. Working
with this legacy data, therefore, presents a range of difficulties. \

\bigskip

2.1 Verbal Descriptions of Figural Graffiti Found With Text

\begin{figure}
\centering
\includegraphics[width=4.493cm,height=9.979cm]{EAGLE2016BenefielSypniewski-img002.jpg}
\end{figure}
A drawing could, for example, be described in the text field of an entry in CIL 4. This occurs in the entries below,
where three drawings of human heads [CIL 4, 2315-2316] and two drawings of gladiators [CIL 4, 2319] are described in
small italics, placed where the images occur alongside the textual inscriptions [Fig. 2].

Fig. 2 \ The entries of CIL 4, 2315-2316 and 2319, representing figural graffiti via brief description

\bigskip

The italics make clear that those descriptions are not part of the texts of the inscriptions themselves, which are
represented in capital letters.

\bigskip

\ \ This practice is common in the original volume of CIL 4, when it seems that the editors documented textual and
figural graffiti that were in close proximity, or that were in some way related to each other. In later supplements,
line-drawings for figural graffiti were sometimes included when the drawings and text were obviously understood as one
inscription, as shown in Fig. 1 above. \ Perhaps due to the complications with preparing and printing such
illustrations, however, it also remained common practice to represent figural graffiti with very brief description in
italics (e.g. CIL 4, 4822, 4823, 5264, 5275, 6624, 6672, 6889).

\bigskip

2.2 Figural Graffiti described in notes or apparatus

The most common strategy, however, for documenting figural graffiti in CIL 4 is by including brief mention of a drawing
in the editorial note that introduces a graffito or in the apparatus that follows it [Fig. 3]. \ \

\bigskip

 \includegraphics[width=10.227cm,height=4.075cm]{EAGLE2016BenefielSypniewski-img003.jpg}

Fig. 3 \ CIL entry for an alphabet near figural graffiti [CIL 4, 10711]\newline

Note that the editorial note above the entry mentions drawings nearby (novem galeas gladiatorias et parvum phallum), but
the text is presented without illustration. This mode becomes more common in the fascicles of CIL 4 produced in the
later twentieth century and so the figural graffiti from Herculaneum are usually represented this way (cf. CIL 4,
10532, 10568, 10586).

\bigskip

2.3 Figural graffiti omitted by CIL

A fourth possibility exists as well, namely, when figural graffiti were not even mentioned in CIL. In each of the
previous scenarios, the editors of CIL include a description of a figural graffito when it was close to a textual
inscription. In contrast, figural graffiti found in isolation tended to be excluded altogether due to the focus of the
Corpus on text. Fortunately there is now a useful resource devoted to figural graffiti: Martin Langner’s Antike
Graffitizeichnungen, a monograph and accompanying database of figural graffiti from across the Mediterranean. His
catalog includes some 600 graffiti drawings from Pompeii and 60 from Herculaneum, including 200 that are not mentioned
in CIL. In addition, whenever possible, Langner will provide a line-drawing of the graffito, either his own or one
found in an earlier source; therefore, his database includes many line drawings that are not included in CIL even when
a drawing is described. However, certain motifs are omitted from Langner’s catalog. While he does catalog the more
interesting Phalluskopfen examples, he generally omits simple drawings of phalli. He also leaves aside the decorative
elements of coronae and palmae, which are sometimes mentioned in CIL. This means that an accurate total of all figural
graffiti in Pompeii and Herculaneum can only be reached by working through the collections of both Langner and CIL. To
create the most comprehensive resource possible for figural graffiti, we include all known drawings in the AGP search
engine. \

\bigskip

2.4 \ Documenting extant figural graffiti

Since the verbal descriptions of figural graffiti provided by legacy data are limited and vague or exceedingly general
(e.g. caput), the best circumstance under which to digitize a graffito is when the drawing itself still remains extant.
\ In such cases, we will use any published data as a starting point, but we are also able to make our own editorial
decisions about the subject matter of the drawing, how to describe it, and its relation to any text that is nearby.
\ The material with which we are working, therefore, includes a range of different information about the figural
graffiti of Pompeii and Herculaneum: from brief verbal descriptions to line-drawings, to the best case scenario when an
inscription is still extant.

3.1 Challenges in working with text and images

Several challenges, therefore, arise when making decisions about how to edit and digitize figural graffiti. These can
depend on how a drawing may or may not relate to a textual graffito, whether or not a drawing is extant, and how to
interpret and standardize legacy data.

\bigskip

Three of our main questions are:

\begin{enumerate}
\item How to define an entry? (Where for example does one entry stop and another begin? Do we catalog series or clusters
of graffiti, or individual images? How do we account for or represent the larger context?)
\item How to describe a drawing? (Here, there arise issues both of standardization and of interpretation, or over
interpretation.)
\item How can we make drawings searchable? (Ideally, we would like to make it possible for users both to browse and to
locate specific images.)
\end{enumerate}

\bigskip

3.2 How to define an epigraphic entry?

One of the first challenges we face in working with figural graffiti is deciding how to define an entry, that is, to
consider whether or not multiple elements should be part of the same EDR record or should be given separate entries.
First, we must ask: can we be assured that certain elements were meant to be understood together? There might be an
issue of accretion or accumulation, where additional graffiti have been added subsequently. A related challenge is
then, if we create individual entries for separate elements, how do we avoid losing information about the relationship
among the graffiti? This collection of drawings including six textual graffiti illustrates our challenge [Fig. 4].

 \includegraphics[width=10.88cm,height=5.761cm]{EAGLE2016BenefielSypniewski-img004.jpg}

Fig. 4 Line-drawing displaying a collection of textual and figural graffiti [CIL 4, 8383-8386; EDR148730]; unpublished
sketch of Matteo Della Corte from the archives of the Corpus Inscriptionum Latinarum at the Berlin-Brandenburg Academy
of Science and Humanities.

\bigskip

You can see a number of different images here including a small gladiator with trident, a face in profile, leaves,
several animals, and geometric shapes as well as the name “Atini” and the greeting “\textgreek{Gelast`h qa~ire.”
}\ Such a collection raises many interpretive questions. What is the relationship, if any, between the figural and the
textual graffiti? How should that relationship be best represented? Fortunately, we have available with this sketch an
overall view of the spatial relationships of this group of graffiti. Because the CIL entries are focused on text, the
figural graffiti are associated with and described in the entry of nearby textual inscriptions. For this cluster, we
have decided, instead, to give each element on the wall a unique identifier. First, it is not clear that the texts are
clearly linked to any of the drawings. \ Secondly, if we create individual entries, the field for measurements in EDR
permits us to give the measurements for each individual element. Thirdly, EDR has provided an additional solution to
the issue of representing context with the use of hyperlinks to other nearby inscriptions, created by including EDR
record numbers in the apparatus field. Additionally, we have decided to upload a series of images to EDR, including
detail illustrations and the composite sketch of all graffiti, to give the context of the entire cluster and the
relationship of the graffiti to each other.

In an example from Herculaneum (Fig. 3, above), the entry for CIL 4, 10711 notes that in addition to a graffito of the
alphabet, a series of nine gladiator helmets and a small phallus were also drawn on the wall. During our field season
in Herculaneum in 2014, we were not successful in finding the graffiti of the alphabet or small phallus, but we did
locate eight of the nine helmets. Here too we have devoted a separate database entry for each helmet. \ By making
individual records, we have a unique identifier for each image, in the form of the EDR number, so that users can cite a
specific parallel precisely. \ Again, we can record the precise measurements for each image. \ Yet, since separating
each image can obscure how the images relate to one another in the group, as with the previous example, we also upload
to EDR an overall image of the group of helmets together for every individual entry (cf. EDR143634).

In these two cases, we are fortunate to have contextual data that informs our understanding of how text and image may
relate. More often, we are left with only legacy data, with brief mention of a figural graffito in the apparatus of a
CIL entry and without illustration. \ Yet, proximity does not always indicate a relationship between the text and
image. \ Indeed, there may be no relationship at all between the figural and textual graffiti; therefore, putting the
two graffiti in the same EDR record may suggest a relationship where none exists. Given these circumstances, we prefer
to create separate EDR entries for the text and the image and to use the EDR hyperlinks to note that each is found near
the other.

\bigskip

3.3.1 How to describe drawings?

A second challenge occurs when we must decide how much interpretation to offer when we describe a graffito for a
database entry. \ When CIL has included mention of a drawing, we generally incorporate that description directly into
our entry. With figural graffiti documented by Martin Langner, we must create a summary in Latin and when doing so, we
attempt to give as full as possible a description of the elements of the image. With this camel (Langner 2001, n.
1443), for instance, we offer a full description in Latin that accounts for all the features of the drawing: camelus
dromedarius cum cauda, lodicem gerens, ad dextram incedens [Fig. 5].

\bigskip

 \includegraphics[width=10.88cm,height=8.608cm]{EAGLE2016BenefielSypniewski-img005.jpg}

Fig. 5 Example of an entry for a figural graffito not found in CIL \ (EDR144514)

\bigskip

3.3.2 \ Questions of interpretation through description\ \

As one might imagine, issues of interpretation can arise even with simple descriptions of images. In truth, we have
encountered more difficulties with interpretation in the case of drawings that have been described by CIL. The first
example comes from a shop in Pompeii and represents text and a drawing [CIL 4, 8185]. \ The plaster has clearly broken
off, so we do not know if this was part of a larger scene. What remains are two lines of text and just one figure,
which would seem to be a drawing of a person facing forward and rendered with head and shoulders. CIL describes it
thus: herma muliebris prospiciens [Fig. 6].

\begin{figure}
\centering
\includegraphics[width=7.1cm,height=9.155cm]{EAGLE2016BenefielSypniewski-img006.jpg}
\end{figure}
Fig. 6 CIL entry for textual and figural graffiti [CIL 4, 8185]

\bigskip

Is this drawing clearly depicting a female? It is difficult to argue from either the hairstyle or the clothing that the
figure is female. Here, we can only assume that the editors of CIL identified the image as female because the text
above mentions the female name Fortunata. But are we sure the image is meant to illustrate the text? Or that the image
and text are meant to be read together? Since the figure is clearly not enacting the verb of the text, could this be
either Fortunata or Antonius?

The head of a woman, described with a textual graffito from the Suburban Baths in Herculaneum [CIL 4, 10676], raises
similar problems with verbal descriptions of figural graffiti. In this case, CIL does not reproduce an image of the
sketch; it only notes that the four-line inscription of CIL 4, 10676 appears below a drawing of a female (infra
mulieris imaginem). In this instance, too, the text nearby includes two names, one female and one male. The drawing has
appeared in multiple publications (Della Corte 1960, Deiss 1989, Maulucci 1993) [Fig. 7].

\bigskip

 \includegraphics[width=5.553cm,height=4.217cm]{EAGLE2016BenefielSypniewski-img007.jpg}

Fig. 7 Apograph of figural graffito referred to in the note at CIL 4, 10676 [Langner 2001, n. 309].

\bigskip

Again, we might question whether this figure should be identified as female. In fact, we not certain that we had located
the correct apograph for the drawing. There is considerable discrepancy between the description of the drawing in CIL
4, 10676 as female, with no mention of the long nose, and this line-drawing.

Another reason we suspected there might be a mistake was that Martin Langner had catalogued the drawing associated with
CIL 4, 10676, describing it as a “Phalluskopf.” There was no mention of gender. And he categorized this drawing among
several examples of drawings of heads with phallic features. The graffito is in a room that is sealed off, with no
access, so we were unable to view it in person. Eventually, however, a photograph published by Antonio Varone in his
recent two-volume work providing images of extant ancient graffiti (Varone 2012, 509) allowed us to confirm that this
is indeed the correct graffito drawing – somewhat above but also drawn partly through the text of CIL 4, 10676.

Neither description offered by CIL or by Langner, however, seems altogether satisfactory. There are no obvious markers
of female identity and even the description of Phalluskopf is less than transparent. \ Thus this one drawing has two
published descriptions that vary greatly and that each lead to a very different understanding of the graffito. What
should we then do in such situations? Do we repeat the identification of CIL? Or do we offer a less specific
description, merely labeling this a hominis figura? \ In the end, our solution is to offer our own description, which
is detailed but less interpretative, with an emphasis on specific features of the image that are readily identifiable.
\ We also document Langner’s description and CIL’s earlier identification in our entry for EDR, but we note our
hesitation with such identification by labeling the image: gryllus? (“caricature?”). \ We are aware, however, that we
also introduce an interpretation with the tentative suggestion this drawing might be a caricature. \ \

The issue of interpretation arises most often in relation to identification. Other examples concern identifying the
particular types of gladiators or the species of animals, who are assuredly quadrupeds but in some drawings could be
any type of animal with four legs (stags, boars, dogs). In such cases, our solution is to describe a drawing with more
generic, yet accurate, terms such as “gladiator,” without further specification, or “animal” rather than cervus, aper,
or canis. \ Similarly, if we cannot determine male or female, we prefer to describe the drawing as “facies hominis.” In
the AGP search engine we can then indicate possible but not certain identification with a descriptor, or tag, that
comes with a question mark: e.g. “stag?”

\bigskip

4.1 How to search for drawings?

\ \ The third challenge that we face is how to search efficiently for inscriptions that are not just text but either are
images or include images. \ In the AGP search engine, we aim to complement the capabilities of EDR by providing another
way to search for these non-textual, figural graffiti. \ Since we describe the content of the figural graffito in the
Textus field of EDR, it is possible for a user to locate a graffito drawing. However, with text-based searching, a user
would need to need to know the vocabulary used to describe the drawing. \ Would someone ever think to search for
“camelus” without prior knowledge that there is a figural graffito of a camel in Herculaneum? \ Similarly, if you
search for “gladiator”, the text field will give you results for all inscriptions that mention gladiators as well as
drawings where we have described gladiators. \ If, however, we’ve described the gladiator more specifically as a
“retiarius” or we have gladiatorial equipment, such drawings will be omitted from the list of search results.

\ \ We are therefore designing AGP with the capacity for locating figural graffiti through a two-prong solution: with
both browsing and searching possibilities.

\bigskip

4.2 \ Browsing capabilities in AGP

\ \ For browsing, we have defined nine broad general categories, which together cover all the types of figural graffiti
we have encountered so far [Fig. 8].

 \includegraphics[width=7.567cm,height=7.586cm]{EAGLE2016BenefielSypniewski-img008.png}

Fig. 8 Screenshot of AGP search engine showing browsing capabilities.

\bigskip

At this point, one can browse by choosing a category, which will return all examples in that category. So, for example,
the category of “Gladiators/equipment” will return sketches of individual gladiators, gladiators fighting in pairs, and
gladiator equipment such as helmets. The category of “Animals” will return all figural graffiti that include drawings
of animals. As we process greater numbers of figural graffiti, the results of these categories will become larger.

\bigskip

4.3 \ Limiting search results in AGP

\ \ It is therefore necessary to design a way to limit results, so that a user could find only gladiatorial equipment,
or graffiti depicting only boars but not graffiti depicting other animals. We are developing a method of filters and
tags that will allow a user to move beyond browsing. \ These filters will allow a user 1) to limit an initial return of
results, 2) to retrieve more specific results, or 3) to perform a secondary level of search. It will certainly be
helpful to refine results of an entire category to include only a subset of that category, for example, only pairs of
gladiators instead of all gladiators and their equipment.

To allow for this, we are creating a list of tags that we can apply to figural graffiti to allow for greater specificity
of searching. By using tags, we can also assign multiple terms to a single image, e.g. stag and dog. Our goal is
ultimately to enable searches by these tags as well, so a user can directly find all drawings with dogs. The search
capacity will allow a user to search the tags or the Latin description, so both “navis” and “boat” will return hits.
Again, standardization is necessary. We are currently developing a list of tags that is comprehensive and flexible
enough to cover all graffiti, but that includes a level of standardization so that the list of tags offers extensible
terms. We are also creating a system of filters to allow a user to limit the initial results or to move directly to a
desired graffito. So, a user can search all graffiti drawings and then limit the search results, for example, to find
all the drawings in a particular property [Fig. 9].

\bigskip

  [Warning: Image ignored] % Unhandled or unsupported graphics:
%\includegraphics[width=8.855cm,height=6.17cm]{EAGLE2016BenefielSypniewski-img009}

Fig. 9 Screenshot showing filters in AGP.

\bigskip

Or, it will be possible to do a broad search for all drawings of animals, and then filter to limit the results to find
what kind of animals are drawn in taverns, for example, but not public buildings or houses. \

\bigskip

5. Conclusion

\ \ This system of tags and filters is in the early stages of the design process. These are our proposed solutions for
confronting the challenges of working with text and image, and our ideas for creating a resource to complement the
strengths of EDR with search capabilities for characteristics that are specific to these heterogeneous, individualized
handwritten inscriptions.

%WHER IS THE BIBLIOGRAPHY??

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/44/Damnjanovic_et_al_EAGLE2016.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%

\author{valentina}
\date{2015-11-26}
\begin{document}
INTEGRATION OF MULTIMEDIA COLLECTIONS AND TOOLS FOR INTERACTION WITH DIGITAL CONTENT. \newline
The case study of the Archaia Kypriaki Grammateia Digital Corpus

Uros Damnjanovic¹, Valentina Vassallo¹, Sorin Hermon¹\newline
¹The Cyprus Institute - STARC\newline
(u.damnjanovic, v.vassallo, s.hermon)@cyi.ac.cy

Abstract

Supporting a discovery, use, and navigation of digital collections is a fundamental part of providing access and
encouraging inquiry, interpretation, and knowledge. In this paper we present our efforts to store and explore
multimedia collections of archaeological data. Particularly, the case study of the Archaia Kypriaki Grammateia
epigraphic collection is presented. Our work can be seen twofold. One aspect of our work is to provide a place where
the data coming from various sources can stored and accessed. Another aspect is to provide users with means to explore
this data. We argue that currently digital libraries are constrained by their webpage-based paradigm, thus not
providing the means for utilizing the full potential of the heritage data.

Keywords

Digital libraries, Data repository, Epigraphy, User interfaces, Data visualization, Interaction.

1. The world going digital

Nowadays digital information format found its ways to all the spheres of our lives. Cultural heritage is no exception to
this. Gigabytes of data are being created on an every-day basis. For some time now, digital libraries are used to store
and provide access to digital heritage data. However, existing digital libraries often fail to provide the ability to
iteratively explore items, compare data trends, and engender the wisdom that comes from exploring data in new ways.
Current digital libraries are limited by their inadequate webpage-based paradigm, and it is easy for even the most
experienced scholar to get lost [Bergstorm and Atkinson, 2009]. To overcome this issue, new generation of digital
libraries should emerge that will serve as an interactive knowledge environment in which knowledge is created by
interacting with the data. Digital collections often lack features for deeper quantitative and qualitative analysis,
and even very useful functions, such as the ability to annotate or bookmark content, are often not supported. After
digitisation, these collections are typically monolithic, difficult to navigate, and can contain text which is of
variable quality in terms of language, spelling, punctuation and consistency of terminology.

According to [Fast and Sedig, 2010] digital libraries are seen as a store of epistemic potential with interaction having
the role to reveal much of the hidden complexity. In [Buchel and Sedig, 2014] the authors demonstrated the role that a
set of interactions can play in supporting users' understanding of spaces and their formation of cognitive maps when
working with map-based visualisations. \ According to [Algee et al., 2012] interactive visualizations empower users to
discover meaning and patterns within digital collections using dynamic, interactive displays. One example of utilizing
the interactive graphics in a real world digital library is presented in [Hienert et al., 2012]. Authors developed
interactive visualization tools that support exploration of search queries and search results and help users in
formulating new queries. Another example, the application called DaisyViz shown in [Ren et al., 2009], tries to help
users acquire better insights of the data by enabling users to rapidly develop domain-specific information
visualizations without traditional programming. In [Nasar et al., 2011] authors proposed a conceptual framework that
uses interactive visualizations for managing personal collections of images and videos with focus on data re-finding
and improved filtering. INVISQUE [Wong et al., 2012] is a novel system designed for interactive information
exploration. Instead of a conventional list-style arrangement, in INVISQUE information is represented by a
two-dimensional spatial canvas, with each dimension representing user-defined semantics. Search results are presented
as index cards, ordered in both dimensions. Intuitive interactions are used to perform tasks such as keyword searching,
results browsing, categorizing, and linking to online resources such as Google and Twitter. An approach for
overview-first exploration of data collections based on user-selected metadata properties is presented in [Bernard et
al., 2012]. \ In a 2D layout representing entities of the selected property are laid out based on their similarity with
respect to the underlying data content. The display is enhanced by compact summarizations of underlying data elements,
and forms the basis for exploratory navigation of users in the data space. In [Nan et al., 2012] authors propose a
novel visual design termed “Whisper” to fulfil the need for tracing information diffusion processes in social media, in
a real time manner. BirdVis [Ferreira et al., 2011] leverages visualization techniques and uses them in a novel way to
better assist users in the exploration of interdependencies among model parameters. Furthermore, the system allows for
comparative visualization through coordinated views, providing an intuitive interface to identify relevant correlations
and patterns.

\bigskip

2. “A place where the past meets the future”: the STARC repository

STARC repository was initially developed to store the data coming from the daily work of the research group and as
content aggregator for various digital libraries projects: data from archaeological excavations, museums artefacts,
epigraphic corpora, etc. Data stored in the repository ranges from high resolution images, 3D models, texts, maps,
videos and audio resulting from various data acquisition procedures such as photogrammetry, laser scanning, 3D
modeling, photography, sketching, drawing and so on. \ Every item in the repository is described using a metadata
schema that depends on the type of data and according to the needs of the project. For most of the items, we are using
our own metadata schema (STARC metadata schemas) \ [Ronzino et al., 2012; Vassallo et al., 2013], that was developed to
thoroughly describe various aspects of data creation process. \ Once data is uploaded and available, it can also be
used in different systems. For example, the repository serves as source of data for aggregation procedures by which our
data feed the Europeana portal (http://www.europeana.eu/portal/) through different digital libraries projects (e.g.
Athena, CARARE, Linked Heritage, AthenaPlus, EAGLE). In order to facilitate the migration of data from heterogeneous
sources, the repository provides ingestion capabilities, that enable easy transfer of data from any available data
end-point. Another aspect of the back-end is the user management. It deals with providing different access levels for
different users’ groups. It also manages personal user space, where users can add their own information to the
repository and have access to it when exploring the data. Data annotations and personal collections are some of the
examples of how users can add information to the repository and use it for data exploration.

\bigskip

2.1. Tools for accessing data

We wanted to provide users with number of tools that they can use while exploring the repository. In traditional web
based system, navigating from one tool to another usually means that we click on a link and then move to another page
where the tool is available. Once the user is on a page s/he can start interacting with the tool by adding or modifying
data or by performing any action relevant to the selected functionality. The process of exploring the repository starts
by running the initial query (Fig.1). \

Once the results are returned and displayed, the exploration process can start. By default the map functionality is
available in the view. When a new tool is added to the screen, it automatically gets an access to all the data
available in the current search. It also gets notified on any other tools that might be active in the view, so the
necessary communication channels are established. Once the results are ready, there are number of filters available to
filter out the data, either by data type, collection. Search form can be used to filter out the data by typing the text
used to filter the data. As the user types, the data is automatically updated to show only those items that are related
to the text in the search field. The user can then run a new search repeatedly, and the search history functionality
will store all those searches for later reference. The map tool (Fig.1) is used to show information about the
geo-referenced data. It provides two main functions. One is that it accompanies the search tool. Whenever the new
search is ran, and the results become available, the map tool updates itself with new data, showing the

 \includegraphics[width=10.896cm,height=6.81cm]{DamnjanovicetalEAGLE2016-img001.png}

Fig. 1. The exploration process starts by running the search. Once the results are available the user can add or remove
exploration tools. By default map tool is present in the view. The map tool is used to mark the geo-referenced data on
the map, and also to perform geo-search by defining the region of interest.

geographical distribution of the search results. The user can then use the map to filter out the data by clicking on the
markers on the map. Also the map can also be used as a search tool. By selecting the search map mode, the rectangle
mask appears on the map, that is used to define the region of interest on the map. This region is then used as an
additional parameter for the search.

Annotation tool is used to attach the information to the objects in the repository and share it with other users, or use
it to facilitate more efficient search. The tool is used by selecting object from search results, then either by
selecting a region in the image files (Fig.2), or by selecting a point for 3D models, and by adding textual description
of the selected part.

 \includegraphics[width=10.896cm,height=6.81cm]{DamnjanovicetalEAGLE2016-img002.png}

Fig. 2. The annotation tool. The user first selects the object in the search results, then selects the region of the
object and then adds the annotation. The annotation can be accessed when the annotated object is accessed, and by
specifying in the search that the search should also include annotations. \

Once the search results are available, and the user selects the collection tool, the user can start adding and removing
the objects from the collection. Finally the collection is described and saved. Same as for the annotations, the
collection information is saved in the user’s personal space, and can be also used in search, by specifying to the
search that it should include the personal collections in the search. This means that the search will look not only for
objects descriptions and metadata, but will also look inside collection descriptions to try and match the query.

In order to help in exploring the datasets, there are statistics and visualization tools. Statistics tool shows the
basic properties of the repository. A visualisation tool is used to accompany the search tool (Fig. 3). Story creation
tool is another tool that supports collaborative user generated content. It provides users with an online document
editor, where users can write their own ideas, notes, insights about the data. The document created in this way is
stored in the repository and can be accessed in the personal workspace, and shared with the others.

\bigskip

 \includegraphics[width=10.896cm,height=6.81cm]{DamnjanovicetalEAGLE2016-img003.png}

Fig. 3. Visualization tool. Visualization tool is used to accompany the search tool by showing live visualizations as
the search results change. When the search is ran and the search results are available the visualization tools
automatically update itself and shows the visualisation related to the current search results.

\bigskip

\bigskip

3. Archaia Kypriaki Grammateia, the STARC Repository and the EAGLE infrastructure

One of the most interesting collection stored in the repository is the Archaia Kypriaki Grammateia Digital Corpus
[Pitzalis et al., 2012]. The collection is composed of Ancient Greek and Latin epigraphic texts produced within a time
span of circa 13 centuries (from the 7th century BC to the 6th century AD). The texts are attributed to Cypriot authors
or were produced in Cyprus. The corpus consists mainly of funerary or dedicatory epigrams published with their
translation in Modern Greek, critical apparatus and philological comments [Voskos, 1997].

The peculiarity of the digital collection consists in the fact that, beyond the epigraphic texts, it is composed of a
series of multimedia digital resources that describe the content in a multidisciplinary way: digital texts, images
related to the epigraphic supports, 3D representation of the inscriptions, video, audio files, and so forth (Fig. 4).

\bigskip

 \includegraphics[width=10.938cm,height=4.129cm]{DamnjanovicetalEAGLE2016-img004.jpg}

Fig. 4. Multimedia data constituting the AKG digital collection stored in the STARC repository.

Such a material needs to be described in the right way, in order to provide detailed information about all the elements
that constitute the collection. For this reason a specific metadata schema has been developed: it covers the
multidisciplinary research carried out on this material in a manner that users comprehend the complexity of such an
approach through access to heterogeneous types of information [Vassallo et al., 2013].

The metadata schema for Ancient Cypriot inscriptions integrates multidisciplinary information regarding the objects and
their multiple digital resources. It is the result of a research line developed within the group [Vassallo et al.,
2013] and of the assessment and comparisons of schemas, models and ontologies already in use in the digital epigraphy
field (e.g. TEI Epidoc, Dublin Core, CIDOC-CRM).

The metadata schema is organized in groups corresponding to different research areas clustered in wrappers and
sub-wrappers, in order to fully describe all information that the ‘asset inscription’ contains. This schema,
distinguished by its multidisciplinary structure, include all disciplines relevant for description and representation
of epigraphies: archaeology (investigating for example the context of the finds), philology \ (analyzing the text, the
writing style, the scripture), chemistry and geology (providing details on the material upon which inscriptions were
carved), conservation (giving information about the state of the artefact), visualization and museology (about the
museums and places of conservation).

Besides the inscriptions, since the Cypriot collection consists of their digital representations (pictures, 3D models of
inscriptions, videos, etc.), the metadata schema takes into consideration their descriptive features, their digital
provenance and other related information. For example, through the metadata it is possible to describe the digital
provenance of the 3D model of an inscription and to give information about the acquisition phase (the technique, the
tool used, the specification of the tool, the specification of the output, etc.) the post-processing (the operative
info, the file specifications, etc.) and the digital output (data format, software), according to the digital resource
obtained (Fig. 5).

The metadata schema for ancient Cypriot inscriptions aims at the following goals:

\begin{itemize}
\item to describe in detail the digital resources and its digital provenance, 
\item to provide related information about the context of the inscription, 
\item to enable harvesting to larger initiatives (e.g. Europeana, Ariadne).
\end{itemize}

\bigskip

 \includegraphics[width=12.435cm,height=13.573cm]{DamnjanovicetalEAGLE2016-img005.pdf}  Fig. 5. Diagram of the metadata
schema components and relationships.

3.1. From STARC Repository to EAGLE infrastructure

Beyond the functionalities developed for an enhanced users’ access to the content, the STARC repository plays as
aggregator for different digital libraries projects that in turn provide content to Europeana portal. The epigraphic
content of the Archaia Kypriaki Grammateia digital corpus is ingested through the EAGLE Project.

The “Europeana network of Ancient Greek and Latin Epigraphy”, is an EU funded project under the umbrella of the CIP-Best
Practice Network and whose aim is to bring together some of the most prominent European institutions and cultural
archives in the field of Classical Latin and Greek epigraphy. One of the aim is to provide Europeana with a
comprehensive collection of unique and curated online edition of historical and archaeological sources.

To encode inscriptions from different epigraphic databases, \ EAGLE developed a metadata format that assessed the
provider’s metadata structures and considered two sets of standards: TEI EpiDoc and CIDOC CRM. EpiDoc allows a full
description of the text of inscriptions while CIDOC CRM enables a further full description object-oriented,
re[FB02?]ecting at the same time the different souls of Epigraphy, the philological and the archaeological one. Content
providers use different models according to the fact their databases are more oriented towards the text of an
inscription or towards the archaeological object that bears the text. According to the epigraphers’ community needs,
EAGLE consortium developed the data model on the base of the two sets of standards [Liuzzo et al., 2014]. Finally, the
EAGLE data model is mapped to the Europeana Data Model (EDM), in order to be ingested into the Europeana portal.

The content provided by the EAGLE consortium, due to the transformation into the EAGLE metadata model, produce three
different groups of metadata according to the digital objects they are connected to: artefacts, documental
manifestation, visual representation. The data that are sent to Europeana belong to the category of the artefacts (the
inscriptions represented as texts, called by Europeana “Cultural Heritage Object” - CHO) and of the visual
representation (the images related to the inscriptions, called by Europeana “WebResource”). The distinction between CHO
and WebResource has been introduced in Europeana as a result of the introduction of the new EDM schema to provide the
users with a better navigation and data retrieval and to aggregate under the same umbrella all the resources available
for a Cultural Heritage Object, avoiding data duplication in the portal.

Most of the consortium data, as well as the majority of the existent epigraphic databases worldwide, are compliant to
Epidoc. This allows to perform a smooth mapping to the EAGLE data model, to have results that are aligned to the
project aims (e.g. the possibility to create dedicated groups of metadata associated with the category of the object)
and to simplify the involvement of other databases in EAGLE through the enlargement of the consortium.

As previously mentioned, the Archaia Kypriaki Grammateia digital corpus data integrated in the STARC Repository are
different from other epigraphic databases data. The study of the epigraphy is not text-centric. The text is one of the
representations of the artefact: the archaeological object, the support, the 2D image or the 3D of the object are all
resources that, even if connected, have their own identity and their own set of information. Moreover, the AKG is a
collection of texts accompanied by their translations, notes and commentaries as published in its last edition by
Voskos [Voskos, 1997], therefore it is a digital form of this important editorial printed work, to which are connected
all a series of digital resources that help to enrich the corpus.

Such a material and the metadata schema used for the description of this content converges in a difficulty to map our
metadata schema to EAGLE data model. This implies a further effort to integrate the content into the EAGLE
infrastructure for a compliant visualization with the other content and retrieval in the EAGLE and Europeana portals.
Concerning the issues connected to the mapping and integration of the AKG data into EAGLE portal, possible solutions
have been investigated and are currently under tests:

\begin{itemize}
\item the creation and use of relations that are able to aggregate all the resources under a specific resource
identified as reference item.
\item the creation of rules for the development of a script that will be able to map to the Eagle data model every time
according to the type of object we are dealing with. This implies also an editing of the metadata, creating single
items that have as head the information about the inscription (e.g. the Ancient Greek text) and to which are attached
all the other metadata sets concerning the related objects (e.g. the Modern Greek text, the image support of the
inscription, the commentary, the 3D of the support-inscription, etc.).
\end{itemize}
Even if the first solution would be much faster from a technical point of view, at the moment is under preparation a
test for the second solution. In fact the latter, even if it is time-consuming in terms of elaboration, seems to be the
best solution for avoiding data duplication and for having more cohesive data. \

4. Conclusions

We presented in this paper a digital data repository and showed number of innovative tools used to access and explore
its collections. One of the most interesting collection stored in the repository is the Archaia Kypriaki Grammateia
Digital Corpus composed of Ancient Greek and Latin epigraphic texts. We argued that by interacting with data in an
innovative ways can help users better understand data and stimulate knowledge creation. With the set of proposed
functionalities we also enabled exploration of the epigraphic text in a way that supports sense making, understanding
and collaboration. We also showed how the collection of epigraphic texts can be used within other projects by mapping
our data to the appropriate data formats. Next step is to set up an evaluation framework that will try to measure and
evaluate the contribution of each tool to the users daily tasks. We want to measure the benefits of using such a tools
by performing various evaluation tasks and measuring the user's performance.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/48/EAGLE16-lame-et-al-teaching.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Marion Lamé}
\date{2015-12-02}
\begin{document}
Teaching (Digital) Epigraphy

Marion Lamé, Ivan Radman-Livaja, Bruce Robertson and Federico Ponchio

Abstract

Although having the experience of directly manipulating an ancient text-bearing artifact is important for developing the
skills of an epigrapher – an experience that books and photo cannot replace – such access to primary sources is often
problematic. In this article we present our experience with teaching students to transcribe and interpret Roman
inscribed lead tags, using a Digital Autoptic Process (DAP) in a web environment, so to develop basic competences in
epigraphy and digital epigraphy.\footnote{\par \ \  We wish to acknowledge the research assistance of T. M. Crandall,
History student at Mount Allison University \ (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
\ M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
{\textquotedbl}La lettre romaine, entre inscriptions monumentales et écritures communes{\textquotedbl}. The following
students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.}

Keywords

Educational project, digital epigraphy, epigraphy, open access, primary sources

\begin{enumerate}
\item Introduction
\end{enumerate}
Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project \ “Pedagogical,
Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the Tesserarum Sisciae Sylloge (TSS) through
a Digital Autoptic Process (DAP)” we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

2. Epigraphic Primary Sources of Information

The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
found in Sisak (Siscia), one of the largest urban centers in south-western Pannonia. Most of them were found during the
dredging of the Kupa river before WWI. Since the dredging was localized in the very centre of the town, i.e. in front
of the old Roman port quarter, it would seem that all the tags come from a limited area. All of those tags are small
lead tablets, of a more or less rectangular shape, pierced with a hole so that the tag could be attached to the bags
containing the merchandise or to the merchandise itself with a small rope or a metal wire. They all carry an
inscription, most of the time on both sides. Those inscriptions are always, up to now, written in capital letters or
the older Roman cursive, sometimes even in a mixture of both. Most of the tags were reused several times and thus they
carry traces of older inscriptions (palimpsests). Those inscriptions usually follow the same model: one side of the tag
carries personal names, the other side carries an inscription mentioning the merchandise, most of the time in an
abbreviated form, as well as a price - normally expressed in denarii or fractions of the denarius - and quite often an
indication of quantity or weight.

A traditional print photograph, such as this one (No. 12582) drawn from the archives of the AMZ, limits the student’s
opportunity to learn to decipher these tags. Even if a professional photographer took this picture with a good quality
camera in a fully equipped photographic studio, the reading of the inscription stays riddled with uncertainties and a
researcher would have to check the tag de visu in order to offer a plausible transcription.

\begin{figure}
\centering
\includegraphics[width=2.944cm,height=2.752cm]{EAGLE16lameetalteaching-img001.jpg}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.944cm,height=2.82cm]{EAGLE16lameetalteaching-img002.jpg}
\end{figure}
Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like lana, pan(n)um, tvnica,
sagvm, p(a)envla, pal(l)a, palliolum, lodix, banata, and abolla, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\par \ \  Radman-Livaja 2007/11/13/14 and 2013b, 165-172 for analogous tags with Mócsy 1956,
97-104; Egger 1967, 195-210; Frei-Stolba 1984, 127-138; Schwinden 1985 121-137/93, 215-222; Römer-Martijnse 1990/97,
5-48; Feugère 1993, 301-305; Weiss 1991, 211-220; Paci 1995: 29-40; Bassi 1996, 207-216; Bizzarini 2005, 121-135; Buchi
\& Buonopane 2005, 43-51; Cresci \& Pettenò 2010: 42-110; Jacques \& Hoët-Cauwenberghe 2010, 295-317; Wedenig 2013,
237-246.}.

Some of the genuine difficulties of the autoptic process are already solved by a static picture in a print publication.
The tag has a given position, \ it is illuminated from one direction only and therefore it already suggests a reading
direction. A DAP should be able to confront the students to some of those difficulties and to overcome them alone.

In order to consider the value and usefulness of the DAP in this particular case study, the fact that some of the
students had not yet a real background in Latin paleography was actually more of an asset. Indeed, if untrained
students with few basic skills can use the digital edition as a specific medium without too much difficulty, albeit
under supervision of specialists (Digital Epigraphy, Ancient History, Archaeology, Epigraphy, \ Computer Graphics and
Digital Humanities) then the whole concept would appear as convenient and appropriate for similar case studies and
research.

3. Digital Epigraphy

In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g., No. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: http://www.labcd.unipi.it/laboratorio .}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\ Terras, 2015
writes about some of the OA issues: “While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.”}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in Lamé 2015, on a dispositive analysis
and the three fundamental systems of an inscription: writing system (wSystem), textual system (tSystem) and contextual
system (cSystem). Three tools compose the DAP: RTI web viewer, TSS viewer and MarkOut tool. The last two were
specifically designed for the digital edition of the corpus TSS. The following paragraphs present those three DAP
tools. Due to the material properties of the lead tags, it is impossible to capture the appearance of the letters with
a single image. When holding a tag it is necessary to change the light angle to enhance the different marks on the
surface. RTI is a computational photography technique that enables the interactive re-lighting of the observed object
from any light direction. This is accomplished combining into a single compact data structure a set of fixed camera
photographs taken under many different illumination conditions. The acquisition process is inexpensive and does not
require labor intensive steps, unlike 3d model acquisition. RTI visualization allows the user to virtually ‘tilt’ the
object and recreate different lighting conditions, mimicking part of the autoptic process. Another advantage is the
magnification provided by the high resolution camera employed. The widespread adoption in browsers of WebGL (Khronos
group 2009) recently enable RTI visualization on the web. We used the RTI web viewer developed by Visual Computing Lab,
ISTI-CNR. High resolution re-illuminable images allow the selection of the best view for each mark, thus allowing the
user to obtain an adequate readability of the lead tags through an intuitive interface. Another important advantage of
having the tags available online is that the students can easily compare a tag with any other tag already in the
database and compare occurrences of a letter in other already transcribed tags. Finally, the Web access facilitates the
discussion amongst students and experts worldwide.

The TSS viewer is used to give access to the raw and genuine data (115 photograms with different lighting angles) and
also to pick the best photo to be used in the linking tool MarkOut, while the RTI viewer can be consulted in parallel
to interactively relight the tag and examine problematic spots.

The drawing process consists in associating some heterogeneous elements of the wSystem and the tSystem. MarkOut is the
linking tool that allows to represent the graphico-textual (g/t) relationship between those two systems. MarkOut allows
the students to work on the wSystem of the inscriptions by drawing the signs over an image of the tag using the mouse.
The line is converted in a curve that can be later fixed using handles. Each mark, that represents a phenomenon of the
wSystem of the inscription, is assigned to some textual information such as a letter or a symbol (Unicode and/or XML
encoding), that represents a textual phenomenon of the tSystem of the inscription. This assignation creates inside the
MarkOut the g/t relationship used for query processing. The final result is an SVG file that can be easily parsed to
recover the linking information between wSystem and tSystem (e.g. browsing by glyph's shape and interpretation,
identifying glyph's position), and easily shared with anyone with a browser (e.g.: fig. 3.1). The linking tool MarkOut
is one of the editing components of the modular DAP. They fit into communication workflows\footnote{This raises, of
course, questions, most of them outside the purview of this article, about the relationship that evolves between
research and our modern societies. We invite the reader to peruse various initiatives such as Sabattini, 2006 on the «
valeur d'ancienneté » of cultural heritage, that participates to social well-being, the \ Civic Epistemologies
http://www.civic-epistemologies.eu and its Berlin charter, as well as Public History approaches
http://public-history-weekly.oldenbourg-verlag.de.} such as crowdsourcing, editing inscriptions, dialoguing with
experts, preserving and promoting cultural heritage and, \ as here, teaching some skills in epigraphy.

Fig. 3.1. MarkOut screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
Latin characters with the position and the shape of the writing: U+0045 with II (identified as E2 by students) and
U+004C with L (L1 or L5).

\begin{figure}
\centering
\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img003.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img004.png}
\end{figure}
Students were involved for a very short period of time: on average two hours per day over a couple of days only (in
Spring 2014 and in Fall 2015). This training was delivered in three steps (“Readings”, “Writings”, “Test \& Taste a
DAP”) during which there were given some academic documentation about the lead tags, and basic knowledge about digital
epigraphy (epigraphic message in a computer, dispositive analysis, existing projects, digital tools, XML and Unicode
encoding, DAP, and issues regarding digital representation of artifacts).

Students encountered their first epigraphic challenges with tag no. 12582 (Radman-Livaja 2014: cat 03.13, 347), found in
the Kupa river in Sisak and offered by Andrija Colussi to the Archaeological Museum in 1898. It is a rather typical
lead tag of an irregular rectangular form, pierced with a hole. Its size (22.7x22.8x2.4 mm) corresponds fairly well to
the average size of other specimens found in Siscia. Although its surface is rather damaged and shows clear traces of
erasure, the most recent inscription remains quite readable. However, many traces of an older record (or several older
records?), complicate significantly the transcription and the interpretation. The last inscription can be read,
nevertheless, but it demands a certain effort and quite a lot of experience in Latin palaeography:

Obverse (palimpsest)\ \ \ \ Reverse\ \  \ \ \ \ \ \ \ \ \ \ \ \ (palimpsest)

\begin{figure}
\centering
\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img005.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img005.png}
\end{figure}
Ate[323?]r[323?]\ \  \ S[323?]i[323?]i[323?]x[323?]t[323?]i[323?]\ \  \ \ \ \ \ la(na) p(ondo) i\ \ . . . s

ivs\ \  \ i \ \ onis \ n\ \ \ \ cor(ticea)\ \ \ \ . . v\ \

\ \  \ i . n i\ \ \ \ \ \ [4FE?] =$-$ £\ \ \ \  \ iir

The inscriptions on the obverse and reverse may not be contemporaneous and we may not affirm with confidence that the
individual named Aterius has something to do with the small quantity of wool mentioned on the obverse. The price is
rather low, since the value of the transaction appears to be only 1 sestertius and 1 dupondius, i.e. 6/16 of a
denarius.

Traces of older inscriptions are hard to read and interpret but several personal names may be surmised. Such a tag was
particularly interesting for our experiment. It is quite a challenging inscription, even for a skilled epigraphist and
paleographer; we therefore were anxious to find out how efficiently the students would tackle it. Thanks to the DAP,
the transcription was not utterly difficult for most students. Obviously, in the first instance, the interpretation was
beyond their means, but we were nevertheless impressed by their abilities to read it far more easily than expected. The
DAP definitely offers the possibility to work on such material online, without the obligation to inspect it personally,
at least in the first stage.

Thus, students had to face similar issues as scholars who first tackled those epigraphic finds. However, the students
had one significant advantage, despite their lack of experience and skill. This major advantage was allowed by DAP
technologies, which considerably improved their odds. Despite their lack of elaborate knowledge on the subject, we
managed to train them to develop some necessary skills in order to be able to read the inscriptions as well as offer
meaningful transcriptions, thanks to the help of digital facsimiles. It was indeed really gratifying to observe their
enthusiasm during the experiment as well as their delight when they realized they could do it. Naturally, despite the
DAP, whatever the information technology may be, such scientific analysis still requires skilled scholars, but with the
help of technology, digital editions may be thoroughly checked and amended far more easily. Besides giving an
opportunity to further study the primary source of information and amend the paper editions and publications, it also
allows university teachers to train future scholars and specialists, wherever their geographic location may be. In this
particular case, the epigraphic material is in Croatia but the analysis was done online by scholars and students from
Canada, France, and Italy. We therefore believe that digital epigraphy has a role to fulfill in university teaching but
that one needs to establish right protocols and rigorous methods in order to warrant access to documents and ensure
further development of such teaching methods aiming to improve the formation of future specialists.

\begin{figure}
\centering
\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
\end{figure}
4. Practicing Epigraphy

Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
transcription and drawings) and to classify the written signs according to the following chart. Such preliminary
classification gives a first idea of the main general shapes one can find on a lead tag. It is not a paleographical
study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering.

4.1 Drawing With Help (No. 12563)

\begin{figure}
\centering
 [Warning: Image ignored] % Unhandled or unsupported graphics:
%\includegraphics[width=7.329cm,height=4.404cm]{https://lh3.googleusercontent.com/HwLa5or_paTAyMS3Mu5XB3NucZXrDws6Jvc5WCC2ONHqBnBugWRtTF-YaKTLhJLhg6fShq3n_nG_37csBdKiKoSFA2tqVqhA0ovfyyT27h6KvGa-4slEmeURVSu_QG2AzjqTevFycw}

\end{figure}
No. 12563 is a rectangular lead tag of irregular form with a badly damaged and scratched surface. Nevertheless, the most
recent inscription is clearly readable while one may still discern traces of older inscriptions. An older inscription
might be visible on the obverse, perhaps series of numbers but its meaning remains unclear. Students have
transcriptions and drawings.

\begin{figure}
\centering
\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img007.png}
\end{figure}
Obverse\ \ \ \ \ \ Reverse

Aeli(i)\ \ \ \ \ \ \ \ l[323?]odicem

Tasti\ \ \ \ \ \ \ \ murtiolam[323?]

\ \ \ \ \ \ \ \ p(ondo) vi s

\ \ \ \ \ \ \ \ [4FE?] vi s

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{1.4399999cm}|m{0.518cm}|m{0.37700003cm}|m{0.361cm}|m{0.297cm}|m{0.361cm}|m{0.518cm}|m{0.486cm}|m{0.361cm}|m{0.30300003cm}|}
\hline
\centering 12563 (obv) &
A &
E &
L &
I &
T &
A &
S &
T &
I\\\hline
\centering Stud.1 &
A10 &
E2 &
L1 &
I1 &
T1 &
A10 &
S3 &
T2 &
I1\centering
\includegraphics[width=2.563cm,height=2.579cm]{EAGLE16lameetalteaching-img008.png}
\\\hline
\centering Stud. 2 &
A10 &
E2 &
L5 &
I1 &
T2 &
A10 &
S4? &
T2 &
I1\\\hline
\end{supertabular}
\end{flushleft}

\begin{figure}
\centering
\includegraphics[width=2.461cm,height=2.104cm]{EAGLE16lameetalteaching-img009.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.268cm,height=2.104cm]{EAGLE16lameetalteaching-img009.png}
\end{figure}

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\bigskip

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{1.393cm}|m{0.361cm}|m{0.40900004cm}|m{0.40900004cm}|m{0.439cm}|m{0.393cm}|m{0.37700003cm}|m{0.444cm}|}
\hline
\centering 12563 (rev)\centering
\includegraphics[width=2.898cm,height=2.656cm]{EAGLE16lameetalteaching-img008.png}
 &
\centering L &
\centering O &
\centering D &
\centering I &
\centering C &
\centering E &
\centering\arraybslash M\\\hline
\centering Stud. 1 &
L2 &
O3 &
D3 &
I1 &
C2 &
E2 &
M5\\\hline
\centering Stud. 2 &
? &
O3 &
D5 &
I2? &
C2 &
E3 &
M5\\\hline
\end{supertabular}
\end{flushleft}

\bigskip

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{0.58cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.361cm}|m{0.297cm}|m{0.40900004cm}|m{0.361cm}|m{0.518cm}|m{0.439cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.297cm}|m{0.34900004cm}|}
\hline
\centering M &
\centering V &
\centering R &
\centering T &
\centering I &
\centering O &
\centering L &
\centering A &
\centering M &
\centering P &
\centering V &
\centering I &
\centering\arraybslash S\\\hline
M3 &
V3 &
R7 &
T2 &
I1 &
O3 &
L2 &
A10 &
M2 &
P1 &
V3 &
I1 &
S5\\\hline
M3? &
V4 &
R6 &
T2 &
I1 &
O3 &
L5 &
A10 &
M5 &
P2 &
V4 &
I1 &
S5\\\hline
\end{supertabular}
\end{flushleft}
4.2 Drawing \& Deciphering Without Help (obv. No. 13053)

No. 13053 is an irregularly shaped rectangular lead tag with a damaged surface as well which makes the reading rather
uncertain. Besides the last, most recent inscription, one still discerns traces of older inscriptions. The reverse is
ignored. Students do not have neither transcriptions nor drawings.

\begin{figure}
\centering
\includegraphics[width=2.009cm,height=1.96cm]{EAGLE16lameetalteaching-img010.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.103cm,height=1.961cm]{EAGLE16lameetalteaching-img010.png}
\end{figure}
\begin{figure}
\centering
\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img011.png}
\end{figure}
Obverse \ \ \ (palimpsest)

Vale\ \  \ \ \ [4FE?] v[323?]i\ \

rius \ lis\ \  \ \ \ \ tiir \ r[323?] i

Martia\ \  \ \ \ \

The RTI file was presented to the students in an unusual reading position in order to reproduce the epigrapher’s initial
challenge of determining the proper orientation in which to read the lead tag. Students overcome this difficulty quite
quickly ({\textless}5 mn).

The last line, MARTIA, gave the student a hard time. Particularly the tail of the damaged S (end of the second line)
generates noise on the last letters.

\begin{flushleft}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{1.513cm}|m{0.37700003cm}|m{0.518cm}|m{0.361cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.283cm}|m{0.37700003cm}|m{0.46900004cm}|m{0.439cm}|m{0.518cm}|m{0.518cm}|m{0.50200003cm}|m{0.297cm}|m{0.961cm}|}
\hline
\centering 13053 (obv.) &
\centering V &
\centering A &
\centering L &
\centering E &
\centering R &
\centering I &
\centering V &
\centering S &
\centering M &
\centering A &
\centering R &
\centering T &
\centering I &
\centering\arraybslash A\\\hline
\centering Stud. 1 &
V2 &
A10 &
L3 &
E3 &
R7 &
I2 &
V1 &
S6 &
M5 &
? &
R7 &
T3 &
? &
Ignored\\\hline
\centering Stud. 2 &
V2 &
A10 &
L3 &
E2 &
R7 &
I1 &
V1 &
S6? &
M5 &
A10 &
R6? &
T3? &
I1 &
A2\\\hline
\end{supertabular}
\end{flushleft}
\begin{enumerate}
\item \begin{enumerate}
\item Palimpsest (obv. No. 13053)
\end{enumerate}
\end{enumerate}
The older, erased layers, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
g/t relationship between wSystem and tSystem was asked to students.

\begin{figure}
\centering
\includegraphics[width=6.313cm,height=2.207cm]{EAGLE16lameetalteaching-img012.png}
\end{figure}
5. Pedagogical Considerations

From the educational standpoint, this pilot project succeeded tremendously. The students who undertook this
transcription in place of the usual research essay were, on average, far more motivated than their peers, and took
great pleasure and pride in their work. One student, whose cumulative grade at university to this point was below
average, excelled in this project and said that this was what he imagined university studies were going to be like. Two
others prepared a presentation based on this effort, and it was accepted as a 15 minute lecture at the university’s
annual undergraduate research seminar.

Some of the uncommon qualities of these materials made them particularly able to be studied by undergraduate students
with little or no Latin. The transcription of the tags’ oddly shaped letters was a skill that any person literate in a
Latin-script language (but not necessarily in Latin) could acquire. Indeed any additional understanding of Latin
grammar and syntax afford a reader little additional advantage, given the tags’ copious abbreviations. The tags’
frequent and regular use of symbols and numbers meant that the latin-less student could nevertheless quickly begin to
‘read’ the tags, or at least glean information from them.

However much the student researchers’ enthusiasm compensated, there were some impediments which will need to be removed
before a large number of students can undertake this work. To unilingual Anglophone students, the research materials on
the tags are daunting because little of it is available in English. Our volunteers were mostly bilingual Canadians who
could read French as well as English.

6. Conclusion

Gaining expertise with material remains requires time. When a museum trains students, this time is often limited due to
constrained facilities and personnel. In addition, the epigraphic material may be fragile and delicate: the less these
are manipulated by hands the best it is for their preservation. We believe that a process similar to the one described
here will allow us and others to train far more, wherever they may be in the world, while well preserving the
materials. It will represent a useful intermediary step that develops both epigraphic and digital skills. However,
adding such a step will require us to renew, in some regards, the way we train learners and to reconsider who teaches
what and how. The benefits of this renewal, though, would be many: it would ensure the transmission of epigraphic
tradition and the acquisition of good digital skills, whether to train eminent specialists or to allow access to the
roots of societies to the greatest number.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/8/EAGLE_2016_submission_8_revised.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%

%all title info missing!
\title{}
\author{Diederichs}
\date{2015-11-17}
\begin{document}
\clearpage\setcounter{page}{1}\section[]{[Warning: Draw object ignored]}
\section{}
Abstract

\bigskip

No existing digital work environment can sufficiently represent the traditional epigraphic workflow ‘documentation,
analysis, interpretation, publication’ for a non-alphabetic writing system. Using the Maya hieroglyphic script, this
workflow will be transferred to a digital epigraphy. Digital methods and tools will be developed and reused in a
Virtual Research Environment to create a freely accessible, annotated textual corpus, including metadata on cultural
and object history and references.

\section[Keywords]{Keywords}
Classic Mayan, Digital Epigraphy, TEI, EpiDoc,

CIDOC-CRM, Metadata Schema, Open Science

\bigskip

\bigskip

1.1. Digitalizing the Epigraphic Workflow

In order to develop a digital work environment that represents the traditional epigraphic workflow for documenting and
studying inscriptions in a non-alphabetic script, the individual steps of this workflow need to be technically
implemented using methods and tools from the Digital Humanities. The workflow begins with the documentation of the text
carriers and compilation of descriptive data; proceeds to epigraphic analysis, including sign classification and
transliteration and transcription of the texts; continues with morphological segmentation and linguistic
interpretation; and concludes with translation and digital publication of the inscription [Diederichs 2015].

The Virtual Research Environment (VRE) TextGrid initially provides the necessary framework for executing the workflow
that technically facilitates documentation and digital registration of the text carriers in the form of images and
drawings, as well as annotation of the objects with metadata. TextGrid offers input forms, annotation tools, and a
Text-Image-Link-Editor, and provides long-term storage, and free access to the data in an open web repository [Neuroth
\& Lohmeier 2011].

Although more detailed steps in an epigraphic workflow, such as detailed linguistic and epigraphic analysis of primary
texts, can be technically carried out within a VRE such as TextGrid, they still necessitate creation and adaptation of
their own XML-based metadata model, as well as annotation schemas for object and textual contexts [Prager 2015]. Using
this approach, the Maya hieroglyphic script will for the first time be able to fulfill a central requirement of corpus
linguistics, namely machine readability [McEnery \& Wilson 2001].

Figure 1.1. provides a schematic illustration of the following points that represents in great detail the steps in
epigraphically analyzing Maya inscriptions, a process which can also be used in modified form to study other
non-alphabetic writing systems. The modulated VRE is thus applicable not only to the study of Maya texts, but also to
researching texts composed in other hieroglyphic, cuneiform, or linear writing systems.

\bigskip

\begin{enumerate}
\item Identification of hieroglyph blocks by alphanumeric classification 
\item Original spelling
\item Reading order of individual signs 
\item Number of signs within one block 
\item Identification and isolation of signs in one block 
\item Classification of signs based on the sign catalog of Eric Thompson [Thompson 1962]
\item Classification of signs in one block [Thompson 1962] 
\item Transliteration of individual signs 
\item Description of sign function (syllables, logographs)
\item Transliteration 
\item Broad transcription
\item Morphological segmentation 
\item Morphological analysis 
\item Determining congruence between block- and word-boundaries 
\item Determining syntactic function 
\item Determining sentence constituents 
\item Translation of single words 
\item Literal translation 
\item Rough translation \ 
\end{enumerate}

\bigskip

 \includegraphics[width=10.881cm,height=7.214cm]{EAGLE2016submission8revisedx-img001.png}

\bigskip

Fig. 1.1. Visualization of the important steps in the workflow of epigraphic analysis, drawings by Christian Prager,
2015

\bigskip

\section[1.1.1. Conceptual Questions]{1.1.1. Conceptual Questions}
When developing and technically implementing the digital epigraphic research environment, the major focus is on pursuing
the following objectives: 1) creation of a markup schema to represent these steps (which in practice are interconnected
and may also occur parallel to each other in the form of alternative interpretations);

2) defining of the precise requirements that a data structure must fulfill in order to sufficiently represent complex
i.e. non-alphabetic textual data for epigraphic analysis and research.

\section[2.1. Prerequisites for Making a Non{}-Alphabetic Writing System Digitally Accessible]{2.1. Prerequisites for
Making a Non-Alphabetic Writing System Digitally Accessible}
In order to make syllabic or logo-syllabic writing systems accessible for corpus linguistics, for instance to carry out
a corpus analysis as part of the basic lexicography required to compile a dictionary of Classic Mayan, the phonology,
morphology, syntax, semantics, and pragmatics of the language in question must be captured, marked up, and saved in the
corpus. As such, the basic characteristics that distinguish the graphematic lexicon of syllabic and logo-syllabic
scripts from those of alphabetic writing systems must be taken into account. Thus, methodological and technical
prerequisites must be fulfilled in order to be able to digitally represent and analyze a non-alphabetic script, such as
Classic Mayan, for the purposes of epigraphic analysis.

\section[2.1.1. Registering Object Data ]{2.1.1. Registering Object Data }
When designing the digital epigraphic work environment, one must also account for the current state of decipherment of
the script and language that are to be documented. Thus, Classic Mayan presents great challenges, as at least one third
of the known graphemes have not yet been completely deciphered. Even when a lexeme can be read, its etymology or
semantic domain is often unknown, and in some cases not even its lexical class can be determined, which causes
difficulties when attempting to read and understand a text. In order to decode an unknown sign, word, or sign sequence
in its respective context of use, for instance, semantic fields are generally investigated by means of applied
substitution. From a structuralist perspective, this method investigates the paradigmatic relation between two
linguistic entities as indicated by whether or not they appear in the same context [Bußmann 2002]. This step yields
results when sufficient quantities of data can be digitally compiled, marked up and investigated using corpus
linguistics.

The VRE currently being planned is the laboratory for this process, where new decipherments will be obtained and
existing readings will be tested against primary source material while also considering all of the contextual data. The
context of the contents that is referenced during decipherment may refer to the text carrier itself. For this reason,
registering data on the object on which the text is written is just as important as documenting the text itself.

The decoding of Ugaritic cuneiform at the beginning of the twentieth century provides an example of a successful
decipherment that resulted from studying the relationship between object and textual data. Scholars suspected that the
word “axe” occurred in repeating, brief sign sequences on inscribed bronze axes. Building on this hypothesis, they
determined the phonetic values of individual signs using data on the Canaantic language and completely deciphered the
script [Day 2002]. Such investigations enable initial association of signs with both content and function in the
relevant language so that the signs may later be grammatically, morphologically, semantically, etc., defined and
eventually deciphered. This work also facilitates more precise typological classification of a writing system
[Gronemeyer 2015].

Like the relationship between writing and the object, the relationship between text and image must also be coded.
Associated depictions can potentially illuminate the contents of an inscription. Working from this analytical basis, a
writing system can be registered in its respective contexts of use in order to derive semantic meanings and, ideally,
linguistic decipherments. As such, contextual information concerning the text carrier as an object must be recorded,
because its physical features may influence the text’s contents and arrangement. In this respect, the size of the
object, for example, may be just as influential as the material itself.

For example, the text carrier may influence scribal economy. The recent discovery of an inscription on very hard jade in
Nim Li Punit (Belize) illustrates how the use of this particular writing surface can lead to a reduced spelling,
meaning that the scribe omitted final syllables or used very simple sign variants that were almost geometric.

The form of a text carrier and its function also influence the text. Different lexemes may thus be determined for
describing round or quadrilateral altars, with the result that conclusions may be drawn concerning the semantics and
possible linguistic decipherments. Physical description of the text carrier as an object and its linguistic,
historical, and cultural contextualization are indispensable for deciphering a previously unknown text or analyzing its
arrangement. These metadata must be consulted in order to grammatologically and linguistically analyze the text’s
contents.

Modelling an object biography and culturally contextualizing the texts must be initially addressed with information
technology in order to facilitate epigraphic analysis of the text’s contents in its extra-linguistic, cultural context.
Thus, it is necessary to create an object metadata schema that captures such information concerning the language’s
cultural area and makes it available for use.

For these purposes, an epigraphic object metadata schema that compiles and represents extra-linguistic information about
the inscriptions in an ontological structure must be created, in addition to an XML-based analytical metadata schema
for linguistics.

The CIDOC-CRM standard for documenting cultural heritage offers a broad foundation that can be supplemented with other
standards, such as Dublin Core or the SKOS vocabulary. In addition to taking advantage of the comprehensive
understandability of CIDOC-CRM, the scheme also requires to reuse data and to connect them with data from other
research projects in the spirit of Linked Open Data [Diederichs 2015 and Prager 2014].

\section[2.1.2. Making Linguistic Data Accessible]{2.1.2. Making Linguistic Data Accessible}
An interdisciplinary and widely accepted standard that takes into account both generalized and special characteristics
of writing systems beyond those of a single script must be developed in order to be able to satisfactorily annotate
records from various writing traditions that use syllabic signs and logograms to represent language.

In most cases, the scripts annotated and encoded in TEI and also those prepared for epigraphic analysis in EpiDoc are
alphabetic, and most of them are linear and arranged in rows. Non-alphabetic writing systems, such as Egyptian, various
cuneiform systems, or Hieroglyphic Luwian, arrange their graphemes principally in groups or blocks, and only
secondarily in rows or columns, with a high rate of metathesis [Lacau 1903] in order to optimize the use of space.
Furthermore, many of these writing systems demonstrate a high degree of allography, which is no longer apparent in a
pure transliteration. Various desiderata become apparent when an epigraphic project attempts to use the standards
provided by TEI and EpiDoc to edit texts composed in these scripts.

Annotating texts composed in such writing systems that use only alphabetic transcription is insufficient; instead, the
original spelling of a lemma should be represented. This approach facilitates studies of paleography or sign usage
across time and space. In addition, it can reveal preferred sign arrangements within a block or preferences for
particular signs that correspond to the material used or to the subject of the text, for instance.

Thus, when creating a standard for non-alphabetic scripts, object metadata and annotations for orthographic and
linguistic analysis must be taken into account, and the creative process should also promote discussion of
terminological and typographic conventions for these annotations [Sachse \& Dürr 2015]. The goal is to create a generic
markup model that can be implemented regardless of the particular script or language in question. The inscriptions of
the Maya hieroglyphic writing system alone feature attestations of various vernacular languages [Lacadena \& Wichmann
2002], a phenomenon which raises several preliminary questions: what exactly are the demands, and what are the goals of
an epigraphic and linguistic annotation? Where do possibilities and limitations exist for annotating these writing
systems using established standards, such as TEI or EpiDoc? Existing standards have to be modified to some extend and
possibly new standards have to be created, to overcome these limitations.

\bigskip

\section[2.1.2.1. Representing the Primary Text Source]{2.1.2.1. Representing the Primary Text Source}
When defining a text, acknowledging an interpretation of the source text itself is unavoidable. The meaning of the
“original text” that resides in the “source text” is an important point of discussion in epigraphy. Thus, the
annotation does not contain a guiding text in the conventional sense, because it is merely reconstructed using all
given text information and one’s own conclusions and interpretations, and then represented with the aid of alphabetic
transliteration and transcription. In this respect, one should always leave open the possibility of separating primary
data (in the case of hieroglyphic texts, photographs of the text carriers, for example) and secondary data (e.g.
drawings or interpretive annotations) [Stührenberg 2012]. This strategy can be implemented using stand-off annotations
for data. Similarly, such an annotation should permit multiple descriptions, i.e. alternative statements concerning the
data [Stührenberg 2012]. For example, sign sequences in Classic Mayan can be variously analyzed depending in a
particular vernacular context [Gronemeyer 2014]. Only in this manner can various ways of thinking be appropriately
recorded and relayed to the scientific community for discussion.

A key objective of the XML-based markup of hieroglyphic, cuneiform, or linear texts should thus be to represent the
original spelling and arrangement of the signs in their respective contexts. A linear transcription alone cannot
represent the original text or primary source in its entirety, as many details remain undocumented. For example, signs
in Maya hieroglyphic texts are not arranged according to their literal reading order, but instead in spatially
distinct, square or rectangular units (so-called “blocks”), each of which in most cases corresponds to a word or
morpheme sequence. A detailed markup of the original text is therefore of great importance, particularly for partially
deciphered and undeciphered graphemes and writing systems in general. In such cases, an alphanumerical or numerical
nomenclature is often used to refer to the signs in order to carry out a corpus linguistic analysis of the texts. The
arrangement of each block, as well as the text and its position on a text carrier, should be documented as well. To
fully understand a writing system – i.e. the language and information expressed in it - a detailed representation of
the primary source and its context must be a key objective in the digital markup of documents. When studying complex
writing systems (respectively non-alphabetic writing systems), digital documentation of the original spelling using
annotation standards like TEI is a basic prerequisite for conducting a detailed graphemic and graphetic analysis of the
relevant script and for providing a basis for a linguistic and corpus linguistic investigation. This need represents a
significant desideratum in epigraphic research, and it also constitutes a core pillar of computational linguistics
[McEnery \& Wilson 2001].

In an interdisciplinary effort, epigraphers and experts of digital markup-languages alike need to discuss methods for
investigating syllabic and logo-syllabic writing systems using the XML-based standards like TEI or EpiDoc. When doing
so, they should discuss the following points of emphasis, among others:

\bigskip

\section[2.1.2.2. Graphetics ]{2.1.2.2. Graphetics }
Graphetics concerns the formal structure of linguistic units and the structure of texts [Crystal 1997]. Classic Maya
hieroglyphic inscriptions, for instance, may display a very high degree of variation in writing styles for arranging
texts, a phenomenon which is deserving of investigation. In this manner, various systems of notation are studied in
their individual, social, and typographic aspects, for instance [Bußmann 2002]. Similarly, in paleography, script
decipherment is examined from graphetic perspective [Bußmann 2002]. The arrangement of texts, for instance, different
grapheme or block sizes of text fields, or different styles of hieroglyphic writing (equivalent to alphabetic font
style and size) are analyzed.

Research addresses the degree to which attested variation in the composition of hieroglyphic texts may be ascribed to
specific scribal schools or to regional forms of expression, rather than to differences in the meaning of the signs’
contents.

\bigskip

2.1.2.3. Graphemics

When analyzing the contents of these texts, it is important to investigate meaningful variation underlying the manner in
which the writing is arranged and to thereby identify distinct units [Bußmann 2002]. Graphemics is devoted to this area
of research, and consequently to exploring the meaningfully distinct features of signs.

\bigskip

\section[2.1.2.3.1. Allography of Graphemes ]{2.1.2.3.1. Allography of Graphemes }
A very important example of a subject of graphemic research is provided by allography, a phenomenon which describes a
1:n relationship between a grapheme and its various graph representations [Crystal 1997].

The process of establishing a suitable annotating-tool thus must be oriented toward a series of questions: What is the
significance of allography for the respective grapheme inventories? Are allographs annotated in the transliteration or
transcription of the respective writing systems?

For example, in Maya writing, there are over twenty different allographs for the vowel sign {\textless}u{\textgreater},
which is used e.g. as the 3rd person ergative pronoun u- “he, she, it”. Additional research questions may comprise 1)
the reading order of the signs in context; 2) the existence of word separators or other graphic aids used to
differentiate between meaningful units or words; 3) multiple possibilities for establishing a reading order of signs
within a block that are equally meaningful (with alternatives documented or marked by epigraphers); 4) the reading
order of meaningful units within the texts; 5) variations or violations of these orders (and if so, with a markup as
well); 6) cases where images appear with the text and their possible relation; 7) integration of texts into the images
or vice versa.

\bigskip

\section[2.1.2.3.2. Graphotactic Strategies of a Writing System ]{2.1.2.3.2. Graphotactic Strategies of a Writing System
}
For analysis, it is important to convey the graphemics of a writing system, which are lost in an alphabetic
transliteration. Graphemics concerns sign function, or graphotactic strategies for constituting a meaningful unit or
word. Representing the reading order of signs in the text and in each block is similarly imperative. To
graphotactically link individual signs in a hieroglyphic block in Maya writing, various representational rules may be
used (see Fig. 1.2.), e.g. affixation, infixation, conflation, or superimposition [Zender 1999]. Additional typical,
functional traits of logo-syllabic writing systems include underspellings, diacritics, phonemic complements and
indicators, semantic classifiers and determinatives, and sign class convergences. However, these characteristics may
have very diverse graphotactical manifestations in different writing systems, which nonetheless need to be standardized
and communicated in a digital structure in order that researchers may thus take note of various interpretations under
discussion.

\begin{figure}
\centering
\includegraphics[width=10.816cm,height=7.054cm]{EAGLE2016submission8revisedx-img002.png}
\end{figure}
\section{}
\section[Fig. 1.2. Some possible sign combinations in Classical Mayan hieroglyphs [after Zender 1999{]}, drawings by C.
Prager 2015 ]{Fig. 1.2. Some possible sign combinations in Classical Mayan hieroglyphs [after Zender 1999], drawings by
C. Prager 2015 }

\bigskip

\section[2.1.2.4. Signary: Sign Classification and Sign Catalogs ]{2.1.2.4. Signary: Sign Classification and Sign
Catalogs }
The graphemic lexicon provides a central reference for marking up texts composed in non-alphabetic writing systems. In
the case of the Maya script, signs whose phonemic reading is unknown or unconfirmed are denoted using reference systems
from various classification catalogues (e.g. Thompson 1962). Use of question marks for indicating unclear readings, for
example, should be avoided, because such signs impair machine readability and sometimes represent control characters or
part of an escape sequence. In this context, it should be determined what significance sign classifications and
nomenclatures have in their respective branches of epigraphic research. Also their role in transliterating and
transcribing texts in the respective writing system, as well as in sign classification is to be examined. Sign
inventories constitute a central authority in an epigraphic VRE, and they need to be modified or supplemented in
accordance with the ever-changing state of research as soon as newly discovered material is entered and signs are
identified that had not previously been isolated or identified as discrete graphemes.

\clearpage\section[2.1.2.5. The State of Decipherment and the Readability of Texts ]{2.1.2.5. The State of Decipherment
and the Readability of Texts }
In the case of unreadable or only partially deciphered writing systems, the question in the forefront is the state of
decipherment of the relevant writing system and the handling of undeciphered signs or text passages. As archaeological
artifacts, texts are also subject to processes of decay that inhibit visible legibility. It has several implications:
1) the treatment of undeciphered signs or text passages; 2) the indication of physical gaps (as texts are also subject
to processes of decay that inhibit visible legibility) and their marking in the transliteration and transcription of
the text; 3) the treatment of hypothetical, yet plausible, readings of singular signs or passages; 4) the handling of
alternative readings of the same passage; and 5) a review to what extent the EpiDoc Metadata standard, into which the
Leiden Conventions have also been incorporated, is able to capture these features of readability and alternative
interpretations of inscriptions.

These and other characteristics of non-alphabetic scripts, which to a certain extent reach beyond the epigraphic
problematic of alphabetic writing systems, illuminate a desideratum here and the need to formulate ideas that
contribute to the design and modelling of appropriate XML-based metadata schemas.

\bigskip

3.1. Open Science Strategy

A Digital Humanities project working in the spirit of Open Science fundamentally intends the work environment that it
develops to be reused. As such, a VRE for non-alphabetic writing system such as that of the Classic Maya should be
developed as generically as possible, in order that structural interoperability may ensure that they can be reused by
similar epigraphic projects.

In this respect, any metadata schema that is created should be widely intelligible and ensure the data’s syntactic and
semantic interoperability by using common, given, XML-based annotation and metadata standards, like TEI and EpiDoc, and
controlled vocabularies, such as SKOS, that are already established in the Digital Humanities [Diederichs 2015].

Another objective of database projects in the Digital Humanities should be to ensure open access to their data and
metadata, and thus to maximize usage of their research by maximizing user access to it. As all current and future
research and innovation stands on the shoulder of giants, an efficient system for broadly disseminating and allowing
uninhibited access to the project’s research (raw data and metadata), as well as for guaranteeing the contents’
productive reuse, must be ensured through the use of free licenses [DFG 2014].

As such, all the contents of a database and data infrastructure that are being planned will be made available under
so-called “Open” licenses (Open Access, Open Source, Open Methodology, Open Data, etc.), with the goal of pursuing a
comprehensive Open Science strategy [Diederichs 2015].

All methods and the process of developing tools must thus be carefully documented to maximize reuse of all schemas and
software. In particular, all research information and results should be made available not only to the scientific
community, but also to the general public, to contribute to the digital safeguard and dissemination of humanity’s
cultural heritage.

\bigskip

\section[4.1. Summary ]{4.1. Summary }
The creation of digital infrastructures and metadata schemas that facilitate recording and further analysis of the
appropriate annotation for non-alphabetic writing systems, including their respective cultural contexts, is a
desideratum in the Digital Humanities that needs to be addressed.

The recommendations articulated here for an object database, as well as for a structure for recording epigraphic and
linguistic data, can make a significant contribution to understanding how traditional epigraphy can be transferred to
digital form. In working towards this goal, methods and tools from the field of Digital Humanities will be suitably
adapted to the needs of epigraphy, and new methods and structures from the digital world will find their way into
digital epigraphy.

\bigskip

\bigskip

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/9/Latin epigraphy for the visually impaired.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{}
\author{Francesca Licordari}
\date{2015-11-19}
\begin{document}
1. Latin epigraphy for the visually impaired.

\ \ \ \ New technologies to favour universal accessibility

\bigskip

Francesca Licordari

Museo Nazionale Etrusco di Villa Giulia, Roma

francesca.licordari@beniculturali.it

\bigskip

Abstract

\bigskip

The problem of accessibility to works of art for the visually impaired can be extended to a system such as Latin
epigraphy, which presents great difficulties since the engraved letters are not clear to the touch. The technology of
3D printing allows the creation of materials that make it easier to read epigraphic texts in following the “design for
all” principle demonstrating an economic, rapid technique that is easy to implement and reasonably robust.

\bigskip

Keywords:

\bigskip

Accessibility, visually impaired, 3D printing, design for all, Braille, epigraphy

\bigskip

1.1. Introduction

\bigskip

The Declaration of the Rights of Deaf-Blind Persons, adopted by the UN in 1979, provides that every blind or deaf person
has the right to enjoy the same privileges, guaranteed to all people and to have their aspirations and abilities
recognized and respected. These principles are reaffirmed by the UN Convention on the Rights of Persons with
Disabilities of 2006, which states (Art. 30) that the Member States must recognize “the right of persons with
disabilities to take part on an equal basis with others in cultural life and should take all appropriate measures to
ensure to persons with disabilities:

(A) enjoy access to cultural materials in accessible formats;

(B) enjoy access to television programs, films, theater and other cultural activities, in accessible formats;

(C) enjoy access to places of cultural activities, such as theaters, museums, cinemas, libraries and tourism services,
and, as far as possible, to monuments and sites of national cultural importance.”

The problem of accessibility and enjoyment of works of art by the visually impaired has long existed. Sculptures, which
are sometimes designed to be touched, and sometimes are permitted to be touched, provide the possibility for blind
viewers to benefit directly from the experience of the original artwork. For two-dimensional paintings this is of
course not possible. It is necessary to create three-dimensional resin reliefs. This same issue has recently been
understood to exist also in other artistic disciplines, including that of archaeology. This of course includes
epigraphy, which is the study of all those materials which have written inscriptions. Most relevant to this is Latin
Epigraphy, simply because it is the one closest to our language and which uses the same graphic signs of our own
writing.

The bibliography in this area of study is unfortunately not very comprehensive, especially as regards the epigraphic
field; but advances are even now being made. In Italy, the most important studies are conducted by the “Museo Tattile
Statale Omero” in Ancona, which, being a museum of statuary, is mainly engaged with reproductions and the accessibility
of tactile sculptures. The same work, involving the reproduction of paintings, is being carried out by the “Museo
Tattile Anteros” in Bologna. It is no coincidence that Italian studies in this field are the most important done at the
international level. This is mainly due to the fact of Italy’s great artistic and epigraphic heritage, which makes it
the country where the problem is most profoundly felt.

The results of this paper are derived primarily from field studies, and also through various assistance and support
provided in partnership with people of both limited vision and complete visual impairment who tested the strengths and
weaknesses of the materials which will be referenced in this paper.

\bigskip

\bigskip

1.2. Problems in Understanding Epigraphy

\bigskip

The field of epigraphy, always considered highly specialized, can be of great fascination to people with visual
impairment. Some may have studied ancient languages at school, especially Latin, and so feel the need to have practical
examples for the understanding of the subject; and some may be attracted out of curiosity. It is obvious that
accessibility to the inscriptions for such people involves a number of difficulties, and not just that of understanding
the texts, which are often full of abbreviations. There is, as well, a certain percentage of the blind who never
learned to read and write. Not infrequently, they may only know capitals, but not cursive characters. The inscriptions,
which appear so sharp to us that we are often tempted to touch them with our hands to follow the course of the
incisions, actually create complicated problems for the blind.

In this regard it should be noted that the sense of touch is neurologically and physiologically more suited to perceive
what is protruding from a surface, rather than to what is engraved upon it. This is a fundamental difference from the
sense of sight, which is suited to perceive contrasts in color and between light and shadow, regardless of the
technique used to create them (reliefs or incisions). The groove of a letter, even if deeply incised, is not always
easily distinguishable to the touch, while a written relief almost always is. Therefore, to allow for the visually
impaired the proper accessibility to an inscription, the original needs to be touched while, however, exploiting
resources to make transliterations in relief and Braille.

The creation of panels and captions in Braille can help in the understanding of a text, but it can not be the only
solution, as only a percentage of the blind understand Braille. A prerequisite for this is the instruction be given at
a young age, something that unfortunately does not always happen. With widespread computer usage, the majority of blind
children learn to write on the computer, which supplies speech synthesizer without the need for Braille. In this way,
the art work is losing its esthetic value.

A visually impaired person, however, has normally never learned Braille because with the right font size and adequate
lighting he can read standard writing - albeit with difficulty. In many cases, their blindness occurs in advanced age,
or as a result of illness or accident. In such situations, the writing that is understood had been taught in their
school years.

As Braille graphically occupies comparatively more written space, there are distances and dimensions that must be met
and therefore their characters cannot be made smaller as can be done with printing presses or computers. It follows,
then, that the explanations in Braille accompanying museum art works need to be more concise, in large part because
their reading needs to be done with both hands, making it uncomfortable to stand reading the texts for very long.

In the case, then, of writings on mosaics, of paintings and designs, it is not even possible to perceive the form and
the depth of the ductus. The surface appears completely flat to the touch.

The best solution to make the epigraphic media available to the visually impaired turns out to be, therefore, one that
involves more visitors who can better understand the artwork. In this regard, it being understood that the chance of
touching an original work wherever possible is a uniquely fascinating act, it becomes necessary to propose
reproductions that can make the material accessible by following the concept of “design for all,” which is adaptable to
all types of art. As rightly noted by the scholar and educator Enzo Tioli (Tioli, 2006 in L’arte a portata di mano),
this principle can be defined as “identical, whenever possible, equivalent when that is not possible.” All that can be
touched and seen in the original work constitutes the “identical”, the use of a copy constitutes the “equivalent”.

\bigskip

\bigskip

1.3. The Techniques of Accessibility

\bigskip

In the world of epigraphy, casts of inscriptions are often made use of, both to facilitate their study by specialists,
and also to replace those exhibits that had been located outside and brought indoors to protect them from weather
conditions. These casts, if used by the visually impaired, present the same problems of accessibility as the original
in terms of the difficulty of understanding the incision’s grooves. It is necessary, therefore, to have a reproduction
of a different type, upon which the letters can be transliterated in relief.

At this time of experimentation in this direction, there are not as yet many constructed, the reason being that works of
touchable statuary without additional aids are given more importance; and masterpieces of painting in art history are
made accessible by means of a process of conversion to relief.

A remarkable attempt, albeit with many limitations, is the one created in Rome’s Capitoline Museums that, for their
epigraphic collection, the Lapidary Gallery, created in 2007 a touch accessible project, in collaboration with the NGO
Museum - Voluntary Association of Museums. Here, works with the most significant inscriptions of the largest sizes and
preferably with reliefs were selected and availed in a way as to give the visually impaired a sense of the monument as
a whole, rather than only of the text itself. The technique utilized led to a Braille book entitled Messages from
Stone, published by Silvio Zamorani, who has specialized in Braille publications. The book was created with images in
relief; but unfortunately this technique is not very perceptible to the touch and quickly becomes outdated.

The just mentioned reliefs, in fact, hardly stands out from the outline of the exhibit and we opted for the choice of
reproducing the epigraphic texts only in print with overlapping respective Braille letters. For reasons of the problems
of adequate space previously mentioned, the transliteration into Braille is not always complete nor accurate, as in the
case of the relief dedicated to Silvanus and to the Genius of the Equites Singulares. In the latter instance, the
missing transliteration of certain words was somewhat justified by the lack of space; but such a choice deprives the
blind of the total understanding of the work. The dedicatee, Marcus Ulpius Fructus, is shown in Braille without
praenomen and nomen, thereby omitting an important element of dating and contextualization. In the caption, the text in
Braille and in writing does not coincide, thereby losing the abbreviation’s clarity and the translation’s exactness.

The technical device for this publication and the panels presented in the museum gallery is that of graphic Braille
print, a special printing technique of simple and rapid implementation, capable of reproducing text and drawings using
closely spaced Braille dots. This means that sometimes the difference between the signs of writing and those of the
drawings is imperceptible to the touch. Moreover, the device was also forced to work on only one level of relief.
Having multiple levels available to distinguish the different elements of the object, is similar to that of using
different tones of different colors. It is in this way possible to recognize the design of one level and the writing of
another - the differences of only a few millimeters would be sufficient.

A technique which solves this problem is that of the Thermoform, that is a relief obtained from a plastic material,
formed in a metal matrix. This type of technique is the one most used for the creation of teaching aids in schools for
the blind (in this regard see www.prociechi.it). But the technique proves to be very expensive in the creation of the
original matrix, as well as for the reproductions of individual pieces. The costs come down significantly when multiple
copies are derived. The ease of obtaining new copies compensates for the rapid deterioration of the material.

An additional system of reproduction, which is still in use but is gradually disappearing, is the Minolta Heater with
printing granules. This enables the quick printing of designs and writing on a paper little thicker than that of normal
stationary. It has never been used in reproduction of inscriptions for the blind and there would be nothing that could
prevent that. But based on experiences and tests made by people with visual disabilities, it can be said that the
resulting relief is not very pronounced to the touch, to this should be added the poor resistance of the characters,
which often, after prolonged use, tend to become damaged or completely lost. The Minolta system is therefore a still
utilizable technique useful in creating images for brief exposure, but it must be avoided for reproductions intended
for permanent collections.

\bigskip

\bigskip

1.4. 3D Printing

\bigskip

The new frontier in the understanding of works of art for the sight impaired is provided by 3D printing, which proves to
be both economical, fast, easy to implement, and durable. It is a single technique, which gathers together the
advantages of the techniques already highlighted and overcomes the defects of poor durability. Moreover, as the name
implies, it allows the introduction of the third dimension, which until now had never been possible, allowing the blind
person to be able to better perceive objects in space. This problem maybe not obvious in the case of inscriptions on
plates, but presents itself when we deal with sculptured monuments containing inscriptions.

Precisely from this technology in 2007, coordinated by Prof. Massimo Bergamasco and the Perceptual Robotics Laboratory
(Percro), was developed the project PURE-FORM, which led to the creation of a virtual museum, which gives visitors the
chance to see and “touch” objects, in particular digital sculptures, belonging to different historical periods and from
different contexts. All of this is based on a digital scan of the artworks stored in a database, thus offering a wide
choice, and reproduced on a touch screen display. In this project the focus was also placed exclusively on sculptures,
which already possess the third dimension, while the epigraphic content has been completely omitted. The creation of a
display, in practice a computer, allows on the one hand more accessibility to artworks, but on the other deprives the
user of the opportunity to travel to a place where they could personally experience the work’s features. It must not be
forgotten that a museum or an archaeological site are not just warehouses and storage sites.

The great advantage of 3D printing is that it materially allows more types of reproduction fundamental to human
comprehension that can be joined together and be touched and perceived as if real:

{}- Original

{}- Original with fractures and abrasions and signs of aging

{}- Integrated original

{}- Reproduction with transliterated reliefs

{}- Reproduction of the original with Braille overlay.

A copy could be thought of in its original 1:1 scale, that can be regularly handled to give an idea of its dimensions;
to another with letters in relief, rather than etched, for a better understanding of the texts; the realization of
individual fragments of inscriptions that could then be put together like a puzzle, just as the epigraphist does. Very
often, in fact, the inscription is the result of a recombination of the fragments. Moreover the inscription may be
created, by modeling it in a different way to the touch where the integration between the fragments is clear. In this
way the visually impaired would be able to approach epigraphy in almost the same way as any sighted person. This is
notwithstanding the need for explanatory captions accompanied by transliteration and the noting of the abbreviations,
which of course may not be clear to the layman.

This technique would be very valid when used in the case of reproduction of mosaics and inscriptions in the mosaic.
These reproductions make for the best situation where the user can differentiate between the various levels. The
background can be made on one level, while the drawn design can be made on another and also possible written upon. The
use of 3D would obtain a result similar to that by the Federazione Nazionale delle Istituzioni Pro Ciechi (National
Federation of Institutions for the Blind) NPO in the Square of the Corporations of Ostia Antica. In this area, paved
with many squares of mosaics, which bear witness to the intense commercial activity of ancient Ostia, the reproduction
of Ostia’s famous lighthouse was created with different levels of elevation of a mosaic in black and white, which allow
the user to not only to understand the image, but also to understand the technique of its realization. In this case,
though, the reproduction does not contain writing, because the chosen image contained no inscriptions. But nothing
prevents the use of the same technique using written entries, notwithstanding the need for an additional explanatory
panel. The disadvantage of this technique is the high cost of implementation, given that it is a reproduction created
exclusively from a specific work, and requiring a rather high level of craftsmanship since it is not merely modeled
from a metal matrix.

A thesis paper, of the Postgraduate School of Archaeological Heritage in Latin Epigraphy (Licordari, 2015), which
supports an exhibition project for the dissemination of Latin Epigraphy to the public, proposes 3D reproduction of two
mosaics of the stationes of that square in Ostia, with images and writing – that of the Libyan boatmen of Sabratha and
of the Carthaginian boatmen. The first, with the inscription STAT. SABRATENSIVM [CIL 14, 4549.14], is characterized by
the image of an elephant, the symbol of the Sabratha ivory trade. The second contains the inscription -NAVICVL.
KARTHAG. DE SVO [CIL 14, 4549.18] along with two ships. The paper’s original idea was based on the fact that not only
should there be brief and easy to understand inscriptions, but that the understanding could be made easier by the
presence of drawings presumably known to blind people. Certainly the case of Ostia Antica is a particularly simple one
as it involves mosaics of only two colors. A more complicated case would be a polychrome mosaic which attempts to
assimilate all the aspects of a colored painting. It is obvious that the sketched designs and writing can be configured
and presented on different levels.

3D printing has the great advantage of allowing the reproduction of colors. But to be made adequately perceptible by the
visually impaired, they should have a marked color contrast.

In the thesis mentioned above was proposed the use of 3D printing for the reproduction of other artifacts containing
inscriptions. It presented the case of a portable timepiece and an Italic mirror, both objects containing images with a
few written words. The inspiration to utilize the mirror came from the creation of a computer model by the former
Superintendence of the Archaeological Heritage of Southern Etruria. Achieving the goal required the difficult
extraction from the print of a mirror with correct reproductions of the images and the script. According to the officer
in charge, it would have been enough only to deepen the furrows of the incisions in order to construct an object that
could be accessed by the blind; but this obviously does not solve the problems already analyzed. The idea in principle
is good, but modifications need to be made to the relief of the writing and the images. Given the small size of the
object in this case, provisions would need to be made for reproductions of larger scale originals in order to enlarge
the space available for the user to engage in tactile perception. It is obvious that in cases involving a different
scale from the original, it is absolutely necessary to accurately render the differences so as not to create a false
perception.

A further advantage of 3D printing is that its process is not a stranger to digital epigraphy, since its starting point
is the obtaining of a virtual model of the inscription, which is then easily printed. While the processes of
thermo-formed and Braille prints require the contacting of the respective authorities in the specialized industries or
in publishing houses, it is much easier to buy and to utilize a 3D printer in a museum, in a superintendence or in a
university. It is important to remember that whatever technique is used, it is always necessary to consult an
appropriate expert to secure the necessary guidance in the preparation of materials. It is not enough to simply take a
photo or to draw a picture and print it in 3D. The technical support of someone experienced in this field, someone who
knows the psychological and physiological characteristics of the experiential perception in the context of its tangible
modeling is absolutely necessary.

Although in the Epigraphy is a marginal aspect, it is worth mentioning the problem of material, which of course does not
solve the issues involved in 3D printing. A reproduction in resin will never give the feeling of actually touching
stone, marble or bronze material, even if treating it’s surface in a way as to give an antique effect and feeling. The
technique of reproduction is therefore very useful to facilitate experiential understanding, but only as support to the
actual experience of the original. One should of course not think that a copy can completely replace the experience of
the original object.

\bigskip

\bigskip

1.5. Conclusions

\bigskip

As can be seen, the studies in this field as well as the examples of works on which to experiment with seeing impaired
users are as yet few in number. We hope that this essay will lead to inspire a dialogue among the relevant experts, and
that more work will be done in the field of epigraphy to make it increasingly accessible for persons with visual
disabilities.

The use of 3D technology for the reproduction of epigraphic material is in our opinion very promising. But it requires a
substantial amount of materials to be physically touched by the blind and visually impaired, so as in this way to
increase our knowledge of how to further improve the current techniques of modeling and reproduction.

\bigskip

Bibliography:

\bigskip

AA.VV., L’accessibilità al patrimonio museale per le persone con minorazione visiva, CD-ROM Le dispense del Museo Omero,
Ancona 2014

AA.VV., L’arte a portata di mano. Verso una pedagogia di accesso ai Beni Culturali senza barriere, Atti del convegno
Portonovo di Ancona 21-23 ottobre 2004, Roma 2006

Bellini, A., Toccare l’Arte. L’educazione estetica di ipovedenti e non vedenti, Roma 2000

Cetorelli Schivo, G., Colcerasa, M., Disabilità e museo. Percorrenze culturali di emergenza, in “Museo e territorio”,
Atti della III giornata di studi (Velletri 7-8 marzo 2003), Roma 2004

Hersh, M, Johnson, M.A. (Eds.), Assistive Technology for Visually Impaired and Blind People, Glasgow 2008

Jansson, G., Bergamasco, M., Frisoli, A., A new option for the visually impaired to experience 3D art at museums: manual
exploration of virtual copies, in “Visual Impairment Research”, 5,1, 2003, pp. 1–12

Licordari, F., Tituli in artis operibus inscripti: proposta di allestimento di una mostra nei locali della Fondazione
Sorgente, tesi di laurea della Scuola di Specializzazione in Beni Archeologici, Roma 2015

Levi, F., L’accessibilità alla cultura per i disabili visivi. Storia e orientamenti, Torino 2015

Velestino, D., Simili, G., Poscolieri, M. (Eds.), Messaggi dalla pietra, Torino 2007
\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-14 source/XX/EAGLE_2016_submission_XX.tex
all files movede from drive

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%

\title{}
\author{claudio prandoni}
\date{2015-11-09}
\begin{document}
[Warning: Draw object ignored]Abstract

\ \ \ \ \ \ This paper describes the functionalities and the technical infrastructure of the EAGLE Portal, the main
gateway into the EAGLE’s massive epigraphic database. The portal can be accessed both by a human through a web browser
and by external applications through a set of APIs. It is possible to perform full-text searches using a simple
interface, or to launch more advanced queries, including the possibility to upload an image and search for inscriptions
that are similar to the provided one. The seven controlled multilingual vocabularies that were created to help aligning
the multilingual metadata of the inscriptions from the different content providers, have been also integrated in the
search engine as well as all the translations the are available on the EAGLE MediaWiki. All this makes of the EAGLE
Portal a very useful tool for the epigraphic community but also for single users and enthusiasts willing to contribute
to the research in this field.

Keywords

Search engine, API, image recognition, controlled vocabulary, multilingual translations, advanced search, citizen
participation.

1.1. Introduction

The EAGLE Portal is the place where the content provided by the epigraphers’ community is aggregated and stored and
where it is made accessible to the users. It is composed by:

\begin{itemize}
\item The Metadata Aggregation System (in the following referred also as Aggregator), which is where all the content is
stored and indexed.
\item A Graphical User Interface (GUI), which is the interface through which users can browse the content and interact
with it.
\end{itemize}
The Aggregator relies on the D-NET software toolkit, developed within the DRIVER project and extended within the
OpenAIRE, OpenAIREplus\footnote{http://www.openaire.eu/}, EFG, EFG1914\footnote{http://www.europeanfilmgateway.eu/it}
and HOPE\footnote{http://www.peoplesheritage.eu/} projects. D-NET is an open source, general-purpose software conceived
to enable the realization and operation of Aggregative Data Infrastructures (ADIs) and to facilitate their evolution
and maintenance over time. D-NET implements a service-oriented framework based on standards, namely Web Services with
SOAP and REST APIs, where ADIs can be constructed in a LEGO-like approach, by selecting, customizing, and properly
combining D-NET services as building blocks. The resulting ADIs are systems, which can be customized, further extended
(e.g. by the integration of new services), and scaled (e.g. storage and index replicas can be maintained and deployed
on remote nodes to tackle multiple concurrent accesses or very-large data size) at run-time [Amato et al., 2013].

The GUI exposes all the content stored in the Aggregator providing to the users the following functionalities [Prandoni
et al., 2014].

\begin{itemize}
\item Search and browse the rich set of data made available by EAGLE content providers by using either a free text
search or a more advanced interface, including faceted browsing through the integration of the EAGLE controlled
vocabularies.
\item Similarity search through the image recognition algorithm that have been integrated in the EAGLE Portal.
\item Access to all the peer-reviewed translations of the epigraphic texts, in several European languages that have been
produced and integrated in the EAGLE MediaWiki.
\item Export to the user own PC the EpiDoc document of an object for further analysis and processing.
\item Annotate and save relevant information in a user Personal Space (e.g. records of inscriptions, search results,
queries), including content saved while using the Flagship Mobile Application.
\item Access to the Flagship Storytelling Application to browse the existing epigraphy-related narratives and to create
a new story.
\end{itemize}
The ingestion and curation of data is performed through a separate dedicated interface made available by the Aggregator.
This interface includes a series of functionalities to support data ingestion and storage as well as for importing,
indexing, enriching and managing the harvested metadata [Amato et al., 2015].

1.2. The EAGLE Collections

Classical Greek and Latin culture is at the very foundation of modern European identity. From philosophy to
architecture, geometry to law, a variety of contemporary subjects and disciplines have their roots in the classical
world. Only a small fraction of the total production of Greco-Roman texts has survived to the present day, leaving wide
gaps in the historiographical record of an epoch that is immensely relevant to our modern day lives.

The collections held by the EAGLE partners have been assembled with the two-fold criterion of historical-cultural
significance and strong thematic unity. They feature a great variety of inscriptions written in Greek, Latin and other
ancient languages, providing scholars with an authoritative resource by which to verify the reliability of historical
reconstructions. Additionally, they equip the broad public with a way to understand and easily appreciate interesting
and geographically dispersed inscriptions.

The EAGLE collections come from a wide range of digital repositories of epigraphic content. The following paragraphs
illustrate these repositories.

Arachne\footnote{ http://arachne.uni-koeln.de } is the central object-database of the German Archaeological Institute
(DAI). Since 2001, Arachne has been integrating negative archives of ancient sculpture that went beyond the specialised
documentation retained in Cologne itself, such as the negatives of ancient sculpture of the German Archaeological
Institute in Rome and the historic glass plate negative collections. Besides this larger project, many additional
activities are going on different levels, for example the online preparations for the »Corpus der Antiken
Sarkophagreliefs«.

Archaia Kypriaki Grammateia Digital Corpus (AKGDC) is a searchable digital library based on the 6 volumes of Archaia
Kypriaki Grammateia (Ancient Cypriot Literature), published by the A. G. Leventis Foundation between 1995-2008. The
AKGDC/ STARC collection of Cypriot inscriptions are dated from the 5th century BC to the 5th century AD and are mostly
funerary or dedicatory epigrams in elegiac couplet or hexameter.

Epigraphic Database Bari (EDB)\footnote{ http://www.edb.uniba.it } is a project specialized in epigraphic documents of
Christian patronage (third to eighth centuries, AD), including those contained in the Inscriptiones Christianae Vrbis
Romae, nova series, voll. I-X, in civitate Vaticana 1922-1992 (= ICVR), and those edited in other bibliographical seats
and/or not contained in the ICVR. Currently the inscriptions present in the EDB (counting those already online and
those awaiting definitive approval) amount to 40,000 items ca, though this number is obviously increasing continually.

The task of the Epigraphic Database Heidelberg (EDH)\footnote{ http://edh-www.adw.uni-heidelberg.de } is the systematic
entry of ancient Latin and bilingual (usually Latin and Greek) inscriptions into a complex database. The Epigraphic
Text Database is the heart of EDH and contains 65,000 inscriptions at present. Almost all of the records present texts,
which have already either been edited in the monumental Inscription corpora or published, revised and discussed in
thousands of scholarly articles.

Epigraphic Database Rome (EDR)\footnote{ http://edr-edr.it } was launched in 1999 as an experimental project aimed at
creating a unified database for ancient epigraphy. EDR is part of the international federation of Epigraphic Databases
called Electronic Archive of Greek and Latin Epigraphy. The federation’s purpose is to collect all published Greek and
Latin inscriptions up to the 7th century A.D. considering their best existing editions, also enclosing, when possible,
a number of additional important data and/or images. As part of the federation, EDR aims at collecting the whole
epigraphy of Rome and of the Italian peninsula including Sardinia and Sicily, with the exception of Christian
inscriptions (under EDB jurisdiction).

Hispania Epigraphica Online\footnote{ http://eda-bea.es} was created in 2002 when an EU grant enabled a joint research
project between the Archivo Epigráfico de Hispania and Mag. K. Schaller, who was developing computing applications for
archaeological purposes. The focus of the collection is the rich epigraphic patrimony of Portugal and Spain, mainly
written in Latin, but with some small pockets of Greek, Semitic and Iberian inscriptions.

Petrae database\footnote{ http://petrae.tge-adonis.fr} is a system for the recording of Latin and Greek inscriptions
developed at the Institut Ausonius, which collects epigraphic texts from the various regions in which its researchers
and collaborators are active. Each record produced has the text of an inscription in both an uppercase and lowercase
version, accompanied by metadata on all aspects of the monument, including support, fragments, epigraphic fields and
text elements (dating, paleography, critical apparatus, translation, notes).

The Last Statues of Antiquity\footnote{ http://laststatues.classics.ox.ac.uk } is a project funded in 2009 by the ‘Arts
and Humanities Research Council’ in Oxford. The project collects and analyses all the evidence for new, newly
dedicated, or newly re-worked statuary in the period circa 284–650. The two main results of the project are a major
database, with over 2600 individual entries, and a book, published in late 2012, discussing in print the entire
phenomenon of the late-antique statue habit.

The picture database VBI ERAT LUPA\footnote{ http://www.ubi-erat-lupa.org } contains stone monuments (sculptures,
reliefs, inscriptions, architectural pieces etc.). The project’s scope ranges from prehistoric stone monuments to
around the time of Justinian (500 AD). Due to its inception in Vienna, most project data at the moment is from the mid-
and south-eastern European region.

Finally, thanks to the great effort in the networking activities, many other partners joined the EAGLE Best Practice
Network and made available their collections that are currently being uploaded in the EAGLE Portal\footnote{ For a full
list of the EAGLE Associated Partners see http://www.eagle-network.eu/about/partners/ }.

1.3. The EAGLE inscription search engine

The EAGLE Inscriptions Search Engine is accessible through the main horizontal navigation bar of the EAGLE
Portal\footnote{ http://www.eagle-network.eu/basic-search/ }. It represents the core functionality of the portal,
through which the entry of keywords and phrases produce matches from EAGLE’s massive epigraphic database.

The EAGLE Portal makes available a “simple search”, an “advanced search” where the user can specify the values of a
number of fields in order to make a more accurate search, and an “image search” where the user can upload an image and
search for inscriptions that look similar to the provided one [Prandoni et al., 2014].

The objects that a user can search in the EAGLE Portal belong to three different categories, in accordance with the
EAGLE conceptual model [Sicila et al., 2015]:

\begin{itemize}
\item “Artefacts”, which contain all the information that is somehow related to the physical carrier of the inscription.
\item “Texts”, which contain all the information that is textual in nature.
\item “Images”, which contain all the information that is visual in nature.
\end{itemize}

\bigskip

1.3.1. How to search inscriptions

\bigskip

The simple search user interface is very straightforward. The text entered in the query box is used to make a full text
search in all the fields of all the EAGLE objects in the category determined when making the query. By default, the
search will be done on all the objects in the “Artefact” category. Two more tabs on the result list (labelled “Texts”
and “Images”) allow the user to perform the same query choosing a different category.

Using the advanced search interface, a user can specify values for a number of fields, in order to have more accurate
results: findspot, bibliography, text of the inscription, type of inscription, decoration, object type, material, type
of writing, state of preservation (Fig. 1).

In the fields having a controlled vocabulary, the user is allowed to enter only values coming from the vocabulary. For
this purpose, those fields have a “drop-down menu” which displays all the defined values for that field in the
“preferred label”, i.e. the text string that has been indicated in the vocabulary as the preferred one for display,
regardless of the language in which the string is defined.

Finally, by choosing the Image Search option, a user can exploit the image recognition algorithm that has been
integrated in the EAGLE Portal and that allows to search for images that are similar to the one a user has uploaded,
ranked in decreasing order of similarity (Fig. 2).

\bigskip

 \includegraphics[width=10.874cm,height=12.885cm]{EAGLE2016submissionXX-img001.png}

Fig. 1. Search for inscriptions in the EAGLE Portal

\bigskip

Content Based Image Retrieval (CBIR) is becoming an effective way for searching digital libraries, as the amount of
available image data is constantly increasing. CBIR applications are increasingly becoming useful in accessing cultural
heritage information, as a complement to metadata based search. In fact, in some cases metadata associated with images
do not describe the content with sufficient details to satisfy the user queries, or metadata are completely missing.
For example, images containing reproductions of works of art contain a lot of implicit information that is not
generally captured in manually or automatically generated metadata [Amato et al., 2013].

The EAGLE Image Retrieval System (IRS) receives in input the image of an epigraph and provides effective and efficient
image similarity search and image content recognition of epigraphs.

\bigskip

 \includegraphics[width=10.848cm,height=10.451cm]{EAGLE2016submissionXX-img002.png}

Fig. 2. Image search in EAGLE

\bigskip

The images provided by the EAGLE content providers and collected in the Aggregator have been processed by a set of IRS
components (Image Feature Extractor, Image Indexer, CBIR Index) to build the index that allows a fast and efficient
similarity search during the query phase.

In addition, for those epigraphs where a set of images is available (training set), each training set has been processed
to extract the main features characterizing the epigraph. The training sets and the characterizing features are the
base for building another index, to be used by the image recognizer in order to decide if a received image (during the
query phase) can be classified as belonging to one of the existing sets or not. In this way, in many cases, the
recognizer is able to properly recognize the content of a query image even if the image given in the query was never
stored in the database.

\bigskip

 \includegraphics[width=10.874cm,height=12.197cm]{EAGLE2016submissionXX-img003.png}

Fig. 3. Results list - Artefacts

\bigskip

1.3.2. Filtering and browsing results

\bigskip

After having submitted a query, the user is presented with a paginated list of results containing some basic information
(Fig. 3).

By using the facets displayed on the left side panel, the user has the possibility to refine the query by applying some
filters based on the fields that are associated with a controlled vocabulary or based on the fact whether the
inscription has an image and/or a translation.

Regardless of the type of query performed by the user and of the category in which the query was done, clicking on one
of the items in the result page will display a summary of all the information available for that item, with links to
get further details (Fig. 4).

\bigskip

 \includegraphics[width=10.874cm,height=8.864cm]{EAGLE2016submissionXX-img004.png}

Fig. 4. Object details

\bigskip

The text in big characters on top of the page is the Title of the object. Below the title there is the local ID of the
object and after that there is a line collecting all the clickable items of the summary page: Bibliography, which opens
a text box with the bibliography associated with the object; Translations, which displays all the translations
available for the inscription of the object; Original source, which links the page of the object in the database of the
Content Provider who provided it; Save, which allows to save the object in the user’s personal space; Export, which
allows to download locally the EpiDoc document describing the object. Finally, all available pictures are displayed,
together with the remaining descriptive metadata.

Duplicates identification has been performed through the Trismegistos platform\footnote{ http://www.trismegistos.org/ }.
By standardizing publication references, collection information, material, provenance and date, the overlap of the
various databases has been mapped to identify records describing the same inscription. Afterwards a unique numeric
identifier, the Trismegistos ID, has been assigned to each document and spread across the partner databases for
inclusion in the metadata record to be uploaded in the EAGLE Portal.

If more than one instance of an object is available, the other instances appear on the summary page as clickable “tabs”,
each tab being labelled with the local ID of the other instances of the same object.

The seven controlled multilingual vocabularies that have been created (Type of Inscription, Object Type, Material,
Execution, Decoration, State of Preservation, Dating Criteria) to help aligning the multilingual metadata of the
inscriptions from the different content providers, have been also integrated in the EAGLE search engine. Every time
that a term which belongs to one of the vocabularies is displayed, it is automatically linked to the specific page
describing the term, where further connections, synonyms and translations can be explored. The EAGLE vocabularies are
accessible also as Linked Open Data\footnote{ http://www.eagle-network.eu/resources/vocabularies/}.

Finally, all the peer-reviewed translations in several European languages that are available, have been integrated in
the EAGLE Portal too through a RESTful API native of Wikibase, the MediaWiki extension used for this purpose\footnote{
http://www.eagle-network.eu/wiki/api.php }. For each inscription it is possible to view the existing translation, to
request a new one and to contribute by connecting to the EAGLE MediaWiki\footnote{ http://www.eagle-network.eu/wiki }.

\bigskip

\ 1.3.3. Annotating and saving objects

\bigskip

A registered user, during a “web” or a “mobile” session, has the capability of annotating and saving (some of) the
information that is being provided by the EAGLE system.

Depending on the information the user is looking at, hitting the save button will save two types of data in the personal
space:

\begin{itemize}
\item A query and its results.
\item Detailed information about an inscription.
\end{itemize}
The data saved by the local user on the EAGLE server is stored internally in a relational database.

A registered user logged in at the EAGLE Portal can access his/her saved data and perform some simple operations on it:

\begin{itemize}
\item Display of the saved queries or objects.
\item Modify the textual annotations associated with the saved item.
\item Delete the saved item from her personal space.
\item View the saved item.
\item Import the data that he/she saved during a mobile session.
\end{itemize}
It has to be noted that, when a user views one of the saved items, the object is displayed exactly as he/she saved it.
The saved data might be different from the data that can be retrieved by issuing the same query at the time of editing,
due to changes in the data stored in the EAGLE database. This allows the user to make comparisons between the
information that was available when he/she first saved the object and what has been added afterwards.

1.4. Conclusion

This paper presented the main functionalities of the EAGLE Portal, giving some highlights of the content that is
available and of the technical infrastructure that allows to search and browse them.

EAGLE aims to build a multi-lingual online collection of millions of digitised items from European museums, libraries,
archives and multi-media collections, which deal with inscriptions from the Greek and Roman World. The aim of the
network is to make available the vast majority of the surviving inscriptions of the Greco-Roman world, completed with
the essential information about them, enriched through the use of seven multilingual vocabularies, and complemented
with a series of peer-reviewed translations in several European languages, which are notoriously unavailable for
inscriptions.

All this makes of the EAGLE Portal a very useful tool for the epigraphic community but also for single users and
enthusiasts willing to contribute to the research in this field.

The participation of Europe’s citizens in scientific research represents an important opportunity for improving European
competitiveness, because of the value that citizens can add in specific areas of research. In particular, the
participation of citizens in the research on CH and humanities has the potential to play an important role in the
development of the European Research Area, and can take the lead in the discovery of new directions of
cross-disciplinary research.

In this framework, the CIVIC EPISTEMOLOGIES\footnote{ http://www.civic-epistemologies.eu } project developed a Roadmap
for the use of e-Infrastructures to support the participation of European citizens in CH practices and humanities
research, where such engagement has a twofold benefit for culture: to be enriched by the citizens’ contributions and to
become more widely used and exploited (also, for example, with the participation of creative industries) [Fresa \&
Justrell, 2015]. The future development of the EAGLE initiative will refer to the CIVIC EPISTEMOLOGIES Roadmap for
investigating how to put in place more advanced services supporting citizen science in epigraphic research.

The EAGLE Portal is only an example of application that makes use of the APIs provided by the Aggregator to access the
EAGLE collections. Other examples are the Flagship Mobile Application\footnote{
http://www.eagle-network.eu/resources/flagship-mobile-app/ }, that allows to access the inscriptions’ database through
a mobile device and to fully exploit the image recognition features integrated in the EAGLE Portal, and the Flagship
Storytelling Application\footnote{ http://www.eagle-network.eu/stories/ }, that allows a user to create stories
starting from the content available in EAGLE, Europeana and many other data sources.

Other applications may be created based on the APIs provided by the Aggregator, which uses a SOLR-base search engine
through which it is possible to search and browse the full set of content\footnote{ For further information see
http://wiki.apache.org/solr/\#Search\_and\_Indexing }. Furthermore, APIs are provided to register a new user, login and
perform an image-based search using the EAGLE image recognition algorithm.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2015-12-16 source/10/Paper proposal for EAGLE final.tex
started work on tex of papers, added folders for CNR papers, changed main stracture to include keynotes and AHE abstracts

@@
 %%%%%%%%
-\title{}
-\author{Abigail S Graham}
+\title{FACE VALUE: THE POWER OF IMAGES AT APHRODISIAS. How Digital Resources Can Transform our use of Palaeography in understanding Inscriptions.}
+\author[war]{Abigail S. Graham\corref{first}}
+\address[war]{The University of Warwick}
+\cortext[first]{Corresponding author. Email: Abigail.graham@warwick.ac.uk}
 \date{2015-11-16}
 \begin{document}
-\begin{enumerate}
-\item FACE VALUE: THE POWER OF IMAGES AT APHRODISIAS
-\end{enumerate}
-How Digital Resources Can Tranform our use of Palaeography in understanding Inscriptions.
-
-\bigskip
-
-\ Dr. Abigail Graham
-
-\bigskip
-
-The University of Warwick (Lecturer)
-
-\bigskip
-
-Abigail.graham@warwick.ac.uk
-
-\bigskip
-
-Possibly for subject panel: Dating criteria for epigraphic material: possible approaches.
-
-\bigskip
-
-Abstract:
-
-\bigskip
-
-\ This paper will consider how the publication of a large digital corpus (The Inscriptions of Aphrodisias [Reynolds,
-Roueche \& Bodard, 2007] has shaped the assessment of inscriptions, particularly regarding palaeography and the dating
+ \maketitle
+\begin{abstract}
+This paper will consider how the publication of a large digital corpus The Inscriptions of Aphrodisias \citet{ReynoldsRouecheBodard2007} has shaped the assessment of inscriptions, particularly regarding palaeography and the dating
 of inscriptions. A case study of dedicatory inscriptions from the Temple of Aphrodite at Aphrodisias will explore how
@@
 
-\bigskip
+\end{abstract}
+\keywords{Dating, Palaeography, Ordinatio, Letter forms, Aphrodisias, Context, Recarving.}
 
-Keywords
+\section{Introduction}
+\subsection{Caveat Lector: Defining Palaeography and traditional approaches}
 
-\bigskip
-
-Dating, Palaeography, Ordinatio, Letter forms, Aphrodisias, Context, Recarving.
-
-\bigskip
-
-\bigskip
-
-1.1 Introduction
-
-\bigskip
-
-\begin{enumerate}
-\item \begin{enumerate}
-\item \begin{enumerate}
-\item Caveat Lector: \ Defining Palaeography and traditional approaches
-\end{enumerate}
-\end{enumerate}
-\end{enumerate}
-
-\bigskip
-
-As the co-ordinator and lecturer on three graduate level courses of Roman Epigraphy, I am invariably asked the same
-question: how do you use letterforms to date inscriptions? My answer is always the same: “very carefully”. \ Studying
-palaeography within the discipline of ancient epigraphy can be a journey into thorny hedge where one can easily “fall
-into that category of human endeavor known as stylistic attribution and inevitably involve subjectivity”.\footnote{
-Tracy, 1995, 3-4. } Stephen Tracy advises that palaeographic surveys should be carried out with caution: “Caveat Lector
-must needs be our motto”. \ The study of palaeography is problematic on a number of levels, both in the way it is
+\noindent  As the co-ordinator and lecturer on three graduate level courses of Roman Epigraphy, I am invariably asked the same
+question: how do you use letterforms to date inscriptions? My answer is always the same: ``very carefully''. Studying
+palaeography within the discipline of ancient epigraphy can be a journey into thorny hedge where one can easily ``fall
+into that category of human endeavor known as stylistic attribution and inevitably involve subjectivity''.\footnote{\citet[3-4]{tracy1995}} Stephen Tracy advises that palaeographic surveys should be carried out with caution: ``Caveat Lector
+must needs be our motto''. The study of palaeography is problematic on a number of levels, both in the way it is
 defined and the ways in which it is employed. Many scholars dismiss lettering as a means of dating and they make an
@@
 problematic when applied generally on a broader scale (e.g. to larger geographical areas or time trame) where
-archaisms, local styles, and variations can create distortions.\footnote{ Susini 1973; Petrucci 1993; Di Stefano
-Manzella 1995 163-181, Cooley 2012, 433; Edmundson 2015 122-125.} Similar caution would be applied in dating a
+archaisms, local styles, and variations can create distortions.\footnote{\citet{susini1973}; \citet{petrucci1993}; \citet[163-181]{Manzella1995}; \citet[433]{cooley2012}; \citet[122-125]{bruun2015}.} Similar caution would be applied in dating a
 sculpture on the basis of a hair fragment. Analysis of statues considers a number of factors, material, hairstyle,
@@
 While there is potentially more information, there is also a greater chance that it will be contradictory. Thus the use
-of lettering, an “imprecise science”, is better used in combination with a number of different factors.\footnote{
-Harris 1989, 26-9; Bodel 2001 3-5; Di Stefano Manzella 2007; 393-418 and Cooley 2012, 432-435. } Despite the
+of lettering, an ``imprecise science'', is better used in combination with a number of different factors.\footnote{\citet[26-9]{harris1989}; \citet[3-5]{bodel2001}; \citet[393-418]{Manzella2007} and \citet[432-435]{cooley2012}.} Despite the
 aforementioned limitations, a number of informative palaeographic studies have been produced. The success of these is
 based on detailed commentaries on a specific corpus of material, a transparent methodology and the incorporation of
-numerous high quality images.\footnote{ For example: Gordon 1957 offers more than 50 plates and figures, Di Stefano
-Manzella 1987 provides 218 illustrations, Tracy 1990, 1995 and 2003 uses over 60 plates and figures in each volume. \ }
+numerous high quality images.\footnote{ For example: \citet{Gordon1957} offers more than 50 plates and figures, \citet{Manzella1987} provides 218 illustrations, \citet{tracy1990,tracy1995,tracy2003} uses over 60 plates and figures in each volume.}
 
-\bigskip
-
-A second issue in the study of palaeography involves the access and publication of epigraphic materials. \ Access to
+A second issue in the study of palaeography involves the access and publication of epigraphic materials. Access to
 large corpora of inscriptions has traditionally been limited to a small audience of scholars and site visitors. Apart
@@
 viewing only the lettering, was often removed from the visual elements of the inscription (e.g. the type of stone, use
-of spacing and decoration, letter size). \ This is a suboptimal way of assessing epigraphic evidence.
-
-\bigskip
+of spacing and decoration, letter size). This is a suboptimal way of assessing epigraphic evidence.
 
@@
 monumentality, including new methodologies, approaches as well as attempts to redefine genre classifications and
-terminology.\footnote{ Woolf 1996, 22-39, Eck 2010, Panciera 2012, 1-10, Graham 2013, 1-17. \ For a recent collection
-and analysis of epigraphic databases, see Tom Elliot’s work \ (chapter 5) in Bruun and Edmundson 2015, 78-85. \ } \ A
+terminology.\footnote{\citet[22-39]{woolf1996}, \citet{eck2010}, \citet[1-10]{panciera2012}, \citet[1-17]{graham2013}.
+For a recent collection and analysis of epigraphic databases, see \citet[78-85]{Elliot2015}.} A
 number of recent studies have used changes in the appearance of inscriptions over time, such as the use of different
-media, decorative and paratextual elements\footnote{ \ One marks in the text see Susini 1973, 26 and Cooley in Jansen
-2014, 143-155.} (e.g. ivy leaves as interpuncts (hederae distinguentes),\footnote{ \ Hommel 1970, 293-303. }
-abbreviations\footnote{ Gordon, “Supralineate Abbreviations” 1957, 111.} and spacing between words\footnote{ On using
-space to reconstruct inscriptions see Alföldy 1995, 195-226, Grasby 2002, 151-176.} alongside lettering, as dating
-criteria.\footnote{ For an extensive list of dating conventions cf..Di Stefano Manzella 1987, Chapter 20. }
-\ Panciera’s recent article, in particular, advocates the significance of the public context and visibility of
-inscriptions.\footnote{ “I would propose to regard as an ‘inscription’ any particular type of written human
+media, decorative and paratextual elements\footnote{On marks in the text see \citet[26]{susini1973} and \citet[143-155]{cooley2014}.} (e.g.
+ivy leaves as interpuncts (hederae distinguentes),\footnote{\citep[293-303]{hommel1970}}
+abbreviations\footnote{\citep[111]{Gordon1957}} and spacing between words\footnote{On using
+space to reconstruct inscriptions see \citet[195-226]{alfoldy1995}, \citet[151-176]{grasby2002}.} alongside lettering, as dating
+criteria.\footnote{ For an extensive list of dating conventions cf. \citet[Chapter 20]{Manzella1987}. }
+Panciera’s recent article, in particular, advocates the significance of the public context and visibility of
+inscriptions.\footnote{ ``I would propose to regard as an ‘inscription’ any particular type of written human
 communication of the sort that we would today call unidirectional…not being addressed to a person or to a group but to
 a collectivity, and for this reason is made with the location, writing technique, graphic form and impagination, mode
-and register of expression chosen because they are most suitable to the attainment of its intended goal.” (Panciera
-2012: 8).} \ In this vein, it is worth considering how the physical characteristics of an inscription belong within a
+and register of expression chosen because they are most suitable to the attainment of its intended goal.'' \citep[8]{panciera
+2012}.} In this vein, it is worth considering how the physical characteristics of an inscription belong within a
 broader assessment of a culture of writing. Is palaeography the study of letterforms alone? Was a focus on letterforms
@@
 
-\bigskip
+\subsection{Outline}
 
-1.1.3 Outline
-
-\bigskip
-
-The focus of this paper is a series of rather unimpressive column dedications from the temple of Aphrodite at
-Aphrodisias. \ While the information recorded in the texts is unremarkable, the journey of these inscriptions from
+\noindent The focus of this paper is a series of rather unimpressive column dedications from the temple of Aphrodite at
+Aphrodisias. While the information recorded in the texts is unremarkable, the journey of these inscriptions from
 their publication as part of an epigraphic corpus to a recent digital publication reveals an epic transformation in the
-format and approach to these materials. \ This study will begin with brief overview of previous published editions
-(e.g. MAMA volume VIII, 1962) followed by an assessment of the information available on the current Inscriptions of
+format and approach to these materials. This study will begin with brief overview of previous published editions
+(e.g. \citet{mama8}) followed by an assessment of the information available on the current Inscriptions of
 Aphrodisias (2007) website. Through careful assessment of text (including formula, vocabulary and spelling) and its
@@
 the amount of information available whilst also adding clarity to the process of dating an inscription. By examining
-how we define and use palaeography to evaluate inscriptions on the Inscriptions of Aphrodisias website (IAph 2007), we
+how we define and use palaeography to evaluate inscriptions on the Inscriptions of Aphrodisias website \citep{ReynoldsRouecheBodard2007}, we
 can observe how the discipline has evolved and what changes may be possible in approaches and the publication of
@@
 
-\bigskip
+%%%%%%%%%
+%IN THE DOC THERE IS A "BOX" WITH this text
+%Key Questions:
+%
+%What elements of the text are available to us in different media?
+%How do we record the dating process and criteria? 
+%How do we represent (and attempt to reconcile) contradictory information?
+%%%%%%%%%%%
 
-\bigskip
+\section{Publishing inscriptions: A Case Study of Column Dedications from The Temple Of Aphrodite at Aphrodisias}
 
-\bigskip
+\subsection{A brief introduction to the inscriptions IAph 1.4-1.6.}
 
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-[Warning: Draw object ignored]
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\begin{enumerate}
-\item \begin{enumerate}
-\item Publishing inscriptions: A Case Study of Column Dedications from The Temple Of Aphrodite at Aphrodisias 
-\end{enumerate}
-\end{enumerate}
-
-\bigskip
-
-1.2.1 A brief introduction to the inscriptions IAph 1.4-1.6.
-
-\bigskip
-
-Three inscriptions, each of which record the dedication of a column at the Temple of Aphrodite, will be the subject of
-this survey. \ The first two texts were noted as early as the 18th century and copied in a notebook by the British
-Architect Deering in the early 19th century. A third version of the text, was \ uncovered during excavations at the
+\noindent Three inscriptions, each of which record the dedication of a column at the Temple of Aphrodite, will be the subject of
+this survey. The first two texts were noted as early as the 18th century and copied in a notebook by the British
+Architect Deering in the early 19th century. A third version of the text, was uncovered during excavations at the
 temple site by the French Engineer Gaudin in the early 20th century.\footnote{ A comprehensive history of the history
 and bibliography, as well as a description of the resources can be found on the IAph website:
-http://insaph.kcl.ac.uk/iaph2007/iAph010006.html. } The inscriptions are recorded on tabella ansata (0.745 x 0.465m) as
+\url{http://insaph.kcl.ac.uk/iaph2007/iAph010006.html}. } The inscriptions are recorded on tabella ansata (0.745 x 0.465m) as
 part of fluted marble columns of the peristasis, some of these have been reconstructed in modern restorations on the
@@
 
-\bigskip
-
-These texts can be dated through a number of different criteria. \ Contextual association with construction of the
+These texts can be dated through a number of different criteria. Contextual association with construction of the
 temple, dates between the end of the 1st century BC and the early 1st century AD, though the land was clearly in use
-well before this time. \footnote{ Smith 1996, 41-42 and Reynolds 1990, 37. References to Sulla’s dedications are in
-Appian BC 1.97. \ Caesar and Augustus’ acknowledgements of the sacred space are evident in IAph 8.27, 8.31 and 1.1. }
-\ This is corroborated by coins depicting the temple, which date from 2 BC- 14 AD.\footnote{ Coin type 41 depicts
+well before this time. \footnote{\citet[41-42]{smith1996} and \citet[37]{reynolds1990}. References to Sulla’s dedications are in
+Appian BC 1.97. Caesar and Augustus’ acknowledgements of the sacred space are evident in IAph 8.27, 8.31 and 1.1. }
+This is corroborated by coins depicting the temple, which date from 2 BC- 14 AD.\footnote{ Coin type 41 depicts
 Augustus (OBV) and the temple of Aphrodite (REV) in MacDonald 1992, Plate V R131. } Prosopography is also informative:
@@
 euergetes, are absent from Zoilos’ other inscriptions and may imply a posthumous dedication, perhaps by the boule and
-demos (Reynolds 1990, 38).} The benefactor of the columns in this case study, Eumachos Diogenes, also comes from an
+demos \citep[38]{reynolds 1990}.} The benefactor of the columns in this case study, Eumachos Diogenes, also comes from an
 established family that flourished well into the 2nd century AD.\footnote{18 The success of Eumachus Diogenes’ family,
 which included the first known Aphrodisian to hold a procuratorship in the 2nd century AD, reveals an enduring
-significance for his family monuments in the city (Reynolds 1999b, 327-334).} \ Finally, the formula of the text the
+significance for his family monuments in the city \citep[327-334]{reynolds1999}.} Finally, the formula of the text the
 inscription and the vocabulary, particularly the term philokaisar, suggest a late Republican or Augustan
@@
 period in Aphrodisias (and Asia Minor) until the early Imperial Period, after which a new formula (beginning with
-recipients (e.g. Aphrodite and Imperial recipients) is predominant (cf. Graham 2013, 4-7). \ The Augustan and early
+recipients (e.g. Aphrodite and Imperial recipients) is predominant (cf. \citet[4-7]{graham2013}). The Augustan and early
 Tiberian uses of the philokaisar are known from a dedication at Ioulis dated ca. 27 BC - AD 14 (SEG XLVIII (1998) no.
 1129) and in a monument to Ti. Cl. Drusus in Patara (SEG XLIV (1994) no. 1205). For significance and date of the title
-see Buraselis 2001, 101-105. } \ Before encountering the inscriptions face to face, the dating, function and meaning of
-these inscriptions appears to be quite straightforward. \ So let us examine the experience of viewing these
+see \citet[101-105]{buraselis2001}.} Before encountering the inscriptions face to face, the dating, function and meaning of
+these inscriptions appears to be quite straightforward. So let us examine the experience of viewing these
 inscriptions various published formats.
 
-\bigskip
-
-1.2.2. Publishing IAPH 1.4 and 1.5 in MAMA VIII (nos. 347 \& 348): A series of copies?
-
-\bigskip
-
-Cormack published the first comprehensive catalogue of inscriptions from Aphrodisias as part of his Monumenta Asiae
-Minoris Antiqua volume VIII in 1962. \ This was a great undertaking that included texts as well as images. Two of the
-three texts \ (IAph 1.4 \& 1.5) were published here as MAMA 437 and MAMA 438 respectively, while the third text was
+\subsection{Publishing IAPH 1.4 and 1.5 in MAMA VIII (nos. 347 \& 348): A series of copies?}
+\noindent Cormack published the first comprehensive catalogue of inscriptions from Aphrodisias as part of his Monumenta Asiae
+Minoris Antiqua volume VIII in 1962. This was a great undertaking that included texts as well as images. Two of the
+three texts (IAph 1.4 \& 1.5) were published here as MAMA 437 and MAMA 438 respectively, while the third text was
 merely mentioned as a further copy.
 
-\bigskip
+\begin{quotation}
+\noindent MAMA 437\\
+\begin{otherlanguage*}{greek}
+\noindent Εὔμαχος Ἀθηναγό-\\
+ρου τοῦ Ἀθηναγόρου\\
+τοῦ Εὐμάχου Διογένη-\\
+ς φιλόκαισαρ καὶ Ἀμιὰς\\
+Διονυσίου φύσι δὲ Ἀδρά<σ>του\\
+τοῦ Μόλωνος Ὀλυνπιὰς\\
+τὸν κίονα θεᾷ Ἀφροδίτῃ\\
+καὶ τῷ δήμῳ.
+\end{otherlanguage*}
+\end{quotation}
 
-\bigskip
+\begin{quotation}
+\noindent MAMA 438 \\
+\begin{otherlanguage*}{greek}
+\noindent Εὔμαχος Ἀθηνα-\\
+γόρου τοῦ Ἀθηναγό-\\
+ρου τοῦ Εὐμάχου Δι-\\
+ογένης φιλόκαισαρ\\
+καὶ Ἀμμιὰς Διονυσί-\\
+ου φύσι δὲ Ἀδράστου\\
+τοῦ Μόλωνος Ὀλυν-\\
+πιὰς τὸν κίονα θεᾷ\\
+Ἀφροδίτῃ καὶ τῷ\\
+δήμῳ.
+\end{otherlanguage*}
+\end{quotation}
 
-\bigskip
-
-\bigskip
-
-[Warning: Draw object ignored][Warning: Draw object ignored]
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-[Warning: Draw object ignored]
-
-\bigskip
+\begin{quotation}
+\emph{Translation by author:}\\
+Eumachos Diogenes, son of Athenagoras, the son of Athenagoras, the son of Eumachos, devoted to Caesar, and Ammias Olympias, daughter of Dionysius, the natural daughter of Adrastus, the son of Molon, (dedicated) a column to the goddess Aphrodite and the People.
+\end{quotation}
 
@@
 their digital format as published the Packhum website \ (Aphrodisias 108 and 109 (respectively)).
-http://epigraphy.packhum.org/inscriptions/main?url=oi\%3Fikey\%3D256987} \ IAph 1.4 (MAMA 437) has a slightly less
+\url{http://epigraphy.packhum.org/inscriptions/main?url=oi\%3Fikey\%3D256987}} IAph 1.4 (MAMA 437) has a slightly less
 impressive ordinatio, particularly in line 4 with an odd line break, a misspelling of Ammias (as Amias) and Adrastus
@@
 The published format of the texts also makes them appear similar in lettering and arrangement (e.g. left indentation).
-\
-
-\bigskip
 
@@
 
-\bigskip
-
-\ The third text version of this text is mentioned in the commentary and based on the information provided (two similar
-texts and a reference to a “copy”) one would expect that it was very similar. The same editorial practice is employed
+The third text version of this text is mentioned in the commentary and based on the information provided (two similar
+texts and a reference to a ``copy'') one would expect that it was very similar. The same editorial practice is employed
 in the parallel column dedications from a different benefactor (I.Aph 7 \& I.Aph 8), where 2nd version of the text is
@@
 late 1st c. BC/early 1st c. AD lettering. When publishing a large corpus of inscriptions, referring to copies is
-understandable. However, treating texts as “copies” as opposed to individual monuments can, and in this case will,
-prove problematic.\footnote{ For another study of “copy” inscriptions at Aphrodisias see Graham 2015 (forthcoming).}
+understandable. However, treating texts as ``copies'' as opposed to individual monuments can, and in this case will,
+prove problematic.\footnote{ For another study of ``copy'' inscriptions at Aphrodisias see Graham 2015 (forthcoming).}
 While MAMA VIII offered a broader audience access to the inscriptions at Aphrodisias, the approach to the inscriptions
 and the quality of the images also imposed limitations that made it difficult to analyse the physical elements of the
-inscription or to question the proposed texts and restorations. \
+inscription or to question the proposed texts and restorations.
 
-1.2.3 \ Living in a Digital World: The Publication of Column Dedications in IAph 2007 (IAph 1.4.)
-
-\bigskip
-
+\subsection{Living in a Digital World: The Publication of Column Dedications in IAph 2007 (IAph 1.4.)}
 The Inscriptions of Aphrodisias project was a groundbreaking endeavour both for publishing epigraphic materials and
@@
 wealth of superlative scholarship).\footnote{ It is worth noting that websites, however they develop, will remain
-supplementary to published scholarship.} \ In addition to search functions, the page is more interactive, the texts
+supplementary to published scholarship.} In addition to search functions, the page is more interactive, the texts
 have been revisited, translations added, as well as commentary with weblinks to parallel texts. The organization of the
 inscriptions by context allows the reader to view the inscriptions in context and to gain a better understanding of the
-area’s “epigraphic habit” (Figure 1: Screenshot of \ IAph 4 webpage).
+area’s ``epigraphic habit'' (Fig.~\ref{fig:1}).
 
-\bigskip
-
- \includegraphics[width=14.605cm,height=8.726cm]{PaperproposalforEAGLEfinal-img001.png}
-
-\bigskip
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{PaperproposalforEAGLEfinal-img001.png}
+\caption{Screenshot of IAph 4 webpage}
+\label{fig:1}
+\end{figure}
 
@@
 
-\bigskip
-
 The published texts of IAph 1.4 and MAMA 437 present one significant variation: the inclusion of space indicators within
@@
 dedications at the Sebasteion IAph 9.1, 9.25, 9.112). This vacat serves little grammatical function, and is likely to
-be a result of the arrangement of the text/ omission in Ammias’ name. } \ The indentation on the last line also reveals
+be a result of the arrangement of the text/ omission in Ammias’ name. } The indentation on the last line also reveals
 a left orientation of the line which is more common in Late Republican/Augustan texts at Aphrodisias (e.g. IAph 1.1,
-1.7, 1.8 and 1.38), as opposed to the “justified” approach (indentations on both sides of the text to accentuate a name
+1.7, 1.8 and 1.38), as opposed to the ``justified'' approach (indentations on both sides of the text to accentuate a name
 or word), which is more common in later Imperial inscriptions at Aphrodisias.
 
-\bigskip
+\begin{quotation}
+\noindent MAMA 437\\
+\begin{otherlanguage*}{greek}
+\noindent Εὔμαχος Ἀθηναγό-\\
+ρου τοῦ Ἀθηναγόρου\\
+τοῦ Εὐμάχου Διογένη-\\
+ς φιλόκαισαρ καὶ Ἀμιὰς\\
+Διονυσίου φύσι δὲ Ἀδρά<σ>του\\
+τοῦ Μόλωνος Ὀλυνπιὰς\\
+τὸν κίονα θεᾷ Ἀφροδίτῃ\\
+καὶ τῷ δήμῳ.
+\end{otherlanguage*}
 
-\bigskip
-
-\bigskip
-
-[Warning: Draw object ignored]
-
-[Warning: Draw object ignored]
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
+\end{quotation}
+\begin{quotation}
+\noindent IAph 1.4.\\
+\begin{otherlanguage*}{greek}
+\noindent Εὔμαχος Ἀθηναγό-\\
+ρου τοῦ Ἀθηναγόρου\\
+τοῦ Εὐμάχου Διογένη-\\
+ς Φιλόκαισαρ καὶ Ἀμιὰς\end{otherlanguage*} v.\\
+\begin{otherlanguage*}{greek}Διονυσίου φύσι δὲ Ἀδρά<σ>του\\
+τοῦ Μόλωνος Ὀλυνπιὰς\\
+τὸν κίονα θεᾷ Ἀφροδίτῃ\\
+καὶ τῷ Δήμῳ
+\end{otherlanguage*} vacat
+\end{quotation}
 
 High quality images on the website facilitate connections between the published text and the inscription. In the image
-of IAph 1.4 (Figure 2. Photograph of IAph 1.4), one can to observe how descriptive elements manifest themselves in an
+of IAph 1.4 (Fig.~\ref{fig:2}), one can to observe how descriptive elements manifest themselves in an
 inscription: how varying letter sizes (2-3 cm) represent a lack of uniformity and crowding as space becomes more
@@
 
-\bigskip
-
-\bigskip
-
- \includegraphics[width=8.758cm,height=7.668cm]{PaperproposalforEAGLEfinal-img002.jpg}
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{PaperproposalforEAGLEfinal-img002.jpg}
+\caption{Photograph of IAph 1.4}
+\label{fig:2}
+\end{figure}
 
@@
 of this letter), the rhos have small legs, making them appear more like their Latin counterpart, and the omicron has an
-oval shape with serifs only at one end of the lower bars.\footnote{ “Legged” rhos are observed primarily in Augustan
+oval shape with serifs only at one end of the lower bars.\footnote{ ``Legged'' rhos are observed primarily in Augustan
 and Julio-Claudian dedications and also on coins (cf. note 16). This style letterform is increasingly less common after
@@
 
-\bigskip
-
-\bigskip
-
-1.2.4. Looks can be deceiving. When Inscriptions defy our expectations: IAph 1.5 and IAph 1.6.
-
-\bigskip
-
-Based on the publication of these inscriptions in MAMA VIII, one could easily come to the conclusion that all the column
+\subsection{Looks can be deceiving. When Inscriptions defy our expectations: IAph 1.5 and IAph 1.6.}
+\noindent Based on the publication of these inscriptions in MAMA VIII, one could easily come to the conclusion that all the column
 dedications at the Temple of Aphrodite looked quite similar. However, upon visiting the site of Aphrodisias in the
@@
 of letter forms. } While significant variations in letter size can be observed in poorly executed inscriptions
-throughout Aphrodisias’ history, one may not expect such a significant variation between two “copies”.\footnote{
-A.Graham 2015 (forthcoming) illustrates a further case study of “copy” inscriptions at Aphrodisias.} \ The published
+throughout Aphrodisias’ history, one may not expect such a significant variation between two ``copies''.\footnote{
+\citep{graham2015} illustrates a further case study of ``copy'' inscriptions at Aphrodisias.} \ The published
 text of IAph 1.5 also indicates indentations on both sides of the word demos in the bottom line. While double
@@
 
-\bigskip
+\begin{quotation}
+\noindent MAMA 438\\
+\begin{otherlanguage*}{greek}
+\noindent Εὔμαχος Ἀθηνα-\\
+γόρου τοῦ Ἀθηναγό-\\
+ρου τοῦ Εὐμάχου Δι-\\
+ογένης φιλόκαισαρ\\
+καὶ Ἀμμιὰς Διονυσί-\\
+ου φύσι δὲ Ἀδράστου\\
+τοῦ Μόλωνος Ὀλυν-\\
+πιὰς τὸν κίονα θεᾷ\\
+Ἀφροδίτῃ καὶ τῷ\\
+δήμῳ.
+\end{otherlanguage*}
+\end{quotation}
 
-[Warning: Draw object ignored][Warning: Draw object ignored]
+\begin{quotation}
+\noindent IAph 1.5.\\
+\begin{otherlanguage*}{greek}
+\noindent Εὔμαχος Ἀθηνα-\\
+γόρου τοῦ Ἀθηναγό-\\
+ρου τοῦ Εὐμάχου Δι-\\
+ογένης Φιλόκαισαρ\\
+ καὶ Ἀμμιὰς Διονυσί-\\
+ου φύσι δὲ Ἀδράστου\\
+τοῦ Μόλωνος Ὀλυν-\\
+πιὰς τὸν κίονα θεᾷ\\
+Ἀφροδίτῃ καὶ τῷ
+\end{otherlanguage*}
+vac. \begin{otherlanguage*}{greek}
+Δήμῳ
+\end{otherlanguage*} vac.
+\end{quotation}
 
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-An image \ (Figure 3 Photograph of IAph 1.5) illustrates further differences between the appearance of this inscription
+An image (Fig.~\ref{fig:3}) illustrates further differences between the appearance of this inscription
 and the comparative materials from the site (IAph 1.1, 1.2, 1.4, 1.7 and 1.8).
 
-\bigskip
-
-\bigskip
-
- \includegraphics[width=9.036cm,height=6.985cm]{PaperproposalforEAGLEfinal-img003.jpg}
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{PaperproposalforEAGLEfinal-img003.jpg}
+\caption{Photograph of IAph 1.5}
+\label{fig:3}
+\end{figure}
 
@@
 later period, seem to suddenly emerge nearly a century early.\footnote{ Archaising and imitation of earlier lettering
-is evident in at least one inscription (IAPH 13.116) Reynolds 1982, 155-166). } Any one of these features, letter
+is evident in at least one inscription (IAPH 13.116) \citet[155-166]{reynolds1982}). } Any one of these features, letter
 variations, stylistic letterforms, use of serifs, indentations in the text, correction of errors in a previous text,
 would not stand well alone, but taken together, a more compelling case can be made for the reassessment of this
-inscription’s date. \
+inscription’s date.
 
-\bigskip
-
-The website records the date of this inscription just like IAph 1.4: citing “context, lettering and prosopography.” For
+The website records the date of this inscription just like IAph 1.4: citing ``context, lettering and prosopography.'' For
 those trying to gain an understanding of letterforms at Aphrodisias, this is somewhat confusing and it is not the only
 example in which the lettering and organization of the inscription do not match date provided.\footnote{ Honours for
-Zoilos (IAph 8.203) is dated by “prosopography” to the 1st c. BC but the letters are square Imperial forms and do not
-match Zoilos’ other dedications (8.1, 1.1,1.2). Honours for P. and M. Vinicius (IAph 3.101) are dated as “Augustan” and
-“Tiberian” by “lettering” though prosopography is known (there is some controversy cf. Reynolds 1982,175). Stylistic
+Zoilos (IAph 8.203) is dated by ``prosopography'' to the 1st c. BC but the letters are square Imperial forms and do not
+match Zoilos’ other dedications (8.1, 1.1,1.2). Honours for P. and M. Vinicius (IAph 3.101) are dated as ``Augustan'' and
+``Tiberian'' by ``lettering'' though prosopography is known (there is some controversy cf. Reynolds 1982,175). Stylistic
 elements (arrangement, spacing and decoration), particularly in M. Vinicius’ base, reflect qualities of
-Claudian-Flavian period inscriptions (which is not excluded by prosopographical dates). \ In both cases, conflicting
-dating criteria are omitted. \ \ \ } The date of the text is not incorrect, insofar as it was originally inscribed at
-this time and the prosopography as well as the formula of the text support this. \ However, one struggles to see how
+Claudian-Flavian period inscriptions (which is not excluded by prosopographical dates). In both cases, conflicting
+dating criteria are omitted. } The date of the text is not incorrect, insofar as it was originally inscribed at
+this time and the prosopography as well as the formula of the text support this. However, one struggles to see how
 the lettering and/or the arrangement of the text corroborate this date. An answer may be found in further examination
@@
 added under the emperor Hadrian, a time during which Athenagoras’ descendants were alive and prospering in
-Aphrodisias.\footnote{ Smith and Ratté, 1995, 43. For an analysis of the temenos plan, see Doruk in Roueche and Erim
-1990, 66-74. \par For Athenagoras descendants, cf. note 18 and Reynolds 1999b, 327-334} \
-
-\bigskip
+Aphrodisias.\footnote{\citep[43]{smith1995}. For an analysis of the temenos plan, see \citet[66-74]{doruk1990}. For Athenagoras descendants, cf. note 18 and \citet[327-334]{reynolds1999b}.}
 
 One benefit of the website is that it facilitates searches for parallel texts in this context and this period. A
-Hadrianic building dedication from the temple, IAph 1.174 (Figure 4: Photograph of IAph 1.174) reveals similarly
+Hadrianic building dedication from the temple, IAph 1.174 (Fig.~\ref{fig:4}) reveals similarly
 rendered letterforms with little variation in size and a number of stylistic similarities: thin strokes with deep
@@
 
-\bigskip
-
-\bigskip
-
- \includegraphics[width=10.836cm,height=4.092cm]{PaperproposalforEAGLEfinal-img004.jpg}
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{PaperproposalforEAGLEfinal-img004.jpg}
+\caption{Photograph of IAph 1.174}
+\label{fig:4}
+\end{figure}
 
@@
 
-The entry for IAph.1.6 represents a different response to a similar problem. \ It demonstrates that scholars are willing
+The entry for IAph.1.6 represents a different response to a similar problem. It demonstrates that scholars are willing
 to use letterforms to support a recarving when the case is sufficiently extreme. The text, only mentioned in MAMA VIII
-as a “copy”, presents a number of irregularities, repeating some errors, correcting others, and making quite a few new
+as a ``copy'', presents a number of irregularities, repeating some errors, correcting others, and making quite a few new
 ones. Whilst maintaining the first 3 lines of the arrangement in IAph 1.4, this inscription compounds the error on line
@@
 
-\bigskip
+\begin{quotation}
+\noindent IAPh 1.6.\\
+\begin{otherlanguage*}{greek}
+\noindent Εὔμαχος Ἀθηναγό-\\
+ρου τοῦ Ἀθηναγόρου\\
+τοῦ Εὐμάχου Διογε-\\
+\{γέ\}νης Φιλόκαισαρ καὶ\\
+  Ἀμιὰς Διονυσίου φύσι\\
+δὲ Ἀδράστου τοῦ Μό-\\
+λω<ν>ος Ὀλυ<ν>πιὰς τὸν κί-\\
+ονα θεᾷ Ἀφροδίτῃ \{τὸν\}\\
+\{κίονα\} καὶ τῷ Δήμῳ
+\end{otherlanguage*}
+\end{quotation}
 
-[Warning: Draw object ignored]
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-Reading this inscription is even more of challenge (Figure 5. Photograph of I.Aph 1.6), one can suffer vertigo as the
+Reading this inscription is even more of challenge (Fig.~\ref{fig:5}), one can suffer vertigo as the
 lines run up and down and the letters run into the margins. There is no use of spaces or decorations to clarify or
 distinguish sections of the text and there are quite a few inadvertent errors to make it difficult, even to the trained
-eye. Although the prosopography and context support an earlier date, the entry suggests that this “inelegant” text was
-“recarved ?” The entry does not, however, propose a date or explain why, in this instance, recarving is a viable
-conclusion. \ While a methodology is evident in the IAph2007 dating format, it is not employed consistently or
-transparently in this case. \
+eye. Although the prosopography and context support an earlier date, the entry suggests that this ``inelegant'' text was
+``recarved ?'' The entry does not, however, propose a date or explain why, in this instance, recarving is a viable
+conclusion. While a methodology is evident in the IAph2007 dating format, it is not employed consistently or
+transparently in this case.
 
-\bigskip
-
-\bigskip
-
-\bigskip
-
- \includegraphics[width=7.23cm,height=5.549cm]{PaperproposalforEAGLEfinal-img005.jpg} 
-\includegraphics[width=7.396cm,height=5.456cm]{PaperproposalforEAGLEfinal-img006.jpg}
-
-\bigskip
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{PaperproposalforEAGLEfinal-img005.jpg}
+  \includegraphics[width=\columnwidth]{PaperproposalforEAGLEfinal-img006.jpg}
+\caption{Photograph of I.Aph 1.6}
+\label{fig:5}
+\end{figure}
 
@@
 that the inscription is from the late Antique period at Aphrodisias, possibly after an earthquake in 359 BC (for
-parallels see ALA 29 and 30), before the conversion of the temple into a Christian church (after 450 AD).\footnote{
-R.R.R. Smith 1996,41-42.} Acknowledging that the text was reinscribed centuries later affords further insight into both
+parallels see ALA 29 and 30), before the conversion of the temple into a Christian church (after 450 AD).\footnote{\citet[41-42]{smith1996}.} Acknowledging that the text was reinscribed centuries later affords further insight into both
 the inscription and act of recarving as a process that often changed an inscription but did not necessarily improve it.
-
-\bigskip
 
@@
 the ancient world but by minimizing the conflicting elements of these inscriptions, one potentially overlooks this
-aspect of an inscription.\footnote{ Thomas and Witschel 1995, 135-177. }
+aspect of an inscription.\footnote{\citet[135-177]{thomas1990}.}
 
-\bigskip
-
-1.2.5 \ Conclusions on the case study
-
-\bigskip
-
-This survey of column dedications has demonstrate how a series of inscriptions, which were represented as a series of
+\subsection{Conclusions on the case study}
+\noindent This survey of column dedications has demonstrate how a series of inscriptions, which were represented as a series of
 similar texts in MAMA VIII can be seen in a fundamentally different way in IAPh 2007. The digital publication has a
@@
 an inscription in a digital world, this is probably as close a person can come to assessing the face value of an
-inscription. \ It is a tremendous step forward in addressing the longstanding limitations inherent in the publication
-of epigraphic evidence, though some challenges remain. \ The final section will consider how we might further use this
+inscription. It is a tremendous step forward in addressing the longstanding limitations inherent in the publication
+of epigraphic evidence, though some challenges remain. The final section will consider how we might further use this
 information, in terms of adding clarity and transparency to the dating process, as well as in our approach to these
-materials. \
+materials.
 
-\bigskip
+\section{Palaeography in a Digital World: Monumental Problems \& Solutions}
 
-\ 1.3. \ Palaeography in a Digital World: Monumental Problems \& Solutions
-
-\bigskip
-
-\begin{enumerate}
-\item \begin{enumerate}
-\item \begin{enumerate}
-\item Digital Epigraphy. New Method: New Methodology?
-\end{enumerate}
-\end{enumerate}
-\end{enumerate}
-
-\bigskip
-
-For students and academics alike, dating by letter forms is an almost “mystic” practice: something upon which experts
-often comment but more rarely explain or demonstrate in practice. \ Recent epigraphic handbooks keep the term
-palaeography at arms length, using it interchangeably with discussions of letterforms.\footnote{ \ In the index (Bruun
-and Edmundson et al., 2015, p. 880) \ “palaeography” is cross-referenced with “Letter forms” and though used by O.
-Salomies (chapter 9), the word is not defined in the preceding chapters (Bruun and Edmundson et al., 2015, 155).
-\ Letterforms are described generally with charts of figure numbers rather than images to directly illustrate changing
-styles (Edmundson 2015, pp. 123-4). Cooley’s manual offers a more detailed assessment of lettering with case studies to
-illustrate the limitations of using lettering (Cooley 2012, 423-33)} \ The caution of these scholars is justified and
+\subsection{Digital Epigraphy. New Method: New Methodology?}
+\noindent For students and academics alike, dating by letter forms is an almost ``mystic'' practice: something upon which experts
+often comment but more rarely explain or demonstrate in practice. Recent epigraphic handbooks keep the term
+palaeography at arms length, using it interchangeably with discussions of letterforms.\footnote{In the index \citep[880]{bruun2015} ``palaeography'' is cross-referenced with ``Letter forms'' and though used by O.
+Salomies (chapter 9), the word is not defined in the preceding chapters \citep[155]{bruun2015}.
+Letterforms are described generally with charts of figure numbers rather than images to directly illustrate changing
+styles (\citep[123-4]{bruun2015}). Cooley’s manual offers a more detailed assessment of lettering with case studies to
+illustrate the limitations of using lettering \citep[423-33]{cooley2012}} The caution of these scholars is justified and
 understandable: conclusions based on a single element of an inscription are precarious and, perhaps more importantly,
@@
 antiquity. Writing, both ancient and modern, is a product of a number of factors, all of which functioned together in
-the image of writing (margins, lettering, indentations punctuation).\footnote{ \ Woolf 1996, 25-27, Bodel 2001, 3-5,
-\ Cooley 2012, 433-437; Panciera 2012, 1-10. \ } As few would look at a document today and say “that’s a fine Helvetica
-10 point!,” we should be cautious in an assessment of palaeography that excludes the visual elements which were
+the image of writing (margins, lettering, indentations punctuation).\footnote{\citet[25-27]{woolf1996}, \citet[3-5]{bodel2001},
+\citet[433-437]{cooley2012}; \citet[1-10]{panciera2012}.} As few would look at a document today and say ``that’s a fine Helvetica
+10 point!,'' we should be cautious in an assessment of palaeography that excludes the visual elements which were
 inextricably linked an inscription’s appearance (e.g. alignment, margins, spacing, punctuation). Modern definitions and
 studies of manuscripts suggest a broader scope of inquiry, which includes handwriting together with decorations and
-spatial organization as well as subsequent comments in the margins.\footnote{ \ Buonocore 2015, 21-37. } These
+spatial organization as well as subsequent comments in the margins.\footnote{\citet[21-37]{buonocore2015}. } These
 practices suggest a discipline with an interest not only in the evolution of lettering but in the appearance and
@@
 
-\bigskip
-
 While traditional text-based modes of publication such as MAMA VIII have, at times, constrained the study of hands to an
 analysis of lettering, digital corpora offer new opportunities to view, analyse and incorporate broader scope of visual
-elements in the interpretation of an inscription. \ This is not to say that the traditional methods or charts of
+elements in the interpretation of an inscription. This is not to say that the traditional methods or charts of
 letterforms development should be abandoned, merely that the methodology could be expanded, as it has been in this case
@@
 
-\bigskip
-
-\bigskip
-
-1.3.2 Employing change: How we might improve presentation of materials on the website.
-
-\bigskip
-
-One difficulty of the current site, observed in the assessment of IAph 1.5 is that despite the clear criteria for
+\subsection{Employing change: How we might improve presentation of materials on the website.}
+\noindent One difficulty of the current site, observed in the assessment of IAph 1.5 is that despite the clear criteria for
 dating, the terms do not reflect a consistent methodology in dating, or the contradictions in the process. This is
@@
 
-\bigskip
-
 \ Further improvements, could be made with additional pages on lettering and context. \ While lettering is undoubtedly
 employed in dating inscriptions at Aphrodisias, one must also consider how this dating is represented in the published
-materials. Epigraphers, who have described monumental lettering as ‘Augustan’, ‘distinctively Julio-Claudian’ or
-‘Domitianic’ have already acknowledged that such distinctions exist at Aphrodisias, but the classifications remain
-undefined.\footnote{ Studies of inscriptions describe lettering as “distinctively Julio-Claudian” (Reynolds 1981 (nos.
-2\&3) 317-318), “Triumviral or Augustan” (Reynolds 1982 docs. 35-37, 159-163), and ‘Domitianic’ \ (Chaniotis 2004, no.
-14), suggest that such distinctions are evident at Aphrodisias. \ } While such definitions are often problematic, so is
+materials. Epigraphers, who have described monumental lettering as `Augustan', `distinctively Julio-Claudian' or
+`Domitianic' have already acknowledged that such distinctions exist at Aphrodisias, but the classifications remain
+undefined.\footnote{ Studies of inscriptions describe lettering as ``distinctively Julio-Claudian'' \citep[nos.
+2\&3) 317-318]{reynolds1981}, ``Triumviral or Augustan'' \citep[docs. 35-37, 159-163]{reynolds1982}, and ‘Domitianic’ \ (Chaniotis 2004, no.
+14), suggest that such distinctions are evident at Aphrodisias.} While such definitions are often problematic, so is
 a situation in which a broader audience accepts and uses a series of dates, without understanding how certain criteria
@@
 
-\bigskip
-
-\ \ Finally, while inscriptions are given a good deal of context in IAph 2007, the epigraphic information remains
+Finally, while inscriptions are given a good deal of context in IAph 2007, the epigraphic information remains
 separated from the archaeological studies. Both of these are factors in the dating process and, as we observed in IAph
 1.5 and 1.6 can often facilitate the study of an inscription, both in reconciling discrepancies and searching for
-parallel inscriptions. \ With inscriptions grouped by context and an excavation history that is referenced (but not
+parallel inscriptions. With inscriptions grouped by context and an excavation history that is referenced (but not
 explained) it might be helpful to have a brief building history for each context. In the case of Aphrodisias, these
-materials and further bibliographies are \ available on NYU excavation websites and could be easily connected with
-weblinks.\footnote{ NYU website. \ http://www.nyu.edu/projects/aphrodisias/home.ti.htm. \ Recent excavation reports are
-reference here as well: \ http://www.nyu.edu/gsas/dept/fineart/academics/aphrodisias/aphrodisias.htm.}
-
-\bigskip
+materials and further bibliographies are available on NYU excavation websites and could be easily connected with
+weblinks.\footnote{ NYU website. \url{http://www.nyu.edu/projects/aphrodisias/home.ti.htm}. Recent excavation reports are
+reference here as well: \url{http://www.nyu.edu/gsas/dept/fineart/academics/aphrodisias/aphrodisias.htm}.}
 
 The benefits and the challenges observed in the IAph 2007 digital publication are applicable to a number of digital
-resources. \ As we attempt to bring corpora of unprecedented size to a digital realm, we must consider not only how we
+resources. As we attempt to bring corpora of unprecedented size to a digital realm, we must consider not only how we
 represent this information but how we engage a wider readership in epigraphic materials. It would not take a great deal
@@
 
+%\nocite{} check after compiling with quoted bibliography
+
 \bibliographystyle{sapauth-eng}

### 2015-12-16 source/13/EAGLE2016_Full_Paper_Verreth.tex
started work on tex of papers, added folders for CNR papers, changed main stracture to include keynotes and AHE abstracts

@@
 %%%%%%%%
-\title{}
-\author{Herbert Verreth}
+\title{Trismegistos Places, a geographical index for all Latin inscriptions}
+\author[KUL]{Herbert Verreth\corref{first}}
+\address[KUL]{Trismegistos Project, Katholieke Universiteit Leuven, Belgium}
+\cortext[first]{Corresponding author. Email: Herbert.Verreth@arts.kuleuven.be}
 \date{2015-11-15}
 \begin{document}
-[Warning: Draw object ignored]Abstract
 
+\maketitle
+\begin{abstract}
 The Trismegistos database has recently created a geographical index for all Latin inscriptions. For the moment we have
@@
 we made for the presentation of the material.
+\end{abstract}
+\keywords{Imperium Romanum, Latin, epigraphy, topography, geography, Trismegistos.}
 
-Keywords
-
-Imperium Romanum, Latin, epigraphy, topography, geography, Trismegistos.
-
-\bigskip
-
-\ \ The Trismegistos (TM) database (http://www.trismegistos.org) of the Katholieke Universiteit Leuven in Belgium
+\noindent The Trismegistos (TM) database (\url{http://www.trismegistos.org}) of the Katholieke Universiteit Leuven in Belgium
 gathers the metadata for all documentary and literary texts from Egypt and the Ancient World in general written in
@@
 
-\ \ Trismegistos is a relational database created within the computer program Filemaker (with as latest version
+ Trismegistos is a relational database created within the computer program Filemaker (with as latest version
 Filemaker Pro 14). Its contents are uploaded weekly to an online MySQL / PHP environment. Attached to the main TM Text
@@
 the relevant information for every card from the main file, so that no double work is needed. With the help of the
-Papyrological Navigator (PN) (http://www.papyri.info) we developed a PHP environment with the assistance of a
+Papyrological Navigator (PN) (\url{http://www.papyri.info}) we developed a PHP environment with the assistance of a
 programmer (Jeroen Clarysse), to tag all words starting with a capital occurring in the published papyrological texts,
@@
 
-\ \ The main credits for the whole set-up of this new project go to Mark Depauw, the Trismegistos director. The general
+ The main credits for the whole set-up of this new project go to Mark Depauw, the Trismegistos director. The general
 idea remained the tagging of all words starting with a capital, but this time Depauw tried a new approach, which was
@@
 
-\ \ Caio Annioleno Cai filio Arnensi Karthaginiensi Galliano
+\begin{table}[htp]
+\begin{center}
+\begin{tabular}{ccccccc}
+Caio & Annioleno & Cai & filio & Arnensi & Karthaginiensi & Galliano\\
+dat & dat & gen & filius & tribus & origo & dat\\
+\end{tabular}
+\end{center}
+\label{Example}
+\end{table}%
 
-\ \ dat\ \ dat \ \ \ \ \ gen \ filius\ \ tribus\ \  \ \ \ \ origo\ \ \ \ dat
-
-\bigskip
-
- \includegraphics[width=10.881cm,height=7.505cm]{EAGLE2016FullPaperVerreth-img001.png}
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img001.png}
+\caption{default}
+\label{fig:1}
+\end{figure}
 
@@
 
-\ \ This is also the phase where the toponyms come in (which can occur both in the first and in the second group).
+This is also the phase where the toponyms come in (which can occur both in the first and in the second group).
 Within the EAGLE project we created already a fairly large reference corpus for toponyms from the Roman empire, but
@@
 toponyms mentioned in Egyptian papyrological and Latin epigraphical sources.\footnote{ For more information on TM Geo,
-cf. Verreth, Herbert, Topography of Egypt online, in Proceedings of the 27th International Congress of Papyrology in
-Warsaw, July-August 2013 (forthcoming).} We played with the idea of automatic matching, like we did for the personal
+\citep{verreth2013}.} We played with the idea of automatic matching, like we did for the personal
 names, but except for the case of the relatively straightforward tribus names, this yielded no satisfactory results. A
@@
 of strings such as Colonia Ulpia Traiana Augusta Fidelis Lepcis Magna (TM 198383) or Municipium Augustum Hipponiensium
-Regiorum (TM 200133) seemed too cumbersome. In the end, we settled for plan B, which shows that in 'Digital
+Regiorum (TM 200133) seemed too cumbersome. In the end, we settled for plan B, which shows that in `Digital
 Humanities', the human component is still essential: we had to go through all 900.000 capital cluster cards manually,
@@
 
-\ \ Phase 1, the identification of toponyms in the capital cluster strings, was finished in the beginning of July 2015,
-and we have now entered phase 2, the incorporation of the capital cluster file into the 'real' TM Georef file. For
+Phase 1, the identification of toponyms in the capital cluster strings, was finished in the beginning of July 2015,
+and we have now entered phase 2, the incorporation of the capital cluster file into the `real' TM Georef file. For
 every place listed in TM Geo we try to list all the ancient text references where that place is attested. The file of
@@
 exported to a separate Georef card. When a toponym exists out of several consecutive elements, like the colonia and
-municipium examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as 'Bythina et
-Ponthus', which cannot belong to the same capital cluster string because of the intermediary 'et', are exported double
+municipium examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as `Bythina et
+Ponthus', which cannot belong to the same capital cluster string because of the intermediary `et', are exported double
 and then afterwards joined manually. For the moment TM Georef lists 67.820 geographical references attested in Latin
@@
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img002.png}
+\caption{default}
+\label{fig:2}
+\end{figure}
 
- \includegraphics[width=10.881cm,height=5.008cm]{EAGLE2016FullPaperVerreth-img002.png}
-
-\bigskip
-
-\ \ In this phase we also start comparing the reading of the toponym in EDCS with the readings of the same passage in
+In this phase we also start comparing the reading of the toponym in EDCS with the readings of the same passage in
 EDH, EDR, HEp and EDB, which can all be shown simultaneously in the Filemaker database. In theory it is possible to
@@
 
-\ \ Since the users of TM must be able to look for specific spelling variants of each toponyms, all these references are
+Since the users of TM must be able to look for specific spelling variants of each toponyms, all these references are
 presented the way they are on the stone, with as little additions or emendations as possible, except of course for any
@@
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img003.png}
+\caption{default}
+\label{fig:3}
+\end{figure}
 
- \includegraphics[width=10.881cm,height=11.984cm]{EAGLE2016FullPaperVerreth-img003.png}
-
-[Fig. 3]
-
-\ \ For every text in TM we try to give a reference to the most authoritative edition, where the user can find the best
+For every text in TM we try to give a reference to the most authoritative edition, where the user can find the best
 and most up to date reading and interpretation of that text. As authoritative editions we preferably use CIL, Année
@@
 
-\ \ A major problem is the dating of the texts. Unfortunately not every edition provides a date for each inscription.
+A major problem is the dating of the texts. Unfortunately not every edition provides a date for each inscription.
 Even if the scholar who publishes the text, has a fairly good idea of the century or range of time to which the text
@@
 
-\ \ The third and final phase involves the context of the toponym. In the field Detail we give a plain translation of
+The third and final phase involves the context of the toponym. In the field Detail we give a plain translation of
 the immediate phrase to which the toponym belongs. The translation should be a standardised as possible, with termini
-technici preferably added in Latin, so that the users can easily search for them in the database; e.g. 'Tiberius Iulius
-Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris', 'praefectus
-Aegypti', 'cohors I Tungrorum'. If the place is explicitly ascribed to a provincia or a region, this provincia is
+technici preferably added in Latin, so that the users can easily search for them in the database; e.g. `Tiberius Iulius
+Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris', `praefectus
+Aegypti', `cohors I Tungrorum'. If the place is explicitly ascribed to a provincia or a region, this provincia is
 listed in the field 'Administrative situation'. If a town is explicitly called an oppidum, a vicus or a civitas, this
@@
 start asking quite specific questions; e.g. the first and last attestation of a place in the sources; the periods in
-which a town was called a colonia or a municipium; the places in which the 'ala I Thracum Mauretana' has been attested.
+which a town was called a colonia or a municipium; the places in which the `ala I Thracum Mauretana' has been attested.
 
-\ \ TM is a relational database, which implies that is possible to get to the information from different angles. If
+ TM is a relational database, which implies that is possible to get to the information from different angles. If
 someone is studying a certain text, he can get a list of all toponyms mentioned in that text. On the other hand, if
@@
 
-\ \ We are aware that this is a very succinct presentation of the new and exciting developments in Trismegistos Places,
+We are aware that this is a very succinct presentation of the new and exciting developments in Trismegistos Places,
 but everybody interested is always more than welcome to ask for more information. Please feel free to provide us with
 any addenda or corrigenda to the database you might have.
+
 \bibliographystyle{sapauth-eng}

### 2015-12-16 source/17/Zuddas Epigraphy in Italian High Schools (final).tex
started work on tex of papers, added folders for CNR papers, changed main stracture to include keynotes and AHE abstracts

@@
 %%%%%%%%
-\title{}
-\author{Utente}
+\title{EPIGRAPHY IN ITALIAN HIGH SCHOOLS}
+\author[up]{Enrico Zuddas\corref{first}}
+\address[up]{University of Perugia}\cortext[first]{Corresponding author. Email: enricozudd@yahoo.com}
 \date{2015-11-16}
 \begin{document}
-Abstract
 
-\begin{figure}
-\centering
-\begin{minipage}{10.901cm}
-1.\ \ EPIGRAPHY IN ITALIAN HIGH SCHOOLS\newline
-
-Enrico Zuddas
-
-University of Perugia
-
-enricozudd@yahoo.com
-\end{minipage}
-\end{figure}
+\maketitle
+\begin{abstract}
 The paper focuses on the possible uses of Epigraphy when teaching Latin to Italian High School students. Inscriptions
-(especially from the school’s territory) make Latin “come to life” because even the simplest of texts is real.
+(especially from the school’s territory) make Latin ``come to life'' because even the simplest of texts is real.
 Inscriptions are an excellent tool to internalize declensions, but they can also broaden the horizons offered by the
@@
 traditional supports.
+\end{abstract}
+\keywords{High School Education, Latin Grammar, Lexicon, Translations, EDR Database, Roman Asisium.}
 
-Keywords
-
-High School Education, Latin Grammar, Lexicon, Translations, EDR Database, Roman Asisium.
-
-\bigskip
-
-1.1. Introduction
-
-High school students often feel that Latin is not a “real” language. Every attempt to use Latin as a modern language, by
+\section{Introduction}
+\noindent High school students often feel that Latin is not a ``real'' language. Every attempt to use Latin as a modern language, by
 the creation of words from the contemporary world or by using it in a conversation, makes Latin seem even more
@@
 quite easily readable. They exist in reality, they can be seen and even touched – which should never be forgotten in a
-society where visual communication is so important. Through epigraphy the Classics become, in a way, “multimedia”: this
-is what scholars mean by the expression “words on stone”. Even if generally, when visiting a museum, inscriptions are
+society where visual communication is so important. Through epigraphy the Classics become, in a way, ``multimedia'': this
+is what scholars mean by the expression ``words on stone''. Even if generally, when visiting a museum, inscriptions are
 not of such interest as the statues or mosaics, they do strike students as something concrete and, in a certain sense,
-“alive”. That is why they are a good tool to improve the ability to read, understand and translate Latin.
+``alive''. That is why they are a good tool to improve the ability to read, understand and translate Latin.
 
@@
 why in this paper I will mainly refer to inscriptions from modern Umbria (regiones VI and VII, from Perusia) and
-especially from Asisium, used at school during my experimental lessons. The local museum, best known as the “Foro
-romano”, hosts a huge collection of inscriptions that were catalogued in 2008 by pupils of the Liceo Properzio, under
+especially from Asisium, used at school during my experimental lessons. The local museum, best known as the ``Foro
+romano'', hosts a huge collection of inscriptions that were catalogued in 2008 by pupils of the Liceo Properzio, under
 my supervision, with details of the type of inscription, material, place of origin and chronology. Previously a booklet
 about Roman Assisi had been
-written\footnote{http://www.liceoassisi.it/index.php?option=com\_content\&view=article\&id=47:assisi-romana\&catid=6:i-nostri-risultati\&Itemid=34.},
+written\footnote{\url{http://www.liceoassisi.it/index.php?option=com\_content\&view=article\&id=47:assisi-romana\&catid=6:i-nostri-risultati\&Itemid=34.}},
 both in Italian and English, comprising a page about the famous tetrastyle with a translation of the inscription on the
@@
 
-1.2. How to use inscriptions
-
-Inscriptions in high school can be used in a variety of ways, but particularly in the following fields:
+\section{How to use inscriptions}
+\noindent Inscriptions in high school can be used in a variety of ways, but particularly in the following fields:
 
 \begin{itemize}
-\item elements of Latin grammar, syntax and linguistics\footnote{ Hartnett 2012; McCarthy 1992.};
+\item elements of Latin grammar, syntax and linguistics\footnote{\citet{Hartnett2012}; \citet{McCarthy1992}.};
 \item culture and history;
@@
 
-\bigskip
-
-1.2.1. Lexicon in context
-
-Although the acquisition of lexicon is essential in order to access the conceptual categories of a culture, it is
+\subsection{Lexicon in context}
+\noindent Although the acquisition of lexicon is essential in order to access the conceptual categories of a culture, it is
 extremely difficult to learn words without using them actively, and it is a terrible mistake, often made by school
@@
 
+\begin{quotation}
 P. Decimius P. l. Eros / Merula, medicus / clinicus, chirurgus, / ocularius, VIvir. / Hic pro libertate dedit
@@
 dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni ((sestertium))... 
+\end{quotation}
 
 not only does a student make contact with the technical lexicon of medicine and words related to building activities
-(statuas ponere, vias sternere), but he can also perceive the “evergetic spirit” of an ancient society: a physician was
+(statuas ponere, vias sternere), but he can also perceive the ``evergetic spirit'' of an ancient society: a physician was
 usually a freedman coming from the East, who could be very rich and spend his money for public utility; whenever a
@@
 about Roman virtues and mos maiorum, the famous Clipeus from Arles (\emph{AE} 1952, 0165 = \emph{AE} 1994, 0227) is a
-perfect complement to chapter 34 of the Res Gestae (formally an inscription, too): \emph{Senatus / populusque Romanus /
+perfect complement to chapter 34 of the Res Gestae (formally an inscription, too): \begin{quotation}
+\emph{Senatus / populusque Romanus /
 imp(eratori) Caesari / Divi f(ilio) / Augusto / co(n)s(uli) VIII dedit clupeum / virtutis, clementiae, / iustitiae,
@@
 
-\bigskip
+\end{quotation}
 
-1.2.2. Grammar
-
-One of the biggest problems in teaching Latin to young students is that they find it really difficult to understand
+\subsection{Grammar}
+\noindent One of the biggest problems in teaching Latin to young students is that they find it really difficult to understand
 declensions. In the preliminary lessons the sentences used are very simple and can be understood and even translated
@@
 
-\textwide{\FilledBigTriangleRight} CIL 11, 05501 = EDR025449: Noniae Privatae, C. Propertius -{}-{}-{}-{}-{}-
+\begin{itemize}
+\item CIL 11, 05501 = EDR025449: \begin{quotation}
+Noniae Privatae, C. Propertius -{}-{}-{}-{}-{}-
+\end{quotation}
 
@@
 
-\textwide{\FilledBigTriangleRight} CIL 11, 05461 = EDR025411: A(n)noru(m) XIX. Calventia C(ai) f(ilia) Polla, L(ucius)
+\item CIL 11, 05461 = EDR025411: \begin{quotation}
+A(n)noru(m) XIX. Calventia C(ai) f(ilia) Polla, L(ucius)
 Vistinius vir, Gavia mater posuer(unt).
 
+\end{quotation}
 This inscription offers the chance to get feedback about mistakes in the use of language, or to reflect on mortality and
@@
 
-\textwide{\FilledBigTriangleRight} CIL 11, 05399 = EDR025349: P(ublius) Decimius P(ubli) l(ibertus) Eros Merula VIvir
+\item CIL 11, 05399 = EDR025349: \begin{quotation}
+P(ublius) Decimius P(ubli) l(ibertus) Eros Merula VIvir
 viam a cisterna ad domum L(uci) Muti stravit ea pecunia -{}-{}-{}-{}-{}- 
+\end{quotation}
 
@@
 clause that, in this case, has evidently been lost).
-
-\bigskip
+\end{itemize}
 
@@
 They give a false image of antiquity: for instance, when you start with the first declension you get the wrong
-impression of a “female” world simply because the male nouns are rare. Students are unimpressed; these sentences have
+impression of a ``female'' world simply because the male nouns are rare. Students are unimpressed; these sentences have
 no significance for them. Those provided by inscriptions are equally easy and short, but they are not banal because
@@
 
+\begin{quotation}
 Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(- - -),
@@
 
+\end{quotation}
 The comprehension of the text is easy, especially with the abbreviations solved. Even so, many considerations can be
@@
 \item for history: the use of the Latin language prior to the Social War, as proof of the intense Romanization of the
-area at the end of the II century B.C.\footnote{ Coarelli 1991.}; the presence of elements in the names that are not
+area at the end of the II century B.C.\footnote{\citet{Coarelli1991}.}; the presence of elements in the names that are not
 Roman but of Umbrian origin;
 \item for archaeology: the building of terraces to create public spaces in a town like Assisi established on a hill; the
-identification of the area around San Rufino as the “acropolis”; the incorporation of the Roman wall into the left nave
+identification of the area around San Rufino as the ``acropolis''; the incorporation of the Roman wall into the left nave
 of the church (the inscription being still in situ).
@@
 
-\bigskip
-
 Schoolbooks rarely present these opportunities offered by inscriptions, so the main problem for a teacher is to access
-appropriate material\footnote{ See the observations made by Carpenter 2006.}. Holding a PhD in Late Antiquity and
+appropriate material\footnote{ See the observations made by \citet{Carpenter2006}.}. Holding a PhD in Late Antiquity and
 Umbrian inscriptions, I am fortunate enough to already know many sources. However, a graduate may not have enough
@@
 
-\textwide{\FilledBigTriangleRight} Barbieri 2015, 39: the inscription of the architect C. Vettius Gratus (CIL X 3392) is
+\begin{itemize}
+\item \citet[39]{Barbieri2015}: the inscription of the architect C. Vettius Gratus (CIL X 3392) is
 merely a decorative element on the page, without any relation to the topic (phonetic changes from Latin to Italian);
 
-\textwide{\FilledBigTriangleRight} Domenici 2012, 40: programmata from Pompeii are introduced to explain the Roman
+\item \citet[40]{Domenici2012}: programmata from Pompeii are introduced to explain the Roman
 naming system;
 
-\textwide{\FilledBigTriangleRight} Gambis Meini, Roffi, Guidotti Bacci 2013, 269-270: tabellae defixionum are used for
+\item \citet[269-270]{Gambis2013}: tabellae defixionum are used for
 different purposes (demonstrative adjectives, functions of subjunctive); the subject (magic in the ancient world) may
@@
 students.
+\end{itemize}
 
-\bigskip
-
-\bigskip
-
-\bigskip
-
-1.3. Translations
-
-Translating inscriptions is a difficult task for everyone and especially for high school students.
+\section{Translations}
+\noindent Translating inscriptions is a difficult task for everyone and especially for high school students.
 
@@
 Examples:
-
-\textwide{\FilledBigTriangleRight} CIL 11, 04431 = EDR025160: [Inf]austo, levis umbra, tuo mihi flebilis hora / sorte
+\begin{itemize}
+\item CIL 11, 04431 = EDR025160: \begin{quotation}
+[Inf]austo, levis umbra, tuo mihi flebilis hora / sorte
 tua certe tempus in omne fuit.
 
+\end{quotation}
 This funerary inscription from Ameria contains two verses, but the position of the words is tricky; infausto may be
@@
 
-\textwide{\FilledBigTriangleRight} CIL 11, 04391 = EDR025123: Iuliae M(arci) f(iliae) Felicitati, / uxori C(ai) Curiati
+ \item CIL 11, 04391 = EDR025123:\begin{quotation}
+ Iuliae M(arci) f(iliae) Felicitati, / uxori C(ai) Curiati
 Eutychetis / IIIIvir(i), magistrae Fortu/nae Mel(ioris), coll(egium) centonarior(um) / ob merita eius. Quo honore /
@@
 
+\end{quotation}
 Specialists are well used to an inscription like this. A high school student may find unexpected difficulties in
@@
 
+\end{itemize}
+
 Nevertheless, this kind of challenge is exactly what makes inscriptions the perfect tool to fully appreciate what the
@@
 
-The article published by F. Bigi in the Proceedings of the First EAGLE International Conference\footnote{ Bigi 2014.}
+The article published by F. Bigi in the Proceedings of the First EAGLE International Conference\footnote{\citet{Bigi2014}.}
 includes many observations regarding the problems that may be encountered, for instance when translating names and
@@
 
-\textwide{\FilledBigTriangleRight} CIL 11, 04213 = EDR130908: Interamna Nahars should be further qualified as “Terni” to
+\begin{itemize}
+\item  CIL 11, 04213 = EDR130908: Interamna Nahars should be further qualified as ``Terni'' to
 help those readers who are not familiar with Umbrian cities;
 
-\textwide{\FilledBigTriangleRight} CIL 11, 01925 = EDR142701: the names of the emperors M. Aurelius Antoninus and M.
+\item CIL 11, 01925 = EDR142701: the names of the emperors M. Aurelius Antoninus and M.
 Antoninus Pius Germaticus Sarmaticus need to be explained (Caracalla and Marcus Aurelius) to avoid confusion. It should
@@
 translated into Italian, even if it is advisable to transcribe them in the nominative case\footnote{
-http://www.eagle-network.eu/wiki/index.php/Guidelines\_for\_Translators.}; the same thing was done with cognomina ex
+\url{http://www.eagle-network.eu/wiki/index.php/Guidelines\_for\_Translators}.}; the same thing was done with cognomina ex
 virtute, considering that they are intuitively interpretable for an Italian.
-
-\bigskip
+\end{itemize}
 
@@
 The ministerial syllabus set out for the course focuses on Literature and culminates in a specific exam requirement,
-the translation of a piece of literary prose. Is this “epigraphical” activity helpful? Does it take up too much of the
-time which should be employed in translating the Classics? The answer to both these questions is “yes”. On the one
-hand, as I said, students have to “jump into translating” without a net (the Net, in fact). On the other hand, if at
+the translation of a piece of literary prose. Is this ``epigraphical'' activity helpful? Does it take up too much of the
+time which should be employed in translating the Classics? The answer to both these questions is ``yes''. On the one
+hand, as I said, students have to ``jump into translating'' without a net (the Net, in fact). On the other hand, if at
 the end of the final year pupils are required to translate a passage from certain authors, then clearly it would be
@@
 
-1.4. Inscriptions and Literature: a few samples
-
-On a more advanced level, inscriptions may also integrate certain aspects related to the study of Latin Literature. The
+\section{Inscriptions and Literature: a few samples}
+\noindent On a more advanced level, inscriptions may also integrate certain aspects related to the study of Latin Literature. The
 most typical example could be a comparison between the Tabula Claudiana (CIL 13, 01668) and Tacitus’ account (Annales
-XI, 23-24)\footnote{ On which see Jahn 1993.}: reading the original document is a privileged occasion to determine how
+XI, 23-24)\footnote{ On which see \citet{Jahn1993}.}: reading the original document is a privileged occasion to determine how
 reliable the historian is when using his sources.
@@
 Highly original suggestions have been provided by Mauro Reali, who is also the author of different school manuals, in a
-paper published online\footnote{http://mediaclassica.loescher.it/nuove-e-\%93vecchie\%94-forme-di-multimedialita.n2799
-; see also Reali, Turazza 2015.}. Being an expert on the subject, he offers a comparison between the “noble” form of
+paper published online\footnote{\url{http://mediaclassica.loescher.it/nuove-e-\%93vecchie\%94-forme-di-multimedialita.n2799}
+; see also \citep{Reali2015}.}. Being an expert on the subject, he offers a comparison between the ``noble'' form of
 the political-philosophical amicitia presented in Cicero’s Laelius and the term \emph{amicus} mentioned in inscriptions
 from the lower levels of society, such as CIL 05, 05300 from \emph{Comum} (a funerary stela made by a freedman for
-Pliny the Younger) or CIL 05, 05923 = EDR124245 regarding a strange case of “friend deletion” long before the Facebook
-era\footnote{ http://www.laricerca.loescher.it/lingue-classiche/327-un-amico-o-amicus-e-per-sempre.html.}.
+Pliny the Younger) or CIL 05, 05923 = EDR124245 regarding a strange case of ``friend deletion'' long before the Facebook
+era\footnote{ \url{http://www.laricerca.loescher.it/lingue-classiche/327-un-amico-o-amicus-e-per-sempre.html}.}.
 
@@
 another cippus for 40 friends (amicis meis, i.e. freedmen probably belonging to the same association):
+\begin{quotation}
 
@@
 
+\end{quotation}
 As for the Augustan age, Reali suggests showing some monumental inscriptions of the princeps (he impressively goes so
@@
 mentioned among the Umbrian magistrates of the late II century B.C., this gens stands out in the town for its social
-influence and wealth\footnote{ Forni 1986; Zuddas 2006.}. It is always very exciting to combine the information on
+influence and wealth\footnote{\citet{Forni1986}; \citet{Zuddas2006}.}. It is always very exciting to combine the information on
 Passennus Paullus Propertius Blaesus given by CIL 11, 05405 = EDR025355 and that contained in two Letters by Pliny (VI,
@@
 
+\begin{quotation}
 [{}- - - I]o[323?]vino consulibb(us) (ante diem) VIII Kal(endas) Martias domum oscilavi Musae. 
+\end{quotation}
 
 This house was still visited in the fourth century A.D., which really supports the theory that it used to be the poet’s
-residence, and remained an object of reverence for centuries\footnote{ Boldrighini 2014, 244-246.}. Through the stories
+residence, and remained an object of reverence for centuries\footnote{\citet[244-246]{Boldrighini2014}.}. Through the stories
 of Passennus Paullus and Sextus Propertius macro-history and local history meet to make the past come alive in every
@@
 
+%\nocite{} non necessario, tutte le voci in bibliografia sono citate
 \bibliographystyle{sapauth-eng}

### 2015-12-16 source/18/EAGLE2016_Full_Paper_Broux.tex
started work on tex of papers, added folders for CNR papers, changed main stracture to include keynotes and AHE abstracts

@@
 %%%%%%%%
-\title{}
-\author{Yanne Broux}
+\title{Towards a Universal Facebook of the Ancient World}
+\author[KU]{Yanne Broux\corref{first}}
+\address[KU]{KU Leuven / Research Foundation Flanders (FWO)}
+\cortext[first]{Corresponding author. Email: Email: yanne.broux@kuleuven.be}
 \date{2015-09-16}
 \begin{document}
-[Warning: Draw object ignored]Abstract
 
+\maketitle
+\begin{abstract}
 Facebooking the past. The idea grew while developing a database of all people mentioned in texts from Greco-Roman Egypt
-(350 BC–AD 800). Thanks to Trismegistos’ role in EAGLE, Named Entity Recognition can now be applied to almost 500,000
+(350 BC–AD 800). Thanks to Trismegistos' role in EAGLE, Named Entity Recognition can now be applied to almost 500,000
 Latin inscriptions from the Roman Empire, and some 400,000 clusters containing personal names can be extracted. This
@@
 changes in society at large.
+\end{abstract}
+\keywords{Latin epigraphy, Trismegistos, Named Entity Recognition, Social Network Analysis, Onomastics, Prosopography}
 
-\bigskip
-
-Keywords
-
-Latin epigraphy, Trismegistos, Named Entity Recognition, Social Network Analysis, Onomastics, Prosopography
-
-\bigskip
-
-1.1. Trismegistos: the early years
+\section{Trismegistos: the early years}
 
@@
 full text in repositories, Named Entity Recognition procedures are used to create lists of toponyms and anthroponyms
-that occur in the ancient sources. This was first applied during the socio-onomastic project ‘Creating Identities in
-Graeco-Roman Egypt’ (KU Leuven OT project 2008-2012), on a corpus of about 50,000 papyri and ostraca in the Duke
-Databank of Documentary Papyri (papyri.info). With the additional support of a Hercules Grant, (‘An Interdisciplinary
-Database of Proper Names in Late Pharaonic, Graeco-Roman and Byzantine Egypt (ca. 800 BC – AD 640)’; 2008-2014) the
+that occur in the ancient sources. This was first applied during the socio-onomastic project `Creating Identities in
+Graeco-Roman Egypt' (KU Leuven OT project 2008-2012), on a corpus of about 50,000 papyri and ostraca in the Duke
+Databank of Documentary Papyri (papyri.info). With the additional support of a Hercules Grant, (`An Interdisciplinary
+Database of Proper Names in Late Pharaonic, Graeco-Roman and Byzantine Egypt (ca. 800 BC – AD 640)'; 2008-2014) the
 work on the core data could be finished in just over two years, resulting in a database with almost half a million
@@
 Greco-Roman Egypt have been published. Two PhDs focused on the longstanding tradition of double names: the first dealt
-with the Ptolemaic period, when this form of polyonymy served to cross ethnic borders [Coussement forthcoming], the
+with the Ptolemaic period, when this form of polyonymy served to cross ethnic borders \citep{Coussement}, the
 other with the Roman period, when the practice was adapted by the local elite to distinguish themselves from the hoi
-polloi and to resemble Roman nomenclature [Broux 2015a]. The data of the fourth century AD provided new insights into
-the spread of Christianity in Egypt [Depauw and Clarysse 2013]. The development of a standardized identification
-cluster, consisting of a person’s name, patronymic and metronymic [Broux and Depauw 2015a], as well as the use of fixed
-expressions to denote illegitimacy [Broux 2015b], name change [Broux 2013a] and official identification [Broux,
-Coussement and Depauw 2010], were all related to the legal reorganization of the population under the Romans and the
-ensuing tax reform [Broux 2013b]. Most recent studies focus on the influence of Roman socio-linguistic practices on
-Greek and Egyptian conventions [Depauw forthcoming] and how network analysis can provide us with new insights regarding
-onomastic habits and what they say about cultural identity and social status [Broux 2015c].
+polloi and to resemble Roman nomenclature \citep{Broux2015a}. The data of the fourth century AD provided new insights into
+the spread of Christianity in Egypt \citep{Depauw2013}. The development of a standardized identification
+cluster, consisting of a person's name, patronymic and metronymic \citep{BrouxDepauw}, as well as the use of fixed
+expressions to denote illegitimacy \citep{Broux2015b}, name change \citep{Broux2013a} and official identification \citep{Broux2010}, were all related to the legal reorganization of the population under the Romans and the
+ensuing tax reform \citep{Broux2013b}. Most recent studies focus on the influence of Roman socio-linguistic practices on
+Greek and Egyptian conventions \citep{Depauw} and how network analysis can provide us with new insights regarding
+onomastic habits and what they say about cultural identity and social status \citep{Broux2015c}.
 
-1.2. Going global: an encompassing source guide for the ancient world
+\section{Going global: an encompassing source guide for the ancient world}
 
-1.2.1. Expanding Trismegistos Texts
+\subsection{Expanding Trismegistos Texts}
 
 Like Facebook, however, Trismegistos wants to grow, and get the entire ancient world on board. To achieve this goal,
-Trismegistos’ core database, the text database, must first be expanded by broadening its chronological and geographic
+Trismegistos' core database, the text database, must first be expanded by broadening its chronological and geographic
 horizon. Since 2013 the team has been actively working toward the inclusion of all texts from antiquity in
 Trismegistos. This implies including Latin and Greek inscriptions, an estimated 700,000 texts. Contacts with the Latin
-epigraphic database in Heidelberg resulted in Trismegistos’ participation in the Europeana EAGLE project, coordinating
+epigraphic database in Heidelberg resulted in Trismegistos' participation in the Europeana EAGLE project, coordinating
 the disambiguation across all partners. This added about 150,000 new texts to the Trismegistos database. The remaining
-300,000 will be integrated from another source (Clauss-Slaby: www.manfredclauss.de), so that the coverage of Latin will
+300,000 will be integrated from another source (Clauss-Slaby: \url{www.manfredclauss.de}), so that the coverage of Latin will
 soon be exhaustive. For Greek, good contacts have been established with the main players in the field (PHI:
-epigraphy.packhum.org/inscriptions/; DC3 [i.e. Duke, SEG, Claros]: blogs.library.duke.edu/dcthree/), and the aim is to
+\url{epigraphy.packhum.org/inscriptions/}; DC3 [i.e. Duke, SEG, Claros]: \url{blogs.library.duke.edu/dcthree/}), and the aim is to
 become partners in a project dealing with the 250,000 or so inscriptions during the next year. At the same time
 Trismegistos is also working towards a complete coverage of indigenous languages and scripts, which are often separate
-fields, isolated from the ‘classical’ world. New partnerships for Etruscan have been set up, and soon also for Punic
-(CIP: cip.cchs.csic.es) and South-Arabian (DASI: dasi.humnet.unipi.it), while for the Italic languages, for Gallic,
+fields, isolated from the `classical' world. New partnerships for Etruscan have been set up, and soon also for Punic
+(CIP: \url{cip.cchs.csic.es}) and South-Arabian (DASI: \url{dasi.humnet.unipi.it}), while for the Italic languages, for Gallic,
 Lepontic, Venetic, and Messapian information has already been digitized on the basis of existing corpora.
 
-\bigskip
-
-1.2.2. Expanding Trismegistos People
+\subsection{Expanding Trismegistos People}
 
@@
 
-\bigskip
-
-1.2.2.1. Named Entity Recognition
+\subsubsection{Named Entity Recognition}
 
@@
 texts, but quickly spread to other fields, such as biology and genetics, and is now gaining momentum in the Digital
-Humanities [van Hooland et al. forthcoming]. The problem with NER-systems, however, is that techniques designed for one
+Humanities \citep{vanHooland}. The problem with NER-systems, however, is that techniques designed for one
 genre or field do not necessarily work for others, due to specific text properties (some follow strict writing
@@
 connected to a set of variants in all possible languages. As a rule, only very minor dialectal or orthographical
-variation is allowed in the ‘native’ language, in this case Greek (e.g. \textgreek{E>is'idwroc }and
+variation is allowed in the `native' language, in this case Greek (e.g. \textgreek{E>is'idwroc }and
 \textgreek{>Is'idoroc). }Many of the variants are actually created by renderings of a name in other languages, e.g.
@@
 40,000 individuals with titles listed in the Prosopographia Ptolemaica (KU Leuven) and was supplemented with new names
-collected during the ‘Creating Identities’ project [Depauw and Van Beek 2009]. Through cooperation with the Lexicon of
+collected during the `Creating Identities' project \citep{Depauw2009}. Through cooperation with the Lexicon of
 Greek Personal Names (Oxford University) the names of some 300,000 individuals can now also be integrated.
@@
 Apart from this gazetteer, rules were developed to cope with the combination of names, or more correctly declined name
-variants, as well as with the combination of names with non-onomastic introductory terms (e.g. ‘son of’) in the
+variants, as well as with the combination of names with non-onomastic introductory terms (e.g. `son of') in the
 identification of individuals. For Greek texts, this is pretty straightforward, as people were identified by their name
@@
 cognomen; sometimes even the papponymic and the voting tribe, e.g. Marcus Tullius Marci filius Marci nepos Cornelia
-Cicero (‘Marcus Tullius Cicero, son of Marcus, grandson of Marcus, [of the] Cornelia [tribe]’). A completely new set of
+Cicero (`Marcus Tullius Cicero, son of Marcus, grandson of Marcus, [of the] Cornelia [tribe]'). A completely new set of
 rules is therefore being developed to apply to the Latin inscriptions.
 
-\bigskip
-
-1.2.2.2. Human quality control and prosopographical identifications
+\subsubsection{Human quality control and prosopographical identifications}
 
@@
 
-All this could be labeled ‘quality control’, but we will also rely exclusively on humans for the logical next step when
+All this could be labeled `quality control', but we will also rely exclusively on humans for the logical next step when
 developing a prosopography: the identification of namesakes as attestations of the same person. Since the systematic
@@
 
-\bigskip
-
-1.2.2.3. Social Network Analysis
+\subsubsection{Social Network Analysis}
 
@@
 structural forms of relations between individuals. It has huge potential for future historical research, not only by
-applying ‘traditional’ network analysis for the study of social interaction, but also by developing new,
+applying `traditional' network analysis for the study of social interaction, but also by developing new,
 non-conventional techniques.
 
-In the wake of the automatic extraction of individuals from Greek papyri during the ‘Creating Identities’ project,
+In the wake of the automatic extraction of individuals from Greek papyri during the `Creating Identities' project,
 people appearing in more than one text could not be identified and were therefore entered in Trismegistos People under
@@
 and Vanbeselaere therefore developed a new procedure to identify individuals with the help of graph visualization and
-SNA [Broux and Vanbeselaere forthcoming]. By linking people based on co-occurrence in the same texts, an overview of
-all the data is presented in a single network, where one can “zoom in” on specific individuals and compare their
-“surroundings”. When specific clusters of names reappear, these are likely the same individuals mentioned in different
-texts (see Fig. 1: each circle [or ‘node’] represents an individual; some of these need to be merged, since they
+SNA \citep{BrouxVanbeselaere}. By linking people based on co-occurrence in the same texts, an overview of
+all the data is presented in a single network, where one can ``zoom in'' on specific individuals and compare their
+``surroundings''. When specific clusters of names reappear, these are likely the same individuals mentioned in different
+texts (see Fig.~\ref{fig:1}: each circle [or `node'] represents an individual; some of these need to be merged, since they
 actually refer to the same person).
 
-\bigskip
-
- \includegraphics[width=9.984cm,height=9.454cm]{EAGLE2016FullPaperBroux-img001.jpg}
-
-Fig. 1.1. Visualizing individuals co-occurring in texts.
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperBroux-img001.jpg}
+\caption{Visualizing individuals co-occurring in texts. }
+\label{fig:1}
+\end{figure}
 
@@
 genealogical relations (since names are the result of conscious choices made by parents), the co-occurrence of names in
-communities can be mapped, which opens up new possibilities for quantification and interpretation [Broux 2015c]. A
-name’s popularity can be calculated by means of its in-degree (i.e. how many other names point to it), while the
+communities can be mapped, which opens up new possibilities for quantification and interpretation \citep{Broux2015c}. A
+name's popularity can be calculated by means of its in-degree (i.e. how many other names point to it), while the
 density of the network, the number of reciprocal links and the weight of these links can tell us something about the
@@
 
-1.3. Goals
-
-1.3.1. Study of naming practices
+\section{Goals}
+\subsection{Study of naming practices}
 
@@
 
-\bigskip
-
-1.3.2. Towards a Facebook of the ancient world
+\subsection{Towards a Facebook of the ancient world}
 
@@
 how these communities responded to impositions from above (bottom-up approach). Social models, such as the six degrees
-of separation theory, can be tested, to check whether our ‘small world’ perception is indeed the result of present-day
+of separation theory, can be tested, to check whether our `small world' perception is indeed the result of present-day
 technology and mass-communication, or if similar structures of interconnectivity existed, and, if so, what the

### 2015-12-16 source/8/EAGLE_2016_submission_8_revised.tex
started work on tex of papers, added folders for CNR papers, changed main stracture to include keynotes and AHE abstracts

@@
 %all title info missing!
-\title{}
-\author{Diederichs}
-\date{2015-11-17}
+\title{A VIRTUAL RESEARCH ENVIRONMENT TO DOCUMENT AND ANALYZE NON-ALPHABETIC WRITING SYSTEMS. A Case Study for Maya Writing.}
+\author[bonn]{Katja Diederichs\corref{first}}
+\author[bonn]{Sven Gronemeyer}
+\author[bonn]{Christian Prager}
+\author[bonn]{Elisabeth Wagner}
+\author[bonn]{Franziska Diehr}
+\author[bonn]{Brodhun Maximilian}
+\author[bonn]{Nikolai Grube}
+
+\address[bonn]{missing}
+
+\cortext[bonn]{missing}
+
 \begin{document}
-\clearpage\setcounter{page}{1}\section[]{[Warning: Draw object ignored]}
-\section{}
-Abstract
 
-\bigskip
+\maketitle
 
-No existing digital work environment can sufficiently represent the traditional epigraphic workflow ‘documentation,
-analysis, interpretation, publication’ for a non-alphabetic writing system. Using the Maya hieroglyphic script, this
+\begin{abstract}
+No existing digital work environment can sufficiently represent the traditional epigraphic workflow `documentation,
+analysis, interpretation, publication' for a non-alphabetic writing system. Using the Maya hieroglyphic script, this
 workflow will be transferred to a digital epigraphy. Digital methods and tools will be developed and reused in a
@@
 and object history and references.
+\end{abstract}
+\keywords{Classic Mayan, Digital Epigraphy, TEI, EpiDoc, CIDOC-CRM, Metadata Schema, Open Science}
 
-\section[Keywords]{Keywords}
-Classic Mayan, Digital Epigraphy, TEI, EpiDoc,
+\section{Digitalizing the Epigraphic Workflow}
 
-CIDOC-CRM, Metadata Schema, Open Science
-
-\bigskip
-
-\bigskip
-
-1.1. Digitalizing the Epigraphic Workflow
-
-In order to develop a digital work environment that represents the traditional epigraphic workflow for documenting and
+\noindent In order to develop a digital work environment that represents the traditional epigraphic workflow for documenting and
 studying inscriptions in a non-alphabetic script, the individual steps of this workflow need to be technically
@@
 transliteration and transcription of the texts; continues with morphological segmentation and linguistic
-interpretation; and concludes with translation and digital publication of the inscription [Diederichs 2015].
+interpretation; and concludes with translation and digital publication of the inscription \citep{Diederichs2015}.
 
@@
 drawings, as well as annotation of the objects with metadata. TextGrid offers input forms, annotation tools, and a
-Text-Image-Link-Editor, and provides long-term storage, and free access to the data in an open web repository [Neuroth
-\& Lohmeier 2011].
+Text-Image-Link-Editor, and provides long-term storage, and free access to the data in an open web repository \citep{NeurothLohmeier2011}.
 
@@
 texts, can be technically carried out within a VRE such as TextGrid, they still necessitate creation and adaptation of
-their own XML-based metadata model, as well as annotation schemas for object and textual contexts [Prager 2015]. Using
+their own XML-based metadata model, as well as annotation schemas for object and textual contexts \citep{Prager2015}. Using
 this approach, the Maya hieroglyphic script will for the first time be able to fulfill a central requirement of corpus
-linguistics, namely machine readability [McEnery \& Wilson 2001].
+linguistics, namely machine readability \citep{McEneryWilson2001}.
 
-Figure 1.1. provides a schematic illustration of the following points that represents in great detail the steps in
+Fig.~\ref{fig:1} provides a schematic illustration of the following points that represents in great detail the steps in
 epigraphically analyzing Maya inscriptions, a process which can also be used in modified form to study other
@@
 researching texts composed in other hieroglyphic, cuneiform, or linear writing systems.
-
-\bigskip
 
@@
 \item Identification and isolation of signs in one block 
-\item Classification of signs based on the sign catalog of Eric Thompson [Thompson 1962]
-\item Classification of signs in one block [Thompson 1962] 
+\item Classification of signs based on the sign catalog of Eric Thompson \citep{Thompson1962}
+\item Classification of signs in one block \citep{Thompson1962} 
 \item Transliteration of individual signs 
@@
 \item Literal translation 
-\item Rough translation \ 
+\item Rough translation
 \end{enumerate}
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016submission8revisedx-img001.png}
+\caption{Visualization of the important steps in the workflow of epigraphic analysis, drawings by \citet{Prager2015}}
+\label{fig:1}
+\end{figure}
 
- \includegraphics[width=10.881cm,height=7.214cm]{EAGLE2016submission8revisedx-img001.png}
+\subsection[Conceptual Questions]{Conceptual Questions}
+When developing and technically implementing the digital epigraphic research environment, the major focus is on pursuing
+the following objectives: 
+\begin{enumerate}
+\item creation of a markup schema to represent these steps (which in practice are interconnected
+and may also occur parallel to each other in the form of alternative interpretations); 
+\item defining of the precise requirements that a data structure must fulfill in order to sufficiently represent complex
+i.e. non-alphabetic textual data for epigraphic analysis and research. 
+\end{enumerate}
 
-\bigskip
-
-Fig. 1.1. Visualization of the important steps in the workflow of epigraphic analysis, drawings by Christian Prager,
-2015
-
-\bigskip
-
-\section[1.1.1. Conceptual Questions]{1.1.1. Conceptual Questions}
-When developing and technically implementing the digital epigraphic research environment, the major focus is on pursuing
-the following objectives: 1) creation of a markup schema to represent these steps (which in practice are interconnected
-and may also occur parallel to each other in the form of alternative interpretations);
-
-2) defining of the precise requirements that a data structure must fulfill in order to sufficiently represent complex
-i.e. non-alphabetic textual data for epigraphic analysis and research.
-
-\section[2.1. Prerequisites for Making a Non{}-Alphabetic Writing System Digitally Accessible]{2.1. Prerequisites for
+\section[Prerequisites for Making a Non{}-Alphabetic Writing System Digitally Accessible]{Prerequisites for
 Making a Non-Alphabetic Writing System Digitally Accessible}
-In order to make syllabic or logo-syllabic writing systems accessible for corpus linguistics, for instance to carry out
+\noindent In order to make syllabic or logo-syllabic writing systems accessible for corpus linguistics, for instance to carry out
 a corpus analysis as part of the basic lexicography required to compile a dictionary of Classic Mayan, the phonology,
@@
 
-\section[2.1.1. Registering Object Data ]{2.1.1. Registering Object Data }
-When designing the digital epigraphic work environment, one must also account for the current state of decipherment of
+\subsection[Registering Object Data ]{Registering Object Data}
+\noindent When designing the digital epigraphic work environment, one must also account for the current state of decipherment of
 the script and language that are to be documented. Thus, Classic Mayan presents great challenges, as at least one third
@@
 substitution. From a structuralist perspective, this method investigates the paradigmatic relation between two
-linguistic entities as indicated by whether or not they appear in the same context [Bußmann 2002]. This step yields
+linguistic entities as indicated by whether or not they appear in the same context \citep{Bußmann2002}. This step yields
 results when sufficient quantities of data can be digitally compiled, marked up and investigated using corpus
@@
 decipherment that resulted from studying the relationship between object and textual data. Scholars suspected that the
-word “axe” occurred in repeating, brief sign sequences on inscribed bronze axes. Building on this hypothesis, they
+word ``axe'' occurred in repeating, brief sign sequences on inscribed bronze axes. Building on this hypothesis, they
 determined the phonetic values of individual signs using data on the Canaantic language and completely deciphered the
-script [Day 2002]. Such investigations enable initial association of signs with both content and function in the
+script \citep{Day2002}. Such investigations enable initial association of signs with both content and function in the
 relevant language so that the signs may later be grammatically, morphologically, semantically, etc., defined and
 eventually deciphered. This work also facilitates more precise typological classification of a writing system
-[Gronemeyer 2015].
+\citep{Gronemeyer2015}.
 
@@
 understandability of CIDOC-CRM, the scheme also requires to reuse data and to connect them with data from other
-research projects in the spirit of Linked Open Data [Diederichs 2015 and Prager 2014].
+research projects in the spirit of Linked Open Data \citep{Diederichs2015, Prager2015}. %corrected from Prager 2014, not in biblio
 
-\section[2.1.2. Making Linguistic Data Accessible]{2.1.2. Making Linguistic Data Accessible}
-An interdisciplinary and widely accepted standard that takes into account both generalized and special characteristics
+\subsection[Making Linguistic Data Accessible]{Making Linguistic Data Accessible}
+\noindent An interdisciplinary and widely accepted standard that takes into account both generalized and special characteristics
 of writing systems beyond those of a single script must be developed in order to be able to satisfactorily annotate
@@
 cuneiform systems, or Hieroglyphic Luwian, arrange their graphemes principally in groups or blocks, and only
-secondarily in rows or columns, with a high rate of metathesis [Lacau 1903] in order to optimize the use of space.
+secondarily in rows or columns, with a high rate of metathesis \citep{Lacau1903} in order to optimize the use of space.
 Furthermore, many of these writing systems demonstrate a high degree of allography, which is no longer apparent in a
@@
 linguistic analysis must be taken into account, and the creative process should also promote discussion of
-terminological and typographic conventions for these annotations [Sachse \& Dürr 2015]. The goal is to create a generic
+terminological and typographic conventions for these annotations \citep{SachseDürr2015}. The goal is to create a generic
 markup model that can be implemented regardless of the particular script or language in question. The inscriptions of
-the Maya hieroglyphic writing system alone feature attestations of various vernacular languages [Lacadena \& Wichmann
-2002], a phenomenon which raises several preliminary questions: what exactly are the demands, and what are the goals of
+the Maya hieroglyphic writing system alone feature attestations of various vernacular languages \citep{LacadenaWichmann2002}, a phenomenon which raises several preliminary questions: what exactly are the demands, and what are the goals of
 an epigraphic and linguistic annotation? Where do possibilities and limitations exist for annotating these writing
@@
 
-\bigskip
-
-\section[2.1.2.1. Representing the Primary Text Source]{2.1.2.1. Representing the Primary Text Source}
+\subsubsection[Representing the Primary Text Source]{Representing the Primary Text Source}
 When defining a text, acknowledging an interpretation of the source text itself is unavoidable. The meaning of the
-“original text” that resides in the “source text” is an important point of discussion in epigraphy. Thus, the
+``original text'' that resides in the ``source text'' is an important point of discussion in epigraphy. Thus, the
 annotation does not contain a guiding text in the conventional sense, because it is merely reconstructed using all
@@
 data (in the case of hieroglyphic texts, photographs of the text carriers, for example) and secondary data (e.g.
-drawings or interpretive annotations) [Stührenberg 2012]. This strategy can be implemented using stand-off annotations
+drawings or interpretive annotations) \citep{Stührenberg2012}. This strategy can be implemented using stand-off annotations
 for data. Similarly, such an annotation should permit multiple descriptions, i.e. alternative statements concerning the
-data [Stührenberg 2012]. For example, sign sequences in Classic Mayan can be variously analyzed depending in a
-particular vernacular context [Gronemeyer 2014]. Only in this manner can various ways of thinking be appropriately
+data \citep{Stührenberg2012}. For example, sign sequences in Classic Mayan can be variously analyzed depending in a
+particular vernacular context \citep{Gronemeyer2014}. Only in this manner can various ways of thinking be appropriately
 recorded and relayed to the scientific community for discussion.
@@
 in Maya hieroglyphic texts are not arranged according to their literal reading order, but instead in spatially
-distinct, square or rectangular units (so-called “blocks”), each of which in most cases corresponds to a word or
+distinct, square or rectangular units (so-called ``blocks''), each of which in most cases corresponds to a word or
 morpheme sequence. A detailed markup of the original text is therefore of great importance, particularly for partially
@@
 significant desideratum in epigraphic research, and it also constitutes a core pillar of computational linguistics
-[McEnery \& Wilson 2001].
+\citep{McEneryWilson2001}.
 
@@
 
-\bigskip
-
-\section[2.1.2.2. Graphetics ]{2.1.2.2. Graphetics }
-Graphetics concerns the formal structure of linguistic units and the structure of texts [Crystal 1997]. Classic Maya
+\subsubsection[Graphetics ]{Graphetics }
+Graphetics concerns the formal structure of linguistic units and the structure of texts \citep{Crystal1997}. Classic Maya
 hieroglyphic inscriptions, for instance, may display a very high degree of variation in writing styles for arranging
 texts, a phenomenon which is deserving of investigation. In this manner, various systems of notation are studied in
-their individual, social, and typographic aspects, for instance [Bußmann 2002]. Similarly, in paleography, script
-decipherment is examined from graphetic perspective [Bußmann 2002]. The arrangement of texts, for instance, different
+their individual, social, and typographic aspects, for instance \citep{Bußmann2002}. Similarly, in paleography, script
+decipherment is examined from graphetic perspective \citep{Bußmann2002}. The arrangement of texts, for instance, different
 grapheme or block sizes of text fields, or different styles of hieroglyphic writing (equivalent to alphabetic font
@@
 
-\bigskip
-
-2.1.2.3. Graphemics
-
+\subsubsection{Graphemics }
 When analyzing the contents of these texts, it is important to investigate meaningful variation underlying the manner in
-which the writing is arranged and to thereby identify distinct units [Bußmann 2002]. Graphemics is devoted to this area
+which the writing is arranged and to thereby identify distinct units \citep{Bußmann2002}. Graphemics is devoted to this area
 of research, and consequently to exploring the meaningfully distinct features of signs.
 
-\bigskip
-
-\section[2.1.2.3.1. Allography of Graphemes ]{2.1.2.3.1. Allography of Graphemes }
+\subsubsection[1. Allography of Graphemes ]{1. Allography of Graphemes }
+% THESE ARE SUBSUB*SUB*SECTIONS...
 A very important example of a subject of graphemic research is provided by allography, a phenomenon which describes a
-1:n relationship between a grapheme and its various graph representations [Crystal 1997].
+1:n relationship between a grapheme and its various graph representations \citep{Crystal1997}.
 
@@
 For example, in Maya writing, there are over twenty different allographs for the vowel sign {\textless}u{\textgreater},
-which is used e.g. as the 3rd person ergative pronoun u- “he, she, it”. Additional research questions may comprise 1)
+which is used e.g. as the 3rd person ergative pronoun u- ``he, she, it''. Additional research questions may comprise 1)
 the reading order of the signs in context; 2) the existence of word separators or other graphic aids used to
@@
 or vice versa.
+%what to do with this lists? another below.
 
-\bigskip
-
-\section[2.1.2.3.2. Graphotactic Strategies of a Writing System ]{2.1.2.3.2. Graphotactic Strategies of a Writing System
+\subsubsection[2. Graphotactic Strategies of a Writing System ]{2. Graphotactic Strategies of a Writing System
 }
@@
 graphotactically link individual signs in a hieroglyphic block in Maya writing, various representational rules may be
-used (see Fig. 1.2.), e.g. affixation, infixation, conflation, or superimposition [Zender 1999]. Additional typical,
+used (see Fig.~\ref{fig:2}), e.g. affixation, infixation, conflation, or superimposition \citep{Zender1999}. Additional typical,
 functional traits of logo-syllabic writing systems include underspellings, diacritics, phonemic complements and
@@
 
-\begin{figure}
+\begin{figure}[!bp]
 \centering
-\includegraphics[width=10.816cm,height=7.054cm]{EAGLE2016submission8revisedx-img002.png}
+ \includegraphics[width=\columnwidth]{EAGLE2016submission8revisedx-img002.png}
+\caption[Some possible sign combinations in Classical Mayan hieroglyphs after \citet{Zender1999}, drawings by
+\citet{Prager2015}]{Fig. 1.2. Some possible sign combinations in Classical Mayan hieroglyphs after \citet{Zender1999}, drawings by \citet{Prager2015}}
+\label{fig:2}
 \end{figure}
-\section{}
-\section[Fig. 1.2. Some possible sign combinations in Classical Mayan hieroglyphs [after Zender 1999{]}, drawings by C.
-Prager 2015 ]{Fig. 1.2. Some possible sign combinations in Classical Mayan hieroglyphs [after Zender 1999], drawings by
-C. Prager 2015 }
 
-\bigskip
-
-\section[2.1.2.4. Signary: Sign Classification and Sign Catalogs ]{2.1.2.4. Signary: Sign Classification and Sign
-Catalogs }
-The graphemic lexicon provides a central reference for marking up texts composed in non-alphabetic writing systems. In
+\subsubsection[Signary: Sign Classification and Sign Catalogs ]{Signary: Sign Classification and Sign Catalogs }
+\noindent The graphemic lexicon provides a central reference for marking up texts composed in non-alphabetic writing systems. In
 the case of the Maya script, signs whose phonemic reading is unknown or unconfirmed are denoted using reference systems
-from various classification catalogues (e.g. Thompson 1962). Use of question marks for indicating unclear readings, for
+from various classification catalogues \citep[e.g.]{Thompson1962}. Use of question marks for indicating unclear readings, for
 example, should be avoided, because such signs impair machine readability and sometimes represent control characters or
@@
 
-\clearpage\section[2.1.2.5. The State of Decipherment and the Readability of Texts ]{2.1.2.5. The State of Decipherment
+\subsubsection[The State of Decipherment and the Readability of Texts ]{The State of Decipherment
 and the Readability of Texts }
-In the case of unreadable or only partially deciphered writing systems, the question in the forefront is the state of
+\noindent In the case of unreadable or only partially deciphered writing systems, the question in the forefront is the state of
 decipherment of the relevant writing system and the handling of undeciphered signs or text passages. As archaeological
@@
 
-\bigskip
+\section{Open Science Strategy}
 
-3.1. Open Science Strategy
-
-A Digital Humanities project working in the spirit of Open Science fundamentally intends the work environment that it
+\noindent A Digital Humanities project working in the spirit of Open Science fundamentally intends the work environment that it
 develops to be reused. As such, a VRE for non-alphabetic writing system such as that of the Classic Maya should be
@@
 semantic interoperability by using common, given, XML-based annotation and metadata standards, like TEI and EpiDoc, and
-controlled vocabularies, such as SKOS, that are already established in the Digital Humanities [Diederichs 2015].
+controlled vocabularies, such as SKOS, that are already established in the Digital Humanities \citep{Diederichs2015}.
 
@@
 uninhibited access to the project’s research (raw data and metadata), as well as for guaranteeing the contents’
-productive reuse, must be ensured through the use of free licenses [DFG 2014].
+productive reuse, must be ensured through the use of free licenses \citep{DFG2014}.
 
 As such, all the contents of a database and data infrastructure that are being planned will be made available under
-so-called “Open” licenses (Open Access, Open Source, Open Methodology, Open Data, etc.), with the goal of pursuing a
-comprehensive Open Science strategy [Diederichs 2015].
+so-called ``Open'' licenses (Open Access, Open Source, Open Methodology, Open Data, etc.), with the goal of pursuing a
+comprehensive Open Science strategy \citep{Diederichs2015}.
 
@@
 
-\bigskip
-
-\section[4.1. Summary ]{4.1. Summary }
-The creation of digital infrastructures and metadata schemas that facilitate recording and further analysis of the
+\section[Summary ]{Summary}
+\noindent The creation of digital infrastructures and metadata schemas that facilitate recording and further analysis of the
 appropriate annotation for non-alphabetic writing systems, including their respective cultural contexts, is a
@@
 
-\bigskip
-
-\bigskip
+\nocite{Gronemeyer2014}
 

### 2015-12-16 source/9/Latin epigraphy for the visually impaired.tex
started work on tex of papers, added folders for CNR papers, changed main stracture to include keynotes and AHE abstracts

@@
 %%%%%%%%
-\title{}
-\author{Francesca Licordari}
+\title{Latin epigraphy for the visually impaired. New technologies to favour universal accessibility}
+\author[mne]{Francesca Licordari\corref{first}}
+\address[mne]{Museo Nazionale Etrusco di Villa Giulia, Roma}
+\cortext[first]{Corresponding author. Email: francesca.licordari@beniculturali.it}
 \date{2015-11-19}
 \begin{document}
-1. Latin epigraphy for the visually impaired.
-
-\ \ \ \ New technologies to favour universal accessibility
-
-\bigskip
-
-Francesca Licordari
-
-Museo Nazionale Etrusco di Villa Giulia, Roma
-
-francesca.licordari@beniculturali.it
-
-\bigskip
-
-Abstract
-
-\bigskip
-
+\maketitle
+\begin{abstract}
 The problem of accessibility to works of art for the visually impaired can be extended to a system such as Latin
 epigraphy, which presents great difficulties since the engraved letters are not clear to the touch. The technology of
-3D printing allows the creation of materials that make it easier to read epigraphic texts in following the “design for
-all” principle demonstrating an economic, rapid technique that is easy to implement and reasonably robust.
+3D printing allows the creation of materials that make it easier to read epigraphic texts in following the ``design for
+all'' principle demonstrating an economic, rapid technique that is easy to implement and reasonably robust.
+\end{abstract}
+\keywords{Accessibility, visually impaired, 3D printing, design for all, Braille, epigraphy}
 
-\bigskip
-
-Keywords:
-
-\bigskip
-
-Accessibility, visually impaired, 3D printing, design for all, Braille, epigraphy
-
-\bigskip
-
-1.1. Introduction
-
-\bigskip
-
-The Declaration of the Rights of Deaf-Blind Persons, adopted by the UN in 1979, provides that every blind or deaf person
+\section{Introduction}
+\noindent The Declaration of the Rights of Deaf-Blind Persons, adopted by the UN in 1979, provides that every blind or deaf person
 has the right to enjoy the same privileges, guaranteed to all people and to have their aspirations and abilities
 recognized and respected. These principles are reaffirmed by the UN Convention on the Rights of Persons with
-Disabilities of 2006, which states (Art. 30) that the Member States must recognize “the right of persons with
+Disabilities of 2006, which states (Art. 30) that the Member States must recognize ``the right of persons with
 disabilities to take part on an equal basis with others in cultural life and should take all appropriate measures to
@@
 
-(A) enjoy access to cultural materials in accessible formats;
+\begin{description}
+\item[A] enjoy access to cultural materials in accessible formats;
 
-(B) enjoy access to television programs, films, theater and other cultural activities, in accessible formats;
+\item[B] enjoy access to television programs, films, theater and other cultural activities, in accessible formats;
 
-(C) enjoy access to places of cultural activities, such as theaters, museums, cinemas, libraries and tourism services,
-and, as far as possible, to monuments and sites of national cultural importance.”
+\item[C] enjoy access to places of cultural activities, such as theaters, museums, cinemas, libraries and tourism services,
+and, as far as possible, to monuments and sites of national cultural importance.''
+
+\end{description}
 
@@
 The bibliography in this area of study is unfortunately not very comprehensive, especially as regards the epigraphic
-field; but advances are even now being made. In Italy, the most important studies are conducted by the “Museo Tattile
-Statale Omero” in Ancona, which, being a museum of statuary, is mainly engaged with reproductions and the accessibility
-of tactile sculptures. The same work, involving the reproduction of paintings, is being carried out by the “Museo
-Tattile Anteros” in Bologna. It is no coincidence that Italian studies in this field are the most important done at the
+field; but advances are even now being made. In Italy, the most important studies are conducted by the ``Museo Tattile
+Statale Omero'' in Ancona, which, being a museum of statuary, is mainly engaged with reproductions and the accessibility
+of tactile sculptures. The same work, involving the reproduction of paintings, is being carried out by the ``Museo
+Tattile Anteros'' in Bologna. It is no coincidence that Italian studies in this field are the most important done at the
 international level. This is mainly due to the fact of Italy’s great artistic and epigraphic heritage, which makes it
@@
 
-\bigskip
-
-\bigskip
-
-1.2. Problems in Understanding Epigraphy
-
-\bigskip
-
-The field of epigraphy, always considered highly specialized, can be of great fascination to people with visual
+\section{Problems in Understanding Epigraphy}
+\noindent The field of epigraphy, always considered highly specialized, can be of great fascination to people with visual
 impairment. Some may have studied ancient languages at school, especially Latin, and so feel the need to have practical
@@
 touching an original work wherever possible is a uniquely fascinating act, it becomes necessary to propose
-reproductions that can make the material accessible by following the concept of “design for all,” which is adaptable to
-all types of art. As rightly noted by the scholar and educator Enzo Tioli (Tioli, 2006 in L’arte a portata di mano),
-this principle can be defined as “identical, whenever possible, equivalent when that is not possible.” All that can be
-touched and seen in the original work constitutes the “identical”, the use of a copy constitutes the “equivalent”.
+reproductions that can make the material accessible by following the concept of ``design for all,'' which is adaptable to
+all types of art. As rightly noted by the scholar and educator Enzo Tioli \citep{Tioli2006},
+this principle can be defined as ``identical, whenever possible, equivalent when that is not possible.'' All that can be
+touched and seen in the original work constitutes the ``identical'', the use of a copy constitutes the ``equivalent''.
 
-\bigskip
-
-\bigskip
-
-1.3. The Techniques of Accessibility
-
-\bigskip
-
-In the world of epigraphy, casts of inscriptions are often made use of, both to facilitate their study by specialists,
+\section{The Techniques of Accessibility}
+\noindent In the world of epigraphy, casts of inscriptions are often made use of, both to facilitate their study by specialists,
 and also to replace those exhibits that had been located outside and brought indoors to protect them from weather
@@
 formed in a metal matrix. This type of technique is the one most used for the creation of teaching aids in schools for
-the blind (in this regard see www.prociechi.it). But the technique proves to be very expensive in the creation of the
+the blind (in this regard see \url{www.prociechi.it}). But the technique proves to be very expensive in the creation of the
 original matrix, as well as for the reproductions of individual pieces. The costs come down significantly when multiple
@@
 
-\bigskip
-
-\bigskip
-
-1.4. 3D Printing
-
-\bigskip
-
-The new frontier in the understanding of works of art for the sight impaired is provided by 3D printing, which proves to
+\section{3D Printing}
+\noindent The new frontier in the understanding of works of art for the sight impaired is provided by 3D printing, which proves to
 be both economical, fast, easy to implement, and durable. It is a single technique, which gathers together the
@@
 (Percro), was developed the project PURE-FORM, which led to the creation of a virtual museum, which gives visitors the
-chance to see and “touch” objects, in particular digital sculptures, belonging to different historical periods and from
+chance to see and ``touch'' objects, in particular digital sculptures, belonging to different historical periods and from
 different contexts. All of this is based on a digital scan of the artworks stored in a database, thus offering a wide
@@
 
-{}- Original
+\begin{itemize}
+\item Original
 
-{}- Original with fractures and abrasions and signs of aging
+\item Original with fractures and abrasions and signs of aging
 
-{}- Integrated original
+\item Integrated original
 
-{}- Reproduction with transliterated reliefs
+\item Reproduction with transliterated reliefs
 
-{}- Reproduction of the original with Braille overlay.
+\item Reproduction of the original with Braille overlay.
+\end{itemize}
 
@@
 
-A thesis paper, of the Postgraduate School of Archaeological Heritage in Latin Epigraphy (Licordari, 2015), which
+A thesis paper, of the Postgraduate School of Archaeological Heritage in Latin Epigraphy \citep{Licordari2015}, which
 supports an exhibition project for the dissemination of Latin Epigraphy to the public, proposes 3D reproduction of two
@@
 
-\bigskip
-
-\bigskip
-
-1.5. Conclusions
-
-\bigskip
-
-As can be seen, the studies in this field as well as the examples of works on which to experiment with seeing impaired
+\section{Conclusions}
+\noindent As can be seen, the studies in this field as well as the examples of works on which to experiment with seeing impaired
 users are as yet few in number. We hope that this essay will lead to inspire a dialogue among the relevant experts, and
@@
 
-\bigskip
+\nocite{accessibilità,Bellini2000,Cetorelli2004,Hersh2008,Jansson2003,Levi2015,Velestino2007}
 
-Bibliography:
-
-\bigskip
-
-AA.VV., L’accessibilità al patrimonio museale per le persone con minorazione visiva, CD-ROM Le dispense del Museo Omero,
-Ancona 2014
-
-AA.VV., L’arte a portata di mano. Verso una pedagogia di accesso ai Beni Culturali senza barriere, Atti del convegno
-Portonovo di Ancona 21-23 ottobre 2004, Roma 2006
-
-Bellini, A., Toccare l’Arte. L’educazione estetica di ipovedenti e non vedenti, Roma 2000
-
-Cetorelli Schivo, G., Colcerasa, M., Disabilità e museo. Percorrenze culturali di emergenza, in “Museo e territorio”,
-Atti della III giornata di studi (Velletri 7-8 marzo 2003), Roma 2004
-
-Hersh, M, Johnson, M.A. (Eds.), Assistive Technology for Visually Impaired and Blind People, Glasgow 2008
-
-Jansson, G., Bergamasco, M., Frisoli, A., A new option for the visually impaired to experience 3D art at museums: manual
-exploration of virtual copies, in “Visual Impairment Research”, 5,1, 2003, pp. 1–12
-
-Licordari, F., Tituli in artis operibus inscripti: proposta di allestimento di una mostra nei locali della Fondazione
-Sorgente, tesi di laurea della Scuola di Specializzazione in Beni Archeologici, Roma 2015
-
-Levi, F., L’accessibilità alla cultura per i disabili visivi. Storia e orientamenti, Torino 2015
-
-Velestino, D., Simili, G., Poscolieri, M. (Eds.), Messaggi dalla pietra, Torino 2007
 \bibliographystyle{sapauth-eng}

### 2015-12-16 source/20/EAGLE2016_BONACINI - Pilot project at Paolo Orsi Museum.tex
added bst files in each folder, started work on bibliography through Zotero, added CNR paper

@@
 
-\title{}
-\author{}
+\title{Digital enhancement of the ``Paolo Orsi'' museum: a Google Street View 360° pilot project tour}
+\author[catania]{Elisa Bonacini\corref{first}}
+\cortext[first]{Corresponding author. Email: e_bonacini@hotmail.com; elisa.bonacini@unict.it}
+
+\address[catania]{Humanities Department, University of Catania; IEMEST (Istituto Euro Mediterraneo di Scienze e Tecnologie), Palermo, Italy}
 \date{2015-12-11}
 \begin{document}
-Digital enhancement of the “Paolo Orsi” museum: a Google Street View 360° pilot project tour \newline
 
-Elisa Bonacini
-
-Humanities Department, University of Catania; IEMEST (Istituto Euro Mediterraneo di Scienze e Tecnologie), Palermo,
-Italy
-
-Email: e\_bonacini@hotmail.com; elisa.bonacini@unict.it
-
-\newline
-Abstract
-
-\ \ \ \ \ \ The aim of this paper is to present the pilot project in progress at the “Paolo Orsi” Archaeological Museum
+\maketitle
+\begin{abstract}
+The aim of this paper is to present the pilot project in progress at the ``Paolo Orsi'' Archaeological Museum
 (Syracuse, Italy). Thanks to a free partnership with Google Business Photos, we have managed to map the entire museum
 for a online 360° tour on Google Street View. A dozen archaeological finds have been selected for 360° virtual tours,
-provided with descriptive sheets. Among them the beautiful ‘inscription of Nassiane’, from the Catacombs of San
+provided with descriptive sheets. Among them the beautiful `inscription of Nassiane', from the Catacombs of San
 Giovanni in Syracuse, has been selected.
+\end{abstract}
+\keywords{Google, Sicily, Sicilian Cultural Heritage, Virtual Museum, Virtual tour, Digital Heritage}
 
-\bigskip
+\section{The ``Paolo Orsi'' Archaeological Museum}
 
-Keywords
-
-Google, Sicily, Sicilian Cultural Heritage, Virtual Museum, Virtual tour, Digital Heritage.
-
-\bigskip
-
-\bigskip
-
-1.1. The “Paolo Orsi” Archaeological Museum
-
-The “Paolo Orsi” Regional Archaeological Museum of Syracuse, together with the “Antonino Salinas” Regional
+The ``Paolo Orsi'' Regional Archaeological Museum of Syracuse, together with the ``Antonino Salinas'' Regional
 Archaeological Museum in Palermo, is the most important Sicilian archaeological museum and it is one of the most
@@
 
-The National Archaeological Museum of Syracuse was born by a royal decree in 1878, known as “national” for its
+The National Archaeological Museum of Syracuse was born by a royal decree in 1878, known as ``national'' for its
 collections’ importance and size. Well-placed inside the historical palace in the Cathedral Square on Ortigia island,
@@
 
-\bigskip
-
-1.2. The project: reasons and birth
+\section{The project: reasons and birth}
 
@@
 Sicilian puppet theatre; 2013: Mediterranean diet (transnational); 2014: Traditional agricultural practice of
-cultivating the ‘vite ad alberello’ (head-trained bush vines) of the community of Pantelleria.}) in Italy and in the
+cultivating the `vite ad alberello' (head-trained bush vines) of the community of Pantelleria.}) in Italy and in the
 world. Infact, Sicily has some of the largest and most important archaeological sites in the world: the temple of
@@
 Despite of this, Sicilian cultural heritage struggles to be present on Google’s platforms such as Google Art Project and
-others, as it should, compared to other Italian cultural sites (Bonacini 2013; Bonacini 2014).
+others, as it should, compared to other Italian cultural sites \citep{Bonacini2013;Bonacini2014}).
 
@@
 
-In Street View Gallery\footnote{https://www.google.com/maps/views/streetview?gl=us.}, which now has contributed a great
+In Street View Gallery\footnote{\url{https://www.google.com/maps/views/streetview?gl=us}.}, which now has contributed a great
 number of users from all over the world, thousands are the spherical and geo located photos of Italian places. However,
-tightening the selection to “Landmarks of Italy”
-\footnote{https://www.google.com/maps/views/streetview/italy-highlights?gl=us.}, Sicily has only 9 spherical photos,
+tightening the selection to ``Landmarks of Italy''
+\footnote{\url{https://www.google.com/maps/views/streetview/italy-highlights?gl=us}.}, Sicily has only 9 spherical photos,
 showing the beaches of the Aeolian Islands (7), the islands of Favignana (1) and Marettimo (1).
 
-Among Art Project’s 605 museum collections\footnote{https://www.google.com/maps/views/streetview/art-project?gl=it.}, 47
+Among Art Project’s 605 museum collections\footnote{\url{https://www.google.com/maps/views/streetview/art-project?gl=it}.}, 47
 are Italian, especially relevant to Rome, Turin, Venice and Milan. One of these, not the best artistic production, is a
@@
 Browsing on the Street View Gallery, 21 sites are all over inscribed on the World Wonders
-Project\footnote{https://www.google.com/maps/views/streetview/world-wonders-project?gl=it.}: for Italy, only Pompeii
+Project\footnote{\url{https://www.google.com/maps/views/streetview/world-wonders-project?gl=it}.}: for Italy, only Pompeii
 and the historic center of Florence are inscribed in this still restricted list. Many more sites can be visited
 virtually from the same project linked on Google Cultural Institute’s
-website\footnote{www.google.com/intl/it/culturalinstitute/worldwonders/. }: 172 sites in the world, 22 in Italy and,
+website\footnote{\url{www.google.com/intl/it/culturalinstitute/worldwonders/}. }: 172 sites in the world, 22 in Italy and,
 finally, 2 of them in Sicily. Until last year the Unesco site of the baroque town Val di Noto was the only one in
@@
 Heritage and Sicilian Identity in 2014. He proposed to realize the Street View mapping of a museum and an
-archaeological site in Sicily, considering the possibility of creating “augmented” virtual tours (3D virtual tours of
+archaeological site in Sicily, considering the possibility of creating ``augmented'' virtual tours (3D virtual tours of
 objects displayed in museum’s windows and virtual aerial tours, with the addition of text and audio descriptions).
 
-The profound reason of this project consists in an effort to bridge this gap, “increasing” fruition and enhancement of
+The profound reason of this project consists in an effort to bridge this gap, ``increasing'' fruition and enhancement of
 Sicilian cultural heritage.
@@
 and the European Coordination of Google Business Photo, it was decided to choose as a sample of this project two
-cultural regional institutions, the “Paolo Orsi” Museum in Syracuse and the Valley of the Temples in Agrigento, both
+cultural regional institutions, the ``Paolo Orsi'' Museum in Syracuse and the Valley of the Temples in Agrigento, both
 UNESCO sites since 2005 and 1997. The project was then structured as a research fellow project at the University of
@@
 
-The first part of the project at the “Paolo Orsi” Museum, which we will discuss here, is about to be completed. The
+The first part of the project at the ``Paolo Orsi'' Museum, which we will discuss here, is about to be completed. The
 second part at Valley of the Temples in Agrigento is going to start.
 
-\bigskip
-
-1.2.1. The 360° tour on Google Maps: some technical data
+\subsection{The 360° tour on Google Maps: some technical data}
 
@@
 The 360° virtual tour makes it possible browsing the entire museum and its collection, between levels connected by
-arrows, placed online at the link https://goo.gl/maps/oagnd8urP1H2 (for the first level) and
-https://goo.gl/maps/vrpDfuPPgwM2 (for the second level). As you can see in fig. 1.1., by clicking on the first link,
+arrows, placed online at the link \url{https://goo.gl/maps/oagnd8urP1H2} (for the first level) and
+\url{https://goo.gl/maps/vrpDfuPPgwM2} (for the second level). As you can see in Fig.~\ref{fig:1}, by clicking on the first link,
 the remote user can enter the museum through the Street view panel: the 360° virtual tour begins at the entrance and
@@
 The project provided the opportunity to carry out 360° virtual tours of some exposed archaeological finds, like an
-“augmented” virtual tour, certainly innovative compared to what Google’s platforms provide.
+``augmented'' virtual tour, certainly innovative compared to what Google’s platforms provide.
 
@@
 
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{m{10.872cm}}
-{\centering  \includegraphics[width=8.911cm,height=4.632cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img001.jpg}
-\par}
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img001.jpg}
+\caption{The virtual entrance to Archaeological Museum ``Paolo Orsi'' on Google Street View. }
+\label{fig:1}
+\end{figure}
 
-~
-
-Fig. 1.1. The virtual entrance to Archaeological Museum “Paolo Orsi” on Google Street View.
-
-~
-\\
-\end{supertabular}
-\end{flushleft}
 As for the displaying on Google Maps of the 360° virtual tour of the objects, we must specify that Maps so far does not
 support the integration of menus, captions, photos, video, info inside the Street View virtual tour technology.
-Infact,it \ is only allowed navigation in 360°, ie a virtual walk.
+Infact,it is only allowed navigation in 360°, ie a virtual walk.
 
@@
 Maps, through a link containing the Google mapping of the Museum. In this way the virtual tour of the Museum - made
-with Google standards - is “augmented” by another virtual walk much more exhaustive, located via link on the Google
+with Google standards - is ``augmented'' by another virtual walk much more exhaustive, located via link on the Google
 Maps board of the Museum, where you can view all these additional items.
@@
 the Regional Department for Cultural Heritage and Sicilian Identity portal
-(http://www.regione.sicilia.it/beniculturali/museopaoloorsi/). Because of the profound principle that this non-profit
+(\url{http://www.regione.sicilia.it/beniculturali/museopaoloorsi/}). Because of the profound principle that this non-profit
 project would improve the visibility and promotion of Sicilia Cultural Heritage, the Museum’s virtual tour and its
@@
 
-\bigskip
-
-1.2.2. The “Paolo Orsi” Museum 360° tour
+\subsection{The ``Paolo Orsi'' Museum 360° tour}
 
@@
 
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{m{10.681cm}}
-{\centering  \includegraphics[width=8.98cm,height=4.697cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img002.jpg}
-\par}
-
-~
-
-Fig. 1.2. 360° tour of Sector A: a couple of casts of dwarf elephants from Spinagallo cave. \\
-\end{supertabular}
-\end{flushleft}
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img002.jpg}
+\caption{360° tour of Sector A: a couple of casts of dwarf elephants from Spinagallo cave. }
+\label{fig:2}
+\end{figure}
 
@@
 
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{m{10.681cm}}
-{\centering  \includegraphics[width=8.89cm,height=4.641cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img003.jpg}
-\par}
-
-~
-
-Fig. 1.3. 360° tour of Sector B2: the findings from Athenaion and Ionic temple in Syracuse. \\
-\end{supertabular}
-\end{flushleft}
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img003.jpg}
+\caption{360° tour of Sector B2: the findings from Athenaion and Ionic temple in Syracuse.}
+\label{fig:3}
+\end{figure}
 
@@
 
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{m{10.681cm}}
-{\centering  \includegraphics[width=8.708cm,height=4.521cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img004.jpg}
-\par}
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img004.jpg}
+\caption{360° tour of Sector D: the Venus Landolina.}
+\label{fig:}
+\end{figure}
 
-~
-
-Fig. 1.4. 360° tour of Sector D: the Venus Landolina. \\
-\end{supertabular}
-\end{flushleft}
 They document the multiple aspects of life in Syracuse and come from urban necropolis from III-II century B.C. Here are
-masterpieces like the wonderful statue of the Venus Anadiomene, called Venus Landolina, here in Fig. 1.4.
+masterpieces like the wonderful statue of the Venus Anadiomene, called Venus Landolina, here in Fig.~\ref{fig:4}.
 
@@
 
-\bigskip
-
-1.2.1.1. 360° virtual tours of archaeological selected objects
+\subsubsection{360° virtual tours of archaeological selected objects}
 
@@
 
-[Warning: Draw object ignored]
+\begin{description}
+\item[From Sector A:]
+
+\begin{description}
+\item[1] A foot-cup dated to the 15th century B.C., relevant to the facies of Rodì-Tindari and coming from Vallelunga, in the province of Caltanissetta. 
+\end{description}
+
+\item[From Sector B1:]
+
+\begin{description}
+\item[2] A red-figured wedding lebes, attributed to Painter of Siracusa 47099, dated to 360-340 B.C. from Lentinoi. 
+\end{description}
+
+\item[From Sector B2:]
+
+\begin{description}
+\item[3] A proto-corinthyan oinochoe, dated to 670 B.C., coming from the excavations in Piazza Duomo; 
+\item[4] A plastic Corinthian vase in the form of a lion, dated 610-590 B.C., from the Garden Spain Necropolis; 
+\item[5] An Attic black-figured calix-crater, by the Antimene Painter in 520 B.C., coming from the Garden Spain Necropolis; 
+\item[6] An Attic black-figured Panathenaic amphora, dated to the middle of the 6th century B.C., coming from excavations in Viale Paolo Orsi; 
+\item[7] A terracotta bidder statuette, dated to the 4th century B.C., coming from the sacred votive deposit in Piazza della Vittoria. 
+\end{description}
+
+\item[From Sector C:]
+
+\begin{description}
+\item[8] A red-figured bell-shaped krater, coming from Camarina, produced in the workshop of the Athenian painter Polignoto, around 440-430 B.C., decorated on the principal side with the Delphic triad (Apollo, Artemis and Latona);
+\item[9] An Attic red-figured lekythos, coming from the necropolis of Capo Soprano near Gela, dated to 470 B.C. and realized according to the manner of the London painter E342; 
+\item[10] The Ephebus of Adrano, a small bronze athlete, dated to the first half of the 5th century B.C., generally thought to be a scaled-down copy of a large bronze original by the famous Greek sculptor Pythagoras.
+\end{description}
+
+\item[From Sectors D:]
+
+\begin{description}
+\item[11] A small Hellenistic terracotta boat in the shape of a pistrix (sea monster), from the Fusco necropolis in Syracuse.
+
+\end{description}
+
+\item[From Sector F:]
+
+\begin{description}
+\item[12] The Nassiane inscription Fig.~\ref{fig:4}, a curious marble disk, from the Catacombs of San Giovanni.
+\end{description}
+
+\end{description}
 
@@
 
-\bigskip
-
-1.2.1.3. 360° virtual tours of \ Nassiane inscription
+\subsubsection{360° virtual tours of Nassiane inscription}
 
 The last selected object we present here is a disk from Sector F: it is a marble disk, discovered by Paolo Orsi in the
-Catacombs of San Giovanni in 1894 (Orsi 1895: 509-510), decorated by a wreath of laurel leaves on one side; the other
+Catacombs of San Giovanni in 1894 \citep[509-510]{Orsi1895}, decorated by a wreath of laurel leaves on one side; the other
 side was reused in the 4th century A.D. for the funerary inscription of Nassiane, woman who died at age 32 in God
@@
 and because it documents the religious syncretism of the early centuries of Christianity. Re-using of architectural is
-well documented in the catacombs of Syracuse (Sgarlata 2013). It was found shattered in front of a burial, distinct
-from others, known as “the Saint's Tomb” (Sgarlata 2004: 40-44).
+well documented in the catacombs of Syracuse \citep{Sgarlata2013}. It was found shattered in front of a burial, distinct
+from others, known as ``the Saint's Tomb'' \citep[40-44]{Sgarlata2004}.
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img005.jpg}
+\caption{360° tour of the Nassiane inscription.}
+\label{fig:5}
+\end{figure}
 
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{m{10.681cm}}
-{\centering  \includegraphics[width=10.029cm,height=5.225cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img005.jpg}
-\par}
+This is the Greek text:
 
-~
+\begin{quotation}
 
-Fig. 1.5. 360° tour of the Nassiane inscription. \\
-\end{supertabular}
-\end{flushleft}
-\ This is the Greek text:
+\begin{otherlanguage*}{greek}
+chrismon ς χριστ<ι>ανῆς σεμνῆς ἀγανóφρονος [ἠδ]ὲ φιλάνδρου Nασσιανῆς τύμβον εἰσορᾷς, φίλε, κίμενον [ὣδε ἥ]τις σεμνοσύνῃσιν [ἐριζε]το Πηνελοπίῃ palma chrismon ἐνθάδε κῖτε [Nασσ]ιανή, ζήσασα [ἐν Θ(ε)ῷ] καλῶς καὶ ἀμέμπτως ἒτη λβ´, μῆνας ι´.
+\end{otherlanguage*}
 
-\bigskip
+\end{quotation}
 
-chrismon \textgreek{c qrist{\textless}i{\textgreater}an~hc semn~hc >aganófronoc [>hd]`e fil'androu
-}N\textgreek{assian~hc t'umbon e>isor~a|c, f'ile, k'imenon [<`wde <'h]tic semnos'unh|sin [>erize]to Phnelop'ih| }palma
-chrismon \textgreek{>enj'ade k~ite [}N\textgreek{ass]ian'h, z'hsasa [>en J(e)~w|] kal~wc ka`i >am'emptwc >`eth
-lb}\textgreek{', }\textgreek{m}\textgreek{~hnac i'.}
-
-\bigskip
-
-The inscription says: “Oh friend, you see here the tomb of Nassiane, Christian, ripe, sweet, fond of her husband, who
+The inscription says: ``Oh friend, you see here the tomb of Nassiane, Christian, ripe, sweet, fond of her husband, who
 competed for virtue with Penelope. Here lies Nassiane, she lived well and blamelessly in God (?), 32 years and 10
-months”. This inscription documents such as epigraphic formulas affected by religious syncretism (syncretism generally
+months''. This inscription documents such as epigraphic formulas affected by religious syncretism (syncretism generally
 means that a complex of phenomena and concepts derive from the meeting and fusion of different religious forms) between
@@
 refrigerium, the funeral feast that symbolically was consumed with the dead, even this syncretic practice \ dates back
-to the pagan world (Sgarlata 2013, Scandurra 2014).
+to the pagan world \citep{Sgarlata2013;Scandurra2014}.
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img006.jpg}
+\caption{ The Nassiane inscription, recto (courtesy by the ``Paolo Orsi'' Museum). }
+\label{fig:6}
+\end{figure}
 
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{m{10.681cm}}
-{\centering  \includegraphics[width=6.205cm,height=6.174cm]{EAGLE2016BONACINIPilotprojectatPaoloOrsiMuseum-img006.jpg}
-\par}
-
-~
-
-Fig. 1.6. The Nassiane inscription, recto (courtesy by the “Paolo Orsi” Museum). \\
-\end{supertabular}
-\end{flushleft}
-
-\bigskip
-
-1.3. Conclusions
+\section{Conclusions}
 
@@
 
-Thanks to this project we hope that Google itself could realize how the time has come to “rejuvenate” the Google Maps
+Thanks to this project we hope that Google itself could realize how the time has come to ``rejuvenate'' the Google Maps
 Street View system, allowing enabled users to apply additional content on the maps.
@@
 However, the wide interoperability between Google software the development of new solutions and the integration of
-geo-referenced results in the page results on the search engine continues unabated: the “Paolo Orsi” Museum - and with
+geo-referenced results in the page results on the search engine continues unabated: the ``Paolo Orsi'' Museum - and with
 the museum, the city of Syracuse and the whole of Sicily – will surely take advantage of this new tool for its
@@
 
-\bigskip
+\section*{Acknowledgement}
 
-[Warning: Draw object ignored]
-
-\bigskip
-
-Acknowledgement
-
-To Mr. Gianfranco Guccione (http://www.airworks.it/) goes my heartfelt and deep thanks for having proposed to me this
+To Mr. Gianfranco Guccione (\url{http://www.airworks.it/}) goes my heartfelt and deep thanks for having proposed to me this
 project and having concluded it together, with the only aim of enhancing the Cultural Heritage in Sicily.
@@
 
-I warmly thank all staff of the “Paolo Orsi” Regional Archaeological Museum in Syracuse for the great availability and
+I warmly thank all staff of the ``Paolo Orsi'' Regional Archaeological Museum in Syracuse for the great availability and
 friendship shown to me and Mr. Guccione: Dr. Gioconda Lamagna, executive head of the Museum, Dr. Angela Maria Manenti,
@@
 
-My deepest gratitude goes to my friend Prof. Annalisa Bonica for the patience she \ has showed in correcting this
+My deepest gratitude goes to my friend Prof. Annalisa Bonica for the patience she has showed in correcting this
 article.
@@
 This publication has been made on concession of the Regional Minister of Cultural Heritage and Sicilian Identity;
-archaeological findings and plans presented here are exclusive property of the “Paolo Orsi” Regional Archaeological
-Museum in Syracuse and it not allowed to be copied or duplicated \ by any means.
+archaeological findings and plans presented here are exclusive property of the ``Paolo Orsi'' Regional Archaeological
+Museum in Syracuse and it not allowed to be copied or duplicated by any means.
 
+\nocite{Farman}
 \bibliographystyle{sapauth-eng}

### 2015-12-16 source/8/EAGLE_2016_submission_8_revised.tex
added bst files in each folder, started work on bibliography through Zotero, added CNR paper

@@
 substitution. From a structuralist perspective, this method investigates the paradigmatic relation between two
-linguistic entities as indicated by whether or not they appear in the same context \citep{Bußmann2002}. This step yields
+linguistic entities as indicated by whether or not they appear in the same context \citep{Bussmann2002}. This step yields
 results when sufficient quantities of data can be digitally compiled, marked up and investigated using corpus
@@
 linguistic analysis must be taken into account, and the creative process should also promote discussion of
-terminological and typographic conventions for these annotations \citep{SachseDürr2015}. The goal is to create a generic
+terminological and typographic conventions for these annotations \citep{SachseDurr2015}. The goal is to create a generic
 markup model that can be implemented regardless of the particular script or language in question. The inscriptions of
@@
 data (in the case of hieroglyphic texts, photographs of the text carriers, for example) and secondary data (e.g.
-drawings or interpretive annotations) \citep{Stührenberg2012}. This strategy can be implemented using stand-off annotations
+drawings or interpretive annotations) \citep{Stuhrenberg2012}. This strategy can be implemented using stand-off annotations
 for data. Similarly, such an annotation should permit multiple descriptions, i.e. alternative statements concerning the
-data \citep{Stührenberg2012}. For example, sign sequences in Classic Mayan can be variously analyzed depending in a
+data \citep{Stuhrenberg2012}. For example, sign sequences in Classic Mayan can be variously analyzed depending in a
 particular vernacular context \citep{Gronemeyer2014}. Only in this manner can various ways of thinking be appropriately
@@
 texts, a phenomenon which is deserving of investigation. In this manner, various systems of notation are studied in
-their individual, social, and typographic aspects, for instance \citep{Bußmann2002}. Similarly, in paleography, script
-decipherment is examined from graphetic perspective \citep{Bußmann2002}. The arrangement of texts, for instance, different
+their individual, social, and typographic aspects, for instance \citep{Bussmann2002}. Similarly, in paleography, script
+decipherment is examined from graphetic perspective \citep{Bussmann2002}. The arrangement of texts, for instance, different
 grapheme or block sizes of text fields, or different styles of hieroglyphic writing (equivalent to alphabetic font
@@
 When analyzing the contents of these texts, it is important to investigate meaningful variation underlying the manner in
-which the writing is arranged and to thereby identify distinct units \citep{Bußmann2002}. Graphemics is devoted to this area
+which the writing is arranged and to thereby identify distinct units \citep{Bussmann2002}. Graphemics is devoted to this area
 of research, and consequently to exploring the meaningfully distinct features of signs.

### 2015-12-16 source/CNR/TheEAGLEDataAggregatorDataQualityMonitoring.tex
added bst files in each folder, started work on bibliography through Zotero, added CNR paper

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\usepackage{float}
\usepackage[table]{xcolor}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
%\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{The EAGLE Data Aggregator:
Data Quality Monitoring}

%%%%%%%%
%AUTHORS:%
%%%%%%%%
\author[CNR-ISTI]{Andrea Mannocci\corref{all}}
\author[CNR-ISTI]{Vittore Casarosa}
\author[CNR-ISTI]{Paolo Manghi}
\author[CNR-ISTI]{Franco Zoppi}

%Additional information about corresponding author, addresses and affiliations:
\cortext[all]{\{andrea.mannocci, vittore.casarosa, paolo.manghi, franco.zoppi\}@isti.cnr.it
}

\address[CNR-ISTI]{ISTI - Area della Ricerca CNR, via G. Moruzzi 1, 56124 Pisa, Italy}

%%%%%%%%%%%%
%PAPER CONTENT %
%start your paper here%
%%%%%%%%%%%%
\begin{document}
\rowcolors{2}{gray!25}{white}
\maketitle

\begin{abstract}
The EAGLE project aggregates epigraphy related content from about 20 different data providers, and makes its content available to both Europeana and to scholars. Data Quality monitoring is a key issue in Aggregative Data Infrastructures, where content is collected from a number of different sources with different data models and quality standards. This paper presents a Monitoring Framework for enabling the observation and monitoring of an aggregative infrastructure focusing on the description of the Data Flow and Dynamics Service, and exemplifying these concepts with a use case tailored to the characteristics of the EAGLE aggregation data flow.

An Infrastructure Quality Manager (IQM) is provided with a Web user interface (WebUI), allowing her to describe the data flows taking place in the infrastructure and to define monitoring scenarios. The scenarios will include the definition of sensors (pieces of software plugged into the data flow), which will provide observations of measured objects. The scenarios include also the definition of controls and analysers, which will store and process the observations received from the sensors and will verify if the values of the measured features comply with some expected behaviour over time.

A monitoring scenario for EAGLE has been defined and tested on simulated data (the monitoring framework is still under development) in order to monitor the ``health'' of different data collections involved in the EAGLE collection and transformation workflows.

\end{abstract}

\keywords{EAGLE, Aggregative Data Infrastructure, Data Quality, Metrics, Monitoring.}

\section{Monitoring framework}\label{framework}

The EAGLE project (fully described in \citep{eagle}) aggregates epigraphy related content from about 20 different data providers (cultural institutions all over Europe), and makes its content available to both Europeana \citep{europeana} (through an OAI-PMH interface) and to scholars and the general public through a Web portal.

Collecting data provided by a number of different sources, very often with different quality standards, presents the challenge of measuring the overall quality of the aggregated data, and how it compares with standards and objectives set by the aggregating institution.

We present here an extension to EAGLE that will take advantage of a \textit{Monitoring Framework} (being developed in the context of another project \citep{openaire}), enabling the observation and monitoring of an Aggregative Data Infrastructure over time.

By modelling ``data processing'' as a manufacturing process involving data \citep{ballou1998}, the Monitoring Framework provides tools for the automatic extraction of observations (numeric indicators about properties of the system) from the context where the data is being processed and stored, providing time series of indicators expressed in user-defined metrics.

The Monitoring Framework offers a \textit{Data Flow and Dynamics Service} (DFDS for short) to the \textit{Infrastructure Quality Manager} (IQM for short), a role that (hopefully) will become standard in data aggregation infrastructures. The DFDS enables the IQM to create one or more monitoring scenarios, each designed to monitor a particular functional area or aspect of the aggregation infrastructure. For example, an EAGLE monitoring scenario could deal with the workflow that builds the aggregated content to be used by the EAGLE Web portal and by Europeana’s OAI-PMH harvesters \citep{mannocci2014}. After the collection and processing of (possibly heterogeneous) records coming from different content providers, the aggregator stores them into different environments, for different purposes. More precisely: \textit{i)} the processed records are indexed as a full-text index (implemented by Apache Solr) to support search and browse queries from the EAGLE Web portal; \textit{ii)} the processed records are stored also in a document store (implemented by MongoDB) to support OAI-PMH requests. In this case, for example, we might be interested in assessing (and keep assessing over time) whether the total number of artifacts indexed by Solr matches the number of artifacts delivered via OAI-PMH. This could easily be accomplished by monitoring the number of records stored in the index and the number of records stored in the document store, and comparing their values. As another example, we might be interested in verifying whether the trend of a certain property of the monitored system complies with a given criterion (e.g. the number of records per content provider should be strictly increasing over time).

Using the Data Flow and Dynamics Service provided by the Monitoring Framework, it is possible to define \textit{monitoring scenarios}, which define a conceptualization of the data flows taking place in the aggregating infrastructure. Similarly to what happens to goods in a manufacturing process, data collections and processes acting over that data can be observed thanks to specially devised \textit{sensors}, which in our case are pieces of code providing numeric values about features of interest. The monitoring scenario can also define controls to verify if the values of such features comply with some expected behaviour over time.

With the concept of sensor, we refer to a piece of software capable of generating \textit{observations} (a numeric value plus some contextual metadata) about a measured object. \textit{Measured objects} can be of different granularity, e.g. single data units (datum) or data collections stored somewhere by the aggregating infrastructure. Each observation is expressed in just one specific metric, intended to measure a specific feature of a measured object (e.g. the number of publications present in the data collection stored in the Solr index). A sensor can generate observations in more than one metric, each one referring to a different feature of the measured object.

A sensor can in principle be plugged anywhere in the aggregating infrastructure; with the understanding that the implementation of the sensor and the point of the data flow where it will be placed are the responsibility of the IQM. When the workflows of the aggregating infrastructure are in execution, the sensors will be activated and will produce a stream of observations. The DFDS will separate the stream of observations into different streams, each one related to a specific metric, and will store them as points of time series. In this way, observations can be queried and examined either as charts or tabular data.

The interaction between the IQM and the Monitoring Framework is done through a user friendly user interface (WebUI) provided by the DFDS. In addition to providing easy access to the observation time series, the WebUI provides the facilities to define the overall Monitoring Scenario, which will include the sensors and the controls.

For the purpose of monitoring, the IQM can define \textit{controls}, i.e. checks that can be scheduled to automatically verify the compliance of one or more metrics (and their observations) with a desired (or un-desired) condition. A control uses an \textit{analyser} for the comparison of observation values, such as, for example, values alignment, less or greater than, (strictly) monotonic increasing or decreasing values, threshold guards, thresholded peak or percentage variation, etc. For example, as mentioned before, we might be interested to check if the number of artefacts indexed by Solr is equal to the number of artefacts exported in OAI-PMH records, or if the total number of EAGLE objects provided by a content provider is steadily increasing over time, or, as a further example, if the ``goodness'' of an EAGLE record (the concept of goodness being defined by the IQM and implemented through sensors and controls) remains above a threshold of 0.8. The Data Flow and Dynamics Service offers some analysers out-of-the-box, but also enables the IQM to develop her own custom analysers.

Finally, the Data Flow and Dynamics Service enables the generation of an exhaustive \textit{report} about the defined metrics and controls providing insights, via the WebUI, in a quick glance about key features and potential issues present in the infrastructure. Given a set of controls, the monitoring service also takes care of raising \textit{alerts} and \textit{notifications} informing the IQM about the status of the infrastructure and its operation.

\section{Architecture}\label{arch}

The Monitoring Framework (see Figure 1) is architected as a client-server application, where a core module (imported and used by the code of the aggregating infrastructure) plays the role of \textit{client}. The infrastructure source code needs to be instrumented in order to put sensors in place and produce observations of the measured objects.

The implementation of a sensor provides the logic to produce a measurement; in general, a sensor is devised to probe a specific object type (thus the typing of the measured object is hardcoded), while its configuration can be defined via WebUI and retrieved at runtime (dynamically). Such a design opens up to configurability and extendibility of the sensor’s collection offered by the DFDS, which in any case comes with off-the-shelf implemented sensors.

The \textit{server} component is a stand-alone web application that receives observations from sensors, stores those observations as time series and runs automatically user-defined controls over this corpus of data. The controls present the outcome of their activity as reports, which are made available to the IMQ via the WebUI. There is also an alert and notification service to warn the IMQ about potential anomalies or wrong operation in the aggregating infrastructure.

\section{The EAGLE use case}\label{usecase}

In order to apply the monitoring framework described above to the EAGLE aggregator, we need to define first a monitoring scenario, based on the actual EAGLE data flow, which must include the following. 
\begin{itemize}
\item Sensors to be used in terms of measured objects and metrics implemented (i.e. functions to apply in order to extract observations). Once deployed and running, a sensor dynamically retrieves its defined configuration from the server whenever it is needed. Metrics identified in this test case are reported in Section \ref{sec:metrics}.
\item Controls representing quality checks to be run against measurements obtained with certain metrics. The framework guides the IQM into the definition of a control according to the scenario defined so far. Controls identified in this test case are reported in Section \ref{sec:controls}.
\end{itemize}

\subsection{Metrics}\label{sec:metrics}

A monitoring scenario has been defined and tested on simulated data in order to monitor the ``health'' of different data collections involved in the EAGLE collection and transformation workflows. In particular, a first collection is stored in a full-text index (Apache Solr) serving search queries, and a second one is stored in a document store (MongoDB) serving data for OAI-PMH export.

In this same scenario, we are interested in monitoring also the collection workflow by inspecting every single native XML record flowing into the EAGLE infrastructure from content providers. Some useful metrics identified in EAGLE are described in Table \ref{tab:metrics}.

\begin{table}
{\small
\addtolength{\tabcolsep}{-0.5mm}
\begin{tabular*}{\textwidth}{ l l }
\toprule
Measured property & Metric\\
\midrule
\parbox[t]{0.6\textwidth}{Total \# of content providers joining the EAGLE infrastructure} & Content providers\\
Total \# of languages for translations & Languages\\
Total \# of EAGLE records & Total records\\
\parbox[t]{0.6\textwidth}{Compliance toward the controlled vocabulary for materials. The values of this metric track the percentage of vocabulary-compliant occurrences in the XML field containing the value for ``material'', over the total amount of occurrences of that field. As an example, if the vocabulary defines the entries ``aaa'', ``bbb'', ``ccc'' and the vocabulary-controlled XML field uses ``aaa'' four times, ``bbb'' three times, and ``xxx'' three times, the metric yields 0.7 (i.e. 7 out of 10 occurrences match the vocabulary).} & Voc:material compliance\\
\parbox[t]{0.6\textwidth}{Completeness of every single collected native XML record. The values of this metric track the percentage of non-empty XML fields among 5 user-defined fields (e.g. title, description, object type, date, material.). The value could be 0\%, 20\%, 40\%, 60\%, 80\%, 100\%, depending on how many non-empty values have been found.} & Completeness\\
\bottomrule
\end{tabular*}}
\caption{Metrics implemented in EAGLE.}
\label{tab:metrics}
\end{table}

\subsection{Controls}\label{sec:controls}

Given the metrics described in \ref{sec:metrics}, Table \ref{tab:controls} reports some controls defined over those metrics.

\begin{table}
{\small
\addtolength{\tabcolsep}{-0.5mm}
\begin{tabular*}{\textwidth}{ l l }
\toprule
Metric & Control\\
\midrule
Content providers & \parbox[t]{0.6\textwidth}{Check if the number of content providers indexed in Solr is monotonic increasing over time (considering the three last observations of the metric)}\\
Content providers & \parbox[t]{0.6\textwidth}{Check whether the number of content providers indexed in Solr equals the number of OAI sets present in MongoDB (considering only the last observation of the metric)}\\
Languages & \parbox[t]{0.6\textwidth}{Check if the number of modern languages present in translations indexed in Solr is steadily increasing over time (considering the two last observations of the metric)}\\
Total records & \parbox[t]{0.6\textwidth}{Check whether the total number of EAGLE records (per content provider) is steadily increasing over time (considering only the three last observations)}\\
Voc:material compliance & \parbox[t]{0.6\textwidth}{Check if such indicator, ranging from 0.0 to 1.0, is above 0.9 threshold (considering only the very last observation of the metric)}\\
Completeness & \parbox[t]{0.6\textwidth}{Check if such indicator (actually its rolling average), ranging from 0.0 to 1.0, is above 0.8 threshold (considering only the very last observation of the metric average)}\\
\bottomrule
\end{tabular*}}
\caption{Controls implemented in EAGLE.}
\label{tab:controls}
\end{table}

\subsection{Sample implementation}\label{sample}

Three different sensors have been defined and implemented for this test on EAGLE: two collection sensors (one for Solr and one for MongoDB) and one single-datum sensor for XML record-by-record inspection. Once the three sensors have been placed in the EAGLE workflow implementation and the EAGLE infrastructure is running, they start to produce observations and deliver them to the server component of the monitoring framework.

As an example, we report in figures included in Section \ref{sec:figures} simulated trends of the defined metrics and relative reports (i.e. the evaluation of controls defined over the metrics).

The metric about the ``Total number of content providers'' is reported in Figure \ref{fig:cps}; as expected, the number of content providers indexed is monotonically (each value is greater or equal to the previous one) increasing, as stated in the leftmost report, and the number of OAI sets present in the data collection stored in MongoDB equals the number of content providers indexed in Solr, as stated in the rightmost report. It is also interesting to notice how the two trends diverged in time back in October. In the simulated data we introduced an \textit{ad-hoc} problem in October, related to the publication of EAGLE records into the OAI-PMH store, which the monitoring service succeeded to discover.

In Figure \ref{fig:lang}, we report the ``Languages'' metric; as expected its trend is strictly increasing over time indicating that the corpus of translations is expanding and that they are correctly integrated into the system.

Figure \ref{fig:tot} shows the total number of EAGLE records for four content providers (CP1, CP2, CP3, CP4) and informs the IQM that everything behaves as expected; in fact, the four reports states that the trends of the metric are always increasing over time.

In Figure \ref{fig:voc}, the metric ``Voc:material compliance'' is depicted. The four trends show that the four content providers (CP1, CP2, CP3, CP4) are in general increasing the quality of their data by enforcing the use of correct values offered by the controlled vocabulary for materials. However, CP1 (in orange), with a 0.83 score, does not meet the threshold requirement (set to 0.9) indicated in the control, thus its report is marked in red notifying the issue.

Figure \ref{fig:comp} reports the oscillations of the record-by-record metric ``Completeness'' (narrowed down to a hundred records sample), while Figure \ref{fig:ravg} reports the associated ``rolling average'' (i.e. each point is the updated average up to that instant). Again, as the last observation of metrics is equal to 0.52, the relative reports is marked in red, as the 0.8 threshold defined in the control is not met.

\section{Conclusions and future work}\label{sec:conclusions}

We have presented here the possible application of the Monitoring Framework concepts to the EAGLE aggregating infrastructure. The Monitoring Framework is ``work in progress'' in another European project related to research data infrastructures \citep{openaire}.

The results presented here are based on simulated data, as the EAGLE infrastructure is not yet instrumented with the sensors needed for collecting observations, but we are planning to instrument it as soon as the development of the Monitoring Framework will reach the \textit{beta} status.

EAGLE represents an ideal \textit{testbed} for this monitoring technology, as the workflows are clean and well defined, the data collected also is well defined, given the mappings that have been defined between the different incoming records and the EAGLE data model.

Finally, when EAGLE will be equipped with the final Monitoring Framework, we expect that it will provide valuable data for ensuring that the epigraphy data made available at the EAGLE portal will be of the highest quality. It will also provide valuable feedback to the content providers, helping them to detect possible inconsistencies and lack of information in their data, in order to improve the quality of the data provided to EAGLE and also, even more important, the quality of the data that each content provider makes available to its users.
\clearpage

\section{Figures}\label{sec:figures}
\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{img/arch.jpg}
\caption{The architectural overview of the monitoring framework.}
\label{fig:arch}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{img/cps.png}
\caption{The trend of the metric ``Total number of content providers''.}
\label{fig:cps}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{img/lang.png}
\caption{The trend of the metric ``Languages''.}
\label{fig:lang}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{img/tot.png}
\caption{The trend of the metric ``Total records''.}
\label{fig:tot}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{img/voc.png}
\caption{The trend of the metric ``Voc:material compliance''.}
\label{fig:voc}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{img/comp.png}
\caption{The trend of the metric ``Completeness''.}
\label{fig:comp}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{img/comp_ravg.png}
\caption{The ``rolling average'' of the metric ``Completeness''.}
\label{fig:ravg}
\end{figure}

\section*{Acknowledgements}
This work is partly funded by the EU EAGLE Best Practice Network project: Grant Agreement CIP 325122, call CIP-ICT-PSP-2012-6.

\bibliographystyle{sapauth-eng}
\bibliography{EAGLE}

\end{document}

### 2015-12-16 source/10/Paper proposal for EAGLE final.tex
more biblio added, papers compiled with bibliography

@@
 media, decorative and paratextual elements\footnote{On marks in the text see \citet[26]{susini1973} and \citet[143-155]{cooley2014}.} (e.g.
-ivy leaves as interpuncts (hederae distinguentes),\footnote{\citep[293-303]{hommel1970}}
-abbreviations\footnote{\citep[111]{Gordon1957}} and spacing between words\footnote{On using
+ivy leaves as interpuncts (hederae distinguentes),\footnote{\citet[293-303]{hommel1970}}
+abbreviations\footnote{\citet[111]{Gordon1957}} and spacing between words\footnote{On using
 space to reconstruct inscriptions see \citet[195-226]{alfoldy1995}, \citet[151-176]{grasby2002}.} alongside lettering, as dating
@@
 
-%\nocite{} check after compiling with quoted bibliography
+\nocite{roueche2004}
 

### 2015-12-16 source/13/EAGLE2016_Full_Paper_Verreth.tex
more biblio added, papers compiled with bibliography

@@
 which yielded in the end a full index of all personal names and toponyms in every Greek and Latin papyrus. This process
-is nowadays called Named-Entity Recognition [NER]\footnote{ For more information on NER, cf. Broux, Yanne - Depauw,
-Mark, Developing onomastic gazetteers and prosopographies for the ancient world through named entity recognition and
-graph visualization: some examples from Trismegistos people, in 2nd International workshop on computational history. 2.
-Barcelona, 11 November 2014 (Lecture notes in computer science, 8852), Berlin - Heidelberg, 2015, p. 304-313.}. These
+is nowadays called Named-Entity Recognition [NER]\footnote{ For more information on NER, \citet[304-313]{BrouxDepauw}.}. These
 TM files within People and Places are freely accessible online and can be looked up, questioned and investigated in a

### 2015-12-16 source/17/Zuddas Epigraphy in Italian High Schools (final).tex
other biblio added till paper 20

@@
 paper published online\footnote{\url{http://mediaclassica.loescher.it/nuove-e-\%93vecchie\%94-forme-di-multimedialita.n2799}
-; see also \citep{Reali2015}.}. Being an expert on the subject, he offers a comparison between the ``noble'' form of
+; see also \citet{Reali2015}.}. Being an expert on the subject, he offers a comparison between the ``noble'' form of
 the political-philosophical amicitia presented in Cicero’s Laelius and the term \emph{amicus} mentioned in inscriptions
@@
 \begin{quotation}
-[{}- - - I]o[323?]vino consulibb(us) (ante diem) VIII Kal(endas) Martias domum oscilavi Musae. 
+
+[{}- - - I]ọvino consulibb(us) (ante diem) VIII Kal(endas) Martias domum oscilavi Musae. 
 \end{quotation}

### 2015-12-16 source/20/EAGLE2016_BONACINI - Pilot project at Paolo Orsi Museum.tex
other biblio added till paper 20

@@
 \cortext[first]{Corresponding author. Email: e_bonacini@hotmail.com; elisa.bonacini@unict.it}
-
 \address[catania]{Humanities Department, University of Catania; IEMEST (Istituto Euro Mediterraneo di Scienze e Tecnologie), Palermo, Italy}
@@
 Despite of this, Sicilian cultural heritage struggles to be present on Google’s platforms such as Google Art Project and
-others, as it should, compared to other Italian cultural sites \citep{Bonacini2013;Bonacini2014}).
+others, as it should, compared to other Italian cultural sites \citep{Bonacini2013} \citep{Bonacini2014}).
 
@@
 refrigerium, the funeral feast that symbolically was consumed with the dead, even this syncretic practice \ dates back
-to the pagan world \citep{Sgarlata2013;Scandurra2014}.
+to the pagan world \citep{Sgarlata2013}, \citep{Scandurra2014}.
 

### 2015-12-17 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
all biblio in EAGLE.bib

@@
 
+CTS = Canonical Text Services (protocol to cite digitally-available texts in a canonical way).
+
+EAGLE = The Europeana network of Ancient Greek and Latin Epigraphy: \ http://www.eagle-network.eu/
+
+IEs = Integrating Digital Epigraphies project.
+
+LGPN = The Lexicon of Greek Personal Names: http://www.lgpn.ox.ac.uk/
+
+Pleiades = A community-built gazetteer and graph of ancient places: \ http://pleiades.stoa.org/
+
+SNAP = Standards for Networking Ancient Prosopographies project: http://snapdrgn.net/
+
+Trismegistos = Interdisciplinary portal of papyrological and epigraphical resources formerly Egypt and the Nile valley
+(800 BC-AD 800): http://www.trismegistos.org/
+
 \bibliographystyle{sapauth-eng}

### 2015-12-17 source/3/EAGLE_2016_submission_3.tex

--- 2015-12-14 source/3/EAGLE2016_Full_Paper_Template.tex
@@
 \usepackage{ucs}
-\usepackage{supertabular}
 \usepackage[greek,english]{babel}
@@
 \newcommand{\bs}{\textbackslash}
-
+\usepackage{xtab}
 %%%%%%%%
@@
 
-\author[VARGA]{Rada Varga}
+\author[CLUJ]{Rada Varga\corref{first}}
 
-\address{Babeș-Bolyai University, Cluj-Napoca}
+\address[CLUJ]{Babeș-Bolyai University, Cluj-Napoca}
 
@@
 
+%\usepackage{longtable}
 \begin{document}
@@
 The database in itself would have three components (see below Fig.~\ref{fig:schemadatabase}), following the best practices in the field
-(\citet{doi:10.3200/HMTS.37.1.34-38}): the sources database (with ``facsimile'' transcription of the sources’ text), the central
-database (the complete, corrected, integrated, standardized and coded form of the sources database) and the data
-releases (destined for on-line usage, allowing easy extraction of data in view of analyses).
+(\citet{doi:10.3200/HMTS.37.1.34-38}): the \emph{sources database} (with ``facsimile'' transcription of the sources’ text), the \emph{central database} (the complete, corrected, integrated, standardized and coded form of the sources database) and the \emph{data releases} (destined for on-line usage, allowing easy extraction of data in view of analyses).
 
@@
 Low Danube provinces and their epigraphic attestations. The database created in this context together with the platform
-www.romans1by1.com represent the skeleton of the future population database of the Roman Empire.
+\url{www.romans1by1.com} represent the skeleton of the future population database of the Roman Empire.
 
@@
 \noindent The core of the database’s epigraphical division is formed by a table used for recording personal
-individual data (labelled Personal data), around which the entire network of components needed to ensure proper
+individual data (labelled \emph{Personal data}), around which the entire network of components needed to ensure proper
 information recording is built. However, when starting data entry, one must begin by recording information about the
@@
 
-The first table to be filled in is the file of the source – Inscriptions. First of all, each inscription gets a source
+The first table to be filled in is the file of the source – \emph{Inscriptions}. First of all, each inscription gets a source
 code, formed of 5 digits and a symbol/acronym of the province’s name (D for Dacia, MS for Moesia Superior, DAL for
-Dalmatia, etc.) – so we have, for example 00001MS. The Inscriptions table has text fields, as well as value lists.
-Then, we have text strings for: Relevant expressions, Stylistic details, Atypical features, Observations, Place of
-discovery, Place of provenience, Ancient name provenience, Timestamp/Timeframe and External links. Although we are
-aware that some of the data (Timestamp/Timeframe) could have benefitted from a standardized form, at this point we
+Dalmatia, etc.) – so we have, for example 00001MS. The \emph{Inscriptions} table has text fields, as well as value lists.
+Then, we have text strings for: \emph{Relevant expressions}, \emph{Stylistic details}, \emph{Atypical features}, \emph{Observations}, \emph{Place of discovery}, \emph{Place of provenience}, \emph{Ancient name provenience}, \emph{Timestamp/Timeframe} and \emph{External links}.
+Although we are
+aware that some of the data (\emph{Timestamp/Timeframe}) could have benefitted from a standardized form, at this point we
 opted for vaster possibilities of expression and adaptation. Other information will be filled using value lists, as
-standardization is more suited: Type of inscription, Language, Material. For these we will use the Eagle
+standardization is more suited: \emph{Type of inscription}, \emph{Language}, \emph{Material}. For these we will use the Eagle
 vocabularies,\footnote{\url{http://www.eagle-network.eu/resources/vocabularies}} because they offer an already
@@
 
-The Coordinates table links each inscription with the latitude and longitude of its place of provenience, in order to
+The \emph{Coordinates} table links each inscription with the latitude and longitude of its place of provenience, in order to
 place it on a map.
 
-The Inscription bibliography was conceived so that extracting complete or selective bibliographical lists would be
+The \emph{Inscription bibliography} was conceived so that extracting complete or selective bibliographical lists would be
 possible. Thus, a normalization table includes all bibliographical titles referred to and from it; through a value
-list, one can select the Bibliography abbreviation. The exact reference is presented and detailed in the Details and
-Comments text strings. Of course, all data are linked to the Inscription code, selected as well from a value list.
+list, one can select the \emph{Bibliography abbreviation}. The exact reference is presented and detailed in the \emph{Details} and
+\emph{Comments} text strings. Of course, all data are linked to the \emph{Inscription code}, selected as well from a value list.
 
-Only after properly documenting the source one can access the main and most complex table, Personal data, where
+Only after properly documenting the source one can access the main and most complex table, \emph{Personal data}, where
 basically the individual file of each person is created. At this point, each new entry represents a singular epigraphic
@@
 
-The Personal data table was built to host a large variety of information offered by epigraphs, and its structure has
+The \emph{Personal data} table was built to host a large variety of information offered by epigraphs, and its structure has
 proved, up to this day, rather stable. However, if need be, it can always be extended in order to accommodate any new
-kind of information. A first set of its variables are name-connected: Praenomen, Nomen, Cognomen/Personal name,
-Father/Master name, Agnomen, Signum, where each also has a drop-down associated for Ethnicity and the Agnomen and
-Signum an Observations string. While we believe the possibilities of actually identifying signa for members of the
+kind of information. A first set of its variables are name-connected: \emph{Praenomen}, \emph{Nomen}, \emph{Cognomen/Personal name},
+\emph{Father/Master} name, \emph{Agnomen}, \emph{Signum}, where each also has a drop-down associated for \emph{Ethnicity} and the \emph{Agnomen} and
+\emph{Signum} an \emph{Observations} string. While we believe the possibilities of actually identifying \emph{signa} for members of the
 middle classes, during the Principate period, is rather reduced, we opted for facilitating their correct registration,
-in case they are discovered. Other data regard Natione, Ethnicity, Origo and Domus – if and how they are mentioned in
+in case they are discovered. Other data regard \emph{Natione}, \emph{Ethnicity}, \emph{Origo} and \emph{Domus} – if and how they are mentioned in
 the inscription. As acknowledged above, some information will be recorded, even if they are deductive and not literally
-written in the source: Gender and Juridical status (though the servile one often is literally recorded). For the
+written in the source: \emph{Gender} and \emph{Juridical status} (though the servile one often is literally recorded). For the
 latter, we have opted for a check box, which, if checked, opens all the available possibilities. The rest of the fields
-accommodate supplementary information, when and how there is the case: Occupation, Collegium, Deities, Age (at death),
-Details of life/death and Observations.
+accommodate supplementary information, when and how there is the case: \emph{Occupation}, \emph{Collegium}, \emph{Deities}, \underline{Age} (at death),
+\emph{Details of life/death} and \emph{Observations}.
 
-The Occupation field requires some special attention, as it has the Occupation code associated with it; as we are trying
+The \emph{Occupation} field requires some special attention, as it has the \emph{Occupation code} associated with it; as we are trying
 to propose a codification of Roman occupations/professions, based on and adapting the HISCO classification
@@
 be a slight challenge, finding a theoretical model might prove to be a more elaborated task. What we aim at is
-constructing the “metadata” for the codification of Roman professions attested on stone and analyse how deep the
+constructing the ``metadata'' for the codification of Roman professions attested on stone and analyse how deep the
 classification can realistically go. Much alike other normalization fields present in the sources component of the
-database, Occupation code was implemented because it helps dealing faster with small and medium size samples (up to
+database, \emph{Occupation code} was implemented because it helps dealing faster with small and medium size samples (up to
 hundreds of people) in view of publishing preliminary results which are vital for dissemination and further financial
@@
 
-Two important variables, for statistics and working with the data, are Dedicated by and Dedicated for, which state the
+Two important variables, for statistics and working with the data, are \emph{Dedicated by} and \emph{Dedicated for}, which state the
 relation of the recorded individual with the epigraph and with other persons mentioned by it. Both appear in the simple
@@
 of the elites. In this case, we have to register a minimum of data on the later as well, in order to build a wider
-image of the characters that we focus on. When checking Later, it opens Status, which at its turn opens the following
-options in the form of check boxes: Senator, Knight, Local magistrate, Decurion, Imperial priest, Imperial slave,
-Imperial freedman, Military personnel. If checked, each of these boxes opens a Details text box; additionally, Local
-magistrate and Decurion open a City/Town value list, while Military personnel, opens Rank and Unit value lists. While
+image of the characters that we focus on. When checking \emph{Later}, it opens \emph{Status}, which at its turn opens the following
+options in the form of check boxes: \emph{Senator}, \emph{Knight}, \emph{Local magistrate}, \emph{Decurion}, \emph{Imperial priest}, \emph{Imperial slave},
+\emph{Imperial freedman}, \emph{Military personnel}. If checked, each of these boxes opens a Details text box; additionally, \emph{Local
+magistrate} and \emph{Decurion} open a \emph{City/Town} value list, while \emph{Military personnel}, opens \emph{Rank} and \emph{Unit} value lists. While
 elite members and military personnel will at some point be added to the database, for the currently running project
@@
 recorded individuals represent enough information. 
+\vspace{20pt}
 
-\begin{center}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{|m{3.711cm}|m{2.941cm}|m{4.046cm}|}
-\hline
-Field label &
-Field id &
-Data type\\\hline
+%\begin{table}[!h]
+\tablefirsthead{\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline }
+\bottomcaption{Metadata of the Personal data table}
+
+%\tablehead{\multicolumn{3}{c}%
+%{{\captionsize\bfseries \tablename\ \thetable{} --
+%continued from previous page}} \\
+%\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline }
+
+\tablehead{\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline}
+
+\tabletail{\hline \multicolumn{3}{|r|}{{\textit{Continued on next page}}} \\ \hline}
+\tablelasttail{\hline}
+
+{\small
+\addtolength{\tabcolsep}{-0.5mm}
+%\begin{supertabular*}{\textwidth}{p{3.5cm}lp{3cm}}
+\begin{xtabular}{|p{.30\textwidth}|p{.30\textwidth}|p{.30\textwidth}|}
+%\begin{tabular*}{p{.30\textwidth}p{.29\textwidth}p{.31\textwidth}}
+%\toprule
+%\textbf{Field label} & \textbf{Field id} & \textbf{Data type}\\\midrule
 Praenomen &
 rperson\_praenomen &
-Text\\\hline
+Text\\ \hline
 Ethnicity praenomen &
 rperson\_eth\_praen &
-Value list\\\hline
+Value list\\ \hline
 Nomen &
 rperson\_nomen &
-Text\\\hline
+Text\\ \hline
 Ethnicity nomen &
 rperson\_eth\_nom &
-Value list\\\hline
+Value list\\ \hline
 Cognomen/Personal name &
 rperson\_cognomen &
-Text\\\hline
+Text\\ \hline
 Ethnicity cognomen/personal name &
 rperson\_eth\_cogn &
-Value list\\\hline
+Value list\\ \hline
 Father/Master name &
 rperson\_father &
-Text\\\hline
+Text\\ \hline
 Ethnicity father/master name &
 rperson\_eth\_fth &
-Value list\\\hline
+Value list\\ \hline
 Agnomen &
 rperson\_agnomen &
-Text\\\hline
+Text\\ \hline
 Ethnicity agnomen &
 rperson\_eth\_agn &
-Value list\\\hline
+Value list\\ \hline
 Observations agnomen &
 rperson\_remark\_agn &
-Text\\\hline
+Text\\ \hline
 Signum &
 rperson\_signum &
-Text\\\hline
+Text\\ \hline
 Ethnicity signum &
 rperson\_eth\_sign &
-Value list\\\hline
+Value list\\ \hline
 Observations signum &
 rperson\_remark\_sign &
-Text\\\hline
+Text\\ \hline
 Natione &
 rperson\_nation &
-Text\\\hline
+Text\\ \hline
 Ethnicity &
 rperson\_nat\_ethnicity &
-Text\\\hline
+Text\\ \hline
 Gender &
 rperson\_rgender\_id &
-Value list\\\hline
+Value list\\ \hline
 Juridical status &
@@
 
-Citizen/Libertus/Peregrine/Slave/
-
-Veteranus (opens Unit/Rank - valuelists)\\\hline
+Citizen, Libertus, Peregrine, Slave 
+Veteranus (opens Unit/Rank - valuelists)\\ \hline
 Tribus &
 rperson\_rtribus\_id &
-Text\\\hline
+Text\\ \hline
 Origo &
 rperson\_origo &
-Text\\\hline
+Text\\ \hline
 Domus &
 rperson\_domus &
-Text\\\hline
+Text\\ \hline
 Collegium &
 rperson\_collegium &
-Text\\\hline
+Text\\ \hline
 Occupation &
 rperson\_occupation &
-Text\\\hline
+Text\\ \hline
 Occupation code &
 rperson\_occ\_code &
-Text\\\hline
+Text\\ \hline
 Deities &
 rperson\_deity &
-Text\\\hline
+Text\\ \hline
 Age &
 rperson\_age &
-Text\\\hline
+Text\\ \hline
 Details of life/death &
 rperson\_death\_details &
-Text\\\hline
-Dedicated for &
-rperson\_dedicated\_for &
-Checkbox\\\hline
-Dedicated by &
-rperson\_dedicated\_by &
-Checkbox\\\hline
-Inscription code &
-rperson\_rinscription\_ids &
-Value list\\\hline
+Text\\ \hline
+Dedicated for & rperson\_dedicated\_for & Checkbox\\ \hline
+Inscription code & rperson\_rinscription\_ids & Value list\\ \hline
 Later &
-rperson\_now\_stat &
-Checkbox. Opens checkbox: Status\\\hline
-Status &
-rperson\_status &
-Checkbox. Opens checkboxes:
-
-Senator (opens Details – text)/
-
-Knight (opens Details – text)/
-
-Local Magistrate (opens Details – text, City – value list)/
-
-Decurion (opens Details – text, City – value list)/
-
-Imperial priest (opens Details – text)/
-
-Imperial slave (opens Details – text)/
-
-Imperial freedman (opens Details – text)/
-
-Military personnel (opens Details – text, Unit/Rank – value lists)\\\hline
+rperson\_now\_stat & Checkbox. Opens checkbox:
+Status\\ \hline
+Status & rperson\_status & Checkbox. Opens checkboxes: 
+Senator (opens Details – text),
+Knight (opens Details – text) 
+Local Magistrate (opens Details – text, City – value list), 
+Decurion (opens Details – text, City – value list), 
+Imperial priest (opens Details – text), 
+Imperial slave (opens Details – text), 
+Imperial freedman (opens Details – text), 
+Military personnel (opens Details – text, Unit/Rank – value lists)\\ \hline
 Observations &
 rperson\_observation &
-Text\\\hline
+Text\\ \hline
+%\bottomrule
+
+\end{xtabular}}
+%\caption{This is a wide table where we have used the \texttt{tabular*} environment.
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.}
+%\label{tab:personaldata}
+%\end{table}
+
 %\caption{Metadata of the Personal data table}
-\end{supertabular}
-\label{tab:personaldata}
-\end{center}
+%\end{supertabular}
+%\label{tab:personaldata}
+%\end{center}
 
-Based on the personal ID given to each individual, the Relationship table will solely name the relationship between
+Based on the personal ID given to each individual, the \emph{Relationship} table will solely name the relationship between
 individuals (A to B and B to A), choosing from a drop-down menu. The relationships have been encoded from the start, in
@@
 
-\noindent Romans 1 by 1 is a first step towards a comprehensive and exhaustive electronic resource for the attested population of
+\noindent \emph{Romans 1 by 1} is a first step towards a comprehensive and exhaustive electronic resource for the attested population of
 the Roman Empire. The following normal steps for expanding and enriching the database is elaborating the fitted

### 2015-12-17 source/21/EAGLE2016_Full_Paper_revised.tex
paper 21 updated
first doc to tex correction done

@@
 %%%%%%%%
-\title{}
-\author{}
-\date{}
+\title{AXON. A Selection of Greek Historical Inscriptions. A Database for Research and Teaching}
+ \author[cafo]{Ivan Matijašić\corref{first}}
+\author[cafo]{Stefania De Vido}
+\author[cafo]{Silvia Palazzo}
+\address[cafo]{Dipartimento di Studi Umanistici, Università Ca' Foscari Venezia}
+\cortext[first]{Corresponding author. Email: ivan.matijasic@unive.it}
 \begin{document}
-AXON. A Selection of Greek Historical Inscriptions
 
-A Database for Research and Teaching
-
-\bigskip
-
-Abstract
-
+\maketitle
+\begin{abstract}
 The AXON Project has developed a database of Greek historical inscriptions, from the birth of the~polis~in the Archaic
@@
 
-Keywords
+\end{abstract}
+\keywords{Online epigraphic editions; interoperability of digital editions of Greek historical inscriptions; images of Greek
+historical inscriptions; digital epigraphy in teaching and research}
 
-Online epigraphic editions; interoperability of digital editions of Greek historical inscriptions; images of Greek
-historical inscriptions; digital epigraphy in teaching and research
-
-\bigskip
-
-1. The AXON Project
+\section{The AXON Project }
 
@@
 
-AXON includes texts from the birth of the Greek polis in the Archaic Age\footnote{ See the Introduction in Hansen \&
-Nielsen 2004; Hansen 2006. } to 31 BC, a chronological frame traditionally related to Greek History (though a future
+AXON includes texts from the birth of the Greek polis in the Archaic Age\footnote{ See the Introduction in \citet{nielsen_inventory_2004}; \citet{hansen_polis:_2006}.
+} to 31 BC, a chronological frame traditionally related to Greek History (though a future
 extension of this chronological limit is not excluded). The epigraphic entries have been prearranged in order to allow
@@
 
-\bigskip
+\section{A unique model-entry for a great diversity of inscriptions: taxonomy and categorisation}
 
-2. A unique model-entry for a great diversity of inscriptions: taxonomy and categorisation
+\subsection{Entry description}
 
-\bigskip
-
-2.1. Entry description
-
-\bigskip
-
-2.1.1. Object’s description
+\subsubsection{Object’s description}
 
@@
 categories – such as Object type, Material, or State of preservation – are directly linked with the corresponding
-sections in the EAGLE Vocabularies (see http://www.eagle-network.eu/resources/vocabularies/). Furthermore, a hyperlink
-has been created between the AXON-entries and Pleiades website (http://pleiades.stoa.org): where the finding site is
+sections in the EAGLE Vocabularies (see \url{http://www.eagle-network.eu/resources/vocabularies/}). Furthermore, a hyperlink
+has been created between the AXON-entries and Pleiades website (\url{http://pleiades.stoa.org}): where the finding site is
 known, each entry offers the geographic coordinates and a Googlemaps visualisation. This gives the possibility of
@@
 
-\bigskip
+\subsubsection{Chronology}
 
-2.1.2. Chronology
+The chronological delimitation of each text is supported by many options, as you can see in Fig.~\ref{fig:1}:
 
-The chronological delimitation of each text is supported by many options, as you can see in fig. 1:
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperrevised-img001.png}
+\caption{Window for the input of data, section Text/Chronology}
+\label{fig:1}
+\end{figure}
 
-\begin{figure}
-\centering
-\includegraphics[width=10.88cm,height=4.156cm]{EAGLE2016FullPaperrevised-img001.png}
-\end{figure}
-Fig. 1. Window for the input of data, section Text/Chronology
-
-\bigskip
-
-2.1.3. Alphabet \& language
-
-\bigskip
+\subsubsection{Alphabet \& language}
 
@@
 \item Type of Inscription (with link to EAGLE’s Vocabularies:
-http://www.eagle-network.eu/resources/vocabularies/typeins/. The categories are those used by Guarducci 1967-1978)
+\url{http://www.eagle-network.eu/resources/vocabularies/typeins/}. The categories are those used by \citet{guarducci_epigrafia_1987})
 \item Text’s structure
-\item Writing (Execution technique\footnote{ With link to http://www.eagle-network.eu/resources/vocabularies/writing/.};
+\item Writing (Execution technique\footnote{ With link to \url{http://www.eagle-network.eu/resources/vocabularies/writing/}.};
 different types of epichoric alphabets according to Kirchhoff’s colour-coded map; Local script\footnote{ Following the
-categorisation in Jeffery 1990.}; Palaeographic features and letters’ form\footnote{ The letter-form and glyph-form are
-based on the symbols of the font Cardo (http://scholarsfonts.net/cardofnt.html), but many have been developed by the
+categorisation in \citet{jeffery_local_1990}.}; Palaeographic features and letters’ form\footnote{ The letter-form and glyph-form are
+based on the symbols of the font Cardo (\url{http://scholarsfonts.net/cardofnt.html}), but many have been developed by the
 AXON Team on the examples of letter-form given in Jeffrey 1990 (see also
-http://poinikastas.csad.ox.ac.uk/browseGlyphs.shtml).}; letters’ heights, description and layout of the text field;
+\url{http://poinikastas.csad.ox.ac.uk/browseGlyphs.shtml}).}; letters’ heights, description and layout of the text field;
 Direction of Text)
@@
 
-\bigskip
-
-\bigskip
-
-2.1.4. Genetic lemma \& apparatus criticus
+\subsubsection{Genetic lemma \& apparatus criticus}
 
 The text of each inscription is preceded by a hierarchically arranged lemma (the so called genetic lemma, according to
-Louis Robert’s definition\footnote{ Robert \& Robert 1954.}) and is followed by the apparatus criticus.
+Louis Robert’s definition\footnote{\citet{robert_carie:_1954}.}) and is followed by the apparatus criticus.
 
-\bigskip
-
-2.1.5. Italian translation \& commentary
+\subsubsection{Italian translation \& commentary}
 
@@
 
-\bigskip
-
-2.1.6. Abstract
+\subsubsection{Abstract}
 
@@
 
-\begin{figure}
+\begin{figure}[!bp]
 \centering
-\includegraphics[width=10.88cm,height=5.449cm]{EAGLE2016FullPaperrevised-img002.png}
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperrevised-img002.png}
+\caption{Example of Abstract in AXON. The different colours allow a selection of words from a drop-down menu. }
+\label{fig:2}
 \end{figure}
-Fig. 2. Example of Abstract in AXON. The different colours allow a selection of words from a drop-down menu.
-
-\bigskip
 
@@
 
-\bigskip
-
-2.1.7. Bibliography
+\subsubsection{Bibliography}
 
@@
 secondary sources. SEG abbreviations have been used for epigraphic corpora and other publications (the section
-“materiali” on the website gives access to a list of all abbreviations, useful for students, as well\footnote{
-http://virgo.unive.it/venicepigraphy/axon/public/axon/pagine/materiali (still being processed).}).
+``materiali'' on the website gives access to a list of all abbreviations, useful for students, as well\footnote{
+\url{http://virgo.unive.it/venicepigraphy/axon/public/axon/pagine/materiali} (still being processed).}).
 
-\bigskip
-
-2.2. Internal \& external interoperability of the AXON database
-
-\bigskip
+\subsection{Internal \& external interoperability of the AXON database}
 
@@
 
-\bigskip
-
-2.3. A simple website interface for the input of data
-
-\bigskip
+\subsection{A simple website interface for the input of data}
 
@@
 
-\bigskip
+\section{Searchability}
 
-\bigskip
+The website is designed to allow for many search options. Beyond the ``full text'' search and another based on the number,
+title and author of the entry (see Fig.~\ref{fig:3}), three other search-possibilities will also be available:
 
-3. Searchability
-
-\bigskip
-
-The website is designed to allow for many search options. Beyond the “full text” search and another based on the number,
-title and author of the entry (see fig. 3), three other search-possibilities will also be available:
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperrevised-img003.png}
+\caption{Search based on ``Full text'', number, title, and author of the entry.}
+\label{fig:3}
+\end{figure}
 
@@
 \item access the entries through an interactive map;
-\end{enumerate}
-
-\begin{figure}
-\centering
-\includegraphics[width=10.88cm,height=7.354cm]{EAGLE2016FullPaperrevised-img003.png}
-\end{figure}
-Fig. 3. Search based on “Full text”, number, title, and author of the entry.
-
-\bigskip
-
+\item perform an advanced search based on different categories: 
 \begin{enumerate}
-\item perform an advanced search based on different categories: 
-\end{enumerate}
-\begin{enumerate}
-\item \begin{enumerate}
 \item bibliography
 \item keywords
-\item object’s description and preservation (see fig. 4)
+\item object’s description and preservation (see Fig.~\ref{fig:4})
 \item chronology
-\item text (single words or phrases, typology, dialect, alphabet, letter-form, etc.) (see fig. 5).
+\item text (single words or phrases, typology, dialect, alphabet, letter-form, etc.) (see Fig.~\ref{fig:5}).
 \end{enumerate}
@@
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperrevised-img004.png}
+\caption{Different searching categories.}
+\label{fig:4}
+\end{figure}
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016FullPaperrevised-img005.png}
+\caption{Advanced text search}
+\label{fig:5}
+\end{figure}
 
-\begin{figure}
-\centering
-\includegraphics[width=10.88cm,height=9.518cm]{EAGLE2016FullPaperrevised-img004.png}
-\end{figure}
-Fig. 4. Different searching categories.
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-Fig. 5. Advanced text search.
-
-\bigskip
-
-\begin{figure}
-\centering
-\includegraphics[width=10.88cm,height=9.888cm]{EAGLE2016FullPaperrevised-img005.png}
-\end{figure}
-Other filters will be employed for each search result. A section entitled “tools” will also be available, and will
+Other filters will be employed for each search result. A section entitled ``tools'' will also be available, and will
 include information on the entries’ structure, tables with the contents of different categories, links to Vocabularies
@@
 
-4. The AXON Project for teaching and academic research
+\section{The AXON Project for teaching and academic research}
 
-\bigskip
-
-4.1. Teaching
-
-\bigskip
+\subsection{Teaching}
 
@@
 
-\bigskip
-
-\bigskip
-
-4.2. Significance for the academic community
-
-\bigskip
+\subsection{Significance for the academic community}
 

### 2015-12-17 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
paper 22 updated
doc to tex conversion first step

@@
 %TITLE:%
-%%%%%%%%\title{}\author{MARTA}
+%%%%%%%%
+
+\title{The Digital Edition of the Archaic Latin Inscriptions (7th{}-5th century B.C.)}
+\author[unimi]{Giovanna Rocca \corref{first}}
+\author[kore]{Giulia Sarullo}
+\author[unimi]{Marta Muscariello}
+\address[kore]{Università degli Studi di Enna "Kore"}
+\address[unimi]{Università IULM Milano}
 \date{2015-09-18}
 \begin{document}
-The Digital Edition of the Archaic Latin Inscriptions
 
-(7th{}-5th century B.C.)
-
-Giovanna Rocca – Giulia Sarullo – Marta Muscariello\footnote{\par \ \ Giovanna Rocca (GR) is responsible for § 1 and 2;
-Giulia Sarullo (GS) is responsible for § 4 and Marta Muscariello (MM) for § 3.}
-
-Abstract
-
+\maketitle
+\begin{abstract}
 This project consists in the digital edition of the archaic Latin inscriptions (7th – 5th century B.C.) according to the
@@
 this led to new discoveries and confirmed old hypotheses. Each text will be presented in an epigraphic chart, enriched
-by photos and illustrations.
+by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for § 1 and 2;
+Giulia Sarullo (GS) is responsible for § 4 and Marta Muscariello (MM) for § 3.}
+\end{abstract}
+\keywords{Archaic Latin Inscriptions, Latin Epigraphy, EpiDoc, Digital Humanities, Epigraphic Edition, Forum inscription.}
 
-\bigskip
-
-Keywords
-
-Archaic Latin Inscriptions, Latin Epigraphy, EpiDoc, Digital Humanities, Epigraphic Edition, Forum inscription.
-
-\bigskip
-
-\bigskip
-
-1. Generalia [GR]
+\section{Generalia [GR]}
 
 The project Iscrizioni Latine Arcaiche (ILA) consists in the digital edition of the inscriptions found in old
-Latium\footnote{\par \ \  With the two exceptions of the Vendia’s Urn, found in Cerveteri, and the Garigliano bowl.}
+Latium\footnote{With the two exceptions of the Vendia's Urn, found in Cerveteri, and the Garigliano bowl.}
 dating back to the period between the 7th and the 5th century B.C. Between the end of the 19th century and the first
@@
 about each text (images, bibliographical references, etc.) immediately and consider the inscription in the context of
-its place of finding; at the same time, it encourages a direct [201B?]in real time’ comparison between texts. Moreover,
+its place of finding; at the same time, it encourages a direct `in real time' comparison between texts. Moreover,
 this kind of structure allows us to obviate the inconvenience of an edition that presents only a specifically
-[201B?]historical’, [201B?]archaeological’ or [201B?]epigraphic’ point of view, as sometimes happens in \ traditional
+`historical', `archaeological' or `epigraphic' point of view, as sometimes happens in traditional
 editions. This is achieved thanks to cross-references to topics analyzed in depth by specialists, thus making our
@@
 present particular features that allow us to investigate the various steps that led to “standard Latin”. The language
-attested in our inscriptions can be considered a Restsprache insofar as it is not [201B?]readable’ through later Latin
-but it can be [201B?]interpreted’. The linguistic commentary will be carried out, in votis, in the second phase of this
+attested in our inscriptions can be considered a Restsprache insofar as it is not `readable' through later Latin
+but it can be `interpreted'. The linguistic commentary will be carried out, in votis, in the second phase of this
 project. Here the research focuses on the epigraphic features, that show a plurality of forms and of alphabets in such
@@
 
-2. Epigraphic News about the Forum Inscription [GR]
+\section{Epigraphic News about the Forum Inscription [GR]}
 
 Thanks to the agreement and collaboration with public authorities, we were invited to take part in an extraordinary
-event (July 3rd 2015), that is the 3D laser scanning of the Forum Cippus (CIL I2, 1)\footnote{Cippus of tuff found
+event (July 3rd 2015), that is the 3D laser scanning of the Forum Cippus (CIL I2, 1).\footnote{Cippus of tuff found
 under the Lapis Niger by Giacomo Boni in 1899. The inscription is cut boustrophedon on the four faces of the pillar and
 on the edge sliced between face D and face A. The stone (late 6th B.C.) is badly damaged in the upper part so that the
-beginnings and the endings of the lines are lost. The text has been interpreted as a sacred law.}. The autopsy, carried
+beginnings and the endings of the lines are lost. The text has been interpreted as a sacred law.} The autopsy, carried
 out with the help of a strong source of light, and the observation of the scanning in real time clarified, hopefully in
@@
 deep and more worn-out, after esed\footnote{The details of significant portions of the inscriptions will be shown on
-the website with several photographs. } (Fig. 1.1). This fact was highlighted by Gamurrini (1899) and the three dots
-appear in the apograph in Hülsen 1899, col. 1003 and they could also be guessed in the photograph Anderson 3192
-(Archivio Alinari). Nevertheless, the post Goidanich (1943) vulgata and the difficulties in checking in person the
+the website with several photographs. } (Fig.~\ref{fig:1}). This fact was highlighted by \citet{gamurrini_paleografia_1899} and the three dots
+appear in the apograph in \citet[col. 1003]{hulsen_neue_1899},  and they could also be guessed in the photograph Anderson 3192
+(Archivio Alinari). Nevertheless, the post \citet{goidanich_rapporti_1943} vulgata and the difficulties in checking in person the
 stone, because of its almost unreachable and scarcely illuminated position that led to the publication of studies not
 supported by an autoptical check, perpetuated a reading influenced by the uncertainty in distinguishing between natural
-cavities or notches due to the nature of the stone and [201B?]significant’ holes caused by the tool used for writing.
+cavities or notches due to the nature of the stone and `significant' holes caused by the tool used for writing.
 
-Fig. 1.1. Forum Cippus, Face A and Face E (by Marta Muscariello).
-
-\begin{figure}
+\begin{figure}[!bp]
 \centering
-\includegraphics[width=4.837cm,height=7.225cm]{IscrizioniLatineArcaicheEAGLE2016-img001.png}
+ \includegraphics[width=\columnwidth]{IscrizioniLatineArcaicheEAGLE2016-img001.png}
+\caption{Forum Cippus, Face A and Face E (by Marta Muscariello).}
+\label{fig:1}
 \end{figure}
 
-\bigskip
-
 The most relevant news arrives from the analysis of line 16 (face E). This has been read up to now as loiuquod()qo
-(Wachter 1987), loi\{u\}quiod (Vine 1993), LOIUQUIOD[QO] (Baldi 1999), LOIUQUIOD,QO/// (Hartmann 2005)\footnote{The
-final QO is based only on Goidanich 1943.}, in order to explain the [201B?]unusual’ shape of V (letter no. 4) that has
-been recently read as F (Prosdocimi 2010). As the scanning showed, the first vertical stroke, read as an I, is much
+\citep{wachter_altlateinische_1987}, loi\{u\}quiod \citep{_studies_1993}, LOIUQUIOD[QO] \citep{baldi_foundations_2002}, LOIUQUIOD,QO/// \citep{hartmann_fruhlateinischen_2005},\footnote{The
+final QO is based only on \citet{goidanich_rapporti_1943}.} in order to explain the `unusual' shape of V (letter no. 4) that has
+been recently read as F \citep{prosdocimi_roma_2010}. As the scanning showed, the first vertical stroke, read as an I, is much
 closer to the sign that looks like a V than it appears on the apographs and the photographs published so far;
@@
 This point can result fundamental in understanding how the inscriber worked on the stone and how he corrected the
-sign\footnote{Other corrections can be found on the last line of face C (kapia on kapa) and on the second line of face
-D where a V was corrected into a koppa. }. Hypothesis no. 1 (which is also the simplest one): the sequence to write was
+sign.\footnote{Other corrections can be found on the last line of face C (kapia on kapa) and on the second line of face
+D where a V was corrected into a koppa. } Hypothesis no. 1 (which is also the simplest one): the sequence to write was
 LOIVQVIOD but, after he cut a first vertical stroke, the inscriber mistakenly started to cut the oblique stroke of a V
@@
 
-3. The Epigraphic Chart [MM]
+\section{The Epigraphic Chart [MM]}
 
@@
 the study of the Forum Inscription; in the past, the collection of all the fragmentary instrumental Latin inscriptions
-up to the 4th century B.C. published by G. Colonna in 1980\footnote{In Stibbe et al. 1980.} supplied, at least
+up to the 4th century B.C. published by G. Colonna in 1980\footnote{In \citet{stibbe_lapis_1980}.} supplied, at least
 partially, the extent of the alphabetization developing in Latium Vetus, subtracting the major inscriptions from a sort
@@
 fundamental, since these are different in languages and alphabets: besides the Etruscan examples in Rome, we can
-remember the case of Satricum, where both Latin and Volscan are attested\footnote{See Rocca 1997, 189-198.}, or the
+remember the case of Satricum, where both Latin and Volscan are attested,\footnote{See \citet[189-198]{rocca_i_1995}.} or the
 Garigliano bowl which bears, together with the Latin inscription engraved inside, an inscription in Italic alphabet and
-language on the external body of the vase\footnote{On the relationship between the two inscriptions, see Antonini
-2012.}, that has also given a hint for a particular interpretation of the Latin text.
+language on the external body of the vase,\footnote{On the relationship between the two inscriptions, see \citet{antonini_osservazioni_2012}.} that has also given a hint for a particular interpretation of the Latin text.
 
@@
 attributed to the object, stratified in time in literature and well known to the scholars (for example “Duenos vase” /
-“Vase of the Quirinal”). We then indicate here, in the item [201B?]Present collocation’, where the inscription is
+“Vase of the Quirinal”). We then indicate here, in the item `Present collocation', where the inscription is
 preserved with, when possible, the inventory number.
 
-The chart continues with a group of three items that form the section dedicated to the [201B?]Archaeological data’: the
-‘Description of the object’ (with the general type, the possible peculiar features, the function and the dimensions of
-the inscribed object); the ‘Provenance’ (that is the place where the object was found and its exact archaeological
-context\footnote{In the case of mobile objects a different place of fabrication can be presumed, as in the case of the
-Vendia’s Urn, found in Cerveteri, but considered by some to have been fabricated somewhere else in Latium.}); the
-‘Date’, which quotes the hypotheses given by scholars about the chronological coordinates of the inscription.
+The chart continues with a group of three items that form the section dedicated to the `Archaeological data': the
+‘Description of the object' (with the general type, the possible peculiar features, the function and the dimensions of
+the inscribed object); the ‘Provenance' (that is the place where the object was found and its exact archaeological
+context);\footnote{In the case of mobile objects a different place of fabrication can be presumed, as in the case of the
+Vendia's Urn, found in Cerveteri, but considered by some to have been fabricated somewhere else in Latium.} the
+‘Date', which quotes the hypotheses given by scholars about the chronological coordinates of the inscription.
 Concerning this item, we must keep in mind that the dating of the antiquissimae is often approximate, and it is based
@@
 dating the objects and the rarity of the findings is surely connected to the difficulty in defining the specific
-features of the language testified by these inscriptions\footnote{On the periodization of Latin, P. Cuzzolin and G.
+features of the language testified by these inscriptions.\footnote{On the periodization of Latin, P. Cuzzolin and G.
 Haverling state: “The division of the history of a language into different periods implies that we have a rather clear
@@
 is to determine what is Latin and what is not: unfortunately there is no overall agreement on whether all of the early
-inscriptions considered to provide us with early examples of Latin actually do that.” (Cuzzolin, \ \ Haverling 2009,
-20).}.
+inscriptions considered to provide us with early examples of Latin actually do that.” \citep[20]{cuzzolin_syntax_2009}.}
 
 The charts present several photographs of the inscriptions, taken during the photographic campaign carried out by the
-project’s team. Enlargements of some useful or problematic details of the inscription have been added: the richness in
+project's team. Enlargements of some useful or problematic details of the inscription have been added: the richness in
 images is related to the participation of the project in the Europeana network, a database of images of the European
@@
 
-We offer two kinds of transcription, the ‘Diplomatic transcription’ and the ‘Interpretative transcription’, which
+We offer two kinds of transcription, the ‘Diplomatic transcription' and the ‘Interpretative transcription', which
 provides the edition of the text. The need for a diplomatic transcription is due to the problematic nature of archaic
@@
 
-In the item [201B?]Textual typology’ inscriptions are classified according to the nature of the text, taking into due
+In the item `Textual typology' inscriptions are classified according to the nature of the text, taking into due
 consideration the peculiarities of this corpus. In the chronological span between the 7th and the 5th century B.C. the
@@
 branches spoken in ancient Italy (with the exception of Greek); for this reason, from a classificatory point of view an
-inscription can be ‘anomalous’ in two ways: on the one hand, it can lack the typical elements of a formulaic scheme
+inscription can be ‘anomalous' in two ways: on the one hand, it can lack the typical elements of a formulaic scheme
 that will be fixed later on, thus requiring a further interpretative effort in order to assign it to a specific textual
@@
 related to more than one textual typology: in this case, the object type and the archaeological context are determinant
-for the overall classification\footnote{On this subject, see two recent publication, Poccetti 2009 and Maras 2015.}.
+for the overall classification.\footnote{On this subject, see two recent publication, \citet{poccetti_paradigmi_2009} and \citet{maras_storie_2015}.}
 
-A further group of items composes the ‘Epigraphic data’ section: the analysis begins from the ‘Position of the
-inscription’ on the object, which indicates the relationship between the text and the inscribed object, an aspect that
-has important consequences on the function and the fruition of the inscription\footnote{An important methodological
-point was established by Susini 1982.}. We then have ‘Scriptura’, where the execution technique is described;
-‘Direction of writing’, i.e. right-to-left, left-to-right, boustrophedon, etc.; ‘Dividing signs’, in which the presence
-of punctuation and its eventual function is signaled; ‘Dimensions of the letters’, which is important as regards the
+A further group of items composes the ‘Epigraphic data' section: the analysis begins from the ‘Position of the
+inscription' on the object, which indicates the relationship between the text and the inscribed object, an aspect that
+has important consequences on the function and the fruition of the inscription.\footnote{An important methodological
+point was established by \citet{susini_epigrafia_1982}.} We then have ‘Scriptura', where the execution technique is described;
+‘Direction of writing', i.e. right-to-left, left-to-right, boustrophedon, etc.; ‘Dividing signs', in which the presence
+of punctuation and its eventual function is signaled; ‘Dimensions of the letters', which is important as regards the
 visibility of the inscription in relation with the object and the observer. This section ends with the ‘Epigraphic
-commentary’, containing the description and analysis of the letters one by one both from the formal (shape-model of the
+commentary', containing the description and analysis of the letters one by one both from the formal (shape-model of the
 letter) and the factual point of view (possible particular features in the execution of the inscription) and some
@@
 
-The ‘Notes and issues’ item gathers historical-bibliographical notes, observations and discussions on the most
+The ‘Notes and issues' item gathers historical-bibliographical notes, observations and discussions on the most
 problematic points of each inscription: the insertion of the discussion at the end of the chart offers the advantage of
@@
 
-The chart is closed by the ‘Bibliography’ section. The ‘Editio princeps’ and the possible ‘First notice’ (if the
+The chart is closed by the ‘Bibliography' section. The ‘Editio princeps' and the possible ‘First notice' (if the
 inscribed object had been mentioned in a publication preceding the first edition of the text) are indicated in two
@@
 
-4. Technical notes [GS]
+\section{Technical notes [GS]}
 
@@
 tackle. Unlike EDR that, as the other projects constituting the Electronic Archive of Greek and Latin Epigraphy
-(EAGLE)\footnote{http://www.eagle-eagle.it/. The other databases related to EAGLE are the Epigraphic Database Bari
-(EDB), http://www.edb.uniba.it/, the Epigraphische Datenbank Heidelberg (EDH),
-http://www.uni-heidelberg.de/institute/sonst/adw/edh/index.html and Hispania Epigraphica (HE),
-http://www.eda-bea.es/.}, is a database, the archaic Latin inscriptions have been digitalized according to the EpiDoc
-Guidelines\footnote{For further information on EpiDoc and its history, see Elliott, Bodard, Cayless et al. 2006-2013.
-The guidelines are available at http://www.stoa.org/epidoc/gl/latest/.}. This is a set of specifications and encoding
+(EAGLE),\footnote{\url{http://www.eagle-eagle.it/}. The other databases related to EAGLE are the Epigraphic Database Bari
+(EDB), \url{http://www.edb.uniba.it/}, the Epigraphische Datenbank Heidelberg (EDH),
+\url{http://www.uni-heidelberg.de/institute/sonst/adw/edh/index.html} and Hispania Epigraphica (HE),
+\url{http://www.eda-bea.es/}.} is a database, the archaic Latin inscriptions have been digitalized according to the EpiDoc
+Guidelines.\footnote{For further information on EpiDoc and its history, see \citet{elliott_epidoc_2007}.
+The guidelines are available at http://www.stoa.org/epidoc/gl/latest/.} This is a set of specifications and encoding
 tools in XML (eXtensible Markup Language) for the scientific edition of ancient texts based on the Text Encoding
 Initiative (TEI), a set of XML specifications designed for the digital publication of texts and manuscripts for
-research purposes\footnote{http://www.tei-c.org/index.xml. See also Burnard, Baumann 2014 (About these Guidelines):
+research purposes.\footnote{http://www.tei-c.org/index.xml. See also \citet{consortium_tei_2008} (About these Guidelines):
 “The TEI encoding scheme is of particular usefulness in facilitating the loss-free interchange of data amongst
-individuals and research groups using different programs, computer systems, or application software”.}. EpiDoc is
+individuals and research groups using different programs, computer systems, or application software”.} EpiDoc is
 becoming more and more a point of reference for digital epigraphic projects\footnote{On the relationship between EAGLE
-projects and EpiDoc see Felle 2012.} and it is also the standard chosen for the aggregation of the archives’ data in
+projects and EpiDoc see \citet{felle_esperienze_2012}.} and it is also the standard chosen for the aggregation of the archives' data in
 the recently constituted network, again called EAGLE (Europeana Network of Ancient Greek and Latin
-Epigraphy)\footnote{http://www.eagle-network.eu. About the new Best Practice Network, co-founded by the European
-Commission, see http://www.europeana.eu.}, in which, besides the data of the EAGLE archives, the archaic Latin
+Epigraphy),\footnote{http://www.eagle-network.eu. About the new Best Practice Network, co-founded by the European
+Commission, see http://www.europeana.eu.} in which, besides the data of the EAGLE archives, the archaic Latin
 inscriptions corpus will also converge. As a matter of fact, inscriptions encoded with EpiDoc are not only compatible
@@
 the information and not to its appearance, it is possible to modify the look of the final result by simply operating on
-the style sheet\footnote{In XML, all information related to the formatting of the text are registered on a separate
-file called style sheet, see Bodard 2010, 104, 110-111.}, not having to revise the single files. This will facilitate
+the style sheet,\footnote{In XML, all information related to the formatting of the text are registered on a separate
+file called style sheet, see \citet[104, 110-111]{bodard_epidoc:_2009}.} not having to revise the single files. This will facilitate
 the integration of the archaic Latin inscriptions into wider digital collections such as EDR and EAGLE-Europeana.
 Moreover, since the file thus encoded can also be translated into another markup language, their survival despite any
-future technical evolution is guaranteed\footnote{Tissoni 2008, 37-38 and Bodard 2010, 104-105.}.
+future technical evolution is guaranteed.\footnote{\citet[37-38]{tissoni_epidoc_2008} and \citet[104-105]{bodard_epidoc:_2009}.}
 
@@
 produce a digital edition of the text that complies to Leiden Conventions\footnote{On Leiden Conventions, the standard
-used to annotate epigraphic documents and papyri in printed editions, see Krummrey, Panciera 1980, Panciera 2006a and
-2006b. About their use in EpiDoc’s files see Elliot 2007, Mahoney 2006, 229 and Bodard 2010, 105.} that will show the
+used to annotate epigraphic documents and papyri in printed editions, see \citet{krummrey_criteri_1980}, \citet{panciera_segni_2006} and \citet{panciera_i_2006}.
+About their use in EpiDoc's files see \citet{elliot_conformance_2007}, \citet[229]{burnard_electronic_2006} and \citet[105]{bodard_epidoc:_2009}.} that will show the
 same typographical marks a printed edition following the Leiden system would have, thus being immediately
@@
 epigraphic documents, it has been necessary to adapt these Guidelines to respond to the peculiar issues of this
-corpus\footnote{See Sarullo 2011, 162-167, where a few examples of markup are quoted. }. For this reason, some elements
+corpus.\footnote{See \citet[162-167]{sarullo_ledizione_2011}, where a few examples of markup are quoted. } For this reason, some elements
 have been adapted and others have been specially designed, and this was possible thanks to the fact that XML is an
@@
 rend={\textquotedbl}down-to-up{\textquotedbl}{\textgreater} elements in order to render the peculiar directions in
-which the text was cut\footnote{The issue of the direction of writing was the subject of much debate at the 6th EAGLE
+which the text was cut.\footnote{The issue of the direction of writing was the subject of much debate at the 6th EAGLE
 International Event Off the beaten track. Epigraphy at the borders (Bari, September 24th{}-25th, 2015). The discussion
@@
 antiquity and the necessity to establish a common standard to encode the instances of “non-standard” directions of
-writing emerged. For this reason, this issue will surely be the object of further consideration in the future.}.
+writing emerged. For this reason, this issue will surely be the object of further consideration in the future.}
 Reversed and upside-down letters are usually left unmarked in traditional epigraphic editions and we decided to comply
@@
 rend={\textquotedbl}reversed{\textquotedbl}{\textgreater} was used, a tag that in the EpiDoc Guidelines is used to
-encode the litterae inversae, enclosed in double round parentheses, such as in ((C)) for mulier\footnote{See Panciera
-2006b, 1722.}. For upside-down letters, a new value was provided, {\textless}hi
+encode the litterae inversae, enclosed in double round parentheses, such as in ((C)) for mulier.\footnote{See \citet[1722]{panciera_i_2006}.} For upside-down letters, a new value was provided, {\textless}hi
 rend={\textquotedbl}upside-down{\textquotedbl}{\textgreater}, since none of the allowed values of @rend for the
@@
 
+\nocite{colonna_iscrizioni_2007}
 \bibliographystyle{sapauth-eng}

### 2015-12-17 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
updated paper 22, updated paper 24
paper 24 doc to tex
added email in paper 22

--- 2015-12-14 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
@@
 %%%%%%%%
-\title{}
-\author{Noemí Moncunill Martí}
+\title{Epigraphy and onomastics in the Hesperia databank}
+\author[barc]{Noemí Moncunill Martí \corref{first}}
+\author[sorb]{Javier Velaza}
+\address[sorb]{Université de Paris-Sorbonne (Paris IV)}
+\address[barc]{University of Barcelona2}
+\cortext[first]{Corresponding author. Email: nmoncunill@gmail.com}
 \date{2015-11-15}
 \begin{document}
-Abstract
 
-\begin{figure}
-\centering
-\begin{minipage}{10.901cm}
-1.\ \ Epigraphy and onomastics in the Hesperia databank\newline
-\newline
-
-Noemí Moncunill1, Javier Velaza2
-
-Université de Paris-Sorbonne (Paris IV)1, University of Barcelona2
-
-Email(s): velaza@ub.edu
-
-nmoncunill@gmail.com
-
-velaza@ub.edu
-
-\bigskip
-\end{minipage}
-\end{figure}
+\maketitle
+\begin{abstract}
 The first part of this work provides a general overview of the features and advantages of the digital epigraphic corpora
@@
 names and divinity names.
+\end{abstract}
+\keywords{Hispania, epigraphy, onomastics, digital corpora, Palaeohispanic languages and writings}
 
-\bigskip
-
-Keywords
-
-Hispania, epigraphy, onomastics, digital corpora, Palaeohispanic languages and writings
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-1. Introduction.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
-(2014SGR63), on the one hand, and 655938 — ZEPHYROS — H2020-MSCA-IF-2014 project, on the other.}
+\section{Introduction}
 
 The aim of this work is to do a general overview of the computer-based epigraphic corpora, to think about their features
-and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.
+and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
+(2014SGR63), on the one hand, and 655938 — ZEPHYROS — H2020-MSCA-IF-2014 project, on the other.}
 
@@
 are imposed by their morphology and format. If summarizing, there are three main and most evident limitations: 
-
-a) One-dimensional format. The monument's description, the textual edition, the critical apparatus and remarks are
+\begin{description}
+\item[a] One-dimensional format. The monument's description, the textual edition, the critical apparatus and remarks are
 displayed one after another, with no wider possibility to link one to another than through some indices, which by the
@@
 
-b) Ephemeral usefulness. It looks also obvious that a traditional corpus will necessarily be always ephemeral for many
+\item[b] Ephemeral usefulness. It looks also obvious that a traditional corpus will necessarily be always ephemeral for many
 reasons: new inscriptions cannot be added, mistakes cannot be corrected, text editions cannot be improved and
@@
 
-c) Authoritative work. In a traditional corpus, editors impose their knowledge in a large range of features, and thus
+\item[c] Authoritative work. In a traditional corpus, editors impose their knowledge in a large range of features, and thus
 they interfere in the utility of other views, to the extend of depreciating them too much. When it comes to an
@@
 
+\end{description}
 In our opinion, these limitations can be solved or, at least, diminished by a proper and efficient use of the tools IT
@@
 
-2. The Hesperia project
+\section{The Hesperia project}
 
@@
 linguistic evidences from Palaeohispanic languages, that is, pre-Roman languages in the Iberian Peninsula\footnote{
-General presentations of the project can be found in Luján - Orduña (forthcoming), Orduña - Luján (forthcoming) and
-Velaza 2014.}. First of all, we need to recall these are very different materials as for their quantity, quality and
+General presentations of the project can be found in \citet{orduna_implementing_????}, \citet{orduna_philology_????} and
+\citet{velaza_hesperia:_2014}.}. First of all, we need to recall these are very different materials as for their quantity, quality and
 reliability. Ancient Hispania left us epigraphic texts in at least four languages: Tartessian, Iberian, Celtiberian and
 Lusitanian -{}-the possibility there are also some texts in Palaeobasque language is still sub iudice\footnote{
-Concerning this subject, see Velaza 2009.}{}-{}-, mainly written in four epichoric scripts -{}-Tartessian,
+Concerning this subject, see \citet{velaza_epigrafiy_2009}.}{}-{}-, mainly written in four epichoric scripts -{}-Tartessian,
 southeastern-Iberian, northeastern-Iberian and Greek-Iberian script.
 
- \includegraphics[width=10.987cm,height=7.835cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img001.jpg}
-
-Fig. 1: Map of the Palaeohispanic inscriptions attested in the Iberian Peninsula.
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img001.jpg}
+\caption{ Map of the Palaeohispanic inscriptions attested in the Iberian Peninsula.}
+\label{fig:1}
+\end{figure}
 
@@
 del País Vasco, the Universidad de Zaragoza and the Universidad de Barcelona. There are currently several sections open
-for searching (http://hesperia.ucm.es/).
+for searching (\url{http://hesperia.ucm.es/}).
 
- \includegraphics[width=10.896cm,height=7.669cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img002.png}
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img002.png}
+\caption{Dispersal areas of Palaeohispanic inscriptions and sections available in Hesperia databank}
+\label{fig:2}
+\end{figure}
 
-Fig. 2. Dispersal areas of Palaeohispanic inscriptions and sections available in Hesperia databank
-
-i) Partial epigraphic corpora from B zone (Iberian inscriptions from the south of France), K zone (Celtiberian
+\begin{description}
+\item[i] Partial epigraphic corpora from B zone (Iberian inscriptions from the south of France), K zone (Celtiberian
 inscriptions) and L zone (Lusitanian inscriptions). All of them follow the same data form made of six tabs
-corresponding to {\textquotedbl}general features{\textquotedbl}, {\textquotedbl}text{\textquotedbl},
-{\textquotedbl}epigraphy and paleography{\textquotedbl}, {\textquotedbl}pictures{\textquotedbl},
-{\textquotedbl}arqueological context{\textquotedbl} and {\textquotedbl}bibliography{\textquotedbl}. The databases are
+corresponding to ``general features'', ``text'',
+``epigraphy and paleography'', ``pictures'',
+``arqueological context'' and ``bibliography''. The databases are
 linked between them through a powerful engine for simple and complex searches, and pdf documents as well as maps can be
@@
 
-ii) A numismatic database with its specific four-tab structure: {\textquotedbl}general features{\textquotedbl},
-{\textquotedbl}inscription{\textquotedbl}, {\textquotedbl}language and writing{\textquotedbl} and
-{\textquotedbl}bibliography{\textquotedbl}, where all Paleohispanic mints are gathered, no matter their scripts or
+\item[ii] A numismatic database with its specific four-tab structure: ``general features'',
+``inscription'', ``language and writing'' and
+``bibliography'', where all Paleohispanic mints are gathered, no matter their scripts or
 languages.
 
-iii) An onomastic database, with a specific search engine allowing combined searches with any of the elements typical
+\item[iii] An onomastic database, with a specific search engine allowing combined searches with any of the elements typical
 from the identified onomastic formulae.
+
+\end{description}
 
@@
 
-3. The new sections on onomastics.
+\section{The new sections on onomastics}
 
@@
 
- \includegraphics[width=11.749cm,height=6.712cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003.png}
-
-Fig. 3. Personal-names and divinity-names record in Hesperia databank.
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003.png}
+\caption{Personal-names and divinity-names record in Hesperia databank.}
+\label{fig:3}
+\end{figure}
 
@@
 
- \includegraphics[width=11.883cm,height=5.66cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img004.png}
-
-Fig. 4. Personal-names and divinity-names record in Hesperia databank.
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img004.png}
+\caption{Personal-names and divinity-names record in Hesperia databank.}
+\label{fig:4}
+\end{figure}
 
@@
 
- \includegraphics[width=7.031cm,height=5.359cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img005.png}
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img005.png}
+\caption{Dispersal area of indigenous personal names.}
+\label{fig:5}
+\end{figure}
 
-Fig. 5: Dispersal area of indigenous personal names.
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img006.png}
+\caption{Dispersal area of indigenous divinity names}
+\label{fig:}
+\end{figure}
 
- \includegraphics[width=6.98cm,height=5.262cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img006.png}
-
-Fig. 6: Dispersal area of indigenous divinity names
-
- \includegraphics[width=6.32cm,height=6.045cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img007.png}
-
-Fig. 7: Disperal areas of indigenous divinity names and personal names
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img007.png}
+\caption{Disperal areas of indigenous divinity names and personal names}
+\label{fig:7}
+\end{figure}
 
@@
 
- \includegraphics[width=6.491cm,height=4.833cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img008.jpg}
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img008.jpg}
+\caption{Dispersal area of the anthroponymic element biur}
+\label{fig:}
+\end{figure}
 
-Fig. 8: Dispersal area of the anthroponymic element biur.
-
- \includegraphics[width=6.092cm,height=4.583cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img009.jpg}
-
-Fig. 9: Dispersal area of the name Tancinus.
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img009.jpg}
+\caption{Dispersal area of the name Tancinuse}
+\label{fig:9}
+\end{figure}
 
@@
 
- \includegraphics[width=10.862cm,height=6.325cm]{EpigraphyandonomasticsinHesperiadatabanktemplate-img010.png}
-
-Fig. 10: record of a possible Greek name attested in an Iberian inscription.
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img010.png}
+\caption{Fig. 10: record of a possible Greek name attested in an Iberian inscription.}
+\label{fig:10}
+\end{figure}
 
@@
 
+CTS = Canonical Text Services (protocol to cite digitally-available texts in a canonical way).
+
+EAGLE = The Europeana network of Ancient Greek and Latin Epigraphy:  \url{http://www.eagle-network.eu/}
+
+IDEs = Integrating Digital Epigraphies project.
+
+LGPN = The Lexicon of Greek Personal Names: \url{http://www.lgpn.ox.ac.uk/}
+
+Pleiades = A community-built gazetteer and graph of ancient places: \url{http://pleiades.stoa.org/}
+
+SNAP = Standards for Networking Ancient Prosopographies project: \url{http://snapdrgn.net/}
+
+Trismegistos = Interdisciplinary portal of papyrological and epigraphical resources formerly Egypt and the Nile valley
+(800 BC-AD 800): \url{http://www.trismegistos.org/}
+
+\nocite{moncunill_ochocientos_????}
+\nocite{untermann_monumenta_1975}
+\nocite{vallejo_antroponimia_????}
+
 \bibliographystyle{sapauth-eng}

### 2015-12-17 source/29/EAGLE2016_camera_ready.tex
paper 29 updated
doc to tex

@@
 %%%%%%%%
-\title{}
-\author{Angelos}
+\title{AUGMENTING THE WORKSPACE OF EPIGRAPHISTS
+An interaction design study}
+
+\author[1]{Angelos Barmpoutis\corref{first}}
+\author[2]{Eleni Bozia}
+\address[1]{Digital Worlds Institute, University of Florida}
+\address[2]{Department of Classics, University of Florida}
+\cortext[first]{Corresponding author. Email: angelos@digitalworlds.ufl.edu}
 \date{2015-11-21}
 \begin{document}
-Abstract
 
-\begin{figure}
-\centering
-\begin{minipage}{10.901cm}
-1.\ \ AUGMENTING THE WORKSPACE OF EPIGRAPHISTS\newline
-An interaction design study\newline
-
-Angelos Barmpoutis1, Eleni Bozia2
-
-Digital Worlds Institute, University of Florida1
-
-Department of Classics, University of Florida2
-
-Email(s): angelos@digitalworlds.ufl.edu, bozia@ufl.edu
-\end{minipage}
-\end{figure}
-\ \ \ \ \ \ This paper presents the results of an interaction design study that focuses on the use of natural user
+\maketitle
+\begin{abstract}
+This paper presents the results of an interaction design study that focuses on the use of natural user
 interfaces for professionals in the fields of epigraphy and archaeology. This study proposes solutions for utilizing
@@
 objects, and interact with metadata or other multi-modal data, such as text and images. 
+\end{abstract}
+\keywords{Mobile Applications, Interaction Design, Natural User Interfaces, 3D models, Archaeology}
 
-Keywords
-
-Mobile Applications, Interaction Design, Natural User Interfaces, 3D models, Archaeology
-
-1.1. Introduction
+\section{Introduction}
 
@@
 record tridimensional data either in the form of geo-spatial coordinates, or in the form of local 3D point coordinates
-needed for augmented-reality interaction, or for the construction of triangular meshes of 3D models. \
+needed for augmented-reality interaction, or for the construction of triangular meshes of 3D models.
 
-\ \ There are several examples in literature that present 3D digitization projects that have been undertaken by museums
-including the Epigraphical Museum of Athens [Sullivan 2011, Papadaki et al. 2015], Museo Arqueológico Nacional de
-Madrid [Ramírez-Sánchez et al. 2014], Museo Nazionale Romano di Palazzo Altemps [Barmpoutis et al. 2015], Museo
-Geologico Giovanni Capellini di Bologna [Abate \& Fanti 2014], National Museums Liverpool [Cooper et al. 2007],
-Smithsonian Institution [Wachowiak and Karas 2009], and several other museums and institutes [Gonizzi Barsanti and
-Guidi 2013, Landon and Seales 2006, Levoy et al. 2000].
+There are several examples in literature that present 3D digitization projects that have been undertaken by museums
+including the Epigraphical Museum of Athens (\citet{sullivan_analytical_2011}, \citet{papadaki_accurate_2015}), Museo Arqueológico Nacional de
+Madrid \citep{ramirez-sanchez_epigrafidigital:_2014}, Museo Nazionale Romano di Palazzo Altemps \citep{barmpoutis_interactive_2015}, Museo
+Geologico Giovanni Capellini di Bologna \citep{abate_valorizzazione_2014}, National Museums Liverpool \citep{cooper_chiswick_2007},
+Smithsonian Institution \citep{wachowiak_3d_2009}, and several other museums and institutes (\citet{barsanti_3d_2013}, \citet{landon_petroglyph_2006}, \citet{levoy_digital_2000}).
 
 Several novel methods for scanning, processing, and analyzing 3D models of inscriptions have been developed, including
-methods for text extraction from inscriptions [Aswatha et al. 2014, Sullivan 2011], accurate 3D scanning of
-inscriptions [Papadaki et al. 2015], visualization of inscriptions [Bozia et al. 2014], as well as 3D applications for
-other archaeological artifacts [Babeu 2011, Pollefeys et al. 2001, Malzbender et al. 2001, Esteban \& Schmitt 2004].
-Comparative studies of 3D scanning methods for cultural herirage can be found in [Pavlidis et al. 2007] and [Böhler \&
-Marbs 2004].
+methods for text extraction from inscriptions (\citet{aswatha_method_2014}, \citet{sullivan_analytical_2011}), accurate 3D scanning of
+inscriptions \citep{papadaki_accurate_2015}, visualization of inscriptions \citep{bozia_open-access_2014}, as well as 3D applications for
+other archaeological artifacts (\citet{babeu__2011}, \citet{pollefeys_image-based_2001}, \citet{malzbender_polynomial_2001}, \citet{esteban_silhouette_2004}).
+Comparative studies of 3D scanning methods for cultural herirage can be found in \citep{pavlidis_methods_2007} and \citep{bohler_3d_2004}.
 
-\ \ The aforementioned examples show that the use of 3D technologies in epigraphy and archaeology has been a
+The aforementioned examples show that the use of 3D technologies in epigraphy and archaeology has been a
 well-studied topic over the past two decades. However, there is a notable disconnect between the research on these
@@
 Popular low-cost NUIs, such as touch screens, marker-less position trackers, motion sensors, and head-mounted displays
-have been recently studied and employed by museums as mechanisms for multi-sensory virtual experiences [Ujitok \&
-Hirota 2015, Soile et al. 2013, Ikei et al. 2015].
+have been recently studied and employed by museums as mechanisms for multi-sensory virtual experiences (\citet{ujitoko_application_2015}, \citet{soile_accurate_2013}, \citet{ikei_experience_2015}).
 
@@
 
-1.2. Understanding the workflow of epigraphists
+\section{Understanding the workflow of epigraphists}
 
 Understanding the users is one of the integral steps of interaction design, which is an iterative process during which
-representative users interact with preliminary designs and provide useful feedback [Preece et al. 2015]. For the
+representative users interact with preliminary designs and provide useful feedback \citep{preece_interaction_2015}. For the
 purposes of this study, our team interacted with early adopters of our prototype system, who were epigraphists and
@@
 goals of our interaction were twofold: a) to study the various forms of physical interaction that epigraphists have
-with an inscription as a real physical object \ and b) to expose epigraphists to a digital interface that imitates
+with an inscription as a real physical object and b) to expose epigraphists to a digital interface that imitates
 their interaction routine, using digital replicas of physical objects.
@@
 \item Change of point of view: Observation of the inscription from different viewing angles assists epigraphists
-understand better the shape of the inscribed letterforms. \ 
+understand better the shape of the inscribed letterforms. 
 \item Change of lighting conditions: Relighting the inscription by introducing artificial shadows or additional light
@@
 special cases of I and II. More specifically, the physical object can be either portable (such as a small fragment of
-stone or other material) or not (when the inscription is on an inscription bearer). \ In the case of a handheld object,
+stone or other material) or not (when the inscription is on an inscription bearer). In the case of a handheld object,
 interactions I and II involve manual movement of the inscription with respect to the fixed observer (case I) or the
 fixed light source (case II), while in the case of large rigid objects the observer and the light source move with
-respect to the fixed inscription. \
+respect to the fixed inscription.
 
 According to the above analysis, in the case of digitized inscriptions a NUI should provide the means for an epigraphist
-to {\textquotedbl}hold{\textquotedbl} the virtual object, {\textquotedbl}move{\textquotedbl} the point of view with
-respect to the virtual object, {\textquotedbl}manipulate{\textquotedbl} the virtual object with respect to the virtual
-light source, and {\textquotedbl}focus{\textquotedbl} on details of interest. The next section presents a NUI-based
+to ``hold'' the virtual object, ``move'' the point of view with
+respect to the virtual object, ``manipulate'' the virtual object with respect to the virtual
+light source, and ``focus'' on details of interest. The next section presents a NUI-based
 interaction design that proposes natural solutions to the aforementioned forms of interaction that seamlessly imitate
-the typical workflow of epigraphists. \ \ \
+the typical workflow of epigraphists.
 
-1.3. Natural User Interface design for epigraphy
+\section{Natural User Interface design for epigraphy}
 
@@
 
-\ An optimal interaction design solution should be intuitive, minimalistic, and non intrusive [Preece et al. 2015].
+\ An optimal interaction design solution should be intuitive, minimalistic, and non intrusive \citep{preece_interaction_2015}.
 Therefore, in order to design interaction for epigraphy one should choose devices that are easily accessible by
@@
 handheld, without loss of generality, in order to generate a multi-sensory experience for the user (i.e. holding the
-device = holding the digital inscription). \ \ Hence, NUI design is possible by utilizing accessible devices and
+device = holding the digital inscription). Hence, NUI design is possible by utilizing accessible devices and
 without the use of external sensors as it is described in details in the following sections.
 
-\bigskip
-
-1.3.1. Natural interactive relighting of 3D models
+\subsection{Natural interactive relighting of 3D models}
 
@@
 with respect to the gravitational vector, is enough to track the slope of the device with respect to the virtual light.
-The top row of Fig. 1.1 shows the approximated real-world orientation of a tablet computer as it was estimated using
+The top row of Fig.~\ref{fig:1} shows the approximated real-world orientation of a tablet computer as it was estimated using
 the gyroscope of the device. The orientation is updated in real-time as the user moves the device.
 
- \includegraphics[width=10.881cm,height=2.72cm]{EAGLE2016cameraready-img001.png} 
-\includegraphics[width=10.881cm,height=2.72cm]{EAGLE2016cameraready-img002.png}
-
-Fig. 1.1. Top row: Illustration of interactive manipulation of the virtual lighting by moving the device. The figures
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img001.png}
+ \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img002.png}
+\caption{ Top row: Illustration of interactive manipulation of the virtual lighting by moving the device. The figures
 show the corresponding field of normal vectors in the 3D space as computed using the gyroscope of the device. Bottom
 row: Demonstration of interactive relighting of a 3D digitized inscription. Different virtual lighting angles reveal
-different inscribed details. \ \ \
+different inscribed details. }
+\label{fig:1}
+\end{figure}
 
 The estimated orientation of the device can be used in order to relight the depicted 3D model of inscription using the
-angle of the device with the direction of the virtual light source. The bottom row of Fig. 1.1 demonstrates interactive
+angle of the device with the direction of the virtual light source. The bottom row of Fig.~\ref{fig:1} demonstrates interactive
 relighting of a digitized paper cast. Epigraphists can relight an inscription by reorienting the tablet as if it were a
@@
 inscriptions and can be intuitively extrapolated to the case of 3D models of large inscriptions that were not handheld
-in the real world (see Fig. 1.3).
+in the real world (see Fig.~\ref{fig:3}).
 
-\bigskip
-
-1.3.2. Natural interactive manipulation of 3D models
+\subsection{Natural interactive manipulation of 3D models }
 
@@
 
-Figure 1.2 shows 15 different projections of the same virtual cube that correspond to the change of perspective caused
+Fig.~\ref{fig:2} shows 15 different projections of the same virtual cube that correspond to the change of perspective caused
 by moving the observer's head parallel to the screen. Note that all cubes are parallel to each other.
 
-\bigskip
-
- \includegraphics[width=10.881cm,height=5.611cm]{EAGLE2016cameraready-img003.png}
-
-Fig. 1.2. Visualization of same-size boxes using 15 different perspective projections with the same FOV angle and
-different cropping parameters. None of the boxes is rotated in the space. \ \
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img003.png}
+\caption{Visualization of same-size boxes using 15 different perspective projections with the same FOV angle and
+different cropping parameters. None of the boxes is rotated in the space. }
+\label{fig:2}
+\end{figure}
 
@@
 eye-tracker should be required), the change of perspective can be realistically achieved by naturally reorienting the
-tablet as shown in the top row of Fig. 1.3. The superimposed boxes in this figure were estimated, using the
+tablet as shown in the top row of Fig.~\ref{fig:3}. The superimposed boxes in this figure were estimated, using the
 accelerometer reading of the device for three different orientations of the tablet.
 
-Interactive manipulation of a 3D digitized inscription bearer is shown in the bottom row of Fig. 1.3. Different sides of
+Interactive manipulation of a 3D digitized inscription bearer is shown in the bottom row of Fig.~\ref{fig:3}. Different sides of
 the bearer can be observed by reorienting the tablet naturally. In this example a large inscription model with its
@@
 scale. It should also be noted that the interactive manipulation of the perspective can be optionally performed
-simultaneously with the interactive relighting as shown in Fig. 1.1 (bottom) in order to perform a more realistic
-interaction that causes relighting and change of point of view at the same time. \ \ \
+simultaneously with the interactive relighting as shown in Fig.~\ref{fig:1} (bottom) in order to perform a more realistic
+interaction that causes relighting and change of point of view at the same time.
 
-\bigskip
-
- \includegraphics[width=10.881cm,height=2.72cm]{EAGLE2016cameraready-img004.png} 
-\includegraphics[width=10.881cm,height=2.72cm]{EAGLE2016cameraready-img005.png}
-
-Fig. 1.3. Top row: Illustration of interactive manipulation of the perspective projection by moving the device and using
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img004.png}
+ \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img005.png}
+\caption{Top row: Illustration of interactive manipulation of the perspective projection by moving the device and using
 the {\textquotedbl}fixed eye{\textquotedbl} assumption. Bottom row: Demonstration of interactive visual inspection of a
 3D digitized inscription along with the inscription bearer. The user can view the object from different perspectives,
-using natural motions. \
+using natural motions.}
+\label{fig:3}
+\end{figure}
 
-\bigskip
-
-1.3.3. {\textquotedbl}Touching{\textquotedbl} the metadata: Interacting with multi-modal data
+\subsection{``Touching'' the metadata: Interacting with multi-modal data}
 
@@
 modalities of information such as text, images, and metadata. The user can interactively browse the different forms of
-data by using intuitive touch gestures as shown in Fig. 1.4. This set of 2D interactions along with the 3D NUI design
+data by using intuitive touch gestures as shown in Fig.~\ref{fig:4}. This set of 2D interactions along with the 3D NUI design
 presented earlier can compose an intuitive yet powerful workspace for an epigraphist who can now perform digitally
-several parts of the epigraphic workflow. \
+several parts of the epigraphic workflow.
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img006.png}
+\caption{Screenshot of our interactive environment. The user interacts with the 3D object, using touch gestures and
+selects one of the regions of interests. This action initiates other data tools such as the image viewer or the edge
+filter as shown in this example}
+\label{fig:4}
+\end{figure}
 
- \includegraphics[width=10.881cm,height=5.754cm]{EAGLE2016cameraready-img006.png}
-
-Fig. 1.4. Screenshot of our interactive environment. The user interacts with the 3D object, using touch gestures and
-selects one of the regions of interests. This action initiates other data tools such as the image viewer or the edge
-filter as shown in this example.
-
-1.3. Conclusions and future directions
+\section{Conclusions and future directions}
 
@@
 perspective, using a set of intuitive gestures that imitate the natural interaction with a physical object. In the
-proposed design, the epigraphist can {\textquotedbl}hold{\textquotedbl} a digital inscription, relight it by
+proposed design, the epigraphist can ``hold'' a digital inscription, relight it by
 reorienting it as a tangible object, observe it from different perspectives, and finally interact with other modalities
@@
 
-\bigskip
+\section*{Acknowledgement}
 
-\begin{figure}
-\centering
-\begin{minipage}{10.901cm}
-Bibliography
-\end{minipage}
-\end{figure}
-
-\bigskip
-
-\begin{figure}
-\centering
-\begin{minipage}{10.901cm}
-Acknowledgement
-
-\bigskip
-\end{minipage}
-\end{figure}
 We would like to acknowledge our collaborators for their invaluable interaction and feedback during the past three

### 2015-12-17 source/31/EAGLE2016_I Sicily_final copy.tex
paper 31 updated
doc to tex. question about IIIP sent to Prag

@@
 %%%%%%%%
-\title{}
-\title{}
-\author{a}
+\title{I.Sicily. An EpiDoc corpus for ancient Sicily}
+\author[OXF]{Jonathan Prag\corref{first}}
+\author[opensky]{James Chartrand}
+\author[OXF]{James Cummings}
+\address[OXF]{University of Oxford}
+\address[opensky]{OpenSky Solutions}
+\cortext[first]{Corresponding author. Email: jonathan.prag@merton.ox.ac.uk}
 \date{2015-11-18}
 \begin{document}
-[Warning: Draw object ignored]Abstract
 
+\maketitle
+\begin{abstract}
 This paper introduces the EpiDoc corpus of inscriptions on stone for ancient Sicily, I.Sicily. The project is one of the
@@
 museum collections, mapping, and data conversion and integration, and these are briefly outlined in the paper.
+\end{abstract}
+\keywords{Sicily, Epigraphy, Epidoc, Greek, Latin, Pleiades, multilingualism}
 
-Keywords
-
-Sicily, Epigraphy, Epidoc, Greek, Latin, Pleiades, multilingualism
-
-\bigskip
-
-1.Introduction: What is I.Sicily
+\section{Introduction: What is I.Sicily}
 
 I.Sicily is an online, open access, digital corpus of the inscriptions on stone from ancient Sicily.\footnote{ The
-corpus will be mounted at www.csad.ox.ac.uk/sicily/isicily, but is presently on a development server. A public face is
-currently maintained via a blog at http://isicily.wordpress.com/, as well as on Facebook at www.facebook.com/ISicily
-and on Twitter via I.Sicily@Sicilyepigraphy.} The corpus aims to include all texts inscribed on stone, in any language,
+corpus will be mounted at \url{www.csad.ox.ac.uk/sicily/isicily}, but is presently on a development server. A public face is
+currently maintained via a blog at \url{http://isicily.wordpress.com/}, as well as on Facebook at \url{www.facebook.com/ISicily}
+and on Twitter via \url{I.Sicily@Sicilyepigraphy}.} The corpus aims to include all texts inscribed on stone, in any language,
 between approximately the seventh century BC and the seventh century AD. The corpus currently contains records for over
 2,500 texts, and when complete is likely to contain c.4,000. The corpus is built upon a conversion from a legacy
-dataset of metadata in MS Access to EpiDoc TEI XML.\footnote{ http://sourceforge.net/p/epidoc/wiki/Home/ [accessed
-26.09.2015].} \ The XML records are held in an eXist database for xQuery access, and additionally indexed for full-text
+dataset of metadata in MS Access to EpiDoc TEI XML.\footnote{\url{http://sourceforge.net/p/epidoc/wiki/Home/} [accessed
+26.09.2015].} The XML records are held in an eXist database for xQuery access, and additionally indexed for full-text
 search using SOLR/Lucene. The corpus and related information (museum list, bibliography) are published as Linked Data,
@@
 Pan, Rotate) image-viewing is provided by the IIIP image server and the OpenSeadragon javascript library.
+%IIIP or IIIF ? I believe this is a typo
 
@@
 
-2. The motivations for and origins of I.Sicily
+\section{The motivations for and origins of I.Sicily}
 
@@
 extensive use throughout antiquity of both Greek and Latin, as well as Oscan, Punic, Sikel, and Hebrew;\footnote{
-Recent overview of much of the linguistic tradition in Tribulato 2012; and of the epigraphic material in Gulletta
-1999.} on the other hand, the publication of this material has a very uneven record and despite an excellent
+Recent overview of much of the linguistic tradition in \citet{tribulato_language_2012}; and of the epigraphic material in \citet{gulletta_sicilia_1999}.} on the other hand, the publication of this material has a very uneven record and despite an excellent
 pre-twentieth century tradition, the existing copora are far from complete and the ability of key journals such as SEG
 or AE to keep pace with local publication has been limited.\footnote{ For an overview of the corpora tradition up to
-the twentieth century, see De Vido 1999.} A limited number of museum-based corpora have been published in recent
+the twentieth century, see \citet{de_vido_corpora_1999}.} A limited number of museum-based corpora have been published in recent
 decades (for Catania, Palermo, Messina, and Termini Imerese, as well as the material from Lipari), but this has not
@@
 dataset in MS Access. This consists of a single table originally constructed in MS Access 2000, and maintained
-erratically from the year 2000 onwards. The original purpose of this table was to gather data to assess the ‘epigraphic
-habit’ of ancient Sicily, and consequently the texts themselves were not the primary focus. However, the extent to
+erratically from the year 2000 onwards. The original purpose of this table was to gather data to assess the `epigraphic
+habit' of ancient Sicily, and consequently the texts themselves were not the primary focus. However, the extent to
 which the dataset facilitated further study made increasingly clear its potential value for the study of Sicilian
-epigraphy.\footnote{ The principal results were published in Prag 2002, revised in Prag 2004: 159-188; cf. Prag 2003,
-2007, 2008, 2010.}
+epigraphy.\footnote{ The principal results were published in \citet{prag_epigraphy_2002}, revised in \citet[159-188]{prag_sicily_2004}; cf.
+\citet{prag_nouveau_2003}, \citet{prag_ciceronian_2007}, \citet{prag_sicilia_2008}, \citet{prag_sicilia_2010}.}
 
@@
 combine a full record of past publication and study with a fully revised edition, and potential for multiple
-individuals to contribute to a process of on-going revision (see Fig. 1 for a draft edition of one inscription (AE
+individuals to contribute to a process of on-going revision (see Fig.~\ref{fig:1} for a draft edition of one inscription (AE
 1962.314 = I.Sicily 820)).
 
-\bigskip
-
-3. The aims of I.Sicily
+\section{The aims of I.Sicily}
 
@@
 
-\bigskip
+\subsection{Multilingualism}
 
-3.1. Multilingualism
-
-Sicily is traditionally described as a ‘melting pot’, the ‘crossroads of the Mediterranean’. The negative consequences
+Sicily is traditionally described as a `melting pot', the `crossroads of the Mediterranean'. The negative consequences
 of the separation of epigraphic material according to linguistic traditions have recently been highlighted and directly
 confronted by the Corpus Inscriptionum Iudaeae/Palaestinae (CIIP), edited by H. Cotton et al.\footnote{ Original notice
-in Cotton et al. 1999; presentation in Cotton and Price 2007.} I.Sicily sets out to follow in that mould, since the
+in \citet{cotton_corpus_1999}; presentation in \citet{cotton_corpus_2010}.} I.Sicily sets out to follow in that mould, since the
 different linguistic traditions of Sicily not only exist side-by-side but interact constantly throughout the island’s
 history, and no study of the epigraphic material can afford to ignore contemporary and parallel material in the other
-languages.\footnote{ See e.g. Manganaro 1993, Prag 2002, Salmeri 2004, Korhonen 2011, Tribulato 2012.} The situation
+languages.\footnote{ See e.g. \citet{manganaro_greco_1993}, \citet{prag_epigraphy_2002}, \citet{salmeri_i_2004}, \citet{korhonen_language_2011}, \citet{tribulato_language_2012}.} The situation
 created by basic technologies such as Unicode and EpiDoc XML mean that there is now no reason not to be language
@@
 Personal Names records most instances of Greek names for the island, but Sicily is no less rich in non-Greek names
-(Latin and others), and at present there is no onomasticon for the island.\footnote{ www.lgpn.ox.ac.uk/ [accessed
+(Latin and others), and at present there is no onomasticon for the island.\footnote{ \url{www.lgpn.ox.ac.uk/} [accessed
 26.09.2015].} Simply by the marking-up and indexing of all names in the island’s inscriptions, I.Sicily will generate a
@@
 
-\bigskip
-
-3.2. Identification and Bibliography
+\subsection{Identification and Bibliography}
 
@@
 traditional corpora for the region, but both are over 125 years old and cover less than 30\% of the material now
-known).\footnote{ Mommsen 1883; Kaibel 1890.} Existing online databases improve on this situation, but the results
+known).\footnote{\citet{_inscriptiones_1883}; \citet{kaibel_inscriptiones_1890}.} Existing online databases improve on this situation, but the results
 obtainable are of very varied value. The most comprehensive, in terms of the range of data recorded, is EDR (with which
-I.Sicily is collaborating), which currently reports 1,906 records for ‘Sicilia’; but this reduces to 833 when limited
-to texts on stone (‘lapis’ or ‘marmor’); contrast I.Sicily, with 2,563 records at the time of writing.\footnote{
-www.edr-edr.it/ [accessed 26.09.2015]} Clauss Slaby reports 4,374 records for ‘Sicilia’ (including Christian
-inscriptions, excluding ‘sigilla impressa’), but the return is inclusive of all kinds of epigraphic material, without
+I.Sicily is collaborating), which currently reports 1,906 records for `Sicilia'; but this reduces to 833 when limited
+to texts on stone (`lapis' or `marmor'); contrast I.Sicily, with 2,563 records at the time of writing.\footnote{
+\url{www.edr-edr.it/} [accessed 26.09.2015]} Clauss Slaby reports 4,374 records for `Sicilia' (including Christian
+inscriptions, excluding `sigilla impressa'), but the return is inclusive of all kinds of epigraphic material, without
 indication or discrimination, contains some duplication, is much harder to reconcile to existing records, and records
-only text.\footnote{ www.manfredclauss.de/ [accessed 26.09.2015].} The PHI database of Greek inscriptions has a rich
+only text.\footnote{ \url{www.manfredclauss.de/} [accessed 26.09.2015].} The PHI database of Greek inscriptions has a rich
 record of published Greek texts, but is text only and limited in outputs.\footnote{
-http://noapplet.epigraphy.packhum.org/regions/1156 [accessed 26.09.2015].} SEG references are available for 733
+\url{http://noapplet.epigraphy.packhum.org/regions/1156} [accessed 26.09.2015].} SEG references are available for 733
 inscriptions on stone and AE references for 328 (data taken from the I.Sicily database and based upon comprehensive
@@
 
-\ \ http://csad.ox.ac.uk/sicily/inscription/ISic1234
+\url{http://csad.ox.ac.uk/sicily/inscription/ISic1234}
 
@@
 Sicilian epigraphy, and an online Zotero library has already been created with over 700 records which are referenced in
-the EpiDoc files.\footnote{ www.zotero.org/isicily (private at the time of writing, but to be made public when the main
+the EpiDoc files.\footnote{\url{www.zotero.org/isicily} (private at the time of writing, but to be made public when the main
 site is launched) [accessed 26.09.2015].} A locally cached version of the bibliography will be presented at the
@@
 in digital format and several are already mounted in the Arachne archive, making direct page-citation
-possible.\footnote{ E.g. Castelli 1784, at http://arachne.uni-koeln.de/books/Castelli1784 [accessed 26.09.2015].}
+possible.\footnote{ E.g. Castelli 1784, at \url{http://arachne.uni-koeln.de/books/Castelli1784} [accessed 26.09.2015].}
 
 The richness of I.Sicily’s records in this area mean that I.Sicily is currently collaborating with both Trismegistos and
-IDES (‘Integrating Digital Epigraphies’).\footnote{ www.trismegistos.org/ [accessed 26.09.2015] and http://ides.io/
+IDES (`Integrating Digital Epigraphies').\footnote{\url{www.trismegistos.org/} [accessed 26.09.2015] and \url{http://ides.io/}
 [accessed 26.09.2105].} The former aims to generate TM numbers for all the Sicilian material (which I.Sicily will
@@
 
-\bigskip
-
-3.3. Identification and Collections
+\subsection{Identification and Collections}
 
@@
 I.Sicily is making use of the TEI {\textless}msIdentifer{\textgreater} element, with its associated sub-elements in
-order to record details of institutional collections and inventory numbers wherever possible.\footnote{
-www.tei-c.org/release/doc/tei-p5-doc/en/html/ref-msIdentifier.html [accessed 26.09.2015].} In order to maximise the
+order to record details of institutional collections and inventory numbers wherever possible.\footnote{\url{www.tei-c.org/release/doc/tei-p5-doc/en/html/ref-msIdentifier.html} [accessed 26.09.2015].} In order to maximise the
 value of this, we have adopted two further courses of action. In the first place, as part of the larger ambition of
@@
 
-\ http://csad.ox.ac.uk/sicily/museum/SicMus123
+\url{http://csad.ox.ac.uk/sicily/museum/SicMus123}
 
@@
 
-\bigskip
-
-3.4. Location, location, location
+\subsection{Location, location, location}
 
@@
 {\textless}geo{\textgreater} elements), and through the use of Pleiades URI references wherever possible.\footnote{
-http://pleiades.stoa.org/ [accessed 26.09.2015].} We are committed to the long-term use of Pleiades as our primary
+\url{http://pleiades.stoa.org/} [accessed 26.09.2015].} We are committed to the long-term use of Pleiades as our primary
 reference for ancient places, and to that end we aim to update and improve the Pleiades data for Sicilian locations, in
 particular name data and sub-locations, in conjunction with the editing of the I.Sicily records.\footnote{ See e.g.
-Wilson et al. \ 2015. Valeria Vitale (KCL) is currently undertaking a significant programme of data improvement in
+\citet{wilson_places:_2015}. Valeria Vitale (KCL) is currently undertaking a significant programme of data improvement in
 Pleiades on behalf of I.Sicily; we are grateful to Tom Elliott and Jeffrey Becker for their support.}
 
-\bigskip
-
-3.5 Searching
+\subsection{Searching}
 
@@
 spreadsheet interacts particularly well with maps: all findspots or museums in a filtered subset of the grid can be
-simultaneously shown on the map (see Fig. 2). The spreadsheet model also provides a very quick and intuitive (since so
+simultaneously shown on the map (see Fig.~\ref{fig:2}). The spreadsheet model also provides a very quick and intuitive (since so
 many people are familiar with spreadsheets) means for editing records (in this case, inscriptions and museums) online.
@@
 
-\bigskip
-
-3.6. Translations
+\subsection{Translations}
 
@@
 translations wherever possible in both English and Italian. We see this as one obvious area where public contribution
-(‘crowd-sourcing’) will be invaluable (see below).
+(`crowd-sourcing') will be invaluable (see below).
 
-4. Limitations and future ambitions
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=0.6\columnwidth]{EAGLE2016ISicilyfinalcopy-img001.jpg}
+\caption{sample edition}
+\label{fig:1}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016ISicilyfinalcopy-img002.png}
+\caption{Screenshot of map-based searching (museum locations) and of part of the spreadsheet/grid search interface
+employed in I.Sicily}
+\label{fig:2}
+\end{figure}
+
+\section{Limitations and future ambitions}
 
@@
 It is our long-term ambition that I.Sicily might become the default location for the publication and dissemination of
-Sicilian inscriptions; in the shorter term, we hope that it will serve as \ valuable portal in the world of Sicilian
+Sicilian inscriptions; in the shorter term, we hope that it will serve as valuable portal in the world of Sicilian
 epigraphy and of ancient world open linked data, greatly improving the accessibility of Sicilian epigraphy and so
-enriching the study of the ‘crossroads of the Mediterranean’.
+enriching the study of the `crossroads of the Mediterranean’.
 
-\bigskip
+\section*{Acknowledgments}
 
-\bigskip
-
- \includegraphics[width=7.197cm,height=18.653cm]{EAGLE2016ISicilyfinalcopy-img001.jpg} Fig. 1: sample edition.
-
- \includegraphics[width=10.881cm,height=7.966cm]{EAGLE2016ISicilyfinalcopy-img002.png}
-
-\bigskip
-
-Fig. 2. Screenshot of map-based searching (museum locations) and of part of the spreadsheet/grid search interface
-employed in I.Sicily.
-
-\bigskip
-
-Bibliography
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-[Warning: Draw object ignored]I.Sicily gratefully acknowledges the financial support of the John Fell Fund of the
+I.Sicily gratefully acknowledges the financial support of the John Fell Fund of the
 University of Oxford, and of the Warden and Scholars of the House or College of Scholars of Merton in the University of

### 2015-12-17 source/18/EAGLE2016_Full_Paper_Broux.tex
other biblio added till paper 20
no indent added

@@
 
-Facebooking the past. The idea grew a couple of years ago, while developing a database of all people mentioned in texts
+\noindent Facebooking the past. The idea grew a couple of years ago, while developing a database of all people mentioned in texts
 from Greco-Roman Egypt. While probably not exactly considered Big Data by those who actually work with BIG data, the
@@
 
-Like Facebook, however, Trismegistos wants to grow, and get the entire ancient world on board. To achieve this goal,
+\noindent Like Facebook, however, Trismegistos wants to grow, and get the entire ancient world on board. To achieve this goal,
 Trismegistos' core database, the text database, must first be expanded by broadening its chronological and geographic
@@
 
-With the experience obtained during the extraction of names from the Greek papyri, Named Entity Recognition will be
+\noindent With the experience obtained during the extraction of names from the Greek papyri, Named Entity Recognition will be
 applied to the Latin inscriptions incorporated in Trismegistos so far. This will result in the addition of an estimated
@@
 
-The collection of references to people and their names in the inscriptions will be carried out by applying Named Entity
+\noindent The collection of references to people and their names in the inscriptions will be carried out by applying Named Entity
 Recognition [NER] to the clusters of capitalized words extracted from the full text repositories of Greek and Latin
@@
 \textgreek{>Is'idoroc). }Many of the variants are actually created by renderings of a name in other languages, e.g.
-[1EC8?]sytrs, [21D?]ysydwrs or [2C93?][2CA5?][2C93?][2C87?][2CB1?][2CA3?][2C89?] in Egyptian and Coptic. In all there
+Ỉsytrs, ȝysydwrs or ⲓⲥⲓⲇⲱⲣⲉ in Egyptian and Coptic. In all there
 are 177,290 variants in the NAMVAR database. Finally, for each of the variants, the declined forms were created, e.g.
@@
 
-Once the clusters of capitalized words have been extracted and have been matched to the onomastic gazetteer and rules
+\noindent Once the clusters of capitalized words have been extracted and have been matched to the onomastic gazetteer and rules
 for name combinations, a human check will be performed. This includes tasks which are not easily automated:
@@
 
-Social Network Analysis [SNA] was developed in the 1960s in mathematics, anthropology and sociology and measures
+\noindent Social Network Analysis [SNA] was developed in the 1960s in mathematics, anthropology and sociology and measures
 structural forms of relations between individuals. It has huge potential for future historical research, not only by
@@
 
-The collection of names from across the entire Mediterranean will lead to a large-scale study of naming practices in the
+\noindent The collection of names from across the entire Mediterranean will lead to a large-scale study of naming practices in the
 ancient world, and how these reflect changes in society at large. A major transition point is of course the steady
@@
 
-Eventually, the goal of Trismegistos is to recreate a prosopography of the Greco-Roman world. Reconstructing social
+\noindent Eventually, the goal of Trismegistos is to recreate a prosopography of the Greco-Roman world. Reconstructing social
 networks of the past will help us gain a better understanding of the mechanisms of interaction in the ancient

### 2015-12-17 source/20/EAGLE2016_BONACINI - Pilot project at Paolo Orsi Museum.tex
no indent added

@@
 
-The ``Paolo Orsi'' Regional Archaeological Museum of Syracuse, together with the ``Antonino Salinas'' Regional
+\noindent The ``Paolo Orsi'' Regional Archaeological Museum of Syracuse, together with the ``Antonino Salinas'' Regional
 Archaeological Museum in Palermo, is the most important Sicilian archaeological museum and it is one of the most
@@
 
-This project born from the desire to fill the deep gap in the promotion and enhancement of Sicilian cultural heritage.
+\noindent This project born from the desire to fill the deep gap in the promotion and enhancement of Sicilian cultural heritage.
 
@@
 
-A large photographic survey began in August 2014, with the aim of mapping all areas of the first and second level open
+\noindent A large photographic survey began in August 2014, with the aim of mapping all areas of the first and second level open
 to visitors (only the collection of coins and medals, for security reason, has not yet been photographed).
@@
 
-Coming from Sector A, after turning around a couple of casts of dwarf elephants from Spinagallo cave (Syracuse), a
+\noindent Coming from Sector A, after turning around a couple of casts of dwarf elephants from Spinagallo cave (Syracuse), a
 remote user can see the displayed artifacts, starting with Neolithic phase of Stentinello (VI millennium B.C.) to reach
@@
 
-The Museum staff has selected 12 objects along the path between the two levels, which can be linked as points of
+\noindent The Museum staff has selected 12 objects along the path between the two levels, which can be linked as points of
 interest placed on the Museum map - where the remote user can identify sectors, plans and POI - with virtual 360° tours
@@
 
-The last selected object we present here is a disk from Sector F: it is a marble disk, discovered by Paolo Orsi in the
+\noindent The last selected object we present here is a disk from Sector F: it is a marble disk, discovered by Paolo Orsi in the
 Catacombs of San Giovanni in 1894 \citep[509-510]{Orsi1895}, decorated by a wreath of laurel leaves on one side; the other
@@
 
-As discussed elsewhere (Bonacini 2013 and 2014), Google is undoubtedly the most active entity in the world committed to
+\noindent As discussed elsewhere (Bonacini 2013 and 2014), Google is undoubtedly the most active entity in the world committed to
 preservation, dissemination and promotion of cultural heritage, well above any public institution, through an
@@
 
-To Mr. Gianfranco Guccione (\url{http://www.airworks.it/}) goes my heartfelt and deep thanks for having proposed to me this
+\noindent To Mr. Gianfranco Guccione (\url{http://www.airworks.it/}) goes my heartfelt and deep thanks for having proposed to me this
 project and having concluded it together, with the only aim of enhancing the Cultural Heritage in Sicily.

### 2015-12-17 source/21/EAGLE2016_Full_Paper_revised.tex
no indent added

@@
 
-AXON. A Selection of Greek Historical Inscriptions is a project conceived within the Greek Epigraphy Laboratory
+\noindent AXON. A Selection of Greek Historical Inscriptions is a project conceived within the Greek Epigraphy Laboratory
 (Director, Prof. Claudia Antonetti), and has been brought into existence with the financial support of the University
@@
 
-The model-entry has been created with an eye on the object’s thorough description. Here is the object’s categorisation:
+\noindent The model-entry has been created with an eye on the object’s thorough description. Here is the object’s categorisation:
 
@@
 
-The chronological delimitation of each text is supported by many options, as you can see in Fig.~\ref{fig:1}:
+\noindent The chronological delimitation of each text is supported by many options, as you can see in Fig.~\ref{fig:1}:
 
@@
 
-Each entry provides all the necessary information on the alphabet and language of each inscription:
+\noindent Each entry provides all the necessary information on the alphabet and language of each inscription:
 
@@
 
-The text of each inscription is preceded by a hierarchically arranged lemma (the so called genetic lemma, according to
+\noindent The text of each inscription is preceded by a hierarchically arranged lemma (the so called genetic lemma, according to
 Louis Robert’s definition\footnote{\citet{robert_carie:_1954}.}) and is followed by the apparatus criticus.
@@
 
-Each entry corresponds to an Italian translation and commentary (in .pdf).
+\noindent Each entry corresponds to an Italian translation and commentary (in .pdf).
 
@@
 
-The Abstract – with a WYSIWYG interface – includes all the keywords for indexing and lemmatisation:
+\noindent The Abstract – with a WYSIWYG interface – includes all the keywords for indexing and lemmatisation:
 
@@
 
-Finally, an updated bibliography highlights any previous edition for each entry, as well as all the appropriate
+\noindent Finally, an updated bibliography highlights any previous edition for each entry, as well as all the appropriate
 secondary sources. SEG abbreviations have been used for epigraphic corpora and other publications (the section
@@
 
-Each entry is related – whenever it seems appropriate – to other entries in the AXON database. A hyperlink connects the
+\noindent Each entry is related – whenever it seems appropriate – to other entries in the AXON database. A hyperlink connects the
 entry with other digital editions of the same text (if available), or with other useful websites, possibly containing
@@
 
-Since the contributors to the project (i.e. the authors of the entries) are experts from different Italian and European
+\noindent Since the contributors to the project (i.e. the authors of the entries) are experts from different Italian and European
 universities (and not all of them are familiar with the Digital Humanities), and given the great number of entries
@@
 
-The website is designed to allow for many search options. Beyond the ``full text'' search and another based on the number,
+\noindent The website is designed to allow for many search options. Beyond the ``full text'' search and another based on the number,
 title and author of the entry (see Fig.~\ref{fig:3}), three other search-possibilities will also be available:
@@
 
-The AXON Project, as an example of a digital edition of inscriptions (see esp. genetic lemma and apparatus) with a high
+\noindent The AXON Project, as an example of a digital edition of inscriptions (see esp. genetic lemma and apparatus) with a high
 degree of clarity for contributors and users, is a useful tool for teaching Greek epigraphy as well as ancient history.
@@
 
-Each entry is created by an expert contributor and is subject to double-blind peer review, thus assuring an important
+\noindent Each entry is created by an expert contributor and is subject to double-blind peer review, thus assuring an important
 contribution to the scholarly community. At the same time, the hyperlinks to other websites and digital editions will

### 2015-12-17 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
updated paper 22, updated paper 24
paper 24 doc to tex
added email in paper 22
no indent added

@@
 \title{The Digital Edition of the Archaic Latin Inscriptions (7th{}-5th century B.C.)}
-\author[unimi]{Giovanna Rocca \corref{first}}
-\author[kore]{Giulia Sarullo}
+\author[unimi]{Giovanna Rocca }
+\author[kore]{Giulia Sarullo \corref{first}}
 \author[unimi]{Marta Muscariello}
@@
 \address[unimi]{Università IULM Milano}
+\cortext[first]{Corresponding author. Email: giulia.sarullo@unikore.it}
 \date{2015-09-18}
@@
 
-The project Iscrizioni Latine Arcaiche (ILA) consists in the digital edition of the inscriptions found in old
+\noindent The project Iscrizioni Latine Arcaiche (ILA) consists in the digital edition of the inscriptions found in old
 Latium\footnote{With the two exceptions of the Vendia's Urn, found in Cerveteri, and the Garigliano bowl.}
@@
 
-Thanks to the agreement and collaboration with public authorities, we were invited to take part in an extraordinary
+\noindent Thanks to the agreement and collaboration with public authorities, we were invited to take part in an extraordinary
 event (July 3rd 2015), that is the 3D laser scanning of the Forum Cippus (CIL I2, 1).\footnote{Cippus of tuff found
@@
 
-In the website, each text of the ancient Latin corpus is presented by an ad hoc designed chart in order to meet the
+\noindent In the website, each text of the ancient Latin corpus is presented by an ad hoc designed chart in order to meet the
 peculiar requirements of this kind of corpus. The chart is organized in items that concern every important aspect
@@
 
-An epigraphic corpus can be digitalized in different ways, according to the specific issues that each project intends to
+\noindent An epigraphic corpus can be digitalized in different ways, according to the specific issues that each project intends to
 tackle. Unlike EDR that, as the other projects constituting the Electronic Archive of Greek and Latin Epigraphy

### 2015-12-17 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
no indent added

@@
 
-The aim of this work is to do a general overview of the computer-based epigraphic corpora, to think about their features
+\noindent The aim of this work is to do a general overview of the computer-based epigraphic corpora, to think about their features
 and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
@@
 
-All above-mentioned principles have been taken into account in Hesperia project, its main goal being to gather all
+\noindent All above-mentioned principles have been taken into account in Hesperia project, its main goal being to gather all
 linguistic evidences from Palaeohispanic languages, that is, pre-Roman languages in the Iberian Peninsula\footnote{
@@
 
-The Hesperia databank has been recently extended with a new resource on indigenous onomastics. More specifically, the
+\noindent The Hesperia databank has been recently extended with a new resource on indigenous onomastics. More specifically, the
 new sections now available are devoted to the anthroponymy and theonymy, leaving for the near future the data referring

### 2015-12-17 source/29/EAGLE2016_camera_ready.tex
no indent added

@@
 
-The technological advances in the last decade have equipped the general public with several handheld electronic
+\noindent The technological advances in the last decade have equipped the general public with several handheld electronic
 commodities that changed significantly daily routine in a personal and professional level and contributed to the user's
@@
 
-Understanding the users is one of the integral steps of interaction design, which is an iterative process during which
+\noindent Understanding the users is one of the integral steps of interaction design, which is an iterative process during which
 representative users interact with preliminary designs and provide useful feedback \citep{preece_interaction_2015}. For the
@@
 
-Natural User Interfaces consist of sensors that track the natural behavior of users and provide a natural form of
+\noindent Natural User Interfaces consist of sensors that track the natural behavior of users and provide a natural form of
 interactivity with computers and other electronic devices. The common forms of NUI sensors are: pressure sensors for
@@
 
-\ An optimal interaction design solution should be intuitive, minimalistic, and non intrusive \citep{preece_interaction_2015}.
+An optimal interaction design solution should be intuitive, minimalistic, and non intrusive \citep{preece_interaction_2015}.
 Therefore, in order to design interaction for epigraphy one should choose devices that are easily accessible by
@@
 
-In order to achieve natural interactive relighting of an inscription, the system should imitate the process of
+\noindent In order to achieve natural interactive relighting of an inscription, the system should imitate the process of
 relighting a handheld physical object (such as a paper cast of inscription) by reorienting the object with respect to
@@
 
-Another important form of interaction in the epigraphic routine is the change of the point of view in order to
+\noindent Another important form of interaction in the epigraphic routine is the change of the point of view in order to
 understand better the structural details of the inscribed letterforms. Assuming that the model of the inscription is
@@
 
-The forms of interactions presented in 1.3.1 and 1.3.2 involved only the motion sensors of a tabled computer without
+\noindent The forms of interactions presented in 1.3.1 and 1.3.2 involved only the motion sensors of a tabled computer without
 utilizing the pressure sensors of the screen of the device. The commonly used touch gestures (such as 2-finger twist to
@@
 
-In this pilot study, a complete set of natural user interactions was designed based on the physical interactions of
+\noindent In this pilot study, a complete set of natural user interactions was designed based on the physical interactions of
 epigraphists with real inscriptions. The proposed interactions utilize the existing sensors in a typical tablet
@@
 
-We would like to acknowledge our collaborators for their invaluable interaction and feedback during the past three
+\noindent We would like to acknowledge our collaborators for their invaluable interaction and feedback during the past three
 years. We would also like to thank the UF College of the Arts and the UF Center for Greek Studies for providing

### 2015-12-17 source/31/EAGLE2016_I Sicily_final copy.tex
no indent added

@@
 \section{Introduction: What is I.Sicily}
-
-I.Sicily is an online, open access, digital corpus of the inscriptions on stone from ancient Sicily.\footnote{ The
+\noindent I.Sicily is an online, open access, digital corpus of the inscriptions on stone from ancient Sicily.\footnote{ The
 corpus will be mounted at \url{www.csad.ox.ac.uk/sicily/isicily}, but is presently on a development server. A public face is
@@
 \section{The motivations for and origins of I.Sicily}
-
-The existing epigraphic landscape in Sicily is extremely diverse in two primary regards: on the one hand, the island has
+\noindent The existing epigraphic landscape in Sicily is extremely diverse in two primary regards: on the one hand, the island has
 a very mixed cultural and linguistic make-up, meaning that the epigraphic material is itself extremely varied, with
@@
 \section{The aims of I.Sicily}
-
-We outline briefly five areas in which I.Sicily aims to develop, facilitate and improve the study of epigraphic material
+\noindent We outline briefly five areas in which I.Sicily aims to develop, facilitate and improve the study of epigraphic material
 from ancient Sicily.
@@
 
-Sicily is traditionally described as a `melting pot', the `crossroads of the Mediterranean'. The negative consequences
+\noindent Sicily is traditionally described as a `melting pot', the `crossroads of the Mediterranean'. The negative consequences
 of the separation of epigraphic material according to linguistic traditions have recently been highlighted and directly
@@
 
-Sicily presents a particularly extreme version of the common problem of identifying a text and its publication record.
+\noindent Sicily presents a particularly extreme version of the common problem of identifying a text and its publication record.
 No existing corpus in either Greek or Latin comes close to full coverage (CIL X and IG XIV are the largest individual
@@
 
-The traditional focus of epigraphic study upon the text, rather than the epigraphic support, means that epigraphic
+\noindent The traditional focus of epigraphic study upon the text, rather than the epigraphic support, means that epigraphic
 publication in the past has frequently been relatively limited in the information which it has recorded about the
@@
 
-I.Sicily is actively generating rich geo-data for the individual inscriptions, both for the original findspot/provenance
+\noindent I.Sicily is actively generating rich geo-data for the individual inscriptions, both for the original findspot/provenance
 and the current location (whether museum-based, on-site, or elsewhere), and we aim to provide map-based searching for
@@
 
-In order to support the aims outlined above, I.Sicily has taken a different approach to search and browse. Although
+\noindent In order to support the aims outlined above, I.Sicily has taken a different approach to search and browse. Although
 standard form-based search with paged results, like that of Google, makes sense for very large result sets, the
@@
 
-As was extensively discussed at the first EAGLE conference (Paris 2014), the creation and availability of translations
+\noindent As was extensively discussed at the first EAGLE conference (Paris 2014), the creation and availability of translations
 is a major goal of the EAGLE project and its collaborators, and I.Sicily is no less committed to that
@@
 
-The scale of the enterprise, and the available resources, mean that in its current form the project has limited itself
+\noindent The scale of the enterprise, and the available resources, mean that in its current form the project has limited itself
 to inscriptions engraved on stone (the coverage of rupestral inscriptions/graffiti and of inscriptions painted on
@@
 
-I.Sicily gratefully acknowledges the financial support of the John Fell Fund of the
+\noindent I.Sicily gratefully acknowledges the financial support of the John Fell Fund of the
 University of Oxford, and of the Warden and Scholars of the House or College of Scholars of Merton in the University of

### 2015-12-17 source/33/TIGLIO.tex
no indent added

@@
 
-Nearly all Greek and Latin epigraphic texts are available, sometimes in multiple versions, after three decades of continuous digitization and online publication.
+\noindent Nearly all Greek and Latin epigraphic texts are available, sometimes in multiple versions, after three decades of continuous digitization and online publication.
 Without counting repeated inscriptions and \textit{instrumenta}, there are about 300.000 Latin inscriptions\footnote{EAGLE disambiguated total: 235.626 ; EDCS not disambiguated total without \textit{instrumenta} and \textit{inscriptiones christianae}: 308.581.
@@
 \section{Problems}
-Let's look at the data we have.
+\noindent Let's look at the data we have.
 The EAGLE project has gathered some insights on small collections of images of inscriptions openly published online, on Wikimedia Commons,\footnote{\url{https://commons.wikimedia.org/wiki/Category:Media_contributed_by_EAGLE}} and on a set of translations of inscriptions, collected in the EAGLE Media Wiki.\footnote{\citet{Liuzzo2014} \url{http://www.eagle-network.eu/wiki/index.php/Main_Page}} We shall compare their impact and reach on the wider public to that of the texts of inscriptions, to underline the urgency for these materials to be produced also in order to bring epigraphy outside its restricted circles.
@@
 \section{Translations}
-To face these challenges with regard to translations, it is the opinion of the project team that tying together existing resources is a better way to tackle the issues rather than trying to superimpose a new tool or system.
+\noindent To face these challenges with regard to translations, it is the opinion of the project team that tying together existing resources is a better way to tackle the issues rather than trying to superimpose a new tool or system.
 The available building blocks for such systematization of existing resources are the following:
@@
 \section{Images}
-Most of the existing images of inscriptions are currently safely stored in private computers.
+\noindent Most of the existing images of inscriptions are currently safely stored in private computers.
 Large collections of images of inscriptions are available at the major databases, and can count up to thousands of photos of inscriptions or drawings.
@@
 \section{User involvement and expert sourcing}
-The problems of all digital projects looking at putting together materials from various sources and contributors are on one side to get people involved, on the other to overlook the work done and take care of the administration.
+\noindent The problems of all digital projects looking at putting together materials from various sources and contributors are on one side to get people involved, on the other to overlook the work done and take care of the administration.
 The possible user scenarios are eventually many more.
@@
 \section{Conclusions}
-Some of the tasks above have been already carried out, some are under way, but there are some general conclusions which can be summarized.
+\noindent Some of the tasks above have been already carried out, some are under way, but there are some general conclusions which can be summarized.
 There is a need to unify the resources, agree on standards for reference and citation and provide stable identifiers and citation structures, providing a comprehensive list of epigraphic publications with the relevant abbreviation in use.
@@
 \section*{Aknowledgements}
-The Andrew W.
+\noindent The Andrew W.
 Mellon Foundation has funded the meetings of the project TIGLIO and the work which will be carried out for Attic Inscriptions Online.

### 2015-12-17 source/36/DatabasesToEpiDoc.tex
no indent added

@@
 
-The Europeana network of Ancient Greek and Latin Epigraphy\footnote{\citet{OrlandiCasarosa2014} \citet{Orlandi2014} \citet{liuzzo2014} \url{http://www.eagle-network.eu/}} brings
+\noindent The Europeana network of Ancient Greek and Latin Epigraphy\footnote{\citet{OrlandiCasarosa2014} \citet{Orlandi2014} \citet{liuzzo2014} \url{http://www.eagle-network.eu/}} brings
 together most repositories of ancient epigraphic material and aims to
@@
 \section{Why XML}
-Choosing XML for an online publication is almost obvious given the current stage of development of the editorial methodology in epigraphy.
+\noindent Choosing XML for an online publication is almost obvious given the current stage of development of the editorial methodology in epigraphy.
 However there is room for a brief explanation of why this is preferable on a large scale to a relational database.
@@
 \section{EAGLE Mappings and Metadata modeling}
-The EAGLE BPN has taken 18 different content providers and has come to
+\noindent The EAGLE BPN has taken 18 different content providers and has come to
 the decision to take the first steps to bring hundreds of thousands of
@@
 
-The occasion of a mapping work allows also for other tasks of curation to be performed.
+\noindent The occasion of a mapping work allows also for other tasks of curation to be performed.
 EAGLE Members had in some case the text of a same inscription with three parallel different types of encoding conventions applied.
@@
 \subsection{Step 1}
-Each project uses different conventions and therefore the regular expressions used to match particular situations are different.
+\noindent Each project uses different conventions and therefore the regular expressions used to match particular situations are different.
 The process of mark up of the string text in div[@type="edition"] is accomplished in several steps to guarantee consistency and precision.
@@
 \subsection{Step 2}
-On the elements which contain information such as Object Type, Material, Execution Technique, etc.
+\noindent On the elements which contain information such as Object Type, Material, Execution Technique, etc.
 which are typically handled with a controlled vocabulary a series of XSL transformations is passed, one for each vocabulary related to that element  to match the content of the element with the vocabulary entry into the SKOS vocabulary stored in git, regularly updated and published as a self standing resource on the EAGLE portal.\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}} What follows is a brief description of such vocabularies and their development.
@@
 \section{Linked Open Data}
-The Europeana LOD practice for metadata recommends the adoption of machine readable vocabularies.
+\noindent The Europeana LOD practice for metadata recommends the adoption of machine readable vocabularies.
 Within the EAGLE BPN Linked Open Data practices and approaches have been taken on board for a process which will bring current material some steps closer to that practice.\footnote{\citet{Bizer2009}} Addressing existing problems in classification and the publication of a machine readable vocabulary of values (controlled vocabulary) is one of these steps:\footnote{\citet[4-5]{Harper2012} for the definition.
@@
 \section{Crosswalking EpiDoc to EDH with XSLT}
-The proof of concept of usability of the XML source file is given by the reverse function being possible, i.e.
+\noindent The proof of concept of usability of the XML source file is given by the reverse function being possible, i.e.
 to transform the XML files into a database data structure.
@@
 \section{Conclusion}
-I have presented in this paper the effort of a large consortium of people and institutions belonging to the same field of interest, that of Ancient Roman and Greek Epigraphy.
+\noindent I have presented in this paper the effort of a large consortium of people and institutions belonging to the same field of interest, that of Ancient Roman and Greek Epigraphy.
 Mappings and harmonization of data have been for the EAGLE project not just a way to allow machines to do more, but an intellectual effort of revision of contents and establishment of best practices.

### 2015-12-17 source/38/EAGLE2016_Final_Paper Modolo.tex
no indent added

@@
 %%%%%%%%
-\title{}
-\author{Mirco Modolo}
+\title{FREE REPRODUCTION OF CULTURAL HERITAGE FOR RESEARCH PURPOSES: REGULATORY ASPECTS AND NEW PROSPECTS FOR PROJECT `EAGLE'}
+\author[free]{Mirco Modolo\corref{first}}
+\address[free]{Free lance archaeologist and archivist}
+\cortext[first]{Corresponding author. Email: mircomodolo@gmail.com}
 \date{2015-11-20}
 \begin{document}
-Abstract
 
-\begin{figure}
-\centering
-\begin{minipage}{10.901cm}
-FREE REPRODUCTION OF CULTURAL HERITAGE FOR RESEARCH PURPOSES: REGULATORY ASPECTS AND NEW PROSPECTS FOR PROJECT
-‘EAGLE’\newline
-
-Mirco Modolo
-
-Free lance archaeologist and archivist
-
-Email(s): mircomodolo@gmail.com
-\end{minipage}
-\end{figure}
+\maketitle
+\begin{abstract}
 This paper traces the genesis and activity of the “Fotografie libere per i Beni Culturali” movement promoting free
@@
 archives and libraries. Simultaneously, evidence highlights the advantages of deregulating cultural heritage images for
-historical research, in particular for the study of antique epigraphy developed by the EAGLE project. \
+historical research, in particular for the study of antique epigraphy developed by the EAGLE project.
+\end{abstract}
+\keywords{Archivi, Biblioteche, Manoscritti, Riproduzioni, Creative Commons}
 
-\bigskip
+Free reproduction of cultural heritage outlined in the “Art Bonus” decree
 
-Keywords
-
-Archivi, Biblioteche, Manoscritti, Riproduzioni, Creative Commons
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\begin{enumerate}
-\item Free reproduction of cultural heritage outlined in the “Art Bonus” decree
-\end{enumerate}
-\ L’entrata in vigore, il primo giugno 2014, del decreto “Art Bonus”, che sanciva la libera riproduzione per finalità
+L’entrata in vigore, il primo giugno 2014, del decreto “Art Bonus”, che sanciva la libera riproduzione per finalità
 scientifiche, purché effettuata a distanza, di tutti i beni culturali rappresentò una novità assoluta per la ricerca
@@
 
-\bigskip
-
 \begin{enumerate}
@@
 dei Beni e delle Attività Culturali e del Turismo’).
-
-\bigskip
 
@@
 treasury\footnote{\ http://www.bianchibandinelli.it/2015/05/25/fotografie-libere-un-comunicato-dellassociazione-bianchi-bandinelli/}.
-
-\bigskip
-
-\bigskip
 
@@
 
-\bigskip
-
-\bigskip
-
 \begin{enumerate}
@@
 
-\bigskip
-
 The reuse of cultural heritage images, briefly mentioned in the proposal scope for “Fotografie libere per i Beni
@@
 release of an appropriate circular which should affirm, without ambiguity:
-
-\bigskip
 

### 2015-12-17 source/38/EAGLE2016_Final_Paper Modolo.tex
updated paper 38
doc to tex. removed italian

@@
 \begin{abstract}
-This paper traces the genesis and activity of the “Fotografie libere per i Beni Culturali” movement promoting free
+This paper traces the genesis and activity of the ``Fotografie libere per i Beni Culturali'' movement promoting free
 remote reproduction, and reuse for scientific purposes, of manuscripts and antique volumes conserved in Italian
@@
 
-Free reproduction of cultural heritage outlined in the “Art Bonus” decree
+\section{Free reproduction of cultural heritage outlined in the ``Art Bonus'' decree}
 
-L’entrata in vigore, il primo giugno 2014, del decreto “Art Bonus”, che sanciva la libera riproduzione per finalità
-scientifiche, purché effettuata a distanza, di tutti i beni culturali rappresentò una novità assoluta per la ricerca
-storica: numerosi archivi e biblioteche resero infatti libera e gratuita la riproduzione con la fotocamera dell’utente
-di manoscritti e volumi storici, con evidente vantaggio per la ricerca storica e in particolare per lo studio
-dell’epigrafia nei manoscritti antiquari.
-
-Appena un mese dopo un emendamento restrittivo, poi divenuto legge, intervenne ad escludere esplicitamente dalla
-liberalizzazione i beni bibliografici e archivistici. Alla sorpresa subentrò allora l’irritazione di numerosi studiosi,
-alimentata dal paradosso che mentre rimaneva libera la fotografia per il turista nel museo, lo studioso impegnato in
-attività di ricerca in archivi e biblioteche non avrebbe invece più potuto beneficiare dei vantaggi della
-liberalizzazione.
-
-On the first of June 2014, the ‘Art Bonus’ decree came into effect, sanctioning the free reproduction of all cultural
+On the first of June 2014, the `Art Bonus' decree came into effect, sanctioning the free reproduction of all cultural
 heritage for scientific purposes (even reproduction carried out at a distance). It represented a new first for
-historical research\footnote{\ Decreto-legge 31 maggio 2014 n. 83, art. 12.}: numerous archives and libraries opened
+historical research\footnote{Decreto-legge 31 maggio 2014 n. 83, art. 12.}: numerous archives and libraries opened
 and made free of charge the reproduction of manuscripts and historical volumes with a user’s camera. This provided a
@@
 later, a restrictive amendment modified the original text of the law, explicitly excluding printed books, manuscripts
-and archival documents from the liberalisation\footnote{\ Legge 29 luglio 2014, n. 106}. Many scholars’ surprise was
+and archival documents from the liberalisation\footnote{Legge 29 luglio 2014, n. 106}. Many scholars’ surprise was
 overtaken by irritation, driven by an apparent paradox: museum photography remained free for the tourist, yet the
 scholar engaged in research activity in archives and libraries would not reap the benefits of
-liberalisation\footnote{\ MODOLO, M., Il sogno infranto delle libere riproduzioni, in Il Giornale dell’Arte, p. 10
-(settembre 2014); Brugnoli, A., Ancora sulla riproduzione dei beni culturali, in ROARS (05/10/2014).\par }.
+liberalisation\footnote{\citet{modolo_il_2014}; \citet{brugnoli_ancora_2014}. }.
 
@@
 Furthermore this is in direct contrast with the Italian Constitution which, according to articles 33 and 9, requires
-the Republic not only to guarantee free research, but also to actively promote it. An effective “tax” on reproduction
+the Republic not only to guarantee free research, but also to actively promote it. An effective ``tax'' on reproduction
 with a personal device becomes a research obstacle; it is absurd to expect a fee for a service provided autonomously by
@@
 
-Regarding ‘reuse’ of the image, publication for scientific purposes remains subject to a precise and formal request to
-have ‘publication authorised’ granted on stamped paper, hugely time consuming for administrator as well as applicant.
-The payment of fees for use (‘canoni di utilizzo’ or royalties) is excluded instead for the publication of images of
+Regarding `reuse' of the image, publication for scientific purposes remains subject to a precise and formal request to
+have `publication authorised' granted on stamped paper, hugely time consuming for administrator as well as applicant.
+The payment of fees for use (`canoni di utilizzo' or royalties) is excluded instead for the publication of images of
 cultural heritage (such as archival documents, ancient printed books or ancient inscriptions) in scholarly books or
-papers with a cover price of less than 77 Euro and up to 2000 printed copies (as per circular 21/2005 of the ‘Ministero
-dei Beni e delle Attività Culturali e del Turismo’).
+papers with a cover price of less than 77 Euro and up to 2000 printed copies (as per circular 21/2005 of the `Ministero
+dei Beni e delle Attività Culturali e del Turismo').
 
-\begin{enumerate}
-\item \begin{enumerate}
-\item \ \ Traditional objections to liberalisation of reproductions
-\end{enumerate}
-\end{enumerate}
+\section{Traditional objections to liberalisation of reproductions}
+
 The principal objections opposing the liberalisation of reproductions of bibliographic and archival records, and which
-must be addressed here, regard both ‘conservationist’ as well as derivational ‘improper use’ risks for the images.
+must be addressed here, regard both `conservationist' as well as derivational `improper use' risks for the images.
 
@@
 the State Archives of Florence (Archivio di Stato di Firenze) directly noted the advantages of free reproduction in the
-month “Art Bonus” came into force (prior to the approval of the restrictive amendment): one could observe not only that
+month ``Art Bonus'' came into force (prior to the approval of the restrictive amendment): one could observe not only that
 the images did not damage the archival material, but rather, being subject to less movement, actually contributed to
-its conservation\footnote{\ https://fotoliberebbcc.wordpress.com/2015/06/06/4724/}.
+its conservation\footnote{\url{https://fotoliberebbcc.wordpress.com/2015/06/06/4724/}}.
 
@@
 compressing the back of the volume and risking damaging the
-binding\footnote{\ http://www.bl.uk/reshelp/inrrooms/stp/copy/selfsrvcopy/book\_photography\_video.mp4.}.
+binding\footnote{\url{http://www.bl.uk/reshelp/inrrooms/stp/copy/selfsrvcopy/book\_photography\_video.mp4}.}.
 
@@
 
-To those fearing ‘abuse’ relating to improper use of images (different from those permitted by authorisation) it is easy
-to respond that prior authorisation had already been eliminated from “Art Bonus” for photographs of all cultural
-heritage other than bibliographical or archival records. In other words, ‘ex post’ control instead of ‘ex ante’
+To those fearing `abuse' relating to improper use of images (different from those permitted by authorisation) it is easy
+to respond that prior authorisation had already been eliminated from ``Art Bonus'' for photographs of all cultural
+heritage other than bibliographical or archival records. In other words, `ex post' control instead of `ex ante'
 control, characterised as the only genuinely effective measure. It is hard to imagine now, in the digital era, strict
 ironclad control over all who request daily permission to reproduce mass of images which are each day authorised only
-for “educational purposes”! This is a problem which doesn’t exist in France: the Archives Nationales does not in fact
+for ``educational purposes''! This is a problem which doesn’t exist in France: the Archives Nationales does not in fact
 provide any specific request for autonomous reproduction, while the Bibliothéque National de France allows anyone to
@@
 generating profits, a pretext designed to discourage reproduction of manuscripts and ancient volumes and to minimise
-the risk of abuse associated with using the images for profit-making activity, in a way which might ‘damage’ the public
-treasury\footnote{\ http://www.bianchibandinelli.it/2015/05/25/fotografie-libere-un-comunicato-dellassociazione-bianchi-bandinelli/}.
+the risk of abuse associated with using the images for profit-making activity, in a way which might `damage' the public
+treasury\footnote{\url{http://www.bianchibandinelli.it/2015/05/25/fotografie-libere-un-comunicato-dellassociazione-bianchi-bandinelli/}}.
 
-\begin{enumerate}
-\item The proposal of \ “Fotografie libere per i Beni Culturali”
-\end{enumerate}
-Sulla scia del decreto “Art Bonus”, per tentare allora di sanare una situazione di disagio diffusa tra gli studiosi nel
-mese di settembre 2014 si è costituito spontaneamente tra i ricercatori un movimento di idee, “Fotografie libere per i
-Beni Culturali”, che ha avviato una petizione di massa al ministro dei Beni Culturali Dario Franceschini finalizzata a
-promuovere la libera riproduzione in archivi e biblioteche, ripristinando lo spirito originario dell’”Art Bonus” Una
-iniziativa senza precedenti non solo per il numero di firme, più di 4200, ma anche per la qualità dei sottoscrittori:
-sono presenti, oltre ai massimi rappresentanti in Italia di tutte le discipline storico-umanistiche impegnati nella
-ricerca storica e nella tutela dei beni archivistici e bibliografici, anche funzionari della pubblica amministrazione,
-studenti, ricercatori e semplici appassionati di storia locale.
+\section{The proposal of ``Fotografie libere per i Beni Culturali''}
 
-Il movimento ha prodotto inoltre una proposta di modifica dell’art. 108 del Codice dei Beni Culturali disciplinante le
-riproduzioni, visualizzabile in una pagina del sito web del
-movimento\footnote{\ https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/}, con la quale si chiede al
-legislatore di limitare i casi di esclusione della liberalizzazione ai beni bibliografici e archivistici
-rispettivamente ai volumi a stampa già tutelati da norme sul diritto di autore e ai documenti di archivio contenenti
-dati sensibili (privacy).
-
-La libera riproduzione dunque è uno strumento fondamentale che, venendo incontro alle esigenze prioritarie della
-conservazione, agevola fortemente l’attività di trascrizione dei documenti e dunque la ricerca storica, consentendo un
-significativo risparmio in termini di tempo e di denaro per lo studioso, soprattutto per chi è costretto a spostarsi in
-archivi distanti dal proprio luogo di residenza.
-
-La liberalizzazione, intesa come gratuità dello scatto con mezzo proprio che si accompagna all’esenzione dall’obbligo di
-autorizzazione preventiva, si potrà estendere, secondo il testo della proposta, non solo alle copie effettuate per
-“studio personale”, ma anche per al riuso ‘non commerciale’ delle immagini, prevedendo una eccezione ‘culturale’ per
-agevolare le pubblicazioni a stampa scientifiche entro precisi termini. “Fotografie libere per i Beni Culturali”
-propone infatti di rendere libera la pubblicazione di immagini di beni culturali in testi scientifici “minori” con
-l’obbligo tuttavia di specificare sempre il nome dell’istituto di provenienza e di consegna di un esemplare della
-pubblicazione. In questi casi la richiesta formale di ‘concessione’ potrà essere sostituita da una semplice
-comunicazione online all’istituto dell’intenzione di pubblicare. Per le pubblicazioni “maggiori”, oltre cioè le soglie
-indicate, ed in generale per attività lucrative, rimarrà in vita il regime di concessione, che prevede sia la richiesta
-formale di autorizzazione sia il pagamento di canoni d’uso (royalties).
-
-Infine si richiede di creare le condizioni normative per facilitare la condivisione sul web delle riproduzioni di
-documenti di archivio, manoscritti e volumi antichi eseguite in sala dagli studiosi.
-
-To meet scholars’ demands, in September 2014 researchers spontaneously formed a movement of ideas, “Fotografie libere
-per i Beni Culturali”. The group launched a petition asking Minister of Culture on. Dario Franceschini for free
-reproduction in archives and libraries, renewing the original spirit of the “Art Bonus”
-decree\footnote{\ https://fotoliberebbcc.wordpress.com/category/adesioni-e-contatti/}. It was an unprecedented
+To meet scholars’ demands, in September 2014 researchers spontaneously formed a movement of ideas, ``Fotografie libere
+per i Beni Culturali''. The group launched a petition asking Minister of Culture on. Dario Franceschini for free
+reproduction in archives and libraries, renewing the original spirit of the ``Art Bonus''
+decree\footnote{\url{https://fotoliberebbcc.wordpress.com/category/adesioni-e-contatti/}}. It was an unprecedented
 initiative given not only the number of signatures (more than 4200), but also the quality of subscribers: a chorus
@@
 state archives and functionaries of the ministry
-itself\footnote{\ http://www.ilgiornaledellarte.com/articoli/2015/4/123892.html.}.
+itself\footnote{\url{http://www.ilgiornaledellarte.com/articoli/2015/4/123892.html}.}.
 
 Moreover, the movement produced a proposal to amend article 108 of the Codice dei Beni Culturali regulating the
-reproduction of cultural heritage (https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/), to eliminate the
+reproduction of cultural heritage (\url{https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/}), to eliminate the
 exclusion of printed books and documents from the liberalisation archive, hence complying with copyright and personal
@@
 Free photography means the ability to use one’s own smartphone or camera in archives and libraries during consultation,
-without the need to request prior written approval (for purposes other than profit). A ‘cultural’ exception should be
+without the need to request prior written approval (for purposes other than profit). A `cultural’ exception should be
 provided for a particular form of profit, scientific publishing, given the belief that research is meaningful only if
-it can be disseminated in the broadest way\footnote{\ Circolare Mibact 21/2005.}. “Fotografie libere per i Beni
-Culturali” in fact proposes to make free the publication of images of cultural heritage in scientific texts in
+it can be disseminated in the broadest way\footnote{Circolare Mibact 21/2005.}. ``Fotografie libere per i Beni
+Culturali'' in fact proposes to make free the publication of images of cultural heritage in scientific texts in
 circulation and of a limited cover price (below 77 Euro and 2000 copies) with the sole duty to always specify the name
@@
 authorisation by ordinary post may be substituted by a simple communication online at the institute with the intent to
-publish. For the “greater” publications, other than the usual indicated above, and in general for moneymaking
+publish. For the ``greater'' publications, other than the usual indicated above, and in general for moneymaking
 activities, the concession regime will remain in force, requiring both a formal authorisation request as well as a
@@
 
-\begin{enumerate}
-\item The free reuse cultural heritage images
-\end{enumerate}
-Il tema del riuso delle immagini di beni culturali, cui si è brevemente accennato nell’ambito della proposta di
-“Fotografie libere per i Beni Culturali”, si incrocia strettamente con l’esigenza di EAGLE di pubblicare online
-iscrizioni latine e greche. A questo scopo EAGLE ha stabilito una convenzione con la Direzione Generale per
-l’innovazione tecnologica e la promozione del Ministero dei Beni Culturali in data 21 novembre 2005 per pubblicare sul
-web le riproduzioni di iscrizioni latine e greche già edite anteriori al VII sec., mentre per le iscrizioni inedite si
-autorizzano nuove riproduzioni “fatti salvi eventuali diritti scientifici” degli studiosi che hanno rinvenuto le
-iscrizioni e solo tramite il “consenso” delle soprintendenze interessate. Una seconda convenzione, stipulata nel 2012
-con l’Archivio di Stato di Roma, autorizza EAGLE invece a pubblicare di immagini di documenti di archivio che riportano
-informazioni su iscrizioni.
+\section{The free reuse cultural heritage images}
 
-\ Prendendo spunto dalla proposta di “Fotografie libere per i Beni Culturali” è utile riflettere sulle modalità per
-rendere effettiva la libera divulgazione con ogni mezzo delle immagini di beni culturali per finalità non lucrative già
-prevista dall’art. 108 comma 3 del Codice dei Beni Culturali, così come modificato dall’”Art Bonus” (è libera “la
-divulgazione con qualsiasi mezzo delle immagini di beni culturali, legittimamente acquisite, in modo da non poter
-essere ulteriormente riprodotte a scopo di lucro”). In altre parole sarebbe importante quindi esplicitare a livello
-normativo che la divulgazione delle riproduzioni per fini non lucrativi (e dunque anche la pubblicazione online a scopo
-scientifico), debba rimanere gratuita e libera dalla procedura burocratica vigente che deve oggi accompagnare qualsiasi
-richiesta di concessione alla pubblicazione delle immagini (sostituendola al più, come nel caso delle pubblicazioni a
-stampa, con una comunicazione che informi l’istituto di appartenenza sull’uso culturale dell’immagine).
-
-La British Library ad esempio, che proprio nei mesi scorsi ha liberalizzato le riproduzioni per assecondare le richieste
-dell’utenza, promuove esplicitamente la condivisione sui social networks delle immagini di manoscritti, riconoscendo il
-‘great benefit in sharing images’.
-
-Il riuso delle immagini di beni culturali per finalità educative e scientifiche potrebbe essere regolato attraverso il
-ricorso alle licenze Creative Commons. Il Ministero dei Beni Culturali potrebbe chiarire l’ambito applicativo dell’art.
-108 comma 3, prevedendo il ricorso a licenze CC-BY-NC-SA by default sulle immagini di beni culturali. In questo modo
-verrebbe sancita la libera diffusione delle immagini per finalità diverse dal lucro purché sia sempre specificata la
-provenienza del materiale originale e a condizione che nei riusi successivi vengano attribuite le stesse licenze di
-origine.
-
-The reuse of cultural heritage images, briefly mentioned in the proposal scope for “Fotografie libere per i Beni
-Culturali”, intersects closely with EAGLE’s need to publish Latin and Greek inscriptions online for cultural purposes.
+The reuse of cultural heritage images, briefly mentioned in the proposal scope for ``Fotografie libere per i Beni
+Culturali'', intersects closely with EAGLE’s need to publish Latin and Greek inscriptions online for cultural purposes.
 To this end, EAGLE established an agreement with the Italian Ministry of Culture on the 21st November 2005, to publish
@@
 difficult it is to publish images of inscription – mostly due to the mistrust of those who still view diffusion of
-digital images as a form of “expropriation” and not an occasion of cultural enrichment.
+digital images as a form of ``expropriation'' and not an occasion of cultural enrichment.
 
@@
 satisfy the demands of users, explicitly promotes the sharing of images and manuscripts on social networks, recognising
-the ‘great benefit in sharing images’. An interesting case is the Metropolitan Museum of New York, which has made
+the `great benefit in sharing images’. An interesting case is the Metropolitan Museum of New York, which has made
 available hundreds of thousands of high-resolution photographs of their own works, even allowing their reuse in
@@
 proposes an even more extreme model in giving anyone the possibility to reuse excellent resolution images for any
-purpose, including commercial (!) with a “CC0” license.
+purpose, including commercial (!) with a ``CC0'' license.
 
@@
 collections. Museums, archives and libraries should therefore be seen as active and dynamic centres of cultural
-promotion, more so than bureaucratic offices for the static conservation of cultural heritage\footnote{\ Cfr. on the
-Creative Commons licences in archaeology: Brugnoli, A., Gardini, S., Fotografia digitale, beni archivistici e utenti:
-l’impiego e la diffusione di una nuova tecnologia nella normativa e nelle iniziative dell’amministrazione archivistica,
-in Archivi e Computer, marzo 2013, pp. 213-256; Gualandi, M. L., L’archeologia italiana di fronte alla sfida dell'Open
-Data. Il MOD - Mappa Open Data Archive, in Atti delle Giornate Gregoriane (VII edizione), Bari 2014, pp. 69-75.}.
+promotion, more so than bureaucratic offices for the static conservation of cultural heritage\footnote{Cfr. on the
+Creative Commons licences in archaeology: \citet{brugnoli_fotografia_2013}; \citet{gualandi}.}.
+
+\section{Encouraging first results and European perspectives for project EAGLE}
+
+The effort underlying the collection of thousands of signatures has already led to encouraging openings. A recent
+parliamentary question to the minister on whether to restore the original sprit of the ``Art Bonus''
+decree\footnote{\url{http://www.senato.it/japp/bgt/showdoc/showText?tipodoc=Sindisp\&leg=17\&id=914146}}, and the
+comparable availability shown by the Ministry of Culture, are in fact events which inspire cautious optimism –
+primarily because they mark a return to this topic in parliament after one year. It begins a course that will hopefully
+soon arrive at the modification of article 108 of the ``Codice dei Beni Culturali'' regulating reproduction and the
+release of an appropriate circular which should affirm, without ambiguity:
 
 \begin{enumerate}
-\item Encouraging first results and European perspectives for project EAGLE
-\end{enumerate}
-The effort underlying the collection of thousands of signatures has already led to encouraging openings. A recent
-parliamentary question to the minister on whether to restore the original sprit of the “Art Bonus”
-decree\footnote{\ http://www.senato.it/japp/bgt/showdoc/showText?tipodoc=Sindisp\&leg=17\&id=914146}, and the
-comparable availability shown by the Ministry of Culture, are in fact events which inspire cautious optimism –
-primarily because they mark a return to this topic in parliament after one year. It begins a course that will hopefully
-soon arrive at the modification of article 108 of the “Codice dei Beni Culturali” regulating reproduction and the
-release of an appropriate circular which should affirm, without ambiguity:
-
-1)\ \ the free reproduction in archives and libraries for research purposes and in accordance with complete respect for
+\item the free reproduction in archives and libraries for research purposes and in accordance with complete respect for
 privacy and copyright law;
 
-2)\ \ the free reuse of images of cultural heritage for purposes outside of financial gain, or free publication online
-and in print within the limits mentioned above\footnote{\ Cfr. n. 10.}.
+\item the free reuse of images of cultural heritage for purposes outside of financial gain, or free publication online
+and in print within the limits mentioned above\footnote{Cfr. n. 10.}.
+
+\end{enumerate}
 
@@
 
+\nocite{brugnoli_riproduzione_2013}
+\nocite{casini_gli_2015}
+\nocite{ciociola_libere_2015}
+\nocite{gallo_il_2014}
+\nocite{giunta_viva_2015}
+\nocite{lupoli_libere_2015}
+\nocite{manacorda_fotografare_2014}
+\nocite{manacorda_tutela._2014}
+\nocite{modolo_ricerca_2015}
+\nocite{pavolini_situazione_2015}
+\nocite{pigliaru_selfie_2015}
+\nocite{stella_gabella_2014}
+\nocite{stella_costose_2015}
+\nocite{stella_biblioteche_2015}
+\nocite{tumicelli_questione_2014}
+\nocite{volpe_limmagine_2014}
+\nocite{_lettera_2015}
+\nocite{_appello_2013}
+
 \bibliographystyle{sapauth-eng}

### 2015-12-17 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
paper 40 updated
doc to tex

@@
 %%%%%%%%
-\title{}
-\author{A\&D\&F\&L}
+\title{VISUAL FEATURES OF INSCRIPTIONS. An issue for EDB (and EAGLE)}
+\author[uniba]{Antonio, Enrico FELLE\corref{first}}
+\address[uniba]{Università degli Studi di Bari ``Aldo Moro'' -Italy} 
+\cortext[first]{Corresponding author. Email: antonio.felle@uniba.it}
 \date{2015-11-15}
 \begin{document}
-Abstract
 
-\begin{figure}
-\centering
-\begin{minipage}{10.901cm}
-VISUAL FEATURES OF INSCRIPTIONS.
-
-An issue for EDB (and EAGLE)\newline
-
-Antonio, Enrico FELLE
-
-Università degli Studi di Bari {\textquotedbl}Aldo Moro{\textquotedbl} -Italy
-
-Email: antonio.felle@uniba.it
-\end{minipage}
-\end{figure}
+\maketitle
+\begin{abstract}
 In these last years the amount of digital images of inscriptions increased very quickly: we do not need accurate textual
@@
 
-The issue of the {\textquotedbl}illustrated inscriptions{\textquotedbl} brings us to consider more broadly all the
+The issue of the ``illustrated inscriptions'' brings us to consider more broadly all the
 visual features of inscriptions, that were conceived as objects to see, not only to read.
+\end{abstract}
+\keywords{Early Christian Epigraphy, Byzantine Epigraphy, Middle Ages Epigraphy, Images, Symbols, Signs, Paleography,
+Stonecutters’ workshops}
 
-\bigskip
-
-\bigskip
-
-Keywords
-
-Early Christian Epigraphy, Byzantine Epigraphy, Middle Ages Epigraphy, Images, Symbols, Signs, Paleography,
-Stonecutters’ workshops
-
-\bigskip
-
-1. {\textquotedbl}Illustrated inscriptions{\textquotedbl} by the Christians of Rome.
-
-\bigskip
-
+\section{``Illustrated inscriptions'' by the Christians of Rome.}
 In 2012, during a conference in Rome about Late Antique plates decorated with engravings, I presented a paper about the
@@
 notion and the use of images (signs, symbols, figures and so on) by the Christians of Rome in Late Antiquity, by
-analyzing the inscriptions stored in the database [Felle 2013].
+analyzing the inscriptions stored in the database \citep{felle_apporto_2013}.
 
 Then, the first datum was that only a quarter of these epigraphs displays images or generical non-alphabetical signs
-[Felle 2013, p. 101] \ (fig. 1).
+\citep[101]{felle_apporto_2013} (Fig.~\ref{fig:1}).
 
- \includegraphics[width=10.119cm,height=5.74cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img001.pdf}
-
-Fig. 1. Percentage in EDB of the {\textquotedbl}illustrated inscriptions{\textquotedbl} \ (chart by A. E. Felle).
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img001.pdf}
+\caption{Percentage in EDB of the ``illustrated inscriptions'' (chart by A. E. Felle).}
+\label{fig:1}
+\end{figure}
 
@@
 Carlo Carletti explained this phenomenon as the result of a will to display explicit signs of a religious identity, such
-the Chi-Rho monogram is [Carletti 2008, 68-72; Felle 2007, 365-366]. But we have to underline that the phenomenon is
-not exclusive of Christian patrons. We observe more and more recurrent similar {\textquotedbl}signs of
-identity{\textquotedbl} also among inscriptions commissioned by Jews, not only in Rome but also in other contexts in
-Late Antiquity world where they were [Felle 2007, passim; Felle, in press]. \ \
+the Chi-Rho monogram is (\citet[68-72]{carletti_epigrafia_2008}; \citet[365-366]{felle_judaism_2007}). But we have to underline that the phenomenon is
+not exclusive of Christian patrons. We observe more and more recurrent similar ``signs of
+identity'' also among inscriptions commissioned by Jews, not only in Rome but also in other contexts in
+Late Antiquity world where they were [\citet[passim]{felle_judaism_2007}; \citet{pesce_documenti_2016}].
 
-Going back to the inscriptions by Christians, the use of signa Christi \ in form of monograms strongly reduces the use
-of \ other christological signs or figures, as for instance the anchor: this one, very recurrent during all III
-century, disappears completely and very quickly, already in the very first decades of IV century [Felle 2012, 103]
-(fig. 2).
+Going back to the inscriptions by Christians, the use of signa Christi in form of monograms strongly reduces the use
+of other christological signs or figures, as for instance the anchor: this one, very recurrent during all III
+century, disappears completely and very quickly, already in the very first decades of IV century \citep[103]{felle_apporto_2013}
+(Fig.~\ref{fig:2}). % changed from Felle 2012, not in Bibliography.
 
-\bigskip
-
- [Warning: Image not found]
-
-Fig. 2. Use of different images in inscriptions by Christians of Rome between III and IV cent. (chart by A. E. Felle)
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}
+\caption{Use of different images in inscriptions by Christians of Rome between III and IV cent. (chart by A. E. Felle)}
+\label{fig:2}
+\end{figure}
 
@@
 
-First of all, \ we have to say that in EDB we faced huge difficulties about the treatment of images other than signa
+First of all, we have to say that in EDB we faced huge difficulties about the treatment of images other than signa
 Christi, or generally other non-alphabetical signs, or also captions related to figures on the slabs, and so on.
@@
 pictures, the so-called anaglypha are concisely described by simplified and repetitive clichées to suggest the depicted
-subjects to the readers (fig. 3) or by brief descriptions in Latin in transcriptions or also in the commentaries (fig.
-4).
+subjects to the readers (Fig.~\ref{fig:3}) or by brief descriptions in Latin in transcriptions or also in the commentaries (Fig.~\ref{fig:4}).
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
+\caption{Rome, coemeterium Hippolyti, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.12.14) and
+edition in ICVR, VII 19820.}
+\label{fig:3}
+\end{figure}
 
-\bigskip
-
- \includegraphics[width=10.59cm,height=5.159cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}
-
-\bigskip
-
- \includegraphics[width=10.268cm,height=2.639cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
-
-Fig. 3. Rome, coemeterium Hippolyti, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.12.14) \ and
-edition \ in ICVR, VII 19820.
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
- \includegraphics[width=8.202cm,height=2.514cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img004.png}
-
- \includegraphics[width=9.622cm,height=2.244cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img005.png}
-
-Fig. 4. Rome, coemeterium Maius, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.8.3) and edition in
-ICVR, VIII 22407.
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img004.png}
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img005.png}
+\caption{Rome, coemeterium Maius, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.8.3) and edition in
+ICVR, VIII 22407.}
+\label{fig:4}
+\end{figure}
 
@@
 Some examples: in EDB different verbal descriptions about the some subject are recorded, such as avis uvam pascitur
-(e.g. ICVR, III 8114a, b, c, e), or avis uvas pascitur (ICVR, III 8004b); or \ also avis racemum carpens (ICVR, IX
+(e.g. ICVR, III 8114a, b, c, e), or avis uvas pascitur (ICVR, III 8004b); or also avis racemum carpens (ICVR, IX
 24020), avis racemum carpit (ICVR V, 14157), avis racemum pascitur (ICVR, V 15194), avis racemum rostro carpit (ICVR,
@@
 
-Moreover, recording in EDB descriptions with different words for the same illustrated subject, \ such as
-{\textquotedbl}avis, racemus{\textquotedbl} \ (EDB 19827: ICVR, III 9311, see fig. 5a) or \ {\textquotedbl}avis cum
-racemo{\textquotedbl} (EDB 24933: ICVR, III 8044, see fig. 5b) we are not able to retrieve all the occurrences of this
+Moreover, recording in EDB descriptions with different words for the same illustrated subject, such as
+``avis, racemus'' (EDB 19827: ICVR, III 9311, see Fig.~\ref{fig:5a}) or ``avis cum
+racemo'' (EDB 24933: ICVR, III 8044, see Fig.~\ref{fig:5b}) we are not able to retrieve all the occurrences of this
 same subject, because they are recorded (both n ICVR and in EDB) in different ways.
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img006.png}
+\caption{Rome, catacomb of Domitilla. Edition of ICVR, III 9311}
+\label{fig:5a}
+\end{figure}
 
- \includegraphics[width=5.076cm,height=2.21cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img006.png}  a
-\ \ \ \ \ \ b 
-\includegraphics[width=4.207cm,height=2.078cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img007.png}
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img007.png}
+\caption{Rome, catacomb of Domitilla. Edition of ICVR, III 8044}
+\label{fig:5b}
+\end{figure}
 
-Fig. 5. Rome, catacomb of Domitilla. Edition of \ ICVR, III 9311 (a) and \ ICVR, \ III 8044 (b). \ \ \ \ \ \
-
-\bigskip
-
-\ This ambiguity prevents to retrieve all the occurrences of the same illustrated subjects and then they adulterate the
+This ambiguity prevents to retrieve all the occurrences of the same illustrated subjects and then they adulterate the
 result of our queries: I think that we have to correct as soon as possible this ambiguity, in order to establish an
@@
 In these last years - also with the kind help by the Photographic Archive of Papal Commission of Sacred Archaeology
-(http://www.archeologiasacra.net/pcas-web/) - \ the amount of images available in EDB increased very quickly: this is
+(\url{http://www.archeologiasacra.net/pcas-web/}) - the amount of images available in EDB increased very quickly: this is
 surely an advantage in respect to the situation of only three years ago. At the present day and more over in the
@@
 possibility to easily view a photo or a drawing of an inscription does not solve the issues related to search and to
-retrieve inscriptions bearing given kinds of image, or specific signs, and so on. \
+retrieve inscriptions bearing given kinds of image, or specific signs, and so on.
 
@@
 
-\bigskip
+\section{Visual features of the 'written monuments'.}
 
-2. Visual features of the 'written monuments'. \
-
-The issue of the {\textquotedbl}illustrated inscriptions{\textquotedbl} brings us to consider more broadly the visual
+The issue of the ``illustrated inscriptions'' brings us to consider more broadly the visual
 features of inscriptions overall. The ancient epigraphs are conceived not only as texts to be read (very few people
@@
 
-In our projects - we have to admit it - \ the notion that inscriptions are essentially texts is still largely
+In our projects - we have to admit it - the notion that inscriptions are essentially texts is still largely
 prevailing: but now we can - consequently, we must - to increase our capability to represent, to record and so to
-investigate \ also other, visual features of inscriptions.
+investigate also other, visual features of inscriptions.
 
-\bigskip
-
-2.1. Positioning
+\subsection{Positioning}
 
@@
 Österreichische Akademie der Wissenschaften and the Deutsches Archäologisches Institut) focused to the frescoes of the
-catacomb of Domitilla along via Ardeatina, the largest one in Rome suburb [Felle - Zimmermann 2014]. The positive
-collaboration between our two projects \ continued: now we are able to offer to EDB users to view on the updated map of
+catacomb of Domitilla along via Ardeatina, the largest one in Rome suburb \citet{zimmermann_epigraphy_2014}. The positive
+collaboration between our two projects continued: now we are able to offer to EDB users to view on the updated map of
 the catacomb of Domitilla the distribution of the inscriptions still in situ in the four levels of the subterranean
@@
 
-\bigskip
-
-2.2. Materials and shape
+\subsection{Materials and shape }
 
 Materials and shape of the written objects communicate im-mediately, to all, before the inscribed texts. Because now we
-can do it, we have to display to the users of our projects the real communicative power of the inscriptions, \ that
+can do it, we have to display to the users of our projects the real communicative power of the inscriptions, that
 assume much more meaningfulness when we can see them than when we can read their only texts in a library. By pictures
 and 3D rendering of the places where the epigraphs were arranged we should be able also to provide virtual images of
-inscriptions \ {}- also lost or incomplete - conceived to be seen and read exactly there. An effective example can be
-the dedication by pope Damasus (366-384) to the martyr Ianuarius in the catacomb of Praetextatus on the via Appia (fig.
-6).
+inscriptions {}- also lost or incomplete - conceived to be seen and read exactly there. An effective example can be
+the dedication by pope Damasus (366-384) to the martyr Ianuarius in the catacomb of Praetextatus on the via Appia (Fig.~\ref{fig:6}).
 
- \includegraphics[width=8.692cm,height=4.21cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img008.jpg}
-
- \includegraphics[width=10.141cm,height=9.059cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img009.jpg}
-
-Fig. 6. Rome, catacomb of Praetextatus. Damasus’ dedicatory inscription (ICVR, V 13871) for the martyr Ianuarius (Photo:
-PCAS; 3D rendering: G. De Felice).
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img008.jpg}
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img009.jpg}
+\caption{Rome, catacomb of Praetextatus. Damasus’ dedicatory inscription (ICVR, V 13871) for the martyr Ianuarius (Photo:
+PCAS; 3D rendering: G. De Felice)}
+\label{fig:6}
+\end{figure}
 
@@
 white marble over the tomb of the martyr, well-lighted by a skylight made on purpose, in strong contrast to the
-darkness of the neighbouring galleries of the subterranean cemetery. \
+darkness of the neighbouring galleries of the subterranean cemetery.
 
@@
 has no need to read, to perceive the actual and effective messages displayed by inscriptions placed within the fabric
-of the walls of Byzantine cities, such as Constantinople or Thessaloniki (fig. 7).
+of the walls of Byzantine cities, such as Constantinople or Thessaloniki (Fig.~\ref{fig:7}).
 
- \includegraphics[width=11.1cm,height=5.673cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}
-
-Fig. 7. Thessaloniki, city walls. Inscriptions in the masonry of a tower near Eptapyrgion (photo: A. E. Felle).
-
-\bigskip
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}
+\caption{Thessaloniki, city walls. Inscriptions in the masonry of a tower near Eptapyrgion (photo: A. E. Felle).}
+\label{fig:7}
+\end{figure}
 
@@
 opinion, this notion is clearly demonstrated by the unnecessary captions in the icons and in the images of martyrs
-(fig. 8), where \ {}- on closer view - the inscriptions are completely useless, if we continue to consider them only as
+(Fig.~\ref{fig:8}), where \ {}- on closer view - the inscriptions are completely useless, if we continue to consider them only as
 texts to be read.
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img011.jpg}
+\caption{Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
+“useless” caption  s(an)c(t)a Agnes (photo: A. E. Felle).}
+\label{fig:8}
+\end{figure}
 
- \includegraphics[width=10.749cm,height=9.208cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img011.jpg}
-
-\bigskip
-
-Fig. 8. Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
-“useless” caption \ s(an)c(t)a Agnes (photo: A. E. Felle).
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-2.3. Relationship with the context
+\subsection{Relationship with the context}
 
@@
 inscriptions: rather, the main condition appears their relationship with their contexts. An incisive example can be
-offered by the Christian inscriptions bearing biblical quotations [cfr. Felle 2006]: in the middle of the bronze
+offered by the Christian inscriptions bearing biblical quotations \citep{felle_biblia_2006}: in the middle of the bronze
 plating of the marble lintel over the Great Door of the Royal Gates of the Haghia Sophia in Constantinople, an empty
-throne is occupied by an open codex, according to the words by Kähler, {\textquotedbl}the only extant plastic
-composition dating from the founding period of the church{\textquotedbl} [Kähler 1967, pp. 29-30; 32 taff. 22; 62]
-(fig. 9).
-
-\ Fig
-
-\bigskip
+throne is occupied by an open codex, according to the words by Kähler, ``the only extant plastic
+composition dating from the founding period of the church'' \citep[29-30; 32][taff. 22; 62]{kahler_hagia_1967}
+(Fig.~\ref{fig:9}).
 
@@
 
-\bigskip
+\begin{quotation}
+John 10.7: \begin{otherlanguage*}{greek}
+Εἶπεν οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι ἐγώ εἰμι ἡ θύρα τῶν προβάτων.
+\end{otherlanguage*} (Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep);
 
-John 10.7: \textgreek{E>~ipen} \textgreek{o>~un p'alin <o >Ihso~uc, >Am`hn >am`hn l'egw <um~in <'oti }\textgreek{>eg'w
-e>imi <h j'ura t~wn prob'atwn}. (Therefore Jesus said again,~‘Very truly I tell you, I am the gate for the sheep);
+John 10, 9: \begin{otherlanguage*}{greek}
+ἐγώ εἰμι ἡ θύρα· δι' ἐμοῦ ἐάν τις εἰσέλθῃ σωθήσεται καὶ εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει
+\end{otherlanguage*}  ( 9 I am the gate;
+whoever enters through me will be saved. [a] They will come in and go out, and find pasture).
 
-[Warning: Draw object ignored]John 10, 9: \textgreek{>eg'w e>imi <h j'ura· }\textgreek{di' >emo~u >e'an tic e>is'eljh|}
-\textgreek{swj'hsetai ka`i }\textgreek{e>isele'usetai ka`i >exele'usetai ka`i nom`hn e<ur'hsei} \ ( 9I am the gate;
-whoever enters through me will be saved.[a]~They will come in and go out, and find pasture).
-
-\begin{figure}
+\end{quotation}
+\begin{figure}[!bp]
 \centering
-\includegraphics[width=4.419cm,height=7.329cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img012.jpg}
-\end{figure}
-\begin{figure}
-\centering
-\includegraphics[width=6.655cm,height=6.509cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img013.jpg}
-\end{figure}
-\begin{figure}
-\centering
-\begin{minipage}{0.688cm}
-
-\bigskip
-\end{minipage}
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img012.jpg}
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img013.jpg}
+\caption{Istanbul, Hagia Sophia. On the left, the Royal Gates in the narthex. On the right, the particular of the image
+of the throne with open inscribed codex above the lintel of the central gate (from Kähler 1967).}
+\label{fig:9}
 \end{figure}
 
-\bigskip
+This the text of the inscription \citep[n. 505]{felle_biblia_2006}:
 
-\bigskip
+\begin{quotation}
+((crux)) \begin{otherlanguage*}{greek}
+εἶπεν ὁ κ(ύριο)ς | ἐγώ εἰμι | ἡ θύρα τῶν | προβάτων· | δι᾿ ἐμοῦ || ἐάν τις | εἰσέλθῃ | εἰσελεύσετ(αι) | κ(αὶ) ἐξελεύσετ(αι) | κ(αὶ) νομὴν | εὑρήσει.
+\end{otherlanguage*}
 
-Fig. 9. Istanbul, Hagia Sophia. On the left, the Royal Gates in the narthex. On the right, the particular of the image
-of the throne with open inscribed codex above the lintel of the central gate (from Kähler 1967).
-
-\bigskip
-
-This the text of the inscription [Felle 2006, n. 505]:
-
-((crux)) \textgreek{e>~ipen <o k('urio)c {\textbar} >eg'w e>imi {\textbar} <h j'ura t~wn {\textbar} prob'atwn;
-{\textbar} di> >emo~u {\textbar}{\textbar} >e'an tic {\textbar} e>is'eljh| {\textbar} e>isele'uset(ai) {\textbar}
-k(a`i) >exele'uset(ai) {\textbar} k(a`i) nom`hn {\textbar} e<ur'hsei.}
-
-\bigskip
+\end{quotation}
 
@@
 (not necessarily the readability) of the sacred words give proper and strong sense to their material support and to
-entire context, the Royal Gates of the Great Church of Constantinoples [Felle 2015, p. 320 and passim].
+entire context, the Royal Gates of the Great Church of Constantinoples \citep[320 and passim]{felle_esporre_2015}.
 
-\bigskip
-
-2.4. Writing
+\subsection{Writing}
 
@@
 intrinsecally meaningful such as visual element of the equipment of a simple or rich funerary monument or of a cultual
-building. Then, we have to face the issue of the description of the writing not only \ from the \ necessary point of
+building. Then, we have to face the issue of the description of the writing not only from the necessary point of
 view of paleography ( we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
 also in order to perceive and to understand its non-verbal significance: by its disposition, direction, shape. The
-notion of the inscriptions in the Islamic world, \ where often the letters are also - and maybe firstly - images (they
+notion of the inscriptions in the Islamic world, where often the letters are also - and maybe firstly - images (they
 are used as decorative friezes, architectural ornaments, figures) and their clearity and readability are not considered
-as necessary, can help us to evocate this feature of the writing (fig. 10).
+as necessary, can help us to evocate this feature of the writing (Fig.~\ref{fig:10}).
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img014.jpg}
+\caption{Granada, Alhambra. An example of the writing as decorative frieze and architectural ornaments (Photo: A. E.
+Felle)}
+\label{fig:10}
+\end{figure}
 
- \includegraphics[width=11.044cm,height=8.281cm]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img014.jpg}
-
-Fig. 10. Granada, Alhambra. An example of the writing as decorative frieze and architectural ornaments (Photo: A. E.
-Felle)
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-3. Conclusions
+\section{Conclusions}
 
@@
 from our standard: to see an inscription with the same point of view of the most part of the citizens in Roman Empire -
-and mostly in Late Antiquity - \ we have to become, in some way, illiterate.
+and mostly in Late Antiquity - we have to become, in some way, illiterate.
 
-In conclusion: we have to consider in digital descriptions of the inscriptions some their {\textquotedbl}visual
-features{\textquotedbl} that in our projects - first of all in EDB, of course - are not too considered, although they
+In conclusion: we have to consider in digital descriptions of the inscriptions some their``visual
+features'' that in our projects - first of all in EDB, of course - are not too considered, although they
 are very significant. We need, about encoding these non-verbal features, the same positive results that by Epidoc we
-reached \ in encoding the texts: a hard challenge.
+reached in encoding the texts: a hard challenge.
 

### 2015-12-17 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
updated paper 42
added no indent to paper 40, doc to tex for paper 42

@@
 \section{``Illustrated inscriptions'' by the Christians of Rome.}
-In 2012, during a conference in Rome about Late Antique plates decorated with engravings, I presented a paper about the
+\noindent In 2012, during a conference in Rome about Late Antique plates decorated with engravings, I presented a paper about the
 potentially very useful contribute that the Epigraphic Database Bari (EDB) could offer \ to study and to interpret the
@@
 
-The issue of the ``illustrated inscriptions'' brings us to consider more broadly the visual
+\noindent The issue of the ``illustrated inscriptions'' brings us to consider more broadly the visual
 features of inscriptions overall. The ancient epigraphs are conceived not only as texts to be read (very few people
@@
 
-First of all, I think to the positioning of the epigraphs in the contexts for which they were created. About EDB, we
+\noindent First of all, I think to the positioning of the epigraphs in the contexts for which they were created. About EDB, we
 already presented in the EAGLE Conference in Paris in 2014 a first attempt to record and to describe the exact
@@
 
-Materials and shape of the written objects communicate im-mediately, to all, before the inscribed texts. Because now we
+\noindent Materials and shape of the written objects communicate im-mediately, to all, before the inscribed texts. Because now we
 can do it, we have to display to the users of our projects the real communicative power of the inscriptions, that
@@
 
-Indeed - more over in Late Antiquity and Middle Ages - readability of the texts is not the main property of
+\noindent Indeed - more over in Late Antiquity and Middle Ages - readability of the texts is not the main property of
 inscriptions: rather, the main condition appears their relationship with their contexts. An incisive example can be
@@
 
-Scarce or null readability does not imply low quality of the appearance of writing: rather, the writing appears
+\noindent Scarce or null readability does not imply low quality of the appearance of writing: rather, the writing appears
 intrinsecally meaningful such as visual element of the equipment of a simple or rich funerary monument or of a cultual
@@
 
-The ancient inscriptions actually belong to civilizations where the literacy - with very few exception - was very far
+\noindent The ancient inscriptions actually belong to civilizations where the literacy - with very few exception - was very far
 from our standard: to see an inscription with the same point of view of the most part of the citizens in Roman Empire -

### 2015-12-17 source/42/Paper_licciardello.tex
updated paper 42
added no indent to paper 40, doc to tex for paper 42

@@
 %TITLE:%
-%%%%%%%%\title{}
-\title{}
-\author{Flavia Licciardello}
+%%%%%%%%
+\title{DEIXIS AND FRAMES OF REFERENCE IN DEDICATORY EPIGRAMS. The use of a database with an interdisciplinary approach}
+\author[topoi]{Flavia Licciardello\corref{first}}
+\address[topoi]{Humboldt Universität zu Berlin, Excellence Cluster TOPOI}
+\cortext[first]{Corresponding author. Email: flavia.licciardello@topoi.org}
 \date{2015-11-20}
 \begin{document}
-[Warning: Draw object ignored]Abstract
 
-\ \ \ \ \ \ This paper aims to present an example of a database designed to combine epigraphic, linguistic and
+\maketitle
+\begin{abstract}
+This paper aims to present an example of a database designed to combine epigraphic, linguistic and
 philological data. This database is part of my project on a study of deictic expressions in dedicatory inscribed and
@@
 extract information on trends and recurrent patterns in the genre.
+\end{abstract}
+\keywords{Greek epigram, dedication, database, deixis, Hellenistic age}
 
-\bigskip
+\section{Introduction}
 
-Keywords
-
-\ Greek epigram, dedication, database, deixis, Hellenistic age.
-
-\bigskip
-
-1.1 Introduction
-
-For a long time, the study of the Greek epigrams has maintained a clear distinction between ‘epigraphic’ and ‘literary’
+\noindent For a long time, the study of the Greek epigrams has maintained a clear distinction between `epigraphic' and `literary'
 epigrams. This distinction, however, fails to understand the complexity and the development of this poetic genre. Only
 in very recent years have scholars started to highlight properly the important relationship between Hellenistic
-‘literary’ epigrams and epigraphic models.\footnote{ The shift in the attitude towards the history of the epigrammatic
-genre is well resumed by [Garulli 2012, 5-34]. For other work, which consider the importance of the dialogue between
-‘epigraphic’ and ‘literary’ epigrams, see e.g. [Meyer 2005], [Tueller 2008]. On archaic and classical epigrams, see
-[Baumbach-Petrovic-Petrovic 2010]. Further bibliography in [Baumbach-Petrovic-Petrovic 2010, 2].}
+`literary' epigrams and epigraphic models.\footnote{ The shift in the attitude towards the history of the epigrammatic
+genre is well resumed by \citep[5-34]{garulli_byblos_2012}. For other work, which consider the importance of the dialogue between
+`epigraphic' and `literary' epigrams, see e.g. \citep{meyer_inszeniertes_2005}, \citep{tueller_look_2008}. On archaic and classical epigrams, see
+\citep{baumbach_archaic_2010}. Further bibliography in \citep[2]{baumbach_archaic_2010}.}
 
 My PhD project follows this new exegetic line and contributes to our understanding of the inscriptional component of the
-Hellenistic epigram.\footnote{ The project is developed within the frame of the C-1 group (‘Deixis and frame of
+Hellenistic epigram.\footnote{ The project is developed within the frame of the C-1 group (`Deixis and frame of
 reference) of the Excellence Cluster TOPOI.} In particular, I focus on the use of deictic expressions in dedicatory
 epigrams, i.e. on the use of all those linguistic elements whose meaning and interpretation depend on the spatial and
-temporal context where they are uttered.\footnote{ For an overview of the concept of deixis see [Diessel 2012] (with
-further bibliography). On deixis in Ancient Greek see e.g. [Felson 2004], [Edmunds 2008] and [Bonifazi 2014].} The
+temporal context where they are uttered.\footnote{ For an overview of the concept of deixis see \citep{diessel_deixis_2012} (with
+further bibliography). On deixis in Ancient Greek see e.g. \citep{felson_poetics_2004}, \citep{edmunds_deixis_2008} and \citep{bonifazi_deixis_2014}.} The
 corpus analysed includes inscribed epigrams from the archaic epoch until the end of the 4th century BC, and the
-Hellenistic epigrams transmitted in the Greek Anthology or on papyrus. The reference editions used are [Hansen CEG 1-2]
-for inscribed epigrams and [Gow-Page 1965] and [Austin-Bastianini 2002] for ‘literary epigrams’.\footnote{ For
-convenience, I will later on refer to this second group as ‘literary epigrams’. The definition simply identifies those
+Hellenistic epigrams transmitted in the Greek Anthology or on papyrus. The reference editions used are \citep[1-2]{hansen_carmina_1983}
+for inscribed epigrams and \citep{gow_greek_1965} and \citep{austin_posidippi_2002} for `literary epigrams'.\footnote{ For
+convenience, I will later on refer to this second group as `literary epigrams'. The definition simply identifies those
 epigrams that are transmitted in the Greek Anthology or on papyrus and does not imply any value judgment.}
 
-\ \ In order to deal with this heterogeneous material, I developed a database to register and organise the data obtained
+In order to deal with this heterogeneous material, I developed a database to register and organise the data obtained
 from the analysis of the texts. The specific aim of this database is the registration of all relevant linguistic
@@
 
-\bigskip
+\section{Greek dedicatory epigrams and the role of deixis}
 
-1.2 Greek dedicatory epigrams and the role of deixis
-
-In the Ancient Greek world it was customary to accompany a dedication to the gods with a – usually – short inscription.
+\noindent In the Ancient Greek world it was customary to accompany a dedication to the gods with a – usually – short inscription.
 This was normally chiselled on the dedicated object (or on its support) and it recorded the main information related to
 the offering. The elements recurring in such inscriptions were the name of the dedicator, the verb of dedication and
-the name of the god receiving the dedication.\footnote{ On archaic dedications, see [Day 2010], in part. [p. 1-14] for
+the name of the god receiving the dedication.\footnote{ On archaic dedications, see \citep[ 1-14]{day_archaic_2010} for
 an introduction to the genre.} The most common and widespread formula for dedications contained exactly these three
-elements: \textgreek{<o de~ina >an'ejhke t~w| je~w|. }A frequent variation was \textgreek{<o de~in'a me >an'ejhke t~w|
-je~w|,}\footnote{ The formula here presented (with its two variants) was \ identified by Maria Letizia Lazzarini, in
-her study on the formulas of archaic dedications [Lazzarini 1976, 58-40]. Her analysis is based on both verse and prose
+elements: \begin{otherlanguage*}{greek}ὁ δεῖνα ἀνέθηκε τῷ θεῷ\end{otherlanguage*}.
+A frequent variation was \begin{otherlanguage*}{greek}ὁ δεῖνά με ἀνέθηκε τῷ θεῷ\end{otherlanguage*}, \footnote{ The formula here presented (with its two variants) was \ identified by Maria Letizia Lazzarini, in
+her study on the formulas of archaic dedications \citep[58-40]{lazzarini_formule_1976}. Her analysis is based on both verse and prose
 inscriptions from the archaic age. However, such basic scheme, with these elements, was normal in epigrams as well and
 it remained substantially unchanged in classical and post-classical time.} \ where the speaker is the object itself, as
-is clear from the employment of the personal pronoun \textgreek{me. }It is interesting to note that since their first
+is clear from the employment of the personal pronoun \begin{otherlanguage*}{greek}με\end{otherlanguage*}. It is interesting to note that since their first
 examples the speaking object was a recurring feature in dedicatory epigrams and it is still frequent in the 4th century
-BC.\footnote{ On the topos of the speaking object, see [Burzachechi 1962], [Svenbro 1988, 36-52], [Tueller 2008,
-16-27], [Furley 2010, 151-166], [Wachter 2010, 250-260], [Christian 2015, 29-107]. \ } Clearly, the basic scheme here
+BC.\footnote{ On the topos of the speaking object, see \citep{burzachechi_oggetti_1962}, \citep[36-52]{svenbro_phrasikleia:_1988}, \citep[16-27]{tueller_look_2008}, \citep[151-166]{furley_life_2010}, \citep[250-260]{wachter_origin_2010}, \citep[29-107]{christian_gebildete_2014}.} Clearly, the basic scheme here
 presented could be varied by omitting some elements (as the name of the dedicator) or adding others (as the generic
-name of the object, like the recurrent \textgreek{>'agalma).}
+name of the object, like the recurrent \begin{otherlanguage*}{greek}ἄγαλμα\end{otherlanguage*}).
 
-\ \ A crucial moment in the evolution of the epigrammatic genre is the beginning of the Hellenistic era, when epigrams
+A crucial moment in the evolution of the epigrammatic genre is the beginning of the Hellenistic era, when epigrams
 began to be considered full literary creations and started to be circulated autonomously, no longer limited to one
 inscription alone.\footnote{ This trend may be linked to the custom prevalent in the 5th and 4th century BC to copy
-epigrams on monuments and then quote them in various texts. See [Gutzwiller 1998, 47ff.]} This development led to the
+epigrams on monuments and then quote them in various texts. See \citep[47ff.]{gutzwiller_poetic_1998}} This development led to the
 composition of literary epigrams not intended for inscription, but which in some way maintained the illusion of a
@@
 be reinvented. The primary reason for such changes is self-evident: the monument intended for inscription and the space
-surrounding it do not exist anymore – they have to be imagined into reader’s mind.\footnote{ This involvement of the
-reader can be put in relation with what [Bing 1995] defines Ergänzungsspiel, i.e. with the process of supplementation
+surrounding it do not exist anymore – they have to be imagined into reader's mind.\footnote{ This involvement of the
+reader can be put in relation with what 
+%\citep{Bing1995} 
+defines Ergänzungsspiel, i.e. with the process of supplementation
 of the text deliberately incited by Hellenistic epigrammatists.}
+%reference to Bing 1995 not in biblio, asked to Licciardello
 
@@
 typically used in inscriptions to lead the gaze of the reader towards the dedicated object. In this epigraphic context
-the deictic expressions point to something in front of reader’s eyes (deixis ad oculos), whereas in a literary context
+the deictic expressions point to something in front of reader's eyes (deixis ad oculos), whereas in a literary context
 the readers will have to imagine in their mind the invisible referents of deictic expressions (deixis am
 Phantasma).\footnote{ The difference between deixis ad oculos and deixis am Phantasma was highlighted and described for
-the first time by Bühler in 1934, see [Bühler 1982, 121-126, 133-135].} If on the one hand this change requires a
+the first time by Bühler in 1934, see \citet[121-126, 133-135]{buhler_sprachtheorie:_1982}.} If on the one hand this change requires a
 particular attention to the verbal reconstruction of the setting, on the other hand, the loss of the material context
@@
 centre, which normally coincides with the origin of the utterance, works as a point of reference for all deictic
-markers and expression.\footnote{ See [Bühler 1982, 102f.] and [Levinson 1983, 63f.].} In the case of dedicatory
+markers and expression.\footnote{See \citet[102f.]{buhler_sprachtheorie:_1982} and \citet[63f.]{levinson_pragmatics_1983}.} In the case of dedicatory
 epigrams on stone, the deictic centre in spatial terms is generally understood to be in the place of the inscription
 itself, which in most cases is the dedicated object. In other words, for all deictic expressions that point to
-something close (such as the proximal \textgreek{<'ode ‘}this here’\footnote{ ‘Proximal’ deictic elements are all those
-elements which refers to the deictic centre: among these demonstrative pronouns or adverbs such as ‘this’ or ‘here’ and
-temporal adverbs such as ‘now’. See [Diessel 2012, 2408f.].}) the occasional reader will look for the referents in the
+something close (such as the proximal \begin{otherlanguage*}{greek}ὅδε\end{otherlanguage*} this here'\footnote{ `Proximal' deictic elements are all those
+elements which refers to the deictic centre: among these demonstrative pronouns or adverbs such as `this' or `here' and
+temporal adverbs such as `now'. See \citet[2408f.]{diessel_deixis_2012}.}) the occasional reader will look for the referents in the
 space close to the inscription. In the evolution of the epigrammatic genre, the loss of the material context produces
@@
 
-\ \ This progressive detachment of the text from its original physical location played a crucial role in the process
+This progressive detachment of the text from its original physical location played a crucial role in the process
 that led, out of trivial verse inscriptions, to the emergence of the epigram as a full literary genre. Deictic
@@
 
-1.3 The database
+\section{The database}
 
-In the first step of my research, the analysis of the texts is followed by the registration into a database of all the
+\noindent In the first step of my research, the analysis of the texts is followed by the registration into a database of all the
 relevant data related to spatial, personal and temporal deixis. This database is conceived first of all as a tool to
@@
 
-The corpus of epigrams analysed includes dedicatory verse-inscriptions collected in Hansen’s Carmina Epigraphica Graeca
+The corpus of epigrams analysed includes dedicatory verse-inscriptions collected in Hansen's Carmina Epigraphica Graeca
 [CEG 1-2], as for epigrams on stone. In this case, I selected those epigrams included by Hansen in the section “Tituli
-dedicatorii”. For literary epigrams, I followed Gow’s and Page’s Hellenistic epigrams [Gow-Page 1965], to which I added
-the epigrams attributed to Posidippus and edited by Austin and Bastianini [Austin-Bastianini 2002]. In this case, the
+dedicatorii”. For literary epigrams, I followed Gow's and Page's Hellenistic epigrams \citep{gow_greek_1965}, to which I added
+the epigrams attributed to Posidippus and edited by Austin and Bastianini \citep{austin_posidippi_2002}. In this case, the
 selection was made according to the contents of the epigrams and I selected all those epigrams that have a clear
@@
 aspects.\footnote{ The software used is FileMaker Pro 13.0v4.} The data were organised in three main tables, connected
-to each other through a central one (See Fig. 1.1).
+to each other through a central one (See Fig.~\ref{fig:1}).
 
-[Warning: Draw object ignored]
-
-Fig. \stepcounter{Figura}{\theFigura}.1
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{schema.png}
+\caption{default}
+\label{fig:1}
+\end{figure}
 
@@
 specific feature (and its interconnection with other related aspects) within the corpus and to leave out information
-that is not immediately consistent with the research done. When necessary, the tool ‘portal’ helps to find the data
+that is not immediately consistent with the research done. When necessary, the tool `portal' helps to find the data
 rapidly and to combine the data contained in different tables. In this way, for example, it is possible to look for the
-recurrence of a specific tense for the verb of dedication (in the table ‘deixis’) and to see if this is associated with
-the use of a specific verb of dedication (in the table ‘dedicatory\_context’) or with a particular epoch (in the table
-‘archeological\_context’).
+recurrence of a specific tense for the verb of dedication (in the table `deixis') and to see if this is associated with
+the use of a specific verb of dedication (in the table `dedicatory\_context') or with a particular epoch (in the table
+`archeological\_context').
 
-\bigskip
+\section{An example: the present tense in dedicatory epigrams}
 
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-1.4 An example: the present tense in dedicatory epigrams
-
-The verbal tenses can work as temporal deictic markers and can play an important role in the definition of the temporal
+\noindent The verbal tenses can work as temporal deictic markers and can play an important role in the definition of the temporal
 frame.\footnote{ Though the grammatical category of tense does not always coincide with the semantic category of time,
@@
 always consider other linguistic elements, which contribute determine the temporal frame of the text. On temporal
-deixis and verbal tense see [Lyons 1977, 677-690]; [Levinson 1983, 73-79]; [Klein 1994, 14-26 and 120-130]. For the
-analysis of temporal deixis in Ancient Greek texts see e.g. [D’Alessio 2005] and [Edmunds 2008, 8f.] (with further
+deixis and verbal tense see \citet[677-690]{lyons_semantics_1977}; \citet[73-79]{levinson_pragmatics_1983}; \citet[14-26 and 120-130]{klein_time_1994}. For the
+analysis of temporal deixis in Ancient Greek texts see e.g. \citep{dalessio_past_2004} and \citet[8f.]{edmunds_deixis_2008} (with further
 bibliography).} In particular, the present tense usually decodes the present time, which is the time of the deictic
-centre (‘now’). Since the deictic centre operates as a point of reference for the spatial and temporal orientation,
+centre (`now'). Since the deictic centre operates as a point of reference for the spatial and temporal orientation,
 individuating the deictic centre helps determine the frame of reference of the epigram. This could lay, for example, in
 the act of dedication celebrated, in the moment of the composition by the poet or in the moment of reading.\footnote{
-On the relation between the temporal frame of the utterance and the deictic centre, see [Levinson 1983, 79f.] and
-[D’Alessio 2004].} \
+On the relation between the temporal frame of the utterance and the deictic centre, see \citet[79f.]{levinson_pragmatics_1983} and
+\citep{dalessio_past_2004}.} \
 
@@
 Out of 422 epigrams analysed, we find only 5 clear examples, from different epochs and geographical areas: [CEG 192i]
-(Athens, ca. 520? BC \textgreek{>anakeima[i]), [}CEG 302] (Attica, found in Ptoion, ca. 540? BC v. 1
-\textgreek{e>im>)}\footnote{ For the verb \textgreek{e>im'i }in dedicatory formulas, see [Lazzarini 1976, 59f.].} [CEG
-251] (Athens, ca. 500-480? BC v.1 \textgreek{e>im`i}), [CEG 390] (Apollonia Illyrica, found in Olympia, ca. 450-440? BC
-v.1 \textgreek{>a\textlatin{[323?]}}[E32D?]\textgreek{nake'imeja}), [CEG 822] (Geronthrai, 4th cent.? BC v.1
-\textgreek{>an'akeitai).}\footnote{ In addition to these, in [CEG 347] and [CEG 775i], the verbs
-\textgreek{>anake'im]eja }and \textgreek{kos]mo~umen }respectively are supplied. In [CEG 830ii] the verb
-\textgreek{>an'akeitai }refers to another dedication, not to the one celebrated by the epigram.} In all these cases,
+(Athens, ca. 520? BC \begin{otherlanguage*}{greek}ἀνακειμα[ι]\end{otherlanguage*}), [CEG 302] (Attica, found in Ptoion, ca. 540? BC v. 1
+\begin{otherlanguage*}{greek}εἰμ᾿\end{otherlanguage*})\footnote{ For the verb \begin{otherlanguage*}{greek}εἰμί\end{otherlanguage*} in dedicatory formulas, see \citet[59f.]{lazzarini_formule_1976}.} [CEG
+251] (Athens, ca. 500-480? BC v.1 \begin{otherlanguage*}{greek}εἰμὶ\end{otherlanguage*}), [CEG 390] (Apollonia Illyrica, found in Olympia, ca. 450-440? BC
+v.1 \begin{otherlanguage*}{greek}ἀ̣νακείμεθα\end{otherlanguage*}), [CEG 822] (Geronthrai, 4th cent.? BC v.1
+\begin{otherlanguage*}{greek}ἀνάκειται\end{otherlanguage*}).\footnote{ In addition to these, in [CEG 347] and [CEG 775i], the verbs
+\begin{otherlanguage*}{greek}ἀνακείμ]εθα\end{otherlanguage*} and \begin{otherlanguage*}{greek}κοσ]μοῦμεν\end{otherlanguage*} respectively are supplied. In [CEG 830ii] the verb
+\begin{otherlanguage*}{greek}ἀνάκειται\end{otherlanguage*} refers to another dedication, not to the one celebrated by the epigram.} In all these cases,
 the grammatical subject of the verb is the dedicated object. The use of the present form refers therefore to the
@@
 dedications on stone, the object is rarely the speaker. It is also interesting to note that the formulaic
-\textgreek{>an'akeimai }is frequently substituted by the simple form \textgreek{ke'imai. }Similarly, in literary
-epigrams the simple \textgreek{t'ijhmi }is preferred to \textgreek{>anat'ijhmi, }which is traditional in epigraphic
+\begin{otherlanguage*}{greek}ἀνάκειμαι\end{otherlanguage*} is frequently substituted by the simple form \begin{otherlanguage*}{greek}κείμαι\end{otherlanguage*}. Similarly, in literary
+epigrams the simple \begin{otherlanguage*}{greek}τίθημι\end{otherlanguage*} is preferred to \begin{otherlanguage*}{greek}ἀνατίθημι\end{otherlanguage*}, which is traditional in epigraphic
 examples.} An important innovation is the fact that in some epigrams the subject of the verb of dedication in present
@@
 
-1.5 Conclusion
+\section{Conclusion}
 
-The database presented here is designed to be a helpful tool in the study of heterogeneous material. This tool will
+\noindent The database presented here is designed to be a helpful tool in the study of heterogeneous material. This tool will
 allow us to apply an interdisciplinary approach, which combines epigraphic, linguistic and philological strategies. It
@@
 
-\ \ It is important to stress that this database aims to enable many possible, different research. The structure is not
+It is important to stress that this database aims to enable many possible, different research. The structure is not
 based on a single, pre-fixed hypothesis of investigation, but is built so as to let the user interrogate the data in
@@
 
-\ \ Moreover, the database is a dynamic tool and suitable for further and continual additions. This feature is
+Moreover, the database is a dynamic tool and suitable for further and continual additions. This feature is
 particularly valuable in the field of the epigrammatic poetry, where continual discoveries of new material (on stone,
@@
 here is not only a response to the recent need, in the specific field of epigrammatic studies, to create corpora that
-combine epigraphic and ‘literary’ materials, it is also open to future additions and new research.
+combine epigraphic and `literary' materials, it is also open to future additions and new research.
 
-Aknowledgments
+\section*{Aknowledgments}
 
-\bigskip
-
-The project is funded by the Excellence Cluster Topoi. I am grateful to Dominik Lukas, who helped me develop the
+\noindent The project is funded by the Excellence Cluster Topoi. I am grateful to Dominik Lukas, who helped me develop the
 structure of the database. I also thank Dr. Valentina Garulli for her helpful comments.

### 2015-12-17 source/43/EAGLE2016_Benefiel_Sypniewski.tex
paper 43 updated
doc to tex, biblio hidden in doc found and added to EAGLE.bib

@@
 %%%%%%%%\title{}
-\title{}
-\author{Rebecca Benefiel}
-\date{2015-11-19}
+\title{Working with Text and Images: The Graffiti of Herculaneum}
+\author[WL]{Rebecca Benefiel\corref{first}}
+\author[Millisaps]{Holly Sypniewski}
+\author[WL]{Sara Sprenkle}\date{2015-11-19}
+\address{Millsaps College, United States of America	}
+\address{Washington \& Lee University, United States of America}
+\cortext[first]{Corresponding author. Email: benefielr@wlu.edu}	
 \begin{document}
-1.1. Introduction
 
-Our project is working with informal, handwritten wall-inscriptions, or ancient graffiti, which were scratched into the
+\maketitle
+\begin{abstract}
+We discuss several challenges encountered by our team as we digitize ancient graffiti, handwritten inscriptions scratched into wall-plaster, for the Epigraphic Database Roma and the Ancient Graffiti Project.
+Here, we focus on decisions we made in editing and digitizing not only textual graffiti but also the figural examples (hand-sketched drawings) that sometimes appear alongside them.
+We also discuss search capabilities that will allow users both to browse and search for figural graffiti.
+\end{abstract}
+\keywords{Ancient graffiti, figural images, contextualization, standards, Herculaneum}
+
+\section{Introduction}
+
+\noindent Our project is working with informal, handwritten wall-inscriptions, or ancient graffiti, which were scratched into the
 wall-plaster of ancient towns. \ Several hundred of these handwritten inscriptions have been documented at Herculaneum
@@
 Among the many texts written on the walls of these two cities, there sometimes also appear graffiti drawings, or figural
-graffiti [Fig. 1].
+graffiti [Fig.~\ref{fig:1}].
 
- \includegraphics[width=8.985cm,height=4.413cm]{EAGLE2016BenefielSypniewski-img001.jpg} \newline
-Fig. 1 Graffito from Pompeii [CIL 4, 5215]\newline
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img001.jpg}
+\caption{ Graffito from Pompeii [CIL 4, 5215]}
+\label{fig:1}
+\end{figure}
 
@@
 gladiatorial contest with athletes, musicians, and perhaps magistrates, sketched by hand on a funerary monument just
-outside the Porta Nocera of Pompeii (CIL 4, 10237; D31 in Cooley and Cooley 2014; cf. also CIL 4, 10236 and 10238 drawn
+outside the Porta Nocera of Pompeii (CIL 4, 10237; \citet[D31]{cooley_pompeii_2013}; cf. also CIL 4, 10236 and 10238 drawn
 nearby). \ More commonly, people made small sketches on the walls around them choosing from roughly a handful of
-popular designs: \ heads in profile, boats, gladiators, birds, and geometric designs (Langner 2001).
-
-\bigskip
+popular designs: \ heads in profile, boats, gladiators, birds, and geometric designs \citep{langner_antike_2001}.
 
@@
 design a way to search for and retrieve such drawings via the Ancient Graffiti Project search engine. In this paper we
-will discuss the challenges we face and some of the strategies we have developed in response. \
+will discuss the challenges we face and some of the strategies we have developed in response.
 
-\bigskip
+\section{Our Material}
 
-2. Our Material
-
-First, a little background on figural graffiti and our sources for this data. \ In Herculaneum, we are fortunate that a
+\noindent First, a little background on figural graffiti and our sources for this data. In Herculaneum, we are fortunate that a
 significant number of graffiti are still extant and in situ, as roofing has been reconstructed for many buildings to
-protect them from the elements. \ Due to the fragile nature of wall plaster, however, especially in Pompeii, many
+protect them from the elements. Due to the fragile nature of wall plaster, however, especially in Pompeii, many
 graffiti that were recorded previously and published in CIL 4, have now been lost. Much of our data, therefore, comes
@@
 supplements used different methods to denote that a drawing was present, and their practices changed over time. Working
-with this legacy data, therefore, presents a range of difficulties. \
+with this legacy data, therefore, presents a range of difficulties.
 
-\bigskip
+\subsection{Verbal Descriptions of Figural Graffiti Found With Text}
 
-2.1 Verbal Descriptions of Figural Graffiti Found With Text
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img002.jpg}
+\caption{The entries of CIL 4, 2315-2316 and 2319, representing figural graffiti via brief description}
+\label{fig:2}
+\end{figure}
 
-\begin{figure}
-\centering
-\includegraphics[width=4.493cm,height=9.979cm]{EAGLE2016BenefielSypniewski-img002.jpg}
-\end{figure}
 A drawing could, for example, be described in the text field of an entry in CIL 4. This occurs in the entries below,
 where three drawings of human heads [CIL 4, 2315-2316] and two drawings of gladiators [CIL 4, 2319] are described in
-small italics, placed where the images occur alongside the textual inscriptions [Fig. 2].
-
-Fig. 2 \ The entries of CIL 4, 2315-2316 and 2319, representing figural graffiti via brief description
-
-\bigskip
+small italics, placed where the images occur alongside the textual inscriptions [Fig.~\ref{fig:2}].
 
@@
 
-\bigskip
-
-\ \ This practice is common in the original volume of CIL 4, when it seems that the editors documented textual and
+This practice is common in the original volume of CIL 4, when it seems that the editors documented textual and
 figural graffiti that were in close proximity, or that were in some way related to each other. In later supplements,
 line-drawings for figural graffiti were sometimes included when the drawings and text were obviously understood as one
-inscription, as shown in Fig. 1 above. \ Perhaps due to the complications with preparing and printing such
+inscription, as shown in Fig. 1 above. Perhaps due to the complications with preparing and printing such
 illustrations, however, it also remained common practice to represent figural graffiti with very brief description in
@@
 
-\bigskip
+\subsection{Figural Graffiti described in notes or apparatus}
 
-2.2 Figural Graffiti described in notes or apparatus
+\noindent The most common strategy, however, for documenting figural graffiti in CIL 4 is by including brief mention of a drawing
+in the editorial note that introduces a graffito or in the apparatus that follows it [Fig.~\ref{fig:3}].
 
-The most common strategy, however, for documenting figural graffiti in CIL 4 is by including brief mention of a drawing
-in the editorial note that introduces a graffito or in the apparatus that follows it [Fig. 3]. \ \
-
-\bigskip
-
- \includegraphics[width=10.227cm,height=4.075cm]{EAGLE2016BenefielSypniewski-img003.jpg}
-
-Fig. 3 \ CIL entry for an alphabet near figural graffiti [CIL 4, 10711]\newline
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img003.jpg}
+\caption{CIL entry for an alphabet near figural graffiti [CIL 4, 10711]}
+\label{fig:3}
+\end{figure}
 
@@
 
-\bigskip
-
-2.3 Figural graffiti omitted by CIL
+\subsection{Figural graffiti omitted by CIL}
 
@@
 create the most comprehensive resource possible for figural graffiti, we include all known drawings in the AGP search
-engine. \
+engine.
 
-\bigskip
+\subsection{Documenting extant figural graffiti}
 
-2.4 \ Documenting extant figural graffiti
-
-Since the verbal descriptions of figural graffiti provided by legacy data are limited and vague or exceedingly general
+\noindent Since the verbal descriptions of figural graffiti provided by legacy data are limited and vague or exceedingly general
 (e.g. caput), the best circumstance under which to digitize a graffito is when the drawing itself still remains extant.
-\ In such cases, we will use any published data as a starting point, but we are also able to make our own editorial
+ In such cases, we will use any published data as a starting point, but we are also able to make our own editorial
 decisions about the subject matter of the drawing, how to describe it, and its relation to any text that is nearby.
-\ The material with which we are working, therefore, includes a range of different information about the figural
+The material with which we are working, therefore, includes a range of different information about the figural
 graffiti of Pompeii and Herculaneum: from brief verbal descriptions to line-drawings, to the best case scenario when an
@@
 
-3.1 Challenges in working with text and images
+\section{Challenges in working with text and images}
 
@@
 interpret and standardize legacy data.
-
-\bigskip
 
@@
 
-\bigskip
-
-3.2 How to define an epigraphic entry?
+\subsection{How to define an epigraphic entry?}
 
@@
 then, if we create individual entries for separate elements, how do we avoid losing information about the relationship
-among the graffiti? This collection of drawings including six textual graffiti illustrates our challenge [Fig. 4].
+among the graffiti? This collection of drawings including six textual graffiti illustrates our challenge [Fig.~\ref{fig:4}].
 
- \includegraphics[width=10.88cm,height=5.761cm]{EAGLE2016BenefielSypniewski-img004.jpg}
-
-Fig. 4 Line-drawing displaying a collection of textual and figural graffiti [CIL 4, 8383-8386; EDR148730]; unpublished
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img004.jpg}
+\caption{Line-drawing displaying a collection of textual and figural graffiti [CIL 4, 8383-8386; EDR148730]; unpublished
 sketch of Matteo Della Corte from the archives of the Corpus Inscriptionum Latinarum at the Berlin-Brandenburg Academy
-of Science and Humanities.
-
-\bigskip
+of Science and Humanities.}
+\label{fig:4}
+\end{figure}
 
 You can see a number of different images here including a small gladiator with trident, a face in profile, leaves,
-several animals, and geometric shapes as well as the name “Atini” and the greeting “\textgreek{Gelast`h qa~ire.”
+several animals, and geometric shapes as well as the name ``Atini'' and the greeting ``\textgreek{Gelast`h qa~ire.''
 }\ Such a collection raises many interpretive questions. What is the relationship, if any, between the figural and the
@@
 
-In an example from Herculaneum (Fig. 3, above), the entry for CIL 4, 10711 notes that in addition to a graffito of the
+In an example from Herculaneum (Fig.~\ref{fig:3}, above), the entry for CIL 4, 10711 notes that in addition to a graffito of the
 alphabet, a series of nine gladiator helmets and a small phallus were also drawn on the wall. During our field season
 in Herculaneum in 2014, we were not successful in finding the graffiti of the alphabet or small phallus, but we did
-locate eight of the nine helmets. Here too we have devoted a separate database entry for each helmet. \ By making
+locate eight of the nine helmets. Here too we have devoted a separate database entry for each helmet. By making
 individual records, we have a unique identifier for each image, in the form of the EDR number, so that users can cite a
-specific parallel precisely. \ Again, we can record the precise measurements for each image. \ Yet, since separating
+specific parallel precisely. \ Again, we can record the precise measurements for each image. Yet, since separating
 each image can obscure how the images relate to one another in the group, as with the previous example, we also upload
@@
 
-\bigskip
-
-3.3.1 How to describe drawings?
+\subsubsection{How to describe drawings? }
 
@@
 our entry. With figural graffiti documented by Martin Langner, we must create a summary in Latin and when doing so, we
-attempt to give as full as possible a description of the elements of the image. With this camel (Langner 2001, n.
-1443), for instance, we offer a full description in Latin that accounts for all the features of the drawing: camelus
-dromedarius cum cauda, lodicem gerens, ad dextram incedens [Fig. 5].
+attempt to give as full as possible a description of the elements of the image.
+With this camel \citep[n.
+1443]{langner_antike_2001}, for instance, we offer a full description in Latin that accounts for all the features of the drawing: camelus
+dromedarius cum cauda, lodicem gerens, ad dextram incedens [Fig.~\ref{fig:5}].
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img005.jpg}
+\caption{Example of an entry for a figural graffito not found in CIL (EDR144514)}
+\label{fig:5}
+\end{figure}
 
- \includegraphics[width=10.88cm,height=8.608cm]{EAGLE2016BenefielSypniewski-img005.jpg}
-
-Fig. 5 Example of an entry for a figural graffito not found in CIL \ (EDR144514)
-
-\bigskip
-
-3.3.2 \ Questions of interpretation through description\ \
+\subsubsection{Questions of interpretation through description}
 
@@
 which would seem to be a drawing of a person facing forward and rendered with head and shoulders. CIL describes it
-thus: herma muliebris prospiciens [Fig. 6].
+thus: herma muliebris prospiciens [Fig.~\ref{fig:6}].
 
-\begin{figure}
+\begin{figure}[!bp]
 \centering
-\includegraphics[width=7.1cm,height=9.155cm]{EAGLE2016BenefielSypniewski-img006.jpg}
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img006.jpg}
+\caption{CIL entry for textual and figural graffiti [CIL 4, 8185]}
+\label{fig:6}
 \end{figure}
-Fig. 6 CIL entry for textual and figural graffiti [CIL 4, 8185]
-
-\bigskip
 
@@
 mulieris imaginem). In this instance, too, the text nearby includes two names, one female and one male. The drawing has
-appeared in multiple publications (Della Corte 1960, Deiss 1989, Maulucci 1993) [Fig. 7].
+appeared in multiple publications (\citet{della_corte_loves_1960}, \citet{deiss_herculaneum_1989}, \citet{vivolo_pompei:_1993}) [Fig.~\ref{fig:7}].
 
-\bigskip
-
- \includegraphics[width=5.553cm,height=4.217cm]{EAGLE2016BenefielSypniewski-img007.jpg}
-
-Fig. 7 Apograph of figural graffito referred to in the note at CIL 4, 10676 [Langner 2001, n. 309].
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img007.jpg}
+\caption{Apograph of figural graffito referred to in the note at CIL 4, 10676 \citet[n. 309]{langner_antike_2001}.}
+\label{fig:7}
+\end{figure}
 
@@
 Another reason we suspected there might be a mistake was that Martin Langner had catalogued the drawing associated with
-CIL 4, 10676, describing it as a “Phalluskopf.” There was no mention of gender. And he categorized this drawing among
+CIL 4, 10676, describing it as a ``Phalluskopf.'' There was no mention of gender. And he categorized this drawing among
 several examples of drawings of heads with phallic features. The graffito is in a room that is sealed off, with no
 access, so we were unable to view it in person. Eventually, however, a photograph published by Antonio Varone in his
-recent two-volume work providing images of extant ancient graffiti (Varone 2012, 509) allowed us to confirm that this
+recent two-volume work providing images of extant ancient graffiti \citep[509]{varone_titulorum_2012} allowed us to confirm that this
 is indeed the correct graffito drawing – somewhat above but also drawn partly through the text of CIL 4, 10676.
@@
 Neither description offered by CIL or by Langner, however, seems altogether satisfactory. There are no obvious markers
-of female identity and even the description of Phalluskopf is less than transparent. \ Thus this one drawing has two
+of female identity and even the description of Phalluskopf is less than transparent. Thus this one drawing has two
 published descriptions that vary greatly and that each lead to a very different understanding of the graffito. What
 should we then do in such situations? Do we repeat the identification of CIL? Or do we offer a less specific
-description, merely labeling this a hominis figura? \ In the end, our solution is to offer our own description, which
+description, merely labeling this a hominis figura? In the end, our solution is to offer our own description, which
 is detailed but less interpretative, with an emphasis on specific features of the image that are readily identifiable.
-\ We also document Langner’s description and CIL’s earlier identification in our entry for EDR, but we note our
-hesitation with such identification by labeling the image: gryllus? (“caricature?”). \ We are aware, however, that we
-also introduce an interpretation with the tentative suggestion this drawing might be a caricature. \ \
+We also document Langner’s description and CIL’s earlier identification in our entry for EDR, but we note our
+hesitation with such identification by labeling the image: gryllus? (``caricature?''). We are aware, however, that we
+also introduce an interpretation with the tentative suggestion this drawing might be a caricature.
 
@@
 any type of animal with four legs (stags, boars, dogs). In such cases, our solution is to describe a drawing with more
-generic, yet accurate, terms such as “gladiator,” without further specification, or “animal” rather than cervus, aper,
-or canis. \ Similarly, if we cannot determine male or female, we prefer to describe the drawing as “facies hominis.” In
+generic, yet accurate, terms such as ``gladiator,'' without further specification, or ``animal'' rather than cervus, aper,
+or canis. Similarly, if we cannot determine male or female, we prefer to describe the drawing as ``facies hominis.'' In
 the AGP search engine we can then indicate possible but not certain identification with a descriptor, or tag, that
-comes with a question mark: e.g. “stag?”
+comes with a question mark: e.g. ``stag?''
 
-\bigskip
+\section{How to search for drawings?}
 
-4.1 How to search for drawings?
+\noindent The third challenge that we face is how to search efficiently for inscriptions that are not just text but either are
+images or include images. In the AGP search engine, we aim to complement the capabilities of EDR by providing another
+way to search for these non-textual, figural graffiti. Since we describe the content of the figural graffito in the
+Textus field of EDR, it is possible for a user to locate a graffito drawing. However, with text-based searching, a user
+would need to need to know the vocabulary used to describe the drawing. Would someone ever think to search for
+``camelus'' without prior knowledge that there is a figural graffito of a camel in Herculaneum? Similarly, if you
+search for ``gladiator'', the text field will give you results for all inscriptions that mention gladiators as well as
+drawings where we have described gladiators. If, however, we’ve described the gladiator more specifically as a
+``retiarius'' or we have gladiatorial equipment, such drawings will be omitted from the list of search results.
 
-\ \ The third challenge that we face is how to search efficiently for inscriptions that are not just text but either are
-images or include images. \ In the AGP search engine, we aim to complement the capabilities of EDR by providing another
-way to search for these non-textual, figural graffiti. \ Since we describe the content of the figural graffito in the
-Textus field of EDR, it is possible for a user to locate a graffito drawing. However, with text-based searching, a user
-would need to need to know the vocabulary used to describe the drawing. \ Would someone ever think to search for
-“camelus” without prior knowledge that there is a figural graffito of a camel in Herculaneum? \ Similarly, if you
-search for “gladiator”, the text field will give you results for all inscriptions that mention gladiators as well as
-drawings where we have described gladiators. \ If, however, we’ve described the gladiator more specifically as a
-“retiarius” or we have gladiatorial equipment, such drawings will be omitted from the list of search results.
-
-\ \ We are therefore designing AGP with the capacity for locating figural graffiti through a two-prong solution: with
+We are therefore designing AGP with the capacity for locating figural graffiti through a two-prong solution: with
 both browsing and searching possibilities.
 
-\bigskip
+\subsection{Browsing capabilities in AGP}
 
-4.2 \ Browsing capabilities in AGP
+\noindent For browsing, we have defined nine broad general categories, which together cover all the types of figural graffiti
+we have encountered so far [Fig.~\ref{fig:8}].
 
-\ \ For browsing, we have defined nine broad general categories, which together cover all the types of figural graffiti
-we have encountered so far [Fig. 8].
-
- \includegraphics[width=7.567cm,height=7.586cm]{EAGLE2016BenefielSypniewski-img008.png}
-
-Fig. 8 Screenshot of AGP search engine showing browsing capabilities.
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img008.png}
+\caption{Screenshot of AGP search engine showing browsing capabilities.}
+\label{fig:8}
+\end{figure}
 
 At this point, one can browse by choosing a category, which will return all examples in that category. So, for example,
-the category of “Gladiators/equipment” will return sketches of individual gladiators, gladiators fighting in pairs, and
-gladiator equipment such as helmets. The category of “Animals” will return all figural graffiti that include drawings
+the category of ``Gladiators/equipment'' will return sketches of individual gladiators, gladiators fighting in pairs, and
+gladiator equipment such as helmets. The category of ``Animals'' will return all figural graffiti that include drawings
 of animals. As we process greater numbers of figural graffiti, the results of these categories will become larger.
 
-\bigskip
+\subsection{Limiting search results in AGP}
 
-4.3 \ Limiting search results in AGP
-
-\ \ It is therefore necessary to design a way to limit results, so that a user could find only gladiatorial equipment,
+\noindent It is therefore necessary to design a way to limit results, so that a user could find only gladiatorial equipment,
 or graffiti depicting only boars but not graffiti depicting other animals. We are developing a method of filters and
-tags that will allow a user to move beyond browsing. \ These filters will allow a user 1) to limit an initial return of
+tags that will allow a user to move beyond browsing. These filters will allow a user 1) to limit an initial return of
 results, 2) to retrieve more specific results, or 3) to perform a secondary level of search. It will certainly be
@@
 ultimately to enable searches by these tags as well, so a user can directly find all drawings with dogs. The search
-capacity will allow a user to search the tags or the Latin description, so both “navis” and “boat” will return hits.
+capacity will allow a user to search the tags or the Latin description, so both ``navis'' and ``boat'' will return hits.
 Again, standardization is necessary. We are currently developing a list of tags that is comprehensive and flexible
@@
 desired graffito. So, a user can search all graffiti drawings and then limit the search results, for example, to find
-all the drawings in a particular property [Fig. 9].
+all the drawings in a particular property [Fig.~\ref{fig:9}].
 
-\bigskip
-
-  [Warning: Image ignored] % Unhandled or unsupported graphics:
-%\includegraphics[width=8.855cm,height=6.17cm]{EAGLE2016BenefielSypniewski-img009}
-
-Fig. 9 Screenshot showing filters in AGP.
-
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img009.png}
+\caption{Screenshot showing filters in AGP.}
+\label{fig:9}
+\end{figure}
 
@@
 
-\bigskip
+\section{Conclusion}
 
-5. Conclusion
-
-\ \ This system of tags and filters is in the early stages of the design process. These are our proposed solutions for
+\noindent  This system of tags and filters is in the early stages of the design process. These are our proposed solutions for
 confronting the challenges of working with text and image, and our ideas for creating a resource to complement the
@@
 
-%WHER IS THE BIBLIOGRAPHY??
+\nocite{wallace-hadrill_herculaneum:_2011}
+\nocite{vivolo_pompei:_1993}
+\nocite{guadagno_supplemento_1978}
+\nocite{guadagno_supplemento_1981}
+\nocite{della_corte_iscrizioni_1958}
 

### 2015-12-17 source/44/Damnjanovic_et_al_EAGLE2016.tex
paper 44 updated
doc to tex, also the bibliography has been updated, Bergstrom gave
troubles until I rewrote it entirely and then it magically worked…

@@
 %%%%%%%%
+\title{INTEGRATION OF MULTIMEDIA COLLECTIONS AND TOOLS FOR INTERACTION WITH DIGITAL CONTENT.
+The case study of the Archaia Kypriaki Grammateia Digital Corpus}
+\author[cyi]{Valentina Vassallo\corref{first}}
+\author[cyi]{Uros Damnjanovic}
+\author[cyi]{Sorin Hermon}
+\date{2015-11-26}
+\cortext[first]{Corresponding author. Email: (u.damnjanovic, v.vassallo, s.hermon)@cyi.ac.cy}
+\begin{document}
 
-\author{valentina}
-\date{2015-11-26}
-\begin{document}
-INTEGRATION OF MULTIMEDIA COLLECTIONS AND TOOLS FOR INTERACTION WITH DIGITAL CONTENT. \newline
-The case study of the Archaia Kypriaki Grammateia Digital Corpus
-
-Uros Damnjanovic¹, Valentina Vassallo¹, Sorin Hermon¹\newline
-¹The Cyprus Institute - STARC\newline
-(u.damnjanovic, v.vassallo, s.hermon)@cyi.ac.cy
-
-Abstract
-
+\maketitle
+\begin{abstract}
 Supporting a discovery, use, and navigation of digital collections is a fundamental part of providing access and
@@
 providing the means for utilizing the full potential of the heritage data.
+\end{abstract}
+\keywords{Digital libraries, Data repository, Epigraphy, User interfaces, Data visualization, Interaction}
 
-Keywords
+\section{The world going digital}
 
-Digital libraries, Data repository, Epigraphy, User interfaces, Data visualization, Interaction.
-
-1. The world going digital
-
-Nowadays digital information format found its ways to all the spheres of our lives. Cultural heritage is no exception to
+\noindent Nowadays digital information format found its ways to all the spheres of our lives. Cultural heritage is no exception to
 this. Gigabytes of data are being created on an every-day basis. For some time now, digital libraries are used to store
@@
 Current digital libraries are limited by their inadequate webpage-based paradigm, and it is easy for even the most
-experienced scholar to get lost [Bergstorm and Atkinson, 2009]. To overcome this issue, new generation of digital
+experienced scholar to get lost \citep{augmenting_2009}.
+To overcome this issue, new generation of digital
 libraries should emerge that will serve as an interactive knowledge environment in which knowledge is created by
@@
 
-According to [Fast and Sedig, 2010] digital libraries are seen as a store of epistemic potential with interaction having
-the role to reveal much of the hidden complexity. In [Buchel and Sedig, 2014] the authors demonstrated the role that a
+According to \citep{fast_interaction_2010} digital libraries are seen as a store of epistemic potential with interaction having
+the role to reveal much of the hidden complexity. In \citep{buchel_making_2014} the authors demonstrated the role that a
 set of interactions can play in supporting users' understanding of spaces and their formation of cognitive maps when
-working with map-based visualisations. \ According to [Algee et al., 2012] interactive visualizations empower users to
+working with map-based visualisations. According to \citep{algee_viewshare_2012} interactive visualizations empower users to
 discover meaning and patterns within digital collections using dynamic, interactive displays. One example of utilizing
-the interactive graphics in a real world digital library is presented in [Hienert et al., 2012]. Authors developed
+the interactive graphics in a real world digital library is presented in \citep{hienert_integrating_2012}. Authors developed
 interactive visualization tools that support exploration of search queries and search results and help users in
-formulating new queries. Another example, the application called DaisyViz shown in [Ren et al., 2009], tries to help
+formulating new queries. Another example, the application called DaisyViz shown in \citep{ren_daisyviz:_2010}, tries to help
 users acquire better insights of the data by enabling users to rapidly develop domain-specific information
-visualizations without traditional programming. In [Nasar et al., 2011] authors proposed a conceptual framework that
+visualizations without traditional programming. In \citep{nasar_conceptual_2011} authors proposed a conceptual framework that
 uses interactive visualizations for managing personal collections of images and videos with focus on data re-finding
-and improved filtering. INVISQUE [Wong et al., 2012] is a novel system designed for interactive information
+and improved filtering. INVISQUE \citep{wong_invisque:_2011} is a novel system designed for interactive information
 exploration. Instead of a conventional list-style arrangement, in INVISQUE information is represented by a
@@
 results browsing, categorizing, and linking to online resources such as Google and Twitter. An approach for
-overview-first exploration of data collections based on user-selected metadata properties is presented in [Bernard et
-al., 2012]. \ In a 2D layout representing entities of the selected property are laid out based on their similarity with
+overview-first exploration of data collections based on user-selected metadata properties is presented in \citep{bernard_content-based_2012}.
+In a 2D layout representing entities of the selected property are laid out based on their similarity with
 respect to the underlying data content. The display is enhanced by compact summarizations of underlying data elements,
-and forms the basis for exploratory navigation of users in the data space. In [Nan et al., 2012] authors propose a
-novel visual design termed “Whisper” to fulfil the need for tracing information diffusion processes in social media, in
-a real time manner. BirdVis [Ferreira et al., 2011] leverages visualization techniques and uses them in a novel way to
+and forms the basis for exploratory navigation of users in the data space. In \citep{cao_whisper:_2012} authors propose a
+novel visual design termed ``Whisper'' to fulfil the need for tracing information diffusion processes in social media, in
+a real time manner. BirdVis \citep{ferreira_birdvis:_2011} leverages visualization techniques and uses them in a novel way to
 better assist users in the exploration of interdependencies among model parameters. Furthermore, the system allows for
@@
 
-\bigskip
+\section{``A place where the past meets the future'': the STARC repository}
 
-2. “A place where the past meets the future”: the STARC repository
-
-STARC repository was initially developed to store the data coming from the daily work of the research group and as
+\noindent STARC repository was initially developed to store the data coming from the daily work of the research group and as
 content aggregator for various digital libraries projects: data from archaeological excavations, museums artefacts,
@@
 schema that depends on the type of data and according to the needs of the project. For most of the items, we are using
-our own metadata schema (STARC metadata schemas) \ [Ronzino et al., 2012; Vassallo et al., 2013], that was developed to
+our own metadata schema (STARC metadata schemas) \citep{ronzino_metadata_2012}; \citep{vassallo_revealing_2013}, that was developed to
 thoroughly describe various aspects of data creation process. \ Once data is uploaded and available, it can also be
 used in different systems. For example, the repository serves as source of data for aggregation procedures by which our
-data feed the Europeana portal (http://www.europeana.eu/portal/) through different digital libraries projects (e.g.
+data feed the Europeana portal (\url{http://www.europeana.eu/portal/}) through different digital libraries projects (e.g.
 Athena, CARARE, Linked Heritage, AthenaPlus, EAGLE). In order to facilitate the migration of data from heterogeneous
@@
 
-\bigskip
+\subsection{Tools for accessing data }
 
-2.1. Tools for accessing data
-
-We wanted to provide users with number of tools that they can use while exploring the repository. In traditional web
+\noindent We wanted to provide users with number of tools that they can use while exploring the repository. In traditional web
 based system, navigating from one tool to another usually means that we click on a link and then move to another page
@@
 data or by performing any action relevant to the selected functionality. The process of exploring the repository starts
-by running the initial query (Fig.1). \
+by running the initial query (Fig.~\ref{fig:1}).
 
@@
 to the text in the search field. The user can then run a new search repeatedly, and the search history functionality
-will store all those searches for later reference. The map tool (Fig.1) is used to show information about the
+will store all those searches for later reference. The map tool (Fig.~\ref{fig:1}) is used to show information about the
 geo-referenced data. It provides two main functions. One is that it accompanies the search tool. Whenever the new
-search is ran, and the results become available, the map tool updates itself with new data, showing the
-
- \includegraphics[width=10.896cm,height=6.81cm]{DamnjanovicetalEAGLE2016-img001.png}
-
-Fig. 1. The exploration process starts by running the search. Once the results are available the user can add or remove
-exploration tools. By default map tool is present in the view. The map tool is used to mark the geo-referenced data on
-the map, and also to perform geo-search by defining the region of interest.
-
-geographical distribution of the search results. The user can then use the map to filter out the data by clicking on the
+search is ran, and the results become available, the map tool updates itself with new data, showing the geographical distribution of the search results.
+The user can then use the map to filter out the data by clicking on the
 markers on the map. Also the map can also be used as a search tool. By selecting the search map mode, the rectangle
@@
 
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img001.png}
+\caption{The exploration process starts by running the search. Once the results are available the user can add or remove
+exploration tools. By default map tool is present in the view. The map tool is used to mark the geo-referenced data on
+the map, and also to perform geo-search by defining the region of interest. }
+\label{fig:1}
+\end{figure}
+
 Annotation tool is used to attach the information to the objects in the repository and share it with other users, or use
 it to facilitate more efficient search. The tool is used by selecting object from search results, then either by
-selecting a region in the image files (Fig.2), or by selecting a point for 3D models, and by adding textual description
+selecting a region in the image files (Fig.~\ref{fig:2}), or by selecting a point for 3D models, and by adding textual description
 of the selected part.
 
- \includegraphics[width=10.896cm,height=6.81cm]{DamnjanovicetalEAGLE2016-img002.png}
-
-Fig. 2. The annotation tool. The user first selects the object in the search results, then selects the region of the
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img002.png}
+\caption{The annotation tool. The user first selects the object in the search results, then selects the region of the
 object and then adds the annotation. The annotation can be accessed when the annotated object is accessed, and by
-specifying in the search that the search should also include annotations. \
+specifying in the search that the search should also include annotations. }
+\label{fig:2}
+\end{figure}
 
@@
 In order to help in exploring the datasets, there are statistics and visualization tools. Statistics tool shows the
-basic properties of the repository. A visualisation tool is used to accompany the search tool (Fig. 3). Story creation
+basic properties of the repository. A visualisation tool is used to accompany the search tool (Fig.~\ref{fig:3}). Story creation
 tool is another tool that supports collaborative user generated content. It provides users with an online document
@@
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img003.png}
+\caption{Visualization tool. Visualization tool is used to accompany the search tool by showing live visualizations as
+the search results change. When the search is ran and the search results are available the visualization tools
+automatically update itself and shows the visualisation related to the current search results. }
+\label{fig:3}
+\end{figure}
 
- \includegraphics[width=10.896cm,height=6.81cm]{DamnjanovicetalEAGLE2016-img003.png}
+\section{Archaia Kypriaki Grammateia, the STARC Repository and the EAGLE infrastructure}
 
-Fig. 3. Visualization tool. Visualization tool is used to accompany the search tool by showing live visualizations as
-the search results change. When the search is ran and the search results are available the visualization tools
-automatically update itself and shows the visualisation related to the current search results.
-
-\bigskip
-
-\bigskip
-
-3. Archaia Kypriaki Grammateia, the STARC Repository and the EAGLE infrastructure
-
-One of the most interesting collection stored in the repository is the Archaia Kypriaki Grammateia Digital Corpus
-[Pitzalis et al., 2012]. The collection is composed of Ancient Greek and Latin epigraphic texts produced within a time
+\noindent One of the most interesting collection stored in the repository is the Archaia Kypriaki Grammateia Digital Corpus
+\citep{pitzalis_building_2012}. The collection is composed of Ancient Greek and Latin epigraphic texts produced within a time
 span of circa 13 centuries (from the 7th century BC to the 6th century AD). The texts are attributed to Cypriot authors
 or were produced in Cyprus. The corpus consists mainly of funerary or dedicatory epigrams published with their
-translation in Modern Greek, critical apparatus and philological comments [Voskos, 1997].
+translation in Modern Greek, critical apparatus and philological comments \citep{voskos__1997}.
 
@@
 series of multimedia digital resources that describe the content in a multidisciplinary way: digital texts, images
-related to the epigraphic supports, 3D representation of the inscriptions, video, audio files, and so forth (Fig. 4).
+related to the epigraphic supports, 3D representation of the inscriptions, video, audio files, and so forth (Fig.~\ref{fig:4}).
 
-\bigskip
-
- \includegraphics[width=10.938cm,height=4.129cm]{DamnjanovicetalEAGLE2016-img004.jpg}
-
-Fig. 4. Multimedia data constituting the AKG digital collection stored in the STARC repository.
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img004.jpg}
+\caption{Multimedia data constituting the AKG digital collection stored in the STARC repository.}
+\label{fig:4}
+\end{figure}
 
@@
 multidisciplinary research carried out on this material in a manner that users comprehend the complexity of such an
-approach through access to heterogeneous types of information [Vassallo et al., 2013].
+approach through access to heterogeneous types of information \citep{vassallo_revealing_2013}.
 
 The metadata schema for Ancient Cypriot inscriptions integrates multidisciplinary information regarding the objects and
-their multiple digital resources. It is the result of a research line developed within the group [Vassallo et al.,
-2013] and of the assessment and comparisons of schemas, models and ontologies already in use in the digital epigraphy
+their multiple digital resources.
+It is the result of a research line developed within the group \citep{liuzzo_networking_2014} and of the assessment and comparisons of schemas, models and ontologies already in use in the digital epigraphy
 field (e.g. TEI Epidoc, Dublin Core, CIDOC-CRM).
@@
 info, the file specifications, etc.) and the digital output (data format, software), according to the digital resource
-obtained (Fig. 5).
+obtained (Fig.~\ref{fig:5}).
 
@@
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img005.pdf}
+\caption{Diagram of the metadata schema components and relationships.}
+\label{fig:5}
+\end{figure}
 
- \includegraphics[width=12.435cm,height=13.573cm]{DamnjanovicetalEAGLE2016-img005.pdf}  Fig. 5. Diagram of the metadata
-schema components and relationships.
-
-3.1. From STARC Repository to EAGLE infrastructure
+\subsection{From STARC Repository to EAGLE infrastructure}
 
@@
 
-The “Europeana network of Ancient Greek and Latin Epigraphy”, is an EU funded project under the umbrella of the CIP-Best
+The ``Europeana network of Ancient Greek and Latin Epigraphy'', is an EU funded project under the umbrella of the CIP-Best
 Practice Network and whose aim is to bring together some of the most prominent European institutions and cultural
@@
 description of the text of inscriptions while CIDOC CRM enables a further full description object-oriented,
-re[FB02?]ecting at the same time the different souls of Epigraphy, the philological and the archaeological one. Content
+reflecting at the same time the different souls of Epigraphy, the philological and the archaeological one. Content
 providers use different models according to the fact their databases are more oriented towards the text of an
 inscription or towards the archaeological object that bears the text. According to the epigraphers’ community needs,
-EAGLE consortium developed the data model on the base of the two sets of standards [Liuzzo et al., 2014]. Finally, the
+EAGLE consortium developed the data model on the base of the two sets of standards \citep{liuzzo_networking_2014}. Finally, the
 EAGLE data model is mapped to the Europeana Data Model (EDM), in order to be ingested into the Europeana portal.
@@
 manifestation, visual representation. The data that are sent to Europeana belong to the category of the artefacts (the
-inscriptions represented as texts, called by Europeana “Cultural Heritage Object” - CHO) and of the visual
-representation (the images related to the inscriptions, called by Europeana “WebResource”). The distinction between CHO
+inscriptions represented as texts, called by Europeana ``Cultural Heritage Object'' - CHO) and of the visual
+representation (the images related to the inscriptions, called by Europeana ``WebResource''). The distinction between CHO
 and WebResource has been introduced in Europeana as a result of the introduction of the new EDM schema to provide the
@@
 collection of texts accompanied by their translations, notes and commentaries as published in its last edition by
-Voskos [Voskos, 1997], therefore it is a digital form of this important editorial printed work, to which are connected
+Voskos \citep{voskos__1997}, therefore it is a digital form of this important editorial printed work, to which are connected
 all a series of digital resources that help to enrich the corpus.
@@
 test for the second solution. In fact the latter, even if it is time-consuming in terms of elaboration, seems to be the
-best solution for avoiding data duplication and for having more cohesive data. \
+best solution for avoiding data duplication and for having more cohesive data.
 
-4. Conclusions
+\section{Conclusions}
 

### 2015-12-17 source/31/EAGLE2016_I Sicily_final copy.tex
fixed IIIP to IIP as per email

@@
 AngularJS and jQuery javascript components. Mapping is provided in the browser by the Google Maps API, and ZPR (Zoom,
-Pan, Rotate) image-viewing is provided by the IIIP image server and the OpenSeadragon javascript library.
-%IIIP or IIIF ? I believe this is a typo
+Pan, Rotate) image-viewing is provided by the IIP image server and the OpenSeadragon javascript library.
 

### 2015-12-17 source/48/EAGLE16-lame-et-al-teaching.tex
started to update 48
this paper needs a lot of work on tables, images placed in tabular
form, etc.

@@
 %%%%%%%%
-\title{}
-\author{Marion Lamé}
+\title{Teaching (Digital) Epigraphy}
+\author[ilc]{Marion Lamé\corref{first}}
+\author[zag]{Ivan Radman-Livaja}
+\author[mau]{Bruce Robertson}
+\author[isti]{Federico Ponchio}
+\address[ilc]{Istituto di Linguistica Computazionale}
+\address[isti]{Visual Computing Laboratory, ISTI-CNR}
+\address[zag]{Archaeological Museum in Zagreb}
+\address[mau]{Mount Allison University}
+\cortext[first]{Corresponding author. Email: federico.ponchio@isti.cnr.it}
+
 \date{2015-12-02}
 \begin{document}
-Teaching (Digital) Epigraphy
 
-Marion Lamé, Ivan Radman-Livaja, Bruce Robertson and Federico Ponchio
-
-Abstract
-
+\maketitle
+\begin{abstract}
 Although having the experience of directly manipulating an ancient text-bearing artifact is important for developing the
@@
 inscribed lead tags, using a Digital Autoptic Process (DAP) in a web environment, so to develop basic competences in
-epigraphy and digital epigraphy.\footnote{\par \ \  We wish to acknowledge the research assistance of T. M. Crandall,
-History student at Mount Allison University \ (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
-\ M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
-Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
-{\textquotedbl}La lettre romaine, entre inscriptions monumentales et écritures communes{\textquotedbl}. The following
-students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
-Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
-Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
-as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
-Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
-present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.}
+epigraphy and digital epigraphy.
+\end{abstract}
+\keywords{Educational project, digital epigraphy, epigraphy, open access, primary sources}
 
-Keywords
+\section{Introduction}
 
-Educational project, digital epigraphy, epigraphy, open access, primary sources
-
-\begin{enumerate}
-\item Introduction
-\end{enumerate}
 Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
@@
 sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
-reserved to their senior colleagues is certainly not a common occurrence. In our small project \ “Pedagogical,
+reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
 Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the Tesserarum Sisciae Sylloge (TSS) through
-a Digital Autoptic Process (DAP)” we aimed to challenge this state of affairs. Using digital methods and a
+a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
 custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
@@
 
-2. Epigraphic Primary Sources of Information
+\section{Epigraphic Primary Sources of Information}
 
@@
 (i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
-correct fee\footnote{\par \ \  Radman-Livaja 2007/11/13/14 and 2013b, 165-172 for analogous tags with Mócsy 1956,
-97-104; Egger 1967, 195-210; Frei-Stolba 1984, 127-138; Schwinden 1985 121-137/93, 215-222; Römer-Martijnse 1990/97,
-5-48; Feugère 1993, 301-305; Weiss 1991, 211-220; Paci 1995: 29-40; Bassi 1996, 207-216; Bizzarini 2005, 121-135; Buchi
-\& Buonopane 2005, 43-51; Cresci \& Pettenò 2010: 42-110; Jacques \& Hoët-Cauwenberghe 2010, 295-317; Wedenig 2013,
-237-246.}.
+correct fee\footnote{\citep{radman_livaja_segestica..._2008}, \citep{radman-livaja_roetiquettes_2010}, \citep{radman-livaja_5._2013}, \citep[165-172]{radman-livaja_two_2013}, \citep{radman-livaja_plombs_2014},  for analogous tags with \citep[97-104]{mocsy_olom_1956}; \citep[195-210]{egger_funf_1968}; \citep[127-138]{frei-stolba_les_2011}; \citep[121-137/93, 215-222]{schwinden_romerzeitliche_1985}; \citep{romer-martijnse_romerzeitliche_1990}; \citep[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citep[301-305]{feugere_etiquette_1993}; \citep[211-220]{weiss_bleietiketten_1991}; \citep[29-40]{paci_etichette_1995}; \citep[207-216]{bassi_tre_1996}; \citep[121-135]{bizzarini_quattro_2005}; \citep[43-51]{buchi_etichette_2005}; \citep[42-110]{cresci_supellex_2010}; \citep[295-317]{jacques_artisanat_2010}; \citep[237-246]{wedenig_bleietikett_2013}.}.
 
 Some of the genuine difficulties of the autoptic process are already solved by a static picture in a print publication.
-The tag has a given position, \ it is illuminated from one direction only and therefore it already suggests a reading
+The tag has a given position, it is illuminated from one direction only and therefore it already suggests a reading
 direction. A DAP should be able to confront the students to some of those difficulties and to overcome them alone.
@@
 
-3. Digital Epigraphy
+\section{Digital Epigraphy}
 
@@
 is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\ Terras, 2015
-writes about some of the OA issues: “While digitisation is not a pre-requisite to gaining access to material (...), and
+writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
 while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
@@
 possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
-the nature of research itself in these areas.”}. Of course, learners could train themselves on the drawing of the tag,
+the nature of research itself in these areas.''}. Of course, learners could train themselves on the drawing of the tag,
 but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
@@
 Students were involved for a very short period of time: on average two hours per day over a couple of days only (in
-Spring 2014 and in Fall 2015). This training was delivered in three steps (“Readings”, “Writings”, “Test \& Taste a
-DAP”) during which there were given some academic documentation about the lead tags, and basic knowledge about digital
+Spring 2014 and in Fall 2015). This training was delivered in three steps (``Readings'', ``Writings'', ``Test \& Taste a
+DAP'') during which there were given some academic documentation about the lead tags, and basic knowledge about digital
 epigraphy (epigraphic message in a computer, dispositive analysis, existing projects, digital tools, XML and Unicode
@@
 \end{figure}
-4. Practicing Epigraphy
+
+\section{Practicing Epigraphy}
 
@@
 
-4.1 Drawing With Help (No. 12563)
+\subsection{Drawing With Help (No. 12563)}
 
@@
 \end{flushleft}
-4.2 Drawing \& Deciphering Without Help (obv. No. 13053)
+
+\subsection{Drawing \& Deciphering Without Help (obv. No. 13053)}
 
@@
 \end{flushleft}
-\begin{enumerate}
-\item \begin{enumerate}
-\item Palimpsest (obv. No. 13053)
-\end{enumerate}
-\end{enumerate}
+
+\subsection{Palimpsest (obv. No. 13053)}
+
 The older, erased layers, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
@@
 \end{figure}
-5. Pedagogical Considerations
+
+\section{Pedagogical Considerations}
 
@@
 
-6. Conclusion
+\section{Conclusion}
 
@@
 
+\section*{Aknowledgment}
+We wish to acknowledge the research assistance of T. M. Crandall,
+History student at Mount Allison University (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
+M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
+Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
+``La lettre romaine, entre inscriptions monumentales et écritures communes''. The following
+students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
+Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
+Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
+as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
+Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
+present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.
+
 \bibliographystyle{sapauth-eng}

### 2015-12-17 source/48/EAGLE16-lame-et-al-teaching.tex
few more changes

@@
 
-Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
+\noindent Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
 on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
@@
 
-The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
+\noindent The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
 found in Sisak (Siscia), one of the largest urban centers in south-western Pannonia. Most of them were found during the
@@
 
-In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
+\noindent In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
 far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
@@
 
-Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
+\noindent Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
 transcription and drawings) and to classify the written signs according to the following chart. Such preliminary
@@
 
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
-\bigskip
-
 \begin{flushleft}
@@
 \end{flushleft}
-
-\bigskip
 
@@
 
-From the educational standpoint, this pilot project succeeded tremendously. The students who undertook this
+\noindent From the educational standpoint, this pilot project succeeded tremendously. The students who undertook this
 transcription in place of the usual research essay were, on average, far more motivated than their peers, and took
@@
 
-Gaining expertise with material remains requires time. When a museum trains students, this time is often limited due to
+\noindent Gaining expertise with material remains requires time. When a museum trains students, this time is often limited due to
 constrained facilities and personnel. In addition, the epigraphic material may be fragile and delicate: the less these
@@
 \section*{Aknowledgment}
-We wish to acknowledge the research assistance of T. M. Crandall,
+\noindent We wish to acknowledge the research assistance of T. M. Crandall,
 History student at Mount Allison University (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of

### 2015-12-17 source/48/EAGLE16-lame-et-al-teaching.tex
added missing ref

@@
 case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
-is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\ Terras, 2015
+is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\citep{gorman_opening_2015}
 writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and

### 2015-12-18 source/42/Paper_licciardello.tex

@@
 the offering. The elements recurring in such inscriptions were the name of the dedicator, the verb of dedication and
-the name of the god receiving the dedication.\footnote{ On archaic dedications, see \citep[ 1-14]{day_archaic_2010} for
+the name of the god receiving the dedication.\footnote{ On archaic dedications, see \citep[spec. 1-14]{day_archaic_2010} for
 an introduction to the genre.} The most common and widespread formula for dedications contained exactly these three
@@
 A frequent variation was \begin{otherlanguage*}{greek}ὁ δεῖνά με ἀνέθηκε τῷ θεῷ\end{otherlanguage*}, \footnote{ The formula here presented (with its two variants) was \ identified by Maria Letizia Lazzarini, in
-her study on the formulas of archaic dedications \citep[58-40]{lazzarini_formule_1976}. Her analysis is based on both verse and prose
+her study on the formulas of archaic dedications \citep[58-60]{lazzarini_formule_1976}. Her analysis is based on both verse and prose
 inscriptions from the archaic age. However, such basic scheme, with these elements, was normal in epigrams as well and
@@
 surrounding it do not exist anymore – they have to be imagined into reader's mind.\footnote{ This involvement of the
-reader can be put in relation with what 
-%\citep{Bing1995} 
-defines Ergänzungsspiel, i.e. with the process of supplementation
+reader can be put in relation with what \citep{Bing1995} defines Ergänzungsspiel, i.e. with the process of supplementation
 of the text deliberately incited by Hellenistic epigrammatists.}
-%reference to Bing 1995 not in biblio, asked to Licciardello
 

### 2015-12-18 source/XX/EAGLE_2016_submission_XX.tex
Added title to XX

@@
 %%%%%%%%
+\title{Searching inscriptions through the EAGLE Portal}
 
-\title{}
-\author{claudio prandoni}
-\date{2015-11-09}
+%%%%%%%%
+%AUTHORS:%
+%%%%%%%%
+\author[PROMOTER]{Claudio Prandoni\corref{first}}
+% put the label for the address (like FENWAY here) in the square brackets; change the addresses below to match the labelled authors
+% mark the corresponding author with the \corref command and fill out the complete address in the \cortext command below
+
+\author[PROMOTER]{Antonella Fresa}
+\author[Gogate]{Nicola Alfarano}
+\author[ISTI]{Giuseppe Amato}
+\author[ISTI]{Franco Zoppi}
+\author[ISTI]{Andrea Mannocci}
+\author[HEI]{Pietro Liuzzo}
+\author[DAI]{Francesco Mambrini}
+\author[ROME]{Silvia Orlandi}
+\author[ROME]{Raffaella Santucci}
+%Promoter Srl1, Gogate Srl2, CNR-ISTI3, Heidelberg University4, German Archeological Institute5, Sapienza University of Rome6
+
+%\author[ENEA]{Francesco Biccari}
+
+%Additional information about corresponding author, addresses and affiliations:
+\cortext[first]{Corresponding author. Email:  prandoni@promoter.it}
+
+\address[PROMOTER]{Promoter Srl, Via Boccioni 2, 56037 Peccioli, Italy}
+\address[Googate]{Gogate Srl, Via Malasoma 24, 56121 Ospedaletto, Italy}
+\address[ISTI]{CNR-ISTI, Via G. Moruzzi 1, 56124 Pisa, Italy }
+\address[HEI]{Ruprecht-Karls-Universität Heidelberg, Marstallstraße 6, 69117, Heidelberg (DE)}
+%\address[ENEA]{ENEA, Casaccia Research Center, via anguillarese 301, 00123 Roma, Italy}
+\address[DAI]{Deutsches Archäologisches Institut, 69-71 Podbielskiallee, 14195 Berlin, Germany}
+\address[ROME]{La Sapienza University of Rome, Piazzale Aldo Moro 5, 00185 Roma }
+
+%%%%%%%%%%%%
+%PAPER CONTENT %
+%start your paper here%
+%%%%%%%%%%%%
+
 \begin{document}
-[Warning: Draw object ignored]Abstract
 
-\ \ \ \ \ \ This paper describes the functionalities and the technical infrastructure of the EAGLE Portal, the main
+\maketitle
+
+\begin{abstract}
+This paper describes the functionalities and the technical infrastructure of the EAGLE Portal, the main
 gateway into the EAGLE’s massive epigraphic database. The portal can be accessed both by a human through a web browser
@@
 to the research in this field.
+\end{abstract}
 
-Keywords
+\keywords{Search engine, API, image recognition, controlled vocabulary, multilingual translations, advanced search, citizen
+participation}
 
-Search engine, API, image recognition, controlled vocabulary, multilingual translations, advanced search, citizen
-participation.
+\section{Introduction}
 
-1.1. Introduction
-
-The EAGLE Portal is the place where the content provided by the epigraphers’ community is aggregated and stored and
+\noindent The EAGLE Portal is the place where the content provided by the epigraphers' community is aggregated and stored and
 where it is made accessible to the users. It is composed by:
@@
 The Aggregator relies on the D-NET software toolkit, developed within the DRIVER project and extended within the
-OpenAIRE, OpenAIREplus\footnote{http://www.openaire.eu/}, EFG, EFG1914\footnote{http://www.europeanfilmgateway.eu/it}
-and HOPE\footnote{http://www.peoplesheritage.eu/} projects. D-NET is an open source, general-purpose software conceived
+OpenAIRE, OpenAIREplus\footnote{\url{http://www.openaire.eu/}}, EFG, EFG1914\footnote{\url{http://www.europeanfilmgateway.eu/it}}
+and HOPE\footnote{\url{http://www.peoplesheritage.eu/}} projects. D-NET is an open source, general-purpose software conceived
 to enable the realization and operation of Aggregative Data Infrastructures (ADIs) and to facilitate their evolution
@@
 (e.g. by the integration of new services), and scaled (e.g. storage and index replicas can be maintained and deployed
-on remote nodes to tackle multiple concurrent accesses or very-large data size) at run-time [Amato et al., 2013].
+on remote nodes to tackle multiple concurrent accesses or very-large data size) at run-time \citep{amato_aim_2013}.
 
-The GUI exposes all the content stored in the Aggregator providing to the users the following functionalities [Prandoni
-et al., 2014].
+The GUI exposes all the content stored in the Aggregator providing to the users the following functionalities \citep{prandoni_eagle_2014}.
 
@@
 This interface includes a series of functionalities to support data ingestion and storage as well as for importing,
-indexing, enriching and managing the harvested metadata [Amato et al., 2015].
+indexing, enriching and managing the harvested metadata \citep{amato_second_2015}.
 
-1.2. The EAGLE Collections
+\section{The EAGLE Collections}
 
-Classical Greek and Latin culture is at the very foundation of modern European identity. From philosophy to
+\noindent Classical Greek and Latin culture is at the very foundation of modern European identity. From philosophy to
 architecture, geometry to law, a variety of contemporary subjects and disciplines have their roots in the classical
@@
 
-Arachne\footnote{ http://arachne.uni-koeln.de } is the central object-database of the German Archaeological Institute
+Arachne\footnote{\url{http://arachne.uni-koeln.de}} is the central object-database of the German Archaeological Institute
 (DAI). Since 2001, Arachne has been integrating negative archives of ancient sculpture that went beyond the specialised
@@
 
-Epigraphic Database Bari (EDB)\footnote{ http://www.edb.uniba.it } is a project specialized in epigraphic documents of
-Christian patronage (third to eighth centuries, AD), including those contained in the Inscriptiones Christianae Vrbis
-Romae, nova series, voll. I-X, in civitate Vaticana 1922-1992 (= ICVR), and those edited in other bibliographical seats
+Epigraphic Database Bari (EDB)\footnote{\url{http://www.edb.uniba.it}} is a project specialized in epigraphic documents of
+Christian patronage (third to eighth centuries, AD), including those contained in the \emph{Inscriptiones Christianae Vrbis
+Romae, nova series}, voll. I-X, in civitate Vaticana 1922-1992 (= ICVR), and those edited in other bibliographical seats
 and/or not contained in the ICVR. Currently the inscriptions present in the EDB (counting those already online and
@@
 
-The task of the Epigraphic Database Heidelberg (EDH)\footnote{ http://edh-www.adw.uni-heidelberg.de } is the systematic
+The task of the Epigraphic Database Heidelberg (EDH)\footnote{\url{http://edh-www.adw.uni-heidelberg.de}} is the systematic
 entry of ancient Latin and bilingual (usually Latin and Greek) inscriptions into a complex database. The Epigraphic
@@
 
-Epigraphic Database Rome (EDR)\footnote{ http://edr-edr.it } was launched in 1999 as an experimental project aimed at
+Epigraphic Database Rome (EDR)\footnote{\url{http://edr-edr.it}} was launched in 1999 as an experimental project aimed at
 creating a unified database for ancient epigraphy. EDR is part of the international federation of Epigraphic Databases
@@
 
-Hispania Epigraphica Online\footnote{ http://eda-bea.es} was created in 2002 when an EU grant enabled a joint research
+Hispania Epigraphica Online\footnote{\url{http://eda-bea.es}} was created in 2002 when an EU grant enabled a joint research
 project between the Archivo Epigráfico de Hispania and Mag. K. Schaller, who was developing computing applications for
@@
 
-Petrae database\footnote{ http://petrae.tge-adonis.fr} is a system for the recording of Latin and Greek inscriptions
+Petrae database\footnote{\url{http://petrae.tge-adonis.fr}} is a system for the recording of Latin and Greek inscriptions
 developed at the Institut Ausonius, which collects epigraphic texts from the various regions in which its researchers
@@
 
-The Last Statues of Antiquity\footnote{ http://laststatues.classics.ox.ac.uk } is a project funded in 2009 by the ‘Arts
+The Last Statues of Antiquity\footnote{\url{http://laststatues.classics.ox.ac.uk}} is a project funded in 2009 by the ‘Arts
 and Humanities Research Council’ in Oxford. The project collects and analyses all the evidence for new, newly
-dedicated, or newly re-worked statuary in the period circa 284–650. The two main results of the project are a major
+dedicated, or newly re-worked statuary in the period \textit{circa} 284–650. The two main results of the project are a major
 database, with over 2600 individual entries, and a book, published in late 2012, discussing in print the entire
@@
 
-The picture database VBI ERAT LUPA\footnote{ http://www.ubi-erat-lupa.org } contains stone monuments (sculptures,
+The picture database VBI ERAT LUPA\footnote{\url{http://www.ubi-erat-lupa.org}} contains stone monuments (sculptures,
 reliefs, inscriptions, architectural pieces etc.). The project’s scope ranges from prehistoric stone monuments to
@@
 Network and made available their collections that are currently being uploaded in the EAGLE Portal\footnote{ For a full
-list of the EAGLE Associated Partners see http://www.eagle-network.eu/about/partners/ }.
+list of the EAGLE Associated Partners see \url{http://www.eagle-network.eu/about/partners/}}.
 
-1.3. The EAGLE inscription search engine
+\section{The EAGLE inscription search engine}
 
-The EAGLE Inscriptions Search Engine is accessible through the main horizontal navigation bar of the EAGLE
-Portal\footnote{ http://www.eagle-network.eu/basic-search/ }. It represents the core functionality of the portal,
+\noindent The EAGLE Inscriptions Search Engine is accessible through the main horizontal navigation bar of the EAGLE
+Portal\footnote{\url{http://www.eagle-network.eu/basic-search/}}. It represents the core functionality of the portal,
 through which the entry of keywords and phrases produce matches from EAGLE’s massive epigraphic database.
 
-The EAGLE Portal makes available a “simple search”, an “advanced search” where the user can specify the values of a
-number of fields in order to make a more accurate search, and an “image search” where the user can upload an image and
-search for inscriptions that look similar to the provided one [Prandoni et al., 2014].
+The EAGLE Portal makes available a ``simple search'', an ``advanced search'' where the user can specify the values of a
+number of fields in order to make a more accurate search, and an ``image search'' where the user can upload an image and
+search for inscriptions that look similar to the provided one \citep{prandoni_eagle_2014}.
 
 The objects that a user can search in the EAGLE Portal belong to three different categories, in accordance with the
-EAGLE conceptual model [Sicila et al., 2015]:
+EAGLE conceptual model \citep{sicilia_eagle_2015}:
 
 \begin{itemize}
-\item “Artefacts”, which contain all the information that is somehow related to the physical carrier of the inscription.
-\item “Texts”, which contain all the information that is textual in nature.
-\item “Images”, which contain all the information that is visual in nature.
+\item ``Artefacts'', which contain all the information that is somehow related to the physical carrier of the inscription.
+\item ``Texts'', which contain all the information that is textual in nature.
+\item ``Images'', which contain all the information that is visual in nature.
 \end{itemize}
 
-\bigskip
+\subsection{How to search inscriptions}
 
-1.3.1. How to search inscriptions
-
-\bigskip
-
-The simple search user interface is very straightforward. The text entered in the query box is used to make a full text
+\noindent The simple search user interface is very straightforward. The text entered in the query box is used to make a full text
 search in all the fields of all the EAGLE objects in the category determined when making the query. By default, the
-search will be done on all the objects in the “Artefact” category. Two more tabs on the result list (labelled “Texts”
-and “Images”) allow the user to perform the same query choosing a different category.
+search will be done on all the objects in the ``Artefact'' category. Two more tabs on the result list (labelled ``Texts''
+and ``Images'') allow the user to perform the same query choosing a different category.
 
@@
 results: findspot, bibliography, text of the inscription, type of inscription, decoration, object type, material, type
-of writing, state of preservation (Fig. 1).
+of writing, state of preservation (Fig. \ref{fig:figure1}).
 
 In the fields having a controlled vocabulary, the user is allowed to enter only values coming from the vocabulary. For
-this purpose, those fields have a “drop-down menu” which displays all the defined values for that field in the
-“preferred label”, i.e. the text string that has been indicated in the vocabulary as the preferred one for display,
+this purpose, those fields have a ``drop-down menu'' which displays all the defined values for that field in the
+``preferred label'', i.e. the text string that has been indicated in the vocabulary as the preferred one for display,
 regardless of the language in which the string is defined.
@@
 integrated in the EAGLE Portal and that allows to search for images that are similar to the one a user has uploaded,
-ranked in decreasing order of similarity (Fig. 2).
+ranked in decreasing order of similarity (Fig. \ref{figure2}).
 
-\bigskip
-
+\begin{figure}[!bp]
+\centering
  \includegraphics[width=10.874cm,height=12.885cm]{EAGLE2016submissionXX-img001.png} 
-
-Fig. 1. Search for inscriptions in the EAGLE Portal
-
-\bigskip
+\caption{Search for inscriptions in the EAGLE Portal}
+\label{fig:figure1}
+\end{figure}
 
@@
 For example, images containing reproductions of works of art contain a lot of implicit information that is not
-generally captured in manually or automatically generated metadata [Amato et al., 2013].
+generally captured in manually or automatically generated metadata \citep{amato_aim_2013}.
 
@@
 
-\bigskip
-
+\begin{figure}[!bp]
+\centering
  \includegraphics[width=10.848cm,height=10.451cm]{EAGLE2016submissionXX-img002.png} 
-
-Fig. 2. Image search in EAGLE
-
-\bigskip
+\caption{Image search in EAGLE}
+\label{figure2}
+\end{figure}
 
@@
 
-\bigskip
+\begin{figure}[!bp]
+\centering
+ \includegraphics[width=10.874cm,height=12.197cm]{EAGLE2016submissionXX-img003.png} 
+\caption{Results list - Artefacts}
+\label{figure3}
+\end{figure}
 
- \includegraphics[width=10.874cm,height=12.197cm]{EAGLE2016submissionXX-img003.png}
+\subsection{Filtering and browsing results}
 
-Fig. 3. Results list - Artefacts
-
-\bigskip
-
-1.3.2. Filtering and browsing results
-
-\bigskip
-
-After having submitted a query, the user is presented with a paginated list of results containing some basic information
-(Fig. 3).
+\noindent After having submitted a query, the user is presented with a paginated list of results containing some basic information (Fig. \ref{figure3}).
 
@@
 of the items in the result page will display a summary of all the information available for that item, with links to
-get further details (Fig. 4).
+get further details (Fig. \ref{figure4}).
 
-\bigskip
-
+\begin{figure}[!bp]
+\centering
 \includegraphics[width=10.874cm,height=8.864cm]{EAGLE2016submissionXX-img004.png}
-
-Fig. 4. Object details
-
-\bigskip
+\caption{Object details}
+\label{figure4}
+\end{figure}
 
@@
 
-Duplicates identification has been performed through the Trismegistos platform\footnote{ http://www.trismegistos.org/ }.
+Duplicates identification has been performed through the Trismegistos platform\footnote{\url{http://www.trismegistos.org/}}.
 By standardizing publication references, collection information, material, provenance and date, the overlap of the
@@
 
-If more than one instance of an object is available, the other instances appear on the summary page as clickable “tabs”,
+If more than one instance of an object is available, the other instances appear on the summary page as clickable ``tabs'',
 each tab being labelled with the local ID of the other instances of the same object.
@@
 describing the term, where further connections, synonyms and translations can be explored. The EAGLE vocabularies are
-accessible also as Linked Open Data\footnote{ http://www.eagle-network.eu/resources/vocabularies/}.
+accessible also as Linked Open Data\footnote{\url{http://www.eagle-network.eu/resources/vocabularies/}}.
 
 Finally, all the peer-reviewed translations in several European languages that are available, have been integrated in
-the EAGLE Portal too through a RESTful API native of Wikibase, the MediaWiki extension used for this purpose\footnote{
-http://www.eagle-network.eu/wiki/api.php }. For each inscription it is possible to view the existing translation, to
-request a new one and to contribute by connecting to the EAGLE MediaWiki\footnote{ http://www.eagle-network.eu/wiki }.
+the EAGLE Portal too through a RESTful API native of Wikibase, the MediaWiki extension used for this purpose\footnote{\url{http://www.eagle-network.eu/wiki/api.php}}.
+For each inscription it is possible to view the existing translation, to
+request a new one and to contribute by connecting to the EAGLE MediaWiki\footnote{\url{http://www.eagle-network.eu/wiki}}.
 
-\bigskip
+\subsection{Annotating and saving objects}
 
-\ 1.3.3. Annotating and saving objects
-
-\bigskip
-
-A registered user, during a “web” or a “mobile” session, has the capability of annotating and saving (some of) the
+A registered user, during a ``web'' or a ``mobile'' session, has the capability of annotating and saving (some of) the
 information that is being provided by the EAGLE system.
@@
 
-1.4. Conclusion
+\section{Conclusion}
 
@@
 
-In this framework, the CIVIC EPISTEMOLOGIES\footnote{ http://www.civic-epistemologies.eu } project developed a Roadmap
+In this framework, the CIVIC EPISTEMOLOGIES\footnote{\url{http://www.civic-epistemologies.eu}} project developed a Roadmap
 for the use of e-Infrastructures to support the participation of European citizens in CH practices and humanities
 research, where such engagement has a twofold benefit for culture: to be enriched by the citizens’ contributions and to
-become more widely used and exploited (also, for example, with the participation of creative industries) [Fresa \&
-Justrell, 2015]. The future development of the EAGLE initiative will refer to the CIVIC EPISTEMOLOGIES Roadmap for
+become more widely used and exploited (also, for example, with the participation of creative industries) \citep{fresa_roadmap_2015}.
+The future development of the EAGLE initiative will refer to the CIVIC EPISTEMOLOGIES Roadmap for
 investigating how to put in place more advanced services supporting citizen science in epigraphic research.
@@
 The EAGLE Portal is only an example of application that makes use of the APIs provided by the Aggregator to access the
-EAGLE collections. Other examples are the Flagship Mobile Application\footnote{
-http://www.eagle-network.eu/resources/flagship-mobile-app/ }, that allows to access the inscriptions’ database through
+EAGLE collections.
+Other examples are the Flagship Mobile Application\footnote{\url{http://www.eagle-network.eu/resources/flagship-mobile-app/}}, that allows to access the inscriptions’ database through
 a mobile device and to fully exploit the image recognition features integrated in the EAGLE Portal, and the Flagship
-Storytelling Application\footnote{ http://www.eagle-network.eu/stories/ }, that allows a user to create stories
+Storytelling Application\footnote{\url{http://www.eagle-network.eu/stories/}}, that allows a user to create stories
 starting from the content available in EAGLE, Europeana and many other data sources.
@@
 through which it is possible to search and browse the full set of content\footnote{ For further information see
-http://wiki.apache.org/solr/\#Search\_and\_Indexing }. Furthermore, APIs are provided to register a new user, login and
+\url{http://wiki.apache.org/solr/\#Search\_and\_Indexing}}. Furthermore, APIs are provided to register a new user, login and
 perform an image-based search using the EAGLE image recognition algorithm.

### 2015-12-20 source/48/EAGLE16-lame-et-al-teaching.tex
Paper 48 (Lamé et al) added, but not checked

--- 2015-12-17 source/48/EAGLE16-lame-et-al-teaching.tex
@@
 \usepackage[utf8x]{inputenc}
+
+\usepackage[normalem]{ulem}
 % if you write in english and in greek
@@
 \usepackage[greek,english]{babel}
+\usepackage{teubner}
 \languageattribute{greek}{polutoniko}
@@
 % remove if not necessary
-\usepackage{siunitx}
+%\usepackage{siunitx}
 %Natbib for bibliography management
@@
 \title{Teaching (Digital) Epigraphy}
-\author[ilc]{Marion Lamé\corref{first}}
+\author[ilc]{Marion Lamé}
 \author[zag]{Ivan Radman-Livaja}
 \author[mau]{Bruce Robertson}
-\author[isti]{Federico Ponchio}
+\author[isti]{Federico Ponchio\corref{first}}
 \address[ilc]{Istituto di Linguistica Computazionale}
@@
 reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
-Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the Tesserarum Sisciae Sylloge (TSS) through
+Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the \emph{Tesserarum Sisciae Sylloge} (TSS) through
 a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
@@
 carries personal names, the other side carries an inscription mentioning the merchandise, most of the time in an
-abbreviated form, as well as a price - normally expressed in denarii or fractions of the denarius - and quite often an
+abbreviated form, as well as a price - normally expressed in \emph{denarii} or fractions of the \emph{denarius} - and quite often an
 indication of quantity or weight.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img001.jpg}
+\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img002.jpg}
+\caption{No. 12582.}
+\label{fig:12582}
+\end{figure}
 
@@
 \centering
-\includegraphics[width=2.944cm,height=2.752cm]{EAGLE16lameetalteaching-img001.jpg}
+
 \end{figure}
-\begin{figure}
-\centering
-\includegraphics[width=2.944cm,height=2.82cm]{EAGLE16lameetalteaching-img002.jpg}
-\end{figure}
-Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like lana, pan(n)um, tvnica,
-sagvm, p(a)envla, pal(l)a, palliolum, lodix, banata, and abolla, appear regularly enough without being abbreviated and
-thus the interpretation of common abbreviations like l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab is
+Most of the tags, if not all, are linked to the wool trade and the textile industry.
+Words like \emph{LANA, PAN(N)UM, TVNICA, SAGVM, P(A)ENVLA, PAL(L)A, PALLIOLUM, LODIX, BANATA}, and \emph{ABOLLA}, appear regularly enough without being abbreviated and
+thus the interpretation of common abbreviations like \emph{l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab} is
 not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
@@
 (i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
-correct fee\footnote{\citep{radman_livaja_segestica..._2008}, \citep{radman-livaja_roetiquettes_2010}, \citep{radman-livaja_5._2013}, \citep[165-172]{radman-livaja_two_2013}, \citep{radman-livaja_plombs_2014},  for analogous tags with \citep[97-104]{mocsy_olom_1956}; \citep[195-210]{egger_funf_1968}; \citep[127-138]{frei-stolba_les_2011}; \citep[121-137/93, 215-222]{schwinden_romerzeitliche_1985}; \citep{romer-martijnse_romerzeitliche_1990}; \citep[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citep[301-305]{feugere_etiquette_1993}; \citep[211-220]{weiss_bleietiketten_1991}; \citep[29-40]{paci_etichette_1995}; \citep[207-216]{bassi_tre_1996}; \citep[121-135]{bizzarini_quattro_2005}; \citep[43-51]{buchi_etichette_2005}; \citep[42-110]{cresci_supellex_2010}; \citep[295-317]{jacques_artisanat_2010}; \citep[237-246]{wedenig_bleietikett_2013}.}.
+correct fee\footnote{\citet{radman_livaja_segestica_2008}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014},  for analogous tags with \citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_2011}; \citet[121-137/93, 215-222]{schwinden_romerzeitliche_1985}; \citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.
 
@@
 didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
-in order to allow an entire team to reach its goal. For more information see: http://www.labcd.unipi.it/laboratorio .}.
+in order to allow an entire team to reach its goal. For more information see: \url{http://www.labcd.unipi.it/laboratorio}.}.
 This experiment showed that their contribution can be useful, because technology may compensate for their lack of
@@
 case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
-is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\ Terras, 2015
+is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\citet{gorman_opening_2015}
 writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
@@
 recover the linking information between wSystem and tSystem (e.g. browsing by glyph's shape and interpretation,
-identifying glyph's position), and easily shared with anyone with a browser (e.g.: fig. 3.1). The linking tool MarkOut
+identifying glyph's position), and easily shared with anyone with a browser (e.g.: fig. \ref{fig:markout}). The linking tool MarkOut
 is one of the editing components of the modular DAP. They fit into communication workflows\footnote{This raises, of
 course, questions, most of them outside the purview of this article, about the relationship that evolves between
-research and our modern societies. We invite the reader to peruse various initiatives such as Sabattini, 2006 on the «
-valeur d'ancienneté » of cultural heritage, that participates to social well-being, the \ Civic Epistemologies
-http://www.civic-epistemologies.eu and its Berlin charter, as well as Public History approaches
-http://public-history-weekly.oldenbourg-verlag.de.} such as crowdsourcing, editing inscriptions, dialoguing with
-experts, preserving and promoting cultural heritage and, \ as here, teaching some skills in epigraphy.
+research and our modern societies. We invite the reader to peruse various initiatives such as \citet{bertocci_documenter_2006} on the «
+valeur d'ancienneté » of cultural heritage, that participates to social well-being, the Civic Epistemologies
+\url{http://www.civic-epistemologies.eu} and its Berlin charter, as well as Public History approaches
+\url{http://public-history-weekly.oldenbourg-verlag.de}.} such as crowdsourcing, editing inscriptions, dialoguing with
+experts, preserving and promoting cultural heritage and, as here, teaching some skills in epigraphy.
 
-Fig. 3.1. MarkOut screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
+Fig. \ref{fig:markout}
+
+\begin{figure}[!bp]
+\centering
+\includegraphics[scale=0.10]{EAGLE16lameetalteaching-img003.png}
+\includegraphics[scale=0.10]{EAGLE16lameetalteaching-img004.png}
+\caption{MarkOut screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
 Latin characters with the position and the shape of the writing: U+0045 with II (identified as E2 by students) and
-U+004C with L (L1 or L5).
+U+004C with L (L1 or L5).}
+\label{fig:markout}
+\end{figure}
 
-\begin{figure}
-\centering
-\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img003.png}
-\end{figure}
-\begin{figure}
-\centering
-\includegraphics[width=4.941cm,height=2.281cm]{EAGLE16lameetalteaching-img004.png}
-\end{figure}
 Students were involved for a very short period of time: on average two hours per day over a couple of days only (in
@@
 
-Students encountered their first epigraphic challenges with tag no. 12582 (Radman-Livaja 2014: cat 03.13, 347), found in
+Students encountered their first epigraphic challenges with tag no. 12582 \citep[cat 03.13, 347]{radman-livaja_plombs_2014}, found in
 the Kupa river in Sisak and offered by Andrija Colussi to the Archaeological Museum in 1898. It is a rather typical
@@
 
-Obverse (palimpsest)\ \ \ \ Reverse\ \  \ \ \ \ \ \ \ \ \ \ \ \ (palimpsest)
+\begin{minipage}[t]{0.25\textwidth}
+\textbf{Obverse}\\
+At\d{e}\d{r}\\
+ivs
+\end{minipage}
+\begin{minipage}[t]{0.25\textwidth}
+(palimpsest)\\
+\d{S}\d{i}\d{i}\d{x}\d{t}\d{i}\\
+i onis n\\
+i . n i 
+\end{minipage}
+\begin{minipage}[c]{0.30\textwidth}
+%\begin{figure}
+%\centering
+\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img005a.png}
+%\end{figure}
+\end{minipage}
 
-\begin{figure}
-\centering
-\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img005.png}
-\end{figure}
-\begin{figure}
-\centering
-\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img005.png}
-\end{figure}
-Ate[323?]r[323?]\ \  \ S[323?]i[323?]i[323?]x[323?]t[323?]i[323?]\ \  \ \ \ \ \ la(na) p(ondo) i\ \ . . . s
-
-ivs\ \  \ i \ \ onis \ n\ \ \ \ cor(ticea)\ \ \ \ . . v\ \
-
-\ \  \ i . n i\ \ \ \ \ \ [4FE?] =$-$ £\ \ \ \  \ iir
+\begin{minipage}[t]{0.25\textwidth}
+\textbf{Reverse}\\
+la(na) p(ondo) i\\
+cor(ticea)\\
+\sout{X} =-\pounds\\
+\end{minipage}
+\begin{minipage}[t]{0.25\textwidth}
+(palimpsest)\\
+. . . s\\
+. . v\\
+  iir 
+\end{minipage}
+\begin{minipage}[c]{0.30\textwidth}
+%\begin{figure}
+%\centering
+\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img005b.png}
+%\end{figure}
+\end{minipage}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
-\end{figure}
-
 \section{Practicing Epigraphy}
@@
 
-\begin{figure}
-\centering
- [Warning: Image ignored] % Unhandled or unsupported graphics:
-%\includegraphics[width=7.329cm,height=4.404cm]{https://lh3.googleusercontent.com/HwLa5or_paTAyMS3Mu5XB3NucZXrDws6Jvc5WCC2ONHqBnBugWRtTF-YaKTLhJLhg6fShq3n_nG_37csBdKiKoSFA2tqVqhA0ovfyyT27h6KvGa-4slEmeURVSu_QG2AzjqTevFycw}
+\begin{minipage}[t]{0.18\textwidth}
+\textbf{Obverse}\\
+Aeli(i)\\
+Tasti
+\end{minipage}
+\begin{minipage}[c]{0.27\textwidth}
+%\begin{figure}
+%\centering
+\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
+%\end{figure}
+\end{minipage}
+\begin{minipage}[t]{0.18\textwidth}
+\textbf{Reverse}\\
+\d{l}odicem\\
+murtiola\d{m}\\
+pondo vi s\\
+\sout{X} vi s\\
+\end{minipage}
+\begin{minipage}[c]{0.25\textwidth}
+%\begin{figure}
+%\centering
+\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img007.png}
+%\end{figure}
+\end{minipage}
 
-\end{figure}
 No. 12563 is a rectangular lead tag of irregular form with a badly damaged and scratched surface. Nevertheless, the most
@@
 
-\begin{figure}
+\begin{table}
+{\small
+\addtolength{\tabcolsep}{-0.7mm}
+\begin{tabular*}{\textwidth}{llllllllll}
+\toprule
+\textbf{12563 (obv)} & \textbf{A} & \textbf{E} & \textbf{L} & \textbf{I} & \textbf{T} & \textbf{A} & \textbf{S} & \textbf{T} & \textbf{I}\\
+\midrule
+\textbf{Stud. 1} & \textit{A10} & \textit{E2} & \textit{L11} & \textit{I1} & \textit{T1} & \textit{A10} & \textit{S3} & \textit{T2} & \textit{I1} \\
+\textbf{Stud. 2} & \textit{A10} & \textit{E2} & \textit{L5} & \textit{I1} & \textit{T2} & \textit{A10} & \textit{S4?} & \textit{T2} & \textit{I1}  \\
+\bottomrule
+\end{tabular*}
+
+\begin{tabular*}{\textwidth}{llllllllllllll}
+ & & & & & & & & & & & & & \\
+\toprule
+\textbf{12563 (rev)} & \textbf{L} & \textbf{O} & \textbf{D} & \textbf{I} & \textbf{C} & \textbf{E} & \textbf{M} & & & & & &\\
+\midrule
+\textbf{Stud. 1} & \textit{L2} & \textit{O3} & \textit{D3} & \textit{I1} & \textit{C2} & \textit{E2} & \textit{M5} &  &  & & & & \\
+\textbf{Stud. 2} & \textit{?} & \textit{O3} & \textit{D5} & \textit{I2?} & \textit{C2} & \textit{E3} & \textit{M5} & & & & & &  \\
+%\bottomrule
+%\toprule
+%& & & & & & & & & & & & & \\
+\midrule
+ & \textbf{M} & \textbf{U} & \textbf{R} & \textbf{T} & \textbf{I} & \textbf{O} & \textbf{L} & \textbf{A} & \textbf{M} & \textbf{P} & \textbf{V} & \textbf{I} & \textbf{S}\\
+ \midrule
+\textbf{Stud.
+1}& \textit{M3} & \textit{V3} & \textit{R7} & \textit{T2} & \textit{I1} & \textit{O3} & \textit{L2} & \textit{A10} & \textit{M2} & \textit{P1} & \textit{V3} & \textit{I1} & \textit{S5} \\
+\textbf{Stud.
+2} & \textit{M3?} & \textit{V4} & \textit{R6} & \textit{T2} & \textit{I1} & \textit{O3} & \textit{L5} & \textit{A10} & \textit{M5} & \textit{P2} & \textit{V4} & \textit{I1} & \textit{S5} \\
+\bottomrule
+\end{tabular*}}
+
+\caption{No. 12563.}
+\label{tab:table1}
+\end{table}
+
+\begin{figure}[!bp]
 \centering
-\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img007.png}
-\end{figure}
-Obverse\ \ \ \ \ \ Reverse
-
-Aeli(i)\ \ \ \ \ \ \ \ l[323?]odicem
-
-Tasti\ \ \ \ \ \ \ \ murtiolam[323?]
-
-\ \ \ \ \ \ \ \ p(ondo) vi s
-
-\ \ \ \ \ \ \ \ [4FE?] vi s
-
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{|m{1.4399999cm}|m{0.518cm}|m{0.37700003cm}|m{0.361cm}|m{0.297cm}|m{0.361cm}|m{0.518cm}|m{0.486cm}|m{0.361cm}|m{0.30300003cm}|}
-\hline
-\centering 12563 (obv) &
-A &
-E &
-L &
-I &
-T &
-A &
-S &
-T &
-I\\\hline
-\centering Stud.1 &
-A10 &
-E2 &
-L1 &
-I1 &
-T1 &
-A10 &
-S3 &
-T2 &
-I1\centering
-\includegraphics[width=2.563cm,height=2.579cm]{EAGLE16lameetalteaching-img008.png}
-\\\hline
-\centering Stud. 2 &
-A10 &
-E2 &
-L5 &
-I1 &
-T2 &
-A10 &
-S4? &
-T2 &
-I1\\\hline
-\end{supertabular}
-\end{flushleft}
-
-\begin{figure}
-\centering
-\includegraphics[width=2.461cm,height=2.104cm]{EAGLE16lameetalteaching-img009.png}
-\end{figure}
-\begin{figure}
-\centering
-\includegraphics[width=2.268cm,height=2.104cm]{EAGLE16lameetalteaching-img009.png}
+\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img009.png}
+\caption{12563 Obverse. The two students' reconstructions}
+\label{fig:1}
 \end{figure}
 
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{|m{1.393cm}|m{0.361cm}|m{0.40900004cm}|m{0.40900004cm}|m{0.439cm}|m{0.393cm}|m{0.37700003cm}|m{0.444cm}|}
-\hline
-\centering 12563 (rev)\centering
-\includegraphics[width=2.898cm,height=2.656cm]{EAGLE16lameetalteaching-img008.png}
- &
-\centering L &
-\centering O &
-\centering D &
-\centering I &
-\centering C &
-\centering E &
-\centering\arraybslash M\\\hline
-\centering Stud. 1 &
-L2 &
-O3 &
-D3 &
-I1 &
-C2 &
-E2 &
-M5\\\hline
-\centering Stud. 2 &
-? &
-O3 &
-D5 &
-I2? &
-C2 &
-E3 &
-M5\\\hline
-\end{supertabular}
-\end{flushleft}
-
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{|m{0.58cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.361cm}|m{0.297cm}|m{0.40900004cm}|m{0.361cm}|m{0.518cm}|m{0.439cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.297cm}|m{0.34900004cm}|}
-\hline
-\centering M &
-\centering V &
-\centering R &
-\centering T &
-\centering I &
-\centering O &
-\centering L &
-\centering A &
-\centering M &
-\centering P &
-\centering V &
-\centering I &
-\centering\arraybslash S\\\hline
-M3 &
-V3 &
-R7 &
-T2 &
-I1 &
-O3 &
-L2 &
-A10 &
-M2 &
-P1 &
-V3 &
-I1 &
-S5\\\hline
-M3? &
-V4 &
-R6 &
-T2 &
-I1 &
-O3 &
-L5 &
-A10 &
-M5 &
-P2 &
-V4 &
-I1 &
-S5\\\hline
-\end{supertabular}
-\end{flushleft}
+\begin{figure}[!bp]
+\centering
+\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img008.png}
+\caption{12563 Reverse. The two students' reconstructions}
+\label{fig:2}
+\end{figure}
 
@@
 
-\begin{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[width=2.009cm,height=1.96cm]{EAGLE16lameetalteaching-img010.png}
+%\end{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[width=2.103cm,height=1.961cm]{EAGLE16lameetalteaching-img010.png}
+%\end{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img011.png}
+%\end{figure}
+
+\begin{minipage}[t]{0.3\textwidth}
+\textbf{Vale}\\
+rius lis\\
+Martia\\
+\end{minipage}
+\begin{minipage}[t]{0.3\textwidth}
+(palimpsest)\\
+\sout{X} \d{v}i\\
+tiir \d{r} i\\
+\end{minipage}
+\begin{minipage}[c]{0.3\textwidth}
+\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img011a.png}
+\end{minipage}
+
+\begin{figure}[!bp]
 \centering
-\includegraphics[width=2.009cm,height=1.96cm]{EAGLE16lameetalteaching-img010.png}
+\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img010.png}
+\caption{13053 Obverse. Students' reconstructions}
+\label{fig:3}
 \end{figure}
-\begin{figure}
-\centering
-\includegraphics[width=2.103cm,height=1.961cm]{EAGLE16lameetalteaching-img010.png}
-\end{figure}
-\begin{figure}
-\centering
-\includegraphics[width=2.044cm,height=1.794cm]{EAGLE16lameetalteaching-img011.png}
-\end{figure}
-Obverse \ \ \ (palimpsest)
-
-Vale\ \  \ \ \ [4FE?] v[323?]i\ \
-
-rius \ lis\ \  \ \ \ \ tiir \ r[323?] i
-
-Martia\ \  \ \ \ \
 
@@
 
-\begin{flushleft}
-\tablefirsthead{}
-\tablehead{}
-\tabletail{}
-\tablelasttail{}
-\begin{supertabular}{|m{1.513cm}|m{0.37700003cm}|m{0.518cm}|m{0.361cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.283cm}|m{0.37700003cm}|m{0.46900004cm}|m{0.439cm}|m{0.518cm}|m{0.518cm}|m{0.50200003cm}|m{0.297cm}|m{0.961cm}|}
-\hline
-\centering 13053 (obv.) &
-\centering V &
-\centering A &
-\centering L &
-\centering E &
-\centering R &
-\centering I &
-\centering V &
-\centering S &
-\centering M &
-\centering A &
-\centering R &
-\centering T &
-\centering I &
-\centering\arraybslash A\\\hline
-\centering Stud. 1 &
-V2 &
-A10 &
-L3 &
-E3 &
-R7 &
-I2 &
-V1 &
-S6 &
-M5 &
-? &
-R7 &
-T3 &
-? &
-Ignored\\\hline
-\centering Stud. 2 &
-V2 &
-A10 &
-L3 &
-E2 &
-R7 &
-I1 &
-V1 &
-S6? &
-M5 &
-A10 &
-R6? &
-T3? &
-I1 &
-A2\\\hline
-\end{supertabular}
-\end{flushleft}
+\begin{table}
+{\small
+\addtolength{\tabcolsep}{-0.8mm}
+
+\begin{tabular*}{\textwidth}{lllllllllllllll}
+ & & & & & & & & & & & & & \\
+\toprule
+%\textbf{12563 (rev)} & \textbf{L} & \textbf{O} & \textbf{D} & \textbf{I} & \textbf{C} & \textbf{E} & \textbf{M} & & & & & &\\
+%\midrule
+%\textbf{Stud. 1} & \textit{L2} & \textit{O3} & \textit{D3} & \textit{I1} & \textit{C2} & \textit{E2} & \textit{M5} &  &  & & & & \\
+%\textbf{Stud. 2} & \textit{?} & \textit{O3} & \textit{D5} & \textit{I2?} & \textit{C2} & \textit{E3} & \textit{M5} & & & & & &  \\
+%\bottomrule
+%\toprule
+%& & & & & & & & & & & & & \\
+%\midrule
+\textbf{13053} & \textbf{V} & \textbf{A} & \textbf{L} & \textbf{E} & \textbf{R} & \textbf{I} & \textbf{U} & \textbf{S} & \textbf{M} & \textbf{A} & \textbf{R} & \textbf{T} & \textbf{I} & \textbf{A}\\ \midrule
+\textbf{St.
+1}& \textit{V2} & \textit{A10} & \textit{L3} & \textit{E3} & \textit{R7} & \textit{I2} & \textit{V1} & \textit{S6} & \textit{M5} & \textit{?} & \textit{R7} & \textit{T3} & \textit{?} & \textit{Ignored} \\
+\textbf{St.
+2}& \textit{V2} & \textit{A10} & \textit{L3} & \textit{E2} & \textit{R7} & \textit{I1} & \textit{V1} & \textit{S6?} & \textit{M5} & \textit{A10} & \textit{R6?} & \textit{T3?} & \textit{?I1} & \textit{A2} \\
+\bottomrule
+\end{tabular*}}
+
+\caption{No. 13053 (obv).}
+\label{tab:table2}
+\end{table}
+
+%begin{flushleft}
+%tablefirsthead{}
+%tablehead{}
+%tabletail{}
+%tablelasttail{}
+%begin{supertabular}{|m{1.513cm}|m{0.37700003cm}|m{0.518cm}|m{0.361cm}|m{0.37700003cm}|m{0.37700003cm}|m{0.283cm}|m{0.37700003cm}|m{0.46900004cm}|m{0.439cm}|m{0.518cm}|m{0.518cm}|m{0.50200003cm}|m{0.297cm}|m{0.961cm}|}
+%hline
+%centering 13053 (obv.) &
+%centering V &
+%centering A &
+%centering L &
+%centering E &
+%centering R &
+%centering I &
+%centering V &
+%centering S &
+%centering M &
+%centering A &
+%centering R &
+%centering T &
+%centering I &
+%centering\arraybslash A\\\hline
+%centering Stud. 1 &
+%2 &
+%10 &
+%3 &
+%3 &
+%7 &
+%2 &
+%1 &
+%6 &
+%5 &
+% &
+%7 &
+%3 &
+% &
+%gnored\\\hline
+%centering Stud. 2 &
+%2 &
+%10 &
+%3 &
+%2 &
+%7 &
+%1 &
+%1 &
+%6? &
+%5 &
+%10 &
+%6? &
+%3? &
+%1 &
+%2\\\hline
+%\end{supertabular}
+%\end{flushleft}
 
@@
 
-\begin{figure}
+\begin{figure}[!bp]
 \centering
-\includegraphics[width=6.313cm,height=2.207cm]{EAGLE16lameetalteaching-img012.png}
+\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img012.png}
+\caption{12563 Obverse (palimpsest). The two students' reconstructions}
+\label{fig:4}
 \end{figure}

### 2016-01-13 source/28/EAGLE2016_Rocco_engrev.tex
worked on Anita's paper n.28
converted to tex, added biblio in master, added citations, compiled

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{EDB 2.0 
How Eagle Europeana project improved the Epigraphic Database Bari}
\author[uniba]{Anita Rocco \corref{first}}
\address[uniba]{Dip. di Scienze dell'Antichità e del Tardo Antico, Università degli Studi di Bari ``Aldo Moro''}
\cortext[first]{Corresponding author. Email: anita.rocco@uniba.it}
\begin{document}
\maketitle
\begin{abstract}
This paper is dedicated to the evolution of the Epigraphic Database Bari (EDB) from the minimalistic design of its
origins to its current status. Although EDB, the database of inscriptions by Christians from Rome, dates back to the
late 1980s , involvement in the EAGLE – Europeana project has had a significantly positive impact on its development.
In fact, maintaining its peculiar character, dictated by its own history and, mostly, by the characteristics of its
documentary base, it has taken advantage of the solutions adopted to integrate different archives and purpose-built
best practices. 
\end{abstract}
\keywords{Epigraphic database, EDB, EAGLE Europeana, Christian inscriptions, Late antique inscriptions }

\section{Introduction}

\noindent The Epigraphic Database Bari (EDB) is an `old' database, it dates back to the late 1980s , when Carlo Carletti\footnote{
He was professor of Epigrafia e antichità cristiane in the former Department of Classical and Christian Studies of Bari
University.}, inspired by Jory's experience indexing CIL VI\footnote{\citet{_corpus_1974}. The six volumes of the
computerized KeyWord-In-Context index to all the approximately 40,000 texts collected in the sixth volume of the Corpus
Inscriptionum Latinarum (CIL) are the outcome of a trailblazing work of arrangement and organization of the
inscriptions in a database, even if limited to their textual part.}, started a project of digitization of the
inscriptions commissioned by Christians from Rome between the third and eighth centuries, collected and edited in the
27,688 lemmas of the ICVR\footnote{\emph{The Corpus of Inscriptiones christianae urbis Romae} started in 1922 by A. Silvagni,
was published, mostly, by A. Ferrua, later supported by D. Mazzoleni and C. Carletti, between 1956 and 1992. Pursuing
the work of G.B. de Rossi of the mid-1800s (IC), ICVR registers the inscriptions by Christians found in the suburban
area of Rome, sorted in topographic order by consular road, then by catacomb. Inscriptions found inside the urban walls
or the recently discovered suburban ones aren’t yet included in the ICVR volumes.}. The data were stored in a data
processing program, called ICVR, running under MS-DOS, later converted to a database for Microsoft Access. It was
originally intended for internal use only.

Since its very beginnings the database has been designed on the basis of a conceptual model, which conveys the
complexity of epigraphs even in the frame of a very simple and basic IT structure.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img001.png} 
\caption{The record of a bilingual inscription (ICVR V, 15147) as it appeared in data processing program ICVR in MS-DOS
in 1988.}
\label{fig:1}
\end{figure}

In addition to the text, the data processing program recorded for each inscription\footnote{ The program didn’t allow
the use of Greek fonts; in order to add texts in the Greek alphabet it was necessary to type Latin equivalent letters
in the MS Word Symbol font, as you can see in Fig.~\ref{fig:1}}: bibliographic data (Progressivo = ICVR Number), origin
(Provenienza), type of support and technique of execution (Supp. e tecn.), function (Funzione), the presence or
absence of Christograms (\emph{Signa Christi}), and dating (Datazione). All this information was expressed by alphanumeric
codes of few characters, according to a limitation imposed by the program (Fig.~\ref{fig:1}).

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img002.jpg} 
\caption{The inscription ICVR V, 15147 as it appears now in EDB, with the acronyms expanded, the image and the text
displayed in Greek characters.}
\label{fig:2}
\end{figure}

Adding metadata to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
the epigraphic object in its widest sense as an inscribed artifact. This feature, in particular relating to
geographical information, was even more meaningful according to the characteristics of a large part of the inscriptions
recorded in EDB. In fact, the original pertinence to a monument/container (catacomb) or to a particular area of it,
allows, with reasonable certainty, to determine the patronage by a member of the Christian community and moreover to
determine the chronology, even without specific references inside the text\footnote{\citet{carletti_inscriptiones_1994}; \citet[viii-ix]{felle_introduzione_1997}.}. Likewise reporting the presence of Christological monograms allows us to identify them as explicit symbols
of Christian faith and as chronological indicators\footnote{ The monogram consisting of the first two letters of the
name \begin{otherlanguage*}{greek}Χριστός\end{otherlanguage*} can be considered the archetype of these signs. They first appear in inscriptions at the
beginning of the fourth century.}.

In the early 2000s, the ICVR database, containing more than 20,000 records, became part of the EAGLE federation of
databases (Electronic Archive of Greek and Latin Epigraphy), under the patronage of the Association Internationale
d’Épigraphie Grecque et Latine (AIEGL), as EDB (Epigraphic Database Bari) and extended its competencies to the
epigraphic documentation of Christian patronage of the city of Rome, published after the volumes of ICVR.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img003.png} 
\caption{The EDB homepage in 2004 and 2009.}
\label{fig:3}
\end{figure}

As a member of the federation, the database became available online through its own website and finally, thanks to the
EAGLE Europeana project, through a common portal.

Obviously this step has resulted in a series of changes and adjustments that led from the original basic structure of
the database to the present one.

\section{The EDB structure}

The current structure of EDB consists of a relational database, based on the open-source program My-SQL, with a complex
scheme drafted according to the most recent advances in epigraphic methodology: reestablishing historical and material
value of the object, identifying each epigraph as a complex and polysemic product consisting partially but not only of
text.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img004.png} 
\caption{The EDB database schema.}
\label{fig:4}
\end{figure}

\subsection{Bibliographic data}

As in the past version, the first pieces of information recorded for each epigraphic document are bibliographic data.

Besides data related to the ICVR publication - volume and edition number of the inscription - now EDB is able to record
other bibliographic references as well as concordances with other Corpora (CIL, IG, IGUR, IGC).

The whole bibliography is structured in metadata and is available both on the EDB website, and on the EAGLE BPN group of
Zotero\footnote{ \url{www.zotero.org/groups/eagleepigraphicbibliography/items }}, a tool for managing bibliographic data that
makes it easy to export entries and to cite them. An additional field allows the user to record references and links to
other online databases.

It's even possible to define the relationship between the epigraph and the cited bibliography (printed or digital):
identity, when it is an edition; integration, if it's the edition of another fragment of the same inscription;
opistographic, if it's the edition of the inscription on the back side; reuse, if it's the edition of another
inscription on the same support; comment, if it's a study on a related topic.

\subsection{Geographic data}

One key element of differentiation of EDB from other similar projects is the structuring of the topographic data.

This is due to the fact that the inscriptions of interest of EDB pertain only to the city of Rome, an area far more
limited than the large geographic ones managed by other epigraphic databases, but also and above all considering that,
as has been said previously, a large number of inscriptions of the Christians of Rome are still preserved in the place
for which they were created, sealing a tomb of an underground cemetery. Moreover, even if a given inscription isn't
still in its place on the grave, it is often still attributable to a specific area of the funerary complex.

Consequently in EDB, geographic indications require a more detailed articulation than in other databases, in which the
maximum level of definition is just the city of provenance, often lacking detailed information about the inscription’s
discovery.

Data on the original context are therefore organized hierarchically in three related fields containing controlled lists.
After selecting the area of the suburb identified by the name of the consular road - or by the number of the Augustan
regio for urban inscriptions - it's possible to select the monument from a list: a catacomb or part of it, if it is a
large and multi-layered one; a church; a public building or an urban area. The third field allows access to a further
associated list where the position of the epigraph inside the monument can be selected. For the catacombs in
particular, it’s possible to use these fields to annotate the gallery or the cubicle, named with the alphanumeric code
used in the maps published in ICVR.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img005.jpg} 
\caption{Original context input fields.}
\label{fig:5}
\end{figure}

Is worth noting that this set of associated fields refers to the original position of the inscriptions and not to the
place where they have been found, unless the two data coincide, that is in the case of inscriptions in situ or suo loco
adplicitae, in accordance with the definitions of the ICVR.

To complete information on spatial data, all cemeterial contexts have been georeferenced, so that clicking on their name
opens a new window in Google Map that shows the modern entrance to the cemetery, with the address and its coordinates
(Fig.~\ref{fig:6}).

 \begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
\caption{The original context maps.}
\label{fig:6}
\end{figure}

As a case of study, for the inscriptions in situ pertaining to the Domitilla catacomb (Via Ardeatina) - the only
cemeterial complex with a georeferenced plan of almost its entire extension\footnote{ START-Projekt: Die
Domitilla-Katakombe in Rome (Institut für Kulturgeschichte der Antike - Österreichische Akademie der Wissenschaften).
\citet{orlandi_case_2014}.} - clicking on the alphanumeric code\footnote{ The codes are made up of a majuscule letter
relating to a region of the catacomb \ and by another element, digit or minuscule letter, relating to a precise
internal position, respectively a gallery or a cubiculum. } associated to the precise position of the inscriptions,
gallery or cubiculum (F04, in Fig.~\ref{fig:6}), opens the plan of the specific area (regio) in which the inscription is found.
In every plan, the inscriptions preserved in situ are placed and marked by ICVR and EDB number (Fig.~\ref{fig:6})\footnote{The
plans are available on START-Projektwebsite: \url{www.oeaw.ac.at/antike/index.php\&id=431}, clicking on EDB number opens a
window with the EDB record. }.

Even the geographic data relating to the place of conservation of the inscriptions are managed with a similar structure.
Since not a few inscriptions, produced in Rome have been taken away and carried to other places in Italy or abroad, the
information have been organized in three related fields reporting respectively the list of the cities, the list of
associated structures, such as museums, churches or catacombs; and the specific positions in the context where the
object is actually preserved.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img007.png}
\caption{Conservation input fields.}
\label{fig:7}
\end{figure}

Adopting the best practice suggested by the EAGLE consortium, georeferencing is guaranteed even for the Conservation
data, linking every City/Town to GeoNames site\footnote{ \url{www.geonames.org}}, which allows the user to pinpoint the
location and to avoid ambiguity between homonyms. A link to the Trismegistos Collection\footnote{
\url{www.trismegistos.org}}, a database of papyrological and epigraphic collections, if available, also helps to identify
uniquely the place of conservation, as well as to obtain additional information, included the geographic
positioning\footnote{\url{http://www.eagle-network.eu/wp-content/uploads/2013/06/EAGLE\_D2.2.2\_Content-harmonisation-guidelines-including-GIS-and-terminologies-Second-Release.pdf}}.

\subsection{The description of the epigraphic object }

The nature of the inscription as material objects carrying textual information is represented by a series of attributes,
responding to the questions “What?” and “How?”: type of support and measures, technique of execution, height of letters
and paleographical features, and cases of reuse.

A survey of terminologies intended for description of epigraphic objects in the ICVR volumes has been carried out and
has generated lists of controlled terms for some of the fields (Fig.~\ref{fig:8}). \ Type of support, Executing technique and
Function vocabularies are aimed at classifying the specific and peculiar materials, methods and functions of the
inscriptions encoded in EDB, as the traditional epigraphic taxonomies do not totally adhere to their features.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img008.png}
\caption{Controlled vocabularies for Type of support, Executing technique and Function fields.}
\label{fig:8}
\end{figure}

The controlled lists have been integrated in the vocabularies of the EAGLE community\footnote{
\url{http://www.eagle-network.eu/resources/vocabularies/}.}, wich align, harmonize, create relations and translate into
various languages the terms used by the various partners, and returns them in a format that allows the user to get a
stable and unique identifier for each term, accessible and reusable by other users.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img009.png}
\caption{The integration with EAGLE Vocabularies, that can be opened clicking on the highlighted terms.}
\label{fig:9}
\end{figure}

\subsection{The text}

The nature of the inscription as a sequence of characters carried by a physical object is represented by a series of
fields related to various features of texts: language and alphabet (Latin, Greek and the multiform combination of
coexistence between them), and metrical structure.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img010.png}
\caption{Text fields in the submission mask.}
\label{fig:10}
\end{figure}

The proper text of the inscription is stored in an apposite field, following the Krummrey - Panciera
conventions\footnote{\citet{krummrey_criteri_1980}; \citet{Panciera1991}.}, with some adjustments to make it possible to describe
specific and peculiar issues of the inscriptions encoded in EDB. In particular the so-called
{\textquotedbl}aberrant{\textquotedbl} forms are not {\textquotedbl}normalized{\textquotedbl} to the
{\textquotedbl}standard{\textquotedbl} model, if they are recognized as grapho-phonetic outcomes of linguistic
modifications of Latin and Greek.

While the fidelity to what is written on the stone - or other type of support - respects and takes into due account the
evolution of Greek and Latin languages, it compromises the comprehension of the text and greatly complicates the
text-based search of terms. A standard query system, in fact, is not able to match a query with all the inscriptions
containing different spellings of a word. To resolve this issue each inscription is stored in its original form and in
a “lemmatized” form, where each term is actually replaced with its corresponding lemma, possibly by taking into account
its inflexed forms\footnote{ On the contrary, if the compiler recognizes aberrant forms as outcomes of misstatements
and material mistakes of the stonecutters, he transcribes them with the appropriate corrections, following the Krummrey
- Panciera conventions. \citet{felle_perspectives_2014} e \citet{orlandi_improving_2014}.}.

In EDB, as in every epigraphic database, the transcription of the text provides a real pre-edition, with systematic
expansion of abbreviations, hypothesis of integration, and, if possible, an update version of the text, according to
recent publications. It includes, moreover, the description of non-alphabetic signs - in double round brackets - such
as figurative elements (fishes, birds, anchors, etc.) and Christological monograms. Other fields record onomastic
notes, critical apparatus and textual comments.

A series of fields responding to the questions “When?” allows to insert a specific date, if recorded in the text, a
specific time span (such as the duration of the reign of a Bishop of Rome or of an emperor), or a generic interval.

\subsection{The images}

Another meaningful improvement of EDB 2.0 is represented by the inclusion of visual representation of inscriptions. It’s
evident how images dramatically enhance analysis of epigraphic materials, showing them in their manifold aspects:
reference with the context, form and quality of the support, graphic forms, peculiarities of technique, layout and
relationship between the text and any figurative or decorative elements\footnote{\citet{panciera_cronistoria_2006}.}.

In the frame of the collaborations with Europeana, the largest online collection of digitized items, EDB has been
encouraged to enlarge the image repository, based on a cooperation agreement established between the EAGLE consortium
and the Ministry of Culture (MIBACT) and the Pontifical Commission for Sacred Archaeology (PCAS). A new uploading
process allowed tripling photos, squeezes and casts to be published online in low resolution and with a visible digital
watermark, respecting the restrictive Italian rules. A large portion of the digital images stored in the EDB repository
have been taken by collaborators during past years, and others have been scanned from publications, while a large
number come from the Photographic Archive of PCAS\footnote{ \url{www.archeologiasacra.net}}.

\section{Users, interface, search engine}

The system manages three kinds of users: editors, compilers and generic, anonymous ones.

The latter of these can navigate in the descriptive section of the website (About EDB, People) and in the list of cited
Publications. They also have access to the entire database using two research masks: a Quick search allows the user to
search in only one of the following fields: identifier EDB, bibliographic data and text; an Advanced Search provides
the opportunity to explore the database through multiple search criteria variously combined.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img011.png}
\caption{Quick and Advanced Search masks.}
\label{fig:11}
\end{figure}

Like today's web search engines, EDB provides an advanced text search in Latin and Greek - an integrated tool
facilitates writing in the Greek alphabet\footnote{ Greek Inputter 2, developed by J. Naughton, allows the user to
write in Greek using his/her usual keyboard and to easily type various Greek diacritical marks
(\url{http://babel.mml.ox.ac.uk/ naughton / polytonic-greek-inputter.html}).} - allowing users to obtain different results
according to a default syntax, in the case of a search either for a single word or for a set of terms, in sequence or
not. Additionally, it's possible to choose whether or not to consider epigraphic diacritical marks, Greek accents and
spirits, and capitals. The textual search can be combined with other metadata related to bibliographic, geographic and
material data, or to function, reuse, language and date, expressed in a single year or in defined intervals.

This wide range of possibilities has been designed to reach users with different needs: the occasional user looking for
a particular inscription could just type one or more words that he is able to read and decipher, and the specialist
user, who can access detailed information about a single epigraph or use the advanced search to query the database
about groups of documents with common characteristics.

The search results are listed in a table showing the EDB identifier, bibliographic data, place of origin and place of
conservation, text of the inscription and a link to the full record.

\section{Conclusions}

This paper briefly describes the growth of Epigraphic Database Bari in nearly the last thirty years, from the first
experimental and minimalistic version, intended just for the use of a small group of researchers of Bari University, to
the present one, open to a large public of curious curious individuals, students and, of courses, specialists.

The involvement in the EAGLE – Europeana project, network of Ancient Greek and Latin Epigraphy, has had a significantly
positive impact on the development of the database of inscriptions by Christians from Rome.

In fact, although EDB, like other partner databases, has maintained its character, dictated by its own history and,
mostly, by the characteristics of its documentary base, it has taken advantage of the solutions adopted to integrate
different archives and purpose-built best practices.

Among the improvements, it’s worth mentioning

\begin{enumerate}
\item The inclusion of EDB bibliography, structured in metadata, in the Zotero Group, where it have been merged with those
of other content providers and has been made directly and publicly available in the most reusable way, giving more
exposition of the bibliographic database; allowing the integration and enrichment with other databases; allowing easy
export of data in multiple formats (bibtex, bookmarks, mods, rdf, xml, etc).

\item Following EAGLE best practice suggestions, data about modern places have been enriched with links to reference
resources, such as GeoNames and the Trismegistos Collection\footnote{
\url{http://www.eagle-network.eu/wp-content/uploads/2013/06/EAGLE\_D2.2.2\_Content-harmonisation-guidelines-including-GIS-and-terminologies-Second-Release.pdf}},
extending the use of stable and unique identifier accessible and reusable by other users (URI).

\item With the same aim, the controlled lists of Type of support, Executing technique and Function have been integrated in
the corresponding vocabularies of the EAGLE community. Among other benefits, such as alignment and relations between
databases, clicking on specific terms opens an EAGLE vocabulary window with a translation into various languages. This
feature is particularly useful in the case of EDB which, following ICVR, uses Latin for definitions, without modern
language translations.

\item Encouraged by the collaborations with Europeana, EDB has tripled the number of images stored in its repository,
including inedited images taken by collaborators over the years. 
\end{enumerate}

On the other hand, EDB, being inside Eagle since its origin, has been a bridgehead for the non standard epigraphies,
proposing issues to the Eagle community and pushing to make the data model more flexible. For example, EDB asked to add
more than one technique of execution for a single support and to add more than one language and / or alphabet for the
same inscription, change that is indispensable to describe bilingual and/or bigraphic texts.

Other solutions adopted in EDB could be, in the future, suitable for other projects, such as the hierarchic and
multi-step organization of topographic data, which could be applied to closed contexts of any age (houses, columbaria,
and so on) as well as the treatment of aberrant forms and the lemmatization process, which could be applied to any non
standard language.

CIL = Corpus Inscriptionum Latinarum, \ Berolini 1863 ss.

CIG = Corpus Inscriptionum Grecarum, I-IV, Berolini 1828-1877

IC = Inscriptiones christiane urbis Romae septimo saeculo antiquiores, I-II, G.B. de Rossi (a c.), Romae 1857-1861;
Supplementum, G. Gatti (a c.), Romae 1915.

IG = Inscriptiones Graecae, Berolini 1873 ss.

ICVR = Inscriptiones christianae urbis Romae septimo saeculo antiquiores. Nova series, I-X, A. Silvagni, A. Ferrua, D.
Mazzoleni, C. Carletti (a c.), Romae - In Civitate Vaticana 1922 ss.

IGVR = Inscriptiones grecae urbis Romae, I-IV, L. Moretti (a c.), Roma 1968-1990.

\section*{Acknowledgement}
The author would like to acknowledge the support of the European Commission through the
EAGLE project - Europeana network of Ancient Greek and Latin Epigraphy (Grant no: 325122). Moreover, the author thanks
Carlo Carletti, Antonio E. Felle, Filippo A. Piazzolla, Marida Pierno, and Gianvito Pio who daily use EDB and manage
the input of inscriptions and metadata, and the entire EAGLE community, which creates a space of real networking,
debate and the sharing ideas and data.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2016-01-13 source/32/EAGLE_2016_submission_32.tex
began conversion of paper 32
the pdf is in a tremendously bad state. conversion began anyway,
waiting for decision about this paper

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Towards the Publication of ICI Siracusa: General Data and Previews.}
\author[unica]{Mariarita Sgarlata\corref{first}}
\address[unica]{Department of Humanities - University of Catania} 
\cortext[first]{Corresponding author. Email: m.sgarlata@unict.it}
\begin{document}
\maketitle
\begin{abstract}
This paper refers to the general track “Epigraphic edition on paper and on line”. Some researchers who has been working on the editing of the Inscriptiones Christianae Italiae, published from the University of Bari, contributed also to the EDR project, that collect on line the epigraphic documents of Roman Christian Period. I propose a preview of the work currently in progress, with a specific reference to the inscriptions that provide us the chronological and topographic data to study the cemetery as well as formularies linked to the society structure and to the identity-making characteristics.
\end{abstract}
\keywords{Sicily, Syracuse, Epigraphy, Topography, Society, Onomastic and Identity}

\section{General Data about Epigraphs}

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{MISSING IMAGE!}
\caption{The main gallery (so called decumanus) of Catacombs of San Giovanni}
\label{fig:1}
\end{figure}

We are far from a complete and thorough research about palaeochristian epigraphs in Sicily. This catalogue represents the first step of a systematic study focused on the inscriptions discovered in the christian cemeteries of Siracusa.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{Sporus.JPG}
\caption{The Sporus’s inscription}
\label{fig:2}
\end{figure}

The epigraphic research related to San Giovanni catacomb can count on certain informations about the discovery place thanks to Paolo Orsi archeological campaigns. However, looking to previous publications, sometimes it’s necessary a review of the presentations made by Mommsen and Kaibel. In this perspective the main sources of the CIL and IG authors were verified so we had a deeper vision of the relationship between Mommsen and Kaibel and the main sources used, sometimes reevaluating the contribution of this sources. The renovated interest about funerary epigraphy generated new studies about new criteria of dating. The inscriptions dated by the indications of the ‘’consular couple’’ represent an exceptions in the issues about chronological datings (FERRUA 1947, 227; 1989, nn. 191-208): two were discovered during the excavation conducted by Paolo Orsi (1985) and to them is associated the latin epigraph ‘’Sporus’’ (356) .
The bigger one is entitled to San Giovanni and gave back almost 800 inscriptions, now located for the most part in the Archeological and Regional Museum ‘’Paolo Orsi’’ of Siracusa.; alongside will be placed the 40 headstones recovered in the adjacent hypogea of Villa Landolina, that report compatible datings and formularies.

1.2 HISTORICAL ASPECTS OF THE CATACOMB OF S. GIOVANNI AT SYRACUSE
The systematic studies of collective cemeteries in Syracuse started many years ago and, beyond the archaeological surveys, the late antique funerary settlements of Syracuse proposed multiple research cues and paths, as the historical- religious, economic and social nature.
Suburban cemeteries, fanned out from the area of Fusco, in the quarter of Neapolis, to the Santa Lucia area, in the southern part of Acradina, this indicates unequivocally what the perimeter of the city must have already been in the early and mid Roman Empire. The History of the area, which was going to hold the catacombs (San Giovanni, Vigna Cassia and Santa Lucia), spanned the centuries between the classical Greek and late antique ages, gradually giving evidence of quarries (Latomie), water supply systems to the city, characterized by cisterns and aqueducts (COLLIN BOUFFIER 1987, 682), handcraft workshops from the beginning of the 4th/3rd century BCE and burials datable to the early and mid Roman Empire. The analysis of the funerary system certifies one hand the dependence on the Roman model, and other the debt in respect of local traditions.
Several interest will be given to structural aspect of the catacomb of S. Giovanni, practice of funeral rituals, ethnic and cultural fruition’s characters, transformation in the use, transformation in the way of using spaces for graves, to complete a general point of view about the phenomena of continuity and innovation as to previous sepulchral arrangements and, in the analysed periods, the facies belonging to the different settling, variegated in the committees’ideological and religious themes, in choosing monumental types (like the rotundae) and decorations, in self-representative aspects, in burial uses. In this perspective we will give particular importance to the study of funerary epigraphy aimed at the writing of ICI Siracusa and the overall interpretation of the monument.
Between 1893 and 1909 the archaeologist Paolo Orsi carried out a series of campaigns in the catacomb of San Giovanni. Detailed reports of those campaigns are recorded at various times in Notizie degli Scavi (see the BIBLIOGRAPHY), which constitute an indefeasible starting point. The first incisive studies on Christian subterranean Sicily pertain to Joseph Führer, who dedicates many pages to the catacomb of San Giovanni, whose study is also epigraphic, is based on the previous literature, especially on Paolo Orsi’s discoveries (FÜHRER 1897, 13-39; FÜHRER- SCHULTZE 1907, 22-26; see also ORSI 1893, 276, n. 2 and SCHULTZE 1882, 60-61, fig. 5). In the study of Syracuse cemeteries Orsi (ORSI 1900a, 189) was the first to see the mixed nature of the burials and their materials (mostly gravestones), openly attesting a kind of pagan-Christian and orthodox-heterodox symbioses (see also AGNELLO 1957a, 242-243). Both Antonio Ferrua and Santi Luigi Agnello within the space of a few years tried to solve the problems relating to the sarcophagus of Adelfia and the more general problem, strictly correlated with the ones just mentioned, of the cemetery genesis and development (FERRUA 1952 and AGNELLO 1956). It is necessary to record that numerous hypotheses on the genesis and development of the various sectors of the cemetery are ascribable to the account given by inscriptions. Syracuse, thanks to its prolific underground cemeteries, has the larger Christian epigraphic heritage after Rome, kept in the storage of the Soprintendenza di Siracusa.
1.3 TOPOGRAPHIC AND EPIGRAFIC DATA TO STUDY THE CEMETERY
Walking across the main gallery, one can retrace the stages of Paolo Orsi’s interventions, being at the entrance of the Catacomb second northern gallery, before the so-called “tomb of the Saint”. This arcosolium is considered a privileged burial on the basis of the following reasons: most of all because of its position, but also because it is a mensatype burial closed by a sole slab (ORSI 1893, 292-294). The signs of an ancient rite are easily traceable on the slab, a rite that preceded the coming of Christianity and persisted for centuries up to the present day: the rite of refrigerium, which literally means refreshment, cooling (GIUNTELLA 1985). In the Christian ceremony the purpose of a funeral banquet is to benefit the soul of the departed on the anniversary of death, a painful event celebrated as dies natalis of the soul to eternal life.
Who was buried in this sepulchre? The question is destined to remain unanswered and only an inscription found nearby could be a clue in this sense. The text says that the owners purchased the sepulchre close to the one of the bishop Cheperion’s (ORSI 1895, 507-508; RIZZONE 2011, PP. 55-58), of whom the scanty written sources never make mention. In any case this constructed sarcophagus represent one of the most eminent burials of the catacomb. In the same gallery one cannot ignore the finding of an inscription in many fragments, which is a singular phenomenon of religious contamination.
This inscription, with a Christological monogram at the top, records Nassiana “Christian, who competed with Penelope in moral virtue”. This inscription can only be compared with three examples in the Roman sepulchral carmina (BÜCHELER 1897). The circular support of this inscription found close to the “tomb of the Saint” has been regarded as a
mensa(table) for refrigerium rite (GIUNTELLA 1985, 47), whose circular form would derive in any case from an evident reuse of a marble disk of classical craftsmanship, with a laurel wreath and berries sculpted on one side (ORSI 1895, n. 234). A wall inscription painted on the extrados of an arcosolium in the third northern gallery of the catacomb could be connected with Nassiana’s text. Two lines of the inscription say that “Sossa outdid (other women) in conjugal love; as for handiworks, without being taught by anyone, Athena herself had taught her how to do marvellous things” (FERRUA 1940, n. 3). Both inscriptions unequivocally make use of female figures from the classical world, interpreted as a model also for Christian women either for a deep-rooted usage or making up for the absence of assimilable figures in the Church. This allows one to notice in a 4th century large community cemetery, such as San Giovanni, some phenomena of religious contamination in a period in which Christian epigraphic praxis consolidated itself by then (SGARLATA 1999, 484).
The rotunda derives its name from the deceased Antiochia recorded in the sarcophagus, set inside the ring of tombs, made of blocks and bearing an engraved and rubricated inscription. Whom did this private space belong to? To answer the question, one can rely on a suggestion provided once more by epigraphic testimonies. The gravestones found – on which they put as a rule the deceased’s name, lifespan, date of death and deposition – according to the first excavators’ reports (CARINI 1873, 23-45), attest that the rotunda of Antiochia could have housed women only and this would give credence to the idea that the mausoleum had been used for a female monastic community. The hypothesis needs to be verified, but it is very seductive.
It is evident that the catacomb of San Giovanni was originally a community cemetery, planned for only one type of burial: the arcosolium with multiple depositions, which does not require great care for decoration, only transennae. In the topographical and architectural development of the catacomb it appears clear that creating the rotundas breaks up the common burials series, destined to a socially homogeneous Christian community. These changes to the original plan – of the creation of subterranean mausolea both to the north and south of the cemetery – spring from the necessity to create appropriate spaces for the members of the Chruch and above all of the Empire, bringing into question the initial egalitarian choice of the arcosolium burials (GRIESHEIMER 1989, 767). In the terminal part of the pious Giovanni’s gallery a monumental sarcophagus, once more hewn out of the rock, can be perceived. Close to the closing wall of the same gallery an inscription was found: it bears the notification of both the consuls of the year 349 (ORSI 1909, 756, 8; AGNELLO 1953, 89). As a rule this date marks the end of the dig works in this sector of the catacomb (FERRUA 1952, 75- 76). A arcosolium extrados, in the eastern region, shows traces of a palinsesto, an overlap of paintings and epigraphs than obstruct the identification of the first person buried in this sepulchre, even if recently Vittorio Rizzone proposed the interpretation Philadelpheia1. In one of the fossa tombs at the terminal stretch of the main gallery the inscription of Euterpe (IG XIV, 112) has been found, reused to cover the bottom of the tomb, which traditionally marks the end of the dig works of the catacomb (FERRUA 1952, 75-76; AGNELLO 1958, 79; GRIESHEIMER 1989, 781). Beyond the deceased’s biometric data, recorded as “companion of the Muses”, the epigraph in the last three lines mentions the consuls’ iteration, which allow us to date back to the consulate of the Emperor Constantius, consul for the tenth time, and Julian- Caesar, consul for the third time. So Euterpe died on November 27th 360 A.D. at the age of 22 years and 3 months (GUARDUCCI 1978, 524-526).
In the southern region,, more than in other ones, that you can note the transformations that have profoundly undermined its community spirit, which had originally inspired the creation of the catacomb. In the first rotunda of the southern region, is a private space, which was given the name of Marina due to an inscription scratched upon the extrados of the arcosolium,on the right of the entrance to the short gallery of the bishop Siracosio. The arcosolium seems to be enframed by a painted prothyrum, as the still visible column and capital attest, confirming the generalized use of architectural elements in this catacomb, already seen in the rotunda of Antiochia. According to the interpretation of the text, Marina could have been the wife of the patriciuset magister militum Sabinianus, sent by the Emperor Honorius to Spain at the time of barbarian invasions presumably between 409 and 423 (FERRUA 1989, 21-22, 40).
This date would agree with the numerous testimonies given by the inscriptions that have been dated thanks to the notification of the two consuls in office in the year of their death.These testimonies are ascribable to the years of the Emperors Arcadius and Honorius in the first quarter of the 5th century and seem to suggest a link between a few burials of the southern region and the aristocrats’ diaspora from Rome after Alaricus’ advance in 410, who took refuge in Sicily and Africa as they did in other provinces of the Empire (SIRAGO 1989, 175; VERA 1988, 157-158). It is necessary to have a look at the slab of the presumed arcosolium of the bishop Siracosio, recorded in one inscription found in an adjacent fossa tomb, which says that the deceased intentionally purchased the sepulchre close to the one of the bishop just mentioned (IG XIV, 123: Ἐνθάδε κῖτε Πολυχ / ρόνιος καὶ Σεραπία. / Ἠγόρασεν τῷτότ / ε καιρῷ / Πολυχρονίου /
 1 RIZZONE 2012, pp. 45-48.
αἱ Σεραπία ἐπὶ τῷκυρί / ῳ μου ἐπισκόπῳ Συρα / κοσίῳ). It is just a hypothesis seeing in the arcosolium with the engraved slab, still in situ, a noble burial for a member of the ecclesiastical hierarchy, for whom, in the absence of other data, according to a letter of pope Gelasius I, an episcopate between 492 and 496 has been proposed (NARCISO 1952, 223, CARLETTI 2008). One can clearly distinguish a Christogram with the apocalyptic letters alpha and omega and two ships in the shape of fish, regarded as making reference to the sacrament of the Eucharist, this is also suggested by the diskettes next to the fishes’ mouths, assimilable to loaves of bread (SGARLATA 2013).
After reading the text of the inscription upon the sarcophagus lid (CIL X, 7123: Ic Adelfia c(larissima) f(emina) / posita conpar / Balericomitis), one can become aware that it refers only to the woman, not to both husband and wife: here lies Adelfia, clarissima femina, wife of the count Valerius. Who were therefore Valerius and Adelfia? The unsolved enigmas of the catacomb of San Giovanni deal with their names and most of all their identification. In reality it is possible to distinguish more than two phases of intervention, which have preceded the creation of the hole for the sarcophagus. At the beginning the internal space of the large niche was scenographically arranged in a terrace pattern with a forceful ascensional effect, not exempt from comparisons (NESTORI 1973, 14, plate VII).The intact Latin inscription of Sporus, dated to 356, which sealed a forma in the main gallery among the ones that join the rotunda of Marina to the one of Adelfia (CAVALLARI 1872, 24; AGNELLO 1953, 90), would attest that the exploitation of the soil in the area gravitating to the rotunda of Adelfia was already underway after the first half of the 4th century, so confirming the evident anteriority of the six graves cut in the floor to the phase of monumentalization. Both the internal and the external organizations of the large niche (nicchione) would correspond to the first and second phases of intervention respectively, according to the current reconstruction. The problems related to the reconstruction of the third phase seem to be more complicated: the monumentalization phase started with the interment of the sarcophagus and concluded with the acquisition of an aspect comparable to the privileged burials of Roman crypts.
The monument seems to have a less wild internal dynamics of development; the new data (archaeological, historical, epigraphical) (SGARLATA 1996, 101-108) allow the scholars to widen the syncopated weave of temporal sequences, to which the analysis of the monument has been pinned by the constant reference to Valerius Proculus’ chronology. A new chronology and a different identification of the comes Valerius can be proposed, considering: 1) the evident reuse of the sarcophagus; 2) the topographical development of the catacomb in the area where the sarcophagus has been discovered; 3) the type of monumental intervention, which followed the Roman counterparts, datable to the second half of the 4th century (FIOCCHI NICOLAI 1997, 132-134). If Valerius were given a different identity, one could postpone him from the age of Constantine to that of Augustine (SGARLATA 1998, 15-51), on the basis of several accounts about the friendship between the saint from Hippo and a comes Valerius, whose physiognomy, is somewhat vague. This friendship, confirmed by the epistles (AUG., Ep. 200, 206, 207; Retr. II, 79 and 88) and the dedication, in 419, of the treatise De nuptiis et concupiscentia, is fed on the fight against Pelagianism, which in eastern Sicily had found a fertile soil. The presence of Pelagians in Sicily is attested for certain, according to Hilarius’ account under the pontificate of Innocent (PIETRI 2000, 429-452) and Honoreficentia’s letter, where a clarissima based at Syracuse is mentioned. On the island the spread of Pelagian movement looks like a direct consequence of the 410 sacking of Rome and the diaspora of the Roman nobility, of which Pelagius and Celestius were spiritual leaders; the short period they both spent in Sicily was not painless for Christian orthodoxy (PIETRI 2001, 436-437) and particularly in Syracuse, as the cemeteries in the area overhanging the Greek theater, intended to serve the communities of the so-called heretics throughout the 5th century (AGNELLO 1990).All the data gathered seems to lead to a different identification of Valerius – who, even if he was not Augustine’s correspondent, would be sought in the list of Valerii reported in the sources of the first quarter of the 5th century (PLRE II, 1143-44) – and a later chronology of the large niche monumental transformation compared to the one traditionally accepted.
Certainly it is not the case that in this catacomb the traces of the members of ecclesiastical hierarchy are so rare: where are the martyrs? Why is the evidence of bishops, presbyters and deacons so scant? Why is this Christianized elite – to which the will to betray the communal matrix of the original project for a new particularistic conception of funerary space can be attributed – so copious?
To answer the questions will take time, which will assure a greater credibility to what so far seems just a series of suggestions, most of all fed by the re-reading of epigraphic testimonies: the episodic character of the references to burials of bishops, presbyters, deacons and noting that the most significant percentage of evidences pertains to the members of the Church, buried in Syracuse away from their own countries and recorded in wall inscriptions written in Latin – Auxentius Hispanus episcopus and Superianus clerecus de Aquileia (FERRUA 1940, 1 and 6) – as a demonstration that the official language is used by foreigners, who were high clients (the remainder of inscriptions is in Greek), these and many other clues lead to thinking of a less incisive control of the Church in the 5th century than one can commonly believe.
In the rotunda of sarcophagi is surprising the concentration of monumental burials, so that one can think members of the Church commissioned the works here too. What clues are the that could support the hypothesis that the clients belonged to a religious community? To tell the truth, they are scant and among them the epigraph of the blessed virgins Fotina and Filomena deserves to be recorded: the former lived 80 years, the latter 84 (IG XIV, 187; FERRUA 1989, 180). The sole dated inscription found within this chamber, which is related to the name of Eucarpio, bearing the notification of both the consuls between 339 and 360 (FERRUA 1983, 3), cannot be used for dating; the discovery data, insisting on the fact that the gravestone was found overturned in a forma tomb (AGNELLO 1960, 30-31), suggest that the gravestone was reused.
The cubiculum of Eusebio has a structure different than other private spaces of the southern region of the catacomb. The name Eusebio derives, once more, from an inscription found on a three level monumental tomb, in the shape of exedra, visible to the left of the cubiculum (IG XIV, 111). The position and monumentality of the grand arcosolium, the formula “of blessed memory”, the paleographical characters, the identity of pope Eusebius month of death (exiled to Sicily by Maxentius, Eusebius died on 17 August 309 or 310) have suggested to Isidoro Carini that this arcosolium could be the pontiff’s temporary sepulchre, whose bones were transferred to Rome and deposited in the catacomb of Callisto. Despite the scholar’s efforts to sustain this theory (CARINI 1873, 134), the physiognomic contours of Eusebius recorded in the inscription remain hazy.
The cubiculum of Eusebio also deserves to be recorded for another testimony, which has a special value for Syracusans: an inscription found in a forma tomb, which testifies the worship of Saint Lucy, the patron of Syracuse..No decoration, no distinctive signs characterize the modest burial of Euskia (ORSI 1895, 299-308), one of the many graves cut in the soil of the cubiculum (GUARDUCCI 1978, 526-528, fig. 164).

\begin{otherlanguage*}{greek}
Εὐσκία ἡἄμενπτος , ζήσα<σα> \\
 χρηστῶς καὶσεμνῶςἔτη \\ 
 πλῖο<ν>ἔλαττον κε' , ἀνε-\\ 
 παύσετοτῇἑορτῇτῆςκυ-\\
ρίας μουΛουκίας, εἰςἢν \\
 οὔκἐστινἐνκώμειον \\
 εἰπεῖν ,Xρηστειανή , πισ-\\ 
 τή ,τέλιοςοὖσα , εὐχα- \\
 ριστοῦσα τῷεἰδίῳἀν- \\
 δρὶ πολλὰςεὐχαρισ-\\
τίας αⳀω εὐομε[ίλητος].
\end{otherlanguage*}

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{Euskia.JPG}
\caption{ The Euskia’s inscription.}
\label{fig:3}
\end{figure}

Euskia, the irreproachable one, who lived her life in good and pure ways for more or less twenty-five years, died on the feast of our lady Lucia, for whom no praise is adequate. She was Christian, faithful (and) perfect, well pleasing to her husband, endued with much grace, affable.
This is the text of the most important Christian epigraph in Syracuse (AGNELLO 1953, 20).The formulae present the typical elogium, the retrospective data of the life of the deceased and the Christological monogram flanked by the apocalyptic letters, which are common elements in the inscriptions of the catacombs; it is important since Euskia had been privileged to die on the same day that was sacred to Lucia, patron of Syracusans, a martyr during Diocletian persecutions on the 13th of December 304. The sole doubt of this reconstruction lurks in the term kyria, which is referred to Lucia in the epigraph; does one have to interpret it as synonym of haghia (saint), which would assure the official character of the worship, or a simple honorary title? Whatever answer one can give, the importance of Euskia’s inscription survives intact, because it testifies if not already the sanctity of Lucia, the local devotion and worship of which the woman was subject in the 5th century in Syracuse.
So this is the first attestation of the cult of Saint Lucia, which confirms the historicity of the Martyrologium Hieronymianum’s account on popular devotion to the Saint, which came through the celebration of a feast from the outset. All of the other evidence refers to successive periods. It is the Greek martyrion dated to the end of the 5th century, whose reliability has been debated for a long time and to this day never evidently established (MILAZZO – RIZZO NERVO 1988, 95-135). The inscription, ascribable to the beginning of the 5th century, so would precede the questioned passio and would confirm the antiquity of the cult of Saint Lucia, whose bones were presumably kept in the homonymous catacomb in Syracusee, before George Maniakes, in 1039, transported them to Costantinople. Along with
the inscription of Iulia Florentina from Catania (RIZZA 1964, 608-610), the epigraph of Euskia is the most ancient Sicilian document that one could relate to the experience of martyrdom.
. 1.4 EPIGRAPHIC POPULATION: FORMULARIES AND IDENTITY-MAKING CHARACTERISTIES
The 800 inscriptions found in the San Giovanni catacomb give us an idea about the epigraphic approach and the society of that time. One single epigraph can mention more than one person so the number of remembered deceased is superior than the number of inscriptions; starting from this number it’s possible to launch a demographic investigation to determine the expectation of life.
In Syracuse all the social classes were affected with inexorable mortality; high percentage among young people (under 20) and extremely low percentage among old people (over 60). The demographic data follow standard hypothesis: male mortality between 15 and 34 years and female mortality between 20 and 24 years (reproductive period). The life expectancy for both genders was around 29 years.

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{Fotina e Filomena.JPG}
\caption{The Fotina and Filomena’s inscription.}
\label{fig:4}
\end{figure}

We focuse on the decesead elogistic formulary that help to increase the study about social aspects. In our research there are few examples of nuptial terminology, all referred to female deceased: compar, coniunx, σύμβιος, νυμφη2. It marks a common characteristic: the priority in honoring the wives more than the husbands. Linked to this is the importance given to the sexual integrity of the dead person, in fact the concept of virginity in the nuptial context is particularly marked trhough the adjective virginius-a o παρθενικός-ή3. The word παρθένος appears seven times and always referred to women. Women who chose monastic life leave behind them all the duties of nuptial life so probably the had a much longer life expectancy than married women. One impressive example is the one of ‘’Fotina e Filomena’’, consecrate
2V. cat. n. XX-XXI
3 See also SGARLATA 1991, pp. 129-130.

virgins (80 and 84 years old, as stated above). Such a long lifespan, for the centuries under examination, finds a justification only in considering that both the women chose a monastic-type life, avoiding the slow attrition due to diverse factors: precocious age weddings, consecutive childbirths since their early adolescence, abortive practices by makeshift means and, lastly, even after the dangerous age of 25/30, the overwork that household management and hygienic-sanitary conditions involve. The biometric data recorded in the inscriptions confirmed in Syracusan sample as well (SGARLATA 1991), testify that life expectancy at birth, considering the high rates of infant mortality, was not over 30 years on average, both for men and women; this datum is not surprising, if one bears in mind that the average span life was about 45 years still in the first half of the nineteenth century.
____________________________________
CARINI 1887, p. 493; CIL X, 7168, 2-4; ORSI 1893, 5, 22, 44; FUHRER 1897, 6, 7; FERRUA 1983, 21/29, 38, 39, 43, 45, 48, 58, 60, 69; IGCVO 127, 655, 778, 946,950, 1036, 1069 e 1366.
Cat. n. XX (ORSI 1896, 214). See. cat. n. XX-XXI
1.5 CONCLUSIONS
The new approach to the catacomb of San Giovanni brings many things into question, such as the fideistic attitude of those, who have studied this monument and have looked at the epigraphic evidence for chronological purpose. Rereading Orsi’s excavations data unequivocally shows how itinerant are the dated inscriptions within the cemetery – with the exception of three, whose discovery data attest their permanence in the original position (ORSI 1896, 43-50; 352-353; AGNELLO 1953, 90, 97) – which advises the scholars against using them to seal chronologically the diverse sectors. It would be profitable to draw a map of reuse, which is certainly the most striking phenomenon detected thus far in Orsi’s accounts, more than continuing to date the works in the galleries on the basis of dated inscriptions found in their terminal part. But, even underestimating this phenomenon, the inscriptions datable to the years around 350 and the ones, more considerable numerically, that bear the notifications of both the consuls between the end of the 4th century and the first half of the 5th century, have been seen both in the northern region and the southern one, as well as the main gallery (SGARLATA 1996, 109, n. 62), which excludes drawing conclusions on the internal development of the catacomb. A datum, however, is worth considering: epigraphic evidence and intense exploitation of funerary space attest the vitality of the area that gravitates around the three southern rotundas after the end of the great works of excavations. What appears episodic in the other sectors of the catacomb – for example the so- called sepulchre of the Saint – becomes constant in the southern region, where diverse types of intervention on pre- existent structures and high percentage of dated inscriptions demonstrate, still in the first half of the 5th century, a special concentration of interest. During the work on the editing of ICI, the first certain fact we acquire studying the inscriptions is that the dated ones are almost always reemployed and itinerant within the catacomb, with the exceptions just mentioned. We have to make also another consideration that will allow us, once ended the work of cataloguing and copying the epigraphs, to deal with the language question in a deeper and more articulated way. Also the linguistic choice recorded in epigraphic evidence is worth considering: the inscriptions in Greek surpass by far, with a rate of about 90%, the ones in Latin. About motivations of language choice not all the researchers agree, discerning between the language example provided by the christian graveyards inscriptions and the one expressed by the whole citizenry. It is worth to mention what Kalle Korhonen affirms: “moreover, it must be painted out that even if 90% of ca. 1.100 inscriptions from the catacombs of Syracuse are in Greek, we are not allowed to conclude that 90% of the population of Syracuse was Greek from the 3rd to the 5th. It is likely that the non-Christian parts of the population, which was notable until the 4th century, were not buried in catacombs and their epithaphs have mostly perished” (KORHONEN 2011, 124- 125).
Assuming this, one could state that in Sicily religious conversion is not a linguistic conversion, proving wrong the theory according to which in urban centers Christianization would bring along an early diffusion of Latin, whereas the pagus, keeping the use of Greek, would keep its distance from Christianity, at least until the early 5th century, during which signs of both linguistic and religious conversion would become a little more evident (MANGANARO 1993, 545).Epigraphic documentation records, for Syracuse and its territory, as for Catania, a marked preponderance of the use of Greek still in the 5th century; so in urban centers Christianization does not bring an early diffusion of Latin, which was prerogative of high and foreign clients, as only the use of Greek in the pagus does not prove the extraneousness to the process of diffusion of the new creed still at the beginning of the 5th century. The idea according
to which “several signs of religious dissidence, in the 5th and 6th centuries, can be recognized only in peripheral rural areas, where the level of ecclesiastical control was slack. In these centuries conventional superstitions were slowly relegated to periphery” (CRACCO RUGGINI 1998) needs to be dampened.
The cemetery of San Giovanni has subtly given back, even if in a more hidden way, several testimonies of ideological commixture, which do not distinguish it from other communal cemeteries in Syracuse, where phenomena of “religious dissidence” are more readable (SGARLATA 2003). The catacomb was created in different cultural and religious contexts (after the Peace of the Church). The cemeteries distribution (both of private and community law) and topography of funerary monuments in the suburban area of Acradina, between the 3rd and 5th centuries, reflect well a diversified situation within a few hundred meters radius. To understand that, one needs to take account of the relationship between paganism and Christianity, orthodoxy and heterodoxy (most of all for the 5th century) (MACMULLEN 1997), which is not only a Sicilian problem, even if it is strongly sensed in the island (GRECO 1999, 59)

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}

\end{document}

### 2016-01-13 keynote/beard/EAGLE_2016_Beard.tex
keynote speaker pages
copied from files, compiled

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Putting ancient inscriptions in the limelight}
\author[]{Mary Beard\corref{first}}

\begin{document}
\maketitle
\begin{abstract}
This lecture will reflect on public engagement with ancient inscriptions: what is it about inscriptions that interests a wider audience, and -- just as important -- what puts them off? I shall be drawing on a series of BBC documentaries, ``Meet the Romans'' (``Ti presento i Romani'') which used Roman epitaphs and other inscribed texts as a way into the life of ordinary Romans, but will reflect more widely on how academics and museum professionals can (and already do) make inscriptions come alive for the public.
\end{abstract}

Mary Beard is one of Britain’s best-known Classicists – a distinguished Professor of Classics at the University of Cambridge and Fellow of Newnham College. She has written numerous books on the Ancient World, including the 2008 Wolfson Prize-winner, Pompeii: The Life of a Roman Town, The Roman Triumph, Classical Art from Greece to Rome, as well as popular books on the Parthenon and Colosseum. In addition she has presented a highly-acclaimed TV series, Meet the Roman as well as documentaries about Pompeii and the Emperor Caligula. Mary’s interests range from the social and cultural life of the Ancient World to Victorian understanding of antiquity. Mary is also Classics editor of the Times Literary Supplement and writes an engaging, and thought-provoking, blog, A Don’s Life. Mary’s academic achievements were acknowledged, in 2010, by the British Academy which elected her as a Fellow and she was made an OBE in the New Year’s Honours List 2013 for services to Classical scholarship. Her latest book is on the subject of Roman Laughter and her forthcoming history of Rome, SPQR, will be published in Autumn 2015.

\end{document}

### 2016-01-13 keynote/eck/EAGLE_2016_Eck.tex
keynote speaker pages
copied from files, compiled

--- 2016-01-13 keynote/beard/EAGLE_2016_Beard.tex
@@
 %%%%%%%%
-\title{Putting ancient inscriptions in the limelight}
-\author[]{Mary Beard\corref{first}}
-
+\title{Keynote Speech}
+\author[]{Werner Eck\corref{first}}
 \begin{document}
 \maketitle
-\begin{abstract}
-This lecture will reflect on public engagement with ancient inscriptions: what is it about inscriptions that interests a wider audience, and -- just as important -- what puts them off?
-I shall be drawing on a series of BBC documentaries, ``Meet the Romans'' (``Ti presento i Romani'') which used Roman epitaphs and other inscribed texts as a way into the life of ordinary Romans, but will reflect more widely on how academics and museum professionals can (and already do) make inscriptions come alive for the public.
-\end{abstract}
+%\begin{abstract}
+%\end{abstract}
 
-Mary Beard is one of Britain’s best-known Classicists – a distinguished Professor of Classics at the University of Cambridge and Fellow of Newnham College.
-She has written numerous books on the Ancient World, including the 2008 Wolfson Prize-winner, Pompeii: The Life of a Roman Town, The Roman Triumph, Classical Art from Greece to Rome, as well as popular books on the Parthenon and Colosseum.
-In addition she has presented a highly-acclaimed TV series, Meet the Roman as well as documentaries about Pompeii and the Emperor Caligula.
-Mary’s interests range from the social and cultural life of the Ancient World to Victorian understanding of antiquity.
-Mary is also Classics editor of the Times Literary Supplement and writes an engaging, and thought-provoking, blog, A Don’s Life.
-Mary’s academic achievements were acknowledged, in 2010, by the British Academy which elected her as a Fellow and she was made an OBE in the New Year’s Honours List 2013 for services to Classical scholarship.
-Her latest book is on the subject of Roman Laughter and her forthcoming history of Rome, SPQR, will be published in Autumn 2015.
+Werner Eck is an internationally renowned historian of the Roman Empire and of Roman epigraphy.
+From February 1979 until his retirement in February 2007, he worked as Professor of Ancient History at the University of Cologne.
+From 1997 till 2002 he was president of the Association of Greek and Latin Epigraphy.
+At the Accademy of Sciences in Berlin he is responsible for the CIL and the PIR.
+With his colleagues of the Berlin-Brandenburgische Akademie der Wissenschaften he organised the 13th International Congress of Greek and Latin Epigraphy.
+In Cologne he is one of the editors of the Zeitschrift für Papyrology und Epigraphik.
+Together with German and Israeli colleagues he is editing the multilingual Corpus Inscriptionum Iudaea/Palaestinae.
+He has published about very many topics of the Imperial period, about the administration of Rome, Italy and the provinces, the social and military history, prosopography and the history of early Christianity.
+For his bibliography see \url{http://histinst.phil-fak.uni-koeln.de/index.php?id=309}
 

### 2016-01-13 keynote/roueché/EAGLE_2016_Roueche.tex
keynote speaker pages
copied from files, compiled

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Where Eagles Dare}
\author[]{Charlotte Roueché\corref{first}}
\begin{document}
\maketitle
\begin{abstract}
The publication of inscriptions in digital form has been evolving since the 1980s, and online publication since the 1990s. It is useful to see where EAGLE is positioned in this evolution – a huge step forward, but not the end of the road. What can we learn from our progress so far? What have the main challenges been? And what does this suggest about the next stages?

\end{abstract}

Charlotte Roueché studied Classics at Cambridge, and Byzantine Studies in Paris and Washington; as a pupil of Joyce Reynolds, she became involved in publishing Greek inscriptions, with a particular focus on material from Late Antiquity. In 1984 she was appointed to King’s College London, where an important centre of Digital Humanities started to develop in the 1990s; this encouraged her to explore the possibilities for publishing inscribed texts online. In 1999 Tom Elliott published the first proposal for EpiDoc; working with him and Gabriel Bodard, Charlotte was able to use EpiDoc to publish several online corpora: \url{http://insaph.kcl.ac.uk/ala2004, http://insaph.kcl.ac.uk/iaph2007}, and \url{http://irt.kcl.ac.uk/irt2009}. At present she is preparing a digital corpus of inscriptions of Roman Cyrenaica, and developing plans for a shared portal, Inscriptions of Libya. She has also recently published an eleventh century Byzantine text online, in a project exploring the uses of RDF to analyse intertextuality: \url{http://ancientwisdoms.ac.uk}.

\end{document}

### 2016-01-13 source/32/EAGLE_2016_submission_32.tex
paper 32 convertito
bibliografia aggiunta in biblio, riferimenti nel testo aggiornati

@@
 \begin{abstract}
-This paper refers to the general track “Epigraphic edition on paper and on line”.
-Some researchers who has been working on the editing of the Inscriptiones Christianae Italiae, published from the University of Bari, contributed also to the EDR project, that collect on line the epigraphic documents of Roman Christian Period.
+This paper refers to the general track ``Epigraphic edition on paper and on line''.
+Some researchers who has been working on the editing of the \emph{Inscriptiones Christianae Italiae}, published from the University of Bari, contributed also to the EDR project, that collect on line the epigraphic documents of Roman Christian Period.
 I propose a preview of the work currently in progress, with a specific reference to the inscriptions that provide us the chronological and topographic data to study the cemetery as well as formularies linked to the society structure and to the identity-making characteristics.
@@
 
-\begin{figure}[hbp]
-\centering
- \includegraphics[width=\columnwidth]{MISSING IMAGE!}
-\caption{The main gallery (so called decumanus) of Catacombs of San Giovanni}
-\label{fig:1}
-\end{figure}
+%\begin{figure}[hbp]
+%\centering
+ %\includegraphics[width=\columnwidth]{MISSING IMAGE!}
+%\caption{The main gallery (so called \emph{decumanus}) of Catacombs of San Giovanni}
+%\label{fig:1}
+%\end{figure}
 
-We are far from a complete and thorough research about palaeochristian epigraphs in Sicily.
+\noindent We are far from a complete and thorough research about palaeochristian epigraphs in Sicily.
 This catalogue represents the first step of a systematic study focused on the inscriptions discovered in the christian cemeteries of Siracusa.
@@
 The renovated interest about funerary epigraphy generated new studies about new criteria of dating.
-The inscriptions dated by the indications of the ‘’consular couple’’ represent an exceptions in the issues about chronological datings (FERRUA 1947, 227; 1989, nn.
-191-208): two were discovered during the excavation conducted by Paolo Orsi (1985) and to them is associated the latin epigraph ‘’Sporus’’ (356) .
+The inscriptions dated by the indications of the ``consular couple'' represent an exceptions in the issues about chronological datings \citep[227]{FERRUA1947} \citep[nn.
+191-208]{FERRUA1989}: two were discovered during the excavation conducted by Paolo Orsi (1985) and to them is associated the latin epigraph ``Sporus'' (356) .
+
 The bigger one is entitled to San Giovanni and gave back almost 800 inscriptions, now located for the most part in the Archeological and Regional Museum ‘’Paolo Orsi’’ of Siracusa.; alongside will be placed the 40 headstones recovered in the adjacent hypogea of Villa Landolina, that report compatible datings and formularies.
 
-1.2 HISTORICAL ASPECTS OF THE CATACOMB OF S. GIOVANNI AT SYRACUSE
-The systematic studies of collective cemeteries in Syracuse started many years ago and, beyond the archaeological surveys, the late antique funerary settlements of Syracuse proposed multiple research cues and paths, as the historical- religious, economic and social nature.
+\section{HISTORICAL ASPECTS OF THE CATACOMB OF S. GIOVANNI AT SYRACUSE}
+
+\noindent The systematic studies of collective cemeteries in Syracuse started many years ago and, beyond the archaeological surveys, the late antique funerary settlements of Syracuse proposed multiple research cues and paths, as the historical- religious, economic and social nature.
 Suburban cemeteries, fanned out from the area of Fusco, in the quarter of Neapolis, to the Santa Lucia area, in the southern part of Acradina, this indicates unequivocally what the perimeter of the city must have already been in the early and mid Roman Empire.
-The History of the area, which was going to hold the catacombs (San Giovanni, Vigna Cassia and Santa Lucia), spanned the centuries between the classical Greek and late antique ages, gradually giving evidence of quarries (Latomie), water supply systems to the city, characterized by cisterns and aqueducts (COLLIN BOUFFIER 1987, 682), handcraft workshops from the beginning of the 4th/3rd century BCE and burials datable to the early and mid Roman Empire.
+The History of the area, which was going to hold the catacombs (San Giovanni, Vigna Cassia and Santa Lucia), spanned the centuries between the classical Greek and late antique ages, gradually giving evidence of quarries (Latomie), water supply systems to the city, characterized by cisterns and aqueducts \citep[682]{COLLINBOUFFIER1987}), handcraft workshops from the beginning of the 4th/3rd century BCE and burials datable to the early and mid Roman Empire.
 The analysis of the funerary system certifies one hand the dependence on the Roman model, and other the debt in respect of local traditions.
+
 Several interest will be given to structural aspect of the catacomb of S.
-Giovanni, practice of funeral rituals, ethnic and cultural fruition’s characters, transformation in the use, transformation in the way of using spaces for graves, to complete a general point of view about the phenomena of continuity and innovation as to previous sepulchral arrangements and, in the analysed periods, the facies belonging to the different settling, variegated in the committees’ideological and religious themes, in choosing monumental types (like the rotundae) and decorations, in self-representative aspects, in burial uses.
+Giovanni, practice of funeral rituals, ethnic and cultural fruition’s characters, transformation in the use, transformation in the way of using spaces for graves, to complete a general point of view about the phenomena of continuity and innovation as to previous sepulchral arrangements and, in the analysed periods, the facies belonging to the different settling, variegated in the committees’ ideological and religious themes, in choosing monumental types (like the rotundae) and decorations, in self-representative aspects, in burial uses.
 In this perspective we will give particular importance to the study of funerary epigraphy aimed at the writing of ICI Siracusa and the overall interpretation of the monument.
 Between 1893 and 1909 the archaeologist Paolo Orsi carried out a series of campaigns in the catacomb of San Giovanni.
-Detailed reports of those campaigns are recorded at various times in Notizie degli Scavi (see the BIBLIOGRAPHY), which constitute an indefeasible starting point.
-The first incisive studies on Christian subterranean Sicily pertain to Joseph Führer, who dedicates many pages to the catacomb of San Giovanni, whose study is also epigraphic, is based on the previous literature, especially on Paolo Orsi’s discoveries (FÜHRER 1897, 13-39; FÜHRER- SCHULTZE 1907, 22-26; see also ORSI 1893, 276, n.
-2 and SCHULTZE 1882, 60-61, fig.
-5).
-In the study of Syracuse cemeteries Orsi (ORSI 1900a, 189) was the first to see the mixed nature of the burials and their materials (mostly gravestones), openly attesting a kind of pagan-Christian and orthodox-heterodox symbioses (see also AGNELLO 1957a, 242-243).
-Both Antonio Ferrua and Santi Luigi Agnello within the space of a few years tried to solve the problems relating to the sarcophagus of Adelfia and the more general problem, strictly correlated with the ones just mentioned, of the cemetery genesis and development (FERRUA 1952 and AGNELLO 1956).
+Detailed reports of those campaigns are recorded at various times in Notizie degli Scavi, which constitute an indefeasible starting point.
+The first incisive studies on Christian subterranean Sicily pertain to Joseph Führer, who dedicates many pages to the catacomb of San Giovanni, whose study is also epigraphic, is based on the previous literature, especially on Paolo Orsi’s discoveries \citep[13-39]{FUHRER1897}; \citep[22-26]{FUHRERSCHULTZE1907}; see also \citep[276, n.
+2]{ORSI1893}
+%and \citep[60-61, fig. 5]{SCHULTZE1882} MISSING IN BIBLIOGRAPHY
+.
+In the study of Syracuse cemeteries Orsi \citep[189]{ORSI1900} was the first to see the mixed nature of the burials and their materials (mostly gravestones), openly attesting a kind of pagan-Christian and orthodox-heterodox symbioses %\citep[242-243]{AGNELLO1957a} MISSING IN BIBLIOGRAPHY
+.
+Both Antonio Ferrua and Santi Luigi Agnello within the space of a few years tried to solve the problems relating to the sarcophagus of Adelfia and the more general problem, strictly correlated with the ones just mentioned, of the cemetery genesis and development \citep{FERRUA1952} and \citep{AGNELLO1956}.
 It is necessary to record that numerous hypotheses on the genesis and development of the various sectors of the cemetery are ascribable to the account given by inscriptions.
 Syracuse, thanks to its prolific underground cemeteries, has the larger Christian epigraphic heritage after Rome, kept in the storage of the Soprintendenza di Siracusa.
-1.3 TOPOGRAPHIC AND EPIGRAFIC DATA TO STUDY THE CEMETERY
-Walking across the main gallery, one can retrace the stages of Paolo Orsi’s interventions, being at the entrance of the Catacomb second northern gallery, before the so-called “tomb of the Saint”.
-This arcosolium is considered a privileged burial on the basis of the following reasons: most of all because of its position, but also because it is a mensatype burial closed by a sole slab (ORSI 1893, 292-294).
-The signs of an ancient rite are easily traceable on the slab, a rite that preceded the coming of Christianity and persisted for centuries up to the present day: the rite of refrigerium, which literally means refreshment, cooling (GIUNTELLA 1985).
+
+\section{TOPOGRAPHIC AND EPIGRAFIC DATA TO STUDY THE CEMETERY}
+
+\noindent Walking across the main gallery, one can retrace the stages of Paolo Orsi’s interventions, being at the entrance of the Catacomb second northern gallery, before the so-called ``tomb of the Saint''.
+This arcosolium is considered a privileged burial on the basis of the following reasons: most of all because of its position, but also because it is a mensatype burial closed by a sole slab \citep[292-294]{ORSI1893}.
+The signs of an ancient rite are easily traceable on the slab, a rite that preceded the coming of Christianity and persisted for centuries up to the present day: the rite of refrigerium, which literally means refreshment, cooling \citep{GIUNTELLA1985}.
 In the Christian ceremony the purpose of a funeral banquet is to benefit the soul of the departed on the anniversary of death, a painful event celebrated as dies natalis of the soul to eternal life.
+
 Who was buried in this sepulchre?
 The question is destined to remain unanswered and only an inscription found nearby could be a clue in this sense.
-The text says that the owners purchased the sepulchre close to the one of the bishop Cheperion’s (ORSI 1895, 507-508; RIZZONE 2011, PP.
-55-58), of whom the scanty written sources never make mention.
+The text says that the owners purchased the sepulchre close to the one of the bishop Cheperion’s \citep[507-508]{ORSI1895}; \citep[55-58]{RIZZONE2011}, of whom the scanty written sources never make mention.
 In any case this constructed sarcophagus represent one of the most eminent burials of the catacomb.
 In the same gallery one cannot ignore the finding of an inscription in many fragments, which is a singular phenomenon of religious contamination.
-This inscription, with a Christological monogram at the top, records Nassiana “Christian, who competed with Penelope in moral virtue”.
-This inscription can only be compared with three examples in the Roman sepulchral carmina (BÜCHELER 1897).
-The circular support of this inscription found close to the “tomb of the Saint” has been regarded as a
-mensa(table) for refrigerium rite (GIUNTELLA 1985, 47), whose circular form would derive in any case from an evident reuse of a marble disk of classical craftsmanship, with a laurel wreath and berries sculpted on one side (ORSI 1895, n.
-234).
+
+This inscription, with a Christological monogram at the top, records Nassiana ``Christian, who competed with Penelope in moral virtue''.
+This inscription can only be compared with three examples in the Roman sepulchral carmina \citep{CLE}.
+The circular support of this inscription found close to the ``tomb of the Saint'' has been regarded as a mensa (table) for refrigerium rite \citep[47]{GIUNTELLA1985}, whose circular form would derive in any case from an evident reuse of a marble disk of classical craftsmanship, with a laurel wreath and berries sculpted on one side \citep[n.
+234]{ORSI1895}.
 A wall inscription painted on the extrados of an arcosolium in the third northern gallery of the catacomb could be connected with Nassiana’s text.
-Two lines of the inscription say that “Sossa outdid (other women) in conjugal love; as for handiworks, without being taught by anyone, Athena herself had taught her how to do marvellous things” (FERRUA 1940, n.
-3).
+Two lines of the inscription say that ``Sossa outdid (other women) in conjugal love; as for handiworks, without being taught by anyone, Athena herself had taught her how to do marvellous things'' \citep[n.
+3]{FERRUA1940}.
 Both inscriptions unequivocally make use of female figures from the classical world, interpreted as a model also for Christian women either for a deep-rooted usage or making up for the absence of assimilable figures in the Church.
-This allows one to notice in a 4th century large community cemetery, such as San Giovanni, some phenomena of religious contamination in a period in which Christian epigraphic praxis consolidated itself by then (SGARLATA 1999, 484).
+This allows one to notice in a 4th century large community cemetery, such as San Giovanni, some phenomena of religious contamination in a period in which Christian epigraphic praxis consolidated itself by then \citep[484]{SGARLATA1999}.
+
 The rotunda derives its name from the deceased Antiochia recorded in the sarcophagus, set inside the ring of tombs, made of blocks and bearing an engraved and rubricated inscription.
@@
 To answer the question, one can rely on a suggestion provided once more by epigraphic testimonies.
-The gravestones found – on which they put as a rule the deceased’s name, lifespan, date of death and deposition – according to the first excavators’ reports (CARINI 1873, 23-45), attest that the rotunda of Antiochia could have housed women only and this would give credence to the idea that the mausoleum had been used for a female monastic community.
+The gravestones found – on which they put as a rule the deceased’s name, lifespan, date of death and deposition – according to the first excavators’ reports \citep[23-45]{CARINI1873}, attest that the rotunda of Antiochia could have housed women only and this would give credence to the idea that the mausoleum had been used for a female monastic community.
 The hypothesis needs to be verified, but it is very seductive.
+
 It is evident that the catacomb of San Giovanni was originally a community cemetery, planned for only one type of burial: the arcosolium with multiple depositions, which does not require great care for decoration, only transennae.
 In the topographical and architectural development of the catacomb it appears clear that creating the rotundas breaks up the common burials series, destined to a socially homogeneous Christian community.
-These changes to the original plan – of the creation of subterranean mausolea both to the north and south of the cemetery – spring from the necessity to create appropriate spaces for the members of the Chruch and above all of the Empire, bringing into question the initial egalitarian choice of the arcosolium burials (GRIESHEIMER 1989, 767).
+These changes to the original plan – of the creation of subterranean mausolea both to the north and south of the cemetery – spring from the necessity to create appropriate spaces for the members of the Chruch and above all of the Empire, bringing into question the initial egalitarian choice of the arcosolium burials \citep[767]{GRIESHEIMER1989}.
 In the terminal part of the pious Giovanni’s gallery a monumental sarcophagus, once more hewn out of the rock, can be perceived.
-Close to the closing wall of the same gallery an inscription was found: it bears the notification of both the consuls of the year 349 (ORSI 1909, 756, 8; AGNELLO 1953, 89).
-As a rule this date marks the end of the dig works in this sector of the catacomb (FERRUA 1952, 75- 76).
-A arcosolium extrados, in the eastern region, shows traces of a palinsesto, an overlap of paintings and epigraphs than obstruct the identification of the first person buried in this sepulchre, even if recently Vittorio Rizzone proposed the interpretation Philadelpheia1.
-In one of the fossa tombs at the terminal stretch of the main gallery the inscription of Euterpe (IG XIV, 112) has been found, reused to cover the bottom of the tomb, which traditionally marks the end of the dig works of the catacomb (FERRUA 1952, 75-76; AGNELLO 1958, 79; GRIESHEIMER 1989, 781).
-Beyond the deceased’s biometric data, recorded as “companion of the Muses”, the epigraph in the last three lines mentions the consuls’ iteration, which allow us to date back to the consulate of the Emperor Constantius, consul for the tenth time, and Julian- Caesar, consul for the third time.
+Close to the closing wall of the same gallery an inscription was found: it bears the notification of both the consuls of the year 349 %\citep[756, 8]{ORSI1909} MISSING IN BIBLIOGRAPHY
+; \citep[89]{AGNELLO1953}.
+As a rule this date marks the end of the dig works in this sector of the catacomb \citep[75-76]{FERRUA1952}.
+A arcosolium extrados, in the eastern region, shows traces of a palinsesto, an overlap of paintings and epigraphs than obstruct the identification of the first person buried in this sepulchre, even if recently Vittorio Rizzone proposed the interpretation Philadelpheia.\footnote{\citet[45-48]{RIZZONE2012}.} In one of the fossa tombs at the terminal stretch of the main gallery the inscription of Euterpe (IG XIV, 112) has been found, reused to cover the bottom of the tomb, which traditionally marks the end of the dig works of the catacomb \citep[75-76]{FERRUA1952}; \citep[79]{AGNELLO1958}; \citep[781]{GRIESHEIMER1989}.
+Beyond the deceased’s biometric data, recorded as ``companion of the Muses'', the epigraph in the last three lines mentions the consuls’ iteration, which allow us to date back to the consulate of the Emperor Constantius, consul for the tenth time, and Julian- Caesar, consul for the third time.
 So Euterpe died on November 27th 360 A.D.
-at the age of 22 years and 3 months (GUARDUCCI 1978, 524-526).
+at the age of 22 years and 3 months \citep[524-526]{GUARDUCCI1978}.
+
 In the southern region,, more than in other ones, that you can note the transformations that have profoundly undermined its community spirit, which had originally inspired the creation of the catacomb.
@@
 The arcosolium seems to be enframed by a painted prothyrum, as the still visible column and capital attest, confirming the generalized use of architectural elements in this catacomb, already seen in the rotunda of Antiochia.
-According to the interpretation of the text, Marina could have been the wife of the patriciuset magister militum Sabinianus, sent by the Emperor Honorius to Spain at the time of barbarian invasions presumably between 409 and 423 (FERRUA 1989, 21-22, 40).
-This date would agree with the numerous testimonies given by the inscriptions that have been dated thanks to the notification of the two consuls in office in the year of their death.These testimonies are ascribable to the years of the Emperors Arcadius and Honorius in the first quarter of the 5th century and seem to suggest a link between a few burials of the southern region and the aristocrats’ diaspora from Rome after Alaricus’ advance in 410, who took refuge in Sicily and Africa as they did in other provinces of the Empire (SIRAGO 1989, 175; VERA 1988, 157-158).
-It is necessary to have a look at the slab of the presumed arcosolium of the bishop Siracosio, recorded in one inscription found in an adjacent fossa tomb, which says that the deceased intentionally purchased the sepulchre close to the one of the bishop just mentioned (IG XIV, 123: Ἐνθάδε κῖτε Πολυχ / ρόνιος καὶ Σεραπία.
-/ Ἠγόρασεν τῷτότ / ε καιρῷ / Πολυχρονίου /
- 1 RIZZONE 2012, pp. 45-48.
-αἱ Σεραπία ἐπὶ τῷκυρί / ῳ μου ἐπισκόπῳ Συρα / κοσίῳ).
-It is just a hypothesis seeing in the arcosolium with the engraved slab, still in situ, a noble burial for a member of the ecclesiastical hierarchy, for whom, in the absence of other data, according to a letter of pope Gelasius I, an episcopate between 492 and 496 has been proposed (NARCISO 1952, 223, CARLETTI 2008).
-One can clearly distinguish a Christogram with the apocalyptic letters alpha and omega and two ships in the shape of fish, regarded as making reference to the sacrament of the Eucharist, this is also suggested by the diskettes next to the fishes’ mouths, assimilable to loaves of bread (SGARLATA 2013).
+According to the interpretation of the text, Marina could have been the wife of the patriciuset magister militum Sabinianus, sent by the Emperor Honorius to Spain at the time of barbarian invasions presumably between 409 and 423 \citep[21-22, 40]{FERRUA1989}.
+
+This date would agree with the numerous testimonies given by the inscriptions that have been dated thanks to the notification of the two consuls in office in the year of their death.These testimonies are ascribable to the years of the Emperors Arcadius and Honorius in the first quarter of the 5th century and seem to suggest a link between a few burials of the southern region and the aristocrats’ diaspora from Rome after Alaricus’ advance in 410, who took refuge in Sicily and Africa as they did in other provinces of the Empire \citep[175]{SIRAGO1989}; %\citep[157-158]{VERA1988} MISSING IN BIBLIOGRAPHY
+.
+It is necessary to have a look at the slab of the presumed arcosolium of the bishop Siracosio, recorded in one inscription found in an adjacent fossa tomb, which says that the deceased intentionally purchased the sepulchre close to the one of the bishop just mentioned (IG XIV, 123: \begin{otherlanguage*}{greek}Ἐνθάδε κῖτε Πολυχ/ρόνιος καὶ Σεραπία.
+/ Ἠγόρασεν τῷ τότ/ε καιρῷ / Πολυχρονίου /αἱ Σεραπία ἐπὶ τῷ κυρί/ῳ μου ἐπισκόπῳ Συρα/κοσίῳ\end{otherlanguage*}).
+It is just a hypothesis seeing in the arcosolium with the engraved slab, still in situ, a noble burial for a member of the ecclesiastical hierarchy, for whom, in the absence of other data, according to a letter of pope Gelasius I, an episcopate between 492 and 496 has been proposed \citep[223]{NARCISO1952} \citep{CARLETTI2008}.
+One can clearly distinguish a Christogram with the apocalyptic letters alpha and omega and two ships in the shape of fish, regarded as making reference to the sacrament of the Eucharist, this is also suggested by the diskettes next to the fishes’ mouths, assimilable to loaves of bread \citep{SGARLATA2013}.
+
 After reading the text of the inscription upon the sarcophagus lid (CIL X, 7123: Ic Adelfia c(larissima) f(emina) / posita conpar / Balericomitis), one can become aware that it refers only to the woman, not to both husband and wife: here lies Adelfia, clarissima femina, wife of the count Valerius.
@@
 In reality it is possible to distinguish more than two phases of intervention, which have preceded the creation of the hole for the sarcophagus.
-At the beginning the internal space of the large niche was scenographically arranged in a terrace pattern with a forceful ascensional effect, not exempt from comparisons (NESTORI 1973, 14, plate VII).The intact Latin inscription of Sporus, dated to 356, which sealed a forma in the main gallery among the ones that join the rotunda of Marina to the one of Adelfia (CAVALLARI 1872, 24; AGNELLO 1953, 90), would attest that the exploitation of the soil in the area gravitating to the rotunda of Adelfia was already underway after the first half of the 4th century, so confirming the evident anteriority of the six graves cut in the floor to the phase of monumentalization.
+At the beginning the internal space of the large niche was scenographically arranged in a terrace pattern with a forceful ascensional effect, not exempt from comparisons \citep[14, plate VII]{NESTORI1973}.The intact Latin inscription of Sporus, dated to 356, which sealed a forma in the main gallery among the ones that join the rotunda of Marina to the one of Adelfia \citep[24]{CAVALLARI 1872}; \citep[90]{AGNELLO1953}, would attest that the exploitation of the soil in the area gravitating to the rotunda of Adelfia was already underway after the first half of the 4th century, so confirming the evident anteriority of the six graves cut in the floor to the phase of monumentalization.
 Both the internal and the external organizations of the large niche (nicchione) would correspond to the first and second phases of intervention respectively, according to the current reconstruction.
 The problems related to the reconstruction of the third phase seem to be more complicated: the monumentalization phase started with the interment of the sarcophagus and concluded with the acquisition of an aspect comparable to the privileged burials of Roman crypts.
-The monument seems to have a less wild internal dynamics of development; the new data (archaeological, historical, epigraphical) (SGARLATA 1996, 101-108) allow the scholars to widen the syncopated weave of temporal sequences, to which the analysis of the monument has been pinned by the constant reference to Valerius Proculus’ chronology.
-A new chronology and a different identification of the comes Valerius can be proposed, considering: 1) the evident reuse of the sarcophagus; 2) the topographical development of the catacomb in the area where the sarcophagus has been discovered; 3) the type of monumental intervention, which followed the Roman counterparts, datable to the second half of the 4th century (FIOCCHI NICOLAI 1997, 132-134).
-If Valerius were given a different identity, one could postpone him from the age of Constantine to that of Augustine (SGARLATA 1998, 15-51), on the basis of several accounts about the friendship between the saint from Hippo and a comes Valerius, whose physiognomy, is somewhat vague.
+
+The monument seems to have a less wild internal dynamics of development; the new data (archaeological, historical, epigraphical) \citep[101-108]{SGARLATA1996} allow the scholars to widen the syncopated weave of temporal sequences, to which the analysis of the monument has been pinned by the constant reference to Valerius Proculus’ chronology.
+A new chronology and a different identification of the comes Valerius can be proposed, considering: 1) the evident reuse of the sarcophagus; 2) the topographical development of the catacomb in the area where the sarcophagus has been discovered; 3) the type of monumental intervention, which followed the Roman counterparts, datable to the second half of the 4th century \citep[132-134]{FIOCCHINICOLAI 1997}.
+If Valerius were given a different identity, one could postpone him from the age of Constantine to that of Augustine \citep[15-51]{SGARLATA1998}, on the basis of several accounts about the friendship between the saint from Hippo and a comes Valerius, whose physiognomy, is somewhat vague.
 This friendship, confirmed by the epistles (AUG., Ep.
@@
 II, 79 and 88) and the dedication, in 419, of the treatise De nuptiis et concupiscentia, is fed on the fight against Pelagianism, which in eastern Sicily had found a fertile soil.
-The presence of Pelagians in Sicily is attested for certain, according to Hilarius’ account under the pontificate of Innocent (PIETRI 2000, 429-452) and Honoreficentia’s letter, where a clarissima based at Syracuse is mentioned.
-On the island the spread of Pelagian movement looks like a direct consequence of the 410 sacking of Rome and the diaspora of the Roman nobility, of which Pelagius and Celestius were spiritual leaders; the short period they both spent in Sicily was not painless for Christian orthodoxy (PIETRI 2001, 436-437) and particularly in Syracuse, as the cemeteries in the area overhanging the Greek theater, intended to serve the communities of the so-called heretics throughout the 5th century (AGNELLO 1990).All the data gathered seems to lead to a different identification of Valerius – who, even if he was not Augustine’s correspondent, would be sought in the list of Valerii reported in the sources of the first quarter of the 5th century (PLRE II, 1143-44) – and a later chronology of the large niche monumental transformation compared to the one traditionally accepted.
+The presence of Pelagians in Sicily is attested for certain, according to Hilarius’ account under the pontificate of Innocent \citep[429-452]{PIETRI2000} and Honoreficentia’s letter, where a clarissima based at Syracuse is mentioned.
+On the island the spread of Pelagian movement looks like a direct consequence of the 410 sacking of Rome and the diaspora of the Roman nobility, of which Pelagius and Celestius were spiritual leaders; the short period they both spent in Sicily was not painless for Christian orthodoxy %\citep[436-437]{PIETRI2001} MISSING IN BIBLIOGRAPHY
+ and particularly in Syracuse, as the cemeteries in the area overhanging the Greek theater, intended to serve the communities of the so-called heretics throughout the 5th century \citep{AGNELLO1990}.All the data gathered seems to lead to a different identification of Valerius – who, even if he was not Augustine’s correspondent, would be sought in the list of Valerii reported in the sources of the first quarter of the 5th century (PLRE II, 1143-44) – and a later chronology of the large niche monumental transformation compared to the one traditionally accepted.
+
 Certainly it is not the case that in this catacomb the traces of the members of ecclesiastical hierarchy are so rare: where are the martyrs?
@@
 Why is this Christianized elite – to which the will to betray the communal matrix of the original project for a new particularistic conception of funerary space can be attributed – so copious?
-To answer the questions will take time, which will assure a greater credibility to what so far seems just a series of suggestions, most of all fed by the re-reading of epigraphic testimonies: the episodic character of the references to burials of bishops, presbyters, deacons and noting that the most significant percentage of evidences pertains to the members of the Church, buried in Syracuse away from their own countries and recorded in wall inscriptions written in Latin – Auxentius Hispanus episcopus and Superianus clerecus de Aquileia (FERRUA 1940, 1 and 6) – as a demonstration that the official language is used by foreigners, who were high clients (the remainder of inscriptions is in Greek), these and many other clues lead to thinking of a less incisive control of the Church in the 5th century than one can commonly believe.
+
+To answer the questions will take time, which will assure a greater credibility to what so far seems just a series of suggestions, most of all fed by the re-reading of epigraphic testimonies: the episodic character of the references to burials of bishops, presbyters, deacons and noting that the most significant percentage of evidences pertains to the members of the Church, buried in Syracuse away from their own countries and recorded in wall inscriptions written in Latin – Auxentius Hispanus episcopus and Superianus clerecus de Aquileia \citep[1 and 6]{FERRUA1940} – as a demonstration that the official language is used by foreigners, who were high clients (the remainder of inscriptions is in Greek), these and many other clues lead to thinking of a less incisive control of the Church in the 5th century than one can commonly believe.
+
 In the rotunda of sarcophagi is surprising the concentration of monumental burials, so that one can think members of the Church commissioned the works here too.
 What clues are the that could support the hypothesis that the clients belonged to a religious community?
-To tell the truth, they are scant and among them the epigraph of the blessed virgins Fotina and Filomena deserves to be recorded: the former lived 80 years, the latter 84 (IG XIV, 187; FERRUA 1989, 180).
-The sole dated inscription found within this chamber, which is related to the name of Eucarpio, bearing the notification of both the consuls between 339 and 360 (FERRUA 1983, 3), cannot be used for dating; the discovery data, insisting on the fact that the gravestone was found overturned in a forma tomb (AGNELLO 1960, 30-31), suggest that the gravestone was reused.
+To tell the truth, they are scant and among them the epigraph of the blessed virgins Fotina and Filomena deserves to be recorded: the former lived 80 years, the latter 84 (IG XIV, 187); \citep[180]{FERRUA 1989}.
+The sole dated inscription found within this chamber, which is related to the name of Eucarpio, bearing the notification of both the consuls between 339 and 360 (\citep[3]{FERRUA1983}), cannot be used for dating; the discovery data, insisting on the fact that the gravestone was found overturned in a forma tomb \citep[30-31]{AGNELLO1960}, suggest that the gravestone was reused.
+
 The cubiculum of Eusebio has a structure different than other private spaces of the southern region of the catacomb.
 The name Eusebio derives, once more, from an inscription found on a three level monumental tomb, in the shape of exedra, visible to the left of the cubiculum (IG XIV, 111).
-The position and monumentality of the grand arcosolium, the formula “of blessed memory”, the paleographical characters, the identity of pope Eusebius month of death (exiled to Sicily by Maxentius, Eusebius died on 17 August 309 or 310) have suggested to Isidoro Carini that this arcosolium could be the pontiff’s temporary sepulchre, whose bones were transferred to Rome and deposited in the catacomb of Callisto.
-Despite the scholar’s efforts to sustain this theory (CARINI 1873, 134), the physiognomic contours of Eusebius recorded in the inscription remain hazy.
-The cubiculum of Eusebio also deserves to be recorded for another testimony, which has a special value for Syracusans: an inscription found in a forma tomb, which testifies the worship of Saint Lucy, the patron of Syracuse..No decoration, no distinctive signs characterize the modest burial of Euskia (ORSI 1895, 299-308), one of the many graves cut in the soil of the cubiculum (GUARDUCCI 1978, 526-528, fig.
-164).
+The position and monumentality of the grand arcosolium, the formula ``of blessed memory'', the paleographical characters, the identity of pope Eusebius month of death (exiled to Sicily by Maxentius, Eusebius died on 17 August 309 or 310) have suggested to Isidoro Carini that this arcosolium could be the pontiff’s temporary sepulchre, whose bones were transferred to Rome and deposited in the catacomb of Callisto.
+Despite the scholar’s efforts to sustain this theory \citep[134]{CARINI1873}, the physiognomic contours of Eusebius recorded in the inscription remain hazy.
+The cubiculum of Eusebio also deserves to be recorded for another testimony, which has a special value for Syracusans: an inscription found in a forma tomb, which testifies the worship of Saint Lucy, the patron of Syracuse..No decoration, no distinctive signs characterize the modest burial of Euskia \citep[299-308]{ORSI1895}, one of the many graves cut in the soil of the cubiculum \citep[526-528, fig.
+164]{GUARDUCCI1978}.
 
+\begin{quotation}
 \begin{otherlanguage*}{greek}
-Εὐσκία ἡἄμενπτος , ζήσα<σα> \\
- χρηστῶς καὶσεμνῶςἔτη \\ 
- πλῖο<ν>ἔλαττον κε' , ἀνε-\\ 
- παύσετοτῇἑορτῇτῆςκυ-\\
-ρίας μουΛουκίας, εἰςἢν \\
- οὔκἐστινἐνκώμειον \\
- εἰπεῖν ,Xρηστειανή , πισ-\\ 
- τή ,τέλιοςοὖσα , εὐχα- \\
- ριστοῦσα τῷεἰδίῳἀν- \\
- δρὶ πολλὰςεὐχαρισ-\\
-τίας αⳀω εὐομε[ίλητος].
+\noindent Εὐσκία ἡ ἄμενπτος, ζήσα<σα> \\
+%Εὐσκία ἡ ἄμενπτος , ζήσα<σα> \\
+χρηστῶς καὶ σεμνῶς ἔτη \\
+% χρηστῶς καὶ σεμνῶς ἔτη \\ 
+πλῖο<ν> ἔλαττον κε᾽, ἀνε-\\
+% πλῖο<ν>ἔλαττον κε' , ἀνε-\\
+παύσετο τῇ ἑορτῇ τῆς κυ-\\ 
+% παύσετο τῇ ἑορτῇ τῆς κυ-\\
+ρίας μοθ Λουκίας, εἰς ἢν \\
+%ρίας μου Λουκίας, εἰς ἢν \\
+οὔκ ἐστιν ἐν κώμειον \\
+% οὔκ ἐστιν ἐν κώμειον \\
+εἰπεῖν, Χρηστειανή, πισ-\\
+% εἰπεῖν ,Xρηστειανή , πισ-\\
+τή, τέλιος οὖσα, εὐχα-\\ 
+% τή ,τέλιοςοὖσα , εὐχα- \\
+ριστοῦσα τῷ εἰδίῳ ἀν-\\
+% ριστοῦσα τῷ εἰδίῳ ἀν- \\
+δρὶ πολλὰς εὐχαρισ-\\
+% δρὶ πολλὰς εὐχαρισ-\\
+τίασ α((chirho))ω εὐομε[ίλητος]
+%τίας αⳀω εὐομε[ίλητος].
 \end{otherlanguage*}
+\end{quotation}
 
@@
 She was Christian, faithful (and) perfect, well pleasing to her husband, endued with much grace, affable.
-This is the text of the most important Christian epigraph in Syracuse (AGNELLO 1953, 20).The formulae present the typical elogium, the retrospective data of the life of the deceased and the Christological monogram flanked by the apocalyptic letters, which are common elements in the inscriptions of the catacombs; it is important since Euskia had been privileged to die on the same day that was sacred to Lucia, patron of Syracusans, a martyr during Diocletian persecutions on the 13th of December 304.
+This is the text of the most important Christian epigraph in Syracuse \citep[20]{AGNELLO1953}.The formulae present the typical elogium, the retrospective data of the life of the deceased and the Christological monogram flanked by the apocalyptic letters, which are common elements in the inscriptions of the catacombs; it is important since Euskia had been privileged to die on the same day that was sacred to Lucia, patron of Syracusans, a martyr during Diocletian persecutions on the 13th of December 304.
 The sole doubt of this reconstruction lurks in the term kyria, which is referred to Lucia in the epigraph; does one have to interpret it as synonym of haghia (saint), which would assure the official character of the worship, or a simple honorary title?
@@
 All of the other evidence refers to successive periods.
-It is the Greek martyrion dated to the end of the 5th century, whose reliability has been debated for a long time and to this day never evidently established (MILAZZO – RIZZO NERVO 1988, 95-135).
+It is the Greek martyrion dated to the end of the 5th century, whose reliability has been debated for a long time and to this day never evidently established \citep[95-135]{MILAZZORIZZONERVO1988}.
 The inscription, ascribable to the beginning of the 5th century, so would precede the questioned passio and would confirm the antiquity of the cult of Saint Lucia, whose bones were presumably kept in the homonymous catacomb in Syracusee, before George Maniakes, in 1039, transported them to Costantinople.
 Along with
-the inscription of Iulia Florentina from Catania (RIZZA 1964, 608-610), the epigraph of Euskia is the most ancient Sicilian document that one could relate to the experience of martyrdom.
-. 1.4 EPIGRAPHIC POPULATION: FORMULARIES AND IDENTITY-MAKING CHARACTERISTIES
+the inscription of Iulia Florentina from Catania \citep[608-610]{RIZZA1964}, the epigraph of Euskia is the most ancient Sicilian document that one could relate to the experience of martyrdom.
+
+\section{EPIGRAPHIC POPULATION: FORMULARIES AND IDENTITY-MAKING CHARACTERISTIES}
+
 The 800 inscriptions found in the San Giovanni catacomb give us an idea about the epigraphic approach and the society of that time.
@@
 \centering
- \includegraphics[width=\columnwidth]{Fotina e Filomena.JPG}
+ \includegraphics[width=\columnwidth]{FotinaeFilomena.JPG}
 \caption{The Fotina and Filomena’s inscription.}
@@
 
-We focuse on the decesead elogistic formulary that help to increase the study about social aspects.
-In our research there are few examples of nuptial terminology, all referred to female deceased: compar, coniunx, σύμβιος, νυμφη2.
-It marks a common characteristic: the priority in honoring the wives more than the husbands.
-Linked to this is the importance given to the sexual integrity of the dead person, in fact the concept of virginity in the nuptial context is particularly marked trhough the adjective virginius-a o παρθενικός-ή3.
-The word παρθένος appears seven times and always referred to women.
+We focus on the decesead elogistic formulary that help to increase the study about social aspects.
+In our research there are few examples of nuptial terminology, all referred to female deceased: compar, coniunx,
+\begin{otherlanguage*}{greek}σύμβιος, νυμφη\end{otherlanguage*}
+.\footnote{V.
+cat.
+n.
+XX-XXI} It marks a common characteristic: the priority in honoring the wives more than the husbands.
+Linked to this is the importance given to the sexual integrity of the dead person, in fact the concept of virginity in the nuptial context is particularly marked trhough the adjective virginius-a o \begin{otherlanguage*}{greek}παρθενικός-ή.\end{otherlanguage*}
+\footnote{See also \citet[129-130]{SGARLATA1991} \citet[493]{CARINI 1887}; CIL X, 7168, 2-4; \citet[5, 22, 44]{ORSI1893}; \citet[6, 7]{FUHRER1897}; \citep[21/29, 38, 39, 43, 45, 48, 58, 60, 69]{FERRUA1983}; IGCVO 127, 655, 778, 946,950, 1036, 1069 e 1366.
+Cat.
+n.
+XX \citep[214]{ORSI1896}.
+See.
+cat.
+n.
+XX-XXI} The word \begin{otherlanguage*}{greek}παρθένος\end{otherlanguage*}
+appears seven times and always referred to women.
 Women who chose monastic life leave behind them all the duties of nuptial life so probably the had a much longer life expectancy than married women.
-One impressive example is the one of ‘’Fotina e Filomena’’, consecrate
-2V. cat. n. XX-XXI
-3 See also SGARLATA 1991, pp. 129-130.
+One impressive example is the one of ``Fotina e Filomena'', consecrate virgins (80 and 84 years old, as stated above).
+Such a long lifespan, for the centuries under examination, finds a justification only in considering that both the women chose a monastic-type life, avoiding the slow attrition due to diverse factors: precocious age weddings, consecutive childbirths since their early adolescence, abortive practices by makeshift means and, lastly, even after the dangerous age of 25/30, the overwork that household management and hygienic-sanitary conditions involve.
+The biometric data recorded in the inscriptions confirmed in Syracusan sample as well \citep{SGARLATA1991}, testify that life expectancy at birth, considering the high rates of infant mortality, was not over 30 years on average, both for men and women; this datum is not surprising, if one bears in mind that the average span life was about 45 years still in the first half of the nineteenth century.
 
-virgins (80 and 84 years old, as stated above).
-Such a long lifespan, for the centuries under examination, finds a justification only in considering that both the women chose a monastic-type life, avoiding the slow attrition due to diverse factors: precocious age weddings, consecutive childbirths since their early adolescence, abortive practices by makeshift means and, lastly, even after the dangerous age of 25/30, the overwork that household management and hygienic-sanitary conditions involve.
-The biometric data recorded in the inscriptions confirmed in Syracusan sample as well (SGARLATA 1991), testify that life expectancy at birth, considering the high rates of infant mortality, was not over 30 years on average, both for men and women; this datum is not surprising, if one bears in mind that the average span life was about 45 years still in the first half of the nineteenth century.
-____________________________________
-CARINI 1887, p. 493; CIL X, 7168, 2-4; ORSI 1893, 5, 22, 44; FUHRER 1897, 6, 7; FERRUA 1983, 21/29, 38, 39, 43, 45, 48, 58, 60, 69; IGCVO 127, 655, 778, 946,950, 1036, 1069 e 1366.
-Cat. n. XX (ORSI 1896, 214). See. cat. n. XX-XXI
-1.5 CONCLUSIONS
+\section{CONCLUSIONS}
+
 The new approach to the catacomb of San Giovanni brings many things into question, such as the fideistic attitude of those, who have studied this monument and have looked at the epigraphic evidence for chronological purpose.
-Rereading Orsi’s excavations data unequivocally shows how itinerant are the dated inscriptions within the cemetery – with the exception of three, whose discovery data attest their permanence in the original position (ORSI 1896, 43-50; 352-353; AGNELLO 1953, 90, 97) – which advises the scholars against using them to seal chronologically the diverse sectors.
+Rereading Orsi’s excavations data unequivocally shows how itinerant are the dated inscriptions within the cemetery – with the exception of three, whose discovery data attest their permanence in the original position \citep[43-50, 352-353]{ORSI1896}; \citep[90, 97]{AGNELLO1953} – which advises the scholars against using them to seal chronologically the diverse sectors.
 It would be profitable to draw a map of reuse, which is certainly the most striking phenomenon detected thus far in Orsi’s accounts, more than continuing to date the works in the galleries on the basis of dated inscriptions found in their terminal part.
-But, even underestimating this phenomenon, the inscriptions datable to the years around 350 and the ones, more considerable numerically, that bear the notifications of both the consuls between the end of the 4th century and the first half of the 5th century, have been seen both in the northern region and the southern one, as well as the main gallery (SGARLATA 1996, 109, n.
-62), which excludes drawing conclusions on the internal development of the catacomb.
+But, even underestimating this phenomenon, the inscriptions datable to the years around 350 and the ones, more considerable numerically, that bear the notifications of both the consuls between the end of the 4th century and the first half of the 5th century, have been seen both in the northern region and the southern one, as well as the main gallery \citep[109, n.
+62]{SGARLATA1996}, which excludes drawing conclusions on the internal development of the catacomb.
 A datum, however, is worth considering: epigraphic evidence and intense exploitation of funerary space attest the vitality of the area that gravitates around the three southern rotundas after the end of the great works of excavations.
@@
 We have to make also another consideration that will allow us, once ended the work of cataloguing and copying the epigraphs, to deal with the language question in a deeper and more articulated way.
-Also the linguistic choice recorded in epigraphic evidence is worth considering: the inscriptions in Greek surpass by far, with a rate of about 90%, the ones in Latin.
+Also the linguistic choice recorded in epigraphic evidence is worth considering: the inscriptions in Greek surpass by far, with a rate of about 90\%, the ones in Latin.
 About motivations of language choice not all the researchers agree, discerning between the language example provided by the christian graveyards inscriptions and the one expressed by the whole citizenry.
-It is worth to mention what Kalle Korhonen affirms: “moreover, it must be painted out that even if 90% of ca.
-1.100 inscriptions from the catacombs of Syracuse are in Greek, we are not allowed to conclude that 90% of the population of Syracuse was Greek from the 3rd to the 5th.
-It is likely that the non-Christian parts of the population, which was notable until the 4th century, were not buried in catacombs and their epithaphs have mostly perished” (KORHONEN 2011, 124- 125).
-Assuming this, one could state that in Sicily religious conversion is not a linguistic conversion, proving wrong the theory according to which in urban centers Christianization would bring along an early diffusion of Latin, whereas the pagus, keeping the use of Greek, would keep its distance from Christianity, at least until the early 5th century, during which signs of both linguistic and religious conversion would become a little more evident (MANGANARO 1993, 545).Epigraphic documentation records, for Syracuse and its territory, as for Catania, a marked preponderance of the use of Greek still in the 5th century; so in urban centers Christianization does not bring an early diffusion of Latin, which was prerogative of high and foreign clients, as only the use of Greek in the pagus does not prove the extraneousness to the process of diffusion of the new creed still at the beginning of the 5th century.
+It is worth to mention what Kalle Korhonen affirms: ``moreover, it must be painted out that even if 90\% of ca.
+1.100 inscriptions from the catacombs of Syracuse are in Greek, we are not allowed to conclude that 90\% of the population of Syracuse was Greek from the 3rd to the 5th.
+It is likely that the non-Christian parts of the population, which was notable until the 4th century, were not buried in catacombs and their epithaphs have mostly perished'' \citep[124- 125]{KORHONEN2011}.
+Assuming this, one could state that in Sicily religious conversion is not a linguistic conversion, proving wrong the theory according to which in urban centers Christianization would bring along an early diffusion of Latin, whereas the pagus, keeping the use of Greek, would keep its distance from Christianity, at least until the early 5th century, during which signs of both linguistic and religious conversion would become a little more evident \citep[545]{MANGANARO1993}.Epigraphic documentation records, for Syracuse and its territory, as for Catania, a marked preponderance of the use of Greek still in the 5th century; so in urban centers Christianization does not bring an early diffusion of Latin, which was prerogative of high and foreign clients, as only the use of Greek in the pagus does not prove the extraneousness to the process of diffusion of the new creed still at the beginning of the 5th century.
 The idea according
-to which “several signs of religious dissidence, in the 5th and 6th centuries, can be recognized only in peripheral rural areas, where the level of ecclesiastical control was slack.
-In these centuries conventional superstitions were slowly relegated to periphery” (CRACCO RUGGINI 1998) needs to be dampened.
-The cemetery of San Giovanni has subtly given back, even if in a more hidden way, several testimonies of ideological commixture, which do not distinguish it from other communal cemeteries in Syracuse, where phenomena of “religious dissidence” are more readable (SGARLATA 2003).
+to which ``several signs of religious dissidence, in the 5th and 6th centuries, can be recognized only in peripheral rural areas, where the level of ecclesiastical control was slack.
+In these centuries conventional superstitions were slowly relegated to periphery'' \citep{CRACCORUGGINI1998} needs to be dampened.
+The cemetery of San Giovanni has subtly given back, even if in a more hidden way, several testimonies of ideological commixture, which do not distinguish it from other communal cemeteries in Syracuse, where phenomena of ``religious dissidence'' are more readable \citep{SGARLATA2003}.
 The catacomb was created in different cultural and religious contexts (after the Peace of the Church).
 The cemeteries distribution (both of private and community law) and topography of funerary monuments in the suburban area of Acradina, between the 3rd and 5th centuries, reflect well a diversified situation within a few hundred meters radius.
-To understand that, one needs to take account of the relationship between paganism and Christianity, orthodoxy and heterodoxy (most of all for the 5th century) (MACMULLEN 1997), which is not only a Sicilian problem, even if it is strongly sensed in the island (GRECO 1999, 59)
+To understand that, one needs to take account of the relationship between paganism and Christianity, orthodoxy and heterodoxy (most of all for the 5th century) \citep{MACMULLEN1997}, which is not only a Sicilian problem, even if it is strongly sensed in the island \citep[59]{GRECO1999}.
 

### 2016-01-13 Posters.tex
finished posters and committee, compiled the all lot
I have Moved all the compiled articles in the pdfs folder and compiled
with all bits and pieces

% !TEX encoding = UTF-8 Unicode
% -*- coding: UTF-8; -*-

\chapter{List of Posters presented}\label{posterslist}

\begin{sloppypar}
All posters can be seen in the digital poster exhibition on the Eagle website.
www.eagle-network.eu/about/events/eagle2016/digital-poster-exhibition/

\end{sloppypar}

\begin{table}
{\small
\addtolength{\tabcolsep}{-0.8mm}

\begin{tabular*}{\textwidth}{p{4.5cm}|p{6cm}}
%    \rowcolor{black!70}
%    \multicolumn{1}{c}{\textcolor{white}{\textbf{Author}}} & 
%    \multicolumn{1}{c}{\textcolor{white}{\textbf{Title}}} 
%    \tabularnewline
%    
%    \endfirsthead
%    \multicolumn{2}{c}
%    {{\bfseries \tablename\ \thetable{} -- continued from previous page}} \tabularnewline
%    \rowcolor{black!70}
%    \multicolumn{1}{c}{\textcolor{white}{\textbf{Author}}} &
%    \multicolumn{1}{c}{\textcolor{white}{\textbf{Title}}}
%    \tabularnewline
%    
%    \endhead
% \rowcolor{black!70}
%    \multicolumn{2}{r}{{ \textcolor{white}{-- continued on next page}}} \tabularnewline 
%
%    \endfoot
%    
%    \endlastfoot
%   
\toprule Jan-Mathieu Carbon, Saskia Peels and Vinciane Pirenne-Delforge & THE COLLECTION OF GREEK RITUAL NORMS PROJECT (CGRN) \\\hline
Stefano Caneva and Guido D'Ippolito & Pasquino: Talking Statues and Inscriptions in Rome\\\hline
Mat Carbon, Mario Paganini and Stella Skaltsa & The Copenhagen Associations Project (CAP)\\\hline
Emilia Mataix Ferrandiz & Bridging the gap Linking commercial epigraphy and Roman law\\\hline
Barbara Mazzei and Agnese Pergola & The PCAS Photographic Digital Archive: Fruition of Epigraphic Images\\\hline
Barbara De Nicolò & “Talking Stones”: from Stone to History. A Latin Epigraphy workshop for Primary School\\\hline
Sara Chiarini and Kirsten Jahn & TheDeMa (Thesaurus Defixionum Magdeburgensis) – A digital companion to ancient curse tablets\\\hline
Daniele Silvi and Fabio Ciotti & Thematic annotation and ontologies: application to digital epigraphy\\\hline
Loredana Francesca Tedeschi & Integration of geomatic techniques for the analysis and geolocation of graphic forms of the Filocalian letter in the Elogia Martyrum by Pope Damasus(366-384)\\\hline
Fulvia Mainardis, Claudio Zaccaria, Andrea Favretto and Giovanni Mauro & A WebGIS for the knowledge of the epigraphic locations in the Friuli Venezia Giulia region (North-East of Italy)\\\hline
Petra Janouchova & INSCRIPTIONS TO GO: Mobile platform for offline epigraphic data collection\\\hline
Tamar Kalkhitashvili and Eka Kvirkvelia & Epigraphic Corpus of Georgia\\\hline
Gaia Trombin and Antonio Bonaldo & Out of the wall. 3D documentation and restoration of an endangered Roman altar\\\hline
Gaia Lembi & Saxa Judaica: Delving Digitally into Jewish Inscriptions\\\hline
Olga Pelcer-Vujacic. Epigraphica Montenegrina & The road so far and the journey ahead\\\hline
Marion Lamé, Federico Ponchio, Ivan Radman-Livaja and Bruce Robertson & Teaching (Digital) Epigraphy\\\hline
Claudio Prandoni, Antonella Fresa & Civic Epistemologies \\\hline
Silvia Marchesini & Ancient European Languages \\
\bottomrule

\end{tabular*}}

\caption{Posters presented at the conference}
\label{tab:table2}
\end{table}

### 2016-01-18 source/21/EAGLE2016_Full_Paper_revised.tex
Paper 21 (De Vido et al) edited to the final stage. Paper also updated in the pdfs folder

@@
 \title{AXON. A Selection of Greek Historical Inscriptions. A Database for Research and Teaching}
+\author[cafo]{Stefania De Vido}
  \author[cafo]{Ivan Matijašić\corref{first}}
-\author[cafo]{Stefania De Vido}
 \author[cafo]{Silvia Palazzo}
@@
 \begin{abstract}
-The AXON Project has developed a database of Greek historical inscriptions, from the birth of the~polis~in the Archaic
+The AXON Project has developed a database of Greek historical inscriptions, from the birth of the~\emph{polis}~in the Archaic
 Age to 31 BC. Each entry is provided with the object’s description, a complete lemma, Greek text with critical
@@
 
-\noindent AXON. A Selection of Greek Historical Inscriptions is a project conceived within the Greek Epigraphy Laboratory
+\noindent \emph{AXON. A Selection of Greek Historical Inscriptions} is a project conceived within the Greek Epigraphy Laboratory
 (Director, Prof. Claudia Antonetti), and has been brought into existence with the financial support of the University
@@
 epigraphy as well as the scientific acquisitions in the Digital Humanities have been taken into account. The selection
-of texts has been made according to a broader notion of ‘historical’ inscription, including not only significant
+of texts has been made according to a broader notion of `historical' inscription, including not only significant
 military, political, and institutional texts, but also those inscriptions which are essential for the social and
@@
 
-AXON includes texts from the birth of the Greek polis in the Archaic Age\footnote{ See the Introduction in \citet{nielsen_inventory_2004}; \citet{hansen_polis:_2006}.
+AXON includes texts from the birth of the Greek \emph{polis} in the Archaic Age\footnote{ See the \emph{Introduction} in \citet{nielsen_inventory_2004}; \citet{hansen_polis:_2006}.
 } to 31 BC, a chronological frame traditionally related to Greek History (though a future
@@
 \begin{enumerate}
+\renewcommand{\theenumi}{\alph{enumi}}
 \item Object type
@@
 \begin{enumerate}
+\renewcommand{\theenumi}{\alph{enumi}}
 \item Type of Inscription (with link to EAGLE’s Vocabularies:
-\url{http://www.eagle-network.eu/resources/vocabularies/typeins/}. The categories are those used by \citet{guarducci_epigrafia_1987})
+\url{http://www.eagle-network.eu/resources/vocabularies/typeins/}. The categories are those used by \citet{Guarducci:1967aa})
 \item Text’s structure
@@
 based on the symbols of the font Cardo (\url{http://scholarsfonts.net/cardofnt.html}), but many have been developed by the
-AXON Team on the examples of letter-form given in Jeffrey 1990 (see also
+AXON Team on the examples of letter-form given in \citealp{jeffery_local_1990} (see also
 \url{http://poinikastas.csad.ox.ac.uk/browseGlyphs.shtml}).}; letters’ heights, description and layout of the text field;
@@
 
-\subsubsection{Genetic lemma \& apparatus criticus}
+\subsubsection{Genetic lemma \& \emph{apparatus criticus}}
 
-\noindent The text of each inscription is preceded by a hierarchically arranged lemma (the so called genetic lemma, according to
+\noindent The text of each inscription is preceded by a hierarchically arranged lemma (the so called \emph{genetic lemma}, according to
 Louis Robert’s definition\footnote{\citet{robert_carie:_1954}.}) and is followed by the apparatus criticus.
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 \begin{enumerate}
-\item Persons and names (mainly for ‘historical’ characters)
+\renewcommand{\theenumi}{\alph{enumi}}
+\item Persons and names (mainly for `historical' characters)
 \item Gods and heroes
@@
 \noindent Finally, an updated bibliography highlights any previous edition for each entry, as well as all the appropriate
-secondary sources. SEG abbreviations have been used for epigraphic corpora and other publications (the section
+secondary sources. SEG abbreviations have been used for epigraphic \emph{corpora} and other publications (the section
 ``materiali'' on the website gives access to a list of all abbreviations, useful for students, as well\footnote{
@@
 
-\noindent The AXON Project, as an example of a digital edition of inscriptions (see esp. genetic lemma and apparatus) with a high
+\noindent The AXON Project, as an example of a digital edition of inscriptions (see esp. genetic lemma and \emph{apparatus}) with a high
 degree of clarity for contributors and users, is a useful tool for teaching Greek epigraphy as well as ancient history.

### 2016-01-18 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
Paper 22 (Sarullo et al) edited up to end of Section 1. NOT FINISHED!

@@
 this led to new discoveries and confirmed old hypotheses. Each text will be presented in an epigraphic chart, enriched
-by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for § 1 and 2;
-Giulia Sarullo (GS) is responsible for § 4 and Marta Muscariello (MM) for § 3.}
+by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for Sec. \ref{sec:generalia} and \ref{sec:news};
+Giulia Sarullo (GS) is responsible for Sec. \ref{sec:3chart} and Marta Muscariello (MM) for Sec. \ref{sec:4tech}.}
 \end{abstract}
@@
 
-\section{Generalia [GR]}
+\section{\emph{Generalia} [GR]}\label{sec:generalia}
 
-\noindent The project Iscrizioni Latine Arcaiche (ILA) consists in the digital edition of the inscriptions found in old
-Latium\footnote{With the two exceptions of the Vendia's Urn, found in Cerveteri, and the Garigliano bowl.}
+\noindent The project \emph{Iscrizioni Latine Arcaiche (ILA)} consists in the digital edition of the inscriptions found in old
+\emph{Latium}\footnote{With the two exceptions of the \emph{Vendia's} Urn, found in Cerveteri, and the Garigliano bowl.}
 dating back to the period between the 7th and the 5th century B.C. Between the end of the 19th century and the first
-decades of the 20th century this corpus consisted of only four inscriptions of a certain length, that is the Duenos
-vase (1880), the Fibula Praenestina (1887), the Forum inscription (1899) and the Tibur pedestal inscription (1926),
+decades of the 20th century this \emph{corpus} consisted of only four inscriptions of a certain length, that is the \emph{Duenos}
+vase (1880), the \emph{Fibula Praenestina} (1887), the \emph{Forum} inscription (1899) and the \emph{Tibur} pedestal inscription (1926),
 besides other shorter but still very interesting texts that offered important cultural information (such as the
-inscription from the Regia, REX, CIL I2 479). In the second half of the 20th century, the corpus grew significantly and
+inscription from the \emph{Regia, REX, CIL} I$^2$ 479). In the second half of the 20th century, the \emph{corpus} grew significantly and
 reached the total of about eighty documents that have not been gathered in a comprehensive edition yet. The updated
-corpus includes new and crucial discoveries, some of which have already been published in a traditional way (in print),
-such as the inscriptions from Satricum and the fragments found at the Palatinum, whereas others are still unpublished,
-as in the case of the recent findings from the Regia. The quantitative and qualitative enrichment provided by recent
+\emph{corpus} includes new and crucial discoveries, some of which have already been published in a traditional way (in print),
+such as the inscriptions from \emph{Satricum} and the fragments found at the \emph{Palatinum}, whereas others are still unpublished,
+as in the case of the recent findings from the \emph{Regia}. The quantitative and qualitative enrichment provided by recent
 new readings of the texts and the new data emerged justify and require a new publication.
 
-Most of the inscriptions in the collection are frustuli or single letters: these texts, though not really relevant from
-the linguistic point of view, are important testimony of the use of writing in Latium since the 7th century B.C.
+Most of the inscriptions in the collection are \emph{frustuli} or single letters: these texts, though not really relevant from
+the linguistic point of view, are important testimony of the use of writing in \emph{Latium} since the 7th century B.C.
 
@@
 collection specifically dedicated to these documents exists. It is well known that, for every kind of publication, a
-digital edition provides several advantages, for example: the possibility to update both the textual corpus and the
+digital edition provides several advantages, for example: the possibility to update both the textual \emph{corpus} and the
 bibliography continuously; the hypertextual structure, that allows the user to utilize the edition in different ways;
 the opening to a public that is heterogeneous and wider than the one a work addressed to specialists can reach. In
-which other ways can the web respond to the specific requirements of a peculiar corpus such as ours with distinguishing
-features, different from the later Latin epigraphy optimally reproduced in the Epigraphic Database Rome (EDR)?
+which other ways can the web respond to the specific requirements of a peculiar \emph{corpus} such as ours with distinguishing
+features, different from the later Latin epigraphy optimally reproduced in the \emph{Epigraphic Database Rome} (EDR)?
 
 First of all, by using the EpiDoc encoding standard, which is compatible with other encoding systems, it will be
-possible to transfer our data to EDR and to the EAGLE-Europeana Network (see § 4).
+possible to transfer our data to EDR and to the EAGLE-Europeana Network (see Sec. \ref{sec:4tech}).
 
@@
 the inscription from different starting points, will allow the user to search the texts according to various
-parameters: dating, place of finding, object type, and textual type (see § 3).
+parameters: dating, place of finding, object type, and textual type (see Sec. \ref{sec:3chart}).
 
@@
 scientific nature and the usability of the edition, especially in the case of inscriptions that have been variously
-interpreted since their discovery. For longer inscriptions (e.g. Forum Inscription, Duenos vase, Tivoli inscription,
+interpreted since their discovery. For longer inscriptions (\emph{e.g.} \emph{Forum} Inscription, \emph{Duenos} vase, Tivoli inscription,
 Garigliano bowl), the numerous readings that have been proposed so far by scholars have been compared and verified in
@@
 The archaic Latin inscriptions play a fundamental role in the study of the first stages of the language, since they
-present particular features that allow us to investigate the various steps that led to “standard Latin”. The language
-attested in our inscriptions can be considered a Restsprache insofar as it is not `readable' through later Latin
-but it can be `interpreted'. The linguistic commentary will be carried out, in votis, in the second phase of this
+present particular features that allow us to investigate the various steps that led to ``standard Latin''. The language
+attested in our inscriptions can be considered a \emph{Restsprache} insofar as it is not `readable' through later Latin
+but it can be `interpreted'. The linguistic commentary will be carried out, \emph{in votis}, in the second phase of this
 project. Here the research focuses on the epigraphic features, that show a plurality of forms and of alphabets in such
-a limited corpus. The chronological and geographical distribution of the signs and of the variants in use in the
+a limited \emph{corpus}. The chronological and geographical distribution of the signs and of the variants in use in the
 inscriptions have been analyzed in order to offer valid elements to the study of the evolution of the alphabetic model
@@
 
-\section{Epigraphic News about the Forum Inscription [GR]}
+\section{Epigraphic News about the Forum Inscription [GR]}\label{sec:news}
 
@@
 
-\section{The Epigraphic Chart [MM]}
+\section{The Epigraphic Chart [MM]}\label{sec:3chart}
 
@@
 
-\section{Technical notes [GS]}
+\section{Technical notes [GS]}\label{sec:4tech}
 

### 2016-01-19 source/13/EAGLE2016_Full_Paper_Verreth.tex
papers revised
minor fixes

@@
 the EDCS. All these texts were 'cut up' in capital clusters, i.e. strings of consecutive words all starting with a
-capital. Words as filius, nepos and libertus were added to the string so that in most cases the full identification of
-a person could be grouped together, e.g. Quintus Caecilius Quinti filius Quirina Mustacus (TM 332260), Caio Annioleno
-Cai filio Arnensi Karthaginiensi Galliano (TM 349961) or Maesiaes Cai libertae Chrysidis (TM 244384). A minor
-disadvantage of this corpus was the capitalisation of the first word of every inscription (which is not done in the
+capital. Words as \emph{filius}, \emph{nepos} and \emph{libertus} were added to the string so that in most cases the full identification of
+a person could be grouped together, e.g. \emph{Quintus Caecilius Quinti filius Quirina Mustacus} (TM 332260), \emph{Caio Annioleno
+Cai filio Arnensi Karthaginiensi Galliano} (TM 349961) or \emph{Maesiaes Cai libertae Chrysidis} (TM 244384). A minor
+disadvantage of this corpus was the capitalization of the first word of every inscription (which is not done in the
 PN), which yields quite a significant number of mere nouns in the group of expected personal names and toponyms. Also
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
  \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img001.png}
-\caption{default}
+\caption{Named Entity Recognition with FileMaker Pro 14}
 \label{fig:1}
@@
 Within the EAGLE project we created already a fairly large reference corpus for toponyms from the Roman empire, but
-this corpus was enlarged by entering the toponyms occurring in the Itinerarium provinciarum Antonini Augusti [3.434]
-and the Tabula Peutingeriana [3.287]. The TM Geo file now contains 46.277 toponyms from all over Egypt and the Roman
+this corpus was enlarged by entering the toponyms occurring in the \emph{Itinerarium provinciarum Antonini Augusti} [3.434]
+and the \emph{Tabula Peutingeriana} [3.287]. The TM Geo file now contains 46.277 toponyms from all over Egypt and the Roman
 empire, both ancient and modern, which cover most of the places where ancient texts have been found, and most of the
@@
 names, but except for the case of the relatively straightforward tribus names, this yielded no satisfactory results. A
-lot of toponyms resemble personal names (e.g. Florentia, Venusia, (Fundus) Bassianus), and the automated identification
-of strings such as Colonia Ulpia Traiana Augusta Fidelis Lepcis Magna (TM 198383) or Municipium Augustum Hipponiensium
-Regiorum (TM 200133) seemed too cumbersome. In the end, we settled for plan B, which shows that in `Digital
+lot of toponyms resemble personal names (e.g. \emph{Florentia}, \emph{Venusia}, \emph{(Fundus) Bassianus}), and the automated identification
+of strings such as \emph{Colonia Ulpia Traiana Augusta Fidelis Lepcis Magna }(TM 198383) or \emph{Municipium Augustum Hipponiensium
+Regiorum} (TM 200133) seemed too cumbersome. In the end, we settled for plan B, which shows that in `Digital
 Humanities', the human component is still essential: we had to go through all 900.000 capital cluster cards manually,
@@
 automatically receives a chronological and a geographical context. Every toponym found in the capital cluster file is
-exported to a separate Georef card. When a toponym exists out of several consecutive elements, like the colonia and
-municipium examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as `Bythina et
-Ponthus', which cannot belong to the same capital cluster string because of the intermediary `et', are exported double
+exported to a separate Georef card. When a toponym exists out of several consecutive elements, like the \emph{colonia} and
+\emph{municipium} examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as `\emph{Bythina et
+Ponthus'}, which cannot belong to the same capital cluster string because of the intermediary `\emph{et}', are exported double
 and then afterwards joined manually. For the moment TM Georef lists 67.820 geographical references attested in Latin
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
  \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img002.png}
-\caption{default}
+\caption{The Georef Card}
 \label{fig:2}
@@
 presented the way they are on the stone, with as little additions or emendations as possible, except of course for any
-abbreviations at the end of the word; e.g. T(h)ra[c(um)] becomes Tra[c(um)], and Rom(a)e becomes Rome. In another field
-the standardized nominative case is given, without brackets or uncertainty dots; e.g. Trax and Roma.
+abbreviations at the end of the word; e.g.
+\emph{T(h)ra[c(um)]} becomes \emph{Tra[c(um)]}, and Rom(a)e becomes \emph{Rome}.
+In another field the standardized nominative case is given, without brackets or uncertainty dots; e.g.
+\emph{Trax} and \emph{Roma}.
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
  \includegraphics[width=\columnwidth]{EAGLE2016FullPaperVerreth-img003.png}
-\caption{default}
+\caption{Philadelphia}
 \label{fig:3}
@@
 the immediate phrase to which the toponym belongs. The translation should be a standardised as possible, with termini
-technici preferably added in Latin, so that the users can easily search for them in the database; e.g. `Tiberius Iulius
-Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris', `praefectus
-Aegypti', `cohors I Tungrorum'. If the place is explicitly ascribed to a provincia or a region, this provincia is
-listed in the field 'Administrative situation'. If a town is explicitly called an oppidum, a vicus or a civitas, this
+technici preferably added in Latin, so that the users can easily search for them in the database; e.g. `\emph{Tiberius Iulius
+Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris}', `\emph{praefectus
+Aegypti}', `\emph{cohors I Tungrorum}'. If the place is explicitly ascribed to a \emph{provincia} or a region, this \emph{provincia} is
+listed in the field 'Administrative situation'. If a town is explicitly called an \emph{oppidum}, a \emph{vicus} or a \emph{civitas}, this
 information is listed in the field Status. By adding this information in searchable fields we hope that the user can
 start asking quite specific questions; e.g. the first and last attestation of a place in the sources; the periods in
-which a town was called a colonia or a municipium; the places in which the `ala I Thracum Mauretana' has been attested.
+which a town was called a \emph{colonia} or a \emph{municipium}; the places in which the `\emph{ala I Thracum Mauretana}' has been attested.
 

### 2016-01-19 source/20/EAGLE2016_BONACINI - Pilot project at Paolo Orsi Museum.tex
papers revised
minor fixes

@@
 \author[catania]{Elisa Bonacini\corref{first}}
-\cortext[first]{Corresponding author. Email: e_bonacini@hotmail.com; elisa.bonacini@unict.it}
+\cortext[first]{Corresponding author. Email: e\_bonacini@hotmail.com; elisa.bonacini@unict.it}
 \address[catania]{Humanities Department, University of Catania; IEMEST (Istituto Euro Mediterraneo di Scienze e Tecnologie), Palermo, Italy}
@@
 
-Sicily has the highest number of UNESCO heritage sites (7/51 in total\footnote{1997: Valley of Temples in Agrigento;
+Sicily has the highest number of UNESCO heritage sites (7/51 in total)\footnote{1997: Valley of Temples in Agrigento;
 1997: Villa del Casale; 2000: Eolian Islands; 2002: Late Baroque Towns of the Val di Noto (South-Eastern Sicily); 2005:
 Syracuse and the rock necropolisof Pantalica; 2013: Mount Etna; 2015: Arab-Norman sites, Palermo and the Cathedral
-Churches of Cefalù and Monreale.}) and of UNESCO intangible cultural heritage (3/6\footnote{2008: Opera dei Pupi,
+Churches of Cefalù and Monreale.} and of UNESCO intangible cultural heritage (3/6)\footnote{2008: Opera dei Pupi,
 Sicilian puppet theatre; 2013: Mediterranean diet (transnational); 2014: Traditional agricultural practice of
-cultivating the `vite ad alberello' (head-trained bush vines) of the community of Pantelleria.}) in Italy and in the
+cultivating the `vite ad alberello' (head-trained bush vines) of the community of Pantelleria.} in Italy and in the
 world. Infact, Sicily has some of the largest and most important archaeological sites in the world: the temple of
@@
 Despite of this, Sicilian cultural heritage struggles to be present on Google’s platforms such as Google Art Project and
-others, as it should, compared to other Italian cultural sites \citep{Bonacini2013} \citep{Bonacini2014}).
+others, as it should, compared to other Italian cultural sites \citep{Bonacini2013, Bonacini2014}.
 
@@
 
-In Street View Gallery\footnote{\url{https://www.google.com/maps/views/streetview?gl=us}.}, which now has contributed a great
+In Street View Gallery,\footnote{\url{https://www.google.com/maps/views/streetview?gl=us}.} which now has contributed a great
 number of users from all over the world, thousands are the spherical and geo located photos of Italian places. However,
-tightening the selection to ``Landmarks of Italy''
-\footnote{\url{https://www.google.com/maps/views/streetview/italy-highlights?gl=us}.}, Sicily has only 9 spherical photos,
+tightening the selection to ``Landmarks of Italy'',
+\footnote{\url{https://www.google.com/maps/views/streetview/italy-highlights?gl=us}.} Sicily has only 9 spherical photos,
 showing the beaches of the Aeolian Islands (7), the islands of Favignana (1) and Marettimo (1).
 
-Among Art Project’s 605 museum collections\footnote{\url{https://www.google.com/maps/views/streetview/art-project?gl=it}.}, 47
+Among Art Project’s 605 museum collections,\footnote{\url{https://www.google.com/maps/views/streetview/art-project?gl=it}.} 47
 are Italian, especially relevant to Rome, Turin, Venice and Milan. One of these, not the best artistic production, is a
@@
 Browsing on the Street View Gallery, 21 sites are all over inscribed on the World Wonders
-Project\footnote{\url{https://www.google.com/maps/views/streetview/world-wonders-project?gl=it}.}: for Italy, only Pompeii
+Project:\footnote{\url{https://www.google.com/maps/views/streetview/world-wonders-project?gl=it}.} for Italy, only Pompeii
 and the historic center of Florence are inscribed in this still restricted list. Many more sites can be visited
 virtually from the same project linked on Google Cultural Institute’s
-website\footnote{\url{www.google.com/intl/it/culturalinstitute/worldwonders/}. }: 172 sites in the world, 22 in Italy and,
+website:\footnote{\url{www.google.com/intl/it/culturalinstitute/worldwonders/}. } 172 sites in the world, 22 in Italy and,
 finally, 2 of them in Sicily. Until last year the Unesco site of the baroque town Val di Noto was the only one in
@@
 objects, described by short captions, are available to users to be rotated and zoomed. The aim of this new Google’s
-project is to begin building the most important and largest database of 3D scan art of works worldwide.
+project is to begin building the most important and largest database of 3D scans of art works worldwide.
 
@@
 support the integration of menus, captions, photos, video, info inside the Street View virtual tour technology.
-Infact,it is only allowed navigation in 360°, ie a virtual walk.
+Infact, it is only allowed navigation in 360°, i.e. a virtual walk.
 
@@
 (Athenaion, Ionic temple, Olympeion and Apollonion), to the statuary and the terracotta findings from urban excavations
-during the last decades (Piazza Duomo, Ortigia, Piazza della Vittoria) and to the urban and ex-traurban necropolis
+during the last decades (Piazza Duomo, Ortigia, Piazza della Vittoria) and to the urban and extraurban necropolis
 (Fusco, Giardino Spagna) with rich funerary kits. \
@@
 \caption{360° tour of Sector D: the Venus Landolina.}
-\label{fig:}
+\label{fig:4}
 \end{figure}
@@
 
-\begin{otherlanguage*}{greek}
-chrismon ς χριστ<ι>ανῆς σεμνῆς ἀγανóφρονος [ἠδ]ὲ φιλάνδρου Nασσιανῆς τύμβον εἰσορᾷς, φίλε, κίμενον [ὣδε ἥ]τις σεμνοσύνῃσιν [ἐριζε]το Πηνελοπίῃ palma chrismon ἐνθάδε κῖτε [Nασσ]ιανή, ζήσασα [ἐν Θ(ε)ῷ] καλῶς καὶ ἀμέμπτως ἒτη λβ´, μῆνας ι´.
+chrismon \begin{otherlanguage*}{greek} ς χριστ\end{otherlanguage*}<\begin{otherlanguage*}{greek} ι\end{otherlanguage*}>\begin{otherlanguage*}{greek}ανῆς σεμνῆς ἀγανóφρονος [ἠδ]ὲ φιλάνδρου Nασσιανῆς τύμβον εἰσορᾷς, φίλε, κίμενον [ὣδε ἥ]τις σεμνοσύνῃσιν [ἐριζε]το Πηνελοπίῃ\end{otherlanguage*} palma chrismon \begin{otherlanguage*}{greek}ἐνθάδε κῖτε [Nασσ]ιανή, ζήσασα [ἐν Θ(ε)ῷ] καλῶς καὶ ἀμέμπτως ἒτη λβ´, μῆνας ι´.
 \end{otherlanguage*}
@@
 refrigerium, the funeral feast that symbolically was consumed with the dead, even this syncretic practice \ dates back
-to the pagan world \citep{Sgarlata2013}, \citep{Scandurra2014}.
+to the pagan world \citep{Sgarlata2013, Scandurra2014}.
 
@@
 
-\noindent As discussed elsewhere (Bonacini 2013 and 2014), Google is undoubtedly the most active entity in the world committed to
+\noindent As discussed elsewhere \citep{Bonacini2013, Bonacini2014}, Google is undoubtedly the most active entity in the world committed to
 preservation, dissemination and promotion of cultural heritage, well above any public institution, through an
@@
 archaeological findings and plans presented here are exclusive property of the ``Paolo Orsi'' Regional Archaeological
-Museum in Syracuse and it not allowed to be copied or duplicated by any means.
+Museum in Syracuse and are not allowed to be copied or duplicated by any means.
 

### 2016-01-19 source/29/EAGLE2016_camera_ready.tex
papers revised
minor fixes

@@
 %%%%%%%%
-\title{AUGMENTING THE WORKSPACE OF EPIGRAPHISTS
+\title{Augmenting the Workspace of Epigraphists.
 An interaction design study}
@@
 In the fields of epigraphy and archaeology, the areas of digital epigraphy and computational archaeology have benefit
-from the use of several of the sensors available in hand-held devices. Crowd-sourcing of photographic data and
+from the use of several of the sensors available in handheld devices. Crowd-sourcing of photographic data and
 geo-spatial information, augmented-reality navigation in archaeological spaces and museums, and 3D scanning of
-historical artifacts, using smart phones and tablet computers,, are few of the exemplar applications of hand-held
+historical artifacts, using smart phones and tablet computers, are few of the exemplar applications of handheld
 sensors in epigraphy and archaeology. One common component in all the aforementioned applications is the ability to
@@
 There are several examples in literature that present 3D digitization projects that have been undertaken by museums
-including the Epigraphical Museum of Athens (\citet{sullivan_analytical_2011}, \citet{papadaki_accurate_2015}), Museo Arqueológico Nacional de
+including the Epigraphical Museum of Athens \citep{sullivan_analytical_2011, papadaki_accurate_2015}, Museo Arqueológico Nacional de
 Madrid \citep{ramirez-sanchez_epigrafidigital:_2014}, Museo Nazionale Romano di Palazzo Altemps \citep{barmpoutis_interactive_2015}, Museo
 Geologico Giovanni Capellini di Bologna \citep{abate_valorizzazione_2014}, National Museums Liverpool \citep{cooper_chiswick_2007},
-Smithsonian Institution \citep{wachowiak_3d_2009}, and several other museums and institutes (\citet{barsanti_3d_2013}, \citet{landon_petroglyph_2006}, \citet{levoy_digital_2000}).
+Smithsonian Institution \citep{wachowiak_3d_2009}, and several other museums and institutes \citep{barsanti_3d_2013, landon_petroglyph_2006, levoy_digital_2000}.
 
 Several novel methods for scanning, processing, and analyzing 3D models of inscriptions have been developed, including
-methods for text extraction from inscriptions (\citet{aswatha_method_2014}, \citet{sullivan_analytical_2011}), accurate 3D scanning of
+methods for text extraction from inscriptions \citep{aswatha_method_2014, sullivan_analytical_2011}, accurate 3D scanning of
 inscriptions \citep{papadaki_accurate_2015}, visualization of inscriptions \citep{bozia_open-access_2014}, as well as 3D applications for
-other archaeological artifacts (\citet{babeu__2011}, \citet{pollefeys_image-based_2001}, \citet{malzbender_polynomial_2001}, \citet{esteban_silhouette_2004}).
+other archaeological artifacts \citep{babeu__2011, pollefeys_image-based_2001, malzbender_polynomial_2001, esteban_silhouette_2004}.
 Comparative studies of 3D scanning methods for cultural herirage can be found in \citep{pavlidis_methods_2007} and \citep{bohler_3d_2004}.
@@
 Popular low-cost NUIs, such as touch screens, marker-less position trackers, motion sensors, and head-mounted displays
-have been recently studied and employed by museums as mechanisms for multi-sensory virtual experiences (\citet{ujitoko_application_2015}, \citet{soile_accurate_2013}, \citet{ikei_experience_2015}).
+have been recently studied and employed by museums as mechanisms for multi-sensory virtual experiences \citep{ujitoko_application_2015, soile_accurate_2013, ikei_experience_2015}.
 
@@
 
-\noindent In order to achieve natural interactive relighting of an inscription, the system should imitate the process of
-relighting a handheld physical object (such as a paper cast of inscription) by reorienting the object with respect to
-the light of the environment. Without loss of generality, we can assume that the default virtual lighting source is
-located on the ceiling, right above the device, which is also very intuitive choice as it is the most probable
-real-world lighting condition. Under this assumption, a gyroscope, a sensor that tracks the orientation of the device
-with respect to the gravitational vector, is enough to track the slope of the device with respect to the virtual light.
-The top row of Fig.~\ref{fig:1} shows the approximated real-world orientation of a tablet computer as it was estimated using
-the gyroscope of the device. The orientation is updated in real-time as the user moves the device.
-
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 
+\noindent In order to achieve natural interactive relighting of an inscription, the system should imitate the process of
+relighting a handheld physical object (such as a paper cast of inscription) by reorienting the object with respect to
+the light of the environment. Without loss of generality, we can assume that the default virtual lighting source is
+located on the ceiling, right above the device, which is also very intuitive choice as it is the most probable
+real-world lighting condition. Under this assumption, a gyroscope, a sensor that tracks the orientation of the device
+with respect to the gravitational vector, is enough to track the slope of the device with respect to the virtual light.
+The top row of Fig.~\ref{fig:1} shows the approximated real-world orientation of a tablet computer as it was estimated using
+the gyroscope of the device. The orientation is updated in real-time as the user moves the device.
+
 The estimated orientation of the device can be used in order to relight the depicted 3D model of inscription using the
@@
 \subsection{Natural interactive manipulation of 3D models }
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img003.png}
+\caption{Visualization of same-size boxes using 15 different perspective projections with the same FOV angle and
+different cropping parameters. None of the boxes is rotated in the space. }
+\label{fig:2}
+\end{figure}
 
@@
 
-\begin{figure}[!bp]
-\centering
- \includegraphics[width=\columnwidth]{EAGLE2016cameraready-img003.png}
-\caption{Visualization of same-size boxes using 15 different perspective projections with the same FOV angle and
-different cropping parameters. None of the boxes is rotated in the space. }
-\label{fig:2}
-\end{figure}
-
 In the case of a tablet computer, the change of perspective can be implemented using the accelerometer of the device,
@@
 
-Interactive manipulation of a 3D digitized inscription bearer is shown in the bottom row of Fig.~\ref{fig:3}. Different sides of
-the bearer can be observed by reorienting the tablet naturally. In this example a large inscription model with its
-bearer was chosen in order to demonstrate that the proposed interaction design remains intuitive independently of the
-scale. It should also be noted that the interactive manipulation of the perspective can be optionally performed
-simultaneously with the interactive relighting as shown in Fig.~\ref{fig:1} (bottom) in order to perform a more realistic
-interaction that causes relighting and change of point of view at the same time.
-
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 \end{figure}
+
+Interactive manipulation of a 3D digitized inscription bearer is shown in the bottom row of Fig.~\ref{fig:3}. Different sides of
+the bearer can be observed by reorienting the tablet naturally. In this example a large inscription model with its
+bearer was chosen in order to demonstrate that the proposed interaction design remains intuitive independently of the
+scale. It should also be noted that the interactive manipulation of the perspective can be optionally performed
+simultaneously with the interactive relighting as shown in Fig.~\ref{fig:1} (bottom) in order to perform a more realistic
+interaction that causes relighting and change of point of view at the same time.
 
@@
 with iOS, Android, and Microsoft RT tablet and smart phone devices. The interface is available through the web-site of
-the project: http://www.digitalepigraphy.org
+the project: \url{http://www.digitalepigraphy.org}
 

### 2016-01-19 source/38/EAGLE2016_Final_Paper Modolo.tex
papers revised
minor fixes

@@
 %%%%%%%%
-\title{FREE REPRODUCTION OF CULTURAL HERITAGE FOR RESEARCH PURPOSES: REGULATORY ASPECTS AND NEW PROSPECTS FOR PROJECT `EAGLE'}
+\title{Free Reproduction of Cultural Heritage for Research Purposes:
+Regulatory Aspects and New Prospects for Project EAGLE}
 \author[free]{Mirco Modolo\corref{first}}
@@
 \end{abstract}
-\keywords{Archivi, Biblioteche, Manoscritti, Riproduzioni, Creative Commons}
+\keywords{Archives, Libraries, Manuscripts, Reproductions, Creative Commons}
 
@@
 heritage for scientific purposes (even reproduction carried out at a distance). It represented a new first for
-historical research\footnote{Decreto-legge 31 maggio 2014 n. 83, art. 12.}: numerous archives and libraries opened
+historical research:\footnote{Decreto-legge 31 maggio 2014 n. 83, art. 12.} numerous archives and libraries opened
 and made free of charge the reproduction of manuscripts and historical volumes with a user’s camera. This provided a
@@
 later, a restrictive amendment modified the original text of the law, explicitly excluding printed books, manuscripts
-and archival documents from the liberalisation\footnote{Legge 29 luglio 2014, n. 106}. Many scholars’ surprise was
+and archival documents from the liberalisation.\footnote{Legge 29 luglio 2014, n. 106} Many scholars’ surprise was
 overtaken by irritation, driven by an apparent paradox: museum photography remained free for the tourist, yet the
 scholar engaged in research activity in archives and libraries would not reap the benefits of
-liberalisation\footnote{\citet{modolo_il_2014}; \citet{brugnoli_ancora_2014}. }.
+liberalisation.\footnote{\citet{modolo_il_2014}; \citet{brugnoli_ancora_2014}. }
 
-\begin{enumerate}
-\item Current state of the problem
-\end{enumerate}
+\section{Current state of the problem}
+
 Excluding bibliographic and archival records from liberalisation has effectively restored the previous regime, still in
@@
 the images did not damage the archival material, but rather, being subject to less movement, actually contributed to
-its conservation\footnote{\url{https://fotoliberebbcc.wordpress.com/2015/06/06/4724/}}.
+its conservation.\footnote{\url{https://fotoliberebbcc.wordpress.com/2015/06/06/4724/}}
 
@@
 compressing the back of the volume and risking damaging the
-binding\footnote{\url{http://www.bl.uk/reshelp/inrrooms/stp/copy/selfsrvcopy/book\_photography\_video.mp4}.}.
+binding.\footnote{\url{http://www.bl.uk/reshelp/inrrooms/stp/copy/selfsrvcopy/book\_photography\_video.mp4}.}
 
@@
 the risk of abuse associated with using the images for profit-making activity, in a way which might `damage' the public
-treasury\footnote{\url{http://www.bianchibandinelli.it/2015/05/25/fotografie-libere-un-comunicato-dellassociazione-bianchi-bandinelli/}}.
+treasury.\footnote{\url{http://www.bianchibandinelli.it/2015/05/25/fotografie-libere-un-comunicato-dellassociazione-bianchi-bandinelli/}}
 
@@
 reproduction in archives and libraries, renewing the original spirit of the ``Art Bonus''
-decree\footnote{\url{https://fotoliberebbcc.wordpress.com/category/adesioni-e-contatti/}}. It was an unprecedented
+decree.\footnote{\url{https://fotoliberebbcc.wordpress.com/category/adesioni-e-contatti/}} It was an unprecedented
 initiative given not only the number of signatures (more than 4200), but also the quality of subscribers: a chorus
@@
 state archives and functionaries of the ministry
-itself\footnote{\url{http://www.ilgiornaledellarte.com/articoli/2015/4/123892.html}.}.
+itself.\footnote{\url{http://www.ilgiornaledellarte.com/articoli/2015/4/123892.html}.}
 
 Moreover, the movement produced a proposal to amend article 108 of the Codice dei Beni Culturali regulating the
-reproduction of cultural heritage (\url{https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/}), to eliminate the
+reproduction of cultural heritage,\footnote{\url{https://fotoliberebbcc.wordpress.com/category/la-nostra-proposta/}} to eliminate the
 exclusion of printed books and documents from the liberalisation archive, hence complying with copyright and personal
@@
 provided for a particular form of profit, scientific publishing, given the belief that research is meaningful only if
-it can be disseminated in the broadest way\footnote{Circolare Mibact 21/2005.}. ``Fotografie libere per i Beni
+it can be disseminated in the broadest way.\footnote{Circolare Mibact 21/2005.} ``Fotografie libere per i Beni
 Culturali'' in fact proposes to make free the publication of images of cultural heritage in scientific texts in
@@
 collections. Museums, archives and libraries should therefore be seen as active and dynamic centres of cultural
-promotion, more so than bureaucratic offices for the static conservation of cultural heritage\footnote{Cfr. on the
-Creative Commons licences in archaeology: \citet{brugnoli_fotografia_2013}; \citet{gualandi}.}.
+promotion, more so than bureaucratic offices for the static conservation of cultural heritage.\footnote{Cfr. on the
+Creative Commons licences in archaeology: \citet{brugnoli_fotografia_2013}; \citet{gualandi}.}
 
@@
 parliamentary question to the minister on whether to restore the original sprit of the ``Art Bonus''
-decree\footnote{\url{http://www.senato.it/japp/bgt/showdoc/showText?tipodoc=Sindisp\&leg=17\&id=914146}}, and the
+decree,\footnote{\url{http://www.senato.it/japp/bgt/showdoc/showText?tipodoc=Sindisp\&leg=17\&id=914146}} and the
 comparable availability shown by the Ministry of Culture, are in fact events which inspire cautious optimism –
@@
 \item the free reuse of images of cultural heritage for purposes outside of financial gain, or free publication online
-and in print within the limits mentioned above\footnote{Cfr. n. 10.}.
+and in print within the limits mentioned above.\footnote{Cfr. n. 10.}
 

### 2016-01-19 source/44/Damnjanovic_et_al_EAGLE2016.tex
papers revised
minor fixes

@@
 %%%%%%%%
-\title{INTEGRATION OF MULTIMEDIA COLLECTIONS AND TOOLS FOR INTERACTION WITH DIGITAL CONTENT.
+\title{Integration of Multimedia Collections and Tools
+for Interaction with Digital Content.
 The case study of the Archaia Kypriaki Grammateia Digital Corpus}
@@
 \date{2015-11-26}
-\cortext[first]{Corresponding author. Email: (u.damnjanovic, v.vassallo, s.hermon)@cyi.ac.cy}
+\cortext[first]{Corresponding author. Email: (u.damnjanovic, v.vassallo, s.hermon@cyi.ac.cy}
 \begin{document}
@@
 epigraphic collection is presented. Our work can be seen twofold. One aspect of our work is to provide a place where
-the data coming from various sources can stored and accessed. Another aspect is to provide users with means to explore
+the data coming from various sources can be stored and accessed. Another aspect is to provide users with means to explore
 this data. We argue that currently digital libraries are constrained by their webpage-based paradigm, thus not
@@
 experienced scholar to get lost \citep{augmenting_2009}.
-To overcome this issue, new generation of digital
+To overcome this issue, a new generation of digital
 libraries should emerge that will serve as an interactive knowledge environment in which knowledge is created by
@@
 
-According to \citep{fast_interaction_2010} digital libraries are seen as a store of epistemic potential with interaction having
-the role to reveal much of the hidden complexity. In \citep{buchel_making_2014} the authors demonstrated the role that a
+According to \citet{fast_interaction_2010} digital libraries are seen as a store of epistemic potential with interaction having
+the role to reveal much of the hidden complexity. In \citet{buchel_making_2014} the authors demonstrated the role that a
 set of interactions can play in supporting users' understanding of spaces and their formation of cognitive maps when
-working with map-based visualisations. According to \citep{algee_viewshare_2012} interactive visualizations empower users to
+working with map-based visualisations. According to \citet{algee_viewshare_2012} interactive visualizations empower users to
 discover meaning and patterns within digital collections using dynamic, interactive displays. One example of utilizing
-the interactive graphics in a real world digital library is presented in \citep{hienert_integrating_2012}. Authors developed
+the interactive graphics in a real world digital library is presented in \citet{hienert_integrating_2012}. Authors developed
 interactive visualization tools that support exploration of search queries and search results and help users in
-formulating new queries. Another example, the application called DaisyViz shown in \citep{ren_daisyviz:_2010}, tries to help
+formulating new queries. Another example, the application called DaisyViz shown in \citet{ren_daisyviz:_2010}, tries to help
 users acquire better insights of the data by enabling users to rapidly develop domain-specific information
-visualizations without traditional programming. In \citep{nasar_conceptual_2011} authors proposed a conceptual framework that
+visualizations without traditional programming. In \citet{nasar_conceptual_2011} authors proposed a conceptual framework that
 uses interactive visualizations for managing personal collections of images and videos with focus on data re-finding
@@
 results browsing, categorizing, and linking to online resources such as Google and Twitter. An approach for
-overview-first exploration of data collections based on user-selected metadata properties is presented in \citep{bernard_content-based_2012}.
+overview-first exploration of data collections based on user-selected metadata properties is presented in \citet{bernard_content-based_2012}.
 In a 2D layout representing entities of the selected property are laid out based on their similarity with
 respect to the underlying data content. The display is enhanced by compact summarizations of underlying data elements,
-and forms the basis for exploratory navigation of users in the data space. In \citep{cao_whisper:_2012} authors propose a
+and forms the basis for exploratory navigation of users in the data space. In \citet{cao_whisper:_2012} authors propose a
 novel visual design termed ``Whisper'' to fulfil the need for tracing information diffusion processes in social media, in
@@
 schema that depends on the type of data and according to the needs of the project. For most of the items, we are using
-our own metadata schema (STARC metadata schemas) \citep{ronzino_metadata_2012}; \citep{vassallo_revealing_2013}, that was developed to
+our own metadata schema (STARC metadata schemas) \citep{ronzino_metadata_2012, vassallo_revealing_2013}, that was developed to
 thoroughly describe various aspects of data creation process. \ Once data is uploaded and available, it can also be
@@
 
-\noindent We wanted to provide users with number of tools that they can use while exploring the repository. In traditional web
+\noindent We wanted to provide users with a number of tools that they can use while exploring the repository. In traditional web
 based system, navigating from one tool to another usually means that we click on a link and then move to another page
@@
 
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img001.png}
+\caption{The exploration process starts by running the search. Once the results are available the user can add or remove
+exploration tools. By default map tool is present in the view. The map tool is used to mark the geo-referenced data on
+the map, and also to perform geo-search by defining the region of interest. }
+\label{fig:1}
+\end{figure}
+
 Once the results are returned and displayed, the exploration process can start. By default the map functionality is
@@
 available in the current search. It also gets notified on any other tools that might be active in the view, so the
-necessary communication channels are established. Once the results are ready, there are number of filters available to
+necessary communication channels are established. Once the results are ready, there are a number of filters available to
 filter out the data, either by data type, collection. Search form can be used to filter out the data by typing the text
 used to filter the data. As the user types, the data is automatically updated to show only those items that are related
-to the text in the search field. The user can then run a new search repeatedly, and the search history functionality
+to the text in the search field.
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img002.png}
+\caption{The annotation tool. The user first selects the object in the search results, then selects the region of the
+object and then adds the annotation. The annotation can be accessed when the annotated object is accessed, and by
+specifying in the search that the search should also include annotations. }
+\label{fig:2}
+\end{figure}
+
+The user can then run a new search repeatedly, and the search history functionality
 will store all those searches for later reference. The map tool (Fig.~\ref{fig:1}) is used to show information about the
@@
 
-\begin{figure}[!bp]
-\centering
- \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img001.png}
-\caption{The exploration process starts by running the search. Once the results are available the user can add or remove
-exploration tools. By default map tool is present in the view. The map tool is used to mark the geo-referenced data on
-the map, and also to perform geo-search by defining the region of interest. }
-\label{fig:1}
-\end{figure}
-
 Annotation tool is used to attach the information to the objects in the repository and share it with other users, or use
@@
 of the selected part.
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[width=\columnwidth]{DamnjanovicetalEAGLE2016-img002.png}
-\caption{The annotation tool. The user first selects the object in the search results, then selects the region of the
-object and then adds the annotation. The annotation can be accessed when the annotated object is accessed, and by
-specifying in the search that the search should also include annotations. }
-\label{fig:2}
-\end{figure}
 
@@
 distinguished by its multidisciplinary structure, include all disciplines relevant for description and representation
-of epigraphies: archaeology (investigating for example the context of the finds), philology \ (analyzing the text, the
+of epigraphies: archaeology (investigating for example the context of the finds), philology s(analyzing the text, the
 writing style, the scripture), chemistry and geology (providing details on the material upon which inscriptions were

### 2016-01-19 source/8/EAGLE_2016_submission_8_revised.tex
papers revised
minor fixes

@@
 
-% dummy text (remove in a normal thesis)
-% remove if not necessary
-\usepackage{siunitx}
 %Natbib for bibliography management
@@
 %all title info missing!
-\title{A VIRTUAL RESEARCH ENVIRONMENT TO DOCUMENT AND ANALYZE NON-ALPHABETIC WRITING SYSTEMS. A Case Study for Maya Writing.}
-\author[bonn]{Katja Diederichs\corref{first}}
-\author[bonn]{Sven Gronemeyer}
-\author[bonn]{Christian Prager}
-\author[bonn]{Elisabeth Wagner}
-\author[bonn]{Franziska Diehr}
-\author[bonn]{Brodhun Maximilian}
-\author[bonn]{Nikolai Grube}
-
-\address[bonn]{missing}
-
-\cortext[bonn]{missing}
+\title{A Virtual Research Environment to 
+Document and Analyze Non-alphabetic Writing Systems. 
+A Case Study for Maya Writing}
+\author[BONN]{Katja Diederichs\corref{first}}
+\author[BONN,MEL]{Sven Gronemeyer}
+\author[BONN]{Christian Prager}
+\author[BONN]{Elisabeth Wagner}
+\author[GOT]{Franziska Diehr}
+\author[GOT]{Maximilian Brodhun}
+\author[BONN]{Nikolai Grube}
+%
+\address[GOT]{Niedersächsische Staats- und Universitätsbibliothek Göttingen}
+\address[MEL]{La Trobe University Melbourne}
+\address[BONN]{Rheinische Friedrich-Wilhelms-Universität Bonn}
+%
+\cortext[first]{Corresponding author. Email: diederichs@uni-bonn.de}
 
 \begin{document}
-
 \maketitle
@@
 \begin{enumerate}
-\item creation of a markup schema to represent these steps (which in practice are interconnected
+\item Creation of a markup schema to represent these steps (which in practice are interconnected
 and may also occur parallel to each other in the form of alternative interpretations); 
-\item defining of the precise requirements that a data structure must fulfill in order to sufficiently represent complex
+\item Defining of the precise requirements that a data structure must fulfill in order to sufficiently represent complex
 i.e. non-alphabetic textual data for epigraphic analysis and research. 
@@
 
-\subsubsection[1. Allography of Graphemes ]{1. Allography of Graphemes }
+\vspace{0.5cm}
+\noindent\textbf{Allography of Graphemes }\\
 % THESE ARE SUBSUB*SUB*SECTIONS...
@@
 
-\subsubsection[2. Graphotactic Strategies of a Writing System ]{2. Graphotactic Strategies of a Writing System
-}
+\vspace{0.5cm}
+\noindent\textbf{Graphotactic Strategies of a Writing System}\\
 For analysis, it is important to convey the graphemics of a writing system, which are lost in an alphabetic

### 2016-01-19 source/9/Latin epigraphy for the visually impaired.tex
more biblio added, papers compiled with bibliography
papers revised
minor fixes

@@
 recognized and respected. These principles are reaffirmed by the UN Convention on the Rights of Persons with
-Disabilities of 2006, which states (Art. 30) that the Member States must recognize ``the right of persons with
-disabilities to take part on an equal basis with others in cultural life and should take all appropriate measures to
+Disabilities of 2006, which states (Art. 30) that the Member States must recognize
+
+\begin{quotation}
+the right of persons with disabilities to take part on an equal basis with others in cultural life and should take all appropriate measures to
 ensure to persons with disabilities:
@@
 \item[C] enjoy access to places of cultural activities, such as theaters, museums, cinemas, libraries and tourism services,
-and, as far as possible, to monuments and sites of national cultural importance.''
+and, as far as possible, to monuments and sites of national cultural importance.
 
@@
 
+\end{quotation}
 The problem of accessibility and enjoyment of works of art by the visually impaired has long existed. Sculptures, which
@@
 The best solution to make the epigraphic media available to the visually impaired turns out to be, therefore, one that
-involves more visitors who can better understand the artwork. In this regard, it being understood that the chance of
+involves more visitors who can better understand the artwork. In this regard, understood that the chance of
 touching an original work wherever possible is a uniquely fascinating act, it becomes necessary to propose
@@
 \section{The Techniques of Accessibility}
-\noindent In the world of epigraphy, casts of inscriptions are often made use of, both to facilitate their study by specialists,
+\noindent In the world of epigraphy, casts of inscriptions are often used, both to facilitate their study by specialists,
 and also to replace those exhibits that had been located outside and brought indoors to protect them from weather
@@
 These reproductions make for the best situation where the user can differentiate between the various levels. The
-background can be made on one level, while the drawn design can be made on another and also possible written upon. The
+background can be made on one level, while the drawn design can be made on another upon which it is possible to write. The
 use of 3D would obtain a result similar to that by the Federazione Nazionale delle Istituzioni Pro Ciechi (National
@@
 of Ostia’s famous lighthouse was created with different levels of elevation of a mosaic in black and white, which allow
-the user to not only to understand the image, but also to understand the technique of its realization. In this case,
+the user not only to understand the image, but also to understand the technique of its realization. In this case,
 though, the reproduction does not contain writing, because the chosen image contained no inscriptions. But nothing
@@
 
-\nocite{accessibilità,Bellini2000,Cetorelli2004,Hersh2008,Jansson2003,Levi2015,Velestino2007}
+\nocite{accessibilita,Bellini2000,Cetorelli2004,Hersh2008,Jansson2003,Levi2015,Velestino2007}
 

### 2016-01-19 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex

@@
 this led to new discoveries and confirmed old hypotheses. Each text will be presented in an epigraphic chart, enriched
-by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for Sec. \ref{sec:generalia} and \ref{sec:news};
-Giulia Sarullo (GS) is responsible for Sec. \ref{sec:3chart} and Marta Muscariello (MM) for Sec. \ref{sec:4tech}.}
+by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for § \ref{sec:generalia} and \ref{sec:news};
+Giulia Sarullo (GS) is responsible for § \ref{sec:3chart} and Marta Muscariello (MM) for § \ref{sec:4tech}.}
 \end{abstract}
@@
 the inscription from different starting points, will allow the user to search the texts according to various
-parameters: dating, place of finding, object type, and textual type (see Sec. \ref{sec:3chart}).
+parameters: dating, place of finding, object type, and textual type (see § \ref{sec:3chart}).
 
@@
 \noindent Thanks to the agreement and collaboration with public authorities, we were invited to take part in an extraordinary
-event (July 3rd 2015), that is the 3D laser scanning of the Forum Cippus (CIL I2, 1).\footnote{Cippus of tuff found
-under the Lapis Niger by Giacomo Boni in 1899. The inscription is cut boustrophedon on the four faces of the pillar and
+event (July 3rd 2015), that is the 3D laser scanning of the Forum Cippus (CIL I$^2$, 1).\footnote{Cippus of tuff found
+under the \emph{Lapis Niger} by Giacomo Boni in 1899. The inscription is cut boustrophedon on the four faces of the pillar and
 on the edge sliced between face D and face A. The stone (late 6th B.C.) is badly damaged in the upper part so that the
@@
 comparison with the other faces in which three vertical dots divide the syntactical units. This lack was particularly
-suspicious in a sequence in which there are no exegetical alternatives (SAKROSESED = sakros esed). As a consequence,
+suspicious in a sequence in which there are no exegetical alternatives (SAKROSESED = \emph{sakros esed}). As a consequence,
 scholars tried to find an explanation for this absence: the inscription was carved by different hands; the inscription
-was made up of different texts, each copied from different drafts; the antigraph was in scriptio continua and the
+was made up of different texts, each copied from different drafts; the antigraph was in \emph{scriptio continua} and the
 inscriber was not familiar with this procedure; the punctuation was not accurately assigned.
 
-As a matter of fact, face A conforms to the others, showing three vertical dots after sakros and also, although less
-deep and more worn-out, after esed\footnote{The details of significant portions of the inscriptions will be shown on
+As a matter of fact, face A conforms to the others, showing three vertical dots after \emph{sakros} and also, although less
+deep and more worn-out, after \emph{esed}\footnote{The details of significant portions of the inscriptions will be shown on
 the website with several photographs. } (Fig.~\ref{fig:1}). This fact was highlighted by \citet{gamurrini_paleografia_1899} and the three dots
 appear in the apograph in \citet[col. 1003]{hulsen_neue_1899},  and they could also be guessed in the photograph Anderson 3192
-(Archivio Alinari). Nevertheless, the post \citet{goidanich_rapporti_1943} vulgata and the difficulties in checking in person the
+(Archivio Alinari). Nevertheless, the \emph{post} \citet{goidanich_rapporti_1943} \emph{vulgata} and the difficulties in checking in person the
 stone, because of its almost unreachable and scarcely illuminated position that led to the publication of studies not
@@
 
-The most relevant news arrives from the analysis of line 16 (face E). This has been read up to now as loiuquod()qo
-\citep{wachter_altlateinische_1987}, loi\{u\}quiod \citep{_studies_1993}, LOIUQUIOD[QO] \citep{baldi_foundations_2002}, LOIUQUIOD,QO/// \citep{hartmann_fruhlateinischen_2005},\footnote{The
-final QO is based only on \citet{goidanich_rapporti_1943}.} in order to explain the `unusual' shape of V (letter no. 4) that has
-been recently read as F \citep{prosdocimi_roma_2010}. As the scanning showed, the first vertical stroke, read as an I, is much
-closer to the sign that looks like a V than it appears on the apographs and the photographs published so far;
+The most relevant news arrives from the analysis of line 16 (face E). This has been read up to now as \emph{loiuquod()qo}
+\citep{wachter_altlateinische_1987}, \emph{loi\{u\}quiod} \citep{_studies_1993}, \emph{LOIUQUIOD[QO}] \citep{baldi_foundations_2002}, \emph{LOIUQUIOD,QO///} \citep{hartmann_fruhlateinischen_2005},\footnote{The
+final \emph{QO} is based only on \citet{goidanich_rapporti_1943}.} in order to explain the `unusual' shape of \emph{V} (letter no. 4) that has
+been recently read as \emph{F} \citep{prosdocimi_roma_2010}. As the scanning showed, the first vertical stroke, read as an \emph{I}, is much
+closer to the sign that looks like a \emph{V} than it appears on the apographs and the photographs published so far;
 especially shots taken from an oblique and not a frontal point of view can be misleading.
@@
 This point can result fundamental in understanding how the inscriber worked on the stone and how he corrected the
-sign.\footnote{Other corrections can be found on the last line of face C (kapia on kapa) and on the second line of face
-D where a V was corrected into a koppa. } Hypothesis no. 1 (which is also the simplest one): the sequence to write was
-LOIVQVIOD but, after he cut a first vertical stroke, the inscriber mistakenly started to cut the oblique stroke of a V
+sign.\footnote{Other corrections can be found on the last line of face C (\emph{kapia} on \emph{kapa}) and on the second line of face
+D where a \emph{V} was corrected into a \emph{koppa}. } Hypothesis no. 1 (which is also the simplest one): the sequence to write was
+\emph{LOIVQVIOD} but, after he cut a first vertical stroke, the inscriber mistakenly started to cut the oblique stroke of a \emph{V}
 instead of a second vertical stroke that would begin the V; he recognized the error and cut a vertical stroke next to
 the first one and finally another oblique stroke that reached the bottom of the second vertical stroke, completing the
-V. Hypothesis no. 2 presumes the same order in cutting the strokes, but for a different reason: the sequence to write
-was LOVQVIOD, the inscriber started cutting the first stroke but he found an obstacle (i.e. a hole in the stone), so he
+\emph{V}. Hypothesis no. 2 presumes the same order in cutting the strokes, but for a different reason: the sequence to write
+was \emph{LOVQVIOD}, the inscriber started cutting the first stroke but he found an obstacle (\emph{i.e.} a hole in the stone), so he
 continued with the oblique stroke on the right (thinner than the others) but he changed his mind and cut a second
 vertical stroke close to the first one and joined it with a new oblique stroke. The short inner stroke has been
-considered as a correction, i.e. a deletion, and caused the expunction of the whole letter or its reading as an F; as a
+considered as a correction, \emph{i.e}. a deletion, and caused the expunction of the whole letter or its reading as an \emph{F}; as a
 matter of fact, it is nothing else but the result of a first try to cut the oblique stroke. Of course, we could be more
 precise after we receive the outcomes of the scanning, that will be ready soon. The following sign (nr. 4 or 5,
-depending on the reading), instead, is surely a koppa and not an O.
+depending on the reading), instead, is surely a \emph{koppa} and not an \emph{O}.
 
 Our reading proposal has the advantage of illustrating the sequence of the inscribing but does not solve the
-interpretative problems: a louquiod instead of a loiuquiod / loiquiod (for which lucus, eloquium, licium and liquidus
+interpretative problems: a \emph{louquiod} instead of a \emph{loiuquiod} / \emph{loiquiod} (for which \emph{lucus}, \emph{eloquium}, \emph{licium} and \emph{liquidus}
 have been proposed) still needs to be explained and both still await a solution.
@@
 
-\noindent In the website, each text of the ancient Latin corpus is presented by an ad hoc designed chart in order to meet the
-peculiar requirements of this kind of corpus. The chart is organized in items that concern every important aspect
+\noindent In the website, each text of the ancient Latin \emph{corpus} is presented by an \emph{ad hoc} designed chart in order to meet the
+peculiar requirements of this kind of \emph{corpus}. The chart is organized in items that concern every important aspect
 regarding the material and cultural contextualization of the inscription, from the archaeological support to the
@@
 In this initial phase of \ Latin literacy, the linguistic data are not sufficient to establish the linguistic features
-of this language – the corpus chiefly consists of hapax. As a consequence, the contextual data of the inscribed object
+of this language – the \emph{corpus} chiefly consists of \emph{hapax}. As a consequence, the contextual data of the inscribed object
 results being of great help to comprehend the text. For example, the new archaeological data found during the recent
-excavation campaign in the Comitium carried on under the direction of P. Fortini are providing helpful information for
+excavation campaign in the \emph{Comitium} carried on under the direction of P. Fortini are providing helpful information for
 the study of the Forum Inscription; in the past, the collection of all the fragmentary instrumental Latin inscriptions
 up to the 4th century B.C. published by G. Colonna in 1980\footnote{In \citet{stibbe_lapis_1980}.} supplied, at least
-partially, the extent of the alphabetization developing in Latium Vetus, subtracting the major inscriptions from a sort
-of “documental isolation”; moreover, the data concerning the interaction with other inscriptions of ancient Italy are
+partially, the extent of the alphabetization developing in \emph{Latium Vetus}, subtracting the major inscriptions from a sort
+of ``documental isolation''; moreover, the data concerning the interaction with other inscriptions of ancient Italy are
 fundamental, since these are different in languages and alphabets: besides the Etruscan examples in Rome, we can
-remember the case of Satricum, where both Latin and Volscan are attested,\footnote{See \citet[189-198]{rocca_i_1995}.} or the
+remember the case of \emph{Satricum}, where both Latin and Volscan are attested,\footnote{See \citet[189-198]{rocca_i_1995}.} or the
 Garigliano bowl which bears, together with the Latin inscription engraved inside, an inscription in Italic alphabet and
@@
 with the find-spot (using the ancient place name whenever possible) followed by a progressive number: for example, the
-Tibur pedestal inscription (CIL I2 2658) is denominated “Tibur 1”, the inscription on the Garigliano bowl is called
-“Garigliano 1”. Given the small number of find-spots, we decided to use the full name of the place, instead of
+\emph{Tibur} pedestal inscription (\emph{CIL} I$^2$ 2658) is denominated ``Tibur 1'', the inscription on the Garigliano bowl is called
+``Garigliano 1''. Given the small number of find-spots, we decided to use the full name of the place, instead of
 abbreviations; possible new findings that could emerge in the future from the same site will be easily added to the
 corpus simply by increasing the number. Beside the ID tag attributed to the inscription we quote the most common names
-attributed to the object, stratified in time in literature and well known to the scholars (for example “Duenos vase” /
-“Vase of the Quirinal”). We then indicate here, in the item `Present collocation', where the inscription is
+attributed to the object, stratified in time in literature and well known to the scholars (for example ``\emph{Duenos} vase'' /
+``Vase of the Quirinal''). We then indicate here, in the item `Present collocation', where the inscription is
 preserved with, when possible, the inventory number.
@@
 context);\footnote{In the case of mobile objects a different place of fabrication can be presumed, as in the case of the
-Vendia's Urn, found in Cerveteri, but considered by some to have been fabricated somewhere else in Latium.} the
+\emph{Vendia}'s Urn, found in Cerveteri, but considered by some to have been fabricated somewhere else in \emph{Latium}.} the
 ‘Date', which quotes the hypotheses given by scholars about the chronological coordinates of the inscription.
-Concerning this item, we must keep in mind that the dating of the antiquissimae is often approximate, and it is based
+Concerning this item, we must keep in mind that the dating of the \emph{antiquissimae} is often approximate, and it is based
 on different criteria (at times archaeological, at others epigraphic or linguistic or with convergences of two or of
@@
 features of the language testified by these inscriptions.\footnote{On the periodization of Latin, P. Cuzzolin and G.
-Haverling state: “The division of the history of a language into different periods implies that we have a rather clear
+Haverling state: ``The division of the history of a language into different periods implies that we have a rather clear
 picture of what language we have dealing with. At two points in the history of Latin we are not quite sure of this: the
@@
 is to determine what is Latin and what is not: unfortunately there is no overall agreement on whether all of the early
-inscriptions considered to provide us with early examples of Latin actually do that.” \citep[20]{cuzzolin_syntax_2009}.}
+inscriptions considered to provide us with early examples of Latin actually do that.'' \citep[20]{cuzzolin_syntax_2009}.}
 
@@
 provides the edition of the text. The need for a diplomatic transcription is due to the problematic nature of archaic
-texts: in the case of the Duenos vase, as it is well known, the second section of the text is almost always given by
+texts: in the case of the \emph{Duenos} vase, as it is well known, the second section of the text is almost always given by
 scholars in diplomatic transcription because of the difficulties in segmenting the phrases into words (but with
@@
 In the item `Textual typology' inscriptions are classified according to the nature of the text, taking into due
-consideration the peculiarities of this corpus. In the chronological span between the 7th and the 5th century B.C. the
-codification of formularies both of possession and of gift/dedication is still in fieri in the various linguistic
+consideration the peculiarities of this \emph{corpus}. In the chronological span between the 7th and the 5th century B.C. the
+codification of formularies both of possession and of gift/dedication is still \emph{in fieri} in the various linguistic
 branches spoken in ancient Italy (with the exception of Greek); for this reason, from a classificatory point of view an
@@
 point was established by \citet{susini_epigrafia_1982}.} We then have ‘Scriptura', where the execution technique is described;
-‘Direction of writing', i.e. right-to-left, left-to-right, boustrophedon, etc.; ‘Dividing signs', in which the presence
+‘Direction of writing', \emph{i.e.} right-to-left, left-to-right, \emph{boustrophedon}, etc.; ‘Dividing signs', in which the presence
 of punctuation and its eventual function is signaled; ‘Dimensions of the letters', which is important as regards the
@@
 
-The chart is closed by the ‘Bibliography' section. The ‘Editio princeps' and the possible ‘First notice' (if the
+The chart is closed by the ‘Bibliography' section. The ‘\emph{Editio princeps}' and the possible ‘First notice' (if the
 inscribed object had been mentioned in a publication preceding the first edition of the text) are indicated in two
@@
 
-\noindent An epigraphic corpus can be digitalized in different ways, according to the specific issues that each project intends to
-tackle. Unlike EDR that, as the other projects constituting the Electronic Archive of Greek and Latin Epigraphy
+\noindent An epigraphic \emph{corpus} can be digitalized in different ways, according to the specific issues that each project intends to
+tackle. Unlike EDR that, as the other projects constituting the \emph{Electronic Archive of Greek and Latin Epigraphy}
 (EAGLE),\footnote{\url{http://www.eagle-eagle.it/}. The other databases related to EAGLE are the Epigraphic Database Bari
@@
 \url{http://www.eda-bea.es/}.} is a database, the archaic Latin inscriptions have been digitalized according to the EpiDoc
-Guidelines.\footnote{For further information on EpiDoc and its history, see \citet{elliott_epidoc_2007}.
-The guidelines are available at http://www.stoa.org/epidoc/gl/latest/.} This is a set of specifications and encoding
-tools in XML (eXtensible Markup Language) for the scientific edition of ancient texts based on the Text Encoding
-Initiative (TEI), a set of XML specifications designed for the digital publication of texts and manuscripts for
-research purposes.\footnote{http://www.tei-c.org/index.xml. See also \citet{consortium_tei_2008} (About these Guidelines):
-“The TEI encoding scheme is of particular usefulness in facilitating the loss-free interchange of data amongst
-individuals and research groups using different programs, computer systems, or application software”.} EpiDoc is
+Guidelines.\footnote{For further information on EpiDoc and its history, see \citet{Elliot:2006aa}.
+The guidelines are available at \url{http://www.stoa.org/epidoc/gl/latest/}.} This is a set of specifications and encoding
+tools in XML (\emph{eXtensible Markup Language}) for the scientific edition of ancient texts based on the \emph{Text Encoding Initiative} (TEI), a set of XML specifications designed for the digital publication of texts and manuscripts for
+research purposes.\footnote{\url{http://www.tei-c.org/index.xml}. See also \citet{Consortium:2008aa} (About these Guidelines):
+``The TEI encoding scheme is of particular usefulness in facilitating the loss-free interchange of data amongst
+individuals and research groups using different programs, computer systems, or application software''.} EpiDoc is
 becoming more and more a point of reference for digital epigraphic projects\footnote{On the relationship between EAGLE
 projects and EpiDoc see \citet{felle_esperienze_2012}.} and it is also the standard chosen for the aggregation of the archives' data in
-the recently constituted network, again called EAGLE (Europeana Network of Ancient Greek and Latin
-Epigraphy),\footnote{http://www.eagle-network.eu. About the new Best Practice Network, co-founded by the European
-Commission, see http://www.europeana.eu.} in which, besides the data of the EAGLE archives, the archaic Latin
-inscriptions corpus will also converge. As a matter of fact, inscriptions encoded with EpiDoc are not only compatible
+the recently constituted network, again called EAGLE (\emph{Europeana Network of Ancient Greek and Latin
+Epigraphy}),\footnote{\url{http://www.eagle-network.eu}. About the new \emph{Best Practice Network}, co-founded by the European
+Commission, see \url{http://www.europeana.eu}.} in which, besides the data of the EAGLE archives, the archaic Latin
+inscriptions \emph{corpus} will also converge. As a matter of fact, inscriptions encoded with EpiDoc are not only compatible
 with other projects created according to these Guidelines, but they can also be transferred from a system into another
@@
 
-Furthermore, the XML edition of an inscription (or of an entire corpus) created according to the EpiDoc Guidelines will
+Furthermore, the XML edition of an inscription (or of an entire \emph{corpus}) created according to the EpiDoc Guidelines will
 produce a digital edition of the text that complies to Leiden Conventions\footnote{On Leiden Conventions, the standard
@@
 epigraphic documents, it has been necessary to adapt these Guidelines to respond to the peculiar issues of this
-corpus.\footnote{See \citet[162-167]{sarullo_ledizione_2011}, where a few examples of markup are quoted. } For this reason, some elements
+\emph{corpus}.\footnote{See \citet[162-167]{sarullo_ledizione_2011}, where a few examples of markup are quoted. } For this reason, some elements
 have been adapted and others have been specially designed, and this was possible thanks to the fact that XML is an
@@
 The major encoding issues concern the rendering of the direction of writing and of reversed and upside-down letters.
-Unlike later texts, predominantly left-to-right, the inscriptions of the ILA corpus show a certain degree of
+Unlike later texts, predominantly left-to-right, the inscriptions of the ILA \emph{corpus} show a certain degree of
 fluctuation in the direction of writing. Besides left-to-right, right-to-left and boustrophedic inscriptions, there are
-also some particular cases, such as the lamina from Lavinium (CIL I2 2833) and the Tibur pedestal inscription (CIL I2
-2658), that requires a specific treatment. For these texts, it was necessary to adopt the {\textless}lb
-rend={\textquotedbl}up-to-down{\textquotedbl}{\textgreater} and {\textless}lb
-rend={\textquotedbl}down-to-up{\textquotedbl}{\textgreater} elements in order to render the peculiar directions in
+also some particular cases, such as the lamina from Lavinium (CIL I$^2$ 2833) and the Tibur pedestal inscription (CIL I$^2$
+2658), that requires a specific treatment. For these texts, it was necessary to adopt the \texttt{{\textless}lb
+rend={\textquotedbl}up-to-down{\textquotedbl}{\textgreater}} and \texttt{{\textless}lb
+rend={\textquotedbl}down-to-up{\textquotedbl}{\textgreater}} elements in order to render the peculiar directions in
 which the text was cut.\footnote{The issue of the direction of writing was the subject of much debate at the 6th EAGLE
-International Event Off the beaten track. Epigraphy at the borders (Bari, September 24th{}-25th, 2015). The discussion
+International Event \emph{Off the beaten track. Epigraphy at the borders} (Bari, September 24th{}-25th, 2015). The discussion
 highlighted how this is a matter of great relevance both for the archaic inscriptions and the testimonies from late
-antiquity and the necessity to establish a common standard to encode the instances of “non-standard” directions of
+antiquity and the necessity to establish a common standard to encode the instances of ``non-standard'' directions of
 writing emerged. For this reason, this issue will surely be the object of further consideration in the future.}
 Reversed and upside-down letters are usually left unmarked in traditional epigraphic editions and we decided to comply
-to this practice. Nevertheless, the {\textless}hi{\textgreater} element has been used to mark up these letters, with
-two different values of @rend. For reversed letters, the {\textless}hi
-rend={\textquotedbl}reversed{\textquotedbl}{\textgreater} was used, a tag that in the EpiDoc Guidelines is used to
-encode the litterae inversae, enclosed in double round parentheses, such as in ((C)) for mulier.\footnote{See \citet[1722]{panciera_i_2006}.} For upside-down letters, a new value was provided, {\textless}hi
-rend={\textquotedbl}upside-down{\textquotedbl}{\textgreater}, since none of the allowed values of @rend for the
-{\textless}hi{\textgreater} element is suitable for this issue.
+to this practice. Nevertheless, the \texttt{{\textless}hi{\textgreater}} element has been used to mark up these letters, with
+two different values of @rend. For reversed letters, the \texttt{{\textless}hi
+rend={\textquotedbl}reversed{\textquotedbl}{\textgreater}} was used, a tag that in the EpiDoc Guidelines is used to
+encode the \emph{litterae inversae}, enclosed in double round parentheses, such as in ((C)) for mulier.\footnote{See \citet[1722]{panciera_i_2006}.} For upside-down letters, a new value was provided, \texttt{{\textless}hi
+rend={\textquotedbl}upside-down{\textquotedbl}{\textgreater}}, since none of the allowed values of \texttt{@rend} for the
+\texttt{{\textless}hi{\textgreater}} element is suitable for this issue.
 
 XML also allows us to encode the semantic structure of the texts. This kind of markup does not influence how the text is
-displayed but it is essential to generate the Index verborum and to allow a word-based search within the corpus. The
+displayed but it is essential to generate the \emph{Index verborum} and to allow a word-based search within the \emph{corpus}. The
 antiquity of the documents compelled to index the words as they appear in the inscription, because in most of the cases

### 2016-01-20 source/17/Zuddas Epigraphy in Italian High Schools (final).tex
fixed few more papers
minor changes. THERE HAS BEEN CONFUSION WITH THE BIBLIO, MASTER AND
SOME OTHER FILES

@@
 about Roman Assisi had been
-written\footnote{\url{http://www.liceoassisi.it/index.php?option=com\_content\&view=article\&id=47:assisi-romana\&catid=6:i-nostri-risultati\&Itemid=34.}},
+written,\footnote{\url{http://www.liceoassisi.it/index.php?option=com\_content\&view=article\&id=47:assisi-romana\&catid=6:i-nostri-risultati\&Itemid=34.}}
 both in Italian and English, comprising a page about the famous tetrastyle with a translation of the inscription on the
@@
 \begin{itemize}
-\item elements of Latin grammar, syntax and linguistics\footnote{\citet{Hartnett2012}; \citet{McCarthy1992}.};
+\item elements of Latin grammar, syntax and linguistics;\footnote{\citet{Hartnett2012, McCarthy1992}.}
 \item culture and history;
@@
 \begin{itemize}
-\item the ones that are relevant for their Italian derivates (e.g. os, oris);
+\item the ones that are relevant for their Italian derivates (e.g. \emph{os, oris});
 \item the ones that belong to the most significant semantic fields (always keeping in mind that some words are more
-important than others: for instance, knowing the difference between bellum and pugna helps to develop the comprehension
-of the two different categories; on the contrary, a non-specialized student does not need to know who a primus pilus
+important than others: for instance, knowing the difference between \emph{bellum} and \emph{pugna} helps to develop the comprehension
+of the two different categories; on the contrary, a non-specialized student does not need to know who a \emph{primus pilus}
 was);
@@
 \begin{quotation}
-P. Decimius P. l. Eros / Merula, medicus / clinicus, chirurgus, / ocularius, VIvir. / Hic pro libertate dedit
+\emph{P. Decimius P. l. Eros / Merula, medicus / clinicus, chirurgus, / ocularius, VIvir. / Hic pro libertate dedit
 ((sestertium)) (quinquaginta milia). / Hic pro seviratu in rem p(ublicam) / dedit ((sestertium)) (duo milia). / Hic in
 statuas ponendas in / aedem Herculis dedit ((sestertium)) (triginta milia). / Hic in vias sternendas in / publicum
-dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni ((sestertium))... 
+dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni ((sestertium))... }
 \end{quotation}
@@
 not only does a student make contact with the technical lexicon of medicine and words related to building activities
-(statuas ponere, vias sternere), but he can also perceive the ``evergetic spirit'' of an ancient society: a physician was
+(\emph{statuas ponere, vias sternere}), but he can also perceive the ``evergetic spirit'' of an ancient society: a physician was
 usually a freedman coming from the East, who could be very rich and spend his money for public utility; whenever a
-person obtained a priesthood, he used to pay a summa honoraria. The text can also be used for linguistic purposes
-(prepositions like pro; partitive genitive; deponent verbs).
+person obtained a priesthood, he used to pay a \emph{summa honoraria}. The text can also be used for linguistic purposes
+(prepositions like \emph{pro}; partitive genitive; deponent verbs).
 
 This does not mean that inscriptions are not useful for seeing specific abstract words in their context: when talking
-about Roman virtues and mos maiorum, the famous Clipeus from Arles (\emph{AE} 1952, 0165 = \emph{AE} 1994, 0227) is a
-perfect complement to chapter 34 of the Res Gestae (formally an inscription, too): \begin{quotation}
+about Roman virtues and \emph{mos maiorum}, the famous \emph{Clipeus} from Arles (\emph{AE} 1952, 0165 = \emph{AE} 1994, 0227) is a
+perfect complement to chapter 34 of the \emph{Res Gestae} (formally an inscription, too): \begin{quotation}
 \emph{Senatus / populusque Romanus /
@@
 
-For instance, if I say Iulia rosas amat, even an Italian who knows no Latin can get the meaning of the sentence.
+For instance, if I say\emph{ Iulia rosas amat}, even an Italian who knows no Latin can get the meaning of the sentence.
 Problems are encountered when the sentences and texts to translate become more complex: only when it is too late does
@@
 dead, who is the dedicant and what the relationship between them is, you have to distinguish dative and nominative
-cases. Moreover, the concordance of the different elements of the onomastics (especially praenomina and filiation) is
+cases. Moreover, the concordance of the different elements of the onomastics (especially \emph{praenomina} and filiation) is
 good way to encourage (and, for a teacher, to test) the learning of declensions not in a merely mnemonic, but also in
@@
 \item CIL 11, 05501 = EDR025449: \begin{quotation}
-Noniae Privatae, C. Propertius -{}-{}-{}-{}-{}-
+\emph{Noniae Privatae, C. Propertius -{}-{}-{}-{}-{}-}
 \end{quotation}
@@
 \item CIL 11, 05461 = EDR025411: \begin{quotation}
-A(n)noru(m) XIX. Calventia C(ai) f(ilia) Polla, L(ucius)
-Vistinius vir, Gavia mater posuer(unt).
+\emph{A(n)noru(m) XIX. Calventia C(ai) f(ilia) Polla, L(ucius)
+Vistinius vir, Gavia mater posuer(unt). }
 
@@
 \item CIL 11, 05399 = EDR025349: \begin{quotation}
-P(ublius) Decimius P(ubli) l(ibertus) Eros Merula VIvir
-viam a cisterna ad domum L(uci) Muti stravit ea pecunia -{}-{}-{}-{}-{}- 
+\emph{P(ublius) Decimius P(ubli) l(ibertus) Eros Merula VIvir
+viam a cisterna ad domum L(uci) Muti stravit ea pecunia -{}-{}-{}-{}-{}- }
 \end{quotation}
@@
 With an inscription like this you can address different topics, both historical (the importance of freedmen, the
-imperial cult) and linguistic ones (indirect complements, uses of is ea id as an adjective and to introduce a relative
+imperial cult) and linguistic ones (indirect complements, uses of \emph{is ea id} as an adjective and to introduce a relative
 clause that, in this case, has evidently been lost).
@@
 \begin{quotation}
-Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(- - -),
+\emph{Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(- - -),
 Ner(o) Babrius T(iti) f(ilius), C(aius) Capidas T(iti) f(ilius) C(ai) n(epos), V(ibius) Voisienus T(iti) f(ilius)
-marones murum ab fornice ad circum et fornicem cisternamq(ue) d(e) s(enatus) s(ententia) faciundum coiravere.
+marones murum ab fornice ad circum et fornicem cisternamq(ue) d(e) s(enatus) s(ententia) faciundum coiravere.}
 
@@
 \begin{itemize}
-\item for grammar: words of third declensions such as maro and fornix; complements of direction; gerundive to express
-purpose; use of the form {}-ere in the perfect tense personal endings;
+\item for grammar: words of third declensions such as \emph{maro} and \emph{fornix}; complements of direction; gerundive to express
+purpose; use of the form {}\emph{-ere} in the perfect tense personal endings;
 \item for history: the use of the Latin language prior to the Social War, as proof of the intense Romanization of the
-area at the end of the II century B.C.\footnote{\citet{Coarelli1991}.}; the presence of elements in the names that are not
+area at the end of the II century B.C.;\footnote{\citet{Coarelli1991}.} the presence of elements in the names that are not
 Roman but of Umbrian origin;
@@
 Schoolbooks rarely present these opportunities offered by inscriptions, so the main problem for a teacher is to access
-appropriate material\footnote{ See the observations made by \citet{Carpenter2006}.}. Holding a PhD in Late Antiquity and
+appropriate material.\footnote{ See the observations made by \citet{Carpenter2006}.} Holding a PhD in Late Antiquity and
 Umbrian inscriptions, I am fortunate enough to already know many sources. However, a graduate may not have enough
-knowledge in epigraphy, may not be familiar with the Corpus Inscriptionum Latinarum; that is why the EAGLE project can
+knowledge in epigraphy, may not be familiar with the \emph{Corpus Inscriptionum Latinarum}; that is why the EAGLE project can
 also offer tools for further insights.
@@
 \begin{itemize}
-\item \citet[39]{Barbieri2015}: the inscription of the architect C. Vettius Gratus (CIL X 3392) is
+\item \citet[39]{Barbieri2015}: the inscription of the architect \emph{C. Vettius Gratus} (CIL X 3392) is
 merely a decorative element on the page, without any relation to the topic (phonetic changes from Latin to Italian);
 
-\item \citet[40]{Domenici2012}: programmata from Pompeii are introduced to explain the Roman
+\item \citet[40]{Domenici2012}: \emph{programmata} from Pompeii are introduced to explain the Roman
 naming system;
 
-\item \citet[269-270]{Gambis2013}: tabellae defixionum are used for
+\item \citet[269-270]{Gambis2013}: \emph{tabellae defixionum} are used for
 different purposes (demonstrative adjectives, functions of subjunctive); the subject (magic in the ancient world) may
@@
 carmina epigraphica. School dictionaries are not intended for interpreting epigraphical lexicon: some words may not be
-present, their meaning may not always be explained. A lot of institutions and formulae which are clear to a specialist
-(e.g. quattuorvir iure dicundo) may be hard to understand or to translate. It is, however, also the case that a teacher
+present, their meaning may not always be explained. A lot of institutions and \emph{formulae} which are clear to a specialist
+(e.g. \emph{quattuorvir iure dicundo}) may be hard to understand or to translate. It is, however, also the case that a teacher
 cannot dedicate too much time to introducing these words to the class because his ultimate aim is different; so it is
@@
 \item CIL 11, 04431 = EDR025160: \begin{quotation}
-[Inf]austo, levis umbra, tuo mihi flebilis hora / sorte
-tua certe tempus in omne fuit.
+\emph{[Inf]austo, levis umbra, tuo mihi flebilis hora / sorte
+tua certe tempus in omne fuit.}
 
 \end{quotation}
-This funerary inscription from Ameria contains two verses, but the position of the words is tricky; infausto may be
-taken as an adjective (as it is most commonly) and its meaning is basically the same as sorte tua; students who are not
-used to poetry may not recognize the anastrophe tempus in omne.
+This funerary inscription from Ameria contains two verses, but the position of the words is tricky; \emph{infausto} may be
+taken as an adjective (as it is most commonly) and its meaning is basically the same as \emph{sorte tua}; students who are not
+used to poetry may not recognize the anastrophe \emph{tempus in omne}.
 
  \item CIL 11, 04391 = EDR025123:\begin{quotation}
- Iuliae M(arci) f(iliae) Felicitati, / uxori C(ai) Curiati
+\emph{ Iuliae M(arci) f(iliae) Felicitati, / uxori C(ai) Curiati
 Eutychetis / IIIIvir(i), magistrae Fortu/nae Mel(ioris), coll(egium) centonarior(um) / ob merita eius. Quo honore /
@@
 eius summae epu/lantes imperpetuum divider(ent), / quod si divisio die s(upra) s(cripta) celebrata non / fuerit tunc
-pertineb(it) omn(is) summa / ad familiam publicam.
+pertineb(it) omn(is) summa / ad familiam publicam. }
 
@@
 Specialists are well used to an inscription like this. A high school student may find unexpected difficulties in
-understanding the meaning of magistra (not teacher but priestess), but also centonarius or even hoc amplius; not to
+understanding the meaning of \emph{magistra} (not teacher but priestess), but also \emph{centonarius} or even \emph{hoc amplius}; not to
 mention the Roman calendar system, which always takes too much time to explain!
@@
 explanations about specific offices rendered with the technical derivative word, or for concepts omitted in the
-original Latin text, very useful. Examples of this could be words such as centonarii or marones seen in inscriptions
+original Latin text, very useful. Examples of this could be words such as \emph{centonarii} or \emph{marones} seen in inscriptions
 mentioned above, but also the following ones:
@@
 
-\item CIL 11, 01925 = EDR142701: the names of the emperors M. Aurelius Antoninus and M.
-Antoninus Pius Germaticus Sarmaticus need to be explained (Caracalla and Marcus Aurelius) to avoid confusion. It should
+\item CIL 11, 01925 = EDR142701: the names of the emperors \emph{M. Aurelius Antoninus} and \emph{M.
+Antoninus Pius Germaticus Sarmaticus} need to be explained (Caracalla and Marcus Aurelius) to avoid confusion. It should
 be noted that in my classes proper names, and not only those of emperors but also of other people, were generally
-translated into Italian, even if it is advisable to transcribe them in the nominative case\footnote{
-\url{http://www.eagle-network.eu/wiki/index.php/Guidelines\_for\_Translators}.}; the same thing was done with cognomina ex
-virtute, considering that they are intuitively interpretable for an Italian.
+translated into Italian, even if it is advisable to transcribe them in the nominative case;\footnote{
+\url{http://www.eagle-network.eu/wiki/index.php/Guidelines\_for\_Translators}.} the same thing was done with \emph{cognomina ex
+virtute}, considering that they are intuitively interpretable for an Italian.
 \end{itemize}
@@
 most typical example could be a comparison between the Tabula Claudiana (CIL 13, 01668) and Tacitus’ account (Annales
-XI, 23-24)\footnote{ On which see \citet{Jahn1993}.}: reading the original document is a privileged occasion to determine how
+XI, 23-24):\footnote{On which see \citet{Jahn1993}.} reading the original document is a privileged occasion to determine how
 reliable the historian is when using his sources.
@@
 Highly original suggestions have been provided by Mauro Reali, who is also the author of different school manuals, in a
-paper published online\footnote{\url{http://mediaclassica.loescher.it/nuove-e-\%93vecchie\%94-forme-di-multimedialita.n2799}
-; see also \citet{Reali2015}.}. Being an expert on the subject, he offers a comparison between the ``noble'' form of
-the political-philosophical amicitia presented in Cicero’s Laelius and the term \emph{amicus} mentioned in inscriptions
+paper published online.\footnote{\url{http://mediaclassica.loescher.it/nuove-e-\%93vecchie\%94-forme-di-multimedialita.n2799}
+; see also \citet{Reali2015}.} Being an expert on the subject, he offers a comparison between the ``noble'' form of
+the political-philosophical \emph{amicitia} presented in Cicero’s \emph{Laelius} and the term \emph{amicus} mentioned in inscriptions
 from the lower levels of society, such as CIL 05, 05300 from \emph{Comum} (a funerary stela made by a freedman for
 Pliny the Younger) or CIL 05, 05923 = EDR124245 regarding a strange case of ``friend deletion'' long before the Facebook
-era\footnote{ \url{http://www.laricerca.loescher.it/lingue-classiche/327-un-amico-o-amicus-e-per-sempre.html}.}.
+era.\footnote{ \url{http://www.laricerca.loescher.it/lingue-classiche/327-un-amico-o-amicus-e-per-sempre.html}.}
 
@@
 
-Viator, resiste et rogo / te et lege. Post annos XXVII ven[i] / Hispellum, in patriam meam. Scio / me oportere colere
+\emph{Viator, resiste et rogo / te et lege. Post annos XXVII ven[i] / Hispellum, in patriam meam. Scio / me oportere colere
 hunc locum / ubi ossa meorum requiescunt et mea / et amicorum meorum. Ex hoc sepulch[ro] / cippi perierunt duo et
-frontes duae sciun[t] / qui surupuit et acturi simus et legimus, / satis est testium etqs.
+frontes duae sciun[t] / qui surupuit et acturi simus et legimus, / satis est testium etqs.}
 
 \end{quotation}
-As for the Augustan age, Reali suggests showing some monumental inscriptions of the princeps (he impressively goes so
+As for the Augustan age, Reali suggests showing some monumental inscriptions of the \emph{princeps} (he impressively goes so
 far as to compare the qualifications \emph{Imperator Caesar Augustus} to a modern logo or even a \emph{hashtag}). On
-this subject, the altars Augusto sacrum put up by Perusia restituta (CIL 11, 01923 = EDR142666, EDR142667, EDR142668,
+this subject, the altars \emph{Augusto sacrum} put up by \emph{Perusia restituta} (CIL 11, 01923 = EDR142666, EDR142667, EDR142668,
 EDR142669), even with a simple text, offer the opportunity to deal with an aspect as crucial as the imperial cult.
 Traditionally, we read that Augustus was worshipped directly only in the Eastern provinces, but not in Rome and Italy.
-The inscriptions from Perusia testify that things are effectively more complex; these documents also offer remarkable
-information about the restitutio of the town, destroyed at the end of the bellum in 40 B.C. A few years later Perusia
-would become Augusta, as you can read on the city gates, especially on the newly restored Etruscan Arch (CIL 11, 01929
+The inscriptions from \emph{Perusia} testify that things are effectively more complex; these documents also offer remarkable
+information about the \emph{restitutio} of the town, destroyed at the end of the \emph{bellum} in 40 B.C. A few years later \emph{Perusia}
+would become \emph{Augusta}, as you can read on the city gates, especially on the newly restored Etruscan Arch (CIL 11, 01929
 = EDR142706).
@@
 Teaching, as I do, in a school named after Propertius, I always stress the importance of reconstructing the origin of
-the poet through the epigraphical data of the gens Propertia: the greatest number of written documents of the family
-having been found in Assisi. For more than two hundred years, beginning with the Vois(ienus) Ner. (filius) Propertius
+the poet through the epigraphical data of the \emph{gens Propertia}: the greatest number of written documents of the family
+having been found in Assisi. For more than two hundred years, beginning with the \emph{Vois(ienus) Ner. (filius) Propertius}
 mentioned among the Umbrian magistrates of the late II century B.C., this gens stands out in the town for its social
-influence and wealth\footnote{\citet{Forni1986}; \citet{Zuddas2006}.}. It is always very exciting to combine the information on
-Passennus Paullus Propertius Blaesus given by CIL 11, 05405 = EDR025355 and that contained in two Letters by Pliny (VI,
+influence and wealth.\footnote{\citet{Forni1986}; \citet{Zuddas2006}.} It is always very exciting to combine the information on
+\emph{Passennus Paullus Propertius Blaesus }given by CIL 11, 05405 = EDR025355 and that contained in two Letters by Pliny (VI,
 15 and IX, 22). Pliny, showing great concern for his friend’s illness but also great esteem for him as an elegiac poet,
 asserts that he is a descendant and fellow citizen of Propertius; the inscription (the front of an honorific base),
-providing the full name, with the tribe Sergia typical of the inhabitants of Asisium, is indirect, but clear, evidence
+providing the full name, with the tribe Sergia typical of the inhabitants of \emph{Asisium}, is indirect, but clear, evidence
 that the Augustan poet was born there. The information provided by the inscription and the literary text integrates
@@
 
-[{}- - - I]ọvino consulibb(us) (ante diem) VIII Kal(endas) Martias domum oscilavi Musae. 
+\emph{[{}- - - I]ọvino consulibb(us) (ante diem) VIII Kal(endas) Martias domum oscilavi Musae.}
 \end{quotation}
@@
 This house was still visited in the fourth century A.D., which really supports the theory that it used to be the poet’s
-residence, and remained an object of reverence for centuries\footnote{\citet[244-246]{Boldrighini2014}.}. Through the stories
+residence, and remained an object of reverence for centuries.\footnote{\citet[244-246]{Boldrighini2014}.} Through the stories
 of Passennus Paullus and Sextus Propertius macro-history and local history meet to make the past come alive in every

### 2016-01-20 source/3/EAGLE_2016_submission_3.tex
fixed few more papers
minor changes. THERE HAS BEEN CONFUSION WITH THE BIBLIO, MASTER AND
SOME OTHER FILES

@@
 The database in itself would have three components (see below Fig.~\ref{fig:schemadatabase}), following the best practices in the field
-(\citet{doi:10.3200/HMTS.37.1.34-38}): the \emph{sources database} (with ``facsimile'' transcription of the sources’ text), the \emph{central database} (the complete, corrected, integrated, standardized and coded form of the sources database) and the \emph{data releases} (destined for on-line usage, allowing easy extraction of data in view of analyses).
+\citep{doi:10.3200/HMTS.37.1.34-38}: the \emph{sources database} (with ``facsimile'' transcription of the sources’ text), the \emph{central database} (the complete, corrected, integrated, standardized and coded form of the sources database) and the \emph{data releases} (destined for on-line usage, allowing easy extraction of data in view of analyses).
 
@@
 minimum of deduced information (gender, juridical status, ethnicity of the name). At the same time, we operate some
-conventional onomastic transcriptions (AEL will be Aelius from the start, etc.). All these basic normalization
+conventional onomastic transcriptions (AEL will be \emph{Aelius} from the start, etc.). All these basic normalization
 procedures are being thoroughly explained and documented in the data entry manual, and the reason why we chose to apply
@@
 The first table to be filled in is the file of the source – \emph{Inscriptions}. First of all, each inscription gets a source
-code, formed of 5 digits and a symbol/acronym of the province’s name (D for Dacia, MS for Moesia Superior, DAL for
-Dalmatia, etc.) – so we have, for example 00001MS. The \emph{Inscriptions} table has text fields, as well as value lists.
+code, formed of 5 digits and a symbol/acronym of the province’s name (D for \emph{Dacia}, MS for \emph{Moesia Superior}, DAL for
+\emph{Dalmatia}, etc.) – so we have, for example 00001MS. The \emph{Inscriptions} table has text fields, as well as value lists.
 Then, we have text strings for: \emph{Relevant expressions}, \emph{Stylistic details}, \emph{Atypical features}, \emph{Observations}, \emph{Place of discovery}, \emph{Place of provenience}, \emph{Ancient name provenience}, \emph{Timestamp/Timeframe} and \emph{External links}.

### 2016-01-20 source/36/DatabasesToEpiDoc.tex
fixed few more papers
minor changes. THERE HAS BEEN CONFUSION WITH THE BIBLIO, MASTER AND
SOME OTHER FILES

@@
 There are several mappings involved in the aggregation work of EAGLE.
-Content providers need to map to a common Eagle Metadata Format, and the data produced will then be mapped to the Europeana Data Model\footnote{\citet{D3.1}} in order to be aggregated in Europeana, for a wider dissemination with a special eye for the general user.
+Content providers need to map to a common Eagle Metadata Format, and the data produced will then be mapped to the Europeana Data Model\footnote{\citet{sicilia_eagle_2015}} in order to be aggregated in Europeana, for a wider dissemination with a special eye for the general user.
 
@@
 
-Given the template described in Part III, and ANNEX II of the EAGLE Metadata Schema (\citet{D3.1}), an XSL Transformation converts from string epigraphic texts in marked up TEI-EPIDOC XML, following the \href{http://www.stoa.org/epidoc/gl/dev/}{EpiDoc guidelines} (\citet{Elliott2007}).
+Given the template described in Part III, and ANNEX II of the EAGLE Metadata Schema (\citet{sicilia_eagle_2015}), an XSL Transformation converts from string epigraphic texts in marked up TEI-EPIDOC XML, following the \href{http://www.stoa.org/epidoc/gl/dev/}{EpiDoc guidelines} (\citet{Elliott2007}).
 

### 2016-01-20 source/48/EAGLE16-lame-et-al-teaching.tex
fixed few more papers
minor changes. THERE HAS BEEN CONFUSION WITH THE BIBLIO, MASTER AND
SOME OTHER FILES

@@
 
-A traditional print photograph, such as this one (No. 12582) drawn from the archives of the AMZ, limits the student’s
+A traditional print photograph, such as this one (No. 12582, Fig.~\ref{fig:12582}) drawn from the archives of the AMZ, limits the student’s
 opportunity to learn to decipher these tags. Even if a professional photographer took this picture with a good quality
 camera in a fully equipped photographic studio, the reading of the inscription stays riddled with uncertainties and a
-researcher would have to check the tag de visu in order to offer a plausible transcription.
+researcher would have to check the tag \emph{de visu} in order to offer a plausible transcription.
 
@@
 
-From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in Lamé 2015, on a dispositive analysis
+From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in \citet{Lame2015}, on a dispositive analysis
 and the three fundamental systems of an inscription: writing system (wSystem), textual system (tSystem) and contextual
@@
 experts, preserving and promoting cultural heritage and, as here, teaching some skills in epigraphy.
-
-Fig. \ref{fig:markout}
 

### 2016-01-20 source/18/EAGLE2016_Full_Paper_Broux.tex
further changes
Bronx not moved to pdfs folder, as it is being compiled by Francesco

@@
 implies time-consuming and painstaking manual labor. Fortunately, in a digital environment, there are other ways, and
-Trismegistos (www.trismegistos.org) forms an ideal starting point. Trismegistos grew out of a long tradition of
+Trismegistos (\url{www.trismegistos.org}) forms an ideal starting point. Trismegistos grew out of a long tradition of
 databases and prosopographies (structured lists of people), as well as various other Ancient History projects at KU
@@
 equivalent of our family name), and a personal cognomen. Often the patronymic was inserted between the gentilicium and
-cognomen; sometimes even the papponymic and the voting tribe, e.g. Marcus Tullius Marci filius Marci nepos Cornelia
-Cicero (`Marcus Tullius Cicero, son of Marcus, grandson of Marcus, [of the] Cornelia [tribe]'). A completely new set of
+cognomen; sometimes even the papponymic and the voting tribe, e.g. \emph{Marcus Tullius Marci filius Marci nepos Cornelia
+Cicero} (`Marcus Tullius Cicero, son of Marcus, grandson of Marcus, [of the] Cornelia [tribe]'). A completely new set of
 rules is therefore being developed to apply to the Latin inscriptions.

### 2016-01-20 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
updated paper 24
including bibliography

@@
 \item[c] Authoritative work. In a traditional corpus, editors impose their knowledge in a large range of features, and thus
-they interfere in the utility of other views, to the extend of depreciating them too much. When it comes to an
+they interfere in the utility of other views, to the extent of depreciating them too much. When it comes to an
 epigraphic corpus that is written in a wide-known language, such as Latin, this auctoritas becomes particularly
@@
 of significant phonetic features. In the following maps it is displayed, in the first place, the attestation area of
-names containing the element biur, which clearly corresponds to the Iberian area, that is, the non indo-European half
+names containing the element \emph{biur}, which clearly corresponds to the Iberian area, that is, the non indo-European half
 of the Iberian peninsula; and, in the second place, it is displayed the dispersal area of the name Tancinus, which
@@
  \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img008.jpg}
-\caption{Dispersal area of the anthroponymic element biur}
+\caption{Dispersal area of the anthroponymic element \emph{biur}}
 \label{fig:}
@@
 considerable amount of foreign names, such as Latin or Greek.\footnote{ Vid. Unterman’s proposal in MLH IV, K.1.3. }
-This could be the case for markos (Marcus), bolora (Flora), kinbiria (Cimbria), antiokos (Antiochus), bilinos
-(Philinus), bilonikos (Philonicus), tais (Thais), tiokenes (Diogenes), and so on. We could finally mention, in this
-same regard, the Gaullish names observed in some Iberian inscriptions from southern France: aśetile (Adsedilus, CIL III
-5373), tesile (Tessillus, CIL III 14368.28), uaśile (Vassil(l)us, CIL XII 2286), katubaŕe (Catumarus, CIL 3, 4263),
+This could be the case for \emph{markos} (\emph{Marcus}), \emph{bolora} (\emph{Flora}), \emph{kinbiria} (\emph{Cimbria}), \emph{antiokos} (\emph{Antiochus}), \emph{bilinos}
+(\emph{Philinus}), \emph{bilonikos} (\emph{Philonicus}), \emph{tais} (\emph{Thais}), \emph{tiokenes} (\emph{Diogenes}), and so on. We could finally mention, in this
+same regard, the Gaullish names observed in some Iberian inscriptions from southern France: \emph{aśetile} (\emph{Adsedilus}, CIL III
+5373), \emph{tesile} (\emph{Tessillus}, CIL III 14368.28), \emph{uaśile} (\emph{Vassil(l)us}, CIL XII 2286), \emph{katubaŕe} (\emph{Catumarus}, CIL 3, 4263),
 etc.

### 2016-01-20 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
Paper 22 synced in the PDFS folder

@@
 \noindent An epigraphic \emph{corpus} can be digitalized in different ways, according to the specific issues that each project intends to
-tackle. Unlike EDR that, as the other projects constituting the \emph{Electronic Archive of Greek and Latin Epigraphy}
+tackle. Unlike EDR that, as t	he other projects constituting the \emph{Electronic Archive of Greek and Latin Epigraphy}
 (EAGLE),\footnote{\url{http://www.eagle-eagle.it/}. The other databases related to EAGLE are the Epigraphic Database Bari

### 2016-01-20 Posters.tex
updated papers list

@@
 Silvia Marchesini & Ancient European Languages \\
+Antonella Ferraro & From Paper to Web \\
+Lavinio Del Monaco & Le iscrizioni di Palestrina \\
 \bottomrule

### 2016-01-20 source/31/EAGLE2016_I Sicily_final copy.tex
Partial editing of paper 31 (Prag): NOT FINISHED

@@
 \begin{abstract}
-This paper introduces the EpiDoc corpus of inscriptions on stone for ancient Sicily, I.Sicily. The project is one of the
+This paper introduces the EpiDoc \emph{corpus} of inscriptions on stone for ancient Sicily, \emph{I.Sicily}. The project is one of the
 first attempts to generate a substantial regional corpus in EpiDoc. The project is confronting a number of challenges
@@
 
-\section{Introduction: What is I.Sicily}
-\noindent I.Sicily is an online, open access, digital corpus of the inscriptions on stone from ancient Sicily.\footnote{ The
+\section{Introduction: What is \emph{I.Sicily}}
+\noindent \emph{I.Sicily} is an online, open access, digital corpus of the inscriptions on stone from ancient Sicily.\footnote{ The
 corpus will be mounted at \url{www.csad.ox.ac.uk/sicily/isicily}, but is presently on a development server. A public face is
@@
 bibliographic searching. Individual inscriptions and individual museums will both be provided with URIs, as will
-personal names and individuals; places will be referenced using Pleiades; epigraphic types, materials, and supports
+personal names and individuals; places will be referenced using \emph{Pleiades}; epigraphic types, materials, and supports
 using the EAGLE vocabularies.
 
-\section{The motivations for and origins of I.Sicily}
+\section{The motivations for and origins of \emph{I.Sicily}}
 \noindent The existing epigraphic landscape in Sicily is extremely diverse in two primary regards: on the one hand, the island has
@@
 Recent overview of much of the linguistic tradition in \citet{tribulato_language_2012}; and of the epigraphic material in \citet{gulletta_sicilia_1999}.} on the other hand, the publication of this material has a very uneven record and despite an excellent
-pre-twentieth century tradition, the existing copora are far from complete and the ability of key journals such as SEG
-or AE to keep pace with local publication has been limited.\footnote{ For an overview of the corpora tradition up to
-the twentieth century, see \citet{de_vido_corpora_1999}.} A limited number of museum-based corpora have been published in recent
+pre-twentieth century tradition, the existing \emph{copora} are far from complete and the ability of key journals such as \emph{SEG}
+or \emph{AE} to keep pace with local publication has been limited.\footnote{ For an overview of the \emph{corpora} tradition up to
+the twentieth century, see \citet{de_vido_corpora_1999}.} A limited number of museum-based \emph{corpora} have been published in recent
 decades (for Catania, Palermo, Messina, and Termini Imerese, as well as the material from Lipari), but this has not
@@
 
-As noted in the introduction, some of the impetus for I.Sicily comes from a desire to exploit a substantial legacy
+As noted in the introduction, some of the impetus for \emph{I.Sicily} comes from a desire to exploit a substantial legacy
 dataset in MS Access. This consists of a single table originally constructed in MS Access 2000, and maintained
@@
 epigraphy.\footnote{ The principal results were published in \citet{prag_epigraphy_2002}, revised in \citet[159-188]{prag_sicily_2004}; cf.
-\citet{prag_nouveau_2003}, \citet{prag_ciceronian_2007}, \citet{prag_sicilia_2008}, \citet{prag_sicilia_2010}.}
+\citet{prag_nouveau_2003, prag_ciceronian_2007, prag_sicilia_2008, prag_sicilia_2010}.}
 
 In its final form the table holds data across 39 different fields, for 2575 records. 17 of these fields detail
-publication history (corpora references and other bibliography); the other fields record information on the language,
+publication history (\emph{corpora} references and other bibliography); the other fields record information on the language,
 date, provenance, current location, epigraphic type, form and material of the inscriptions, together with a free-text
@@
 XML as well as normalisation of dating and bibliographic records. This conversion is not meant to be repeated as the
-dataset, once converted to EpiDoc XML, will be edited in the I.Sicily website. While the conversion preserves the data
+dataset, once converted to EpiDoc XML, will be edited in the \emph{I.Sicily} website. While the conversion preserves the data
 from the MS Access dataset, it restructures and where possible improves or normalises it.
+%\emph{I.Sicily}
 
-By virtue of the fact that I.Sicily begins from such rich metadata, to which texts, images, and further data will be
+By virtue of the fact that \emph{I.Sicily} begins from such rich metadata, to which texts, images, and further data will be
 added over time, and because this is in turn being supplemented by an on-going programme of autopsy, the form and
-content of I.Sicily is intended to be more akin to that of a true corpus than simply a text-database, seeking to
+content of \emph{I.Sicily} is intended to be more akin to that of a true corpus than simply a text-database, seeking to
 combine a full record of past publication and study with a fully revised edition, and potential for multiple
-individuals to contribute to a process of on-going revision (see Fig.~\ref{fig:1} for a draft edition of one inscription (AE
-1962.314 = I.Sicily 820)).
+individuals to contribute to a process of on-going revision (see Fig.~\ref{fig:1} for a draft edition of one inscription (\emph{AE}
+1962.314 = \emph{I.Sicily} 820)).
 
-\section{The aims of I.Sicily}
-\noindent We outline briefly five areas in which I.Sicily aims to develop, facilitate and improve the study of epigraphic material
+\section{The aims of \emph{I.Sicily}}
+\noindent We outline briefly five areas in which \emph{I.Sicily} aims to develop, facilitate and improve the study of epigraphic material
 from ancient Sicily.
@@
 of the separation of epigraphic material according to linguistic traditions have recently been highlighted and directly
-confronted by the Corpus Inscriptionum Iudaeae/Palaestinae (CIIP), edited by H. Cotton et al.\footnote{ Original notice
-in \citet{cotton_corpus_1999}; presentation in \citet{cotton_corpus_2010}.} I.Sicily sets out to follow in that mould, since the
+confronted by the \emph{Corpus Inscriptionum Iudaeae/Palaestinae} (\emph{CIIP}), edited by H. Cotton et al.\footnote{ Original notice
+in \citet{cotton_corpus_1999}; presentation in \citet{cotton_corpus_2010}.} \emph{I.Sicily} sets out to follow in that mould, since the
 different linguistic traditions of Sicily not only exist side-by-side but interact constantly throughout the island’s
@@
 created by basic technologies such as Unicode and EpiDoc XML mean that there is now no reason not to be language
-agnostic in the inclusion of material (the point may be obvious, but the tendency towards language-specific corpora is
+agnostic in the inclusion of material (the point may be obvious, but the tendency towards language-specific \emph{corpora} is
 still marked). The opportunities and possibilities offered by these technologies are considerable, even at the most
 basic level, since, for example, searching can be made language specific or language neutral. One obvious area where
-Sicilian studies are currently hampered by this partitioning is in the study of onomastics. The Lexicon of Greek
-Personal Names records most instances of Greek names for the island, but Sicily is no less rich in non-Greek names
-(Latin and others), and at present there is no onomasticon for the island.\footnote{ \url{www.lgpn.ox.ac.uk/} [accessed
-26.09.2015].} Simply by the marking-up and indexing of all names in the island’s inscriptions, I.Sicily will generate a
-powerful tool for future study. Although I.Sicily in its first phase is not undertaking morphological or syntactical
+Sicilian studies are currently hampered by this partitioning is in the study of onomastics. The \emph{Lexicon of Greek
+Personal Names} records most instances of Greek names for the island, but Sicily is no less rich in non-Greek names
+(Latin and others), and at present there is no \emph{onomasticon} for the island.\footnote{ \url{www.lgpn.ox.ac.uk/} [accessed
+26.09.2015].} Simply by the marking-up and indexing of all names in the island’s inscriptions, \emph{I.Sicily} will generate a
+powerful tool for future study. Although \emph{I.Sicily} in its first phase is not undertaking morphological or syntactical
 mark-up, the encoding of all these text in XML constitutes the necessary first stage in such a development, and we see
@@
 chronology, etc. will likewise allow detailed analysis of cultural patterns and their relationship to language-use over
-time.\footnote{ See Prag 2002 for a first effort in this direction.}
+time.\footnote{ See \citet{prag_epigraphy_2002} for a first effort in this direction.}
 
@@
 \noindent Sicily presents a particularly extreme version of the common problem of identifying a text and its publication record.
-No existing corpus in either Greek or Latin comes close to full coverage (CIL X and IG XIV are the largest individual
-traditional corpora for the region, but both are over 125 years old and cover less than 30\% of the material now
+No existing \emph{corpus} in either Greek or Latin comes close to full coverage (\emph{CIL} X and \emph{IG} XIV are the largest individual
+traditional \emph{corpora} for the region, but both are over 125 years old and cover less than 30\% of the material now
 known).\footnote{\citet{_inscriptiones_1883}; \citet{kaibel_inscriptiones_1890}.} Existing online databases improve on this situation, but the results
-obtainable are of very varied value. The most comprehensive, in terms of the range of data recorded, is EDR (with which
-I.Sicily is collaborating), which currently reports 1,906 records for `Sicilia'; but this reduces to 833 when limited
-to texts on stone (`lapis' or `marmor'); contrast I.Sicily, with 2,563 records at the time of writing.\footnote{
-\url{www.edr-edr.it/} [accessed 26.09.2015]} Clauss Slaby reports 4,374 records for `Sicilia' (including Christian
+obtainable are of very varied value. The most comprehensive, in terms of the range of data recorded, is \emph{EDR} (with which
+\emph{I.Sicily} is collaborating), which currently reports 1,906 records for `Sicilia'; but this reduces to 833 when limited
+to texts on stone (`lapis' or `marmor'); contrast \emph{I.Sicily}, with 2,563 records at the time of writing.\footnote{
+\url{www.edr-edr.it/} [accessed 26.09.2015]} \emph{Clauss Slaby} reports 4,374 records for `Sicilia' (including Christian
 inscriptions, excluding `sigilla impressa'), but the return is inclusive of all kinds of epigraphic material, without
 indication or discrimination, contains some duplication, is much harder to reconcile to existing records, and records
-only text.\footnote{ \url{www.manfredclauss.de/} [accessed 26.09.2015].} The PHI database of Greek inscriptions has a rich
+only text.\footnote{ \url{www.manfredclauss.de/} [accessed 26.09.2015].} The \emph{PHI} database of Greek inscriptions has a rich
 record of published Greek texts, but is text only and limited in outputs.\footnote{
-\url{http://noapplet.epigraphy.packhum.org/regions/1156} [accessed 26.09.2015].} SEG references are available for 733
-inscriptions on stone and AE references for 328 (data taken from the I.Sicily database and based upon comprehensive
-manual trawls of SEG and AE).
+\url{http://noapplet.epigraphy.packhum.org/regions/1156} [accessed 26.09.2015].} \emph{SEG} references are available for 733
+inscriptions on stone and \emph{AE} references for 328 (data taken from the \emph{I.Sicily} database and based upon comprehensive
+manual trawls of \emph{SEG} and \emph{AE}).
 
-One major aim of I.Sicily, therefore, is to generate unique identifiers for each inscription - the I.Sicily number, in
-the form ISic 1234. These will be maintained as URIs, of the form:
+One major aim of \emph{I.Sicily}, therefore, is to generate unique identifiers for each inscription - the \emph{I.Sicily} number, in
+the form \emph{ISic} 1234. These will be maintained as URIs, of the form:
 
@@
 
-I.Sicily is well placed to do this since its initial dataset is primarily a bibliographic concordance of the lapidary
+\emph{I.Sicily} is well placed to do this since its initial dataset is primarily a bibliographic concordance of the lapidary
 inscriptions of Sicily. One of the associated outputs of the project will therefore be an online bibliography for
@@
 site is launched) [accessed 26.09.2015].} A locally cached version of the bibliography will be presented at the
-I.Sicily site to facilitate detailed bibliographic searching (including the identification of inscriptions by
+\emph{I.Sicily} site to facilitate detailed bibliographic searching (including the identification of inscriptions by
 publication) and to allow the generation of customised concordances.
 
-A further element of bibliographic information which I.Sicily will include is the cross-referencing and linking to
-online editions of major antiquarian corpora of Sicilian inscriptions. A growing number of these are already available
+A further element of bibliographic information which \emph{I.Sicily} will include is the cross-referencing and linking to
+online editions of major antiquarian \emph{corpora} of Sicilian inscriptions. A growing number of these are already available
 in digital format and several are already mounted in the Arachne archive, making direct page-citation
-possible.\footnote{ E.g. Castelli 1784, at \url{http://arachne.uni-koeln.de/books/Castelli1784} [accessed 26.09.2015].}
+possible.\footnote{ E.g. \citet{Castelli:1784aa}, at \url{http://arachne.uni-koeln.de/books/Castelli1784} [accessed 26.09.2015].}
 
-The richness of I.Sicily’s records in this area mean that I.Sicily is currently collaborating with both Trismegistos and
-IDES (`Integrating Digital Epigraphies').\footnote{\url{www.trismegistos.org/} [accessed 26.09.2015] and \url{http://ides.io/}
-[accessed 26.09.2105].} The former aims to generate TM numbers for all the Sicilian material (which I.Sicily will
-include); the latter is to assist IDES in the refining of links between, e.g., PHI and SEG records, and to improve
-I.Sicily’s own recording of PHI numbers.
+The richness of \emph{I.Sicily}’s records in this area mean that \emph{I.Sicily} is currently collaborating with both \emph{Trismegistos} and
+\emph{IDES} (`Integrating Digital Epigraphies').\footnote{\url{www.trismegistos.org/} [accessed 26.09.2015] and \url{http://ides.io/}
+[accessed 26.09.2105].} The former aims to generate \emph{TM} numbers for all the Sicilian material (which \emph{I.Sicily} will
+include); the latter is to assist \emph{IDES} in the refining of links between, e.g., \emph{PHI} and \emph{SEG} records, and to improve
+\emph{I.Sicily}’s own recording of PHI numbers.
 
@@
 publication in the past has frequently been relatively limited in the information which it has recorded about the
-object on which the inscription is inscribed. This is a familiar complaint, and one which I.Sicily will address
+object on which the inscription is inscribed. This is a familiar complaint, and one which \emph{I.Sicily} will address
 wherever possible through full object description and a rich photographic record. However, a corollary of this general
@@
 
-I.Sicily is making use of the TEI {\textless}msIdentifer{\textgreater} element, with its associated sub-elements in
+\emph{I.Sicily} is making use of the TEI \texttt{{\textless}msIdentifer{\textgreater}} element, with its associated sub-elements in
 order to record details of institutional collections and inventory numbers wherever possible.\footnote{\url{www.tei-c.org/release/doc/tei-p5-doc/en/html/ref-msIdentifier.html} [accessed 26.09.2015].} In order to maximise the
@@
 dott. R. Burgio at Messina; and of dott.ssa M.G. Branciforte at Catania.} It is hoped that this work will be of
-considerable value to the museums themselves, since access to the I.Sicily records should facilitate the curation,
+considerable value to the museums themselves, since access to the \emph{I.Sicily} records should facilitate the curation,
 display and accessibility of the inscriptions (see below also on translations), and we welcome future collaboration
@@
 
-Secondly, in collaboration with Dr Michael Metcalfe, I.Sicily has developed a database of Sicilian archaeological
+Secondly, in collaboration with Dr Michael Metcalfe, \emph{I.Sicily} has developed a database of Sicilian archaeological
 collections (130 at the time of writing). This database is mounted online alongside the epigraphic corpus, in a
@@
 
-\noindent I.Sicily is actively generating rich geo-data for the individual inscriptions, both for the original findspot/provenance
+\noindent \emph{I.Sicily} is actively generating rich geo-data for the individual inscriptions, both for the original findspot/provenance
 and the current location (whether museum-based, on-site, or elsewhere), and we aim to provide map-based searching for
@@
 information for each find-spot and the inscription’s current location, through a combination of library and map-based
-research and the use of autopsy and GIS recording. At present geo-data is being recorded in two forms, both through the
+research and the use of autopsy and \emph{GIS} recording. At present geo-data is being recorded in two forms, both through the
 use of explicit geographical locations in the form of longitude and latitude records in decimal degree form (using
-{\textless}geo{\textgreater} elements), and through the use of Pleiades URI references wherever possible.\footnote{
+\texttt{{\textless}geo{\textgreater}} elements), and through the use of Pleiades URI references wherever possible.\footnote{
 \url{http://pleiades.stoa.org/} [accessed 26.09.2015].} We are committed to the long-term use of Pleiades as our primary
 reference for ancient places, and to that end we aim to update and improve the Pleiades data for Sicilian locations, in
-particular name data and sub-locations, in conjunction with the editing of the I.Sicily records.\footnote{ See e.g.
+particular name data and sub-locations, in conjunction with the editing of the \emph{I.Sicily} records.\footnote{ See e.g.
 \citet{wilson_places:_2015}. Valeria Vitale (KCL) is currently undertaking a significant programme of data improvement in
-Pleiades on behalf of I.Sicily; we are grateful to Tom Elliott and Jeffrey Becker for their support.}
+Pleiades on behalf of \emph{I.Sicily}; we are grateful to Tom Elliott and Jeffrey Becker for their support.}
 
@@
 
-\noindent In order to support the aims outlined above, I.Sicily has taken a different approach to search and browse. Although
+\noindent In order to support the aims outlined above, \emph{I.Sicily} has taken a different approach to search and browse. Although
 standard form-based search with paged results, like that of Google, makes sense for very large result sets, the
-comparatively small number of records in I.Sicily (thousands versus the estimated 30 trillion web pages indexed in
+comparatively small number of records in \emph{I.Sicily} (thousands versus the estimated 30 trillion web pages indexed in
 Google) lends itself to a more direct and interactive approach - a spreadsheet/grid model (similar to Microsoft Excel)
@@
 accurately to more easily interpreted results. Further, any subset of the spreadsheet, generated from interactive
-filtering, can, with a single button push, be exported to CSV (comma separated values) for use outside I.Sicily. The
+filtering, can, with a single button push, be exported to CSV (comma separated values) for use outside \emph{I.Sicily}. The
 spreadsheet interacts particularly well with maps: all findspots or museums in a filtered subset of the grid can be
@@
 \noindent As was extensively discussed at the first EAGLE conference (Paris 2014), the creation and availability of translations
-is a major goal of the EAGLE project and its collaborators, and I.Sicily is no less committed to that
-ambition.\footnote{ See Orlandi et al. 2014: Part II.} Translations are rarely available for any of the published
-Sicilian inscriptions.\footnote{ French translations appear in Dubois 1989 and Dubois 2008.} It is obvious that the
+is a major goal of the EAGLE project and its collaborators, and \emph{I.Sicily} is no less committed to that
+ambition.\footnote{See \citet[Part II]{OrlandiCasarosa2014}.} Translations are rarely available for any of the published
+Sicilian inscriptions.\footnote{French translations appear in \citet{dubois_inscriptions_1989} and \citet{dubois_inscriptions_2008}.} It is obvious that the
 inclusion of translations will make the material much more accessible to a wider audience both of students and the
 general public. Equally, provision of translations will add to the value of the database as a resource for museums and
-others curating the inscriptions recorded in the database. To that end, a long-term ambition of I.Sicily is to include
+others curating the inscriptions recorded in the database. To that end, a long-term ambition of \emph{I.Sicily} is to include
 translations wherever possible in both English and Italian. We see this as one obvious area where public contribution
@@
 \caption{Screenshot of map-based searching (museum locations) and of part of the spreadsheet/grid search interface
-employed in I.Sicily}
+employed in \emph{I.Sicily}}
 \label{fig:2}
@@
 will be kept of editorial responsibility, autopsy and authorship as appropriate. In order to speed up the development
-of the corpus, and to encourage those working on the material to take ownership of it for themselves, we aim to enable
+of the \emph{corpus}, and to encourage those working on the material to take ownership of it for themselves, we aim to enable
 individuals to submit new records and emendations or additions to existing records (such as translations, images,
@@
 with those undertaking epigraphic projects in Sicily, and aim to offer the ability for other projects to publish their
-editions through I.Sicily.\footnote{ We are currently establishing just such a collaboration with Prof.ssa M. Sgarlata
+editions through \emph{I.Sicily}.\footnote{ We are currently establishing just such a collaboration with Prof.ssa M. Sgarlata
 and Drs Lo Faro and Gradante, in support of their project to produce the volume of Inscriptiones Christianae Italiae
-for the San Giovanni Catacombs in Siracusa.} We are also exploring the potential of the corpus as a teaching resource
+for the San Giovanni Catacombs in Siracusa.} We are also exploring the potential of the \emph{corpus} as a teaching resource
 both for epigraphy in general and for the teaching of EpiDoc. This latter aspect has already been initiated through a
@@
 
-It is our long-term ambition that I.Sicily might become the default location for the publication and dissemination of
+It is our long-term ambition that \emph{I.Sicily} might become the default location for the publication and dissemination of
 Sicilian inscriptions; in the shorter term, we hope that it will serve as valuable portal in the world of Sicilian
@@
 
-\noindent I.Sicily gratefully acknowledges the financial support of the John Fell Fund of the
+\noindent \emph{I.Sicily} gratefully acknowledges the financial support of the John Fell Fund of the
 University of Oxford, and of the Warden and Scholars of the House or College of Scholars of Merton in the University of

### 2016-01-21 source/31/EAGLE2016_I Sicily_final copy.tex
Paper 31 (Prag) edited and copied in the pdfs folder as EAGLE_2016_submission_31.pdf

@@
 confronted by the \emph{Corpus Inscriptionum Iudaeae/Palaestinae} (\emph{CIIP}), edited by H. Cotton et al.\footnote{ Original notice
-in \citet{cotton_corpus_1999}; presentation in \citet{cotton_corpus_2010}.} \emph{I.Sicily} sets out to follow in that mould, since the
+in \citet{cotton_corpus_1999}; presentation in \citet{Cotton:2007aa}.} \emph{I.Sicily} sets out to follow in that mould, since the
 different linguistic traditions of Sicily not only exist side-by-side but interact constantly throughout the island’s
@@
 traditional \emph{corpora} for the region, but both are over 125 years old and cover less than 30\% of the material now
-known).\footnote{\citet{_inscriptiones_1883}; \citet{kaibel_inscriptiones_1890}.} Existing online databases improve on this situation, but the results
+known).\footnote{\citet{Mommsen:1883aa}; \citet{kaibel_inscriptiones_1890}.} Existing online databases improve on this situation, but the results
 obtainable are of very varied value. The most comprehensive, in terms of the range of data recorded, is \emph{EDR} (with which

### 2016-01-21 source/32/EAGLE_2016_submission_32.tex
Paper 32 (Sgarlata) added, with dreadful bibliography (not checked)

@@
 \languageattribute{greek}{polutoniko}
+
+\usepackage{graphicx}
+\newlength{\xbreite}
+\settowidth{\xbreite}{x}
+\newcommand*{\chirho}{\scalebox{2}[1.2]{x}\hspace*{-1.25\xbreite}\scalebox{.5}[1.5]{\bfseries P}}
 
@@
 The renovated interest about funerary epigraphy generated new studies about new criteria of dating.
-The inscriptions dated by the indications of the ``consular couple'' represent an exceptions in the issues about chronological datings \citep[227]{FERRUA1947} \citep[nn.
-191-208]{FERRUA1989}: two were discovered during the excavation conducted by Paolo Orsi (1985) and to them is associated the latin epigraph ``Sporus'' (356) .
+The inscriptions dated by the indications of the ``consular couple'' represent an exceptions in the issues about chronological datings (\citealp{FERRUA1947}, nn.
+191-208; \citeyear[nn.
+191-208]{FERRUA1989}): two were discovered during the excavation conducted by Paolo Orsi (1985) and to them is associated the latin epigraph ``Sporus'' (356) .
 
@@
 
-\section{HISTORICAL ASPECTS OF THE CATACOMB OF S. GIOVANNI AT SYRACUSE}
+\section{Historical Aspects Of The Catacomb Of S. Giovanni At Syracuse}
 
@@
 Suburban cemeteries, fanned out from the area of Fusco, in the quarter of Neapolis, to the Santa Lucia area, in the southern part of Acradina, this indicates unequivocally what the perimeter of the city must have already been in the early and mid Roman Empire.
-The History of the area, which was going to hold the catacombs (San Giovanni, Vigna Cassia and Santa Lucia), spanned the centuries between the classical Greek and late antique ages, gradually giving evidence of quarries (Latomie), water supply systems to the city, characterized by cisterns and aqueducts \citep[682]{COLLINBOUFFIER1987}), handcraft workshops from the beginning of the 4th/3rd century BCE and burials datable to the early and mid Roman Empire.
+The History of the area, which was going to hold the catacombs (San Giovanni, Vigna Cassia and Santa Lucia), spanned the centuries between the classical Greek and late antique ages, gradually giving evidence of quarries (Latomie), water supply systems to the city, characterized by cisterns and aqueducts \citep[682]{COLLINBOUFFIER1987}, handcraft workshops from the beginning of the 4th/3rd century BCE and burials datable to the early and mid Roman Empire.
 The analysis of the funerary system certifies one hand the dependence on the Roman model, and other the debt in respect of local traditions.
@@
 Several interest will be given to structural aspect of the catacomb of S.
-Giovanni, practice of funeral rituals, ethnic and cultural fruition’s characters, transformation in the use, transformation in the way of using spaces for graves, to complete a general point of view about the phenomena of continuity and innovation as to previous sepulchral arrangements and, in the analysed periods, the facies belonging to the different settling, variegated in the committees’ ideological and religious themes, in choosing monumental types (like the rotundae) and decorations, in self-representative aspects, in burial uses.
+Giovanni, practice of funeral rituals, ethnic and cultural fruition’s characters, transformation in the use, transformation in the way of using spaces for graves, to complete a general point of view about the phenomena of continuity and innovation as to previous sepulchral arrangements and, in the analysed periods, the facies belonging to the different settling, variegated in the committees’ ideological and religious themes, in choosing monumental types (like the \emph{rotundae}) and decorations, in self-representative aspects, in burial uses.
 In this perspective we will give particular importance to the study of funerary epigraphy aimed at the writing of ICI Siracusa and the overall interpretation of the monument.
+
 Between 1893 and 1909 the archaeologist Paolo Orsi carried out a series of campaigns in the catacomb of San Giovanni.
-Detailed reports of those campaigns are recorded at various times in Notizie degli Scavi, which constitute an indefeasible starting point.
-The first incisive studies on Christian subterranean Sicily pertain to Joseph Führer, who dedicates many pages to the catacomb of San Giovanni, whose study is also epigraphic, is based on the previous literature, especially on Paolo Orsi’s discoveries \citep[13-39]{FUHRER1897}; \citep[22-26]{FUHRERSCHULTZE1907}; see also \citep[276, n.
-2]{ORSI1893}
+Detailed reports of those campaigns are recorded at various times in \emph{Notizie degli Scavi}, which constitute an indefeasible starting point.
+The first incisive studies on Christian subterranean Sicily pertain to Joseph Führer, who dedicates many pages to the catacomb of San Giovanni, whose study is also epigraphic, is based on the previous literature, especially on Paolo Orsi’s discoveries (\citealp[13-39]{FUHRER1897}; \citealp[22-26]{FUHRERSCHULTZE1907}; see also \citealp[276, n.
+2]{ORSI1893})
 %and \citep[60-61, fig. 5]{SCHULTZE1882} MISSING IN BIBLIOGRAPHY
@@
 .
-Both Antonio Ferrua and Santi Luigi Agnello within the space of a few years tried to solve the problems relating to the sarcophagus of Adelfia and the more general problem, strictly correlated with the ones just mentioned, of the cemetery genesis and development \citep{FERRUA1952} and \citep{AGNELLO1956}.
+Both Antonio Ferrua and Santi Luigi Agnello within the space of a few years tried to solve the problems relating to the sarcophagus of Adelfia and the more general problem, strictly correlated with the ones just mentioned, of the cemetery genesis and development \citep{FERRUA1952,AGNELLO1956}.
 It is necessary to record that numerous hypotheses on the genesis and development of the various sectors of the cemetery are ascribable to the account given by inscriptions.
@@
 
-\section{TOPOGRAPHIC AND EPIGRAFIC DATA TO STUDY THE CEMETERY}
+\section{Topographic And Epigrafic Data To Study The Cemetery}
 
 \noindent Walking across the main gallery, one can retrace the stages of Paolo Orsi’s interventions, being at the entrance of the Catacomb second northern gallery, before the so-called ``tomb of the Saint''.
-This arcosolium is considered a privileged burial on the basis of the following reasons: most of all because of its position, but also because it is a mensatype burial closed by a sole slab \citep[292-294]{ORSI1893}.
-The signs of an ancient rite are easily traceable on the slab, a rite that preceded the coming of Christianity and persisted for centuries up to the present day: the rite of refrigerium, which literally means refreshment, cooling \citep{GIUNTELLA1985}.
-In the Christian ceremony the purpose of a funeral banquet is to benefit the soul of the departed on the anniversary of death, a painful event celebrated as dies natalis of the soul to eternal life.
+This \emph{arcosolium} is considered a privileged burial on the basis of the following reasons: most of all because of its position, but also because it is a \emph{mensa}type burial closed by a sole slab \citep[292-294]{ORSI1893}.
+The signs of an ancient rite are easily traceable on the slab, a rite that preceded the coming of Christianity and persisted for centuries up to the present day: the rite of \emph{refrigerium}, which literally means refreshment, cooling \citep{GIUNTELLA1985}.
+In the Christian ceremony the purpose of a funeral banquet is to benefit the soul of the departed on the anniversary of death, a painful event celebrated as \emph{dies natalis} of the soul to eternal life.
 
@@
 The question is destined to remain unanswered and only an inscription found nearby could be a clue in this sense.
-The text says that the owners purchased the sepulchre close to the one of the bishop Cheperion’s \citep[507-508]{ORSI1895}; \citep[55-58]{RIZZONE2011}, of whom the scanty written sources never make mention.
+The text says that the owners purchased the sepulchre close to the one of the bishop \emph{Cheperion’s} (\citealp[507-508]{ORSI1895ab}; \citealp[55-58]{RIZZONE2011}), of whom the scanty written sources never make mention.
 In any case this constructed sarcophagus represent one of the most eminent burials of the catacomb.
@@
 
-This inscription, with a Christological monogram at the top, records Nassiana ``Christian, who competed with Penelope in moral virtue''.
-This inscription can only be compared with three examples in the Roman sepulchral carmina \citep{CLE}.
-The circular support of this inscription found close to the ``tomb of the Saint'' has been regarded as a mensa (table) for refrigerium rite \citep[47]{GIUNTELLA1985}, whose circular form would derive in any case from an evident reuse of a marble disk of classical craftsmanship, with a laurel wreath and berries sculpted on one side \citep[n.
-234]{ORSI1895}.
-A wall inscription painted on the extrados of an arcosolium in the third northern gallery of the catacomb could be connected with Nassiana’s text.
-Two lines of the inscription say that ``Sossa outdid (other women) in conjugal love; as for handiworks, without being taught by anyone, Athena herself had taught her how to do marvellous things'' \citep[n.
+This inscription, with a Christological monogram at the top, records \emph{Nassiana} ``Christian, who competed with Penelope in moral virtue''.
+This inscription can only be compared with three examples in the Roman sepulchral \emph{carmina} \citep{CLE}.
+The circular support of this inscription found close to the ``tomb of the Saint'' has been regarded as a \emph{mensa} (table) for \emph{refrigerium} rite \citep[47]{GIUNTELLA1985}, whose circular form would derive in any case from an evident reuse of a marble disk of classical craftsmanship, with a laurel wreath and berries sculpted on one side \citep[n.
+234]{ORSI1895ab}.
+A wall inscription painted on the extrados of an \emph{arcosolium} in the third northern gallery of the catacomb could be connected with Nassiana’s text.
+Two lines of the inscription say that ``Sossa outdid (other women) in conjugal love; as for handiworks, without being taught by anyone, Athena herself had taught her how to do marvelous things'' \citep[n.
 3]{FERRUA1940}.
@@
 
-It is evident that the catacomb of San Giovanni was originally a community cemetery, planned for only one type of burial: the arcosolium with multiple depositions, which does not require great care for decoration, only transennae.
+It is evident that the catacomb of San Giovanni was originally a community cemetery, planned for only one type of burial: the \emph{arcosolium} with multiple depositions, which does not require great care for decoration, only transennae.
 In the topographical and architectural development of the catacomb it appears clear that creating the rotundas breaks up the common burials series, destined to a socially homogeneous Christian community.
-These changes to the original plan – of the creation of subterranean mausolea both to the north and south of the cemetery – spring from the necessity to create appropriate spaces for the members of the Chruch and above all of the Empire, bringing into question the initial egalitarian choice of the arcosolium burials \citep[767]{GRIESHEIMER1989}.
+These changes to the original plan – of the creation of subterranean mausolea both to the north and south of the cemetery – spring from the necessity to create appropriate spaces for the members of the Chruch and above all of the Empire, bringing into question the initial egalitarian choice of the \emph{arcosolium} burials \citep[767]{GRIESHEIMER1989}.
 In the terminal part of the pious Giovanni’s gallery a monumental sarcophagus, once more hewn out of the rock, can be perceived.
@@
 As a rule this date marks the end of the dig works in this sector of the catacomb \citep[75-76]{FERRUA1952}.
-A arcosolium extrados, in the eastern region, shows traces of a palinsesto, an overlap of paintings and epigraphs than obstruct the identification of the first person buried in this sepulchre, even if recently Vittorio Rizzone proposed the interpretation Philadelpheia.\footnote{\citet[45-48]{RIZZONE2012}.} In one of the fossa tombs at the terminal stretch of the main gallery the inscription of Euterpe (IG XIV, 112) has been found, reused to cover the bottom of the tomb, which traditionally marks the end of the dig works of the catacomb \citep[75-76]{FERRUA1952}; \citep[79]{AGNELLO1958}; \citep[781]{GRIESHEIMER1989}.
+A arcosolium extrados, in the eastern region, shows traces of a palinsesto, an overlap of paintings and epigraphs than obstruct the identification of the first person buried in this sepulchre, even if recently Vittorio Rizzone proposed the interpretation \emph{Philadelpheia}\footnote{\citet[45-48]{RIZZONE2012}.}.
+In one of the fossa tombs at the terminal stretch of the main gallery the inscription of Euterpe (IG XIV, 112) has been found, reused to cover the bottom of the tomb, which traditionally marks the end of the dig works of the catacomb (\citealp[75-76]{FERRUA1952}; \citealp[79]{AGNELLO1958}; \citealp[781]{GRIESHEIMER1989}).
 Beyond the deceased’s biometric data, recorded as ``companion of the Muses'', the epigraph in the last three lines mentions the consuls’ iteration, which allow us to date back to the consulate of the Emperor Constantius, consul for the tenth time, and Julian- Caesar, consul for the third time.
@@
 
-In the southern region,, more than in other ones, that you can note the transformations that have profoundly undermined its community spirit, which had originally inspired the creation of the catacomb.
-In the first rotunda of the southern region, is a private space, which was given the name of Marina due to an inscription scratched upon the extrados of the arcosolium,on the right of the entrance to the short gallery of the bishop Siracosio.
-The arcosolium seems to be enframed by a painted prothyrum, as the still visible column and capital attest, confirming the generalized use of architectural elements in this catacomb, already seen in the rotunda of Antiochia.
+In the southern region, more than in other ones, that you can note the transformations that have profoundly undermined its community spirit, which had originally inspired the creation of the catacomb.
+In the first rotunda of the southern region, is a private space, which was given the name of Marina due to an inscription scratched upon the extrados of the \emph{arcosolium},on the right of the entrance to the short gallery of the bishop Siracosio.
+The \emph{arcosolium} seems to be enframed by a painted prothyrum, as the still visible column and capital attest, confirming the generalized use of architectural elements in this catacomb, already seen in the rotunda of Antiochia.
 According to the interpretation of the text, Marina could have been the wife of the patriciuset magister militum Sabinianus, sent by the Emperor Honorius to Spain at the time of barbarian invasions presumably between 409 and 423 \citep[21-22, 40]{FERRUA1989}.
@@
 .
-It is necessary to have a look at the slab of the presumed arcosolium of the bishop Siracosio, recorded in one inscription found in an adjacent fossa tomb, which says that the deceased intentionally purchased the sepulchre close to the one of the bishop just mentioned (IG XIV, 123: \begin{otherlanguage*}{greek}Ἐνθάδε κῖτε Πολυχ/ρόνιος καὶ Σεραπία.
-/ Ἠγόρασεν τῷ τότ/ε καιρῷ / Πολυχρονίου /αἱ Σεραπία ἐπὶ τῷ κυρί/ῳ μου ἐπισκόπῳ Συρα/κοσίῳ\end{otherlanguage*}).
-It is just a hypothesis seeing in the arcosolium with the engraved slab, still in situ, a noble burial for a member of the ecclesiastical hierarchy, for whom, in the absence of other data, according to a letter of pope Gelasius I, an episcopate between 492 and 496 has been proposed \citep[223]{NARCISO1952} \citep{CARLETTI2008}.
-One can clearly distinguish a Christogram with the apocalyptic letters alpha and omega and two ships in the shape of fish, regarded as making reference to the sacrament of the Eucharist, this is also suggested by the diskettes next to the fishes’ mouths, assimilable to loaves of bread \citep{SGARLATA2013}.
+It is necessary to have a look at the slab of the presumed \emph{arcosolium} of the bishop Siracosio, recorded in one inscription found in an adjacent fossa tomb, which says that the deceased intentionally purchased the sepulchre close to the one of the bishop just mentioned (\emph{IG} XIV, 123: \begin{otherlanguage*}{greek}Ἐνθάδε κῖτε Πολυχ/ρόνιος καὶ Σεραπία.
+/ Ἠγόρασεν τῷτότ/ε καιρῷ / Πολυχρονίου /αἱ Σεραπία ἐπὶ τῷ κυρί/ῳ μου ἐπισκόπῳ Συρα/κοσίῳ\end{otherlanguage*}).
+It is just a hypothesis seeing in the \emph{arcosolium} with the engraved slab, still in situ, a noble burial for a member of the ecclesiastical hierarchy, for whom, in the absence of other data, according to a letter of pope Gelasius I, an episcopate between 492 and 496 has been proposed (\citealp[223]{NARCISO1952}; \citealp{Carletti:2008aa}).
+One can clearly distinguish a Christogram with the apocalyptic letters \emph{alpha} and \emph{omega} and two ships in the shape of fish, regarded as making reference to the sacrament of the Eucharist, this is also suggested by the diskettes next to the fishes’ mouths, assimilable to loaves of bread \citep{Sgarlata2013}.
 
-After reading the text of the inscription upon the sarcophagus lid (CIL X, 7123: Ic Adelfia c(larissima) f(emina) / posita conpar / Balericomitis), one can become aware that it refers only to the woman, not to both husband and wife: here lies Adelfia, clarissima femina, wife of the count Valerius.
-Who were therefore Valerius and Adelfia?
+After reading the text of the inscription upon the sarcophagus lid (\emph{CIL} X, 7123: \emph{Ic Adelfia c(larissima) f(emina) / posita conpar / Balericomitis}), one can become aware that it refers only to the woman, not to both husband and wife: here lies \emph{Adelfia, clarissima femina}, wife of the count \emph{Valerius}.
+Who were therefore \emph{Valerius} and \emph{Adelfia}?
 The unsolved enigmas of the catacomb of San Giovanni deal with their names and most of all their identification.
 In reality it is possible to distinguish more than two phases of intervention, which have preceded the creation of the hole for the sarcophagus.
-At the beginning the internal space of the large niche was scenographically arranged in a terrace pattern with a forceful ascensional effect, not exempt from comparisons \citep[14, plate VII]{NESTORI1973}.The intact Latin inscription of Sporus, dated to 356, which sealed a forma in the main gallery among the ones that join the rotunda of Marina to the one of Adelfia \citep[24]{CAVALLARI 1872}; \citep[90]{AGNELLO1953}, would attest that the exploitation of the soil in the area gravitating to the rotunda of Adelfia was already underway after the first half of the 4th century, so confirming the evident anteriority of the six graves cut in the floor to the phase of monumentalization.
+At the beginning the internal space of the large niche was scenographically arranged in a terrace pattern with a forceful ascensional effect, not exempt from comparisons \citep[14, plate VII]{NESTORI1973}.The intact Latin inscription of Sporus, dated to 356, which sealed a forma in the main gallery among the ones that join the rotunda of Marina to the one of Adelfia (\citealp[24]{CAVALLARI1872}; \citealp[90]{AGNELLO1953}), would attest that the exploitation of the soil in the area gravitating to the rotunda of Adelfia was already underway after the first half of the 4th century, so confirming the evident anteriority of the six graves cut in the floor to the phase of monumentalization.
 Both the internal and the external organizations of the large niche (nicchione) would correspond to the first and second phases of intervention respectively, according to the current reconstruction.
@@
 
-The monument seems to have a less wild internal dynamics of development; the new data (archaeological, historical, epigraphical) \citep[101-108]{SGARLATA1996} allow the scholars to widen the syncopated weave of temporal sequences, to which the analysis of the monument has been pinned by the constant reference to Valerius Proculus’ chronology.
-A new chronology and a different identification of the comes Valerius can be proposed, considering: 1) the evident reuse of the sarcophagus; 2) the topographical development of the catacomb in the area where the sarcophagus has been discovered; 3) the type of monumental intervention, which followed the Roman counterparts, datable to the second half of the 4th century \citep[132-134]{FIOCCHINICOLAI 1997}.
-If Valerius were given a different identity, one could postpone him from the age of Constantine to that of Augustine \citep[15-51]{SGARLATA1998}, on the basis of several accounts about the friendship between the saint from Hippo and a comes Valerius, whose physiognomy, is somewhat vague.
+The monument seems to have a less wild internal dynamics of development; the new data (archaeological, historical, epigraphical) \citep[101-108]{SGARLATA1996} allow the scholars to widen the syncopated weave of temporal sequences, to which the analysis of the monument has been pinned by the constant reference to \emph{Valerius Proculus}’ chronology.
+A new chronology and a different identification of the comes Valerius can be proposed, considering: 1) the evident reuse of the sarcophagus; 2) the topographical development of the catacomb in the area where the sarcophagus has been discovered; 3) the type of monumental intervention, which followed the Roman counterparts, datable to the second half of the 4th century \citep[132-134]{FIOCCHINICOLAI1997}.
+If Valerius were given a different identity, one could postpone him from the age of Constantine to that of Augustine \citep[15-51]{SGARLATA1998}, on the basis of several accounts about the friendship between the saint from Hippo and a \emph{comes Valerius}, whose physiognomy, is somewhat vague.
 This friendship, confirmed by the epistles (AUG., Ep.
 200, 206, 207; Retr.
-II, 79 and 88) and the dedication, in 419, of the treatise De nuptiis et concupiscentia, is fed on the fight against Pelagianism, which in eastern Sicily had found a fertile soil.
-The presence of Pelagians in Sicily is attested for certain, according to Hilarius’ account under the pontificate of Innocent \citep[429-452]{PIETRI2000} and Honoreficentia’s letter, where a clarissima based at Syracuse is mentioned.
+II, 79 and 88) and the dedication, in 419, of the treatise \emph{De nuptiis et concupiscentia}, is fed on the fight against Pelagianism, which in eastern Sicily had found a fertile soil.
+The presence of Pelagians in Sicily is attested for certain, according to \emph{Hilarius}’ account under the pontificate of Innocent \citep[429-452]{PIETRI2000} and \emph{Honoreficentia}’s letter, where a \emph{clarissima} based at Syracuse is mentioned.
 On the island the spread of Pelagian movement looks like a direct consequence of the 410 sacking of Rome and the diaspora of the Roman nobility, of which Pelagius and Celestius were spiritual leaders; the short period they both spent in Sicily was not painless for Christian orthodoxy %\citep[436-437]{PIETRI2001} MISSING IN BIBLIOGRAPHY
- and particularly in Syracuse, as the cemeteries in the area overhanging the Greek theater, intended to serve the communities of the so-called heretics throughout the 5th century \citep{AGNELLO1990}.All the data gathered seems to lead to a different identification of Valerius – who, even if he was not Augustine’s correspondent, would be sought in the list of Valerii reported in the sources of the first quarter of the 5th century (PLRE II, 1143-44) – and a later chronology of the large niche monumental transformation compared to the one traditionally accepted.
+ and particularly in Syracuse, as the cemeteries in the area overhanging the Greek theater, intended to serve the communities of the so-called heretics throughout the 5th century \citep{AGNELLO1990}.All the data gathered seems to lead to a different identification of \emph{Valerius} – who, even if he was not Augustine’s correspondent, would be sought in the list of \emph{Valerii} reported in the sources of the first quarter of the 5th century (PLRE II, 1143-44) – and a later chronology of the large niche monumental transformation compared to the one traditionally accepted.
 
@@
 
-To answer the questions will take time, which will assure a greater credibility to what so far seems just a series of suggestions, most of all fed by the re-reading of epigraphic testimonies: the episodic character of the references to burials of bishops, presbyters, deacons and noting that the most significant percentage of evidences pertains to the members of the Church, buried in Syracuse away from their own countries and recorded in wall inscriptions written in Latin – Auxentius Hispanus episcopus and Superianus clerecus de Aquileia \citep[1 and 6]{FERRUA1940} – as a demonstration that the official language is used by foreigners, who were high clients (the remainder of inscriptions is in Greek), these and many other clues lead to thinking of a less incisive control of the Church in the 5th century than one can commonly believe.
+To answer the questions will take time, which will assure a greater credibility to what so far seems just a series of suggestions, most of all fed by the re-reading of epigraphic testimonies: the episodic character of the references to burials of bishops, presbyters, deacons and noting that the most significant percentage of evidences pertains to the members of the Church, buried in Syracuse away from their own countries and recorded in wall inscriptions written in Latin – \emph{Auxentius Hispanus episcopus} and \emph{Superianus clerecus de Aquileia} \citep[1 and 6]{FERRUA1940} – as a demonstration that the official language is used by foreigners, who were high clients (the remainder of inscriptions is in Greek), these and many other clues lead to thinking of a less incisive control of the Church in the 5th century than one can commonly believe.
 
@@
 What clues are the that could support the hypothesis that the clients belonged to a religious community?
-To tell the truth, they are scant and among them the epigraph of the blessed virgins Fotina and Filomena deserves to be recorded: the former lived 80 years, the latter 84 (IG XIV, 187); \citep[180]{FERRUA 1989}.
-The sole dated inscription found within this chamber, which is related to the name of Eucarpio, bearing the notification of both the consuls between 339 and 360 (\citep[3]{FERRUA1983}), cannot be used for dating; the discovery data, insisting on the fact that the gravestone was found overturned in a forma tomb \citep[30-31]{AGNELLO1960}, suggest that the gravestone was reused.
+To tell the truth, they are scant and among them the epigraph of the blessed virgins Fotina and Filomena deserves to be recorded: the former lived 80 years, the latter 84 (\emph{IG} XIV, 187; \citealp[180]{FERRUA1989}).
+The sole dated inscription found within this chamber, which is related to the name of Eucarpio, bearing the notification of both the consuls between 339 and 360 \citep[3]{FERRUA1983}, cannot be used for dating; the discovery data, insisting on the fact that the gravestone was found overturned in a \emph{forma} tomb \citep[30-31]{AGNELLO1960}, suggest that the gravestone was reused.
 
-The cubiculum of Eusebio has a structure different than other private spaces of the southern region of the catacomb.
-The name Eusebio derives, once more, from an inscription found on a three level monumental tomb, in the shape of exedra, visible to the left of the cubiculum (IG XIV, 111).
-The position and monumentality of the grand arcosolium, the formula ``of blessed memory'', the paleographical characters, the identity of pope Eusebius month of death (exiled to Sicily by Maxentius, Eusebius died on 17 August 309 or 310) have suggested to Isidoro Carini that this arcosolium could be the pontiff’s temporary sepulchre, whose bones were transferred to Rome and deposited in the catacomb of Callisto.
+The \emph{cubiculum} of Eusebio has a structure different than other private spaces of the southern region of the catacomb.
+The name Eusebio derives, once more, from an inscription found on a three level monumental tomb, in the shape of exedra, visible to the left of the \emph{cubiculum} (\emph{IG} XIV, 111).
+The position and monumentality of the grand \emph{arcosolium}, the formula ``of blessed memory'', the paleographical characters, the identity of pope Eusebius month of death (exiled to Sicily by Maxentius, Eusebius died on 17 August 309 or 310) have suggested to Isidoro Carini that this \emph{arcosolium} could be the pontiff’s temporary sepulchre, whose bones were transferred to Rome and deposited in the catacomb of Callisto.
 Despite the scholar’s efforts to sustain this theory \citep[134]{CARINI1873}, the physiognomic contours of Eusebius recorded in the inscription remain hazy.
-The cubiculum of Eusebio also deserves to be recorded for another testimony, which has a special value for Syracusans: an inscription found in a forma tomb, which testifies the worship of Saint Lucy, the patron of Syracuse..No decoration, no distinctive signs characterize the modest burial of Euskia \citep[299-308]{ORSI1895}, one of the many graves cut in the soil of the cubiculum \citep[526-528, fig.
+
+The cubiculum of Eusebio also deserves to be recorded for another testimony, which has a special value for Syracusans: an inscription found in a \emph{forma} tomb, which testifies the worship of Saint Lucy, the patron of Syracuse..No decoration, no distinctive signs characterize the modest burial of Euskia \citep[299-308]{ORSI1895ab}, one of the many graves cut in the soil of the cubiculum \citep[526-528, fig.
 164]{GUARDUCCI1978}.
@@
 \begin{otherlanguage*}{greek}
-\noindent Εὐσκία ἡ ἄμενπτος, ζήσα<σα> \\
+\noindent Εὐσκία ἡ ἄμενπτος, ζήσα$<$σα$>$ \\
 %Εὐσκία ἡ ἄμενπτος , ζήσα<σα> \\
@@
 % χρηστῶς καὶ σεμνῶς ἔτη \\ 
-πλῖο<ν> ἔλαττον κε᾽, ἀνε-\\
+πλῖο$<$ν$>$ ἔλαττον κε᾽, ἀνε-\\
 % πλῖο<ν>ἔλαττον κε' , ἀνε-\\
@@
 % δρὶ πολλὰς εὐχαρισ-\\
-τίασ α((chirho))ω εὐομε[ίλητος]
+τίασ α \textlatin{\chirho}{} ω εὐομε[ίλητος]
 %τίας αⳀω εὐομε[ίλητος].
@@
 She was Christian, faithful (and) perfect, well pleasing to her husband, endued with much grace, affable.
-This is the text of the most important Christian epigraph in Syracuse \citep[20]{AGNELLO1953}.The formulae present the typical elogium, the retrospective data of the life of the deceased and the Christological monogram flanked by the apocalyptic letters, which are common elements in the inscriptions of the catacombs; it is important since Euskia had been privileged to die on the same day that was sacred to Lucia, patron of Syracusans, a martyr during Diocletian persecutions on the 13th of December 304.
-The sole doubt of this reconstruction lurks in the term kyria, which is referred to Lucia in the epigraph; does one have to interpret it as synonym of haghia (saint), which would assure the official character of the worship, or a simple honorary title?
+
+This is the text of the most important Christian epigraph in Syracuse \citep[20]{AGNELLO1953}.The formulae present the typical \emph{elogium}, the retrospective data of the life of the deceased and the Christological monogram flanked by the apocalyptic letters, which are common elements in the inscriptions of the catacombs; it is important since Euskia had been privileged to die on the same day that was sacred to Lucia, patron of Syracusans, a martyr during Diocletian persecutions on the 13th of December 304.
+The sole doubt of this reconstruction lurks in the term \emph{kyria}, which is referred to Lucia in the epigraph; does one have to interpret it as synonym of \emph{haghia} (saint), which would assure the official character of the worship, or a simple honorary title?
 Whatever answer one can give, the importance of Euskia’s inscription survives intact, because it testifies if not already the sanctity of Lucia, the local devotion and worship of which the woman was subject in the 5th century in Syracuse.
+
 So this is the first attestation of the cult of Saint Lucia, which confirms the historicity of the Martyrologium Hieronymianum’s account on popular devotion to the Saint, which came through the celebration of a feast from the outset.
 All of the other evidence refers to successive periods.
-It is the Greek martyrion dated to the end of the 5th century, whose reliability has been debated for a long time and to this day never evidently established \citep[95-135]{MILAZZORIZZONERVO1988}.
+It is the Greek \emph{martyrion} dated to the end of the 5th century, whose reliability has been debated for a long time and to this day never evidently established \citep[95-135]{MILAZZORIZZONERVO1988}.
 The inscription, ascribable to the beginning of the 5th century, so would precede the questioned passio and would confirm the antiquity of the cult of Saint Lucia, whose bones were presumably kept in the homonymous catacomb in Syracusee, before George Maniakes, in 1039, transported them to Costantinople.
 Along with
-the inscription of Iulia Florentina from Catania \citep[608-610]{RIZZA1964}, the epigraph of Euskia is the most ancient Sicilian document that one could relate to the experience of martyrdom.
+the inscription of \emph{Iulia Florentina} from Catania \citep[608-610]{RIZZA1964}, the epigraph of Euskia is the most ancient Sicilian document that one could relate to the experience of martyrdom.
 
-\section{EPIGRAPHIC POPULATION: FORMULARIES AND IDENTITY-MAKING CHARACTERISTIES}
+\section{Epigraphic Population: Formularies And Identity-Making Characteristics}
 
@@
 One single epigraph can mention more than one person so the number of remembered deceased is superior than the number of inscriptions; starting from this number it’s possible to launch a demographic investigation to determine the expectation of life.
+
 In Syracuse all the social classes were affected with inexorable mortality; high percentage among young people (under 20) and extremely low percentage among old people (over 60).
@@
 We focus on the decesead elogistic formulary that help to increase the study about social aspects.
-In our research there are few examples of nuptial terminology, all referred to female deceased: compar, coniunx,
+In our research there are few examples of nuptial terminology, all referred to female deceased: \emph{compar, coniunx},
 \begin{otherlanguage*}{greek}σύμβιος, νυμφη\end{otherlanguage*}
@@
 Linked to this is the importance given to the sexual integrity of the dead person, in fact the concept of virginity in the nuptial context is particularly marked trhough the adjective virginius-a o \begin{otherlanguage*}{greek}παρθενικός-ή.\end{otherlanguage*}
-\footnote{See also \citet[129-130]{SGARLATA1991} \citet[493]{CARINI 1887}; CIL X, 7168, 2-4; \citet[5, 22, 44]{ORSI1893}; \citet[6, 7]{FUHRER1897}; \citep[21/29, 38, 39, 43, 45, 48, 58, 60, 69]{FERRUA1983}; IGCVO 127, 655, 778, 946,950, 1036, 1069 e 1366.
+\footnote{See also \citet[129-130]{SGARLATA1991} \citet[493]{CARINI1887}; CIL X, 7168, 2-4; \citet[5, 22, 44]{ORSI1893}; \citet[6, 7]{FUHRER1897}; \citep[21/29, 38, 39, 43, 45, 48, 58, 60, 69]{FERRUA1983}; IGCVO 127, 655, 778, 946,950, 1036, 1069 e 1366.
 Cat.
@@
 
-\section{CONCLUSIONS}
+\section{Conclusions}
 
 The new approach to the catacomb of San Giovanni brings many things into question, such as the fideistic attitude of those, who have studied this monument and have looked at the epigraphic evidence for chronological purpose.
-Rereading Orsi’s excavations data unequivocally shows how itinerant are the dated inscriptions within the cemetery – with the exception of three, whose discovery data attest their permanence in the original position \citep[43-50, 352-353]{ORSI1896}; \citep[90, 97]{AGNELLO1953} – which advises the scholars against using them to seal chronologically the diverse sectors.
+Rereading Orsi’s excavations data unequivocally shows how itinerant are the dated inscriptions within the cemetery – with the exception of three, whose discovery data attest their permanence in the original position (\citealp[43-50, 352-353]{ORSI1896}; \citealp[90, 97]{AGNELLO1953}) – which advises the scholars against using them to seal chronologically the diverse sectors.
 It would be profitable to draw a map of reuse, which is certainly the most striking phenomenon detected thus far in Orsi’s accounts, more than continuing to date the works in the galleries on the basis of dated inscriptions found in their terminal part.
@@
 It is likely that the non-Christian parts of the population, which was notable until the 4th century, were not buried in catacombs and their epithaphs have mostly perished'' \citep[124- 125]{KORHONEN2011}.
-Assuming this, one could state that in Sicily religious conversion is not a linguistic conversion, proving wrong the theory according to which in urban centers Christianization would bring along an early diffusion of Latin, whereas the pagus, keeping the use of Greek, would keep its distance from Christianity, at least until the early 5th century, during which signs of both linguistic and religious conversion would become a little more evident \citep[545]{MANGANARO1993}.Epigraphic documentation records, for Syracuse and its territory, as for Catania, a marked preponderance of the use of Greek still in the 5th century; so in urban centers Christianization does not bring an early diffusion of Latin, which was prerogative of high and foreign clients, as only the use of Greek in the pagus does not prove the extraneousness to the process of diffusion of the new creed still at the beginning of the 5th century.
+
+Assuming this, one could state that in Sicily religious conversion is not a linguistic conversion, proving wrong the theory according to which in urban centers Christianization would bring along an early diffusion of Latin, whereas the pagus, keeping the use of Greek, would keep its distance from Christianity, at least until the early 5th century, during which signs of both linguistic and religious conversion would become a little more evident \citep[545]{manganaro_greco_1993}.
+Epigraphic documentation records, for Syracuse and its territory, as for Catania, a marked preponderance of the use of Greek still in the 5th century; so in urban centers Christianization does not bring an early diffusion of Latin, which was prerogative of high and foreign clients, as only the use of Greek in the \emph{pagus} does not prove the extraneousness to the process of diffusion of the new creed still at the beginning of the 5th century.
 The idea according
@@
 In these centuries conventional superstitions were slowly relegated to periphery'' \citep{CRACCORUGGINI1998} needs to be dampened.
+
 The cemetery of San Giovanni has subtly given back, even if in a more hidden way, several testimonies of ideological commixture, which do not distinguish it from other communal cemeteries in Syracuse, where phenomena of ``religious dissidence'' are more readable \citep{SGARLATA2003}.
@@
 
+\nocite{Orsi:1896aa}
+
 \bibliographystyle{sapauth-eng}

### 2016-01-22 source/28/EAGLE2016_Rocco_engrev.tex
Paper 28 (Rocco) edited and added to the pdfs folder

@@
 \noindent The Epigraphic Database Bari (EDB) is an `old' database, it dates back to the late 1980s , when Carlo Carletti\footnote{
-He was professor of Epigrafia e antichità cristiane in the former Department of Classical and Christian Studies of Bari
+He was professor of \emph{Epigrafia e antichità cristiane} in the former \emph{Department of Classical and Christian Studies} of Bari
 University.}, inspired by Jory's experience indexing CIL VI\footnote{\citet{_corpus_1974}. The six volumes of the
-computerized KeyWord-In-Context index to all the approximately 40,000 texts collected in the sixth volume of the Corpus
-Inscriptionum Latinarum (CIL) are the outcome of a trailblazing work of arrangement and organization of the
-inscriptions in a database, even if limited to their textual part.}, started a project of digitization of the
+computerized KeyWord-In-Context index to all the approximately 40,000 texts collected in the sixth volume of the \emph{Corpus
+Inscriptionum Latinarum} (CIL) are the outcome of a trailblazing work of arrangement and organization of the
+inscriptions in a \emph{database}, even if limited to their textual part.}, started a project of digitization of the
 inscriptions commissioned by Christians from Rome between the third and eighth centuries, collected and edited in the
@@
 the use of Greek fonts; in order to add texts in the Greek alphabet it was necessary to type Latin equivalent letters
-in the MS Word Symbol font, as you can see in Fig.~\ref{fig:1}}: bibliographic data (Progressivo = ICVR Number), origin
+in the MS Word \emph{Symbol} font, as you can see in Fig.~\ref{fig:1}}: bibliographic data (Progressivo = ICVR Number), origin
 (Provenienza), type of support and technique of execution (Supp. e tecn.), function (Funzione), the presence or
@@
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img002.jpg} 
+ \includegraphics[scale=0.4]{EAGLE2016Roccoengrev-img002.jpg} 
 \caption{The inscription ICVR V, 15147 as it appears now in EDB, with the acronyms expanded, the image and the text
@@
 
-Adding metadata to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
+Adding \emph{metadata} to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
 the epigraphic object in its widest sense as an inscribed artifact. This feature, in particular relating to
@@
 It's even possible to define the relationship between the epigraph and the cited bibliography (printed or digital):
-identity, when it is an edition; integration, if it's the edition of another fragment of the same inscription;
-opistographic, if it's the edition of the inscription on the back side; reuse, if it's the edition of another
-inscription on the same support; comment, if it's a study on a related topic.
+\emph{identity}, when it is an edition; \emph{integration}, if it's the edition of another fragment of the same inscription;
+\emph{opistographic}, if it's the edition of the inscription on the back side; reuse, if it's the edition of another
+inscription on the same support; \emph{comment}, if it's a study on a related topic.
 
@@
 
-Data on the original context are therefore organized hierarchically in three related fields containing controlled lists.
+Data on the \emph{original context} are therefore organized hierarchically in three related fields containing controlled lists.
 After selecting the area of the suburb identified by the name of the consular road - or by the number of the Augustan
-regio for urban inscriptions - it's possible to select the monument from a list: a catacomb or part of it, if it is a
+\emph{regio} for urban inscriptions - it's possible to select the monument from a list: a catacomb or part of it, if it is a
 large and multi-layered one; a church; a public building or an urban area. The third field allows access to a further
@@
 Is worth noting that this set of associated fields refers to the original position of the inscriptions and not to the
-place where they have been found, unless the two data coincide, that is in the case of inscriptions in situ or suo loco
-adplicitae, in accordance with the definitions of the ICVR.
+place where they have been found, unless the two data coincide, that is in the case of inscriptions \emph{in situ} or \emph{suo loco
+adplicitae}, in accordance with the definitions of the ICVR.
 
 To complete information on spatial data, all cemeterial contexts have been georeferenced, so that clicking on their name
-opens a new window in Google Map that shows the modern entrance to the cemetery, with the address and its coordinates
+opens a new window in \emph{Google Map} that shows the modern entrance to the cemetery, with the address and its coordinates
 (Fig.~\ref{fig:6}).
@@
  \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
-\caption{The original context maps.}
+\caption{The \emph{original context} maps.}
 \label{fig:6}
@@
 
-As a case of study, for the inscriptions in situ pertaining to the Domitilla catacomb (Via Ardeatina) - the only
-cemeterial complex with a georeferenced plan of almost its entire extension\footnote{ START-Projekt: Die
-Domitilla-Katakombe in Rome (Institut für Kulturgeschichte der Antike - Österreichische Akademie der Wissenschaften).
+As a case of study, for the inscriptions \emph{in situ} pertaining to the Domitilla catacomb (Via Ardeatina) - the only
+cemeterial complex with a georeferenced plan of almost its entire extension\footnote{\emph{START-Projekt: Die
+Domitilla-Katakombe in Rome (Institut für Kulturgeschichte der Antike - Österreichische Akademie der Wissenschaften)}.
 \citet{orlandi_case_2014}.} - clicking on the alphanumeric code\footnote{ The codes are made up of a majuscule letter
 relating to a region of the catacomb \ and by another element, digit or minuscule letter, relating to a precise
-internal position, respectively a gallery or a cubiculum. } associated to the precise position of the inscriptions,
-gallery or cubiculum (F04, in Fig.~\ref{fig:6}), opens the plan of the specific area (regio) in which the inscription is found.
+internal position, respectively a gallery or a \emph{cubiculum}. } associated to the precise position of the inscriptions,
+gallery or \emph{cubiculum} (F04, in Fig.~\ref{fig:6}), opens the plan of the specific area (\emph{regio}) in which the inscription is found.
 In every plan, the inscriptions preserved in situ are placed and marked by ICVR and EDB number (Fig.~\ref{fig:6})\footnote{The
@@
 A survey of terminologies intended for description of epigraphic objects in the ICVR volumes has been carried out and
-has generated lists of controlled terms for some of the fields (Fig.~\ref{fig:8}). \ Type of support, Executing technique and
-Function vocabularies are aimed at classifying the specific and peculiar materials, methods and functions of the
+has generated lists of controlled terms for some of the fields (Fig.~\ref{fig:8}). \ \emph{Type of support}, \emph{Executing technique} and
+\emph{Function} vocabularies are aimed at classifying the specific and peculiar materials, methods and functions of the
 inscriptions encoded in EDB, as the traditional epigraphic taxonomies do not totally adhere to their features.
@@
 specific and peculiar issues of the inscriptions encoded in EDB. In particular the so-called
-{\textquotedbl}aberrant{\textquotedbl} forms are not {\textquotedbl}normalized{\textquotedbl} to the
-{\textquotedbl}standard{\textquotedbl} model, if they are recognized as grapho-phonetic outcomes of linguistic
+``aberrant'' forms are not ``normalized'' to the
+``standard'' model, if they are recognized as grapho-phonetic outcomes of linguistic
 modifications of Latin and Greek.
@@
 containing different spellings of a word. To resolve this issue each inscription is stored in its original form and in
-a “lemmatized” form, where each term is actually replaced with its corresponding lemma, possibly by taking into account
+a ``lemmatized'' form, where each term is actually replaced with its corresponding lemma, possibly by taking into account
 its inflexed forms\footnote{ On the contrary, if the compiler recognizes aberrant forms as outcomes of misstatements
@@
 as figurative elements (fishes, birds, anchors, etc.) and Christological monograms. Other fields record onomastic
-notes, critical apparatus and textual comments.
+notes, \emph{critical apparatus} and textual comments.
 
@@
 The latter of these can navigate in the descriptive section of the website (About EDB, People) and in the list of cited
-Publications. They also have access to the entire database using two research masks: a Quick search allows the user to
-search in only one of the following fields: identifier EDB, bibliographic data and text; an Advanced Search provides
+Publications. They also have access to the entire database using two research masks: a \emph{Quick search} allows the user to
+search in only one of the following fields: identifier EDB, bibliographic data and text; an \emph{Advanced Search} provides
 the opportunity to explore the database through multiple search criteria variously combined.
@@
 Like today's web search engines, EDB provides an advanced text search in Latin and Greek - an integrated tool
-facilitates writing in the Greek alphabet\footnote{ Greek Inputter 2, developed by J. Naughton, allows the user to
+facilitates writing in the Greek alphabet\footnote{ \emph{Greek Inputter 2}, developed by J. Naughton, allows the user to
 write in Greek using his/her usual keyboard and to easily type various Greek diacritical marks
@@
 
-\item With the same aim, the controlled lists of Type of support, Executing technique and Function have been integrated in
+\item With the same aim, the controlled lists of \emph{Type of support}, \emph{Executing technique} and \emph{Function} have been integrated in
 the corresponding vocabularies of the EAGLE community. Among other benefits, such as alignment and relations between
@@
 
-CIL = Corpus Inscriptionum Latinarum, \ Berolini 1863 ss.
-
-CIG = Corpus Inscriptionum Grecarum, I-IV, Berolini 1828-1877
-
-IC = Inscriptiones christiane urbis Romae septimo saeculo antiquiores, I-II, G.B. de Rossi (a c.), Romae 1857-1861;
-Supplementum, G. Gatti (a c.), Romae 1915.
-
-IG = Inscriptiones Graecae, Berolini 1873 ss.
-
-ICVR = Inscriptiones christianae urbis Romae septimo saeculo antiquiores. Nova series, I-X, A. Silvagni, A. Ferrua, D.
-Mazzoleni, C. Carletti (a c.), Romae - In Civitate Vaticana 1922 ss.
-
-IGVR = Inscriptiones grecae urbis Romae, I-IV, L. Moretti (a c.), Roma 1968-1990.
+\nocite{felle_fenomeni_2007,felle_esperienze_2012,jory_problems_1975}
 
@@
 
+\section*{Corpora}
+\begin{description}
+\item[ CIL ] = \emph{Corpus Inscriptionum Latinarum}, \ Berolini 1863 ss.
+\item[CIG ] = \emph{Corpus Inscriptionum Grecarum}, I-IV, Berolini 1828-1877
+\item[IC] = \emph{Inscriptiones christiane urbis Romae septimo saeculo antiquiores}, I-II, G.B. de Rossi (a c.), Romae 1857-1861; \emph{Supplementum}, G. Gatti (a c.), Romae 1915. 
+\item[IG] = \emph{Inscriptiones Graecae}, Berolini 1873 ss.
+\item[ICVR] = \emph{Inscriptiones christianae urbis Romae septimo saeculo antiquiores.
+Nova series}, I-X, A.
+Silvagni, A.
+Ferrua, D.
+Mazzoleni, C.
+Carletti (a c.), Romae - In Civitate Vaticana 1922 ss.
+\item[IGVR] = \emph{Inscriptiones grecae urbis Romae}, I-IV, L. Moretti (a c.), Roma 1968-1990.
+
+\end{description}
+
 \bibliographystyle{sapauth-eng}

### 2016-01-23 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
Paper 40 (Felle) edited

@@
 In these last years the amount of digital images of inscriptions increased very quickly: we do not need accurate textual
-descriptions of the so-called anaglypha, because we can directly see them. But we have to build a search-by-image,
+descriptions of the so-called \emph{anaglypha}, because we can directly see them. But we have to build a search-by-image,
 using photos and drawings but also tagging them with standardized - and shared - labels.
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!htbp]
 \centering
- \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img001.pdf}
+ \includegraphics[scale=0.5]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img000.png}
 \caption{Percentage in EDB of the ``illustrated inscriptions'' (chart by A. E. Felle).}
@@
 The second datum defined by the 2012 survey was the proportional decreasing of the use of different kind of images from
-the first decades of the IV century (the age of Constantine), when a huge and pervasive use of the so-called signa
-Christi - first of all the Chi-Rho monogram, with all its variations - prevails on all the other signs and figures
-[Felle 2013, 101-102].
+the first decades of the IV century (the age of Constantine), when a huge and pervasive use of the so-called \emph{signa
+Christi} - first of all the Chi-Rho monogram, with all its variations - prevails on all the other signs and figures
+\citep[101-102]{felle_apporto_2013}.
 
 Carlo Carletti explained this phenomenon as the result of a will to display explicit signs of a religious identity, such
-the Chi-Rho monogram is (\citet[68-72]{carletti_epigrafia_2008}; \citet[365-366]{felle_judaism_2007}). But we have to underline that the phenomenon is
+the Chi-Rho monogram is (\citealp[68-72]{carletti_epigrafia_2008}; \citealp[365-366]{felle_judaism_2007}). But we have to underline that the phenomenon is
 not exclusive of Christian patrons. We observe more and more recurrent similar ``signs of
 identity'' also among inscriptions commissioned by Jews, not only in Rome but also in other contexts in
-Late Antiquity world where they were [\citet[passim]{felle_judaism_2007}; \citet{pesce_documenti_2016}].
+Late Antiquity world where they were (\citealp[passim]{felle_judaism_2007}; \citealp{pesce_documenti_2016}).
 
@@
 \centering
- \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img001.pdf}
 \caption{Use of different images in inscriptions by Christians of Rome between III and IV cent. (chart by A. E. Felle)}
@@
 
-Since its conception, EDB recorded the presence and the different kinds of various signa Christi, both by a checkbox and
-in the text field, with standardized descriptions, with the aim to easily retrieve them in the database and to get
+Since its conception, EDB recorded the presence and the different kinds of various \emph{signa Christi}, both by a checkbox and
+in the text field, with \emph{standardized} descriptions, with the aim to easily retrieve them in the database and to get
 valuable results from structured queries about their recurrence.
 
-The results of the 2012 study were mainly of quantitative nature; today, by the existing large repositories of images in
-EDB – as like in general in EAGLE and in other similar epigraphic projects – we can improve the qualitative analysis of
+The results of the 2012 study were mainly of \emph{quantitative} nature; today, by the existing large repositories of images in
+EDB – as like in general in EAGLE and in other similar epigraphic projects – we can improve the \emph{qualitative} analysis of
 this and other features of Late Antique inscriptions.
 
-First of all, we have to say that in EDB we faced huge difficulties about the treatment of images other than signa
-Christi, or generally other non-alphabetical signs, or also captions related to figures on the slabs, and so on.
-Indeed, at the moment we are still not able to automatically obtain by EDB a structured index of the repertoire of the
+First of all, we have to say that in EDB we faced huge difficulties about the treatment of images other than \emph{signa
+Christi}, or generally other non-alphabetical signs, or also captions related to figures on the slabs, and so on.
+Indeed, at the moment we are still not able to automatically obtain by EDB a \emph{structured} index of the repertoire of the
 images. As in other epigraphic databases, in EDB they are recorded directly reporting their descriptions as like they
 are in printed editions (sometimes very old, as the first volumes of ICVR, for instance); there, in absence of
-pictures, the so-called anaglypha are concisely described by simplified and repetitive clichées to suggest the depicted
+pictures, the so-called \emph{anaglypha} are concisely described by simplified and repetitive \emph{clichées} to suggest the depicted
 subjects to the readers (Fig.~\ref{fig:3}) or by brief descriptions in Latin in transcriptions or also in the commentaries (Fig.~\ref{fig:4}).
@@
 \centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}
  \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
-\caption{Rome, coemeterium Hippolyti, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.12.14) and
+\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.12.14) and
 edition in ICVR, VII 19820.}
@@
   \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img005.png}
-\caption{Rome, coemeterium Maius, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.8.3) and edition in
+
+\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.8.3) and edition in
 ICVR, VIII 22407.}
@@
 Often, these descriptions are different although indicating the same subjects: in the ICVR, the reason of this
-disomogeneity is not only the longue durée of the realization of the corpus (seventy years, since 1922 to 1992, when
-the last published volume, the tenth, appeared) but also a refined textual variatio. Surely it can be appreciated in
+disomogeneity is not only the \emph{longue durée} of the realization of the corpus (seventy years, since 1922 to 1992, when
+the last published volume, the tenth, appeared) but also a refined textual \emph{variatio}. Surely it can be appreciated in
 printed editions but, for the aim of our digital archives, produces real difficulties.
 
-Some examples: in EDB different verbal descriptions about the some subject are recorded, such as avis uvam pascitur
-(e.g. ICVR, III 8114a, b, c, e), or avis uvas pascitur (ICVR, III 8004b); or also avis racemum carpens (ICVR, IX
-24020), avis racemum carpit (ICVR V, 14157), avis racemum pascitur (ICVR, V 15194), avis racemum rostro carpit (ICVR,
-IV 10934): it is not easy to perceive some difference). However, this variatio prevents right results in retrieving
+Some examples: in EDB different verbal descriptions about the some subject are recorded, such as \emph{avis uvam pascitur}
+(e.g. ICVR, III 8114a, b, c, e), or \emph{avis uvas pascitur} (ICVR, III 8004b); or also \emph{avis racemum carpens} (ICVR, IX
+24020), \emph{avis racemum carpit} (ICVR V, 14157), \emph{avis racemum pascitur} (ICVR, V 15194), \emph{avis racemum rostro carpit} (ICVR,
+IV 10934): it is not easy to perceive some difference). However, this \emph{variatio} prevents right results in retrieving
 data in our database.
@@
 Moreover, recording in EDB descriptions with different words for the same illustrated subject, such as
-``avis, racemus'' (EDB 19827: ICVR, III 9311, see Fig.~\ref{fig:5a}) or ``avis cum
-racemo'' (EDB 24933: ICVR, III 8044, see Fig.~\ref{fig:5b}) we are not able to retrieve all the occurrences of this
+``\emph{avis, racemus}'' (EDB 19827: ICVR, III 9311, see Fig.~\ref{fig:5}a) or ``\emph{avis cum
+racemo}'' (EDB 24933: ICVR, III 8044, see Fig.~\ref{fig:5}b) we are not able to retrieve all the occurrences of this
 same subject, because they are recorded (both n ICVR and in EDB) in different ways.
 
-\begin{figure}[!bp]
+\begin{figure}[!htp]
 \centering
+\begin{minipage}[b]{0.45\textwidth}
+%\begin{figure}
+%\centering
  \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img006.png}
-\caption{Rome, catacomb of Domitilla. Edition of ICVR, III 9311}
-\label{fig:5a}
+\begin{center}
+ \footnotesize(a)
+ \end{center}
+%\end{figure}
+\end{minipage}
+\begin{minipage}[b]{0.45\textwidth}
+%\begin{figure}
+%\centering
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img007.png}
+ \begin{center}
+ \footnotesize(b)
+ \end{center}
+%\end{figure}
+\end{minipage}
+
+\caption{Rome, catacomb of Domitilla. Edition of ICVR, III 9311 (a) and ICVR, III 8044 (b)}
+\label{fig:5}
 \end{figure}
 
-\begin{figure}[!bp]
-\centering
- \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img007.png}
-\caption{Rome, catacomb of Domitilla. Edition of ICVR, III 8044}
-\label{fig:5b}
-\end{figure}
+%\begin{figure}[!bp]
+%\centering
+%
+%\caption{Rome, catacomb of Domitilla. Edition of ICVR, III 8044}
+%\label{fig:5b}
+%\end{figure}
 
@@
 result of our queries: I think that we have to correct as soon as possible this ambiguity, in order to establish an
-unique way to describe the anaglypha.
+unique way to describe the \emph{anaglypha}.
 
@@
 
-In these last years - also with the kind help by the Photographic Archive of Papal Commission of Sacred Archaeology
-(\url{http://www.archeologiasacra.net/pcas-web/}) - the amount of images available in EDB increased very quickly: this is
+In these last years - also with the kind help by the Photographic Archive of Papal Commission of Sacred Archaeology\footnote{\url{http://www.archeologiasacra.net/pcas-web/}}
+ - the amount of images available in EDB increased very quickly: this is
 surely an advantage in respect to the situation of only three years ago. At the present day and more over in the
-future, we no longer need accurate textual descriptions of the anaglypha: we can directly see them. But, the
+future, we no longer need accurate textual descriptions of the \emph{anaglypha}: we can directly see them. But, the
 possibility to easily view a photo or a drawing of an inscription does not solve the issues related to search and to
@@
 mostly needs a logical, structured, hierarchically ordered taxonomy of all non textual elements defined by controlled,
-firm and shared vocabulary: a thesaurus imaginum\footnote{ Surely the experiences of other projects can be useful to
+firm and shared vocabulary: a \emph{thesaurus imaginum}\footnote{ Surely the experiences of other projects can be useful to
 this aim: I think for instance to the solutions presented during the VIth EAGLE International Event in Bari by the
-lecture offered by Rebecca Benefiel and Holly Sypniewski about the Ancient Graffiti of Herculaneum project.}, that can
+lecture offered by Rebecca Benefiel and Holly Sypniewski about the \emph{Ancient Graffiti of Herculaneum project}.}, that can
 be a suitable tool also to trail the activity of different stonecutters’ workshops that served – in the case of EDB
@@
 perspective - common in Christian epigraphy studies after all - to understand the communication power of this
-communicative medium. Indeed, the topic has been assumed as main theme of the last International Congress of Greek and
-Latin Epigraphy in Berlin in 2012 [Öffentlichkeit 2014]; very recently, a just published volume collects various essays
+communicative \emph{medium}. Indeed, the topic has been assumed as main theme of the last International Congress of Greek and
+Latin Epigraphy in Berlin in 2012 \citep{eck_vol_2014}; very recently, a just published volume collects various essays
 about this same topic just about the Late Antique, Medieval (both Christian and Islamic) world, where this notion of
-the inscriptions as 'written monuments' stands out with strong evidence [see Eastmond 2015].
+the inscriptions as 'written monuments' stands out with strong evidence \citep[see][]{eastmond_viewing_2015}.
 
-In our projects - we have to admit it - the notion that inscriptions are essentially texts is still largely
+In our projects - we have to admit it - the notion that inscriptions are \emph{essentially} texts is still largely
 prevailing: but now we can - consequently, we must - to increase our capability to represent, to record and so to
@@
 Österreichische Akademie der Wissenschaften and the Deutsches Archäologisches Institut) focused to the frescoes of the
-catacomb of Domitilla along via Ardeatina, the largest one in Rome suburb \citet{zimmermann_epigraphy_2014}. The positive
+catacomb of Domitilla along via Ardeatina, the largest one in Rome suburb \citep{zimmermann_epigraphy_2014}. The positive
 collaboration between our two projects continued: now we are able to offer to EDB users to view on the updated map of
-the catacomb of Domitilla the distribution of the inscriptions still in situ in the four levels of the subterranean
+the catacomb of Domitilla the distribution of the inscriptions still \emph{in situ} in the four levels of the subterranean
 cemetery: one can use them as reliable documents to (re-)consider the history of the complex, to confirm or to deny the
@@
 
-\noindent Materials and shape of the written objects communicate im-mediately, to all, before the inscribed texts. Because now we
+\noindent Materials and shape of the written objects communicate immediately, \emph{to all}, before the inscribed texts. Because now we
 can do it, we have to display to the users of our projects the real communicative power of the inscriptions, that
@@
 inscriptions {}- also lost or incomplete - conceived to be seen and read exactly there. An effective example can be
-the dedication by pope Damasus (366-384) to the martyr Ianuarius in the catacomb of Praetextatus on the via Appia (Fig.~\ref{fig:6}).
+the dedication by pope Damasus (366-384) to the martyr \emph{Ianuarius} in the catacomb of Praetextatus on the via Appia (Fig.~\ref{fig:6}).
 
@@
 \centering
- \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img008.jpg}
- \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img009.jpg}
-\caption{Rome, catacomb of Praetextatus. Damasus’ dedicatory inscription (ICVR, V 13871) for the martyr Ianuarius (Photo:
+ \includegraphics[scale=0.4]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img008.jpg}
+ \includegraphics[scale=0.3]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img009.jpg}
+\caption{Rome, catacomb of Praetextatus. Damasus’ dedicatory inscription (ICVR, V 13871) for the martyr \emph{Ianuarius} (Photo:
 PCAS; 3D rendering: G. De Felice)}
@@
 
-The bishop of Rome reaches his aim to capture the gaze also of illiterate faithfuls by placing a very large slab of
+The bishop of Rome reaches his aim to capture the gaze also of \emph{illiterate} faithfuls by placing a very large slab of
 white marble over the tomb of the martyr, well-lighted by a skylight made on purpose, in strong contrast to the
@@
 \centering
- \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}
+ \includegraphics[scale=0.06]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}
 \caption{Thessaloniki, city walls. Inscriptions in the masonry of a tower near Eptapyrgion (photo: A. E. Felle).}
@@
 There, the inscriptions are not carved in marble slabs or stone blocks, but they are realized with the same materials of
-the walls: simple bricks, but disposed to obtain letters and signs and symbols, visible also from afar. The
-inscriptions explain the walls; the walls speak its raisons d'être by the inscriptions, that are in different cases
-rich of abbreviations, closed to the reading but open to the sight: writing appears intrinsically significant. In my
+the walls: simple bricks, but disposed to obtain letters \emph{and} signs \emph{and} symbols, visible also from afar. The
+inscriptions explain the walls; the walls speak its \emph{raisons d'être} by the inscriptions, that are in different cases
+rich of abbreviations, closed to the reading but open to the sight: writing appears \emph{intrinsically} significant. In my
 opinion, this notion is clearly demonstrated by the unnecessary captions in the icons and in the images of martyrs
@@
 plating of the marble lintel over the Great Door of the Royal Gates of the Haghia Sophia in Constantinople, an empty
-throne is occupied by an open codex, according to the words by Kähler, ``the only extant plastic
+throne is occupied by an open \emph{codex}, according to the words by Kähler, ``the only extant plastic
 composition dating from the founding period of the church'' \citep[29-30; 32][taff. 22; 62]{kahler_hagia_1967}
@@
 
-On the open codex is inscribed a focused - but barely readable - quotation from John 10, verses 7 and 9, where Jesus
+On the open \emph{codex} is inscribed a focused - but barely readable - quotation from John 10, verses 7 and 9, where Jesus
 indicates himself as the gate:
@@
 Εἶπεν οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι ἐγώ εἰμι ἡ θύρα τῶν προβάτων.
-\end{otherlanguage*} (Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep);
+\end{otherlanguage*} (\emph{Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep});
 
@@
 ἐγώ εἰμι ἡ θύρα· δι' ἐμοῦ ἐάν τις εἰσέλθῃ σωθήσεται καὶ εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει
-\end{otherlanguage*}  ( 9 I am the gate;
-whoever enters through me will be saved. [a] They will come in and go out, and find pasture).
+\end{otherlanguage*}  ( $^\textit{9}$ \emph{I am the gate;
+whoever enters through me will be saved.} $^{[a]}$ \emph{They will come in and go out, and find pasture}).
 
@@
 \centering
+\begin{minipage}[c]{0.42\textwidth}
+%\begin{figure}
+%\centering
  \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img012.jpg}
+%\end{figure}
+\end{minipage}
+\begin{minipage}[c]{0.57\textwidth}
+%\begin{figure}
+%\centering
  \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img013.jpg}
+%\end{figure}
+\end{minipage}
 \caption{Istanbul, Hagia Sophia. On the left, the Royal Gates in the narthex. On the right, the particular of the image
@@
 ((crux)) \begin{otherlanguage*}{greek}
-εἶπεν ὁ κ(ύριο)ς | ἐγώ εἰμι | ἡ θύρα τῶν | προβάτων· | δι᾿ ἐμοῦ || ἐάν τις | εἰσέλθῃ | εἰσελεύσετ(αι) | κ(αὶ) ἐξελεύσετ(αι) | κ(αὶ) νομὴν | εὑρήσει.
+εἶπεν ὁ κ(ύριο)ς \textlatin{|} ἐγώ εἰμι \textlatin{|} ἡ θύρα τῶν | προβάτων· \textlatin{|} δι᾿ ἐμοῦ \textlatin{||} ἐάν τις \textlatin{|} εἰσέλθῃ \textlatin{|} εἰσελεύσετ(αι) \textlatin{|} κ(αὶ) ἐξελεύσετ(αι) \textlatin{|} κ(αὶ) νομὴν \textlatin{|} εὑρήσει.
 \end{otherlanguage*}

### 2016-01-23 source/43/EAGLE2016_Benefiel_Sypniewski.tex
Paper 43 (Benefiel et al.) edited and added also to the pdfs folder

@@
 \author[WL]{Sara Sprenkle}\date{2015-11-19}
-\address{Millsaps College, United States of America	}
-\address{Washington \& Lee University, United States of America}
+\address[WL]{Washington \& Lee University, United States of America}
+\address[Millisaps]{Millsaps College, United States of America	}
 \cortext[first]{Corresponding author. Email: benefielr@wlu.edu}	
@@
 and more than 6000 are known from Pompeii. We are contributing these inscriptions to the Epigraphic Database Roma
-(www.edr-edr.it), and are creating a linked resource, the Ancient Graffiti Project (ancientgraffiti.wlu.edu), that will
+(\url{www.edr-edr.it}), and are creating a linked resource, the Ancient Graffiti Project (\url{ancientgraffiti.wlu.edu}), that will
 allow users to conduct location-specific searches for graffiti.
@@
  \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img001.jpg}
-\caption{ Graffito from Pompeii [CIL 4, 5215]}
+\caption{ Graffito from Pompeii [\emph{CIL} 4, 5215]}
 \label{fig:1}
@@
 gladiatorial contest with athletes, musicians, and perhaps magistrates, sketched by hand on a funerary monument just
-outside the Porta Nocera of Pompeii (CIL 4, 10237; \citet[D31]{cooley_pompeii_2013}; cf. also CIL 4, 10236 and 10238 drawn
+outside the Porta Nocera of Pompeii (\emph{CIL} 4, 10237; \citet[D31]{cooley_pompeii_2013}; cf. also \emph{CIL} 4, 10236 and 10238 drawn
 nearby). \ More commonly, people made small sketches on the walls around them choosing from roughly a handful of
@@
 \noindent First, a little background on figural graffiti and our sources for this data. In Herculaneum, we are fortunate that a
-significant number of graffiti are still extant and in situ, as roofing has been reconstructed for many buildings to
+significant number of graffiti are still extant and \emph{in situ}, as roofing has been reconstructed for many buildings to
 protect them from the elements. Due to the fragile nature of wall plaster, however, especially in Pompeii, many
-graffiti that were recorded previously and published in CIL 4, have now been lost. Much of our data, therefore, comes
-from verbal descriptions of graffiti that have since disappeared. Furthermore, the different editors of CIL 4 and its
+graffiti that were recorded previously and published in \emph{CIL} 4, have now been lost. Much of our data, therefore, comes
+from verbal descriptions of graffiti that have since disappeared. Furthermore, the different editors of \emph{CIL} 4 and its
 supplements used different methods to denote that a drawing was present, and their practices changed over time. Working
@@
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img002.jpg}
-\caption{The entries of CIL 4, 2315-2316 and 2319, representing figural graffiti via brief description}
+ \includegraphics[scale=0.15]{EAGLE2016BenefielSypniewski-img002.jpg}
+\caption{The entries of \emph{CIL} 4, 2315-2316 and 2319, representing figural graffiti via brief description}
 \label{fig:2}
@@
 
-A drawing could, for example, be described in the text field of an entry in CIL 4. This occurs in the entries below,
-where three drawings of human heads [CIL 4, 2315-2316] and two drawings of gladiators [CIL 4, 2319] are described in
+A drawing could, for example, be described in the text field of an entry in \emph{CIL} 4. This occurs in the entries below,
+where three drawings of human heads [\emph{CIL} 4, 2315-2316] and two drawings of gladiators [\emph{CIL} 4, 2319] are described in
 small italics, placed where the images occur alongside the textual inscriptions [Fig.~\ref{fig:2}].
@@
 
-This practice is common in the original volume of CIL 4, when it seems that the editors documented textual and
+This practice is common in the original volume of \emph{CIL} 4, when it seems that the editors documented textual and
 figural graffiti that were in close proximity, or that were in some way related to each other. In later supplements,
@@
 illustrations, however, it also remained common practice to represent figural graffiti with very brief description in
-italics (e.g. CIL 4, 4822, 4823, 5264, 5275, 6624, 6672, 6889).
+italics (e.g. \emph{CIL} 4, 4822, 4823, 5264, 5275, 6624, 6672, 6889).
 
@@
 
-\noindent The most common strategy, however, for documenting figural graffiti in CIL 4 is by including brief mention of a drawing
+\noindent The most common strategy, however, for documenting figural graffiti in \emph{CIL} 4 is by including brief mention of a drawing
 in the editorial note that introduces a graffito or in the apparatus that follows it [Fig.~\ref{fig:3}].
@@
  \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img003.jpg}
-\caption{CIL entry for an alphabet near figural graffiti [CIL 4, 10711]}
+\caption{\emph{CIL} entry for an alphabet near figural graffiti [\emph{CIL} 4, 10711]}
 \label{fig:3}
@@
 
-Note that the editorial note above the entry mentions drawings nearby (novem galeas gladiatorias et parvum phallum), but
-the text is presented without illustration. This mode becomes more common in the fascicles of CIL 4 produced in the
-later twentieth century and so the figural graffiti from Herculaneum are usually represented this way (cf. CIL 4,
+Note that the editorial note above the entry mentions drawings nearby (\emph{novem galeas gladiatorias et parvum phallum}), but
+the text is presented without illustration. This mode becomes more common in the fascicles of \emph{CIL} 4 produced in the
+later twentieth century and so the figural graffiti from Herculaneum are usually represented this way (cf. \emph{CIL} 4,
 10532, 10568, 10586).
 
-\subsection{Figural graffiti omitted by CIL}
+\subsection{Figural graffiti omitted by \emph{CIL}}
 
-A fourth possibility exists as well, namely, when figural graffiti were not even mentioned in CIL. In each of the
-previous scenarios, the editors of CIL include a description of a figural graffito when it was close to a textual
+A fourth possibility exists as well, namely, when figural graffiti were not even mentioned in \emph{CIL}. In each of the
+previous scenarios, the editors of \emph{CIL} include a description of a figural graffito when it was close to a textual
 inscription. In contrast, figural graffiti found in isolation tended to be excluded altogether due to the focus of the
-Corpus on text. Fortunately there is now a useful resource devoted to figural graffiti: Martin Langner’s Antike
-Graffitizeichnungen, a monograph and accompanying database of figural graffiti from across the Mediterranean. His
+\emph{Corpus} on text. Fortunately there is now a useful resource devoted to figural graffiti: Martin Langner’s \emph{Antike
+Graffitizeichnungen}, a monograph and accompanying database of figural graffiti from across the Mediterranean. His
 catalog includes some 600 graffiti drawings from Pompeii and 60 from Herculaneum, including 200 that are not mentioned
-in CIL. In addition, whenever possible, Langner will provide a line-drawing of the graffito, either his own or one
-found in an earlier source; therefore, his database includes many line drawings that are not included in CIL even when
+in \emph{CIL}. In addition, whenever possible, Langner will provide a line-drawing of the graffito, either his own or one
+found in an earlier source; therefore, his database includes many line drawings that are not included in \emph{CIL} even when
 a drawing is described. However, certain motifs are omitted from Langner’s catalog. While he does catalog the more
-interesting Phalluskopfen examples, he generally omits simple drawings of phalli. He also leaves aside the decorative
-elements of coronae and palmae, which are sometimes mentioned in CIL. This means that an accurate total of all figural
-graffiti in Pompeii and Herculaneum can only be reached by working through the collections of both Langner and CIL. To
+interesting \emph{Phalluskopfen} examples, he generally omits simple drawings of \emph{phalli}. He also leaves aside the decorative
+elements of \emph{coronae} and \emph{palmae}, which are sometimes mentioned in \emph{CIL}. This means that an accurate total of all figural
+graffiti in Pompeii and Herculaneum can only be reached by working through the collections of both Langner and \emph{CIL}. To
 create the most comprehensive resource possible for figural graffiti, we include all known drawings in the AGP search
@@
 \noindent Since the verbal descriptions of figural graffiti provided by legacy data are limited and vague or exceedingly general
-(e.g. caput), the best circumstance under which to digitize a graffito is when the drawing itself still remains extant.
+(e.g. \emph{caput}), the best circumstance under which to digitize a graffito is when the drawing itself still remains extant.
  In such cases, we will use any published data as a starting point, but we are also able to make our own editorial
@@
 \begin{enumerate}
-\item How to define an entry? (Where for example does one entry stop and another begin? Do we catalog series or clusters
+\item \textbf{How to define an entry?} (Where for example does one entry stop and another begin? Do we catalog series or clusters
 of graffiti, or individual images? How do we account for or represent the larger context?)
-\item How to describe a drawing? (Here, there arise issues both of standardization and of interpretation, or over
+\item \textbf{How to describe a drawing?} (Here, there arise issues both of standardization and of interpretation, or over
 interpretation.)
-\item How can we make drawings searchable? (Ideally, we would like to make it possible for users both to browse and to
+\item \textbf{How can we make drawings searchable?} (Ideally, we would like to make it possible for users both to browse and to
 locate specific images.)
@@
 consider whether or not multiple elements should be part of the same EDR record or should be given separate entries.
-First, we must ask: can we be assured that certain elements were meant to be understood together? There might be an
+First, we must ask: \emph{can we be assured} that certain elements were meant to be understood together? There might be an
 issue of accretion or accumulation, where additional graffiti have been added subsequently. A related challenge is
@@
  \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img004.jpg}
-\caption{Line-drawing displaying a collection of textual and figural graffiti [CIL 4, 8383-8386; EDR148730]; unpublished
-sketch of Matteo Della Corte from the archives of the Corpus Inscriptionum Latinarum at the Berlin-Brandenburg Academy
+\caption{Line-drawing displaying a collection of textual and figural graffiti [\emph{CIL} 4, 8383-8386; EDR148730]; unpublished
+sketch of Matteo Della Corte from the archives of the \emph{Corpus Inscriptionum Latinarum} at the Berlin-Brandenburg Academy
 of Science and Humanities.}
@@
 You can see a number of different images here including a small gladiator with trident, a face in profile, leaves,
-several animals, and geometric shapes as well as the name ``Atini'' and the greeting ``\textgreek{Gelast`h qa~ire.''
-}\ Such a collection raises many interpretive questions. What is the relationship, if any, between the figural and the
+several animals, and geometric shapes as well as the name ``Atini'' and the greeting ``\textgreek{Γελαστὴ χαῖρε}.'' Such a collection raises many interpretive questions.
+What is the relationship, if any, between the figural and the
 textual graffiti? How should that relationship be best represented? Fortunately, we have available with this sketch an
-overall view of the spatial relationships of this group of graffiti. Because the CIL entries are focused on text, the
+overall view of the spatial relationships of this group of graffiti. Because the \emph{CIL} entries are focused on text, the
 figural graffiti are associated with and described in the entry of nearby textual inscriptions. For this cluster, we
@@
 
-In an example from Herculaneum (Fig.~\ref{fig:3}, above), the entry for CIL 4, 10711 notes that in addition to a graffito of the
+In an example from Herculaneum (Fig.~\ref{fig:3}, above), the entry for \emph{CIL} 4, 10711 notes that in addition to a graffito of the
 alphabet, a series of nine gladiator helmets and a small phallus were also drawn on the wall. During our field season
@@
 individual records, we have a unique identifier for each image, in the form of the EDR number, so that users can cite a
-specific parallel precisely. \ Again, we can record the precise measurements for each image. Yet, since separating
+\emph{specific} parallel precisely. \ Again, we can record the precise measurements for each image. Yet, since separating
 each image can obscure how the images relate to one another in the group, as with the previous example, we also upload
@@
 relate. More often, we are left with only legacy data, with brief mention of a figural graffito in the apparatus of a
-CIL entry and without illustration. \ Yet, proximity does not always indicate a relationship between the text and
+\emph{CIL} entry and without illustration. \ Yet, proximity does not always indicate a relationship between the text and
 image. \ Indeed, there may be no relationship at all between the figural and textual graffiti; therefore, putting the
@@
 A second challenge occurs when we must decide how much interpretation to offer when we describe a graffito for a
-database entry. \ When CIL has included mention of a drawing, we generally incorporate that description directly into
+database entry. \ When \emph{CIL} has included mention of a drawing, we generally incorporate that description directly into
 our entry. With figural graffiti documented by Martin Langner, we must create a summary in Latin and when doing so, we
@@
 With this camel \citep[n.
-1443]{langner_antike_2001}, for instance, we offer a full description in Latin that accounts for all the features of the drawing: camelus
-dromedarius cum cauda, lodicem gerens, ad dextram incedens [Fig.~\ref{fig:5}].
+1443]{langner_antike_2001}, for instance, we offer a full description in Latin that accounts for all the features of the drawing: \emph{camelus dromedarius cum cauda, lodicem gerens, ad dextram incedens} [Fig.~\ref{fig:5}].
 
@@
  \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img005.jpg}
-\caption{Example of an entry for a figural graffito not found in CIL (EDR144514)}
+\caption{Example of an entry for a figural graffito not found in \emph{CIL} (EDR144514)}
 \label{fig:5}
@@
 As one might imagine, issues of interpretation can arise even with simple descriptions of images. In truth, we have
-encountered more difficulties with interpretation in the case of drawings that have been described by CIL. The first
-example comes from a shop in Pompeii and represents text and a drawing [CIL 4, 8185]. \ The plaster has clearly broken
+encountered more difficulties with interpretation in the case of drawings that have been described by \emph{CIL}. The first
+example comes from a shop in Pompeii and represents text and a drawing [\emph{CIL} 4, 8185]. \ The plaster has clearly broken
 off, so we do not know if this was part of a larger scene. What remains are two lines of text and just one figure,
-which would seem to be a drawing of a person facing forward and rendered with head and shoulders. CIL describes it
-thus: herma muliebris prospiciens [Fig.~\ref{fig:6}].
+which would seem to be a drawing of a person facing forward and rendered with head and shoulders. \emph{CIL} describes it
+thus: \emph{herma \underline{muliebris} prospiciens} [Fig.~\ref{fig:6}].
 
@@
  \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img006.jpg}
-\caption{CIL entry for textual and figural graffiti [CIL 4, 8185]}
+\caption{\emph{CIL} entry for textual and figural graffiti [\emph{CIL} 4, 8185]}
 \label{fig:6}
@@
 Is this drawing clearly depicting a female? It is difficult to argue from either the hairstyle or the clothing that the
-figure is female. Here, we can only assume that the editors of CIL identified the image as female because the text
+figure is female. Here, we can only assume that the editors of \emph{CIL} identified the image as female because the text
 above mentions the female name Fortunata. But are we sure the image is meant to illustrate the text? Or that the image
@@
 
-The head of a woman, described with a textual graffito from the Suburban Baths in Herculaneum [CIL 4, 10676], raises
-similar problems with verbal descriptions of figural graffiti. In this case, CIL does not reproduce an image of the
-sketch; it only notes that the four-line inscription of CIL 4, 10676 appears below a drawing of a female (infra
-mulieris imaginem). In this instance, too, the text nearby includes two names, one female and one male. The drawing has
-appeared in multiple publications (\citet{della_corte_loves_1960}, \citet{deiss_herculaneum_1989}, \citet{vivolo_pompei:_1993}) [Fig.~\ref{fig:7}].
+The head of a woman, described with a textual graffito from the Suburban Baths in Herculaneum [\emph{CIL} 4, 10676], raises
+similar problems with verbal descriptions of figural graffiti. In this case, \emph{CIL} does not reproduce an image of the
+sketch; it only notes that the four-line inscription of \emph{CIL} 4, 10676 appears below a drawing of a female (\emph{infra
+mulieris imaginem}). In this instance, too, the text nearby includes two names, one female and one male. The drawing has
+appeared in multiple publications \citep{della_corte_loves_1960,deiss_herculaneum_1989,vivolo_pompei:_1993} [Fig.~\ref{fig:7}].
 
@@
  \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img007.jpg}
-\caption{Apograph of figural graffito referred to in the note at CIL 4, 10676 \citet[n. 309]{langner_antike_2001}.}
+\caption{Apograph of figural graffito referred to in the note at \emph{CIL} 4, 10676 \citet[n. 309]{langner_antike_2001}.}
 \label{fig:7}
@@
 Again, we might question whether this figure should be identified as female. In fact, we not certain that we had located
-the correct apograph for the drawing. There is considerable discrepancy between the description of the drawing in CIL
+the correct apograph for the drawing. There is considerable discrepancy between the description of the drawing in \emph{CIL}
 4, 10676 as female, with no mention of the long nose, and this line-drawing.
@@
 Another reason we suspected there might be a mistake was that Martin Langner had catalogued the drawing associated with
-CIL 4, 10676, describing it as a ``Phalluskopf.'' There was no mention of gender. And he categorized this drawing among
+\emph{CIL} 4, 10676, describing it as a ``Phalluskopf.'' There was no mention of gender. And he categorized this drawing among
 several examples of drawings of heads with phallic features. The graffito is in a room that is sealed off, with no
@@
 recent two-volume work providing images of extant ancient graffiti \citep[509]{varone_titulorum_2012} allowed us to confirm that this
-is indeed the correct graffito drawing – somewhat above but also drawn partly through the text of CIL 4, 10676.
+is indeed the correct graffito drawing – somewhat above but also drawn partly \emph{through} the text of \emph{CIL} 4, 10676.
 
-Neither description offered by CIL or by Langner, however, seems altogether satisfactory. There are no obvious markers
-of female identity and even the description of Phalluskopf is less than transparent. Thus this one drawing has two
+Neither description offered by \emph{CIL} or by Langner, however, seems altogether satisfactory. There are no obvious markers
+of female identity and even the description of \emph{Phalluskopf} is less than transparent. Thus this one drawing has two
 published descriptions that vary greatly and that each lead to a very different understanding of the graffito. What
-should we then do in such situations? Do we repeat the identification of CIL? Or do we offer a less specific
-description, merely labeling this a hominis figura? In the end, our solution is to offer our own description, which
+should we then do in such situations? Do we repeat the identification of \emph{CIL}? Or do we offer a less specific
+description, merely labeling this a \emph{hominis figura}? In the end, our solution is to offer our own description, which
 is detailed but less interpretative, with an emphasis on specific features of the image that are readily identifiable.
-We also document Langner’s description and CIL’s earlier identification in our entry for EDR, but we note our
-hesitation with such identification by labeling the image: gryllus? (``caricature?''). We are aware, however, that we
+We also document Langner’s description and \emph{CIL}’s earlier identification in our entry for EDR, but we note our
+hesitation with such identification by labeling the image: \emph{gryllus}? (``caricature?''). We are aware, however, that we
 also introduce an interpretation with the tentative suggestion this drawing might be a caricature.
@@
 any type of animal with four legs (stags, boars, dogs). In such cases, our solution is to describe a drawing with more
-generic, yet accurate, terms such as ``gladiator,'' without further specification, or ``animal'' rather than cervus, aper,
-or canis. Similarly, if we cannot determine male or female, we prefer to describe the drawing as ``facies hominis.'' In
+generic, yet accurate, terms such as ``gladiator,'' without further specification, or ``animal'' rather than \emph{cervus}, \emph{aper},
+or \emph{canis}. Similarly, if we cannot determine male or female, we prefer to describe the drawing as ``\emph{facies hominis}.'' In
 the AGP search engine we can then indicate possible but not certain identification with a descriptor, or tag, that
@@
 would need to need to know the vocabulary used to describe the drawing. Would someone ever think to search for
-``camelus'' without prior knowledge that there is a figural graffito of a camel in Herculaneum? Similarly, if you
-search for ``gladiator'', the text field will give you results for all inscriptions that mention gladiators as well as
+``\emph{camelus}'' without prior knowledge that there is a figural graffito of a camel in Herculaneum? Similarly, if you
+search for ``\emph{gladiator}'', the text field will give you results for all inscriptions that mention gladiators as well as
 drawings where we have described gladiators. If, however, we’ve described the gladiator more specifically as a
-``retiarius'' or we have gladiatorial equipment, such drawings will be omitted from the list of search results.
+``\emph{retiarius}'' or we have gladiatorial equipment, such drawings will be omitted from the list of search results.
 
@@
 ultimately to enable searches by these tags as well, so a user can directly find all drawings with dogs. The search
-capacity will allow a user to search the tags or the Latin description, so both ``navis'' and ``boat'' will return hits.
+capacity will allow a user to search the tags or the Latin description, so both ``\emph{navis}'' and ``boat'' will return hits.
 Again, standardization is necessary. We are currently developing a list of tags that is comprehensive and flexible

### 2016-01-24 source/CNR/TheEAGLEDataAggregatorDataQualityMonitoring.tex
finished posters and committee, compiled the all lot
I have Moved all the compiled articles in the pdfs folder and compiled
with all bits and pieces
Paper CNR (Mannocci et al) edited and added to the pdfs folder

@@
 \author[CNR-ISTI]{Andrea Mannocci\corref{all}}
-\author[CNR-ISTI]{Vittore Casarosa}
-\author[CNR-ISTI]{Paolo Manghi}
-\author[CNR-ISTI]{Franco Zoppi}
+\author[CNR-ISTI]{Vittore Casarosa\corref{all}}
+\author[CNR-ISTI]{Paolo Manghi\corref{all}}
+\author[CNR-ISTI]{Franco Zoppi\corref{all}}
 
 %Additional information about corresponding author, addresses and affiliations:
-\cortext[all]{\{andrea.mannocci, vittore.casarosa, paolo.manghi, franco.zoppi\}@isti.cnr.it
+\cortext[all]{Corresponding authors: \{andrea.mannocci, vittore.casarosa, paolo.manghi, franco.zoppi\}@isti.cnr.it
 }
@@
 
-The EAGLE project (fully described in \citep{eagle}) aggregates epigraphy related content from about 20 different data providers (cultural institutions all over Europe), and makes its content available to both Europeana \citep{europeana} (through an OAI-PMH interface) and to scholars and the general public through a Web portal.
+The EAGLE project \citep[fully described in][]{eagle} aggregates epigraphy related content from about 20 different data providers (cultural institutions all over Europe), and makes its content available to both Europeana \citep{europeana} (through an OAI-PMH interface) and to scholars and the general public through a Web portal.
 
@@
 The interaction between the IQM and the Monitoring Framework is done through a user friendly user interface (WebUI) provided by the DFDS.
-In addition to providing easy access to the observation time series, the WebUI provides the facilities to define the overall Monitoring Scenario, which will include the sensors and the controls.
+In addition to providing easy access to the observation time series, the WebUI provides the facilities to define the overall Monitoring Scenario, which will include the sensors and the \emph{controls}.
 
@@
 \bibliographystyle{sapauth-eng}
-\bibliography{EAGLE}
+\bibliography{../../EAGLE}
 

### 2016-01-24 source/18/EAGLE2016_Full_Paper_Broux-XeLaTEX.tex
Paper 18 (Broux) fixed with the Coptic and Egyptian fonts (using XeLaTeX); file EAGLE2016_Full_Paper_Broux-XeLaTEX.tex is now the master file. The compiled version (Broux-XeLaTEX.pdf) updated to the pdfs folder as submission_18.pdf

--- 2016-01-20 source/18/EAGLE2016_Full_Paper_Broux.tex
@@
 % !TEX encoding = UTF-8 Unicode
-% !TEX program = pdflatex
+% !TEX TS-program = pdflatex
 % !TEX spellcheck = en_US
@@
 % if you write in italian
-%\setmainlanguage{italian}
+%\setmainlanguage{english}
 % If you want put some ancient greek:
@@
 
+\usepackage{fontspec}
+
+\usepackage{polyglossia}
+
+\setmainlanguage{english}
+\setotherlanguage{greek}
+
+\newfontfamily\greekfont[Script=Greek]{CMU Serif}
+
+%\setmainfont{lmr}
 %%%%%%%%
@@
 \maketitle
+
 \begin{abstract}
@@
 Graeco-Roman Egypt' (KU Leuven OT project 2008-2012), on a corpus of about 50,000 papyri and ostraca in the Duke
-Databank of Documentary Papyri (papyri.info). With the additional support of a Hercules Grant, (`An Interdisciplinary
+Databank of Documentary Papyri (\url{papyri.info}). With the additional support of a Hercules Grant, (`An Interdisciplinary
 Database of Proper Names in Late Pharaonic, Graeco-Roman and Byzantine Egypt (ca. 800 BC – AD 640)'; 2008-2014) the
@@
 with the Ptolemaic period, when this form of polyonymy served to cross ethnic borders \citep{Coussement}, the
-other with the Roman period, when the practice was adapted by the local elite to distinguish themselves from the hoi
-polloi and to resemble Roman nomenclature \citep{Broux2015a}. The data of the fourth century AD provided new insights into
+other with the Roman period, when the practice was adapted by the local elite to distinguish themselves from the \emph{hoi
+polloi} and to resemble Roman nomenclature \citep{Broux2015a}. The data of the fourth century AD provided new insights into
 the spread of Christianity in Egypt \citep{Depauw2013}. The development of a standardized identification
@@
 connected to a set of variants in all possible languages. As a rule, only very minor dialectal or orthographical
-variation is allowed in the `native' language, in this case Greek (e.g. \textgreek{E>is'idwroc }and
-\textgreek{>Is'idoroc). }Many of the variants are actually created by renderings of a name in other languages, e.g.
-Ỉsytrs, ȝysydwrs or ⲓⲥⲓⲇⲱⲣⲉ in Egyptian and Coptic. In all there
+variation is allowed in the `native' language, in this case Greek (e.g.
+\textgreek{Εἰσίδωρος} and \textgreek{Ἰσίδορος}).
+Many of the variants are actually created by renderings of a name in other languages, e.g.
+%{\fontspec{Liberation Sans}ȝysydwrs}
+Ỉsytrs, {\fontspec{Gentium}ȝysydwrs} or {\fontspec{ArialCoptic} ⲓⲥⲓⲇⲱⲣⲉ} in Egyptian and Coptic. In all there
 are 177,290 variants in the NAMVAR database. Finally, for each of the variants, the declined forms were created, e.g.
-\textgreek{>Is'idwrou (}genitive) or \textgreek{>Is'idwrwi (}dative). This NAMVARCASE database is the largest with
+\textgreek{Ἰσίδωρου}) or \textgreek{Ἰσίδωρωι} (dative). This NAMVARCASE database is the largest with
 667,677 entries, and this set is used as a gazetteer for NER. It was developed on the basis of the set of names of some

### 2016-01-24 source/18/EAGLE2016_Full_Paper_Broux-XeLaTEX.tex
Some overfull hboxes fixed on paper 18 (Broux)

@@
 \usepackage{polyglossia}
+\usepackage{ucharclasses}
+\usepackage{microtype}
 
@@
 \newfontfamily\greekfont[Script=Greek]{CMU Serif}
+
+%\begin{hyphenrules}{english}
+%\hyphenation{na-mes/in-di-vi-du-als}
+%\end{hyphenrules}
 
@@
 fields, isolated from the `classical' world. New partnerships for Etruscan have been set up, and soon also for Punic
-(CIP: \url{cip.cchs.csic.es}) and South-Arabian (DASI: \url{dasi.humnet.unipi.it}), while for the Italic languages, for Gallic,
+(CIP: \url{http://cip.cchs.csic.es/}) and South-Arabian (DASI: \url{http://dasi.humnet.unipi.it/}), while for the Italic languages, for Gallic,
 Lepontic, Venetic, and Messapian information has already been digitized on the basis of existing corpora.
@@
 applied to the Latin inscriptions incorporated in Trismegistos so far. This will result in the addition of an estimated
-500,000 or more new references to names/individuals to Trismegistos People. Network analysis will be used to help with
+500,000 or more new references to names / individuals to Trismegistos People. Network analysis will be used to help with
 the disambiguation of individuals.

### 2016-01-25 panels/EAGLE_2016_submission_47.tex
added panels
copying from pdf panel presentations are horrible, especially the first
one.

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}
\usepackage{xtab}
%%%%%%%%
%TITLE:%
%%%%%%%%

\title{Epigraphic Echoes in Epigrams}

\author[CLUJ]{Marion Lam\'e\corref{first}}

\address[CLUJ]{Centre Camille Jullian, MMSH, CNRS, France}

\cortext[first]{Corresponding author. Email: mlame@mmsh.univ-aix.fr}

%\usepackage{longtable}
\begin{document}

\maketitle

\keywords{interoperability between editions, epigram, traditional and digital epigraphy, cross-disciplinary studies of inscriptions}

\section{Description}

For several years, the Memorata Poetis project (PRIN 2010/11) aims at studying the intertextual relationships between literary and epigraphic epigrams in several languages (Ancient Greek, Latin, Italian, Arabic and English). In order to identify some patterns that echo across space, time, culture and languages, one of the main activity consists in manually identifying Themes \& Motifs within the corpus of poems. Such Themes \& Motifs are organized, for now, as a hierarchical index of metadata that are associated with the text of the poem only. However, part of this corpus is composed of epigraphic primary sources. This means that such epigrams were set up in a deliberate context of communication and that part of their message is based on linguistic and non linguistic elements of their contextual, textual and writing systems. Some examples are iconographic programs and relationships with other poems, type of verses, inscriptions, themes and motifs (ex. AE, 1967, 85 ; De Hoz, 2014, No 355, Banti, No 5 and 51). In this panel, participants would like to discuss convergences and divergences between textual editing that tends to follow philological methods, and document editing, that focuses on the unicity of the materiality of the carrier (Pierazzo, 2015) and that appears more adapted to epigraphic edition. Both types of edition would facilitate cross-disciplinary studies of epigraphic epigrams. Panelists will focus on how connecting both types of edition thanks to digital technologies, i.e. integrating digital edition of epigraphic epigrams in the wider context of digital philology on one hand (e.g.: giving access to high quality images thanks to computer graphics tools) and in the one of public history on the other hand (e.g.: crowdsourcing with historical method).

DE HOZ M. 2014, Inscripciones griegas de España y Portugal, Madrid. 
PIERAZZO E. 2015, Digital Scholarly Editing: Theories, Models and Methods, Farnham: Ashgate.
BANTI, O. 2000, Monumenta epigraphica pisana saeculi XV antiquiora, Pisa.

\section{Panelists}
\begin{description}
\item[Marion Lamé]\emph{Centre Camille Jullian, MMSH, CNRS, France Email: mlame@mmsh.univ-aix.fr} Collaborates with the Laboratorio di Cultura Digitale of the University of Pisa. She has worked and works on several digital epigraphic projects (IGLouvre, TSS, Memorata Poetis...) and collaborates to collective dissemination of digital practices (EpiDoc). Her researches are dedicated to Digital Epigraphy applied especially to complex epigraphical situation and device (multilingualism, multialphabetism...).

\item[Pr. Paolo Mastrandrea] \emph{University of Venice Email: mast@unive.it}
P. Mastandrea is specialized in Latin Philology. He is the Project Coordinator of important national projects in digital humanities: Memorata Poetis and Musisque Deoque.

\item[Pr. Flavia De Rubeis] \emph{University of Venice.
Email: flavia.derubeis@unive.it} Specialized in Latin Paleography and is in charge of the epigraphic epigrams of the Middle Ages in Memorata Poetis Project.

\item[Dr. Massimo Manca] \emph{University of Torino
Email: massimo.manca@unito.it} Researcher specialized in Latin Language and Literature, his interests are Latin literature in Late antiquity, didactics of Ancient Greek and Latin, training of teachers, digital tools 
for classics.

\item[Pr. Alfredo Morelli] \emph{Università degli Studi di Cassino Email: alfmorel@unicas.it} Professor of Latin language and literature at the University of Cassino (Italy), he works on the history of Greco-Roman literary and epigraphic epigram of Hellenistic period until the age of Martial, the Roman elegy at the time of Augustus and the tragedies of Seneca. He has been active for years on the Carmina Latina Epigraphica in cooperation with Epigraphic Database Roma.

\item[Pr. Enrica Salvatori] \emph{University of Pisa
Email: enrica.salvatori@unipi.it} Specialized in Medieval history and Director of the Laboratorio di Cultura Digitale of the University of Pisa. She works in the digitization process of inscriptions of Pisa, included epigrams, and in projects of public history concering the valorization of epigraphs.
[http://pisaeislam.humnet.unipi.it/ - http://epigrapisa.humnet.unipi.it/]

\item[Dr. Eleonora Santin] \emph{CNRS - Laboratoire HiSoMA (UMR 5189) Email: eleonora.santin@mom.fr} Specialized in Greek Epigraphy and Greek Philology., she is a CNRS researcher with a personal program entitled Musa Epigraphica: new approaches for the study and the edition of the Greek epigraphic poetry and works on the digitization of epigraphic epigrams.

\end{description}
\end{document}

### 2016-01-25 panels/EAGLE_2016_submission_6.tex
added panels
copying from pdf panel presentations are horrible, especially the first
one.

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english,francais]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}
\usepackage{xtab}
%%%%%%%%
%TITLE:%
%%%%%%%%

\title{Assessing the Role of Digital Libraries of Squeezes in Epigraphic Studies. Digitization, visualization, and metadata}

\author[Flo]{Eleni Bozia\corref{first}}

\address[Flo]{University of Florida}
\cortext[first]{Corresponding author. Email: bozia@ufl.edu}

%\usepackage{longtable}
\begin{document}

\maketitle

\keywords{Squeezes, Online Libraries, 3D Visualizations, Contextuality}

\section{Description}
This panel will discuss issues and research questions regarding the use and efficiency of digital libraries of squeezes of inscriptions—structure of online libraries, nature of the metadata, 3D digitization, and visualization methods. The panel that consists of epigraphists, digital epigraphists, and computer engineers will engage in a dialogue, addressing the above issues from different perspectives.

Digital libraries of squeezes have become an integral part in epigraphic studies and research. Whether it is the issue of collecting data for more efficient use, increasing their accessibility, or simply digitally preserving them, digital libraries are a new scholarly medium. Such projects con- front the challenge of having to determine the types of data that should be included—traditional information, digital metadata, formats and tools for articulating all the specificities of the metadata—as well as decide whether
 the type of the project should predetermine the ontology of the data.

The aforementioned primary issue begs also the question of the need for interoperability of e-libraries in an attempt to combine, contrast, and comparatively appraise and study the artifacts themselves and their metadata. Thus far several projects feature digital libraries of squeezes; howe- ver, how should one proceed about perusing them? Whether this would be facilitated by means of homogeneity of the libraries, or whether the variegated nature of the material could potentially increase the number of users and the amount of information that is available.

However, one should consider the limitations of e-libraries that focus solely on squeezes. When discussing the nature of the digital libraries’ content and the implication that the metadata may have in the interpretation of a squeeze, we need to consider its contextuality—whether the absence of the inscription bearer from a squeeze’s record may prove reductive with regards to the holistic approach to the text. However, notwithstanding that high-resolution digitization of squeezes opens the possibilities for enhanced study on a level that until now could not be achieved, techniques that focus solely on the digitization of the squeezes usually fail to deal with the inscription bearer. This panel will discuss how the digital records of squeezes could be augmented to encapsulate their contextual information.

\section{Panelists}
\begin{description}
\item[Michèle Brunet]  \emph{Université Lyon 2, UMR HiSoMA \& Ecole française 
d'Athènes  Email: Michele.Brunet@univ-lyon2.fr} Prof. Brunet has been the Chair of Greek and Latin Epigraphy at the
Université Louis Lumière Lyon 2 since 2006. She has been a
l'École Normale Supérieure de Paris (1979-1984) and l'École française d'Athènes (1984-1988), She received her HDR on the Study of Antiquity at the Sorbonne. She taught Archaeology and Greek Art a Bordeaux III and Paris I Panthéon-Sorbonne. She also served as the Director of Ancient Studies at l’École française d'Athènes of member.
\item[Adeline Levivier] \emph{Université Lyon 2, UMR HiSoMA & Ecole française d’Athènes Email: adeline.levivier@gmail.com} Adeline is a doctoral candidate in l’École française d'Athènes - Université Lyon 2. She works on the digitization of squeezes, and her thesis is entitled ‘Recherches sur l'écriture grecque à partir des collections d'estampages d'inscriptions.’ She collaborated on the project ANR E-pigramme (Épigraphie et Muséographie - Édition numérique et valorisation de la Collection des inscriptions grecques du Musée du Louvre). She is currently the program manager of the project E-  stampages (Numérisation et diffusion web en 3D des collections de  l'UMR 5189 HiSoMA et de l'Ecole française d'Athènes).

\item[Manuel Ramírez Sánchez] \emph{Universidad de Las Palmas de Gran Canaria Email: manuel.ramirez@ulpgc.es} Prof. Sánchez is a Professor of Historiographic Sciences and Techniques at the Department of Historical Sciences. He works on the 3D digitization and advanced visualizations of inscriptions and has published extensively in the area of inscriptions from Ancient Hispania.

\item[Jose-Pablo Suárez-Rivero] \emph{Universidad de Las Palmas de Gran Canaria Email: Josepablo.suarez@ulpgc.es} Prof. Suárez-Rivero is the Director of Política Informática at the Cartography and Graphic Engineering Department. He has a Ph.D. in Applied Mathematics and works on mesh generation, algorithms and data structures, and computational geometry.

\item[Angelos Barmpoutis] \emph{University of Florida Email: angelos@digitalworlds.ufl.edu} Prof. Barmpoutis is an Associate Professor in the On-line Institute and the Digital Worlds Institute at the University if Florida. He is also the coordinator of research and technology at the Institute and affiliate faculty at the Computer Science and Engineering Department. His research focuses on interdisciplinary applications of computer science to the service of broad areas of learning and training.
\end{description}

\section{Notes}
The intent of this panel of a multifarious cohort of scholars is to pose questions that have derived from the advancements of technology and digital epigraphy research. The panel will consider possible answers and ways to utilize traditional knowledge with the advantageous flexibility of digital tools that will ultimately not only facilitate, but also enhance epigraphic studies.
\end{document}

### 2016-02-05 Posters.tex
added panels
copying from pdf panel presentations are horrible, especially the first
one.
corrections files added

@@
 % !TEX encoding = UTF-8 Unicode
-% -*- coding: UTF-8; -*-
+% !TEX program = pdflatex
 
@@
 All posters can be seen in the digital poster exhibition on the Eagle website.
-www.eagle-network.eu/about/events/eagle2016/digital-poster-exhibition/
+\url{www.eagle-network.eu/about/events/eagle2016/digital-poster-exhibition/}
 
@@
 %   
-\toprule Jan-Mathieu Carbon, Saskia Peels and Vinciane Pirenne-Delforge & THE COLLECTION OF GREEK RITUAL NORMS PROJECT (CGRN) \\\hline
+\toprule Jan-Mathieu Carbon, Saskia Peels and Vinciane Pirenne-Delforge & The Collection of Greek Ritual Norms Project (CGRN) \\\hline
 Stefano Caneva and Guido D'Ippolito & Pasquino: Talking Statues and Inscriptions in Rome\\\hline

### 2016-02-05 source/29/EAGLE2016_camera_ready.tex
paper 29 biblio updated
corrections made to 43, 33, 29

@@
 
-In the fields of epigraphy and archaeology, the areas of digital epigraphy and computational archaeology have benefit
+In the fields of epigraphy and archaeology, the areas of digital epigraphy and computational archaeology have benefited
 from the use of several of the sensors available in handheld devices. Crowd-sourcing of photographic data and
@@
 There are several examples in literature that present 3D digitization projects that have been undertaken by museums
-including the Epigraphical Museum of Athens \citep{sullivan_analytical_2011, papadaki_accurate_2015}, Museo Arqueológico Nacional de
+including the Epigraphic Museum of Athens \citep{sullivan_analytical_2011, papadaki_accurate_2015}, Museo Arqueológico Nacional de
 Madrid \citep{ramirez-sanchez_epigrafidigital:_2014}, Museo Nazionale Romano di Palazzo Altemps \citep{barmpoutis_interactive_2015}, Museo
@@
 other archaeological artifacts \citep{babeu__2011, pollefeys_image-based_2001, malzbender_polynomial_2001, esteban_silhouette_2004}.
-Comparative studies of 3D scanning methods for cultural herirage can be found in \citep{pavlidis_methods_2007} and \citep{bohler_3d_2004}.
+Comparative studies of 3D scanning methods for cultural heritage can be found in \citep{pavlidis_methods_2007} and \citep{bohler_3d_2004}.
 
@@
 technologies and the actual use in the professional epigraphic and archaeological practice as it has been hard for
-non-technology-oriented audiences to handle and manipulate tridimensional data, using conventional computer equipment.
+non technology-oriented audiences to handle and manipulate tridimensional data, using conventional computer equipment.
 Furthermore, without mechanisms for proper user interaction, a 3D model that is projected on a 2D screen is not
@@
 
-\noindent The forms of interactions presented in 1.3.1 and 1.3.2 involved only the motion sensors of a tabled computer without
+\noindent The forms of interactions presented in 1.3.1 and 1.3.2 involved only the motion sensors of a tablet computer without
 utilizing the pressure sensors of the screen of the device. The commonly used touch gestures (such as 2-finger twist to
@@
 \caption{Screenshot of our interactive environment. The user interacts with the 3D object, using touch gestures and
-selects one of the regions of interests. This action initiates other data tools such as the image viewer or the edge
+selects one of the regions of interests. This action initiates other data tools, such as the image viewer or the edge
 filter as shown in this example}
@@
 by following a set of 2D touch gestures. The prototype system was developed as part of the Digital Epigraphy and
-Archaeology (DEA) project using the open-source library VisiNeat for 3D visualization and interaction and is compatible
+Archaeology (DEA) project, using the open-source library VisiNeat for 3D visualization and interaction, and is compatible
 with iOS, Android, and Microsoft RT tablet and smart phone devices. The interface is available through the web-site of
@@
 
-In the future we plan to quantitatively evaluate the designed interface by tracking the user activities and analyze
+In the future, we plan to quantitatively evaluate the designed interface by tracking the user activities and analyze
 their motion patterns in the 3D space while they are interacting with their handheld device. 
+\nocite{osman_exploratory_2011}
 

### 2016-02-05 source/33/TIGLIO.tex
corrections made to 43, 33, 29

@@
 This comparison confirms also that the usability of resources is measured at a different level when they are made open, and that images and translations have an undeniable higher relevance as an online resource, thus attracting interest also to the transcriptions, while this does not happen the other way around and only those who know what they are looking for will stumble upon an ancient inscription published online.
-Nothing new in these observations, but this obvious observation is in contrast with the actual situation in which photos are few and translation even fewer.
-Why so, then?
+Nothing new in these observations, but this obvious observation is in contrast with the actual situation in which photos are few and translation even fewer.\footnote{Photographs are very useful  for simple stones inscribed monuments in large lettering (as e.g.
+Roman monumental texts), but for most longer documentary inscriptions, including most Attic material, after autopsy, squeezes are the most important editorial tool, because photos are mostly 2 D images of 3 D objects.} Why so, then?
 The possible reasons are:
@@
 \item the lack of an entry point which is easy to access and use
-\item people get easily worried by copyright due diligence and find diffucult sometime to trak back who is the author or the copyright owner of a photo
+\item people get easily worried by copyright due diligence and find difficult sometime to trak back who is the author or the copyright owner of a photo
 \item a lack of coordinated effort, planning and management of the storage of both these types of content

### 2016-02-05 source/8/EAGLE_2016_submission_8_revised.tex
fixed biblio for maya

@@
 %
+
+\address[BONN]{Rheinische Friedrich-Wilhelms-Universität Bonn}
+\address[MEL]{La Trobe University Melbourne}
 \address[GOT]{Niedersächsische Staats- und Universitätsbibliothek Göttingen}
-\address[MEL]{La Trobe University Melbourne}
-\address[BONN]{Rheinische Friedrich-Wilhelms-Universität Bonn}
+
 %
@@
 \item Identification and isolation of signs in one block 
-\item Classification of signs based on the sign catalog of Eric Thompson \citep{Thompson1962}
-\item Classification of signs in one block \citep{Thompson1962} 
+\item Classification of signs based on the sign catalog Eric Thompson
+\item Classification of signs in one block 
 \item Transliteration of individual signs 
@@
 \item Determining syntactic function
-\item Determining sentence constituents 
-\item Translation of single words 
-\item Literal translation 
-\item Rough translation
+
 \end{enumerate}
@@
 substitution. From a structuralist perspective, this method investigates the paradigmatic relation between two
-linguistic entities as indicated by whether or not they appear in the same context \citep{Bussmann2002}. This step yields
+linguistic entities as indicated by whether or not they appear in the same context. 
+%\citep{Bussmann2002} 
+This step yields
 results when sufficient quantities of data can be digitally compiled, marked up and investigated using corpus
@@
 word ``axe'' occurred in repeating, brief sign sequences on inscribed bronze axes. Building on this hypothesis, they
-determined the phonetic values of individual signs using data on the Canaantic language and completely deciphered the
-script \citep{Day2002}. Such investigations enable initial association of signs with both content and function in the
+determined the phonetic values of individual signs using data on the %Canaantic 
+language and completely deciphered the script \citep{Day2002}. Such investigations enable initial association of signs with both content and function in the
 relevant language so that the signs may later be grammatically, morphologically, semantically, etc., defined and
@@
 texts, a phenomenon which is deserving of investigation. In this manner, various systems of notation are studied in
-their individual, social, and typographic aspects, for instance \citep{Bussmann2002}. Similarly, in paleography, script
-decipherment is examined from graphetic perspective \citep{Bussmann2002}. The arrangement of texts, for instance, different
+their individual, social, and typographic aspects, for instance.% \citep{Bussmann2002}. 
+Similarly, in paleography, script
+decipherment is examined from graphetic perspective. %\citep{Bussmann2002}. 
+The arrangement of texts, for instance, different
 grapheme or block sizes of text fields, or different styles of hieroglyphic writing (equivalent to alphabetic font
@@
 When analyzing the contents of these texts, it is important to investigate meaningful variation underlying the manner in
-which the writing is arranged and to thereby identify distinct units \citep{Bussmann2002}. Graphemics is devoted to this area
+which the writing is arranged and to thereby identify distinct units. 
+%\citep{Bussmann2002}. 
+Graphemics is devoted to this area
 of research, and consequently to exploring the meaningfully distinct features of signs.
@@
 the case of the Maya script, signs whose phonemic reading is unknown or unconfirmed are denoted using reference systems
-from various classification catalogues \citep[e.g.]{Thompson1962}. Use of question marks for indicating unclear readings, for
+from various classification catalogues. 
+%\citep[e.g.]{Thompson1962}. 
+Use of question marks for indicating unclear readings, for
 example, should be avoided, because such signs impair machine readability and sometimes represent control characters or

### 2016-02-05 source/13/EAGLE2016_Full_Paper_Verreth.tex
fixed biblio Broux
altri paper sistemati secondo istruzioni

@@
 (both Greek and Egyptian), and the last few years - through the collaboration with EAGLE - we have also been
-incorporating Latin epigraphical texts from the whole Roman word. For the moment TM shows the metadata for 479.629
-texts, 306.925 of them written in Latin or containing Latin passages, but the number keeps on growing. Our direct EAGLE
+incorporating Latin epigraphical texts from the whole Roman word. For the moment TM shows the metadata for 618.917
+texts, 439.858 of them written in Latin or containing Latin passages, but the number keeps on growing. Our direct EAGLE
 partners Epigraphic Database Heidelberg (EDH), Epigraphic Database Roma (EDR), Hispania Epigraphica Online Database
@@
 which yielded in the end a full index of all personal names and toponyms in every Greek and Latin papyrus. This process
-is nowadays called Named-Entity Recognition [NER]\footnote{ For more information on NER, \citet[304-313]{BrouxDepauw}.}. These
+is nowadays called Named-Entity Recognition [NER]\footnote{ For more information on NER, \citet[304-313]{BrouxDepauw2}.}. These
 TM files within People and Places are freely accessible online and can be looked up, questioned and investigated in a
@@
 this corpus was enlarged by entering the toponyms occurring in the \emph{Itinerarium provinciarum Antonini Augusti} [3.434]
-and the \emph{Tabula Peutingeriana} [3.287]. The TM Geo file now contains 46.277 toponyms from all over Egypt and the Roman
+and the \emph{Tabula Peutingeriana} [3.287]. The TM Geo file now contains 46.707 toponyms from all over Egypt and the Roman
 empire, both ancient and modern, which cover most of the places where ancient texts have been found, and most of the
@@
 Phase 1, the identification of toponyms in the capital cluster strings, was finished in the beginning of July 2015,
-and we have now entered phase 2, the incorporation of the capital cluster file into the `real' TM Georef file. For
+and we have now almost completed phase 2, the incorporation of the capital cluster file into the `real' TM Georef file. For
 every place listed in TM Geo we try to list all the ancient text references where that place is attested. The file of
@@
 exported to a separate Georef card. When a toponym exists out of several consecutive elements, like the \emph{colonia} and
-\emph{municipium} examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as `\emph{Bythina et
-Ponthus'}, which cannot belong to the same capital cluster string because of the intermediary `\emph{et}', are exported double
+\emph{municipium} examples mentioned before, they are automatically grouped on one card. Twofold toponyms such as `\emph{Bithynia et
+Pontus}', which cannot belong to the same capital cluster string because of the intermediary `\emph{et}', are exported double
 and then afterwards joined manually. For the moment TM Georef lists 67.820 geographical references attested in Latin
@@
 abbreviations at the end of the word; e.g.
-\emph{T(h)ra[c(um)]} becomes \emph{Tra[c(um)]}, and Rom(a)e becomes \emph{Rome}.
+\emph{T(h)ra[c(um)]} becomes \emph{Tra[c(um)]}, and \emph{Rom(a)e }becomes \emph{Rome}.
 In another field the standardized nominative case is given, without brackets or uncertainty dots; e.g.
@@
 the immediate phrase to which the toponym belongs. The translation should be a standardised as possible, with termini
-technici preferably added in Latin, so that the users can easily search for them in the database; e.g. `\emph{Tiberius Iulius
-Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris}', `\emph{praefectus
-Aegypti}', `\emph{cohors I Tungrorum}'. If the place is explicitly ascribed to a \emph{provincia} or a region, this \emph{provincia} is
+technici preferably added in Latin, so that the users can easily search for them in the database; e.g. `Tiberius Iulius
+Martialis son of Tiberius of (the tribus) Claudia from Savaria, soldier (miles) of legio XV Apollinaris', `praefectus
+Aegypti', `cohors I Tungrorum'. If the place is explicitly ascribed to a \emph{provincia} or a region, this \emph{provincia} is
 listed in the field 'Administrative situation'. If a town is explicitly called an \emph{oppidum}, a \emph{vicus} or a \emph{civitas}, this
@@
 
- TM is a relational database, which implies that is possible to get to the information from different angles. If
+ TM is a relational database, which implies that it is possible to get to the information from different angles. If
 someone is studying a certain text, he can get a list of all toponyms mentioned in that text. On the other hand, if

### 2016-02-05 source/17/Zuddas Epigraphy in Italian High Schools (final).tex
altri paper sistemati secondo istruzioni

@@
 \author[up]{Enrico Zuddas\corref{first}}
-\address[up]{University of Perugia}\cortext[first]{Corresponding author. Email: enricozudd@yahoo.com}
+\address[up]{University of Perugia, Liceo Properzio Assisi}\cortext[first]{Corresponding author. Email: enricozudd@yahoo.com}
 \date{2015-11-16}
@@
 Inscriptions are an excellent tool to internalize declensions, but they can also broaden the horizons offered by the
-literary sources. The Eagle databases offer ready to use materials to teachers who are less familiar with the
+literary sources. The Eagle databases offer ready-to-use materials to teachers who are less familiar with the
 traditional supports.
 \end{abstract}
-\keywords{High School Education, Latin Grammar, Lexicon, Translations, EDR Database, Roman Asisium.}
+\keywords{High School Education, Latin Grammar, Lexicon, Translations, EDR Database, Roman \emph{Asisium}.}
 
@@
 Literature and reading literary texts, they generally realize that it was worth making the effort. Hence, one can
-imagine the frustration of Liceo Linguistico (Foreign Languages High School) pupils, who only do Latin for the first
+imagine the frustration of \emph{Liceo Linguistico} (Foreign Languages High School) pupils, who only do Latin for the first
 two years, and will never see the results of such hard work.
@@
 The use of local materials, which can be checked out personally, may more easily arouse the learners’ attention. That is
-why in this paper I will mainly refer to inscriptions from modern Umbria (regiones VI and VII, from Perusia) and
-especially from Asisium, used at school during my experimental lessons. The local museum, best known as the ``Foro
-romano'', hosts a huge collection of inscriptions that were catalogued in 2008 by pupils of the Liceo Properzio, under
+why in this paper I will mainly refer to inscriptions from modern Umbria (\emph{regiones} VI and VII, from \emph{Perusia}) and
+especially from \emph{Asisium}, used at school during my experimental lessons. The local museum, best known as the ``Foro
+romano'', hosts a huge collection of inscriptions that were catalogued in 2008 by pupils of the \emph{Liceo Properzio}, under
 my supervision, with details of the type of inscription, material, place of origin and chronology. Previously a booklet
@@
 both in Italian and English, comprising a page about the famous tetrastyle with a translation of the inscription on the
-base (CIL 11, 05372 = EDR025323, later inserted in the MediaWiki page), which recalls the official inauguration of the
-aedicule: at the time, as was the Roman custom, money was given to decurions, seviri Augustales and the common people.
-Thanks to these few lines, the class is taken on a journey to ancient Asisium, and gets a lot of feedback about
+base (CIL 11, 05372 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025323}{EDR025323}, later inserted in the MediaWiki page), which recalls the official inauguration of the
+\emph{aedicule}: at the time, as was the Roman custom, money was given to \emph{decurions}, \emph{seviri Augustales} and the common people.
+Thanks to these few lines, the class is taken on a journey to ancient \emph{Asisium}, and gets a lot of feedback about
 religion, architecture, social life and economy in the first century A.D.
@@
 was);
-\item the abstract ones that are fundamental to understanding the Roman way of thinking (e.g. imperium, virtus, fas).
+\item the abstract ones that are fundamental to understanding the Roman way of thinking (e.g. \emph{imperium, virtus, fas}).
 \end{itemize}
 Accordingly, a teacher should concentrate on those aspects of Roman society that are still pre-eminent for us:
-archaeology (Roman buildings, domus, theatres, roads), myths and religion, politics (Empire, war and globalization).
+archaeology (Roman buildings, \emph{domus}, theatres, roads), myths and religion, politics (Empire, war and globalization).
 Inscriptions certainly offer a less varied lexical repertoire than a literary text, but can still help to give
@@
 
-In an inscription such as CIL 11, 05400 = EDR025350:
+In an inscription such as CIL 11, 05400 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025350}{EDR025350}:
 
@@
 statuas ponendas in / aedem Herculis dedit ((sestertium)) (triginta milia). / Hic in vias sternendas in / publicum
-dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni ((sestertium))... }
+dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni / ((sestertium)) --- }
 \end{quotation}
@@
 For instance, if I say\emph{ Iulia rosas amat}, even an Italian who knows no Latin can get the meaning of the sentence.
-Problems are encountered when the sentences and texts to translate become more complex: only when it is too late does
+Problems are encountered when the texts to translate become more complex: only when it is too late does
 an ill-prepared teacher realize that the class has not internalized the language system and the patterns.
@@
 \begin{itemize}
-\item CIL 11, 05501 = EDR025449: \begin{quotation}
-\emph{Noniae Privatae, C. Propertius -{}-{}-{}-{}-{}-}
+\item CIL 11, 05501 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025449}{EDR025449}: \begin{quotation}
+\emph{Noniae / Privatae, / C(aius) Propertius / -----}
 \end{quotation}
@@
 
-\item CIL 11, 05461 = EDR025411: \begin{quotation}
-\emph{A(n)noru(m) XIX. Calventia C(ai) f(ilia) Polla, L(ucius)
-Vistinius vir, Gavia mater posuer(unt). }
+\item CIL 11, 05461 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025411}{EDR025411}: \begin{quotation}
+\emph{A(n)noru(m) XIX. / Calventia / C(ai) f(ilia)Polla, / L(ucius) Vistinius vir, / Gavia mater / posuer(unt).}
 
@@
 
-\item CIL 11, 05399 = EDR025349: \begin{quotation}
-\emph{P(ublius) Decimius P(ubli) l(ibertus) Eros Merula VIvir
-viam a cisterna ad domum L(uci) Muti stravit ea pecunia -{}-{}-{}-{}-{}- }
+\item CIL 11, 05399 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025349}{EDR025349}: \begin{quotation}
+\emph{P(ublius) Decimius P(ubli) l(ibertus Eros /  Merula VIvir / viam a cisterna / ad domum L(uci) Muti / stravit ea pecunia / -----}
 \end{quotation}
@@
 
-At the beginner’s level, the sentences used for examples, exercises and translations are fictitious and often banal.
+At the beginner’s level, the sentences used for examplifications, exercises and translations are often fictitious and banal.
 They give a false image of antiquity: for instance, when you start with the first declension you get the wrong
-impression of a ``female'' world simply because the male nouns are rare. Students are unimpressed; these sentences have
-no significance for them. Those provided by inscriptions are equally easy and short, but they are not banal because
+impression of a ``female'' world simply because the masculine nouns are rare. Students are unimpressed; these sentences have
+no significance for them. Those provided by inscriptions may be equally easy and short (without any editing), but they are not banal because
 they are a mirror of a society, they offer a historical perspective; there is always a story behind them. Even the
@@
 
-Let us consider a famous inscription from the Cathedral in Assisi (CIL 11, 05390 = EDR025340):
+Let us consider a famous inscription from the Cathedral in Assisi (CIL 11, 05390 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025340}{EDR025340}):
 
 \begin{quotation}
-\emph{Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(- - -),
+\emph{Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(---), /
 Ner(o) Babrius T(iti) f(ilius), C(aius) Capidas T(iti) f(ilius) C(ai) n(epos), V(ibius) Voisienus T(iti) f(ilius)
-marones murum ab fornice ad circum et fornicem cisternamq(ue) d(e) s(enatus) s(ententia) faciundum coiravere.}
+marones / murum ab fornice ad circum et fornicem cisternamq(ue) d(e) s(enatus) s(ententia) faciundum coiravere.}
 
 \end{quotation}
-The comprehension of the text is easy, especially with the abbreviations solved. Even so, many considerations can be
+The text is plain, especially with the abbreviations solved. Even so, many considerations can be
 made:
@@
 \begin{itemize}
-\item for grammar: words of third declensions such as \emph{maro} and \emph{fornix}; complements of direction; gerundive to express
-purpose; use of the form {}\emph{-ere} in the perfect tense personal endings;
+\item for grammar: third declension nouns such as \emph{maro} and \emph{fornix}; complements of direction; gerundive to express
+purpose; subject/verb agreement; use of the form {}\emph{-ere} in the perfect tense personal endings;
 \item for history: the use of the Latin language prior to the Social War, as proof of the intense Romanization of the
@@
 identification of the area around San Rufino as the ``acropolis''; the incorporation of the Roman wall into the left nave
-of the church (the inscription being still in situ).
+of the church (the inscription being still \emph{in situ}).
 \end{itemize}
 
-Schoolbooks rarely present these opportunities offered by inscriptions, so the main problem for a teacher is to access
-appropriate material.\footnote{ See the observations made by \citet{Carpenter2006}.} Holding a PhD in Late Antiquity and
-Umbrian inscriptions, I am fortunate enough to already know many sources. However, a graduate may not have enough
-knowledge in epigraphy, may not be familiar with the \emph{Corpus Inscriptionum Latinarum}; that is why the EAGLE project can
-also offer tools for further insights.
+Schoolbooks rarely seize these opportunities offered by inscriptions, so the main problem for a teacher is to access
+appropriate material.\footnote{ See the observations made by \citet{Carpenter2006}.
+One of the most successful attempts to teach Latin through Epigraphy comes from the Anglo-Saxon area: \citet{LaFleur2010}.} Holding a PhD in Roman History, I am fortunate enough to already know many sources.
+However, a graduate may not have enough
+knowledge in epigraphy, may not be familiar with the \emph{Corpus Inscriptionum Latinarum}; that is why the EAGLE project can offer tools for further insights.
+It also goes without saying that the multimedia approach is very suitable to students.
 
@@
 \begin{itemize}
-\item \citet[39]{Barbieri2015}: the inscription of the architect \emph{C. Vettius Gratus} (CIL X 3392) is
+\item \citet[39]{Barbieri2015}: the inscription of the architect \emph{C. Vettius Gratus} (CIL 10, 03392) is
 merely a decorative element on the page, without any relation to the topic (phonetic changes from Latin to Italian);
@@
 First of all, they will not find any help elsewhere: every other Latin text can easily be found – even if not always
-correctly translated – on student internet sites and in blogs (such as Splash Latino). But we all know very well that
+correctly translated – on student internet sites and in blogs (such as \emph{Splash Latino}). But we all know very well that
 the Internet still lacks many Italian translations of inscriptions.
@@
 abbreviation, except the easy ones), the word order cannot immediately be reconstructed, especially in decrees and
-carmina epigraphica. School dictionaries are not intended for interpreting epigraphical lexicon: some words may not be
-present, their meaning may not always be explained. A lot of institutions and \emph{formulae} which are clear to a specialist
+\emph{carmina epigraphica}. School dictionaries are not intended for interpreting epigraphical lexicon: some words may not be
+present (such as \emph{maro}, seen above), their meaning may not always be explained 
+(e.g. \emph{centonarius}, \emph{sevir}). A lot of institutions and \emph{formulae} which are clear to a specialist
 (e.g. \emph{quattuorvir iure dicundo}) may be hard to understand or to translate. It is, however, also the case that a teacher
@@
 \begin{itemize}
-\item CIL 11, 04431 = EDR025160: \begin{quotation}
+\item CIL 11, 04431 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025160}{EDR025160}: \begin{quotation}
 \emph{[Inf]austo, levis umbra, tuo mihi flebilis hora / sorte
@@
 \end{quotation}
-This funerary inscription from Ameria contains two verses, but the position of the words is tricky; \emph{infausto} may be
+This funerary inscription from \emph{Ameria} contains two verses, but the position of the words is tricky; \emph{infausto} may be
 taken as an adjective (as it is most commonly) and its meaning is basically the same as \emph{sorte tua}; students who are not
@@
 
- \item CIL 11, 04391 = EDR025123:\begin{quotation}
+ \item CIL 11, 04391 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025123}{EDR025123}:\begin{quotation}
 \emph{ Iuliae M(arci) f(iliae) Felicitati, / uxori C(ai) Curiati
@@
 explanations about specific offices rendered with the technical derivative word, or for concepts omitted in the
-original Latin text, very useful. Examples of this could be words such as \emph{centonarii} or \emph{marones} seen in inscriptions
-mentioned above, but also the following ones:
+original Latin text, very useful. 
+It might be worthwhile to provide a few more tips from personal experience:
 
 \begin{itemize}
-\item  CIL 11, 04213 = EDR130908: Interamna Nahars should be further qualified as ``Terni'' to
+\item  CIL 11, 04213 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR130908}{EDR130908}: \emph{Interamna Nahars} should be further qualified as ``Terni'' to
 help those readers who are not familiar with Umbrian cities;
 
-\item CIL 11, 01925 = EDR142701: the names of the emperors \emph{M. Aurelius Antoninus} and \emph{M.
-Antoninus Pius Germaticus Sarmaticus} need to be explained (Caracalla and Marcus Aurelius) to avoid confusion. It should
+\item CIL 11, 01925 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR142701}{EDR142701}: the names of the emperors \emph{M. Aurelius Antoninus} and \emph{M.
+Antoninus Pius Germanicus Sarmaticus} need to be explained (Caracalla and Marcus Aurelius) to avoid confusion. It should
 be noted that in my classes proper names, and not only those of emperors but also of other people, were generally
@@
 virtute}, considering that they are intuitively interpretable for an Italian.
+We are also faced here with a case of nouns that do not exist in our language (\emph{abnepos}, \emph{adnepos}), which makes the translation less fluent than the original.
 \end{itemize}
 
-This type of activity works better with Liceo Classico (Grammar School) pupils, who do translations from Latin and Greek
+This type of activity works better with \emph{Liceo Classico} (Grammar School) pupils, who do translations from Latin and Greek
 almost every day and who are more at ease with the use of dictionaries. However, there are still hurdles to overcome.
@@
 more appropriate for them to concentrate on this activity as much as possible during the months prior to the exam. For
-this reason, epigraphy can only be a supplement to traditional assignments; the Italian national curricula are
+this reason, epigraphy can only be a supplement to traditional assignments; the Italian national \emph{curricula} are
 apparently very free, but at the same time they are very rigid. Yet, a few forays into epigraphy can be stimulating,
@@
 \noindent On a more advanced level, inscriptions may also integrate certain aspects related to the study of Latin Literature. The
-most typical example could be a comparison between the Tabula Claudiana (CIL 13, 01668) and Tacitus’ account (Annales
+most typical example could be a comparison between the \emph{Tabula Claudiana} (CIL 13, 01668) and Tacitus’ account (\emph{Annales},
 XI, 23-24):\footnote{On which see \citet{Jahn1993}.} reading the original document is a privileged occasion to determine how
@@
 from the lower levels of society, such as CIL 05, 05300 from \emph{Comum} (a funerary stela made by a freedman for
-Pliny the Younger) or CIL 05, 05923 = EDR124245 regarding a strange case of ``friend deletion'' long before the Facebook
+Pliny the Younger) or CIL 05, 05923 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR124245}{EDR124245} regarding a strange case of ``friend deletion'' long before the Facebook
 era.\footnote{ \url{http://www.laricerca.loescher.it/lingue-classiche/327-un-amico-o-amicus-e-per-sempre.html}.}
 
-An interesting example from Umbria can be found in AE 1992, 0560-0561 = EDR150769 and EDR150784. The first gravestone
-recalls the acquisition of a tomb - which had previously been despoiled - by an heir of the founder, who then installed
-another cippus for 40 friends (amicis meis, i.e. freedmen probably belonging to the same association):
+An interesting example from Umbria can be found in AE 1992, 0560-0561 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR150769}{EDR150769} and \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR150784}{EDR150784}.
+The first gravestone
+recalls the acquisition of a tomb --- which had previously been despoiled --- by an heir of the founder, who then installed
+another \emph{cippus} for 40 friends (\emph{amicis meis}, i.e. freedmen probably belonging to the same association):
 \begin{quotation}
@@
 hunc locum / ubi ossa meorum requiescunt et mea / et amicorum meorum. Ex hoc sepulch[ro] / cippi perierunt duo et
-frontes duae sciun[t] / qui surupuit et acturi simus et legimus, / satis est testium etqs.}
+frontes duae. Sciun[t] / qui surupuit et acturi simus et legimus, / satis est testium etqs.}
 
@@
 far as to compare the qualifications \emph{Imperator Caesar Augustus} to a modern logo or even a \emph{hashtag}). On
-this subject, the altars \emph{Augusto sacrum} put up by \emph{Perusia restituta} (CIL 11, 01923 = EDR142666, EDR142667, EDR142668,
-EDR142669), even with a simple text, offer the opportunity to deal with an aspect as crucial as the imperial cult.
+this subject, the altars \emph{Augusto sacrum} put up by \emph{Perusia restituta} (CIL 11, 01923 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR142666}{EDR142666}, \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR142667}{EDR142667}, \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR142668}{EDR142668},
+\href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR142669}{EDR142669}), even with a simple text, offer the opportunity to deal with an aspect as crucial as the imperial cult.
 Traditionally, we read that Augustus was worshipped directly only in the Eastern provinces, but not in Rome and Italy.
@@
 would become \emph{Augusta}, as you can read on the city gates, especially on the newly restored Etruscan Arch (CIL 11, 01929
-= EDR142706).
+= \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR142706}{EDR142706}).
 
 Teaching, as I do, in a school named after Propertius, I always stress the importance of reconstructing the origin of
-the poet through the epigraphical data of the \emph{gens Propertia}: the greatest number of written documents of the family
+the poet through the epigraphical data of the \emph{gens Propertia}, the greatest number of written documents of the family
 having been found in Assisi. For more than two hundred years, beginning with the \emph{Vois(ienus) Ner. (filius) Propertius}
-mentioned among the Umbrian magistrates of the late II century B.C., this gens stands out in the town for its social
+mentioned among the Umbrian magistrates of the late II century B.C., this \emph{gens} stands out in the town for its social
 influence and wealth.\footnote{\citet{Forni1986}; \citet{Zuddas2006}.} It is always very exciting to combine the information on
-\emph{Passennus Paullus Propertius Blaesus }given by CIL 11, 05405 = EDR025355 and that contained in two Letters by Pliny (VI,
+\emph{Passennus Paullus Propertius Blaesus }given by CIL 11, 05405 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025355}{EDR025355} and that contained in two Letters by Pliny (VI,
 15 and IX, 22). Pliny, showing great concern for his friend’s illness but also great esteem for him as an elegiac poet,
@@
 perfectly. Students find it fascinating to look for traces of the poet inside the town, especially when they read the
-following graffito on an interior wall of a Roman house underneath the church of Santa Maria Maggiore (EDR028769):
+following graffito on an interior wall of a Roman house underneath the church of Santa Maria Maggiore (\href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR028769}{EDR028769}):\footnote{\citet[244-246]{Boldrighini2014}.}
 
@@
 This house was still visited in the fourth century A.D., which really supports the theory that it used to be the poet’s
-residence, and remained an object of reverence for centuries.\footnote{\citet[244-246]{Boldrighini2014}.} Through the stories
-of Passennus Paullus and Sextus Propertius macro-history and local history meet to make the past come alive in every
+residence, and remained an object of reverence for centuries. Through the stories
+of \emph{Passennus Paullus} and \emph{Sextus Propertius} macro-history and local history meet to make the past come alive in every
 corner of the modern town.

### 2016-02-05 source/48/EAGLE16-lame-et-al-teaching.tex
altri paper sistemati secondo istruzioni

@@
 \title{Teaching (Digital) Epigraphy}
-\author[ilc]{Marion Lamé}
+\author[ilc]{Marion Lamé}\corref{first}
 \author[zag]{Ivan Radman-Livaja}
 \author[mau]{Bruce Robertson}
-\author[isti]{Federico Ponchio\corref{first}}
-\address[ilc]{Istituto di Linguistica Computazionale}
+\author[isti]{Federico Ponchio}
+\address[ilc]{Centre Camille Jullian, MMSH, CNRS and Laboratorio di Cultura Digitale, University of Pisa}
 \address[isti]{Visual Computing Laboratory, ISTI-CNR}
@@
 \address[mau]{Mount Allison University}
-\cortext[first]{Corresponding author. Email: federico.ponchio@isti.cnr.it}
+\cortext[first]{Corresponding author. Email: mlame@mmsh.univ-aix.fr}
 
@@
 problematic. In this article we present our experience with teaching students to transcribe and interpret Roman
-inscribed lead tags, using a Digital Autoptic Process (DAP) in a web environment, so to develop basic competences in
+inscribed lead tags, using a Digital Autoptic Process (DAP) in a Web environment, so to develop basic competences in
 epigraphy and digital epigraphy.
 \end{abstract}
-\keywords{Educational project, digital epigraphy, epigraphy, open access, primary sources}
+\keywords{educational project, digital epigraphy, epigraphy, Open Access, primary sources}
 
@@
 reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
-Scientific \& Technical Experiment in Digital Epigraphy: the Study Case of the \emph{Tesserarum Sisciae Sylloge} (TSS) through
+Scholarly \& Technical Experiment in Digital Epigraphy: the Study Case of the \emph{Tesserarum Sisciae Sylloge} (TSS) through
 a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
@@
 \centering
+\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img002.jpg}
 \includegraphics[scale=0.25]{EAGLE16lameetalteaching-img001.jpg}
-\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img002.jpg}
-\caption{No. 12582.}
+\caption{No. 12582, obverse and reverse.}
 \label{fig:12582}
@@
 
-A traditional print photograph, such as this one (No. 12582, Fig.~\ref{fig:12582}) drawn from the archives of the AMZ, limits the student’s
+ A traditional print photograph, such as Fig.~\ref{fig:12582} (No. 12582), drawn from the archives of the AMZ, limits the student’s
 opportunity to learn to decipher these tags. Even if a professional photographer took this picture with a good quality
@@
 Words like \emph{LANA, PAN(N)UM, TVNICA, SAGVM, P(A)ENVLA, PAL(L)A, PALLIOLUM, LODIX, BANATA}, and \emph{ABOLLA}, appear regularly enough without being abbreviated and
-thus the interpretation of common abbreviations like \emph{l, la, lan, pan, t, sag, paenv, pal, lo, lod, lodi, bana, ab} is
+thus the interpretation of common abbreviations like \emph{L, LA, LAN, PAN, T, SAG, PAENV, PAL, LO, LOD, LODI, BANA, AB} is
 not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
@@
 (i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
-correct fee\footnote{\citet{radman_livaja_segestica_2008}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014},  for analogous tags with \citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_2011}; \citet[121-137/93, 215-222]{schwinden_romerzeitliche_1985}; \citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.
+correct fee\footnote{\citet{radman_livaja_segestica_2008}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014}, for analogous tags see:\citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_2011}; \citet[121-137]{schwinden_romerzeitliche_1985}; \citet[215-222]{schwinden_zwei_1993};\citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.
 
@@
 small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
-inscriptions, including both easily readable specimens and more challenging graffiti (e.g., No. 12582). One may also
+inscriptions, including both easily readable specimens and more challenging graffiti (e.g.: Fig.~\ref{fig:12582}, no. 12582). One may also
 profit from the assistance of undergraduate and graduate students, which could not often be the case
-before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa applies the
+before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa, applies the
 didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
@@
 
-From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in \citet{Lame2015}, on a dispositive analysis
+From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in \citet{lame_primary_2014}, on a dispositive analysis
 and the three fundamental systems of an inscription: writing system (wSystem), textual system (tSystem) and contextual
-system (cSystem). Three tools compose the DAP: RTI web viewer, TSS viewer and MarkOut tool. The last two were
+system (cSystem). Three tools compose the DAP: RTI Web viewer, TSS viewer and MarkOut tool. The last two were
 specifically designed for the digital edition of the corpus TSS. The following paragraphs present those three DAP
@@
 magnification provided by the high resolution camera employed. The widespread adoption in browsers of WebGL (Khronos
-group 2009) recently enable RTI visualization on the web. We used the RTI web viewer developed by Visual Computing Lab,
+group 2009) recently enable RTI visualization on the Web. We used the RTI Web viewer developed by Visual Computing Lab,
 ISTI-CNR. High resolution re-illuminable images allow the selection of the best view for each mark, thus allowing the
@@
 %\centering
-\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img005a.png}
+%2.965cm,height=1.79cm
+\includegraphics[width=\columnwidth]{EAGLE16lameetalteaching-img005a.png}
 %\end{figure}
@@
 %\centering
-\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img005b.png}
+%2.845cm,height=1.896cm
+\includegraphics[width=\columnwidth]{EAGLE16lameetalteaching-img005b.png}
 %\end{figure}
@@
 
+\newpage
 \section{Practicing Epigraphy}
@@
 classification gives a first idea of the main general shapes one can find on a lead tag. It is not a paleographical
-study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering.
+study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering (see: Fig.~\ref{fig:table1}).
 
-\subsection{Drawing With Help (No. 12563)}
+\begin{figure}[!hbp]
+\centering
+\includegraphics[width=\columnwidth]{4_glyphs.png}
+\caption{Glyph Browser: quick access to all the label associated with the glyph's form.}
+\label{fig:table1}
+\end{figure}
 
+\subsection{Drawing with help (No. 12563)}
+
+\vspace{0,5cm}
 \begin{minipage}[t]{0.18\textwidth}
@@
 %\centering
-\includegraphics[width=2.965cm,height=1.79cm]{EAGLE16lameetalteaching-img006.png}
+%2.965cm,height=1.79cm
+\includegraphics[width=\columnwidth]{EAGLE16lameetalteaching-img006.png}
 %\end{figure}
@@
 %\centering
-\includegraphics[width=2.845cm,height=1.896cm]{EAGLE16lameetalteaching-img007.png}
+%2.845cm,height=1.896cm
+\includegraphics[width=\columnwidth]{EAGLE16lameetalteaching-img007.png}
 %\end{figure}
@@
 
-No. 12563 is a rectangular lead tag of irregular form with a badly damaged and scratched surface. Nevertheless, the most
+\vspace{0,5cm}
+
+The tag no. 12563 is a rectangular lead tag of irregular form with a badly damaged and scratched surface. Nevertheless, the most
 recent inscription is clearly readable while one may still discern traces of older inscriptions. An older inscription
 might be visible on the obverse, perhaps series of numbers but its meaning remains unclear. Students have
-transcriptions and drawings.
+transcriptions and drawings (see Fig.~\ref{fig:1} and Fig.~\ref{fig:2}).
 
-\begin{table}
+\begin{table}[!hbp]
 {\small
@@
 \midrule
-\textbf{Stud. 1} & \textit{A10} & \textit{E2} & \textit{L11} & \textit{I1} & \textit{T1} & \textit{A10} & \textit{S3} & \textit{T2} & \textit{I1} \\
+\textbf{Stud. 1} & \textit{A10} & \textit{E2} & \textit{L1} & \textit{I1} & \textit{T1} & \textit{A10} & \textit{S3} & \textit{T2} & \textit{I1} \\
 \textbf{Stud. 2} & \textit{A10} & \textit{E2} & \textit{L5} & \textit{I1} & \textit{T2} & \textit{A10} & \textit{S4?} & \textit{T2} & \textit{I1}  \\
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
 \includegraphics[scale=0.25]{EAGLE16lameetalteaching-img009.png}
-\caption{12563 Obverse. The two students' reconstructions}
+\caption{No. 12563 Obverse: students' drawings of the most recent inscription.}
 \label{fig:1}
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
 \includegraphics[scale=0.25]{EAGLE16lameetalteaching-img008.png}
-\caption{12563 Reverse. The two students' reconstructions}
+\caption{No. 12563 Reverse: students' drawings of the most recent inscription.}
 \label{fig:2}
@@
 
-\subsection{Drawing \& Deciphering Without Help (obv. No. 13053)}
+\subsection{Drawing \& Deciphering without help (obv. No. 13053)}
 
@@
 \begin{minipage}[t]{0.3\textwidth}
-\textbf{Vale}\\
+Vale\\
 rius lis\\
@@
 \begin{minipage}[c]{0.3\textwidth}
-\includegraphics[scale=0.25]{EAGLE16lameetalteaching-img011a.png}
+\includegraphics[scale=0.20]{EAGLE16lameetalteaching-img011a.png}
 \end{minipage}
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
 \includegraphics[scale=0.25]{EAGLE16lameetalteaching-img010.png}
-\caption{13053 Obverse. Students' reconstructions}
+\caption{No. 13053 Obverse: students' drawings of the most recent inscription.}
 \label{fig:3}
@@
 The last line, MARTIA, gave the student a hard time. Particularly the tail of the damaged S (end of the second line)
-generates noise on the last letters.
+generates noise on the last letters (see Fig.~\ref{fig:3}).
 
@@
 
-\subsection{Palimpsest (obv. No. 13053)}
+The older, erased layers, on the obverse of tag No. 13053, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
+g/t relationship between wSystem and tSystem was asked to students (see Fig.~\ref{fig:4}).
 
-The older, erased layers, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
-g/t relationship between wSystem and tSystem was asked to students.
-
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
 \includegraphics[scale=0.25]{EAGLE16lameetalteaching-img012.png}
-\caption{12563 Obverse (palimpsest). The two students' reconstructions}
+\caption{No. 13053 Obverse: students' drawing of some palimpsest inscriptions.}
 \label{fig:4}
@@
 
-\noindent From the educational standpoint, this pilot project succeeded tremendously. The students who undertook this
+\noindent From the educational standpoint, this pilot project succeeded tremendously. The Canadian students who undertook this
 transcription in place of the usual research essay were, on average, far more motivated than their peers, and took
@@
 annual undergraduate research seminar.
+Three students are currently working on an exercise of online edition and publication (\emph{Épigraphie en Réseau} - ISSN: 2108-7156) of a lead tag using the Markout file created (SVG).
 
@@
 present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.
-
+\nocite{Lame2015b}
+\nocite{cherry_storytelling_2014}
 \bibliographystyle{sapauth-eng}

### 2016-02-09 source/17/Zuddas Epigraphy in Italian High Schools (final).tex
zuddas further changes

@@
 statuas ponendas in / aedem Herculis dedit ((sestertium)) (triginta milia). / Hic in vias sternendas in / publicum
-dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni / ((sestertium)) --- }
+dedit ((sestertium)) (triginta septem milia). / Hic pridie quam mortuus est / reliquit patrimoni / ((sestertium)) - - - }
 \end{quotation}
@@
 \item CIL 11, 05501 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025449}{EDR025449}: \begin{quotation}
-\emph{Noniae / Privatae, / C(aius) Propertius / -----}
+\emph{Noniae / Privatae, / C(aius) Propertius / - - - - -}
 \end{quotation}
 
-You can use an inscription as easy as this one to revise first and second declensions, and at the same time to teach the
+You can use an inscription as easy as this one to revise first and second declension, and at the same time to teach the
 structure of a Roman name.
@@
 \item CIL 11, 05399 = \href{http://www.edr-edr.it/edr\_programmi/res\_complex\_comune.php?do=book\&id\_nr=EDR025349}{EDR025349}: \begin{quotation}
-\emph{P(ublius) Decimius P(ubli) l(ibertus Eros /  Merula VIvir / viam a cisterna / ad domum L(uci) Muti / stravit ea pecunia / -----}
+\emph{P(ublius) Decimius P(ubli) l(ibertus) Eros /  Merula VIvir / viam a cisterna / ad domum L(uci) Muti / stravit ea pecunia / - - - - - }
 \end{quotation}
@@
 \begin{quotation}
-\emph{Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(---), /
+\emph{Post(umus) Mimesius C(ai) f(ilius), T(itus) Mimesius Sert(oris) f(ilius), Ner(o) Capidas C(ai) f(ilius) Ruf(- - -), /
 Ner(o) Babrius T(iti) f(ilius), C(aius) Capidas T(iti) f(ilius) C(ai) n(epos), V(ibius) Voisienus T(iti) f(ilius)

### 2016-02-09 panels/EAGLE_2016_submission_47.tex

@@
 \begin{description}
-\item[Marion Lamé]\emph{Centre Camille Jullian, MMSH, CNRS, France Email: mlame@mmsh.univ-aix.fr} Collaborates with the Laboratorio di Cultura Digitale of the University of Pisa.
+\item[Dr.
+Marion Lam\'e]\emph{Centre Camille Jullian, MMSH, CNRS, France Email: mlame@mmsh.univ-aix.fr} Collaborates with the Laboratorio di Cultura Digitale of the University of Pisa.
 She has worked and works on several digital epigraphic projects (IGLouvre, TSS, Memorata Poetis...) and collaborates to collective dissemination of digital practices (EpiDoc).
@@
 Email: flavia.derubeis@unive.it} Specialized in Latin Paleography and is in charge of the epigraphic epigrams of the Middle Ages in Memorata Poetis Project.
+
+\item[Dr.
+Mia Trentin] \emph{University of Venice Email: trentin.mia@unive.it} Mia Trentin is specialized in Medieval Epigraphy and Latin Palaeography, focusing on medieval ways of communication and expression through the analysis of historical graffiti writing.
+She collaborates with prof.
+Flavia De Rubeis for the project Memorata Poetis.
 
@@
 She works in the digitization process of inscriptions of Pisa, included epigrams, and in projects of public history concering the valorization of epigraphs.
-[http://pisaeislam.humnet.unipi.it/ - http://epigrapisa.humnet.unipi.it/]
-
-\item[Dr.
-Eleonora Santin] \emph{CNRS - Laboratoire HiSoMA (UMR 5189) Email: eleonora.santin@mom.fr} Specialized in Greek Epigraphy and Greek Philology., she is a CNRS researcher with a personal program entitled Musa Epigraphica: new approaches for the study and the edition of the Greek epigraphic poetry and works on the digitization of epigraphic epigrams.
+[\url{http://pisaeislam.humnet.unipi.it/} - \url{http://epigrapisa.humnet.unipi.it/}]
 
 \end{description}
+
+\section*{Acknowldgements}
+Eleorona Santin for her suggestions while preparing this panel.
+
 \end{document}

### 2016-02-09 panels/EAGLE_2016_submission_6.tex
updated panel Bozia

@@
 \section{Description}
-This panel will discuss issues and research questions regarding the use and efficiency of digital libraries of squeezes of inscriptions—structure of online libraries, nature of the metadata, 3D digitization, and visualization methods.
-The panel that consists of epigraphists, digital epigraphists, and computer engineers will engage in a dialogue, addressing the above issues from different perspectives.
+This panel discussed issues and research questions regarding the use and efficiency of digital libraries of squeezes of inscriptions—structure of online libraries, nature of the metadata, 3D digitization, and visualization methods.
+The panel that consisted of epigraphists, digital epigraphists, and computer engineers engaged in a dialogue, addressing the above issues from different perspectives.
 
@@
 Whether it is the issue of collecting data for more efficient use, increasing their accessibility, or simply digitally preserving them, digital libraries are a new scholarly medium.
-Such projects con- front the challenge of having to determine the types of data that should be included—traditional information, digital metadata, formats and tools for articulating all the specificities of the metadata—as well as decide whether
- the type of the project should predetermine the ontology of the data.
+Such projects confront the challenge of having to determine the types of data that should be included—traditional information, digital metadata, formats and tools for articulating all the specificities of the metadata—as well as decide whether the type of the project should predetermine the ontology of the data.
 
 The aforementioned primary issue begs also the question of the need for interoperability of e-libraries in an attempt to combine, contrast, and comparatively appraise and study the artifacts themselves and their metadata.
-Thus far several projects feature digital libraries of squeezes; howe- ver, how should one proceed about perusing them?
-Whether this would be facilitated by means of homogeneity of the libraries, or whether the variegated nature of the material could potentially increase the number of users and the amount of information that is available.
+Thus far several projects, including but not limited to the EAGLE consortium, the Center for Epigraphical and Palaeographical Studies, the Aleshire Collection at the University of California, Berkeley, and the US Epigraphy Project among others feature digital libraries of squeezes; however, how should one proceed about perusing them?
+Issues that need to be addressed include: whether their use would be facilitated by means of homogeneity of the libraries, or whether the variegated nature of the material could potentially increase the number of users and the amount of information that is available.
 
@@
 However, notwithstanding that high-resolution digitization of squeezes opens the possibilities for enhanced study on a level that until now could not be achieved, techniques that focus solely on the digitization of the squeezes usually fail to deal with the inscription bearer.
-This panel will discuss how the digital records of squeezes could be augmented to encapsulate their contextual information.
+Furthermore, there are other types of data—geo-spatial, prosopographic, linguistic—that would not only enhance our understanding of any one particular squeeze, but also contribute to our overall apprehension of classical, archeological, and epigraphic studies.
+This panel discussed how the digital records of squeezes could be augmented to encapsulate their contextual information.
+
+Questions that were addressed:
+
+\begin{enumerate}
+\item How to preserve the traditional nature of the data—specificity, terminology—while retaining the possibility of keyword search?
+
+\item Have we started thinking/searching/researching differently. Is it the database that determines the type of research that one conducts?
+
+\item Do we create authenticity via our digitization selections? In turn does this mean that we validate?
+
+\item How can computer scientists and humanists communicate to find a common point of reference between creating efficient algorithms and databases while retaining the nature of epigraphic and archaeological studies?
+
+\item TEI has created a common form of expression and advanced the homogeneity of data. Could we proceed with a similar trajectory for other forms of data?
+
+\item Would such a scenario also resolve the issue of contextuality.
+One database need not be an artifact’s concordance, but interlinked databases could preserve their individuality while advancing search possibilities?
+
+\item What are other aspects of the artifact (inscription bearer etc.) that could enhance the record of the squeeze?
+
+\item What kind of information would produce a more holistic record for the squeezes.
+\end{enumerate}
+
+	Eleni Bozia gave a brief introduction to the history of online databases of squeezes.
+She presented the type of data and metadata that they favor and proceeded to furnish the problems that surface when research groups do not opt for interlinked databases or structure the metadata according to the tools available, rather than according to the research needs and the nature of the database.
+Prof.
+Bozia also argued in favor of flexibility of databases and projects with the intent to consider the needs of each database rather than favor an existing ontological structure that restricts the content, and contextuality of the database, thus limiting its effectiveness.
+Adeline Levivier presented the collaboration between Université Lyon 2, UMR HiSoMA \& Ecole française d’Athènes and the Digital Epigraphy and Archaeology Project at the University of Florida.
+The project invoves the digitization of approximately 9000 squeezes in 3D as well as the creation of an online library with metadata of the collection.
+The squeezes date from Archaic Greek period to the Roman Imperial era and originate from Delos, Thasos, Delphi, and Asia Minor.
+The group pursues dematerialization for long-term preservation, remote consultation of data, standardization for interoperability with other resources related to cultural heritage and digital epigraphy, and creation of a new resource for research: a big data corpus to  be queried by advanced systems.
+Issues of selection were brought forward—how do research groups make choices— as well as the usability of 3D models of squeezes that give the option of advanced visual manipulations and automatic measurements.
+The major matter in question that the group intends to address in this project is the need to create a database that fits the researcher’s needs instead of merely reusing existing expertise and ontology.
+
+	Manuel Sánchez and Jose Pablo Su\'arez-Rivero advanced the discussion by presenting options for the enhancement of the digital libraries of squeezes.
+They presented their project titled Epigraphia 3D in which they perform 3D digitization of the entire artifact/inscription bearer.
+This admittedly provides a more holistic record of the squeeze and a context for the inscription itself.
+Profs.
+Sánchez and Su\'arez-Rivero presented their methodology that employs photogrammetry, their results, and their online database with the artifacts.
+An issue that was brought forward is how one can combine within the same project a detailed 3D digitization of the letters of the inscription/squeeze as well as a large-scale artifact/inscription bearer.
+The advantages, however, of a more complete record of the artifact and the squeeze became apparent.
+
+Finally, Angelos Barmpoutis concluded the panel with a discussion on the Digital Epigraphy Project and its metadata perspective, arguing in favor of a collected record as well as the flexibility to incorporate fields and metadata particularities as needed in its case.
+He discussed the need to treat both the physical squeeze as well as the digital artifact as objects that require metadata.
+He presented the sandbox upon which the Digital Epigraphy project functions with regards to the addition of metadata information.
+Prof.
+Barmpoutis also discussed the significance of considering every user’s unique perspective should one wishes to increase the usability and importance of their research database.
+
+In conclusion, the panel discussed theoretical issues that derive from the exigency to communicate and consider the advancements of technology alongside the needs of traditional research and presented current projects, databases, and actual case scenaria that deal with such issues.
+The intent of this multifarious cohort of scholars was to emphasize the importance of transdisciplinary collaboration and the pivotal role of the individual researcher/user.
+The panel considered ways to utilize traditional knowledge with the advantageous flexibility of digital tools that will ultimately not only facilitate, but also ultimately enhance epigraphic studies.
 
@@
 \begin{description}
-\item[Michèle Brunet]  \emph{Université Lyon 2, UMR HiSoMA \& Ecole française 
-d'Athènes  Email: Michele.Brunet@univ-lyon2.fr} Prof. Brunet has been the Chair of Greek and Latin Epigraphy at the
-Université Louis Lumière Lyon 2 since 2006. She has been a
-l'École Normale Supérieure de Paris (1979-1984) and l'École française d'Athènes (1984-1988), She received her HDR on the Study of Antiquity at the Sorbonne.
-She taught Archaeology and Greek Art a Bordeaux III and Paris I Panthéon-Sorbonne.
-She also served as the Director of Ancient Studies at l’École française d'Athènes of member.
-\item[Adeline Levivier] \emph{Université Lyon 2, UMR HiSoMA & Ecole française d’Athènes Email: adeline.levivier@gmail.com} Adeline is a doctoral candidate in l’École française d'Athènes - Université Lyon 2.
-She works on the digitization of squeezes, and her thesis is entitled ‘Recherches sur l'écriture grecque à partir des collections d'estampages d'inscriptions.’ She collaborated on the project ANR E-pigramme (Épigraphie et Muséographie - Édition numérique et valorisation de la Collection des inscriptions grecques du Musée du Louvre).
-She is currently the program manager of the project E-  stampages (Numérisation et diffusion web en 3D des collections de  l'UMR 5189 HiSoMA et de l'Ecole française d'Athènes).
+\item[Michèle Brunet]  \emph{Université Lyon 2, UMR HiSoMA \& Ecole française d’Athènes  Email: Michele.Brunet@univ-lyon2.fr} Prof.
+Brunet has been the Chair of Greek and Latin Epigraphy at the Université Louis Lumière Lyon 2 since 2006.
+She has been a member of l'École Normale Supérieure de Paris (1979-1984) and l'École française d'Athènes (1984-1988).
 
-\item[Manuel Ramírez Sánchez] \emph{Universidad de Las Palmas de Gran Canaria Email: manuel.ramirez@ulpgc.es} Prof.
-Sánchez is a Professor of Historiographic Sciences and Techniques at the Department of Historical Sciences.
+\item[Adeline Levivier] \emph{Université Lyon 2, UMR HiSoMA \& Ecole française d’Athènes Email: adeline.levivier@gmail.com} Adeline is a doctoral candidate in l’École française d'Athènes - Université Lyon 2.
+She works on the digitization of squeezes.
+She collaborated on the project ANR E-pigramme (Épigraphie et Muséographie - Édition numérique et valorisation de la Collection des inscriptions grecques du Musée du Louvre).
+She is currently the program manager of the project E-stampages (Numérisation et diffusion web en 3D des collections de l'UMR 5189 HiSoMA et de l'Ecole française d'Athènes).
+
+\item[Manuel Ramírez Sánchez] \emph{Universidad de Las Palmas de Gran Canaria Email: manuel.ramirez@ulpgc.es} Prof.
+Sánchez is a Professor of Historiographic Sciences and Techniques at the Department of Historical Sciences.
 He works on the 3D digitization and advanced visualizations of inscriptions and has published extensively in the area of inscriptions from Ancient Hispania.
 
-\item[Jose-Pablo Suárez-Rivero] \emph{Universidad de Las Palmas de Gran Canaria Email: Josepablo.suarez@ulpgc.es} Prof.
-Suárez-Rivero is the Director of Política Informática at the Cartography and Graphic Engineering Department.
+\item[Jose-Pablo Suárez-Rivero] \emph{Universidad de Las Palmas de Gran Canaria Email: Josepablo.suarez@ulpgc.es} Prof.
+Suárez-Rivero is the Director of Política Informática at the Cartography and Graphic Engineering Department.
 He has a Ph.D.
@@
 Barmpoutis is an Associate Professor in the On-line Institute and the Digital Worlds Institute at the University if Florida.
-He is also the coordinator of research and technology at the Institute and affiliate faculty at the Computer Science and Engineering Department.
+He is also the coordinator of research and technology at the Institute and affilitate faculty at the Computer Science and Engineering Department.
 His research focuses on interdisciplinary applications of computer science to the service of broad areas of learning and training.
@@
 
-\section{Notes}
-The intent of this panel of a multifarious cohort of scholars is to pose questions that have derived from the advancements of technology and digital epigraphy research.
-The panel will consider possible answers and ways to utilize traditional knowledge with the advantageous flexibility of digital tools that will ultimately not only facilitate, but also enhance epigraphic studies.
 \end{document}

### 2016-02-09 source/24/Epigraphy and onomastics in Hesperia databank-template.tex
fixes to Varga e Moucunill

@@
 \title{Epigraphy and onomastics in the Hesperia databank}
-\author[barc]{Noemí Moncunill Martí \corref{first}}
-\author[sorb]{Javier Velaza}
+\author[barc]{Noemí Moncunill Martí}
+\author[sorb]{Javier Velaza \corref{first}}
 \address[sorb]{Université de Paris-Sorbonne (Paris IV)}
-\address[barc]{University of Barcelona2}
-\cortext[first]{Corresponding author. Email: nmoncunill@gmail.com}
+\address[barc]{University of Barcelona}
+\cortext[first]{Corresponding author. Email: noemi.moncunill\_marti@paris-sorbonne.fr and velaza@ub.edu}
 \date{2015-11-15}
@@
 and advantages, taking the experience gained in the last years within the Hesperia project as a starting point.\footnote{ This paper is an output from the FFI2012-25113 project and the Senior Research Team LITTERA
-(2014SGR63), on the one hand, and 655938 — ZEPHYROS — H2020-MSCA-IF-2014 project, on the other.}
+(2014SGR63), on the one hand, and it has received funding from the European Union’s Horizon 2020 research and innovation programme under the Marie Sklodowska-Curie grant agreement No 655938, on the other.}
 
@@
 inscriptions through such relevant data as their date, palaeography or onomastics are set clear. Obviously, those
-connexions can be set by the users in their mind, but when the corpus displays thousands or tens of thousands of
+connexions can be made by the users in their mind, but when the corpus displays thousands or tens of thousands of
 inscriptions, it looks too optimistic to take for granted that most readers will have the aptitude such work takes.
@@
 In our opinion, these limitations can be solved or, at least, diminished by a proper and efficient use of the tools IT
-puts to our disposal. Although this is no longer anything of new, any attempt to use them so far has brought to an
+puts to our disposal. Although this is no longer anything of new, many attempts to use them so far have brought to an
 output too close to the traditional, non-innovative epigraphic corpora, considering the features of such new tools. To
@@
 reliability. Ancient Hispania left us epigraphic texts in at least four languages: Tartessian, Iberian, Celtiberian and
-Lusitanian -{}-the possibility there are also some texts in Palaeobasque language is still sub iudice\footnote{
-Concerning this subject, see \citet{velaza_epigrafiy_2009}.}{}-{}-, mainly written in four epichoric scripts -{}-Tartessian,
+Lusitanian --- the possibility there are also some texts in Palaeobasque language is still sub iudice\footnote{
+Concerning this subject, see \citet{velaza_epigrafiy_2009}.} --- , mainly written in four epichoric writing systems --- Tartessian,
 southeastern-Iberian, northeastern-Iberian and Greek-Iberian script.
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 The current knowledge of each of these scripts is in very different stages, just as when it comes to their languages
-deciphering. Along with the epigraphic texts there are also other evidences, as those from onomastics -{}-anthroponymy
-and theonymy preserved in Roman inscriptions found in these territories, and toponymy from Classic sources-{}- and some
+deciphering. Along with the epigraphic texts there are also other evidences, as those from onomastics --- anthroponymy
+and theonymy preserved in Roman inscriptions found in these territories, and toponymy from Classic sources --- and some
 notes transmitted by several authors.
@@
 Hesperia was born as a natural successor in the digital era of the main corpus for Palaeohispanic languages, that is
-Monumenta linguarum Hispanicarum, published between 1975 and 2001 by Jürgen Untermann. The project conception is due to
+\emph{Monumenta Linguarum Hispanicarum}, published between 1975 and 2001 by Jürgen Untermann. The project conception is due to
 Javier de Hoz and its digital platform has mainly been developed by Eduardo Orduña. Nowadays it is coordinated by an
@@
 ``epigraphy and paleography'', ``pictures'',
-``arqueological context'' and ``bibliography''. The databases are
+``archaeological context'' and ``bibliography''. The databases are
 linked between them through a powerful engine for simple and complex searches, and pdf documents as well as maps can be
 created with the resulting outcomes. Work is currently being focused on C and D zones (Iberian inscriptions in
-Catalonia), which is scheduled to be open for search by the end of 2015.
+Catalonia), which is scheduled to be open for search by the end of 2016.
 
@@
 
-\item[iii] An onomastic database, with a specific search engine allowing combined searches with any of the elements typical
-from the identified onomastic formulae.
+\item[iii] An onomastic database, with a specific search engine allowing combined searches with any of the elements from the identified onomastic formulae.
 
@@
 as lexicon, as well as on creating several kinds of tools, as the possibility of displaying the corpus according to
-alternative readings -{}-which turns out to be essential in epigraphic contexts with a lower level of decoding, as in
-Tartessian, although its applicability to other fields is still being tested-{}-.
+alternative readings --- which turns out to be essential in epigraphic contexts with a lower level of decoding, as in
+Tartessian, although its applicability to other fields is still being tested ---.
 
 Besides, the team has also decided to consider the potential applicability of this corpus structure to other cultures of
-the western Mediterranean. With such aim, prof. Francisco Beltrán leads the AELAW project (Ancient European Languages
-and Writings), with the participation of researchers from ten different countries, and that has been granted a COST
-program in order to build in future a single corpus with the evidences from all fragmentary languages documented in the
+the western Mediterranean. With such aim, prof. Francisco Beltrán leads the AELAW (Ancient European Languages
+and Writings), with the participation of researchers from ten different countries to build in future a single corpus with the evidences from all fragmentary languages documented in the
 western Mediterranean for the Ancient times.
@@
 In this new section we have compiled all the pre-Roman personal names and divinity names attested both by direct and
-indirect tradition. This represents, so far, a total of nearly 6.000 records. Names attested in direct sources are
-those that can be identified in epicoric epigraphies; the second group, in its turn, contains the names that can be
+indirect tradition. This represents, so far, a total of nearly 6,000 records. Names attested in direct sources are
+those that can be identified in epichoric epigraphies; the second group, in its turn, contains the names that can be
 identified in the so-called colonial epigraphy, as well as in literary sources. Regarding the chronology of the
@@
 
-As said above, the database contains about 6.000 records so far; each of them reports a divinity name and/or the whole
+As said above, the database contains about 6,000 records so far; each of them reports a divinity name and/or the whole
 onomastic formula of individuals whose complete name presents at least one indigenous element. This means that each
@@
 
-The following could be a good example of one of the above-described records. This Latin inscription from southern Spain
+The following (Fig.~\ref{fig:3}) could be a good example of one of the above-described records. This Latin inscription from southern Spain
 mentions a woman with a Latin nomen, Aelia, followed by an Iberian cognomen, Belesiar, as well as the name of an
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
- \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003.png}
+ \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img003new.png}
 \caption{Personal-names and divinity-names record in Hesperia databank.}
@@
 A slightly different kind of record is conceived to compile the personal names attested in literary sources. As shown
-below, in these cases the bibliographical field is merely used to mention the ancient passages in which the name is
+below (Fig.~\ref{fig:4}), in these cases the bibliographical field is merely used to mention the ancient passages in which the name is
 attested; there is no geographical information, and the “observations” field is often used just to report the different
@@
 with the information available a map can be easily generated with all the spots where at least an indigenous name is
-attested (in green); or a divinity name (in blue); or, finally, combine those maps in a single interactive map (in
-green and blue), where each point is directly linked with the corresponding records.
+attested (Fig.~\ref{fig:5}); or a divinity name (Fig.~\ref{fig:6}); or, finally, combine those maps in a single interactive map (Fig.~\ref{fig:7}), where each point is directly linked with the corresponding records.
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 \caption{Dispersal area of indigenous divinity names}
-\label{fig:}
+\label{fig:6}
 \end{figure}
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 isoglosses, on the basis, for instance, of the attestation of a well-known anthroponymic element or the dispersal area
-of significant phonetic features. In the following maps it is displayed, in the first place, the attestation area of
+of significant phonetic features. In the following maps it is displayed, in the first place (Fig.~\ref{fig:8}), the attestation area of
 names containing the element \emph{biur}, which clearly corresponds to the Iberian area, that is, the non indo-European half
-of the Iberian peninsula; and, in the second place, it is displayed the dispersal area of the name Tancinus, which
+of the Iberian peninsula; and, in the second place (Fig.~\ref{fig:9}), it is displayed the dispersal area of the name Tancinus, which
 clearly corresponds to the indo-European part.
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 \caption{Dispersal area of the anthroponymic element \emph{biur}}
-\label{fig:}
+\label{fig:8}
 \end{figure}
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 
-The following file contains information on a person’s or god’s name observed in an Iberian rock inscription in the
+The following file (Fig.~\ref{fig:10}) contains information on a person’s or god’s name observed in an Iberian rock inscription in the
 Pyrenees. Nevertheless, the structure of the word as well as its phonetic features show no possible interpretation in
@@
 Greek name. As a matter of fact, the same name is attested as a mythological character in literary sources (Parthenius
-of Nicaea XXX 1; Ovid, Ibis 434), as well as record as a personal name in open-access databanks such as LGPN
+of Nicaea XXX 1; Ovid, Ibis 434), as well as recorded as a personal name in open-access databanks such as LGPN
 (V5a-20577) or Trismegistos (Per 222620). Therefore, in this particular case, the linking between open-access data
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
  \includegraphics[width=\columnwidth]{EpigraphyandonomasticsinHesperiadatabanktemplate-img010.png}
-\caption{Fig. 10: record of a possible Greek name attested in an Iberian inscription.}
+\caption{Record of a possible Greek name attested in an Iberian inscription.}
 \label{fig:10}

### 2016-02-09 source/9/Latin epigraphy for the visually impaired.tex
added images for Licordari

@@
 
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{1tactiletourVillaGiulia.JPG}
+\caption{Sculptures are permitted to be touched in the National Etruscan Museum of Villa Giulia in Rome}
+\label{fig:1}
+\end{figure}
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{2resinreliefMonaLisa.JPG}
+\caption{Resin relief of the Anteros Museum: Mona Lisa}
+\label{fig:2}
+\end{figure}
+
 \section{Problems in Understanding Epigraphy}
@@
 
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{3graphicprintBrailleMarcusUlpiusFructus.jpg}
+\caption{The technical device of graphic Braille print for the panel of Marcus Ulpius Fructus}
+\label{fig:3}
+\end{figure}
+
 A technique which solves this problem is that of the Thermoform, that is a relief obtained from a plastic material,
@@
 copies are derived. The ease of obtaining new copies compensates for the rapid deterioration of the material.
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{4thermoform.JPG}
+\caption{The technique of the Thermoform with caption in Braille}
+\label{fig:4}
+\end{figure}
 
@@
 for permanent collections.
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{5Minoltaheather.jpg}
+\caption{Reproduction with printing granules (Minolta Heather) of the Cortona Triptych by Fra Angelico}
+\label{fig:5}
+\end{figure}
 
@@
 
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{6smallmodel3Dprinting.JPG}
+\caption{Small model in 3D printing of the Sarcophagus of the Spouses}
+\label{fig:6}
+\end{figure}
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{7CloneSarcophagusoftheSpouses.jpg}
+\caption{The clone of the Sarcophagus of the Spouses (scale 1:1) starting from the digital tridimensional scan of the
+original Sarcophagus, Italdesign Giugiaro}
+\label{fig:7}
+\end{figure}
+
 This technique would be very valid when used in the case of reproduction of mosaics and inscriptions in the mosaic.
@@
 
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{8reproductionOstiaslighthouse.JPG}
+\caption{The reproduction of the mosaic of Ostia’s lighthouse by the Federazione Nazionale delle Istituzioni Pro Ciechi NPO in the Square of the Corporations of Ostia Antica}
+\label{fig:8}
+\end{figure}
+
 A thesis paper, of the Postgraduate School of Archaeological Heritage in Latin Epigraphy \citep{Licordari2015}, which
@@
 and presented on different levels.
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{9mosaicoftheLibyanboatmen.jpg}
+\caption{The mosaic of the Libyan boatmen in Ostia Antica}
+\label{fig:9}
+\end{figure}
 
@@
 
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{10italicmirror.jpg}
+\caption{A mirror from Praeneste  in the National Etruscan Museum of Villa Giulia in Rome}
+\label{fig:10}
+\end{figure}
+
 \section{Conclusions}

### 2016-02-09 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
corrections files added
fixes to Sarullo

@@
 by photos and illustrations.\footnote{Giovanna Rocca (GR) is responsible for § \ref{sec:generalia} and \ref{sec:news};
-Giulia Sarullo (GS) is responsible for § \ref{sec:3chart} and Marta Muscariello (MM) for § \ref{sec:4tech}.}
+Giulia Sarullo (GS) is responsible for § \ref{sec:4tech} and Marta Muscariello (MM) for § \ref{sec:3chart}.}
 \end{abstract}
@@
 
-As a matter of fact, face A conforms to the others, showing three vertical dots after \emph{sakros} and also, although less
-deep and more worn-out, after \emph{esed}\footnote{The details of significant portions of the inscriptions will be shown on
+As a matter of fact, face A seems to conform to the others, showing three vertical dots after \emph{sakros}\footnote{The details of significant portions of the inscriptions will be shown on
 the website with several photographs. } (Fig.~\ref{fig:1}). This fact was highlighted by \citet{gamurrini_paleografia_1899} and the three dots
@@
 (EDB), \url{http://www.edb.uniba.it/}, the Epigraphische Datenbank Heidelberg (EDH),
-\url{http://www.uni-heidelberg.de/institute/sonst/adw/edh/index.html} and Hispania Epigraphica (HE),
+\url{http://edh-www.adw.uni-heidelberg.de} and Hispania Epigraphica (HE),
 \url{http://www.eda-bea.es/}.} is a database, the archaic Latin inscriptions have been digitalized according to the EpiDoc
@@
 also some particular cases, such as the lamina from Lavinium (CIL I$^2$ 2833) and the Tibur pedestal inscription (CIL I$^2$
-2658), that requires a specific treatment. For these texts, it was necessary to adopt the \texttt{{\textless}lb
-rend={\textquotedbl}up-to-down{\textquotedbl}{\textgreater}} and \texttt{{\textless}lb
-rend={\textquotedbl}down-to-up{\textquotedbl}{\textgreater}} elements in order to render the peculiar directions in
+2658), that requires a specific treatment. For these texts, it was necessary to create several new specific elements in order to render the peculiar directions in
 which the text was cut.\footnote{The issue of the direction of writing was the subject of much debate at the 6th EAGLE
@@
 antiquity and the necessity to establish a common standard to encode the instances of ``non-standard'' directions of
-writing emerged. For this reason, this issue will surely be the object of further consideration in the future.}
+writing emerged. This issue is discussed in \citet{SarulloBari}, where the newly created elements are presented.}
 Reversed and upside-down letters are usually left unmarked in traditional epigraphic editions and we decided to comply

### 2016-02-09 source/42/Paper_licciardello.tex
lower case titles
fixes to licciardello

@@
 %%%%%%%%
-\title{DEIXIS AND FRAMES OF REFERENCE IN DEDICATORY EPIGRAMS. The use of a database with an interdisciplinary approach}
+\title{Deixis and Frames of Reference in Dedicatory Epigrams. 
+The use of a database with an interdisciplinary approach}
 \author[topoi]{Flavia Licciardello\corref{first}}
@@
 \begin{abstract}
-This paper aims to present an example of a database designed to combine epigraphic, linguistic and
-philological data. This database is part of my project on a study of deictic expressions in dedicatory inscribed and
+This paper presents an example of a database designed to combine epigraphic, linguistic and philological data.
+This database is part of my project on a study of deictic expressions in dedicatory inscribed and
 literary epigrams. It includes the results of the analysis of around 600 dedicatory epigrams and it will be used to
@@
 epigrams, i.e. on the use of all those linguistic elements whose meaning and interpretation depend on the spatial and
-temporal context where they are uttered.\footnote{ For an overview of the concept of deixis see \citep{diessel_deixis_2012} (with
-further bibliography). On deixis in Ancient Greek see e.g. \citep{felson_poetics_2004}, \citep{edmunds_deixis_2008} and \citep{bonifazi_deixis_2014}.} The
+temporal context where they are uttered.\footnote{ For an overview of the concept of deixis see \citet[with further bibliography)]{diessel_deixis_2012}.
+On deixis in Ancient Greek see e.g.
+\citet{felson_poetics_2004}, \citet{edmunds_deixis_2008} and \citep{bonifazi_deixis_2014}.} The
 corpus analysed includes inscribed epigrams from the archaic epoch until the end of the 4th century BC, and the
@@
 the offering. The elements recurring in such inscriptions were the name of the dedicator, the verb of dedication and
-the name of the god receiving the dedication.\footnote{ On archaic dedications, see \citep[spec. 1-14]{day_archaic_2010} for
+the name of the god receiving the dedication.\footnote{ On archaic dedications, see \citet[spec. 1-14]{day_archaic_2010} for
 an introduction to the genre.} The most common and widespread formula for dedications contained exactly these three
@@
 A frequent variation was \begin{otherlanguage*}{greek}ὁ δεῖνά με ἀνέθηκε τῷ θεῷ\end{otherlanguage*}, \footnote{ The formula here presented (with its two variants) was \ identified by Maria Letizia Lazzarini, in
-her study on the formulas of archaic dedications \citep[58-60]{lazzarini_formule_1976}. Her analysis is based on both verse and prose
+her study on the formulas of archaic dedications \citet[58-60]{lazzarini_formule_1976}. Her analysis is based on both verse and prose
 inscriptions from the archaic age. However, such basic scheme, with these elements, was normal in epigrams as well and
@@
 examples the speaking object was a recurring feature in dedicatory epigrams and it is still frequent in the 4th century
-BC.\footnote{ On the topos of the speaking object, see \citep{burzachechi_oggetti_1962}, \citep[36-52]{svenbro_phrasikleia:_1988}, \citep[16-27]{tueller_look_2008}, \citep[151-166]{furley_life_2010}, \citep[250-260]{wachter_origin_2010}, \citep[29-107]{christian_gebildete_2014}.} Clearly, the basic scheme here
+BC.\footnote{ On the topos of the speaking object, see \citet{burzachechi_oggetti_1962}, \citet[36-52]{svenbro_phrasikleia:_1988}, \citet[16-27]{tueller_look_2008}, \citet[151-166]{furley_life_2010}, \citet[250-260]{wachter_origin_2010}, \citet[29-107]{christian_gebildete_2014}.} Clearly, the basic scheme here
 presented could be varied by omitting some elements (as the name of the dedicator) or adding others (as the generic
@@
 
-A crucial moment in the evolution of the epigrammatic genre is the beginning of the Hellenistic era, when epigrams
-began to be considered full literary creations and started to be circulated autonomously, no longer limited to one
+A crucial moment in the evolution of the epigrammatic genre is the beginning of the Hellenistic era, when the genre becomes more and more popular and epigrams started to be circulated autonomously, no longer limited to one
 inscription alone.\footnote{ This trend may be linked to the custom prevalent in the 5th and 4th century BC to copy
-epigrams on monuments and then quote them in various texts. See \citep[47ff.]{gutzwiller_poetic_1998}} This development led to the
+epigrams on monuments and then quote them in various texts. See \citet[47ff.]{gutzwiller_poetic_1998}} This development led to the
 composition of literary epigrams not intended for inscription, but which in some way maintained the illusion of a
@@
 be reinvented. The primary reason for such changes is self-evident: the monument intended for inscription and the space
-surrounding it do not exist anymore – they have to be imagined into reader's mind.\footnote{ This involvement of the
-reader can be put in relation with what \citep{Bing1995} defines Ergänzungsspiel, i.e. with the process of supplementation
+surrounding it do not exist anymore – they have to be imagined in the reader's mind.\footnote{ This involvement of the
+reader can be put in relation with what \citet{Bing1995} defines Ergänzungsspiel, i.e. with the process of supplementation
 of the text deliberately incited by Hellenistic epigrammatists.}
@@
 dedicated object itself. In this case, the loss of the original context forces the author to re-elaborate traditional
-models, by adding some elements (e.g. the specific name of the dedicated object) which cannot be retrieved from the
+models, by adding some elements (e.g. the specific name of the dedicated object) that cannot be retrieved from the
 surrounding setting.
@@
 
-The analysis of such deictic expressions can help determine the deictic centre in dedicatory epigrams. The deictic
+The analysis of deictic expressions can help determine the deictic centre in dedicatory epigrams. The deictic
 centre, which normally coincides with the origin of the utterance, works as a point of reference for all deictic
@@
 something close (such as the proximal \begin{otherlanguage*}{greek}ὅδε\end{otherlanguage*} this here'\footnote{ `Proximal' deictic elements are all those
-elements which refers to the deictic centre: among these demonstrative pronouns or adverbs such as `this' or `here' and
+elements that refers to something that is close to the deictic centre: among these demonstrative pronouns or adverbs such as `this' or `here' and
 temporal adverbs such as `now'. See \citet[2408f.]{diessel_deixis_2012}.}) the occasional reader will look for the referents in the
 space close to the inscription. In the evolution of the epigrammatic genre, the loss of the material context produces
-an important change, since the strong relation that connects the text with its object starts to fade. This means that
+an important change, since the strong relation that connects the text with its object starts to fade.
+In other words, when the epigram is found in a book, the point of reference --- or deictic centre --- is not immediately clear to the reader, since it is not immediately clear the relation of the text with the spatial setting to which it refers.
+This means that
 the deictic centre is somehow released from its traditional location on the object and the poet is free to consider and
@@
 support. Later on, only the possibility of the separation from a unique location allows the epigram to be circulated
-autonomously and reach a wider audience. In this passage, the poet explores new uses of deictic markers in order to
-widen the possible references of his text.
+autonomously and reach a wider audience. In this passage, poets explore new uses of deictic markers in order to
+widen the possible references of their text.
 
@@
 The corpus of epigrams analysed includes dedicatory verse-inscriptions collected in Hansen's Carmina Epigraphica Graeca
-[CEG 1-2], as for epigrams on stone. In this case, I selected those epigrams included by Hansen in the section “Tituli
-dedicatorii”. For literary epigrams, I followed Gow's and Page's Hellenistic epigrams \citep{gow_greek_1965}, to which I added
+[CEG 1-2], as for epigrams on stone. In this case, I selected those epigrams included by Hansen in the section ``Tituli
+dedicatorii''. For literary epigrams, I followed Gow's and Page's edition \citep{gow_greek_1965}, to which I added
 the epigrams attributed to Posidippus and edited by Austin and Bastianini \citep{austin_posidippi_2002}. In this case, the
@@
 for each epigram a complete profile, which includes linguistic, literary, historical and archaeological
-aspects.\footnote{ The software used is FileMaker Pro 13.0v4.} The data were organised in three main tables, connected
-to each other through a central one (See Fig.~\ref{fig:1}).
+aspects.\footnote{ The software used is FileMaker Pro 13.0v4.} The data were organised in various tables, which are all interconnected (See Fig.~\ref{fig:1}).
 
@@
 of the dedicated object (dedicatory\_context).
-
 Such structure offers a balance between the need to record a detailed picture for each epigram analysed and the
@@
 specific feature (and its interconnection with other related aspects) within the corpus and to leave out information
-that is not immediately consistent with the research done. When necessary, the tool `portal' helps to find the data
-rapidly and to combine the data contained in different tables. In this way, for example, it is possible to look for the
+that is not immediately consistent with the research done.
+However, the fact that all the tables are interconnected enables the user to formulate queries that combine fields contained in different tables.
+In this way, for example, it is possible to look for the
 recurrence of a specific tense for the verb of dedication (in the table `deixis') and to see if this is associated with
@@
 deixis and verbal tense see \citet[677-690]{lyons_semantics_1977}; \citet[73-79]{levinson_pragmatics_1983}; \citet[14-26 and 120-130]{klein_time_1994}. For the
-analysis of temporal deixis in Ancient Greek texts see e.g. \citep{dalessio_past_2004} and \citet[8f.]{edmunds_deixis_2008} (with further
+analysis of temporal deixis in Ancient Greek texts see e.g. \citet{dalessio_past_2004} and \citet[8f.]{edmunds_deixis_2008} (with further
 bibliography).} In particular, the present tense usually decodes the present time, which is the time of the deictic
@@
 On the relation between the temporal frame of the utterance and the deictic centre, see \citet[79f.]{levinson_pragmatics_1983} and
-\citep{dalessio_past_2004}.} \
+\citet{dalessio_past_2004}.} \
 
@@
 
-These figures, which obviously requires more in-depth analysis, are an interesting sign of the transformation and
+These figures, which obviously require more in-depth analysis, are an interesting sign of the transformation and
 development of the epigrammatic genre in the Hellenistic epoch. Though the poets still move along the path of the
@@
 
-\noindent The database presented here is designed to be a helpful tool in the study of heterogeneous material. This tool will
-allow us to apply an interdisciplinary approach, which combines epigraphic, linguistic and philological strategies. It
-not only helps to manage a large quantity of data, it also organises the results of the analysis on the texts in a
-practical way.
-
-The coherent organisation of the data into a database has two clear advantages. On the one hand, the database provides
-an overview of the whole corpus. This could be used, for example, to detect easily specific trends and recurring
-elements in the corpus, or to consult and combine the data from different points of view. On the other hand, it is
-possible to retrieve rapidly the data connected to each specific text. This comprehensive look, on the whole as well as
-in the particular, would be difficult to obtain otherwise and it makes the database a fundamental tool for dealing with
-complex materials and applying different levels of analysis.
-
-It is important to stress that this database aims to enable many possible, different research. The structure is not
-based on a single, pre-fixed hypothesis of investigation, but is built so as to let the user interrogate the data in
-different ways, without defining a priori the direction of the research.
-
-Moreover, the database is a dynamic tool and suitable for further and continual additions. This feature is
-particularly valuable in the field of the epigrammatic poetry, where continual discoveries of new material (on stone,
-but also on papyrus) requires us to enlarge and re-work the corpora constantly. In this regard, the database presented
-here is not only a response to the recent need, in the specific field of epigrammatic studies, to create corpora that
-combine epigraphic and `literary' materials, it is also open to future additions and new research.
+\noindent The database presented here is designed to be a helpful tool in the study of heterogeneous material.
+This tool will allow us to apply an interdisciplinary approach, which combines epigraphic, linguistic and philological strategies.
+It not only helps to manage a large quantity of data, it also organises the results of the analysis on the texts in a practical way.
+The coherent organisation of the data into a database has two clear advantages.
+On the one hand, the database provides an overview of the whole corpus.
+This could be used, for example, to detect easily specific trends and recurring elements in the corpus, or to consult and combine the data from different points of view.
+On the other hand, it is possible to retrieve rapidly the data connected to each specific text.
+This comprehensive look, on the whole as well as in the particular, would be difficult to obtain otherwise and it makes the database a fundamental tool for dealing with complex materials and applying different levels of analysis.
+It is important to stress that this database aims to enable many possible, different research.
+The structure is not based on a single, pre-fixed hypothesis of investigation, but is built so as to let the user interrogate the data in different ways, without defining a priori the direction of the research.
+Moreover, the database is a dynamic tool and suitable for further and continual additions.
+This feature is particularly valuable in the field of the epigrammatic poetry, where continual discoveries of new material (on stone, but also on papyrus) require us to enlarge and re-work the corpora constantly.
+In this regard, the database presented here is not only a response to the recent need, in the specific field of epigrammatic studies, to create corpora that combine epigraphic and `literary' materials, it is also open to future additions and new research.
 

### 2016-02-09 source/31/EAGLE2016_I Sicily_final copy.tex
corrections to Prag

@@
 %%%%%%%%
-\title{I.Sicily. An EpiDoc corpus for ancient Sicily}
+\title{I.Sicily. An EpiDoc Corpus for Ancient Sicily}
 \author[OXF]{Jonathan Prag\corref{first}}
@@
 \noindent \emph{I.Sicily} is an online, open access, digital corpus of the inscriptions on stone from ancient Sicily.\footnote{ The
-corpus will be mounted at \url{www.csad.ox.ac.uk/sicily/isicily}, but is presently on a development server. A public face is
+corpus will be mounted at \url{www.sicily.classics.ox.ac.uk/isicily/}, but is presently on a development server. A public face is
 currently maintained via a blog at \url{http://isicily.wordpress.com/}, as well as on Facebook at \url{www.facebook.com/ISicily}
@@
 At the time of writing (September 2015), the main conversion routine is being refined and the epigraphic texts are being
-collated for incorporation into the records. An ancillary database of museum collections in Sicily has been constructed
+collated for incorporation into the records. An ancillary database of museum collections and archaeological sites in Sicily has been constructed
 and bibliography is held in a Zotero library. Extensive search facilities will be provided, including map-based and
@@
 Recent overview of much of the linguistic tradition in \citet{tribulato_language_2012}; and of the epigraphic material in \citet{gulletta_sicilia_1999}.} on the other hand, the publication of this material has a very uneven record and despite an excellent
-pre-twentieth century tradition, the existing \emph{copora} are far from complete and the ability of key journals such as \emph{SEG}
+pre-twentieth-century tradition, the existing \emph{copora} are far from complete and the ability of key journals such as \emph{SEG}
 or \emph{AE} to keep pace with local publication has been limited.\footnote{ For an overview of the \emph{corpora} tradition up to
@@
 powerful tool for future study. Although \emph{I.Sicily} in its first phase is not undertaking morphological or syntactical
-mark-up, the encoding of all these text in XML constitutes the necessary first stage in such a development, and we see
+mark-up, the encoding of all these texts in XML constitutes the necessary first stage in such a development, and we see
 this as a highly desirable future project, and the possibilities for the field of historical linguistics are
@@
 
-\url{http://csad.ox.ac.uk/sicily/inscription/ISic1234}
+\url{sicily.classics.ox.ac.uk/isicily/inscriptions/1234}
 
@@
 
-The richness of \emph{I.Sicily}’s records in this area mean that \emph{I.Sicily} is currently collaborating with both \emph{Trismegistos} and
+The richness of \emph{I.Sicily}’s records in this area means that \emph{I.Sicily} is currently collaborating with both \emph{Trismegistos} and
 \emph{IDES} (`Integrating Digital Epigraphies').\footnote{\url{www.trismegistos.org/} [accessed 26.09.2015] and \url{http://ides.io/}
@@
 include); the latter is to assist \emph{IDES} in the refining of links between, e.g., \emph{PHI} and \emph{SEG} records, and to improve
-\emph{I.Sicily}’s own recording of PHI numbers.
+\emph{I.Sicily}’s own recording of \emph{PHI} numbers.
 
@@
 
-\url{http://csad.ox.ac.uk/sicily/museum/SicMus123}
+\url{sicily.classics.ox.ac.uk/isicily/museums/123}
 

### 2016-02-09 source/21/EAGLE2016_Full_Paper_revised.tex
axon corrections done

@@
 %%%%%%%%
-\title{AXON. A Selection of Greek Historical Inscriptions. A Database for Research and Teaching}
+\title{AXON. A Database for Greek Historical Inscriptions.}
 \author[cafo]{Stefania De Vido}
@@
 
-\noindent \emph{AXON. A Selection of Greek Historical Inscriptions} is a project conceived within the Greek Epigraphy Laboratory
-(Director, Prof. Claudia Antonetti), and has been brought into existence with the financial support of the University
-Ca’ Foscari of Venice (University Project 2013, Project Coordinator, Prof. Stefania De Vido).
+\noindent \emph{AXON.
+A Selection of Greek Historical Inscriptions} is a project conceived within the Greek Epigraphy Lab (Director, Prof.
+Claudia Antonetti), and has been brought into existence with the financial support of the University
+Ca’ Foscari of Venice (University Project 2013, Project Coordinator, Prof. Stefania De Vido; Scientific Team: Ivan Matijašić, Silvia Palazzo, Michela Socal, Luigi Tessarolo [IT Project]).
 
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016FullPaperrevised-img001.png}
+ \includegraphics[width=\columnwidth]{Fig1.png}
 \caption{Window for the input of data, section Text/Chronology}
@@
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016FullPaperrevised-img002.png}
+ \includegraphics[width=\columnwidth]{Fig2.png}
 \caption{Example of Abstract in AXON. The different colours allow a selection of words from a drop-down menu. }
@@
 Our aim is to establish a growing community of experts, students, and enthusiasts to increase the number of contributors
-through lists of inscriptions which have not yet been assigned. At the same time it will be the possible to suggest
+through lists of inscriptions which have not yet been assigned. At the same time it will be possible to suggest
 other texts which are not included in the lists. To achieve these aims, the project follows an EpiDoc-friendly
@@
 \begin{enumerate}
+\renewcommand{\theenumi}{\alph{enumi}}
 \item browse all the entries according to the inscriptions’ a) typology, b) chronology, and c) area of origin;
@@
 \begin{enumerate}
+\renewcommand{\theenumi}{\alph{enumi}}
 \item bibliography
@@
 
-Other filters will be employed for each search result. A section entitled ``tools'' will also be available, and will
-include information on the entries’ structure, tables with the contents of different categories, links to Vocabularies
+Other filters will be employed for each search result. A section entitled ``tools'' is also available, and it
+includes information on the entries’ structure, tables with the contents of different categories, links to Vocabularies
 and websites, etc.

### 2016-02-09 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
Corrections to Felle

@@
 
-After storing in EDB other 10000 inscriptions, since 2012 to the present day, the percentage of figured inscriptions is
+After storing in EDB other around 10000 inscriptions, since 2012 to the present day, the percentage of figured inscriptions is
 still the same: then, I think that we are able to consider this datum enough sure; so we are able to partially correct
@@
 the Chi-Rho monogram is (\citealp[68-72]{carletti_epigrafia_2008}; \citealp[365-366]{felle_judaism_2007}). But we have to underline that the phenomenon is
-not exclusive of Christian patrons. We observe more and more recurrent similar ``signs of
-identity'' also among inscriptions commissioned by Jews, not only in Rome but also in other contexts in
+not exclusive of Christian patrons.
+We observe more and more recurrent similar ``signs of religious identity'' also among inscriptions commissioned by Jews, not only in Rome but also in other contexts in
 Late Antiquity world where they were (\citealp[passim]{felle_judaism_2007}; \citealp{pesce_documenti_2016}).
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!htbp]
 \centering
@@
  \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
-\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.12.14) and
+\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from \emph{Iscrizioni} 1997, sch. 3.12.14) and
 edition in ICVR, VII 19820.}
@@
 
-\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.8.3) and edition in
+\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from \emph{Iscrizioni} 1997, sch. 3.8.3) and edition in
 ICVR, VIII 22407.}
@@
 were able to do it) but also - and, maybe, firstly - as objects to be seen. I think that we have to realize this
-perspective - common in Christian epigraphy studies after all - to understand the communication power of this
+perspective - common in Christian epigraphy studies, after all - to understand the communication power of this
 communicative \emph{medium}. Indeed, the topic has been assumed as main theme of the last International Congress of Greek and
@@
 In our projects - we have to admit it - the notion that inscriptions are \emph{essentially} texts is still largely
-prevailing: but now we can - consequently, we must - to increase our capability to represent, to record and so to
+prevailing: but now we can --- consequently, \emph{we must} --- to increase our capability to represent, to record and so to
 investigate also other, visual features of inscriptions.
@@
 positioning of the inscriptions still found in their original spot, by sharing data with the Domitilla-Projekt (by the
-Österreichische Akademie der Wissenschaften and the Deutsches Archäologisches Institut) focused to the frescoes of the
+Österreichi\-sche Akademie der Wissenschaften and the Deutsches Archäologisches Institut) focused to the frescoes of the
 catacomb of Domitilla along via Ardeatina, the largest one in Rome suburb \citep{zimmermann_epigraphy_2014}. The positive
@@
 inscriptions {}- also lost or incomplete - conceived to be seen and read exactly there. An effective example can be
-the dedication by pope Damasus (366-384) to the martyr \emph{Ianuarius} in the catacomb of Praetextatus on the via Appia (Fig.~\ref{fig:6}).
+the dedication by pope Damasus (366-384) to the martyr \emph{Ianuarius} in the catacomb of Praetextatus along the via Appia (Fig.~\ref{fig:6}).
 
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
- \includegraphics[scale=0.06]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}
-\caption{Thessaloniki, city walls. Inscriptions in the masonry of a tower near Eptapyrgion (photo: A. E. Felle).}
+ \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}
+\caption{Thessaloniki, city walls. Inscriptions in the masonry of the so-called Lapardas Tower near `Eptapyrgion' (Komnenian period, XII cent.) (photo: A. E. Felle).}
 \label{fig:7}
@@
 \caption{Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
-“useless” caption  s(an)c(t)a Agnes (photo: A. E. Felle).}
+``useless'' caption  \emph{s(an)c(t)a Agnes}, age of pope Honorius I, 625/638. (photo: A. E. Felle).}
 \label{fig:8}
@@
 
-On the open \emph{codex} is inscribed a focused - but barely readable - quotation from John 10, verses 7 and 9, where Jesus
-indicates himself as the gate:
-
-\begin{quotation}
-John 10.7: \begin{otherlanguage*}{greek}
-Εἶπεν οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι ἐγώ εἰμι ἡ θύρα τῶν προβάτων.
-\end{otherlanguage*} (\emph{Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep});
-
-John 10, 9: \begin{otherlanguage*}{greek}
-ἐγώ εἰμι ἡ θύρα· δι' ἐμοῦ ἐάν τις εἰσέλθῃ σωθήσεται καὶ εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει
-\end{otherlanguage*}  ( $^\textit{9}$ \emph{I am the gate;
-whoever enters through me will be saved.} $^{[a]}$ \emph{They will come in and go out, and find pasture}).
-
-\end{quotation}
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 
+\newpage
+On the open \emph{codex} is inscribed a focused - but barely readable - quotation from John 10, verses 7 and 9, where Jesus
+indicates himself as the gate:
+
+\begin{quotation}
+John 10.7: \begin{otherlanguage*}{greek}
+Εἶπεν οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι ἐγώ εἰμι ἡ θύρα τῶν προβάτων.
+\end{otherlanguage*} (\emph{Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep});
+
+John 10, 9: \begin{otherlanguage*}{greek}
+ἐγώ εἰμι ἡ θύρα· δι' ἐμοῦ ἐάν τις εἰσέλθῃ σωθήσεται καὶ εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει
+\end{otherlanguage*}  ( $^\textit{9}$ \emph{I am the gate;
+whoever enters through me will be saved.} $^{[a]}$ \emph{They will come in and go out, and find pasture}).
+
+\end{quotation}
+
 This the text of the inscription \citep[n. 505]{felle_biblia_2006}:
@@
 ((crux)) \begin{otherlanguage*}{greek}
-εἶπεν ὁ κ(ύριο)ς \textlatin{|} ἐγώ εἰμι \textlatin{|} ἡ θύρα τῶν | προβάτων· \textlatin{|} δι᾿ ἐμοῦ \textlatin{||} ἐάν τις \textlatin{|} εἰσέλθῃ \textlatin{|} εἰσελεύσετ(αι) \textlatin{|} κ(αὶ) ἐξελεύσετ(αι) \textlatin{|} κ(αὶ) νομὴν \textlatin{|} εὑρήσει.
+εἶπεν ὁ κ(ύριο)ς \textlatin{|} ἐγώ εἰμι \textlatin{|} ἡ θύρα τῶν \textlatin{|} προβάτων· \textlatin{|} δι᾿ ἐμοῦ \textlatin{||} ἐάν τις \textlatin{|} εἰσέλθῃ \textlatin{|} εἰσελεύσετ(αι) \textlatin{|} κ(αὶ) ἐξελεύσετ(αι) \textlatin{|} κ(αὶ) νομὴν \textlatin{|} εὑρήσει.
 \end{otherlanguage*}
@@
 (not necessarily the readability) of the sacred words give proper and strong sense to their material support and to
-entire context, the Royal Gates of the Great Church of Constantinoples \citep[320 and passim]{felle_esporre_2015}.
+entire context, the Royal Gates of the Great Church of Constantinoples \citep[320 and \emph{passim}]{felle_esporre_2015}.
+
+\vspace{2cm}
 
@@
 building. Then, we have to face the issue of the description of the writing not only from the necessary point of
-view of paleography ( we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
+view of palaeography ( we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
 also in order to perceive and to understand its non-verbal significance: by its disposition, direction, shape. The
-notion of the inscriptions in the Islamic world, where often the letters are also - and maybe firstly - images (they
-are used as decorative friezes, architectural ornaments, figures) and their clearity and readability are not considered
-as necessary, can help us to evocate this feature of the writing (Fig.~\ref{fig:10}).
+notion of the inscriptions in the Islamic world --- where often the letters are also and maybe firstly images (they
+are used as decorative friezes, architectural ornaments, figures) and their clarity and readability are not considered
+as necessary --- can help us to evocate this feature of the writing (Fig.~\ref{fig:10}).
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 features'' that in our projects - first of all in EDB, of course - are not too considered, although they
-are very significant. We need, about encoding these non-verbal features, the same positive results that by Epidoc we
+are very significant. We need, about encoding these non-verbal features, the same positive results that by EpiDoc we
 reached in encoding the texts: a hard challenge.
+
+\nocite{Manzella1997Felle}
 

### 2016-02-09 source/28/EAGLE2016_Rocco_engrev.tex
rocco updated

@@
 \title{EDB 2.0 
-How Eagle Europeana project improved the Epigraphic Database Bari}
+How EAGLE improved the Epigraphic Database Bari}
 \author[uniba]{Anita Rocco \corref{first}}
-\address[uniba]{Dip. di Scienze dell'Antichità e del Tardo Antico, Università degli Studi di Bari ``Aldo Moro''}
+\address[uniba]{Università degli Studi di Bari ``Aldo Moro''}
 \cortext[first]{Corresponding author. Email: anita.rocco@uniba.it}
@@
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img001.png} 
+ \includegraphics[width=\columnwidth]{Fig1NEW.png} 
 \caption{The record of a bilingual inscription (ICVR V, 15147) as it appeared in data processing program ICVR in MS-DOS
@@
 codes of few characters, according to a limitation imposed by the program (Fig.~\ref{fig:1}).
+%
+%\begin{figure}[hbp]
+%\centering
+% \includegraphics[scale=0.4]{Fig.2.png} 
+%\caption{The inscription ICVR V, 15147 as it appears now in EDB, with the acronyms expanded, the image and the text
+%displayed in Greek characters.}
+%\label{fig:2}
+%\end{figure}
+%
 
-\begin{figure}[hbp]
-\centering
- \includegraphics[scale=0.4]{EAGLE2016Roccoengrev-img002.jpg} 
-\caption{The inscription ICVR V, 15147 as it appears now in EDB, with the acronyms expanded, the image and the text
-displayed in Greek characters.}
-\label{fig:2}
-\end{figure}
-
-Adding \emph{metadata} to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
+Adding metadata to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
 the epigraphic object in its widest sense as an inscribed artifact. This feature, in particular relating to
@@
 
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{Fig4.png}
+\caption{Bibliographic data fields}
+\label{fig:3b}
+\end{figure}
+
 The whole bibliography is structured in metadata and is available both on the EDB website, and on the EAGLE BPN group of
@@
 To complete information on spatial data, all cemeterial contexts have been georeferenced, so that clicking on their name
-opens a new window in \emph{Google Map} that shows the modern entrance to the cemetery, with the address and its coordinates
-(Fig.~\ref{fig:6}).
+opens a new window in \emph{Google Map} that shows the modern entrance to the cemetery, with the address and its coordinates.
 
@@
  \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
-\caption{The \emph{original context} maps.}
+\caption{The original context maps.}
 \label{fig:6}
@@
 gallery or \emph{cubiculum} (F04, in Fig.~\ref{fig:6}), opens the plan of the specific area (\emph{regio}) in which the inscription is found.
-In every plan, the inscriptions preserved in situ are placed and marked by ICVR and EDB number (Fig.~\ref{fig:6})\footnote{The
+In every plan, the inscriptions preserved \emph{in situ} are placed and marked by ICVR and EDB number (Fig.~\ref{fig:6})\footnote{The
 plans are available on START-Projektwebsite: \url{www.oeaw.ac.at/antike/index.php\&id=431}, clicking on EDB number opens a
@@
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img007.png}
+ \includegraphics[width=\columnwidth]{Fig7.png}
 \caption{Conservation input fields.}
@@
 uniquely the place of conservation, as well as to obtain additional information, included the geographic
-positioning\footnote{\url{http://www.eagle-network.eu/wp-content/uploads/2013/06/EAGLE\_D2.2.2\_Content-harmonisation-guidelines-including-GIS-and-terminologies-Second-Release.pdf}}.
+positioning.\footnote{\url{http://www.eagle-network.eu/wp-content/uploads/2013/06/EAGLE\_D2.2.2\_Content-harmonisation-guidelines-including-GIS-and-terminologies-Second-Release.pdf}}
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{Fig8.png}
+\caption{Conservation maps and links}
+\label{fig:8new}
+\end{figure}
 
@@
 
-\begin{figure}[hbp]
+\begin{figure}[!hbp]
 \centering
@@
 
-\nocite{felle_fenomeni_2007,felle_esperienze_2012,jory_problems_1975}
+\nocite{felle_fenomeni_2007}
+\nocite{felle_esperienze_2012}
+\nocite{jory_problems_1975}
+\nocite{Rocco2015a}
 
@@
 \begin{description}
-\item[ CIL ] = \emph{Corpus Inscriptionum Latinarum}, \ Berolini 1863 ss.
-\item[CIG ] = \emph{Corpus Inscriptionum Grecarum}, I-IV, Berolini 1828-1877
+\item[CIL] = \emph{Corpus Inscriptionum Latinarum}, \ Berolini 1863 ss.
+\item[CIG] = \emph{Corpus Inscriptionum Grecarum}, I-IV, Berolini 1828-1877
 \item[IC] = \emph{Inscriptiones christiane urbis Romae septimo saeculo antiquiores}, I-II, G.B. de Rossi (a c.), Romae 1857-1861; \emph{Supplementum}, G. Gatti (a c.), Romae 1915. 

### 2016-02-09 source/10/Paper proposal for EAGLE final.tex
graham and licciardello biblio updated
lower case titles
graham updated

@@
 %%%%%%%%
-\title{FACE VALUE: THE POWER OF IMAGES AT APHRODISIAS. How Digital Resources Can Transform our use of Palaeography in understanding Inscriptions.}
+\title{The Power of Images at Aphrodisias: 
+How Digital Resources Can Transform our assessment of Palaeography}
 \author[war]{Abigail S. Graham\corref{first}}
@@
 into that category of human endeavor known as stylistic attribution and inevitably involve subjectivity''.\footnote{\citet[3-4]{tracy1995}} Stephen Tracy advises that palaeographic surveys should be carried out with caution: ``Caveat Lector
-must needs be our motto''. The study of palaeography is problematic on a number of levels, both in the way it is
+must needs be our motto''. The study of palaeography is problematic, as demonstrated by W. Eck and C. Roueché during the conference, on a number of levels, both in the way it is
 defined and the ways in which it is employed. Many scholars dismiss lettering as a means of dating and they make an
@@
 a collectivity, and for this reason is made with the location, writing technique, graphic form and impagination, mode
-and register of expression chosen because they are most suitable to the attainment of its intended goal.'' \citep[8]{panciera
-2012}.} In this vein, it is worth considering how the physical characteristics of an inscription belong within a
+and register of expression chosen because they are most suitable to the attainment of its intended goal.'' \citep[8]{panciera2012}.} In this vein, it is worth considering how the physical characteristics of an inscription belong within a
 broader assessment of a culture of writing. Is palaeography the study of letterforms alone? Was a focus on letterforms
@@
 \noindent The focus of this paper is a series of rather unimpressive column dedications from the temple of Aphrodite at
-Aphrodisias. While the information recorded in the texts is unremarkable, the journey of these inscriptions from
-their publication as part of an epigraphic corpus to a recent digital publication reveals an epic transformation in the
+Aphrodisias.
+While the information recorded in the texts is unremarkable, the journey of these inscriptions from a paper epigraphic corpus to their recent publication as a digital corpus reveals an epic transformation in the
 format and approach to these materials. This study will begin with brief overview of previous published editions
-(e.g. \citet{mama8}) followed by an assessment of the information available on the current Inscriptions of
+(e.g. \citet{calder1962}) followed by an assessment of the information available on the current Inscriptions of
 Aphrodisias (2007) website. Through careful assessment of text (including formula, vocabulary and spelling) and its
@@
 euergetes, are absent from Zoilos’ other inscriptions and may imply a posthumous dedication, perhaps by the boule and
-demos \citep[38]{reynolds 1990}.} The benefactor of the columns in this case study, Eumachos Diogenes, also comes from an
-established family that flourished well into the 2nd century AD.\footnote{18 The success of Eumachus Diogenes’ family,
+demos \citep[38]{reynolds1990}.} The benefactor of the columns in this case study, Eumachos Diogenes, also comes from an
+established family that flourished well into the 2nd century AD.\footnote{The success of Eumachus Diogenes’ family,
 which included the first known Aphrodisian to hold a procuratorship in the 2nd century AD, reveals an enduring
@@
 1129) and in a monument to Ti. Cl. Drusus in Patara (SEG XLIV (1994) no. 1205). For significance and date of the title
-see \citet[101-105]{buraselis2001}.} Before encountering the inscriptions face to face, the dating, function and meaning of
+see \citet[101-105]{buraselis2000}.} Before encountering the inscriptions face to face, the dating, function and meaning of
 these inscriptions appears to be quite straightforward. So let us examine the experience of viewing these
@@
 ς φιλόκαισαρ καὶ Ἀμιὰς\\
-Διονυσίου φύσι δὲ Ἀδρά<σ>του\\
+Διονυσίου φύσι δὲ Ἀδρά\textlatin{<}σ\textlatin{>}του\\
 τοῦ Μόλωνος Ὀλυνπιὰς\\
@@
 The Inscriptions of Aphrodisias project was a groundbreaking endeavour both for publishing epigraphic materials and
-employing Epi Doc conventions with XML for marking up ancient documents. Its success has inspired a number of
+employing Epi Doc conventions with XML for marking up the text of an inscription. Its success has inspired a number of
 significantly larger projects with international scope and collaboration. The approach to the materials is detailed
@@
 wealth of superlative scholarship).\footnote{ It is worth noting that websites, however they develop, will remain
-supplementary to published scholarship.} In addition to search functions, the page is more interactive, the texts
+supplementary to published scholarship such as J. Reynolds wonderfully detailed \emph{Aphrodisias and Rome} (1982).} In addition to search functions, the page is more interactive, the texts
 have been revisited, translations added, as well as commentary with weblinks to parallel texts. The organization of the
@@
 
+\newpage
 In addition to the format of the webpage, a number of images have been published along with a drawing from Deering’s
@@
 decorative function (e.g. giving distinction to a name or key elements/individuals in the text (cf. building
-dedications at the Sebasteion IAph 9.1, 9.25, 9.112). This vacat serves little grammatical function, and is likely to
+dedications at the Sebasteion IAph 9.1, 9.25, 9.112)). This vacat serves little grammatical function, and is likely to
 be a result of the arrangement of the text/ omission in Ammias’ name. } The indentation on the last line also reveals
@@
 or word), which is more common in later Imperial inscriptions at Aphrodisias. 
-
+\newpage
 \begin{quotation}
@@
 oval shape with serifs only at one end of the lower bars.\footnote{ ``Legged'' rhos are observed primarily in Augustan
-and Julio-Claudian dedications and also on coins (cf. note 16). This style letterform is increasingly less common after
+and Julio-Claudian dedications and also on coins (cf. note 15). This style letterform is increasingly less common after
 Flavian period at Aphrodisias. } Comparing these letterforms with parallel dedications at the temple (IAph 1.1, 1.2
@@
 Ἀφροδίτῃ καὶ τῷ
-\end{otherlanguage*}
+\end{otherlanguage*} \\
 vac. \begin{otherlanguage*}{greek}
@@
 
-An image (Fig.~\ref{fig:3}) illustrates further differences between the appearance of this inscription
-and the comparative materials from the site (IAph 1.1, 1.2, 1.4, 1.7 and 1.8).
+A high quality image (Fig.~\ref{fig:3}) illustrates further differences between the appearance of this inscription
+and the comparative materials from the site (IAph 1.1, 1.2, 1.4, 1.7 and 1.8), revealing regular letterforms (e.g. omicrons are less oval and more precisely rendered
+throughout) with a clear contrast between slender strokes and large deep triangulate serifs. There are no legged rhos,
+the middle bars of the epsilons are separated from the stem, and the omegas are circular with two large bars that are
+heavily serifed on both ends (e.g. compare the omega on the bottom line of IAph 1.4 with its counterpart of IAph.1.5).
 
-\begin{figure}[!bp]
+\newpage
+
+\begin{figure}[!hbp]
 \centering
@@
 
-The high quality image reveals square and regular letterforms (e.g. omicrons are less oval and more precisely rendered
-throughout) with a clear contrast between slender strokes and large deep triangulate serifs. There are no legged rhos,
-the middle bars of the epsilons are separated from the stem, and the omegas are circular with two large bars that are
-heavily serifed on both ends (e.g. compare the omega on the bottom line of IAph 1.4 with its counterpart of IAph.1.5).
 These traits along with others (e.g. letter size and use of spacing noted above) are commonly attributed to late 1st
@@
 added under the emperor Hadrian, a time during which Athenagoras’ descendants were alive and prospering in
-Aphrodisias.\footnote{\citep[43]{smith1995}. For an analysis of the temenos plan, see \citet[66-74]{doruk1990}. For Athenagoras descendants, cf. note 18 and \citet[327-334]{reynolds1999b}.}
+Aphrodisias.\footnote{\citep[43]{smith1995}. For an analysis of the temenos plan, see \citet[66-74]{doruk1990}. For Athenagoras descendants, cf. note 17 and \citet[327-334]{reynolds1999}.}
 
@@
 δὲ Ἀδράστου τοῦ Μό-\\
-λω<ν>ος Ὀλυ<ν>πιὰς τὸν κί-\\
+λω\textlatin{<}ν\textlatin{>}ος Ὀλυ\textlatin{<}ν\textlatin{>}πιὰς τὸν κί-\\
 ονα θεᾷ Ἀφροδίτῃ \{τὸν\}\\
@@
 \subsection{Conclusions on the case study}
-\noindent This survey of column dedications has demonstrate how a series of inscriptions, which were represented as a series of
+\noindent This survey of column dedications has demonstrated how a series of inscriptions, which were represented as a series of
 similar texts in MAMA VIII can be seen in a fundamentally different way in IAPh 2007. The digital publication has a
 number of advantages: it applies a more rigorous approach to the text in XML thus illustrating the arrangement of the
-inscription more clearly, the date is given some transparency through a list of applied criteria, the descriptors, such
-as the lettering size of the inscriptions, provide more information. The images of the inscription in drawing and its
+inscription more clearly, the date is given some transparency through a list of applied criteria, and the descriptors, such
+as the lettering size of the inscriptions, provide more information. The images of the inscription in drawings and its
 context provide an invaluable resource that allow the reader to better understand the relationship between text,
 inscription and context, including references and better accessibility to parallel inscriptions. In terms experiencing
-an inscription in a digital world, this is probably as close a person can come to assessing the face value of an
+an inscription in a digital world, this is probably as close as a person can come to assessing the face value of an
 inscription. It is a tremendous step forward in addressing the longstanding limitations inherent in the publication
@@
 undefined.\footnote{ Studies of inscriptions describe lettering as ``distinctively Julio-Claudian'' \citep[nos.
-2\&3) 317-318]{reynolds1981}, ``Triumviral or Augustan'' \citep[docs. 35-37, 159-163]{reynolds1982}, and ‘Domitianic’ \ (Chaniotis 2004, no.
+2\&3) 317-318]{reynolds1982}, ``Triumviral or Augustan'' \citep[docs.
+35-37, 159-163]{reynolds1982}, and ‘Domitianic’ \ (Chaniotis 2004, no.
 14), suggest that such distinctions are evident at Aphrodisias.} While such definitions are often problematic, so is
@@
 of work to augment the scope of IAph 2007 from an academic resource into one also achieves a didactic aim of
-illustrating how we use epigraphic evidence to date inscriptions. We have the potential now, to present inscriptions at
+illustrating how we use different elements of an inscription (including letter forms) to suggest a date for an inscription.
+The point is not necessarily to establish a firm date, but to provide greater transparency to process by which we formulate dates.
+We have the potential now, to present inscriptions at
 face value, not only as texts but as contradictory objects whose stories, whether conveyed on stone or a computer app,
@@
 \nocite{roueche2004} 
+\nocite{buraselis2000}
+\nocite{bruun2015}
+\nocite{calabi1991}
+\nocite{chaniotis2004}
+\nocite{calder1962}
+\nocite{panciera1995}
 

### 2016-02-10 source/28/EAGLE2016_Rocco_engrev.tex
pdfs new folder, fixes to Rocco
removed main.pdf, file too large
updated committee
updated main
updated biblio

@@
 \author[uniba]{Anita Rocco \corref{first}}
-\address[uniba]{Università degli Studi di Bari ``Aldo Moro''}
+\address[uniba]{Dip. di Studi Umanistici, Università degli Studi di Bari ``Aldo Moro''}
 \cortext[first]{Corresponding author. Email: anita.rocco@uniba.it}
@@
 \caption{The record of a bilingual inscription (ICVR V, 15147) as it appeared in data processing program ICVR in MS-DOS
-in 1988.}
+in 1988 and as it appears now in EDB, with the acronyms expanded, the image and the text displayed in Greek characters.}
 \label{fig:1}
@@
 d’Épigraphie Grecque et Latine (AIEGL), as EDB (Epigraphic Database Bari) and extended its competencies to the
-epigraphic documentation of Christian patronage of the city of Rome, published after the volumes of ICVR.
+epigraphic documentation of Christian patronage of the city of Rome, published after the volumes of ICVR (Fig.~\ref{fig:3}).
 
@@
 value of the object, identifying each epigraph as a complex and polysemic product consisting partially but not only of
-text.
+text (Fig.~\ref{fig:4}).
 
@@
 Besides data related to the ICVR publication - volume and edition number of the inscription - now EDB is able to record
-other bibliographic references as well as concordances with other Corpora (CIL, IG, IGUR, IGC).
+other bibliographic references as well as concordances with other Corpora (CIL, IG, IGUR, IGC) (Fig.~\ref{fig:3b}).
 
+\newpage
 \begin{figure}[!hbp]
@@
  \includegraphics[width=\columnwidth]{Fig4.png}
-\caption{Bibliographic data fields}
+\caption{Bibliographic data fields.}
 \label{fig:3b}
@@
 
-Data on the \emph{original context} are therefore organized hierarchically in three related fields containing controlled lists.
+Data on the \emph{original context} are therefore organized hierarchically in three related fields containing controlled lists (Fig.~\ref{fig:5}). 
 After selecting the area of the suburb identified by the name of the consular road - or by the number of the Augustan
@@
 
- \begin{figure}[hbp]
-\centering
- \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
-\caption{The original context maps.}
-\label{fig:6}
-\end{figure}
-
 As a case of study, for the inscriptions \emph{in situ} pertaining to the Domitilla catacomb (Via Ardeatina) - the only
@@
 
+ \begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
+\caption{The original context maps.}
+\label{fig:6}
+\end{figure}
+
 Even the geographic data relating to the place of conservation of the inscriptions are managed with a similar structure.
@@
 associated structures, such as museums, churches or catacombs; and the specific positions in the context where the
-object is actually preserved.
+object is actually preserved (Fig.~\ref{fig:7}).
 
@@
 uniquely the place of conservation, as well as to obtain additional information, included the geographic
-positioning.\footnote{\url{http://www.eagle-network.eu/wp-content/uploads/2013/06/EAGLE\_D2.2.2\_Content-harmonisation-guidelines-including-GIS-and-terminologies-Second-Release.pdf}}
+positioning (Fig.~\ref{fig:8new}).\footnote{\url{http://www.eagle-network.eu/wp-content/uploads/2013/06/EAGLE\_D2.2.2\_Content-harmonisation-guidelines-including-GIS-and-terminologies-Second-Release.pdf}}
 
+\newpage
 \begin{figure}[!hbp]
@@
  \includegraphics[width=\columnwidth]{Fig8.png}
-\caption{Conservation maps and links}
+\caption{Conservation maps and links.}
 \label{fig:8new}
@@
 various languages the terms used by the various partners, and returns them in a format that allows the user to get a
-stable and unique identifier for each term, accessible and reusable by other users.
+stable and unique identifier for each term, accessible and reusable by other users (Fig.~\ref{fig:9}).
 
-\begin{figure}[hbp]
+\begin{figure}[!hbp]
 \centering
@@
 
+\newpage
+
 \subsection{The text}
@@
 coexistence between them), and metrical structure.
-
-\begin{figure}[hbp]
-\centering
- \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img010.png}
-\caption{Text fields in the submission mask.}
-\label{fig:10}
-\end{figure}
 
@@
 as figurative elements (fishes, birds, anchors, etc.) and Christological monograms. Other fields record onomastic
-notes, \emph{critical apparatus} and textual comments.
+notes, \emph{critical apparatus} and textual comments (Fig.~\ref{fig:10}).
 
@@
 specific time span (such as the duration of the reign of a Bishop of Rome or of an emperor), or a generic interval.
+
+\newpage
+
+\begin{figure}[!hbp]
+\centering
+ \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img010.png}
+\caption{Text fields in the submission mask.}
+\label{fig:10}
+\end{figure}
 
@@
 reference with the context, form and quality of the support, graphic forms, peculiarities of technique, layout and
-relationship between the text and any figurative or decorative elements\footnote{\citet{panciera_cronistoria_2006}.}.
+relationship between the text and any figurative or decorative elements\footnote{\citet{panciera_questioni_2006}.}.
 
@@
 search in only one of the following fields: identifier EDB, bibliographic data and text; an \emph{Advanced Search} provides
-the opportunity to explore the database through multiple search criteria variously combined.
+the opportunity to explore the database through multiple search criteria variously combined (Fig.~\ref{fig:11}).
 
@@
 \begin{enumerate}
-\item The inclusion of EDB bibliography, structured in metadata, in the Zotero Group, where it have been merged with those
+\item The inclusion of EDB bibliography, structured in metadata, in the Zotero Group, where it has been merged with those
 of other content providers and has been made directly and publicly available in the most reusable way, giving more

### 2016-02-10 source/10/Paper proposal for EAGLE final.tex
further changes to verge, Graham and fixes to Vassallo

@@
 understandable. However, treating texts as ``copies'' as opposed to individual monuments can, and in this case will,
-prove problematic.\footnote{ For another study of ``copy'' inscriptions at Aphrodisias see Graham 2015 (forthcoming).}
+prove problematic.\footnote{ For another study of ``copy'' inscriptions at Aphrodisias  \citet{Graham2015}.}
 While MAMA VIII offered a broader audience access to the inscriptions at Aphrodisias, the approach to the inscriptions
@@
 represent this information but how we engage a wider readership in epigraphic materials. It would not take a great deal
-of work to augment the scope of IAph 2007 from an academic resource into one also achieves a didactic aim of
+of work to augment the scope of IAph 2007 from an academic resource into one that also achieves a didactic aim of
 illustrating how we use different elements of an inscription (including letter forms) to suggest a date for an inscription.

### 2016-02-10 source/44/Damnjanovic_et_al_EAGLE2016.tex
further changes to verge, Graham and fixes to Vassallo

@@
 \title{Integration of Multimedia Collections and Tools
-for Interaction with Digital Content.
-The case study of the Archaia Kypriaki Grammateia Digital Corpus}
+for Interaction with Digital Content. The case study of the Archaia Kypriaki Grammateia Digital Corpus}
+\author[cyi]{Uros Damnjanovic}
 \author[cyi]{Valentina Vassallo\corref{first}}
-\author[cyi]{Uros Damnjanovic}
 \author[cyi]{Sorin Hermon}
 \date{2015-11-26}
-\cortext[first]{Corresponding author. Email: (u.damnjanovic, v.vassallo, s.hermon@cyi.ac.cy}
+\cortext[first]{Corresponding author. Email: u.damnjanovic, v.vassallo, s.hermon@cyi.ac.cy}
 \begin{document}
@@
 distinguished by its multidisciplinary structure, include all disciplines relevant for description and representation
-of epigraphies: archaeology (investigating for example the context of the finds), philology s(analyzing the text, the
+of epigraphies: archaeology (investigating for example the context of the finds), philology (analyzing the text, the
 writing style, the scripture), chemistry and geology (providing details on the material upon which inscriptions were
@@
 identified as reference item.
-\item the creation of rules for the development of a script that will be able to map to the Eagle data model every time
+\item the creation of rules for the development of a script that will be able to map to the EAGLE data model every time
 according to the type of object we are dealing with. This implies also an editing of the metadata, creating single
@@
 We presented in this paper a digital data repository and showed number of innovative tools used to access and explore
-its collections. One of the most interesting collection stored in the repository is the Archaia Kypriaki Grammateia
-Digital Corpus composed of Ancient Greek and Latin epigraphic texts. We argued that by interacting with data in an
-innovative ways can help users better understand data and stimulate knowledge creation. With the set of proposed
+its collections. We argued that by interacting with data in an
+innovative ways can help users better understand data and stimulate knowledge creation. One of the most interesting collection stored in the repository is the Archaia Kypriaki Grammateia
+Digital Corpus composed of Ancient Greek and Latin epigraphic texts. With the set of proposed
 functionalities we also enabled exploration of the epigraphic text in a way that supports sense making, understanding

### 2016-02-10 source/10/Paper proposal for EAGLE final.tex
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 of spacing and decoration, letter size). This is a suboptimal way of assessing epigraphic evidence. 
-
+\vspace{5pt}
 The advent of online corpora have increased both the access and the development of discussions regarding epigraphic
@@
 
+\vspace{5pt}
 \section{Publishing inscriptions: A Case Study of Column Dedications from The Temple Of Aphrodite at Aphrodisias}
@@
 
+\newpage
 The two published texts reveal similar inscriptions with a few variations.\footnote{ MAMA 437 and 438 are represented in
@@
 \centering
- \includegraphics[width=\columnwidth]{PaperproposalforEAGLEfinal-img001.png}
+ \includegraphics[width=0.7\columnwidth]{PaperproposalforEAGLEfinal-img001.png}
 \caption{Screenshot of IAph 4 webpage}
@@
 
+\newpage
 While traditional text-based modes of publication such as MAMA VIII have, at times, constrained the study of hands to an

### 2016-02-10 source/22/Iscrizioni Latine Arcaiche - EAGLE 2016.tex
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 be provided, clearly related with the transcription provided by the editor. 
-
+\newpage
 We offer two kinds of transcription, the ‘Diplomatic transcription' and the ‘Interpretative transcription', which

### 2016-02-10 source/28/EAGLE2016_Rocco_engrev.tex
further changes to verge, Graham and fixes to Vassallo
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 %%%%%%%%
-\title{EDB 2.0 
+\title{EDB 2.0.
 How EAGLE improved the Epigraphic Database Bari}
@@
 
+\newpage
 Obviously this step has resulted in a series of changes and adjustments that led from the original basic structure of
@@
 
-\newpage
 \begin{figure}[!hbp]
@@
 adplicitae}, in accordance with the definitions of the ICVR.
-
+\newpage
 To complete information on spatial data, all cemeterial contexts have been georeferenced, so that clicking on their name
@@
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
+ \includegraphics[width=0.9\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
 \caption{The original context maps.}

### 2016-02-10 source/3/EAGLE_2016_submission_3.tex
further changes
Bronx not moved to pdfs folder, as it is being compiled by Francesco
fixes to Varga e Moucunill
further changes to verge, Graham and fixes to Vassallo
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 
+\vspace{10pt}
 \section{Motivation}
@@
 
+\begin{figure}[!hbp]
+\centering
+\includegraphics[width=\columnwidth]{graph.png}
+\caption{Schema of the database}
+\label{fig:schemadatabase}
+\end{figure}
+
 Alongside these components, which actually represent various steps in the data preparation process, a population
@@
 
+\newpage
 The construction of the database will follow a series of steps imposed by best practices: creating a repository of
@@
 
-\begin{figure}[!bp]
-\centering
-\includegraphics[scale=0.50]{1.png}
-\caption{Schema of the database}
-\label{fig:schemadatabase}
-\end{figure}
-
 \section{The sources}
@@
 \url{www.romans1by1.com} represent the skeleton of the future population database of the Roman Empire.
-
 For now, we are solely focusing on characters attested epigraphically – thus on inscriptions as sources. As constructing
@@
 not their families), as their appurtenance to the army creates their social status.
-
 Regarding the data entry, we are trying to remain faithful to the source and to record, during the first phase, only the
@@
 support of the project. 
-
+\newpage
 Two important variables, for statistics and working with the data, are \emph{Dedicated by} and \emph{Dedicated for}, which state the
@@
 recorded individuals represent enough information. 
-\vspace{20pt}
+\newpage
 
 %\begin{table}[!h]
+\tablehead{\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline}
+
+{\small
+\addtolength{\tabcolsep}{-0.5mm}
+%\begin{supertabular*}{\textwidth}{p{3.5cm}lp{3cm}}
+\begin{xtabular}{|p{.30\textwidth}|p{.33\textwidth}|p{.33\textwidth}|}
+
+\tablehead{\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline}
+
 \tablefirsthead{\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline }
@@
 
-\tablehead{\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline}
-
 \tabletail{\hline \multicolumn{3}{|r|}{{\textit{Continued on next page}}} \\ \hline}
@@
 
-{\small
-\addtolength{\tabcolsep}{-0.5mm}
-%\begin{supertabular*}{\textwidth}{p{3.5cm}lp{3cm}}
-\begin{xtabular}{|p{.30\textwidth}|p{.30\textwidth}|p{.30\textwidth}|}
 %\begin{tabular*}{p{.30\textwidth}p{.29\textwidth}p{.31\textwidth}}
@@
 
-This work began being developed during the period of a residential scholarship at the Hardt Fondation, Vandoeuvre. Its
-expansion to the area of the whole Latin language Empire is being undertaken with the support of a postdoctoral
-fellowship of the Fritz Thyssen Stiftung.
-
 \bibliographystyle{../../sapauth-eng}

### 2016-02-10 source/31/EAGLE2016_I Sicily_final copy.tex
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 
+\vspace{20pt}
 \section{The motivations for and origins of \emph{I.Sicily}}
@@
 
+\vspace{5pt}
+
 The combination of these two factors already means that locating, identifying, or working with a Sicilian inscription,
@@
 collections, whether held in museums, in archaeological stores, or elsewhere, and the lack of consistency in the
-publication of \ new material.
+publication of new material.
+
+\vspace{5pt}
 
@@
 
+\vspace{5pt}
+
 In its final form the table holds data across 39 different fields, for 2575 records. 17 of these fields detail
@@
 this data is derived from existing publications.
+
+\vspace{5pt}
 
@@
 %\emph{I.Sicily}
+
+\vspace{5pt}
 
@@
 \emph{I.Sicily}’s own recording of \emph{PHI} numbers.
+
+\vspace{20pt}
 
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
  \includegraphics[width=0.6\columnwidth]{EAGLE2016ISicilyfinalcopy-img001.jpg}
-\caption{sample edition}
+\caption{Sample edition}
 \label{fig:1}
@@
 the open availability of the underlying XML records means that such a project would be entirely possible in the future.
-
+\newpage
 A core principal of the project is that wherever possible an inscription record should be supported by recent autopsy

### 2016-02-10 source/38/EAGLE2016_Final_Paper Modolo.tex
Paper 38 (Modolo): long url in n.6 fixed
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 
+\usepackage{url}
+\def\UrlBreaks{\do\/\do-}
 %%%%%%%%
@@
 privacy for archival documents.
-
+\newpage
 Free photography is therefore a fundamental tool which, without contradicting the demands of conservation, greatly

### 2016-02-10 source/42/Paper_licciardello.tex
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 
+\vspace{5pt}
 My research will try to detect different trends in the use of deictic expressions in order to highlight the similarities
@@
 
+\vspace{5pt}
 The corpus of epigrams analysed includes dedicatory verse-inscriptions collected in Hansen's Carmina Epigraphica Graeca
@@
 
+\vspace{5pt}
 The structure of the database was designed to include all important data related to the text examined, in order to have

### 2016-02-10 source/43/EAGLE2016_Benefiel_Sypniewski.tex
corrections made to 43, 33, 29
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 thus: \emph{herma \underline{muliebris} prospiciens} [Fig.~\ref{fig:6}].
+\newpage
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
@@
 
-\begin{figure}[!bp]
+Again, we might question whether this figure should be identified as female. In fact, we not certain that we had located
+the correct apograph for the drawing. There is considerable discrepancy between the description of the drawing in \emph{CIL}
+4, 10676 as female, with no mention of the long nose, and this line-drawing.
+
+\begin{figure}[!hbp]
 \centering
@@
 \end{figure}
-
-Again, we might question whether this figure should be identified as female. In fact, we not certain that we had located
-the correct apograph for the drawing. There is considerable discrepancy between the description of the drawing in \emph{CIL}
-4, 10676 as female, with no mention of the long nose, and this line-drawing.
 
@@
 ``\emph{retiarius}'' or we have gladiatorial equipment, such drawings will be omitted from the list of search results.
-
+\newpage
 We are therefore designing AGP with the capacity for locating figural graffiti through a two-prong solution: with
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img008.png}
+ \includegraphics[width=0.9\columnwidth]{EAGLE2016BenefielSypniewski-img008.png}
 \caption{Screenshot of AGP search engine showing browsing capabilities.}
@@
 \subsection{Limiting search results in AGP}
-
 \noindent It is therefore necessary to design a way to limit results, so that a user could find only gladiatorial equipment,
@@
 
-\begin{figure}[!bp]
+\begin{figure}[!hbp]
 \centering
- \includegraphics[width=\columnwidth]{EAGLE2016BenefielSypniewski-img009.png}
+ \includegraphics[width=0.9\columnwidth]{EAGLE2016BenefielSypniewski-img009.png}
 \caption{Screenshot showing filters in AGP.}
@@
 Or, it will be possible to do a broad search for all drawings of animals, and then filter to limit the results to find
-what kind of animals are drawn in taverns, for example, but not public buildings or houses. \
+what kind of animals are drawn in taverns, for example, but not public buildings or houses.
 

### 2016-02-10 source/48/EAGLE16-lame-et-al-teaching.tex
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 
+\newpage
  A traditional print photograph, such as Fig.~\ref{fig:12582} (No. 12582), drawn from the archives of the AMZ, limits the student’s
@@
 the transcription was not utterly difficult for most students. Obviously, in the first instance, the interpretation was
-beyond their means, but we were nevertheless impressed by their abilities to read it far more easily than expected. The
-DAP definitely offers the possibility to work on such material online, without the obligation to inspect it personally,
+beyond their means, but we were nevertheless impressed by their abilities to read it far more easily than expected.
+
+\newpage
+The DAP definitely offers the possibility to work on such material online, without the obligation to inspect it personally,
 at least in the first stage.
@@
 
+\newpage
 The older, erased layers, on the obverse of tag No. 13053, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
@@
 
+\vspace{10pt}
 \section*{Aknowledgment}

### 2016-02-10 source/8/EAGLE_2016_submission_8_revised.tex
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 contents. 
-
+\newpage
 Modelling an object biography and culturally contextualizing the texts must be initially addressed with information
@@
 particular signs that correspond to the material used or to the subject of the text, for instance.
-
 Thus, when creating a standard for non-alphabetic scripts, object metadata and annotations for orthographic and
@@
 
+\newpage
 \section[Summary ]{Summary}

### 2016-02-10 source/CNR/TheEAGLEDataAggregatorDataQualityMonitoring.tex
orfans and widows
percorsi tutti i file per eliminare orfane e vedove inserendo \newpage
o \vspace per spostare leggermente i paragrafi

@@
 
-The EAGLE project \citep[fully described in][]{eagle} aggregates epigraphy related content from about 20 different data providers (cultural institutions all over Europe), and makes its content available to both Europeana \citep{europeana} (through an OAI-PMH interface) and to scholars and the general public through a Web portal.
+\noindent The EAGLE project \citep[fully described in][]{eagle} aggregates epigraphy related content from about 20 different data providers (cultural institutions all over Europe), and makes its content available to both Europeana \citep{europeana} (through an OAI-PMH interface) and to scholars and the general public through a Web portal.
 
@@
 
+\vspace{5pt}
 Using the Data Flow and Dynamics Service provided by the Monitoring Framework, it is possible to define \textit{monitoring scenarios}, which define a conceptualization of the data flows taking place in the aggregating infrastructure.
@@
 
+\vspace{5pt}
 With the concept of sensor, we refer to a piece of software capable of generating \textit{observations} (a numeric value plus some contextual metadata) about a measured object.
@@
 
+\vspace{5pt}
 A sensor can in principle be plugged anywhere in the aggregating infrastructure; with the understanding that the implementation of the sensor and the point of the data flow where it will be placed are the responsibility of the IQM.
@@
 
+\vspace{5pt}
 The interaction between the IQM and the Monitoring Framework is done through a user friendly user interface (WebUI) provided by the DFDS.
@@
 
+\vspace{5pt}
 For the purpose of monitoring, the IQM can define \textit{controls}, i.e.
@@
 
+\vspace{5pt}
 Finally, the Data Flow and Dynamics Service enables the generation of an exhaustive \textit{report} about the defined metrics and controls providing insights, via the WebUI, in a quick glance about key features and potential issues present in the infrastructure.
@@
 
+\vspace{5pt}
 \section{Architecture}\label{arch}
 
-The Monitoring Framework (see Figure 1) is architected as a client-server application, where a core module (imported and used by the code of the aggregating infrastructure) plays the role of \textit{client}.
+\noindent The Monitoring Framework (see Figure 1) is architected as a client-server application, where a core module (imported and used by the code of the aggregating infrastructure) plays the role of \textit{client}.
 The infrastructure source code needs to be instrumented in order to put sensors in place and produce observations of the measured objects.
 
+\vspace{5pt}
 The implementation of a sensor provides the logic to produce a measurement; in general, a sensor is devised to probe a specific object type (thus the typing of the measured object is hardcoded), while its configuration can be defined via WebUI and retrieved at runtime (dynamically).
@@
 
+\vspace{5pt}
 The \textit{server} component is a stand-alone web application that receives observations from sensors, stores those observations as time series and runs automatically user-defined controls over this corpus of data.
@@
 
-In order to apply the monitoring framework described above to the EAGLE aggregator, we need to define first a monitoring scenario, based on the actual EAGLE data flow, which must include the following.
+\noindent In order to apply the monitoring framework described above to the EAGLE aggregator, we need to define first a monitoring scenario, based on the actual EAGLE data flow, which must include the following.
 \begin{itemize}
@@
 
-A monitoring scenario has been defined and tested on simulated data in order to monitor the ``health'' of different data collections involved in the EAGLE collection and transformation workflows.
+\noindent A monitoring scenario has been defined and tested on simulated data in order to monitor the ``health'' of different data collections involved in the EAGLE collection and transformation workflows.
 In particular, a first collection is stored in a full-text index (Apache Solr) serving search queries, and a second one is stored in a document store (MongoDB) serving data for OAI-PMH export.
@@
 
-Given the metrics described in \ref{sec:metrics}, Table \ref{tab:controls} reports some controls defined over those metrics.
+\noindent Given the metrics described in \ref{sec:metrics}, Table \ref{tab:controls} reports some controls defined over those metrics.
 
@@
 
-Three different sensors have been defined and implemented for this test on EAGLE: two collection sensors (one for Solr and one for MongoDB) and one single-datum sensor for XML record-by-record inspection.
+\noindent Three different sensors have been defined and implemented for this test on EAGLE: two collection sensors (one for Solr and one for MongoDB) and one single-datum sensor for XML record-by-record inspection.
 Once the three sensors have been placed in the EAGLE workflow implementation and the EAGLE infrastructure is running, they start to produce observations and deliver them to the server component of the monitoring framework.
 
+\newpage
 As an example, we report in figures included in Section \ref{sec:figures} simulated trends of the defined metrics and relative reports (i.e. the evaluation of controls defined over the metrics).
@@
 
+\newpage
 \section{Conclusions and future work}\label{sec:conclusions}
 
-We have presented here the possible application of the Monitoring Framework concepts to the EAGLE aggregating infrastructure.
+\noindent We have presented here the possible application of the Monitoring Framework concepts to the EAGLE aggregating infrastructure.
 The Monitoring Framework is ``work in progress'' in another European project related to research data infrastructures \citep{openaire}.

### 2016-02-15 panels/EAGLE_2016_submission_47.tex
fixes for lame
panel fixed, added tex version of paper 48 by authors and compiled

@@
 Panelists will focus on how connecting both types of edition thanks to digital technologies, i.e.
-integrating digital edition of epigraphic epigrams in the wider context of digital philology on one hand (e.g.: giving access to high quality images thanks to computer graphics tools) and in the one of public history on the other hand (e.g.: crowdsourcing with historical method).
+integrating digital edition of epigraphic epigrams in the wider context of digital philology on one hand (e.g.: giving access to high quality images thanks to computer graphics tools) and in the one of public history on the other hand (e.g.: crowdsourcing with historical method).\footnote{The texts of authors are open to discussion on the blog Épigraphie En Réseau: \url{ http://eer.hypotheses.org}}
 

### 2016-02-15 source/48/EAGLE16-lame-et-al-teachingAUTORI.tex
fixes for lame
panel fixed, added tex version of paper 48 by authors and compiled

--- 2016-02-10 source/48/EAGLE16-lame-et-al-teaching.tex
@@
 (i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
-correct fee\footnote{\citet{radman_livaja_segestica_2008}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014}, for analogous tags see:\citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_2011}; \citet[121-137]{schwinden_romerzeitliche_1985}; \citet[215-222]{schwinden_zwei_1993};\citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.
+correct fee\footnote{\citet{radman_livaja_segestica_2007}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014}, for analogous tags see:\citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_1984}; \citet[331-334]{frei-stolba_les_2011}; \citet[121-137]{schwinden_romerzeitliche_1985}; \citet[215-222]{schwinden_zwei_1993};\citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.
 
@@
 
-From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in \citet{lame_primary_2014}, on a dispositive analysis
+From a digital epigraphy standpoint, this consists of an online DAP \ modeled, in \citet{lame_primary_2014,Lame2015b}, on a dispositive analysis
 and the three fundamental systems of an inscription: writing system (wSystem), textual system (tSystem) and contextual
@@
 recover the linking information between wSystem and tSystem (e.g. browsing by glyph's shape and interpretation,
-identifying glyph's position), and easily shared with anyone with a browser (e.g.: fig. \ref{fig:markout}). The linking tool MarkOut
+identifying glyph's position), and easily shared with anyone with a browser (e.g.: Fig. \ref{fig:markout}). The linking tool MarkOut
 is one of the editing components of the modular DAP. They fit into communication workflows\footnote{This raises, of
@@
 \includegraphics[scale=0.10]{EAGLE16lameetalteaching-img004.png}
-\caption{MarkOut screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
+\caption{MarkOut Assisted screenshots (detailed part of tag no. 12563) showing the association, expressed in red, of the Unicode
 Latin characters with the position and the shape of the writing: U+0045 with II (identified as E2 by students) and
@@
 
-\newpage
-\section{Practicing Epigraphy}
-
-\noindent Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
-transcription and drawings) and to classify the written signs according to the following chart. Such preliminary
-classification gives a first idea of the main general shapes one can find on a lead tag. It is not a paleographical
-study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering (see: Fig.~\ref{fig:table1}).
-
-\begin{figure}[!hbp]
+\begin{figure}[!htp]
 \centering
@@
 
-\subsection{Drawing with help (No. 12563)}
+\section{Practicing Epigraphy}
 
-\vspace{0,5cm}
+\noindent Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
+transcription and drawings) and to classify the written signs according to the following chart (see: Fig.~\ref{fig:table1}). Such preliminary
+classification gives a first idea of the main general shapes one can find on a lead tag. It is not a paleographical
+study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering.
+
+\subsection{Drawing with Hints (No. 12563) with MarkOut Assisted}
+
+\vspace{0,3cm}
 \begin{minipage}[t]{0.18\textwidth}
@@
 
-\vspace{0,5cm}
+\vspace{0,3cm}
 
@@
 
-\subsection{Drawing \& Deciphering without help (obv. No. 13053)}
+\subsection{Drawing \& Deciphering without Help (obv. No. 13053)}
 
@@
 uncertain. Besides the last, most recent inscription, one still discerns traces of older inscriptions. The reverse is
-ignored. Students do not have neither transcriptions nor drawings.
+ignored. Students do not have neither transcriptions nor drawings (see MarkOut Expert Fig. ~\ref{fig:4.1}).
+
+\begin{figure}[!hbp]
+\centering
+\includegraphics[scale=0.25]{10_expert.png}
+\caption{Markout Expert allows free interpretation, specification of the type of glyph's form without providing extra information.}
+\label{fig:4.1}
+\end{figure}
 
@@
 
-\newpage
 The older, erased layers, on the obverse of tag No. 13053, drawn with dashed lies on the drawings of the academic edition, are more difficult to read. No
@@
 \includegraphics[scale=0.25]{EAGLE16lameetalteaching-img012.png}
-\caption{No. 13053 Obverse: students' drawing of some palimpsest inscriptions.}
+\caption{No. 13053 Obverse (palimpsest): students' drawing of some palimpsest inscriptions.}
 \label{fig:4}
@@
 present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.
-\nocite{Lame2015b}
+
 \nocite{cherry_storytelling_2014}

### 2016-02-15 panels/EAGLEpanel.tex
eagle panel description add

--- 2016-01-25 panels/EAGLE_2016_submission_6.tex
@@
 
-\title{Assessing the Role of Digital Libraries of Squeezes in Epigraphic Studies. Digitization, visualization, and metadata}
+\title{EAGLE Featured Panel}
 
-\author[Flo]{Eleni Bozia\corref{first}}
+\author[ISTI]{Vittore Casarosa\corref{first}}
 
-\address[Flo]{University of Florida}
-\cortext[first]{Corresponding author. Email: bozia@ufl.edu}
+\address[ISTI]{Istituto di Scienza e Tecnologie
+dell’Informazione ``A. Faedo''}
+\cortext[first]{Corresponding author. Email: casarosa@isti.cnr.it}
 
@@
 
-\keywords{Squeezes, Online Libraries, 3D Visualizations, Contextuality}
+\section{Description}
+The EAGLE project is really a multidisciplinary project,
+which can be looked at from three different perspectives,
+each one with many different facets.
+\begin{itemize}
+\item The epigraphy dimension, (inscriptions, transcriptions,
+critical editions, translations, historical context,
+geographical context, materials)
+\item  The technology dimension (data model, data aggregation,
+search engine, 3D representations, disambiguation,
+image recognition)
+\item  The application dimension (mobile application,
+storytelling, virtual reality, epigraphy in schools)
+\end{itemize}
 
-\section{Description}
-This panel will discuss issues and research questions regarding the use and efficiency of digital libraries of squeezes of inscriptions—structure of online libraries, nature of the metadata, 3D digitization, and visualization methods.
-The panel that consists of epigraphists, digital epigraphists, and computer engineers will engage in a dialogue, addressing the above issues from different perspectives.
-
-Digital libraries of squeezes have become an integral part in epigraphic studies and research.
-Whether it is the issue of collecting data for more efficient use, increasing their accessibility, or simply digitally preserving them, digital libraries are a new scholarly medium.
-Such projects con- front the challenge of having to determine the types of data that should be included—traditional information, digital metadata, formats and tools for articulating all the specificities of the metadata—as well as decide whether
- the type of the project should predetermine the ontology of the data.
-
-The aforementioned primary issue begs also the question of the need for interoperability of e-libraries in an attempt to combine, contrast, and comparatively appraise and study the artifacts themselves and their metadata.
-Thus far several projects feature digital libraries of squeezes; howe- ver, how should one proceed about perusing them?
-Whether this would be facilitated by means of homogeneity of the libraries, or whether the variegated nature of the material could potentially increase the number of users and the amount of information that is available.
-
-However, one should consider the limitations of e-libraries that focus solely on squeezes.
-When discussing the nature of the digital libraries’ content and the implication that the metadata may have in the interpretation of a squeeze, we need to consider its contextuality—whether the absence of the inscription bearer from a squeeze’s record may prove reductive with regards to the holistic approach to the text.
-However, notwithstanding that high-resolution digitization of squeezes opens the possibilities for enhanced study on a level that until now could not be achieved, techniques that focus solely on the digitization of the squeezes usually fail to deal with the inscription bearer.
-This panel will discuss how the digital records of squeezes could be augmented to encapsulate their contextual information.
+In this panel, we will touch on some of those facets,
+focussing more on the ones that are not going to be
+presented during the conference. Our aim is to stimulate
+some discussion and exchange of ideas, to try and
+understand what we have missed, what could have been
+done better, and what you (the audience) see as the
+strong points of EAGLE.
 
@@
 \begin{description}
-\item[Michèle Brunet]  \emph{Université Lyon 2, UMR HiSoMA \& Ecole française 
-d'Athènes  Email: Michele.Brunet@univ-lyon2.fr} Prof. Brunet has been the Chair of Greek and Latin Epigraphy at the
-Université Louis Lumière Lyon 2 since 2006. She has been a
-l'École Normale Supérieure de Paris (1979-1984) and l'École française d'Athènes (1984-1988), She received her HDR on the Study of Antiquity at the Sorbonne.
-She taught Archaeology and Greek Art a Bordeaux III and Paris I Panthéon-Sorbonne.
-She also served as the Director of Ancient Studies at l’École française d'Athènes of member.
-\item[Adeline Levivier] \emph{Université Lyon 2, UMR HiSoMA & Ecole française d’Athènes Email: adeline.levivier@gmail.com} Adeline is a doctoral candidate in l’École française d'Athènes - Université Lyon 2.
-She works on the digitization of squeezes, and her thesis is entitled ‘Recherches sur l'écriture grecque à partir des collections d'estampages d'inscriptions.’ She collaborated on the project ANR E-pigramme (Épigraphie et Muséographie - Édition numérique et valorisation de la Collection des inscriptions grecques du Musée du Louvre).
-She is currently the program manager of the project E-  stampages (Numérisation et diffusion web en 3D des collections de  l'UMR 5189 HiSoMA et de l'Ecole française d'Athènes).
-
-\item[Manuel Ramírez Sánchez] \emph{Universidad de Las Palmas de Gran Canaria Email: manuel.ramirez@ulpgc.es} Prof.
-Sánchez is a Professor of Historiographic Sciences and Techniques at the Department of Historical Sciences.
-He works on the 3D digitization and advanced visualizations of inscriptions and has published extensively in the area of inscriptions from Ancient Hispania.
-
-\item[Jose-Pablo Suárez-Rivero] \emph{Universidad de Las Palmas de Gran Canaria Email: Josepablo.suarez@ulpgc.es} Prof.
-Suárez-Rivero is the Director of Política Informática at the Cartography and Graphic Engineering Department.
-He has a Ph.D.
-in Applied Mathematics and works on mesh generation, algorithms and data structures, and computational geometry.
-
-\item[Angelos Barmpoutis] \emph{University of Florida Email: angelos@digitalworlds.ufl.edu} Prof.
-Barmpoutis is an Associate Professor in the On-line Institute and the Digital Worlds Institute at the University if Florida.
-He is also the coordinator of research and technology at the Institute and affiliate faculty at the Computer Science and Engineering Department.
-His research focuses on interdisciplinary applications of computer science to the service of broad areas of learning and training.
+\item[Anita Rocco] EDB 2.0
+\item[Valentina Vassallo] 3D epigraphy
+\item[Andrea Mannocci] data quality
+\item[Francesco Mambrini] storytelling
+\item[Giuseppe Amato] mobile application
+\item[Sorin Hermon] virtual exhibition
+\item[Luca Giberti] promotional video
 \end{description}
 
-\section{Notes}
-The intent of this panel of a multifarious cohort of scholars is to pose questions that have derived from the advancements of technology and digital epigraphy research.
-The panel will consider possible answers and ways to utilize traditional knowledge with the advantageous flexibility of digital tools that will ultimately not only facilitate, but also enhance epigraphic studies.
 \end{document}

### 2016-02-15 intro.tex
added intro

\chapter{EAGLE: PAST, PRESENT AND FUTURE}

\section{Past and Present [Silvia Orlandi]}
Like any other European project, EAGLE had to be reviewed, at the end of each year, by the Project Officer of the European Commission and two external reviewers. Both in 2014 and in 2015, after 12 and 24 months respectively, our project was evaluated as ``excellent'', even if we were invited to consider some critical points. From the very formal point of view of the European Commission, this means essentially that a given budget was used and reported in the correct ways and time, that documents and items were delivered by the deadline, and that tasks and milestones declared in the Description of Work were achieved.
Actually we are approaching the final goal of the project, that is to make accessible to the public --- not only to scholars --- a huge amount of texts and images with related metadata pertaining to ancient Greek and Latin inscriptions. The technical aspects and the positive results of the project will be illustrated by Claudio Prandoni, EAGLE Technical Coordinator, in the next presentation, and then more precisely described during the panel that is scheduled in the second part of this afternoon. As Scientific Coordinator, I would like instead to make some more theoretical observations and share some thoughts on the significance of this ``excellent'' evaluation and the reasons that lay behind what we can call a successful case.
In fact, EAGLE is not only a project that is careful in spending money and sending deliverables and reports, ready to pass any review: EAGLE was born to be much more than this.
The possibility of free online access to all the Greek and Latin inscriptions of the ancient world has been a real need of the academic community for a long time. Our aim is to reach this goal not with the creation of a new, single database, but with the aggregation of the digital archives of different institutions around a common project and a common idea of what an inscription is and how to read it; this idea was born after years of preliminary discussions and agreements that preceded and fostered the European project. Knowing clearly the need that we wanted to meet, and the way in which we wanted to do it, has surely helped us to focus our energies on the project’s main goals: the harmonization of very different materials and the creation of a new portal that could make them accessible through a search form specifically designed for inscriptions. We can say that --- as you will see --- both goals have been achieved, even if corrections and improvements are still possible and necessary. However, strange as this statement may seem, I don’t think that this is the real indicator of the success of the project. In my opinion, what actually shows that EAGLE is meeting a real need, is that our modus operandi is becoming an international standard: a larger and larger number of projects are using the controlled vocabularies that are one of the most interesting and immediately re-usable products of the harmonization job that EAGLE has done, are adopting an EAGLE compatible metadata model, and are sharing their content through the same system.
This scenario was already clear during the first international EAGLE conference, which was held in Paris in autumn 2014, and it is now confirmed by the papers and posters selected to be presented in the coming days, and by the huge networking activity of the project, which has widely enlarged the EAGLE consortium. Among the new partners --- too many to be listed here --- I would like to mention at least two examples that are --- for different reasons --- particularly important: the Inscriptions of Greek Cyrenaica and the EpiNet project, which is currently digitising the \emph{tituli picti} on the Roman amphoras from the Monte Testaccio in Rome. In the first case, one can clearly see the importance of a project dealing with inscriptions from a region corresponding to present-day Lybia: thanks to this digital archive, they will not only be better known, but also virtually preserved and protected against war damages and illegal commerce. In the second case, EpNet is an important step towards a digital archive for inscribed \emph{instrumentum}, a type of inscription that is fundamental to our knowledge of the ancient economy, but that is still lacking a policy of aggregation and harmonization of the existing digital resources.
This role as project of reference, of course, gives EAGLE a great responsibility, not only for the present, but also for the future, since now we must not only continue to meet the need for which the project was born, but also to show the direction in which we want to continue.
In this sense, and not by chance, I think, Euroepana has also shown an evolution in its policy: as has been recently stated during the final conference of the Athena Plus project, in the future Europeana will pay much more attention to the quality over the quantity of its digital items. This decision may have been made possible by the influence of projects like EAGLE, that are particularly careful about the curation of content and its usability in research activity.
The same observation can be made with respect to another aspect of the project. One of the main goals of the EAGLE project is to make inscriptions accessible not only to scholars, but also to a broader public, made up of students, teachers, tourists, and curious and interested people. To reach this audience it’s necessary to overcome the barriers represented by ancient languages and sometime ancient alphabets, epigraphic formulas and abbreviations, but also by the characteristics of traditional academic language and means of communication, using the huge potential of images, social media, and storytelling techniques. The results of this effort are particularly interesting: the realization of a mobile application using an image-based recognition system, the creation of the EAGLE MediaWiki platform to collect and organize thousands of translations in modern languages of epigraphic texts of varying complexity, a new storytelling application to illustrate the narrative content of many inscriptions, not to mention a virtual exhibition and a promotional video, which will be shown in the next few hours.
All this was included in the proposal submitted to the European Commission, but --- once again --- the fulfillment of promises is not the only indicator of the success of our project.  In this case too, I think that much more significant is what went beyond the promises: the unexpected, but no less interesting, developments. 
The possibility to have in a single online archive most of the existing translations of Greek and Latin inscriptions has raised a whole series of theoretical reflections about the often underestimated problems and difficulties that a translator has to face. On this subject, scientific contributions and practical solutions have been proposed, but new questions still must be answered. And this shows that even a task born as a dissemination activity can successfully interact, if seriously undertaken, with research activity.
In the same way, the enormous work of enriching the content of the project with images has led to an effort to to clarify the different laws that in different countries of the European Union govern the use of photographs of cultural heritage for both educational and commercial purposes. In this framework, the EAGLE Consortium has signed the assessment presented by Europeana Fundation to the European Parlament to ask for a revision and an update of the regulations concerning the re-use of digital images on the web, especially within projects related to the digitalization of cultural heritage. We also hope that in the future this assessment will be accepted by all the institutions that don’t yet recognize the civic value of scientific projects like ours.
In this field, too, the effort originally intended to enlarge the accessibility of the epigraphic material through the EAGLE portal is having --- and will have in the future --- interesting repercussions for scholars.
The importance of non textual elements for the correct and complete understanding of epigraphic messages has been recognized for some time. This implies the need not only to read, but also to look at the inscription. Not by chance, the theme chosen for the last International Conference of Greek and Latin Epigraphy, held in Berlin in 2012, was Öffentlichkeit --- Monument --- Text, and the same approach can be found in some of the papers that will be presented in the next days, dealing with the relationship between form and content in epigraphic studies.
You can imagine how many possibilities in the field of palaeography and writing technique can be opened by the ability to search for ``similar images'' through the EAGLE portal. Once again, a project can be recognized as successful if it not only meets a present need, but also shows new directions for future research, that will benefit from a constant and closer relationship between texts and images.
Finally, let me say something that is not strictly epigraphic, but is no less important. Among all the things that I have learned during these three years of work as coordinator, there is the persuasion that all these results would have never been possible without the help of all the people who, in different ways and with different roles, are involved in the project. People first of all curious and keen to ask questions, ready to listen and observe, who don’t use problems as a pretext not to do things, but rather try to solve them. People, above all, able to connect ideas, places, projects, and other people in the awareness that every success is not a point of arrival, but part of a continuing journey.
We can learn something from this experience, not only for our present satisfaction but also as a suggestion for the future: even in the field of digital epigraphy we have to move, I think, towards a wider connection and interoperability of projects, allowing us not only to progressively fill the still existing gaps, but also to better use our human and financial resources.
In the XIXth century, the \emph{Corpus Inscriptionum Latinarum} would have never been completed without the huge net of collaborators, correspondents and scholars from the whole of Europe with whom Theodor Mommsen intensively exchanged letters and documents. In the same way, I think, the new frontier of epigraphy is the broadening of studies and research made more and more open, collaborative and constantly updated thanks to a clever use of technology and digital resources.

\section{The Future [Silvio Panciera]}
There is no doubt that mine is the most difficult of the tasks assigned for this introduction. When we speak about past and present, we already know what has happened or what is happening. And this can also be a very pleasant task if, as in our case, there are many achievements and good results that can be praised.
On the contrary, it’s very different to speak about the future, that is something that has not happened yet, and that we don’t even know with certainty will ever happen. 
When I organized the XIth International Conference of Greek and Latin Epigraphy, which was held in Rome about 20 years ago, I asked Giancarlo Susini to open the conference, with the lecture ``Ten conferences plus one: the way of epigraphy''; on the opposite side, I asked Géza ALföldy to close the conference with a paper on ``The future of epigraphy'', and he, with the same embaressment that I have now, observed: ``As historians, we already have problems understanding the past. What could we say about the future?''
As a matter of fact, the future doesn’t lean on solid, measurable facts: it’s the reign of the unknown, in which one projects rational and irrational desires, mixed with fears or even anguish. But it’s also a temporal space in which many of our previsions are going to be swept away by facts that nobody had foreseen or even imagined.
It would have seemed more logical for the future of the project to be presented by a young person, who will have the time to experience it, and not by an old man, who will never see it. By the way, I’m rather upset when I see the media invaded every day by some old men --- I hope not to be one of them --- who, after having had a long time to express their opinion, and having failed, think that they still have the right to give rules for the future, while --- maybe - it’s the moment to open the floor to others.
But let’s speak of lighter things!
I don’t think that you want me to become a ``futurologist''  and to make predictions of what is going to happen in the world, or, more modestly, in EAGLE, in a more or less distant future. But we can’t do without the future.
Every action presupposes a future, and this feeling leads our steps and gives them a meaning. But we can also say that the future is nothing but the present of yesterday. To speak about the future of EAGLE, therefore, there is no need to imagine complex scenarios for the coming years. We can just ask ourselves: What am I going to do tomorrow? In English, there is a convenient distinction between the simple future, that is used for actions that happen spontaneously, and the intentional future, that is used for actions that are consequences of a plan, an intention. That’s what we should talk about now, but we can’t do that without considering:
a) What was our original goal and how much of it has been achieved
b) What has worked well and is to be kept, and what, according to this experience, should be improved
c) What has been set aside or not originally included, but should be planned for the future
Since the beginning, I’ve thought that my role in the project was not to define every detail, but to establish some key points, and check that they would not be forgotten or changed. That’s why today I would like not to illustrate every aspect of the project, but just to say a couple of words about the points already mentioned.
``Old fashioned'' EAGLE was born as a federation of databases with the goal of ``recording all Greek and Latin inscriptions older than the VIIth century AD, according to the best existing edition, possibly checked and improved, along with some fundamental data and images''. We can say that this goal has been adopted by the ``new version'' EAGLE, too. But, due to a difficult coordination between Greek and Latin epigraphers (an old problem…) most attention has actually been paid to Latin inscriptions. We have known since the beginning that the complete recording of all known inscriptions was practically impossible. Nevertheless, we can probably state that in a short time the number of inscriptions searchable through the EAGLE portal will reach about 550.000, thanks to the enlargement of the consortium and to the enormous work of harmonization and disambiguation of content, and above all thanks to the inclusion of the epigraphic texts put at our disposal by Manfred Clauss and his database (I would like to thank Pietro Maria Liuzzo for this and other information).
It’s not all that we need (I’ll come back on this point later) and, above all, the metadata set of the texts and their degree of elaboration is not homogeneous in all the records. But nobody can deny that the original plan has been mostly fulfilled, and looking at what has been done we can be rather confident in the work that awaits us in the future.
Our tasks for tomorrow are the subject of my second point, concerning what to keep, and what to change according to our past experience. In my opinion, and --- as it seems --- according to the European Commission too, the general structure of the project, what we can call its philosophy, has been successfully tested --- as the facts testify --- and should, therefore, be maintained. The philosophy is based on two fundamental principles.
The first is common to every project that aims to be a scientific research project, and it’s the awareness that we do not have definitive solutions, but just hypotheses, which must always be checked because every attempt involves the possibility of errors that can and must be corrected. The large number of changes and improvements made during the project are not a proof of weakness, but a sign of its strength.
The second point is the clear need for as wide a collaboration as possible, to ensure not only a large quantity of data, but above all a high quality of content, checked by experts in different geographic regions and thematic fields. This aspect has been particularly curated in these last years, so that the number of institutions and single content providers has been greatly increased. In this way, and thanks to great technical work including the fundamental creation of controlled vocabularies, the EAGLE portal now gives access to many different databases, originally independent and with different characteristics and purposes. In my opinion, this is the path to follow in the future as well, in order to face and solve bigger and bigger problems.
EAGLE looks different from other similar projects because, since the beginning, it has paid much attention not only to the quality of information, but also to the combination of textual and non-textual data, according to the current definition of an inscription as an inscribed monument. Moreover, in recent years, w have seen a huge increase in the visual documentation available online, and now EAGLE includes about 250.000 digital photos of inscriptions. This is another aspect that should be maintained, and --- possibly- even strengthened. In fact, thanks to projects like ours, the problems related to the legal treatment of images of cultural heritage seem to be at the moment under discussion, both nationally and internationally (papers on this subject will be presented in the coming days). Silvia Orlandi is also right to underline how important the inclusion of images of inscriptions and the technical possibility to search them will be for palaeographic studies. I have recently discussed with Silvia Evangelisti how to improve the ``\emph{scriptura}'' field of EDR with more detailed information about writing techniques, materials and tools. But the analysis of the graphic forms of inscriptions has not yet been adequately confronted, as it still lacks the contribution of professional palaeographers. Maybe this conference will give us the chance to begin this kind of discussion.
Speaking about old goals to be taken up again, or new goals to be achieved, I have to come back to the problem of Greek inscriptions and the inscribed \emph{instrumentum}. They are both essential components of our research field, with a number of independent digital repositories, but since the beginning we have had problems with their inclusion in the project. At the moment, EAGLE includes more than 7000 Greek inscriptions, while a specific commission of the International Association of Greek and Latin Epigraphy is currently working on the digital \emph{instrumentum}. Now it seems that, after a long period of inactivity, we are having in recent years an awakening on both sides, and particularly on the side of Greek epigraphy, thanks above all, to the IG. Now it’s time to aggregate this fundamental material too, with the help of new technologies. I would just suggest that we first reach a preliminary agreement among the participants using past experience rather than starting from scratch both in terms of base requirements, and of open and flexible structure.
All this is about the future: not a vague and undefined future, but a very positive future, modeled according to our past and present plans, constantly checked and renewed. But at least part of the future doesn’t depend on us.
For example: Will there still be somebody who will trust and fund projects like ours?
And will there still be, in Italy and elsewhere, enough scholars, old and young, adequately educated and motivated, who will take care of it?
The tendencies that we can see in Italian and European policy in the field of culture and the university, or at least some of them, could lead us to pessimism, but I don’t think that we should give up. First of all because there are also signs of hope. The European Union, for example, with other important national institutions whose moral and financial support should never be forgotten --- has trusted and supported the proposal that we submitted three years ago, and, during annual reviews, has appreciated the way in which the project has been carried out under the guidance of Silvia Orlandi and her many collaborators. Why should we exclude the possibility that something similar will happen again in the coming years? I continue to believe that EAGLE is not an ordinary project and that its cultural importance, both for the scientific community and for civic life, will be adequately recognized.
Moreover, my confidence is increased by the enthusiasm with which so many young people have taken part in the project, giving and receiving so much, not only on the professional and cultural side, but also in terms of education and ethics, learning to work not only for themselves, but also for others.
We live in a time of very quick changes --- and this is even more true for a university, where the population of students is almost completely renewed every five years. Therefore, to foresee the worst is not necessarily more realistic than having some hopes.
Anyway, I think that not only a scholar, but any person who believes in something, should not give up only because of a mere calculation of probability: luckily we are not working only for the market, so that we can leave to others the task of dealing with risks and probabilities.

\sign{Silvia Orlandi, Silvio Panciera}

### 2016-02-19 intro.tex
intro updated

@@
 
-\section{Past and Present [Silvia Orlandi]}
-Like any other European project, EAGLE had to be reviewed, at the end of each year, by the Project Officer of the European Commission and two external reviewers.
-Both in 2014 and in 2015, after 12 and 24 months respectively, our project was evaluated as ``excellent'', even if we were invited to consider some critical points.
+\section*{Past and Present}
+Like any other European project, EAGLE had to be reviewed, at the end of each year, by the Project Officer of the European Commission and two external reviewers.\footnote{This texts reproduces without substantial changes the speech held on January 27th, at the opening of the EAGLE Final Conference.} Both in 2014 and in 2015, after 12 and 24 months respectively, our project was evaluated as ``excellent'', even if we were invited to consider some critical points.
 From the very formal point of view of the European Commission, this means essentially that a given budget was used and reported in the correct ways and time, that documents and items were delivered by the deadline, and that tasks and milestones declared in the Description of Work were achieved.
-Actually we are approaching the final goal of the project, that is to make accessible to the public --- not only to scholars --- a huge amount of texts and images with related metadata pertaining to ancient Greek and Latin inscriptions.
-The technical aspects and the positive results of the project will be illustrated by Claudio Prandoni, EAGLE Technical Coordinator, in the next presentation, and then more precisely described during the panel that is scheduled in the second part of this afternoon.
+Actually we are approaching the final goal of the project, that is to make accessible to the public --- not only to scholars --- a huge amount of texts and images, with related metadata, pertaining to ancient Greek and Latin inscriptions.
+The technical aspects and the positive results are illustrated by Claudio Prandoni, EAGLE Technical Coordinator, in a dedicated presentation and panel (see the EAGLE panel p.
+\pageref{peaglepanel}, and the contributions by Mannocci p.
+\pageref{pCNR}, Prandoni p.
+\pageref{pXX}, Rocco p.
+\pageref{p28} and Vassallo-Damnjanovic p.
+\pageref{p44}).
 As Scientific Coordinator, I would like instead to make some more theoretical observations and share some thoughts on the significance of this ``excellent'' evaluation and the reasons that lay behind what we can call a successful case.
+
 In fact, EAGLE is not only a project that is careful in spending money and sending deliverables and reports, ready to pass any review: EAGLE was born to be much more than this.
+
 The possibility of free online access to all the Greek and Latin inscriptions of the ancient world has been a real need of the academic community for a long time.
-Our aim is to reach this goal not with the creation of a new, single database, but with the aggregation of the digital archives of different institutions around a common project and a common idea of what an inscription is and how to read it; this idea was born after years of preliminary discussions and agreements that preceded and fostered the European project.
+Our aim is to reach this goal not with the creation of a new, single database, but with the aggregation of the digital archives of different institutions around a common project and a common idea of what an inscription is, and how to read it; this idea was born after years of preliminary discussions and agreements that preceded and fostered the European project.
 Knowing clearly the need that we wanted to meet, and the way in which we wanted to do it, has surely helped us to focus our energies on the project’s main goals: the harmonization of very different materials and the creation of a new portal that could make them accessible through a search form specifically designed for inscriptions.
-We can say that --- as you will see --- both goals have been achieved, even if corrections and improvements are still possible and necessary.
+We can say that both goals have been achieved, even if corrections and improvements are still possible and necessary.
 However, strange as this statement may seem, I don’t think that this is the real indicator of the success of the project.
 In my opinion, what actually shows that EAGLE is meeting a real need, is that our modus operandi is becoming an international standard: a larger and larger number of projects are using the controlled vocabularies that are one of the most interesting and immediately re-usable products of the harmonization job that EAGLE has done, are adopting an EAGLE compatible metadata model, and are sharing their content through the same system.
-This scenario was already clear during the first international EAGLE conference, which was held in Paris in autumn 2014, and it is now confirmed by the papers and posters selected to be presented in the coming days, and by the huge networking activity of the project, which has widely enlarged the EAGLE consortium.
-Among the new partners --- too many to be listed here --- I would like to mention at least two examples that are --- for different reasons --- particularly important: the Inscriptions of Greek Cyrenaica and the EpiNet project, which is currently digitising the \emph{tituli picti} on the Roman amphoras from the Monte Testaccio in Rome.
+
+This scenario was already clear during the First EAGLE International Conference, which was held in Paris in autumn 2014, and it is now confirmed by the papers and posters selected to be presented in this volume, and by the huge networking activity of the project, which has widely enlarged the EAGLE consortium.
+Among the new partners --- too many to be listed here --- I would like to mention at least two examples that are --- for different reasons --- particularly important: the Inscriptions of Greek Cyrenaica (\url{https://igcyr.unibo.it/}) and the EPNet project (\url{www.roman-ep.net}), which is currently digitising the \emph{tituli picti} on the Roman amphoras from the Monte Testaccio in Rome.
 In the first case, one can clearly see the importance of a project dealing with inscriptions from a region corresponding to present-day Lybia: thanks to this digital archive, they will not only be better known, but also virtually preserved and protected against war damages and illegal commerce.
-In the second case, EpNet is an important step towards a digital archive for inscribed \emph{instrumentum}, a type of inscription that is fundamental to our knowledge of the ancient economy, but that is still lacking a policy of aggregation and harmonization of the existing digital resources.
+In the second case, EPNet is an important step towards a digital archive for inscribed \emph{instrumentum}, a type of inscription that is fundamental to our knowledge of the ancient economy, but that is still lacking a policy of aggregation and harmonization of the existing digital resources.
+
 This role as project of reference, of course, gives EAGLE a great responsibility, not only for the present, but also for the future, since now we must not only continue to meet the need for which the project was born, but also to show the direction in which we want to continue.
-In this sense, and not by chance, I think, Euroepana has also shown an evolution in its policy: as has been recently stated during the final conference of the Athena Plus project, in the future Europeana will pay much more attention to the quality over the quantity of its digital items.
+
+In this sense, and not by chance, I think, Euroepana has also shown an evolution in its policy: as has been recently stated by Joris Pekel during the final conference of the Athena Plus project (\url{http://www.athenaplus.eu/index.php?en/202/athenaplus-final-conference}), in the future Europeana will pay much more attention to the quality over the quantity of its digital items.
 This decision may have been made possible by the influence of projects like EAGLE, that are particularly careful about the curation of content and its usability in research activity.
+
 The same observation can be made with respect to another aspect of the project.
@@
 To reach this audience it’s necessary to overcome the barriers represented by ancient languages and sometime ancient alphabets, epigraphic formulas and abbreviations, but also by the characteristics of traditional academic language and means of communication, using the huge potential of images, social media, and storytelling techniques.
-The results of this effort are particularly interesting: the realization of a mobile application using an image-based recognition system, the creation of the EAGLE MediaWiki platform to collect and organize thousands of translations in modern languages of epigraphic texts of varying complexity, a new storytelling application to illustrate the narrative content of many inscriptions, not to mention a virtual exhibition and a promotional video, which will be shown in the next few hours.
+The results of this effort are particularly interesting: the realization of a mobile application using an image-based recognition system, the creation of the EAGLE MediaWiki platform to collect and organize thousands of translations in modern languages of epigraphic texts of varying complexity, a new storytelling application to illustrate the narrative content of many inscriptions, not to mention a virtual exhibition and a promotional video (see the EAGLE featured panel p.
+\pageref{peaglepanel}).
+
 All this was included in the proposal submitted to the European Commission, but --- once again --- the fulfillment of promises is not the only indicator of the success of our project.
 In this case too, I think that much more significant is what went beyond the promises: the unexpected, but no less interesting, developments.
+
 The possibility to have in a single online archive most of the existing translations of Greek and Latin inscriptions has raised a whole series of theoretical reflections about the often underestimated problems and difficulties that a translator has to face.
-On this subject, scientific contributions and practical solutions have been proposed, but new questions still must be answered.
+On this subject, scientific contributions and practical solutions have been proposed (see, for example, the Guidelines for Translators in EAGLE MediaWiki: \url{http://www.eagle-network.eu/wiki/index.php/Guidelines_for_Translators}), but new questions still must be answered.
 And this shows that even a task born as a dissemination activity can successfully interact, if seriously undertaken, with research activity.
+
 In the same way, the enormous work of enriching the content of the project with images has led to an effort to to clarify the different laws that in different countries of the European Union govern the use of photographs of cultural heritage for both educational and commercial purposes.
-In this framework, the EAGLE Consortium has signed the assessment presented by Europeana Fundation to the European Parlament to ask for a revision and an update of the regulations concerning the re-use of digital images on the web, especially within projects related to the digitalization of cultural heritage.
+In this framework, the EAGLE Consortium has shared the position of Europeana on the review of the EU copyright rules (\url{http://pro.europeana.eu/blogpost/a-first-glance-into-the-future-of-eu-copyright-reform}) and the re-use of digital images on the web, especially within projects related to the digitalization of cultural heritage.
 We also hope that in the future this assessment will be accepted by all the institutions that don’t yet recognize the civic value of scientific projects like ours.
+
 In this field, too, the effort originally intended to enlarge the accessibility of the epigraphic material through the EAGLE portal is having --- and will have in the future --- interesting repercussions for scholars.
+
 The importance of non textual elements for the correct and complete understanding of epigraphic messages has been recognized for some time.
 This implies the need not only to read, but also to look at the inscription.
-Not by chance, the theme chosen for the last International Conference of Greek and Latin Epigraphy, held in Berlin in 2012, was Öffentlichkeit --- Monument --- Text, and the same approach can be found in some of the papers that will be presented in the next days, dealing with the relationship between form and content in epigraphic studies.
+Not by chance, the theme chosen for the last International Conference of Greek and Latin Epigraphy, held in Berlin in 2012, was Öffentlichkeit --- Monument --- Text, and the same approach can be found in some of the papers presented in this volume, dealing with the relationship between form and content in epigraphic studies (see papers by Felle p.
+\pageref{p40}, Benefiel p.
+\pageref{p43}, Graham p.
+\pageref{p10}).
+
 You can imagine how many possibilities in the field of palaeography and writing technique can be opened by the ability to search for ``similar images'' through the EAGLE portal.
 Once again, a project can be recognized as successful if it not only meets a present need, but also shows new directions for future research, that will benefit from a constant and closer relationship between texts and images.
+
 Finally, let me say something that is not strictly epigraphic, but is no less important.
@@
 People, above all, able to connect ideas, places, projects, and other people in the awareness that every success is not a point of arrival, but part of a continuing journey.
+
 We can learn something from this experience, not only for our present satisfaction but also as a suggestion for the future: even in the field of digital epigraphy we have to move, I think, towards a wider connection and interoperability of projects, allowing us not only to progressively fill the still existing gaps, but also to better use our human and financial resources.
-In the XIXth century, the \emph{Corpus Inscriptionum Latinarum} would have never been completed without the huge net of collaborators, correspondents and scholars from the whole of Europe with whom Theodor Mommsen intensively exchanged letters and documents.
+
+In the XIXth century, the \emph{Corpus Inscriptionum Latinarum} would have never been completed without the huge net of collaborators, correspondants and scholars from the whole of Europe with whom Theodor Mommsen intensively exchanged letters and documents.
 In the same way, I think, the new frontier of epigraphy is the broadening of studies and research made more and more open, collaborative and constantly updated thanks to a clever use of technology and digital resources.
 
-\section{The Future [Silvio Panciera]}
+\sign{Silvia Orlandi}
+
+\section*{Future}
 There is no doubt that mine is the most difficult of the tasks assigned for this introduction.
@@
 And this can also be a very pleasant task if, as in our case, there are many achievements and good results that can be praised.
+
 On the contrary, it’s very different to speak about the future, that is something that has not happened yet, and that we don’t even know with certainty will ever happen.
-When I organized the XIth International Conference of Greek and Latin Epigraphy, which was held in Rome about 20 years ago, I asked Giancarlo Susini to open the conference, with the lecture ``Ten conferences plus one: the way of epigraphy''; on the opposite side, I asked Géza ALföldy to close the conference with a paper on ``The future of epigraphy'', and he, with the same embaressment that I have now, observed: ``As historians, we already have problems understanding the past.
+
+When I organized the XIth International Conference of Greek and Latin Epigraphy, which was held in Rome about 20 years ago, I asked Giancarlo Susini to open the conference, with the lecture ``Ten conferences plus one: the way of epigraphy''; on the opposite side, I asked Géza Alföldy to close the conference with a paper on ``The future of epigraphy'', and he, with the same embaressment that I have now, observed: ``As historians, we already have problems understanding the past.
 What could we say about the future?''
+
 As a matter of fact, the future doesn’t lean on solid, measurable facts: it’s the reign of the unknown, in which one projects rational and irrational desires, mixed with fears or even anguish.
 But it’s also a temporal space in which many of our previsions are going to be swept away by facts that nobody had foreseen or even imagined.
+
 It would have seemed more logical for the future of the project to be presented by a young person, who will have the time to experience it, and not by an old man, who will never see it.
 By the way, I’m rather upset when I see the media invaded every day by some old men --- I hope not to be one of them --- who, after having had a long time to express their opinion, and having failed, think that they still have the right to give rules for the future, while --- maybe - it’s the moment to open the floor to others.
+
 But let’s speak of lighter things!
+
 I don’t think that you want me to become a ``futurologist'' and to make predictions of what is going to happen in the world, or, more modestly, in EAGLE, in a more or less distant future.
 But we can’t do without the future.
+
 Every action presupposes a future, and this feeling leads our steps and gives them a meaning.
@@
 That’s what we should talk about now, but we can’t do that without considering:
-a) What was our original goal and how much of it has been achieved
-b) What has worked well and is to be kept, and what, according to this experience, should be improved
-c) What has been set aside or not originally included, but should be planned for the future
+\begin{description}
+\item[a] What was our original goal and how much of it has been achieved
+\item[b] What has worked well and is to be kept, and what, according to this experience, should be improved
+\item[c] What has been set aside or not originally included, but should be planned for the future
+\end{description}
+
 Since the beginning, I’ve thought that my role in the project was not to define every detail, but to establish some key points, and check that they would not be forgotten or changed.
 That’s why today I would like not to illustrate every aspect of the project, but just to say a couple of words about the points already mentioned.
-``Old fashioned'' EAGLE was born as a federation of databases with the goal of ``recording all Greek and Latin inscriptions older than the VIIth century AD, according to the best existing edition, possibly checked and improved, along with some fundamental data and images''.
+
+``Old fashioned'' EAGLE was born as a federation of databases with the goal of ``recording all Greek and Latin inscriptions older than the VIIth century AD, according to the best existing edition, possibly checked and improved, along with some fundamental data and images'' (see the documents collected in \url{http://www.eagle-eagle.it/Italiano/documenti_it.htm}).
 We can say that this goal has been adopted by the ``new version'' EAGLE, too.
@@
 Nevertheless, we can probably state that in a short time the number of inscriptions searchable through the EAGLE portal will reach about 550.000, thanks to the enlargement of the consortium and to the enormous work of harmonization and disambiguation of content, and above all thanks to the inclusion of the epigraphic texts put at our disposal by Manfred Clauss and his database (I would like to thank Pietro Maria Liuzzo for this and other information).
+
 It’s not all that we need (I’ll come back on this point later) and, above all, the metadata set of the texts and their degree of elaboration is not homogeneous in all the records.
 But nobody can deny that the original plan has been mostly fulfilled, and looking at what has been done we can be rather confident in the work that awaits us in the future.
+
 Our tasks for tomorrow are the subject of my second point, concerning what to keep, and what to change according to our past experience.
 In my opinion, and --- as it seems --- according to the European Commission too, the general structure of the project, what we can call its philosophy, has been successfully tested --- as the facts testify --- and should, therefore, be maintained.
-The philosophy is based on two fundamental principles.
-The first is common to every project that aims to be a scientific research project, and it’s the awareness that we do not have definitive solutions, but just hypotheses, which must always be checked because every attempt involves the possibility of errors that can and must be corrected.
+
+The philosophy is based on two fundamental principles:
+\begin{enumerate}
+\item The first is common to every project that aims to be a scientific research project, and it’s the awareness that we do not have definitive solutions, but just hypotheses, which must always be checked because every attempt involves the possibility of errors that can and must be corrected.
 The large number of changes and improvements made during the project are not a proof of weakness, but a sign of its strength.
-The second point is the clear need for as wide a collaboration as possible, to ensure not only a large quantity of data, but above all a high quality of content, checked by experts in different geographic regions and thematic fields.
+\item The second point is the clear need for as wide a collaboration as possible, to ensure not only a large quantity of data, but above all a high quality of content, checked by experts in different geographic regions and thematic fields.
 This aspect has been particularly curated in these last years, so that the number of institutions and single content providers has been greatly increased.
@@
 In my opinion, this is the path to follow in the future as well, in order to face and solve bigger and bigger problems.
+\end{enumerate}
+
 EAGLE looks different from other similar projects because, since the beginning, it has paid much attention not only to the quality of information, but also to the combination of textual and non-textual data, according to the current definition of an inscription as an inscribed monument.
-Moreover, in recent years, w have seen a huge increase in the visual documentation available online, and now EAGLE includes about 250.000 digital photos of inscriptions.
-This is another aspect that should be maintained, and --- possibly- even strengthened.
-In fact, thanks to projects like ours, the problems related to the legal treatment of images of cultural heritage seem to be at the moment under discussion, both nationally and internationally (papers on this subject will be presented in the coming days).
-Silvia Orlandi is also right to underline how important the inclusion of images of inscriptions and the technical possibility to search them will be for palaeographic studies.
+Moreover, in recent years, we have seen a huge increase in the visual documentation available online, and now EAGLE includes about 250.000 digital photos of inscriptions.
+This is another aspect that should be maintained, and --- possibly - even strengthened.
+In fact, thanks to projects like ours, the problems related to the legal treatment of images of cultural heritage seem to be at the moment under discussion, both nationally and internationally (see paper by Modolo p.
+\pageref{p38}).
+Silvia Orlandi is also right to underline how important the inclusion of images of inscriptions and the technical possibility to search them will be for paleographic studies.
 I have recently discussed with Silvia Evangelisti how to improve the ``\emph{scriptura}'' field of EDR with more detailed information about writing techniques, materials and tools.
-But the analysis of the graphic forms of inscriptions has not yet been adequately confronted, as it still lacks the contribution of professional palaeographers.
+But the analysis of the graphic forms of inscriptions has not yet been adequately confronted, as it still lacks the contribution of professional paleographers.
 Maybe this conference will give us the chance to begin this kind of discussion.
+
 Speaking about old goals to be taken up again, or new goals to be achieved, I have to come back to the problem of Greek inscriptions and the inscribed \emph{instrumentum}.
@@
 At the moment, EAGLE includes more than 7000 Greek inscriptions, while a specific commission of the International Association of Greek and Latin Epigraphy is currently working on the digital \emph{instrumentum}.
-Now it seems that, after a long period of inactivity, we are having in recent years an awakening on both sides, and particularly on the side of Greek epigraphy, thanks above all, to the IG.
+Now it seems that, after a long period of inactivity, we are having in recent years an awakening on both sides, and particularly on the side of Greek epigraphy, thanks above all, to the \emph{Inscriptiones Graecae}.
 Now it’s time to aggregate this fundamental material too, with the help of new technologies.
 I would just suggest that we first reach a preliminary agreement among the participants using past experience rather than starting from scratch both in terms of base requirements, and of open and flexible structure.
+
 All this is about the future: not a vague and undefined future, but a very positive future, modeled according to our past and present plans, constantly checked and renewed.
 But at least part of the future doesn’t depend on us.
+
 For example: Will there still be somebody who will trust and fund projects like ours?
+
 And will there still be, in Italy and elsewhere, enough scholars, old and young, adequately educated and motivated, who will take care of it?
+
 The tendencies that we can see in Italian and European policy in the field of culture and the university, or at least some of them, could lead us to pessimism, but I don’t think that we should give up.
@@
 I continue to believe that EAGLE is not an ordinary project and that its cultural importance, both for the scientific community and for civic life, will be adequately recognized.
+
 Moreover, my confidence is increased by the enthusiasm with which so many young people have taken part in the project, giving and receiving so much, not only on the professional and cultural side, but also in terms of education and ethics, learning to work not only for themselves, but also for others.
+
 We live in a time of very quick changes --- and this is even more true for a university, where the population of students is almost completely renewed every five years.
 Therefore, to foresee the worst is not necessarily more realistic than having some hopes.
+
 Anyway, I think that not only a scholar, but any person who believes in something, should not give up only because of a mere calculation of probability: luckily we are not working only for the market, so that we can leave to others the task of dealing with risks and probabilities.
 
-\sign{Silvia Orlandi, Silvio Panciera}
+\sign{Silvio Panciera}

### 2016-02-19 source/48/EAGLE16-lame-et-al-teaching (1).tex
intro updated

--- 2016-02-15 source/48/EAGLE16-lame-et-al-teachingAUTORI.tex
@@
 \title{Teaching (Digital) Epigraphy}
-\author[ilc]{Marion Lamé}\corref{first}
+\author[ilc]{Marion Lamé\corref{first}}
+\author[isti]{Federico Ponchio}
+\author[mau]{Bruce Robertson}
 \author[zag]{Ivan Radman-Livaja}
-\author[mau]{Bruce Robertson}
-\author[isti]{Federico Ponchio}
 \address[ilc]{Centre Camille Jullian, MMSH, CNRS and Laboratorio di Cultura Digitale, University of Pisa}
 \address[isti]{Visual Computing Laboratory, ISTI-CNR}
+\address[mau]{Mount Allison University}
 \address[zag]{Archaeological Museum in Zagreb}
-\address[mau]{Mount Allison University}
-\cortext[first]{Corresponding author. Email: mlame@mmsh.univ-aix.fr}
+\cortext[first]{Corresponding author.
+Email: mlame@mmsh.univ-aix.fr.
+The poster pertaining to this article can be viewed in the \emph{poster} section of the academic blog \emph{Épigraphie En Réseau} -  \url{http://eer.hypotheses.org}.}
 
@@
 
-\noindent Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
+Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
 on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
@@
 
-\noindent The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
+The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
 found in Sisak (Siscia), one of the largest urban centers in south-western Pannonia. Most of them were found during the
@@
 
-\begin{figure}
-\centering
-
-\end{figure}
 Most of the tags, if not all, are linked to the wool trade and the textile industry.
@@
 
-\noindent In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
+In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
 far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
@@
 profit from the assistance of undergraduate and graduate students, which could not often be the case
-before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa, applies the
+before\footnote{The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa, applies the
 didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
@@
 help of technology, digital editions may be thoroughly checked and amended far more easily. Besides giving an
-opportunity to further study the primary source of information and amend the paper editions and publications, it also
-allows university teachers to train future scholars and specialists, wherever their geographic location may be. In this
+opportunity to further study the primary source of information and amend the paper editions and publications, it also allows university teachers to train future scholars and specialists, wherever their geographic location may be.
+In this
 particular case, the epigraphic material is in Croatia but the analysis was done online by scholars and students from
@@
 
-\noindent Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
+Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
 transcription and drawings) and to classify the written signs according to the following chart (see: Fig.~\ref{fig:table1}). Such preliminary
@@
 
-\noindent From the educational standpoint, this pilot project succeeded tremendously. The Canadian students who undertook this
+From the educational standpoint, this pilot project succeeded tremendously. The Canadian students who undertook this
 transcription in place of the usual research essay were, on average, far more motivated than their peers, and took
@@
 
-\noindent Gaining expertise with material remains requires time. When a museum trains students, this time is often limited due to
+Gaining expertise with material remains requires time. When a museum trains students, this time is often limited due to
 constrained facilities and personnel. In addition, the epigraphic material may be fragile and delicate: the less these
@@
 \section*{Aknowledgment}
-\noindent We wish to acknowledge the research assistance of T. M. Crandall,
+We wish to acknowledge the research assistance of T. M. Crandall,
 History student at Mount Allison University (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
@@
 
+%SOSTITUIRE, PRIMA DI MANDARE, CON LA RIGA \bibliography{../../EAGLE}
+
 \end{document}

### 2016-03-01 keynote/roueché/EAGLE_2016_Roueche.tex
added paper route
EAGLE bib updated
paper converted
corrected
pdf copied to pdfs folder

@@
 The publication of inscriptions in digital form has been evolving since the 1980s, and online publication since the 1990s.
-It is useful to see where EAGLE is positioned in this evolution – a huge step forward, but not the end of the road.
+It is useful to see where EAGLE is positioned in this evolution --- a huge step forward, but not the end of the road.
 What can we learn from our progress so far?
@@
 
-Charlotte Roueché studied Classics at Cambridge, and Byzantine Studies in Paris and Washington; as a pupil of Joyce Reynolds, she became involved in publishing Greek inscriptions, with a particular focus on material from Late Antiquity.
-In 1984 she was appointed to King’s College London, where an important centre of Digital Humanities started to develop in the 1990s; this encouraged her to explore the possibilities for publishing inscribed texts online.
-In 1999 Tom Elliott published the first proposal for EpiDoc; working with him and Gabriel Bodard, Charlotte was able to use EpiDoc to publish several online corpora: \url{http://insaph.kcl.ac.uk/ala2004, http://insaph.kcl.ac.uk/iaph2007}, and \url{http://irt.kcl.ac.uk/irt2009}.
+{\small \emph{Charlotte Roueché studied Classics at Cambridge, and Byzantine Studies in Paris and Washington; as a pupil of Joyce Reynolds, she became involved in publishing Greek inscriptions, with a particular focus on material from Late Antiquity.
+In 1984 she was appointed to King's College London, where an important centre of Digital Humanities started to develop in the 1990s; this encouraged her to explore the possibilities for publishing inscribed texts online.
+In 1999 Tom Elliott published the first proposal for EpiDoc; working with him and Gabriel Bodard, Charlotte was able to use EpiDoc to publish several online corpora: \url{http://insaph.kcl.ac.uk/ala2004}, \url{http://insaph.kcl.ac.uk/iaph2007}, and \url{http://irt.kcl.ac.uk/irt2009}.
 At present she is preparing a digital corpus of inscriptions of Roman Cyrenaica, and developing plans for a shared portal, Inscriptions of Libya.
-She has also recently published an eleventh century Byzantine text online, in a project exploring the uses of RDF to analyse intertextuality: \url{http://ancientwisdoms.ac.uk}.
+She has also recently published an eleventh century Byzantine text online, in a project exploring the uses of RDF to analyse intertextuality: \url{http://ancientwisdoms.ac.uk}.}}
+
+\vspace{1cm}
+I couldn't resist this title, taken from Shakespeare's \emph{Richard III} to use for a Hollywood film, full of daring wartime
+adventures. As academics we may underestimate the importance of courage in our undertakings. But the early epigraphers
+required a good deal of \textbf{daring}; while recording inscriptions was straightforward enough in Italy or north-western
+Europe, the eastern and southern regions of the Greco-Roman world remained difficult and dangerous to visit until the
+late nineteenth century or even later --- as is again becoming true. The early travellers also suffered from practical
+constraints; in the 18\textsuperscript{th} and 19\textsuperscript{th} centuries they were limited by the quantity of paper that they had brought with them
+just as in the 20\textsuperscript{th} century; Robert `Palmyra' Wood recorded inscriptions on pages of his copy of Homer. He was
+travelling for several months, and was saving his paper for sketches, plans and drawings of buildings.\footnote{ Robert
+Wood (1716/17–1771): doi:10.1093/ref:odnb/29891} Similarly John Deering recorded texts in a notebook with very small
+pages.\footnote{ John Peter Deering (1787–1850): doi:10.1093/ref:odnb/7420} Twentieth century travellers were to
+experience similar constraints on the amount of photographic film available to them.
+
+At the same time, epigraphers have always been ingenious in their use of technical solutions. The most dramatic of these
+is perhaps the development of squeezes, which have turned out to provide records of enduring and continuing value. The
+driving force here was the need to record inscriptions in languages --- hieroglyphics, cuneiform --- which could not be
+interpreted, and even incised designs; copyists who know Greek or Latin could record texts in those languages with
+relative ease, but they too came to find squeezes useful in representing what they saw. Over time the technology was
+improved; while the plaster casts of Egyptian paintings could seriously damage the monument, paper squeezes improved in
+quality.\footnote{\url{ http://www.asia.si.edu/research/squeezeproject/sq\_making.asp}}
+
+During the 19\textsuperscript{th} century, \textbf{travel} became steadily safer, and also easier --- most dramatically as railways began to open up
+new regions. Over the same period, the creation of the first great corpora of inscriptions encouraged an increasing
+standardisation of records. Early travellers recorded as much as they could, often in haste: this normally took the
+form of a simple transcription, or drawing, of the text, with sometimes a brief mention of its support. \ Gradually,
+measurements start to appear --- often only of letters, then of the monument or fragment itself. In 1890 the Austrian
+government set up the Kleinasiatische Kommission,\footnote{
+\url{http://www.oeaw.ac.at/en/science-and-society/commissions/kleinasiatische-kommission/geschichte/}} which provided
+travellers with special notebooks: these were pre-printed with headings for Location and Position, Material, Height,
+Width, Depth, Letter heights, Shape and Condition, Number and location of squeeze, When copied and by whom.\footnote{
+Fund- und Standort, Material, Höhe, Breite, Dicke, Buchstabenhöhe, Form und Erhaltung, Nummer und Ortsangabe des
+Abklatsches, Wann und von wem copirt} On return to Vienna, the notebooks could be filed, and the squeezes stored, in
+bookshelves and drawers specially designed for the purpose.
+
+At the same time, the publication of the corpora, and the great projects from Boeckh's \emph{CIG} to Mommsen's \emph{CIL} was
+revealing the \textbf{volume} of material. The Ottoman world was becoming increasingly accessible; in the western part of the
+Roman Empire development and industrialisation --- particularly the redevelopment of Rome as a capital city - increased
+the torrent of material. Publishing an abundance of texts, accompanied by increasingly detailed information, required a
+systematised response. Organisation could be thematic: Christian inscriptions, for example, were identified as a
+separate category, requiring different expertise, although this division has remained problematic. It could be
+geographic: the Berlin Academy took responsibility for publishing the material from Italy and the west, while the
+Austrian Academy was to deal with Asia Minor and the East; but national interests also played a part, with Italian and
+French scholars publishing materials from the epigraphically prolific north African regions which their governments
+controlled. The situation also demanded finding aids --- the \emph{PIR} can be seen as a tool for accessing the material in \emph{CIL}.
+And the pressure for standardisation continued, although it was not until the 1930s that use of the Leiden conventions
+for publishing inscribed texts was agreed (and modifications continued).\footnote{\citet[262–9.7]{van1932projet} ; for further modifications, see
+\citet{dow1969conventions}; \citet{krummrey_criteri_1980}; \citet{Panciera1991}.}
+
+Much of this reflected a response to the increased volume of material becoming available as the world changed. But the
+end of the nineteenth century saw the beginning of a further technological revolution with the arrival of \textbf{photography}.
+The use of cameras for archaeological records was at first limited to established excavation sites, or cities such as
+Athens, with a reasonable amount of infrastructure and protection for cumbersome equipment; but by the 1920s cameras
+were sufficiently portable to be taken out into the countryside. More and more inscriptions were photographed; but the
+traditions which had already developed meant that they were not immediately seen as essential elements in publication.
+The \emph{Monumenta Asiae Minoris} Antiqua represent an honourable exception, established with the specific aim of taking and
+publishing photographs; but they had the support of American funding.\footnote{\citet{rouechethon2013}}
+
+Gradually, however, photographs began to effect transformations in scholarly practice. They permitted a far better
+understanding of regional, cultural and chronological distinctions, which could be communicated to readers who would
+have no opportunity to see the stones themselves. And they provided an increasing understanding of \textbf{context} --- whose
+importance was emphasised by both Louis Robert and, more recently, Werner Eck. The photograph can present the support,
+and, when applicable, the setting of that support; Robert emphasised the importance of visualising the landscape
+surrounding a particular community.
+
+Robert also demanded ever higher \textbf{standards} in the accompanying commentaries on inscriptions. But all of this raised a
+huge logistical problem: a text accompanied by a detailed description, a detailed commentary and one or more
+photographs requires a good deal of space --- and more and more texts were appearing. From the 1980s onwards it was also
+becoming standard to provide a translation into a modern language. Publication in book form was becoming increasingly
+expensive and burdensome.
+
+It was therefore changes which had been brought about by a series of technological developments which led far-sighted
+scholars --- in particular Silvio Panciera and Geza Alföldy --- to look to yet another technology. As early as the 1980s
+they both saw the value of \textbf{computers} as tools for holding, organising and searching large volumes of text; others
+quickly followed. Panciera also understood early on that working in this medium required collaboration, and the use of
+agreed standards, convening meetings to discuss such matters from 1989 onwards. With the arrival of the web, and the
+resultant possibilities for communication, these requirements became ever more important; in the early 2000s the agreed
+conventions of epigraphy were translated into a set of machine readable instructions by Tom Elliott, when he developed
+EpiDoc.\footnote{ \url{http://epidoc.sourceforge.net/}}
+
+At the same time, the steady improvements in technology were making it possible, by the 2000s, to exchange images as
+well as texts; and the arrival of the digital camera was transforming the possibilities for photographing texts. The
+epigrapher in the field no longer depends on a finite supply of film: the traditional shot of several fragments
+photographed together for reasons of economy is disappearing. Instead, the epigrapher should be expected to present
+images of every side of a monument and its setting. All these developments both enabled and necessitated the first
+large scale publications of inscriptions on line: \emph{Vindolanda Tablets Online},\footnote{
+\url{http://vindolanda.csad.ox.ac.uk/}} (2003), the \emph{U.S. Epigraphy project},\footnote{ \url{http://usepigraphy.brown.edu}} (2003–),
+\emph{Aphrodisias in Late Antiquity} (2004), \footnote{ \url{http://insaph.kcl.ac.uk/ala2004/}} the \emph{Inscriptions of Aphrodisias}
+(2007),\footnote{ \url{http://insaph.kcl.ac.uk/iaph2007/}} and others in preparation.
+
+While we were working on the materials from Aphrodisias, new possibilities were opening up: connections were becoming
+faster and more ubiquitous. More and more relevant material was being published online: what Tim Berners-Lee calls the
+\emph{next Web of open, \textbf{linked data}}. In 2008 we received a grant to start exploring the use of geodata with inscriptions, in
+the\emph{ Inscriptions of Roman Tripolitania} (2009);\footnote{ \url{http://inslib.kcl.ac.uk/irt2009/}} this approach has been
+further developed in \emph{Monumenta Asiae Minoris Antiqua XI},\footnote{ \url{http://mama.csad.ox.ac.uk/}}  and other kinds of
+interconnection are being actively explored. Linking is closely connected to \textbf{sharing}: it is becoming increasingly clear
+that one way of ensuring the survival of our materials is by making them openly available for others to reuse and share
+as widely as possible.
+
+One of the aims of both Alföldy and Panciera had been to develop collaborative corpora, places where large quantities of
+texts could be shared to enable extensive searching. After their projects moved onto the web it also became possible to
+include more and more material --- photographs and other images, and geodata. It was Silvia Orlandi who realised that the
+next step was a portal, to offer access across these and many other online epigraphic collections. Work over several
+years, by many different scholars, had established EpiDoc as a robust system, particularly for the exchange of
+information, and it was therefore available to build \textbf{EAGLE}. The spirit of this enterprise was exactly the spirit behind
+Europeana --- a project to present high quality records of heritage materials to a worldwide audience.
+
+This is an account, therefore, of an academic discipline which has evolved by engagement with a series of technological
+developments over two centuries, and is continuing to do so; it is also a story of developing steadily higher standards
+for the publication of heritage materials. The current challenge is to confront the fact that such materials will now
+be universally available, and must therefore be presented in a way that helps and supports \textbf{new audiences}. For this the
+EAGLE project has been developing valuable new resources --- such as the mobile app ---and, very importantly, encouraging
+translations. The crucial thing to realise is that it will not be possible to revert to earlier models: this project
+sets new, higher standards for epigraphic publication. This is a project which will take the subject into the future
+`on \emph{eagles' wings}', as the Bible puts it --- in another phrase used by Hollywood.\footnote{ Exodus 19.4} They
+should be taking us all with them.
+
+\bibliographystyle{sapauth-eng}
+\bibliography{../../EAGLE}
 

### 2016-04-04 keynote/eck/EAGLE_2016_Eck.tex
pdfs new folder, fixes to Rocco
removed main.pdf, file too large
updated committee
updated main
updated biblio
Eck's paper converted to LaTeX2e and added to the pdfs folder (no bibliography, though; refs are still in footnote stile -- and so they will remain, afaic)

@@
 
+\usepackage{teubner}
 % IF YOU USE XELATEX
@@
 %%%%%%%%
-\title{Keynote Speech}
-\author[]{Werner Eck\corref{first}}
+\title{Tradition and progress. The Roman World in the Digital Age - seen through Inscriptions}
+%\author[]{Werner Eck\corref{first}}
+\author[koe]{Werner Eck\corref{first}}
+\address[koe]{Universität zu Kölln} 
+\cortext[first]{Corresponding author. Email: Werner.Eck@uni-koeln.de}
 \begin{document}
@@
 
-Werner Eck is an internationally renowned historian of the Roman Empire and of Roman epigraphy.
-From February 1979 until his retirement in February 2007, he worked as Professor of Ancient History at the University of Cologne.
-From 1997 till 2002 he was president of the Association of Greek and Latin Epigraphy.
-At the Accademy of Sciences in Berlin he is responsible for the CIL and the PIR.
-With his colleagues of the Berlin-Brandenburgische Akademie der Wissenschaften he organised the 13th International Congress of Greek and Latin Epigraphy.
-In Cologne he is one of the editors of the Zeitschrift für Papyrology und Epigraphik.
-Together with German and Israeli colleagues he is editing the multilingual Corpus Inscriptionum Iudaea/Palaestinae.
-He has published about very many topics of the Imperial period, about the administration of Rome, Italy and the provinces, the social and military history, prosopography and the history of early Christianity.
-For his bibliography see \url{http://histinst.phil-fak.uni-koeln.de/index.php?id=309}
+Epigraphy has caught up with the digital age. If this point needs to be proven, 
+this very conference speaks for itself.  Its celebration in Rome does not happen 
+by chance. Indeed, the first real attempt to create a collective compilation and 
+presentation of all the Latin (and Greek) inscriptions in a single database or 
+as a connection of the existing databases was made here in Rome in 1997 through 
+Silvio Panciera during the 11\textsuperscript{th} International Congress of Greek 
+and Latin Epigraphy\footnote{See S.
+Panciera, Informatica, in: idem, Epigrafi, epigrafia, epigrafisti, vol.
+2, Rome 2006, 1913ff.
+– I would like to thank Aitor Blanco Perez for the translation of the paper.}.
+His fundamental initiative set in motion a process, which
+step by step will lead to a complete database. Very important steps in this direction 
+have been made in the meantime and the development is progressing with particular 
+emphasis at the Sapienza thanks to the collaboration between traditional epigraphists 
+such as Silvia Orlandi and IT-people, who are affiliated to the Centro Linceo Beniamino 
+Segre. These two and all the other partners in many European countries have to 
+be thanked for these present steps and we wish them both courage and success in 
+the following phase.
+
+Whoever is dealing with inscriptions, at least since the 16\textsuperscript{th} 
+century, has to deal with a continuously growing mass phenomenon. One of the reasons, 
+which instigated Mommsen to establish the \textit{CIL}, was the amount of epigraphic 
+texts, which were known at that time but had not been collected in one single place 
+but were, instead, spread in countless manuscripts and publications. Even when 
+one looks into a representative selection such as the \textit{Inscriptiones Latinae 
+Selectae }by Hermann Dessau – completed in 1916 – there are more than 9500 
+texts. Since the time of Mommsen and Dessau, however, the number of epigraphic 
+documents has increased to a size previously unknown. On the website of EAGLE one 
+can read of more than 1.5 million items, currently scattered across 25 EU countries 
+as well as the east and south Mediterranean, which should be collected in the EAGLE 
+database. In order to illustrate this growth in a more accessible way, let me refer 
+to the following example. Since the conclusion of the first edition of \textit{CIL 
+II}, namely with the supplement in 1892, the number of inscriptions from the Iberian 
+Peninsula has tripled. For some types of inscriptions, the increase is even bigger. 
+When Herbert Nesselhauf collected all the military diplomata in 1935, he could 
+put together a total of 187 documents in \textit{CIL XVI. }Today, we know more 
+than 1100 such texts, i.e. an increase of 7 times. Not in all regions of the early 
+Roman Empire and for all types of inscriptions is the increase as dramatic as in 
+the aforementioned examples. Even so, a doubling can be identified in many Roman 
+places. Just to mention some examples: in Sarmizegetusa, the first Roman colony 
+following the conquest of Dacia, around 330 inscriptions were collected in the 
+different volumes of \textit{CIL III.} Since then, more than the same number of 
+new texts has appeared, perhaps around 400\footnote{According to information by Ioan Piso}. The growth is almost always dramatic 
+when intensive excavations lead to a massive increase in the amount of inscriptions 
+available, as somehow happens in the cases of Sagalassos or Perge in Pamphylia\footnote{In Sagalassos almost all the new inscriptions are in Greek}. 
+Perge provided only 4 Latin inscriptions for the \textit{CIL III} and today there 
+are at least 42\footnote{See the Clauss-Slaby database}. Even more dramatic is the growth in Caesarea Maritima. In \textit{CIL 
+}there were only 3 Latin inscriptions known from the capital of the province of 
+Judea, while now we have in the Corpus Inscriptionum Iudaeae/Palaestinae more than 
+270 Latin documents\footnote{W. Eck, Comprendere contesti: Il nuovo multilingue Corpus Inscriptionum Iudaeae / Palaestinae, in: Rationes Rerum 1, 2013, 17 ff.}.
+
+Each of us is familiar with this massive increase in epigraphic monuments and we 
+are reminded of this every year by the always more extensive volumes of \textit{AE} and \textit{SEG}. Even if it is impossible to provide an exact number of all the 
+Latin inscriptions, one can still take as a starting point the fact that there 
+are approximately 500.000 texts available for our work, including the so-called 
+\textit{instrumentum domesticum}. There are more than 495.000 inscriptions in the 
+Database Clauss-Slaby. Around 72.000 we can find in the Heidelberg Database and 
+around the same number in the Epigraphic Database of Rome, from which the Project 
+EAGLE emerged, not to mention other participants as, among others, \textit{Hispania 
+Epigraphica }or the Bari Epigraphic Database \footnote{See the individual databases}. With regard to the Greek inscriptions, 
+it is much more difficult to give an approximate number because their digital record 
+is not as advanced as in the case of the Latin inscriptions. As for inscriptions 
+in other languages – such as Punic, Hebrew-Aramaic, Demotic or Nabatean – let 
+me exclude them from this discussion, even though they are just as part of the 
+epigraphic tradition of the Greek and Roman antiquity as Latin and Greek inscriptions 
+– we should not forget this very important fact\footnote{\label{fn:7}At the 14.
+International Congress of Greek and Latin Epigraphy there was a special section for this phenomenon: W.
+Eck, Ph.
+Hyse, J.
+Price, St.
+J.
+Seidlmayer, I.-X.
+Adiego, in: ÖFFENTLICHKEIT – MONUMENT – TEXT.
+XIV Congressus Internationalis Epigraphiae Graecae et Latinae Akten, hg.
+W.
+Eck - P.
+Funke, Berlin 2014, 159 ff}.
+Our work as historians or philologists
+is based on all those texts. But, the bigger the number of documents providing 
+information is, the more difficult it becomes to find all texts. Even more difficult 
+is to collect selectively those texts related to the topic on which one wants to 
+work. And yet, the complete collection is crucial for the scientific result.
+
+Let's go back only two or maybe three decades, a time which many of you in this 
+room have vividly gone through. In that time it was clear, what we had to do, when 
+we wanted to explore a topic on the basis of epigraphic evidence. We were to search 
+– sometimes in countless volumes of inscriptions and partly supported through 
+more or less detailed indexes. Depending on the type of subject on which we were 
+working, these indexes were not always perfect tools to find the material. Very 
+often they could not even be such perfect tools, since many phenomena are not accessible 
+through precise termini in the epigraphic material, but rather through the entire 
+content of the text. I am sure that you all have your own experiences. When I worked 
+on the administration of Italy in the high Roman Empire, the indices were helpful 
+but not sufficient. I had to go through each and every of the epigraphic corpora, 
+which means that at least the inscriptions had to be read. In total, the look through 
+all the necessary volumes took me almost two years.
+
+Is that different today because there are numerous databases available? At least 
+with regard to the Latin inscriptions almost all of them are collected, although 
+with very different individual information. With few clicks, we can find an answer 
+for many questions in few seconds – sometimes one has to wait a bit longer, since 
+the server does not work properly or some other problems exist. Indeed, as an already 
+deceased colleague once told me long before the digital period in 1973: \textit{Pazienza} 
+is the greatest virtue of the epigraphist. Today we are ready to forget this virtue 
+thanks to our computers. At any rate, most of the relevant material available in 
+databases appears quickly on the screen. It is not necessary to give examples. 
+And yet, something which we all know: it is by no means guaranteed that we have 
+found absolutely all the relevant material which exists; the reason for these gaps 
+are inter alia, that not a small part of our epigraphic testimonies survives only 
+as fragments and can be restored in different ways. For example, if we would try 
+to find all the \textit{quaestores }of the province of Achaea in the imperial age, 
+we would not find the following fragmentary inscription from Corinth, the capital 
+of the province, because the text has read as follows\footnote{\emph{CIL III} 537}:
+
+\begin{quote}
+
+{\it Q(uinto) Vili[o --] / Titia[no] / Quadra[to --] / IIIIvir(o) vi[ar(um) cur(andarum), trib(uno)] / la\mqs{t}icl(avio) \mqs{le}g(ionis) [--, \textbf{leg(ato) prov(inciae)] / Ach(aiae)}, ob ius[titiam et fidem,] / quam circa [universam] / Achaia[m exhibuit,] / qu\mqs{a}m pro pr(aetore) [administravit.] / \mqs{M}(arcus) An\mqs{to}nius M(arci) [f(ilius)] / Nigrinus [--].}
+\end{quote}
+
+Our search in a database would have recorded a proconsular legate of Achaea whose 
+function in the inscription is nevertheless restored. The restoration of the position 
+of provincial legate is not impossible, but it is also not very likely. The restoration 
+of the office of quaestor:\textit{ [-- quaest(ori) prov(inciae)] / Ach(aiae) }instead 
+of \textit{[-- leg(ato) prov(inciae)] / Ach(aiae) }fits in the lacuna much better\footnote{In the database Clauss-Slaby there are around 68 examples in Latin inscriptions, in the database of the Packard Institute on can find 10 more in Greek texts.}.
+For a young senator who had completed an office of the \textit{vigintiviri }and 
+then a tribunate in the legion, the \textit{quaestura }should actually follow in 
+a normal career and not the position of a proconsular legate under a proconsul. 
+Of course, we know that exceptionally young senators already before the \textit{quaestura 
+}were acting as legates of a proconsul in his province. In total, there are only 
+four of such cases. By contrast, much more senators state in their inscriptions 
+that they have held the office of provincial \textit{quaestor} as\textit{ pro praetore}. 
+The \textit{pro pr(aetore) }preserved in the inscription could therefore refer 
+to the \textit{quaestura}. As the \textit{quaestura }normally follows the\textit{ 
+}viginvirate and the legionary tribunate, this rule should automatically be observed 
+in a fragmentary career\footnote{See now the new reconstruction of the inscription by W.
+Eck, Quaestor, nicht legatus provinciae Achaiae.
+Zu einer Inschrift des Q.
+Vilius Quadratus aus Korinth, (in preparation).}.
+This kind of result is not achieved through an inquiry
+into a database but through a constructive consideration on the basis of personal 
+experience. Such knowledge will remain also the basis of our discipline in the 
+future. Databases complement our knowledge but they cannot replace it. This can't 
+be their business.
+
+Inscriptions are in the first place texts which contain certain evidence as such. 
+But inscriptions do not only consist of text, although this may commonly seem to 
+be the case in the literature. In German, one speaks of ``\textit{Inschriften 
+errichten}'' and in Italy one can say: \textit{una iscrizione eretta}\footnote{It seems not necessary to give examples for this type of formulation; they are abundant.}. In the Prosopographia 
+Imperii Romani,\textit{ }the expression \textit{titulum ponere }was long in use 
+and applied to inscriptions of all kinds. With these expressions, it was not clear, 
+what specific function was connected with an epigraphic text and the support on 
+which the inscription appears. Nonetheless, the support already contains its own 
+message and that was, on many occasions, the primary message for the Greek and 
+Roman public\footnote{\label{fn:12}Cf.
+now W.
+Eck, Wie ehrt man Mitglieder der staatlich-städtischen Elite?
+Inschriftenbasen unerkannter Reiterdenkmäler, Kolloquium Kassel März 2015, hg.
+K.
+Ruffing – H.
+Schneider (in print)}.
+In order to understand what an inscription can today convey as evidence,
+it is fundamental to include the \textbf{monument }with which the inscription was 
+connected. This can be an altar, an architrave of a temple, a mausoleum or, for 
+instance, the base for a statue, which is explained by the inscription. In enquiring 
+about the content of inscriptions, we muss decisively take into consideration the 
+function that inscription and monument shared in the Graeco-Roman world.
+
+However, when one looks for the function of an inscription in a database, the difficulties 
+begin – if I were to speak about my own experiences. On many occasions, this 
+function is not found, if one enquires about what is directly conveyed in the inscription, 
+because this exact function is commonly not mentioned in the epigraphic text. This 
+is not surprising. To mention the function was not particularly necessary because 
+the ancient observer completely saw, on what an inscription was written. Given 
+that today this context is mostly not preserved, we must infer it from the support 
+whose form and particular details are connected with it. This necessity has meanwhile 
+become almost common knowledge – in contrast with Mommsen's time\footnote{W.
+Eck, Mommsen e il metodo epigrafico, in: Concordia e la X Regio.
+Giornate di Studio in onore di Dario Bertolini, Atti del Convegno Portogruaro 22-23 ottobre 1994, Padova 1995, 107 ff.}.
+On the EAGLE
+website, this is specifically taken into account with the option \textit{advanced 
+search, }in which it is possible to browse the database according to object type. 
+With this option, one can find the categories, for example, of \textit{statue, 
+statue base, statue base shaft or plinth}. Hence it is already possible to obtain 
+a result with a high degree of differentiation, if one browses selectively. However, 
+not all the different categories of statues are grasped with these search terms. 
+Just to mention an example: \textit{bigae }or \textit{quadrigae, }which were erected 
+to honour a person, can only be found if these terms would also appear in the inscription. 
+But in the epigraphic text the object is directly mentioned only in a limited number 
+of cases. The ancient observer saw the object which did not need any additional 
+reference. For us, on the other hand, the specific types of statues are generally 
+recognisable only if we include the support of the inscription in the analysis. 
+The original statues are almost always lost or, at least, not connected with the 
+inscription. Enquiring about the types of statues is not a child's play. For the 
+most part, they rather say fundamental things for both the honoured and the honouring 
+individuals.
+
+One example from Volsinii (modern Bolsena) can illustrate this\footnote{AE 1980, 426.}. It is an honorific 
+monument set up to commemorate a senator of the Hadrianic-Antonine age. The surviving 
+fragmentary \textit{titulus honorarius }records his names: Pompeius Vopiscus C. 
+Arruntius Catellius Celer Allius Sabinus, followed by his \textit{cursus }and, 
+finally, the city of Carthage as dedicator. The text, however, does not specify 
+in which figurative way the senator was presented to the audience of ancient Volsinii. 
+This is also not attempted in the epigraphic database of Rome; there it is only 
+said (as in the Epigraphischen Datenbank Heidelberg) that the text is to be read 
+on a base\footnote{EDR077846 = HD005099 (M.G. Schmidt).}. While this is correct, the plain term ``base'' does not allow to recognise 
+the most important element, namely with what type of honour had the city of Carthage 
+honoured the senator coming from Volsinii. The particular quality of the honour 
+is, nevertheless, revealed by the size and shape of the inscribed plaque. The plaque 
+is 76.5 cm high and 1.90 m wide\footnote{P. Gros, MEFRA 92, 1980, 977 ff.}. The monument was so wide that a normal statue 
+could not have stood on it; this would have appeared completely out of place. Given 
+the width of the inscribed slab, it is to be inferred that a \textit{biga }or perhaps 
+even a \textit{quadriga} was set up for the senator\textit{.} He should therefore 
+have completed an action beyond his official duty on behalf of Carthage for which 
+the city decided to recognise his activity in this lavish way and to send ambassadors 
+to Volsinii to erect the monument.
+
+A similar case can be inferred for Larcius Priscus, a legate of the \textit{legio 
+III Augusta}\footnote{\emph{CIL VIII} 17891 = Dessau 1055.}. The city of Thamugadi honoured him in a similar way to that of Carthago 
+for the senator in Volsinii. On the forum of the colony founded by Trajan, not 
+a normal honorary statue not even an equestrian statue – as previously suspected\footnote{\label{note:18}J.
+Bergemann, Römische Reiterstatuen.
+Ehrendenkmäler im öffentlichen Bereich, Mainz 1990, 147 Nr.
+E 90:  ``equestrian statue''.}
+– was set up but rather a \textit{biga }on which the statue of Larcius Priscus 
+was presented to the citizens of Timgad\footnote{The inscription is 188 cm wide, much more than necessary for an equestrian statue.}. But in EAGLE one finds the note:
+
+Type of object: unbestimmt = undefined (= EDH)\footnote{EDH031159 (B. Gräf).}.
+
+But one can see the type of the object on the photo, published by Zimmer in 1989\footnote{G.
+Wesch-Klein, Katalog der Inschriften, in: G.
+Zimmer, Locus datus decreto decurionum.
+Zur Statuenaufstellung zweier Forumsanlagen im römischen Africa, München 1989, 71 Nr.
+T 4.}.
+More examples of this kind could be given. In my opinion, an important consequence 
+is therefore clear. If databases were to fulfil their function completely, then 
+it would be not only useful but rather very necessary to record an indication of 
+the concrete form of the honorific monument in such cases. Otherwise, a crucial 
+part of the potential information would be lost.
+
+Of course such a degree of precision is not always possible, but very likely when 
+– as shown by the two previous examples – the form of the support provides 
+evidence beyond the text of the inscription that can be searched in the database. 
+This happens for example in many inscriptions which in antiquity were to be seen 
+under \textit{statuae equestres}. Sometimes this form of honour is recorded in 
+the text of the inscription itself. In the database Clauss, there are around 45 
+records if one looks for \textit{statuae equestres }or \textit{statuam equestrem},\textit{ 
+}in the database EAGLE around 27. That EAGLE shows fewer cases is naturally due 
+to the fact that this database is still under construction. Either with 45 or 22 
+inscriptions, this only records in any case a minority of the texts which were 
+once to be read under such \textit{statuae equestres}. In the monograph on equestrian 
+statues written by Johannes Bergemann in 1990, 128 inscriptions are collected, 
+which were once connected with an equestrian statue according to the investigations 
+of the author \footnote{See n. \ref{note:18}.}. Besides the statues directly referred to as such in the text, there 
+are also those containing other hints which allow us to recognise that the honoured 
+person was once presented on a horse.
+
+Some examples should illustrate which equestrian statues cannot be recognised as 
+such without these extra hints. In the colonia Aelia Capitolina, modern Jerusalem, 
+an equestrian statue was erected for Antoninus Pius through the council of decurions, 
+apparently straight after his accession in the year 138/9. In the text of the honorific 
+inscription, the type of statue is not mentioned but it can be inferred from the 
+size of the inscription\footnote{Corpus Inscriptionum Iudaeae/Palaestinae I 2, 718.}. Therefore, it must be registered as an equestrian statue 
+in a database. The same applies to the honours bestowed in Lepcis Magna upon the 
+governor of the late-antique province Tripolitana, Flavius Nepotianus. In the inscription 
+itself, it is only spoken of a \textit{statua marmorea} that should memorialise 
+the services of the magistrate\footnote{IRT 565.}. However, the base enables to recognise that a \textit{statua 
+equestris }once stood there because this base is 170 cm deep\footnote{Bergemann (n.
+\ref{note:18}) no.
+83.
+Cf.
+the inscription for the flamen provinciae Pompeius Cerealis Salvianus in Lepcis Magna: the basis is 70 cm high, but 160 cm deep (IRT 602).}.
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.5]{images/fig01ReiterstatueZeichnung.pdf}
+\caption{Equestrian statue: drawing}
+\label{fig:1}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+\includegraphics[scale=0.5]{images/fig02ReiterstatueVonDerSeite.pdf}
+\caption{Equestrian statue: view from the side}
+\label{fig:2}
+\end{figure}
+
+Equestrian statues are, nevertheless, not uniform. They can also provide, in turn, 
+additional information by their specific appearance. They were dedicated in very 
+different forms; above all in very varied sizes. For example: Alfius Secundus, 
+a \textit{flamen perpetuus }in Africa proconsularis, set up two or three equestrian 
+statues of the emperor Septimius Severus\footnote{\emph{CIL VIII} 14370 (Avedda).}. Even though these statues represent the 
+emperor, they must have been small ones because the base is only 54 cm high and 
+35 cm wide\footnote{Bergemann (n.
+\ref{note:18}) no.
+81.}.
+These small equestrian statues fitted probably into the context in which they stood.
+On the other hand, gigantic equestrian statues were erected of
+all the emperors – and not only of them. An exact relation between the size of 
+a monument and social rank did not exist. Various factors could be relevant in 
+such cases. Nonetheless, the size can already tell us something about the status 
+of the honorand and the intention of the dedicator.
+
+\begin{figure}[!bp]
+\centering
+\includegraphics[scale=0.5]{images/fig03}
+\caption{Brescia Basis with 2 pillars}
+\label{fig:3}
+\end{figure}
+
+Most \textit{statuae equestres }either rose from bricked bases or the base was 
+made of solid stone. The base for the statue of C. Minicius Italus in Aquileia 
+was built with bricks and then covered with marble slabs\footnote{\emph{CIL V} 875 = Dessau 1374 = Inscr.
+Aquileiae I 495; G.
+Alföldy, Römische Statuen in Venetia et Histria, Heidelberg 1984, 98 f.
+no.
+87.
+For a photo see EDCS-01600153.}.
+The one of the young
+senator L. Fabius Severus in Tergeste was made of a single solid stone\footnote{\emph{CIL V} 532 = Dessau 6680 = Inscr. Italiae X 31. Photo : EDCS-04200621.}. The latter 
+applies to the majority of these monuments. Even so, at least in the first century 
+AD there was also a type of an equestrian statue that remained unattended in research 
+until recently. For there existed equestrian statues, which seemed much lighter 
+and did not stand on an apparently solid basis\footnote{\label{fn:30}For the following discussion see W.
+Eck - H.
+von Hesberg Tische als Statuenträger, MDAI (R) 111, 2004 [2006], 143 ff.}.
+Sometimes the base was simply made
+of a foundation slab, two supporting pillars and a cover plate on top, on which 
+the equestrian statue stood (Fig. \ref{fig:1} and \ref{fig:2}). To the best of my knowledge, there 
+is only one fully surviving example of this type, which is today kept in the museum 
+of Brescia. (Fig. \ref{fig:3}) This example presents a posthumous honour for a 6-year old 
+boy decreed by the \textit{ordo} \textit{decurionum }of Brixia. The setting up 
+of the equestrian statue was executed by the father of the deceased\footnote{\emph{CIL V} 4441 = Inscr. Italiae X 5, 232.}. In this case, 
+we recognise such a particular type of equestrian statue only because the entire 
+monument survives, in the inscription the father mentioned only a \textit{statua 
+equestris}. Yet research did not consider the piece which can be seen in the museum 
+at Brescia as a special type of honorary monument, but rather as a unique object. 
+There are, however, not a few inscriptions that were connected with this statue 
+type in the Roman age. The central feature of this type are always two supporting 
+pillars and a cover plate on top, on which the equestrian statue stood. Two kinds 
+of pillars can be distinguished, and they differ clearly. There are pillars such 
+as those used as inscribed support in the form of the example from Brescia and 
+there are the so-called \textit{trapezophora }on which an inscription is not rarely 
+found\footnote{For more information see Eck - von Hesberg (n.
+\ref{fn:30}).}.
+Just to mention some: we know for the young senator P.
+Numicius Pica Caesianus\footnote{\emph{CIL VI} 3835 = D 911= VI 31742 = 31743.}
+two \textit{trapezophora }from Rome, quite a number from Torino for Q. Glitius 
+Atilius Agricola\footnote{\emph{CIL V} 6974 – 6987; Eck - von Hesberg (n. \ref{fn:30}) 186 f.}, a single one for T. Flavius Cimber, a municipal magistrate from 
+Urvinum Mautaurense\footnote{\emph{CIL XI} 6062: \emph{T(ito) Flavio L(uci) f(ilio) Ste(llatina) Cimbro pont(ifici), aed(ili) bis, IIIIvir(o) i(ure) d(icundo), quinq(uennali), praef(ecto) fabr(um), d(ecreto) d(ecurionum)}.} and many others\footnote{See Eck - von Hesberg (n.
+\ref{fn:30}) 180 ff.: a list of the inscribed trapezophora which were known to us in 2003/4.}.
+
+These epigraphic monuments can partly be found in the EAGLE database too. Here 
+one has to search for the term \textit{trapezophorus }as object type\footnote{\emph{Trapezophoron} is the normal \emph{terminus technicus}.}. Table feet 
+of this kind, which contain inscriptions, are known in relatively big numbers because 
+of their special shape; they have always been categorised individually\footnote{For the literature see Eck - von Hesberg (n. \ref{fn:30}) notes 1-3.}. However, 
+they were not hitherto considered as parts of statue bases but rather of tables, 
+real \textit{mensae}. Today it is no longer questionable that these \textit{trapezophora 
+}with inscriptions were in reality parts of statue dedications, with an equestrian 
+statue which stood on this specially arranged base\footnote{Eck - von Hesberg (n. \ref{fn:30}) with the general discussion of this type of honorary monuments.}. As long as these inscriptions 
+are marked up with the keyword \textit{trapezophorus }(sic!) in the EAGLE database, 
+they can be found without problems. In the Heidelberg database, however, with the 
+keyword \textit{trapezophorus/um} one will find nothing, although there are three 
+such objects in the database; but there they are\textit{ }registered under the 
+keyword \textit{mensa}. If one is entering \textit{mensa }in the EDH, numerous 
+and extremely varied objects appear, which have by no means the same function\footnote{For an example of a mensa in the concrete sense: IRT 590: \emph{Ti(berius) Cl(audius) Amicus M(arcus) Heliodorius Apollonides aed(iles) mensas p(ecunia) s(ua) d(ono) d(ederunt)}; the inscription is engraved on the frame of the table; cf.
+also \emph{CIL III} 15184, 18 = AIJ 310.
+On the other side also a mensa ponderaria can be found by surging for a mensa: AE 1905, 37 = HD030144; \emph{CIL III} 15025 = HD005744.}.
+If the keyword \textit{mensa }is entered in the EDH and connected with the search 
+parameter ``honorific inscription'', it is possible to find three \textit{trapezophora} 
+– in the category \textit{mensa}\footnote{HD002671 = EDR111713; HD006112 (Ruck) = EDR076783; HD025725 (Féraudi) = EDR073154 (also described as \emph{mensa}).}.
+But who would imagine that the combination
+of \textit{mensa} and honorific inscription is necessary to find this type of monument? 
+Also in EAGLE not all the objects that belong to this form of monument can be found 
+with a single search term. That means as a consequence, that in the databases combined 
+by EAGLE the terms for specific objects should be uniform; for the moment this 
+is not the case.
+
+To give one example here: We know one \textit{trapezophoron } with an inscription 
+from Cosa in which Drusus Caesar, son of Tiberius, is mentioned; it appears in 
+EDR076783 with no specific characteristics for the object, because the information 
+comes from the EDH. In EDH006112, the object of this inscription appears as a \textit{mensa} 
+and an ``Ehreninschrift'', but not as a trapezophoron. On the other side: In EAGLE, 
+\textit{mensa }can almost only be used as long as the word occurs in the inscribed 
+text, not as term for an archaeological object\footnote{For an exception see the preceding note.}. Again the harmonization of search 
+terms becomes extremely important in order to enable a quick and safe search\footnote{In EDR077116 an inscription with the text: \emph{L(ucius) Ansius Quintill[i]anus m\d{e}[nsam -{}-{}-]} is called trapezophorus, although in reality it is a \emph{mensa}, as the text itself tells us and as the photo in EDCS-10700899 clearly shows.
+The inscription is engraved on the frame of the tabula for a \emph{mensa}.}.
+
+As mentioned before, a total of 23 \textit{trapezophora}, which can be described 
+as honorific monuments for single persons, can be found in EAGLE. These are by 
+far not all the inscriptions, which were once connected with equestrian statues, 
+which did not stand on a solid base but rather on a plate supported by two pillars. 
+The posthumous honours in Brescia for the 6-year old P. Matienus Proculus is, as 
+already mentioned, such an example\footnote{\emph{CIL V} 4441 = Inscr. Italiae X 5, 232.}. But in the databases, inscriptions for such 
+monuments are not described with their specific features\textbf{ }as\textbf{ }the 
+entry for the monument from Brescia makes clear; of course normally they can be 
+found with the term honorific inscription, i.e. as \textit{titulus honorarius}; 
+but that does not really help, there are too many \textit{tituli honorarii }in 
+the database. The text for Publius Matienus Proculus one would not even find with 
+the word \textit{titulus honorarius}, because it is categorised as sepulcralis\footnote{EDR090232. The precise terminus would be: \emph{titulus honorarius postumus}.}. 
+To single out the different types, one has to describe in the data bases the special 
+singularities, which identify the particular functionality of such monuments. Here 
+one example.
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.6]{images/fig04}
+\caption{Herennius Caecilianus -- frontside}
+\label{fig:4}
+\end{figure}
+
+In Sirmione (ancient Sirmio), at the Lago di Garda, an inscription was found in 
+1960. It once belonged to an honorific monument for the young senator C. Herennius 
+Caecilianus.
+The following text was published by Alberto Albertini in 1973 (Fig.
+\ref{fig:4})\footnote{\label{fn:46}A.
+Albertini, Un patrono di Verona del secondo secolo d.C.: G.
+Erennio Ceciliano, in: Il territorio Veronese in età romana Verona 1973, 439ff.
+= Alföldy, Römische Statuen (n.
+28) 253; R.
+Bertolazzi - V.
+Guidorizzi, Supplementa Italiae 28, nr.
+7.}:
+
+\begin{quote}
+\textit{C(aio) Herennio} \\
+\textit{C(ai) f(ilio) Pob(lilia)} \\
+\textit{Caeciliano,} \\
+\textit{adlect(o) in senat(um)} \\
+\textit{ab imp(eratore) Hadriano} \\
+\textit{Aug(usto), q(uaestori) prov(inciae) Narb(onensis),} \\
+\textit{trib(uno) pleb(is), $\overline{IIII}$viro i(ure) d(icundo)} \\
+\textit{Veronae,} \\
+\textit{patrono} \\
+\textit{d(ecreto) d(ecurionum)}. \\
+\end{quote}
+
+The text is not particularly interesting with regard to its content. It only records 
+the beginning of a senatorial career in the Hadrianic period. The young senator 
+came apparently from Verona where he was also \textit{IIIIvir iure dicundo }and 
+patron of the city. For this very reason the city wanted to honour him, naturally 
+with a statue. This was also assumed by Albertini, who suggested a bronze statue. 
+The text also appears in the EDR and EDH\footnote{EDR093835 and HD033596.}.
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.5]{images/fig05}
+\caption{Herennius Caecilianus -- backside}
+\label{fig:5}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.6]{images/fig06}
+\caption{Herennius Caecilianus -- lateral}
+\label{fig:6}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.46]{images/fig07}
+\caption{Herennius Caecilianus -- from above}
+\label{fig:7}
+\end{figure}
+
+The inscribed support consists of a slab, 100 cm high, 59 cm wide and 29 cm deep. 
+The text is surrounded with a frame on the front side. The same frame surrounds 
+also both laterals and, above all, the backside, which is crucial (Fig. \ref{fig:5} and \ref{fig:6}). 
+For this shows that the backside was elaborated with the intention to be exposed, 
+a detail also observed by Albertini as he accordingly commented that the base (with 
+the statue directly standing on it, in his view) was not \textit{adossata a una 
+parete, ma eretta in uno spazzio}\footnote{Albertini (n. \ref{fn:46}).}. It is correct that the inscription could not 
+have been \textit{adossata a una parete}. However, both the first editor and all 
+the others who dealt with the inscription thereafter have simply not wondered how, 
+then, could the statue stand on a slab which is only 29 cm deep (Fig. \ref{fig:7}). Furthermore, 
+the three holes for the corresponding dowel on the top side of the slab show that 
+no statue was fixed there, but rather, something completely different, another 
+horizontal slab. If one compares the evidence concerning this inscribed support 
+with the equestrian statue of Matienus Proculus in the Museum of Brescia, shown 
+before, which also had the backside of the front pillar elaborated with the intention 
+to be seen from both sides like also the second uninscribed pillar – the following 
+result becomes immediately clear: Herennius Caecilianus was not simply honoured 
+with a statue by the people of Verona, but with an equestrian monument standing 
+on an cover plate, which was resting on two pillars, whose front side with the 
+inscription was 29 cm deep (the second pillar is lost). This ``lighter'' 
+version of an equestrian statue was perhaps chosen by Verona because the monument 
+should probably be set up in the estate of the senator. The fact that this type 
+of monuments was not unusual in this region is shown not only by the posthumous 
+monument of Matienus Proculus in the Museum of Brescia, but also by two other pillars 
+in the same museum, which are very similar to the monument of Herennius Caecilianus; 
+these too are elaborated on the backside with the intention to be exposed. One 
+of the pillars bore once an inscription that was later erased\footnote{A more detailed argumentation in Eck, Wie ehrt man Mitglieder der staatlich-städtischen Elite?
+(n.
+\ref{fn:12})}, which makes it impossible
+to know who was honoured in such a way (Fig. \ref{fig:8}, \ref{fig:9}, \ref{fig:10}).
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.15]{images/fig08}
+\caption{Brescia -- erased inscription on a pillar}
+\label{fig:8}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.15]{images/fig09}
+\caption{Brescia -- pillar}
+\label{fig:9}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.05]{images/fig10}
+\caption{Brescia -- pillar, reused later}
+\label{fig:10}
+\end{figure}
+
+What are the consequences of these observations? Databases are by now indispensable 
+in the epigraphic work. They speed up not only the work but allow, above all, to 
+recognise evolutions through the possibility of examining texts systematically: 
+e.g. formulae or forms of abbreviations. Previously, this was only possible through 
+the arduous and time-consuming examination of endless volumes. Therefore very often 
+this systematic search was not be done or the result was supported only by a slim 
+documentary basis. This is now much easier especially when it is possible – like 
+now in EAGLE – to access many databases at the same time. In order to achieve 
+an even more effective and extensive access, it appears to me, that a stricter 
+coordination between the different databases is necessary, a harmonization that 
+should also concern the question, which search terms are necessary and possible. 
+If the same phenomena, i.e. inscriptions, which had the same function, are shown 
+with different terms, a uniform search becomes necessarily difficult, if not completely 
+impossible. I referred to the already examined terms \textit{trapezophoron} and 
+\textit{mensa}. Under the same term, phenomena and documents with very different 
+functions should not appear together. A \textit{mensa }should not refer both to 
+a \textit{mensa ponderaria} and, at the same time, to the pillars of an equestrian 
+statue of the type described above, as somehow occurred in Cosa with the \textit{trapezophoron 
+}for Drusus Caesar.
+
+The honours for Herennius Caecilianus introduce another further possibility to 
+make the utilization of databases for the users even more diverse. In EDR093835, 
+a photograph of the monument of Herennius Caecilianus was published, naturally 
+of the front face with the inscription. Yet a photograph of the backside would 
+be equally necessary to recognise the specific function of the stone and to make 
+it immediately clear that the slab was elaborated with the intention to be exposed. 
+In this way the essential evidence for its function would be provided. Of course 
+such photographs are not always available. However, during the preparation of entries 
+one should always check whether more pictures are available and not only those 
+of the side with the text. The text remains essential, but it must be completed 
+with exact observations about the support of the inscription. Often the meaning 
+of the monument can only be reached in this way. This was the central theme of 
+the 14\textsuperscript{th} Congress of Epigraphy. Databases have the capacity to 
+provide all those details which are necessary for the complete interpretation of 
+an epigraphic monument\footnote{For the acts of the Congress see n. \ref{fn:7} above.}. The high costs that the inclusion of many pictures in previous 
+publications entailed are not a crucial problem any longer.
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.06]{images/fig11}
+\caption{Base, Sarapis dedication -- Front}
+\label{fig:11}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.06]{images/fig12}
+\caption{Base, Sarapis dedication -- Eagle}
+\label{fig:12}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.06]{images/fig13}
+\caption{Base, Sarapis dedication -- Victoria}
+\label{fig:13}
+\end{figure}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.06]{images/fig14}
+\caption{Base, Sarapis dedication -- Eagle}
+\label{fig:14}
+\end{figure}
+
+At the end of my presentation let me once more demonstrate the necessity to describe 
+clearly the monumental features of an inscription and the photographic documentation 
+now with an example that comes from the material of the \textit{Corpus Inscriptionum 
+Iudaeae/Palaestinae}\footnote{This monument will be included in volume V of the Corpus Inscriptionum Iudaeae/Palaestinae.}. A short time ago one could find the inscription only in the 
+database Clauss-Slaby and only the text\footnote{EDCS-15200169; now there are also photos connected to the text.
+Under the adresse: \url{http://www.antiquities.org.il/t/item_en.aspx?CurrentPageKe y=33&rock=0} the monument can be seen on the homepage of the Rockefellermuseum in Jerusalem; there is only mentioned that an inscription is written on the monument; but the text is not given}.
+Avi Yonah, who greatly contributed to
+the collection and publication of epigraphic monuments during the time of the British 
+mandate in Palestine, published in 1946 an inscription found near the legionary 
+camp of the \textit{legio VI Ferrata}, near Caparcotna = Legio\footnote{M. Avi-Yonah, Newly discovered Latin and Greek inscriptions, QDAP 12, 1946, 89 = AE 1948, 145.}. It is a round monument, 
+1.05 m high, which he – like many epigraphists later – presented as an altar\footnote{For example.
+L.
+Vidman, Sylloge inscriptionum religionis Isiacae et Sarapiacae, Berlin 1969, 182f.
+no.
+361; F.
+Mora, Prosopographia Isiaca.
+I.
+Corpus Prosopographicum Religionis Isiacae, Leiden – New York – Kopenhagen – Köln 1990, 243 Nr.
+577; N.
+Belayche, Iudaea-Palaestina, The Pagan Cults in Roman Palestine (Second to Fourth Century), Tübingen 2001, 59 ff.; L.
+Bricault, Recueil des inscriptions concernant les cultes isiaques (RICIS), 3 vols., Paris 2005, 508f.
+no.
+403/0201; N.
+Belayche, Les dévotions à Isis et Sérapis dans la Judée-Palestine romaine, in: Nile into Tiber.
+Egypt in the Roman World, hg.
+L.
+Bricault – M.J.
+Versluys – P.G.P.
+Mayboom, Leiden 2007, 451f.; W.
+Eck, Rom und Judaea.
+Fünf Vorträge zur römischen Herrschaft in Palaestina, Tübingen 2007, 186f.; P.
+Figueras, The Pagan Image of Greco-Roman Palestine and Surrounding Lands, Oxford 2013, 78.
+96.}.
+The monument shows three perfectly elaborated relieves on three sides: a Victoria 
+standing on a globe with a tropaion as well as a victory crown in the hands, and 
+two eagles that carry a thunderbolt in the crawls and a crown in the beak. The 
+inscription on the front face reads (Fig.
+\ref{fig:11}, \ref{fig:12}, \ref{fig:13}, \ref{fig:14})\footnote{The text of the inscription is the one corrected by W.
+Eck, Sarapis und die legio VI Ferrata.
+Die Weihung einer Sarapisbüste für das Wohl des Kaisers, ZPE 198, 2016 (in print).}:
+
+\begin{quotation}
+\textit{Pro salute et incolumitate / domini nostri [[Imp(eratoris) Caes(aris) M(arci) 
+Aur(eli) Antonini Aug(usti)]] / praesentissimum deum Mag(num) Sarapidem / leg(io) 
+VI Ferrat(a) F(idelis) C(onstans) [[Antoniniana]] / Iulius Isidorianus p(rimus) 
+p(ilus).}
+\end{quotation}
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.06]{images/fig15}
+\caption{Base, Sarapis dedication -- Upper side}
+\label{fig:15}
+\end{figure}
+
+In the scholarly discussion, the monument was almost universally presented as an 
+altar and logically considered as a dedication \textbf{to} the god Sarapis.\textbf{ 
+}Nevertheless, this understanding did not take into consideration the clear testimony 
+of the inscription, in which it is written: \textit{praesentissim}\textit{\textbf{um}}\textit{ 
+de}\textit{\textbf{um}}\textit{ Mag(num) Sarapid}\textit{\textbf{em}}\footnote{One exception: O.
+Stoll, Zwischen Integration und Abgrenzung.
+Die Religion des römischen Heeres im Nahen Osten.
+Studien zum Verhältnis von Armee und Zivilbevölkerung im römischen Syrien und den Nachbargebieten, St.
+Katharinen 2001, 280, who saw the consequences of the accusativ for the interpretation of the monument.}.
+This evidently
+means that Sarapis is not mentioned here as the one to which something was dedicated 
+but rather that his figurative representation is the dedicated object. There is 
+no doubt that a representation of Sarapis was set up as a votive gift, probably 
+in a shrine, perhaps for Egyptian gods near the camp of the legio. The fact that 
+a representation of Sarapis was dedicated means, consequently, that we are not 
+dealing with an altar but with a base on which the representation found its place. 
+Above all, it should not have been omitted from the beginning that there is a remarkable 
+peculiarity on the upper side of the base,\textbf{ }namely a completely rounded 
+hollow with a diameter of 29 cm and a depth of 9 cm (Fig. \ref{fig:15}). It is an almost 
+half sphere that was perfectly chiselled and smoothed from the marble. Sometimes 
+earlier scholars noticed this hollow, as already Avi Yonah, but concluded that 
+over the ``altar'' a focus would have stood in the hollow, on which the offerings 
+could be given. However, a shallow cavity can be seen in such cases at the most, 
+but not the half spherical hollow found here. As this half sphere can only be carved 
+with a considerable amount of work, it must have a specific meaning, which should 
+be connected, as the entire monument, with the depiction of the god Sarapis.
+
+\begin{figure}[!bp]
+\centering
+ \includegraphics[scale=0.5]{images/fig16}
+\caption{Base, Sarapis dedication -- Reconstruction of the bust of Sarapis}
+\label{fig:16}
+\end{figure}
+
+The representation of the god is not preserved, only the basis with the inscription 
+like in most other dedications. Nevertheless, if one checks the forms 
+in which Sarapis is figuratively represented, one immediately comes upon busts 
+of the god which sits on a globe. With this observation, we also have immediately 
+an explanation for the spherical form of the hollow carved on the upper part of 
+the base. Here the lower part of a globe would sink, on which the bust 
+of the god would likewise rest. This entire monument, the basis and the bust of 
+the god was set up in a shrine and probably took an important place there (Fig. \ref{fig:16})\footnote{Drawing of the reconstruction by Gisela Michel.}.
+
+As already mentioned, this extra-unusual object appeared until a short time ago 
+only as a text in the database Clauss-Slaby. The way in which it should be presented, 
+so that users of databases can decipher its complete meaning, follows with necessity 
+after the previous discussions: not only must all the sides be depicted but even 
+more important is the upper surface for which also the dimensions should be given 
+in this case. At least several photos of the monument are now visible in the database 
+Clauss-Slaby. Indeed, only all this information together can reveal as much of 
+the context as possible.
+
+These observations in one way or another are valid for all of our epigraphic texts. 
+The text alone is not enough, but needs - as far as possible – all the other 
+concrete details and photos not only of the text, but of the monument itself. In 
+such a way the access to ancient reality becomes easier, as, for example, in the 
+villa of Herennius Caecilianus in the area of Sirmione or at the shrine near Caparcotna/Legio 
+in northern Galilee. It is clear that we cannot completely reconstruct ancient 
+reality, but we should come as close as possible to the former reality. The digital 
+presentation is a crucial premise for this purpose.
 

### 2016-04-04 keynote/eck/Tradition and progress.tex
Eck's paper converted to LaTeX2e and added to the pdfs folder (no bibliography, though; refs are still in footnote stile -- and so they will remain, afaic)

--- 2016-04-04 keynote/eck/EAGLE_2016_Eck.tex
@@
+%&pdfLaTeX
 % !TEX encoding = UTF-8 Unicode
-% !TEX program = pdflatex
-% !TEX spellcheck = en_US
+\documentclass{article}
+\usepackage{ifxetex}
+\ifxetex
+\usepackage{fontspec}
+\setmainfont[Mapping=tex-text]{STIXGeneral}
+\else
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\fi
+\usepackage{textcomp}
 
-% In order to correctly compile this document,
-% execute the following commands:
-% 1. pdflatex
-% 2. pdflatex
-% 3. pdflatex
+\usepackage{amssymb}
+\usepackage{fancyhdr}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrulewidth}{0pt}
 
-\documentclass[amsthm,ebook]{saparticle}
+\pagestyle{fancy}
+\rhead{}
+\rfoot{}
+\chead{}
+\cfoot{}
+\lhead{\thepage{}}
+\begin{document}
 
-% IF YOU USE PDFLATEX
-\usepackage[utf8x]{inputenc}
-% if you write in english and in greek
-\usepackage{ucs}
-\usepackage[greek,english]{babel}
-\languageattribute{greek}{polutoniko}
+[2/13/2016 2:10 PM; 41021]
 
-\usepackage{teubner}
-% IF YOU USE XELATEX
-%\usepackage{polyglossia}
-% if you write in italian
-%\setmainlanguage{italian}
-% If you want put some ancient greek:
-%\setotherlanguage[variant=polytonic]{greek}
-%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}
+\begin{center}
+\textbf{Tradition and progress.}
 
-% dummy text (remove in a normal thesis)
-% remove if not necessary
-\usepackage{siunitx}
-%Natbib for bibliography management
-\usepackage[authoryear]{natbib}
-% custom commands
-\newcommand{\bs}{\textbackslash}
-
-%%%%%%%%
-%TITLE:%
-%%%%%%%%
-\title{Tradition and progress. The Roman World in the Digital Age - seen through Inscriptions}
-%\author[]{Werner Eck\corref{first}}
-\author[koe]{Werner Eck\corref{first}}
-\address[koe]{Universität zu Kölln} 
-\cortext[first]{Corresponding author. Email: Werner.Eck@uni-koeln.de}
-\begin{document}
-\maketitle
-%\begin{abstract}
-%\end{abstract}
+\textbf{The Roman World in the Digital Age - seen through Inscriptions. }
+\end{center}
 
@@
 Silvio Panciera during the 11\textsuperscript{th} International Congress of Greek 
-and Latin Epigraphy\footnote{See S.
-Panciera, Informatica, in: idem, Epigrafi, epigrafia, epigrafisti, vol.
-2, Rome 2006, 1913ff.
-– I would like to thank Aitor Blanco Perez for the translation of the paper.}.
-His fundamental initiative set in motion a process, which
+and Latin Epigraphy. His fundamental initiative set in motion a process, which 
 step by step will lead to a complete database. Very important steps in this direction 
@@
 different volumes of \textit{CIL III.} Since then, more than the same number of 
-new texts has appeared, perhaps around 400\footnote{According to information by Ioan Piso}. The growth is almost always dramatic 
+new texts has appeared, perhaps around 400. The growth is almost always dramatic 
 when intensive excavations lead to a massive increase in the amount of inscriptions 
-available, as somehow happens in the cases of Sagalassos or Perge in Pamphylia\footnote{In Sagalassos almost all the new inscriptions are in Greek}. 
+available, as somehow happens in the cases of Sagalassos or Perge in Pamphylia. 
 Perge provided only 4 Latin inscriptions for the \textit{CIL III} and today there 
-are at least 42\footnote{See the Clauss-Slaby database}. Even more dramatic is the growth in Caesarea Maritima. In \textit{CIL 
+are at least 42. Even more dramatic is the growth in Caesarea Maritima. In \textit{CIL 
 }there were only 3 Latin inscriptions known from the capital of the province of 
 Judea, while now we have in the Corpus Inscriptionum Iudaeae/Palaestinae more than 
-270 Latin documents\footnote{W. Eck, Comprendere contesti: Il nuovo multilingue Corpus Inscriptionum Iudaeae / Palaestinae, in: Rationes Rerum 1, 2013, 17 ff.}.
+270 Latin documents.
 
 Each of us is familiar with this massive increase in epigraphic monuments and we 
-are reminded of this every year by the always more extensive volumes of \textit{AE} and \textit{SEG}. Even if it is impossible to provide an exact number of all the 
+are reminded of this every year by the always more extensive volumes of \textit{AE 
+}and \textit{SEG}. Even if it is impossible to provide an exact number of all the 
 Latin inscriptions, one can still take as a starting point the fact that there 
@@
 EAGLE emerged, not to mention other participants as, among others, \textit{Hispania 
-Epigraphica }or the Bari Epigraphic Database \footnote{See the individual databases}. With regard to the Greek inscriptions, 
+Epigraphica }or the Bari Epigraphic Database. With regard to the Greek inscriptions, 
 it is much more difficult to give an approximate number because their digital record 
@@
 epigraphic tradition of the Greek and Roman antiquity as Latin and Greek inscriptions 
-– we should not forget this very important fact\footnote{\label{fn:7}At the 14.
-International Congress of Greek and Latin Epigraphy there was a special section for this phenomenon: W.
-Eck, Ph.
-Hyse, J.
-Price, St.
-J.
-Seidlmayer, I.-X.
-Adiego, in: ÖFFENTLICHKEIT – MONUMENT – TEXT.
-XIV Congressus Internationalis Epigraphiae Graecae et Latinae Akten, hg.
-W.
-Eck - P.
-Funke, Berlin 2014, 159 ff}.
-Our work as historians or philologists
+– we should not forget this very important fact. Our work as historians or philologists 
 is based on all those texts. But, the bigger the number of documents providing 
@@
 we would not find the following fragmentary inscription from Corinth, the capital 
-of the province, because the text has read as follows\footnote{\emph{CIL III} 537}:
+of the province, because the text has read as follows:
 
-\begin{quote}
-
-{\it Q(uinto) Vili[o --] / Titia[no] / Quadra[to --] / IIIIvir(o) vi[ar(um) cur(andarum), trib(uno)] / la\mqs{t}icl(avio) \mqs{le}g(ionis) [--, \textbf{leg(ato) prov(inciae)] / Ach(aiae)}, ob ius[titiam et fidem,] / quam circa [universam] / Achaia[m exhibuit,] / qu\mqs{a}m pro pr(aetore) [administravit.] / \mqs{M}(arcus) An\mqs{to}nius M(arci) [f(ilius)] / Nigrinus [--].}
-\end{quote}
+\textit{Q(uinto) Vili[o --] / Titia[no] / Quadra[to --] / IIIIvir(o) vi[ar(um) 
+cur(andarum), trib(uno)] / la}\textsuperscript{\textit{┌}}\textit{t}\textsuperscript{\textit{┐}}\textit{icl(avio) 
+}\textsuperscript{\textit{┌}}\textit{le}\textsuperscript{\textit{┐}}\textit{g(ionis) 
+[--, }\textit{\textbf{leg(ato) prov(inciae)] / Ach(aiae),}}\textit{ ob ius[titiam 
+et fidem,] / quam circa [universam] / Achaia[m exhibuit,] / qu}\textsuperscript{\textit{┌}}\textit{a}\textsuperscript{\textit{┐}}\textit{m 
+pro pr(aetore) [administravit.] / }\textsuperscript{\textit{┌}}\textit{M}\textsuperscript{\textit{┐}}\textit{(arcus) 
+An}\textsuperscript{\textit{┌}}\textit{to}\textsuperscript{\textit{┐}}\textit{nius 
+M(arci) [f(ilius)] / Nigrinus [--].}
 
@@
 of the office of quaestor:\textit{ [-- quaest(ori) prov(inciae)] / Ach(aiae) }instead 
-of \textit{[-- leg(ato) prov(inciae)] / Ach(aiae) }fits in the lacuna much better\footnote{In the database Clauss-Slaby there are around 68 examples in Latin inscriptions, in the database of the Packard Institute on can find 10 more in Greek texts.}.
+of \textit{[-- leg(ato) prov(inciae)] / Ach(aiae) }fits in the lacuna much better. 
 For a young senator who had completed an office of the \textit{vigintiviri }and 
@@
 }viginvirate and the legionary tribunate, this rule should automatically be observed 
-in a fragmentary career\footnote{See now the new reconstruction of the inscription by W.
-Eck, Quaestor, nicht legatus provinciae Achaiae.
-Zu einer Inschrift des Q.
-Vilius Quadratus aus Korinth, (in preparation).}.
-This kind of result is not achieved through an inquiry
+in a fragmentary career. This kind of result is not achieved through an inquiry 
 into a database but through a constructive consideration on the basis of personal 
@@
 But inscriptions do not only consist of text, although this may commonly seem to 
-be the case in the literature. In German, one speaks of ``\textit{Inschriften 
-errichten}'' and in Italy one can say: \textit{una iscrizione eretta}\footnote{It seems not necessary to give examples for this type of formulation; they are abundant.}. In the Prosopographia 
+be the case in the literature. In German, one speaks of \textit{„Inschriften 
+errichten``}and in Italy one can say: \textit{una iscrizione eretta}. In the Prosopographia 
 Imperii Romani,\textit{ }the expression \textit{titulum ponere }was long in use 
@@
 message and that was, on many occasions, the primary message for the Greek and 
-Roman public\footnote{\label{fn:12}Cf.
-now W.
-Eck, Wie ehrt man Mitglieder der staatlich-städtischen Elite?
-Inschriftenbasen unerkannter Reiterdenkmäler, Kolloquium Kassel März 2015, hg.
-K.
-Ruffing – H.
-Schneider (in print)}.
-In order to understand what an inscription can today convey as evidence,
+Roman public. In order to understand what an inscription can today convey as evidence, 
 it is fundamental to include the \textbf{monument }with which the inscription was 
@@
 whose form and particular details are connected with it. This necessity has meanwhile 
-become almost common knowledge – in contrast with Mommsen's time\footnote{W.
-Eck, Mommsen e il metodo epigrafico, in: Concordia e la X Regio.
-Giornate di Studio in onore di Dario Bertolini, Atti del Convegno Portogruaro 22-23 ottobre 1994, Padova 1995, 107 ff.}.
-On the EAGLE
+become almost common knowledge – in contrast with Mommsen's time. On the EAGLE 
 website, this is specifically taken into account with the option \textit{advanced 
@@
 
-One example from Volsinii (modern Bolsena) can illustrate this\footnote{AE 1980, 426.}. It is an honorific 
+One example from Volsinii (modern Bolsena) can illustrate this. It is an honorific 
 monument set up to commemorate a senator of the Hadrianic-Antonine age. The surviving 
@@
 said (as in the Epigraphischen Datenbank Heidelberg) that the text is to be read 
-on a base\footnote{EDR077846 = HD005099 (M.G. Schmidt).}. While this is correct, the plain term ``base'' does not allow to recognise 
+on a base. While this is correct, the plain term ``base'' does not allow to recognise 
 the most important element, namely with what type of honour had the city of Carthage 
@@
 is, nevertheless, revealed by the size and shape of the inscribed plaque. The plaque 
-is 76.5 cm high and 1.90 m wide\footnote{P. Gros, MEFRA 92, 1980, 977 ff.}. The monument was so wide that a normal statue 
+is 76.5 cm high and 1.90 m wide. The monument was so wide that a normal statue 
 could not have stood on it; this would have appeared completely out of place. Given 
@@
 A similar case can be inferred for Larcius Priscus, a legate of the \textit{legio 
-III Augusta}\footnote{\emph{CIL VIII} 17891 = Dessau 1055.}. The city of Thamugadi honoured him in a similar way to that of Carthago 
+III Augusta.} The city of Thamugadi honoured him in a similar way to that of Carthago 
 for the senator in Volsinii. On the forum of the colony founded by Trajan, not 
-a normal honorary statue not even an equestrian statue – as previously suspected\footnote{\label{note:18}J.
-Bergemann, Römische Reiterstatuen.
-Ehrendenkmäler im öffentlichen Bereich, Mainz 1990, 147 Nr.
-E 90:  ``equestrian statue''.}
+a normal honorary statue not even an equestrian statue – as previously suspected 
 – was set up but rather a \textit{biga }on which the statue of Larcius Priscus 
-was presented to the citizens of Timgad\footnote{The inscription is 188 cm wide, much more than necessary for an equestrian statue.}. But in EAGLE one finds the note:
+was presented to the citizens of Timgad. But in EAGLE one finds the note:
 
-Type of object: unbestimmt = undefined (= EDH)\footnote{EDH031159 (B. Gräf).}.
+Type of object: unbestimmt = undefined (= EDH).
 
-But one can see the type of the object on the photo, published by Zimmer in 1989\footnote{G.
-Wesch-Klein, Katalog der Inschriften, in: G.
-Zimmer, Locus datus decreto decurionum.
-Zur Statuenaufstellung zweier Forumsanlagen im römischen Africa, München 1989, 71 Nr.
-T 4.}.
+But one can see the type of the object on the photo, published by Zimmer in 1989. 
 More examples of this kind could be given. In my opinion, an important consequence 
@@
 which were once connected with an equestrian statue according to the investigations 
-of the author \footnote{See n. \ref{note:18}.}. Besides the statues directly referred to as such in the text, there 
+of the author. Besides the statues directly referred to as such in the text, there 
 are also those containing other hints which allow us to recognise that the honoured 
@@
 inscription, the type of statue is not mentioned but it can be inferred from the 
-size of the inscription\footnote{Corpus Inscriptionum Iudaeae/Palaestinae I 2, 718.}. Therefore, it must be registered as an equestrian statue 
+size of the inscription. Therefore, it must be registered as an equestrian statue 
 in a database. The same applies to the honours bestowed in Lepcis Magna upon the 
@@
 itself, it is only spoken of a \textit{statua marmorea} that should memorialise 
-the services of the magistrate\footnote{IRT 565.}. However, the base enables to recognise that a \textit{statua 
-equestris }once stood there because this base is 170 cm deep\footnote{Bergemann (n.
-\ref{note:18}) no.
-83.
-Cf.
-the inscription for the flamen provinciae Pompeius Cerealis Salvianus in Lepcis Magna: the basis is 70 cm high, but 160 cm deep (IRT 602).}.
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.5]{images/fig01ReiterstatueZeichnung.pdf}
-\caption{Equestrian statue: drawing}
-\label{fig:1}
-\end{figure}
-
-\begin{figure}[!bp]
-\centering
-\includegraphics[scale=0.5]{images/fig02ReiterstatueVonDerSeite.pdf}
-\caption{Equestrian statue: view from the side}
-\label{fig:2}
-\end{figure}
+the services of the magistrate. However, the base enables to recognise that a \textit{statua 
+equestris }once stood there because this base is 170 cm deep.
 
@@
 a \textit{flamen perpetuus }in Africa proconsularis, set up two or three equestrian 
-statues of the emperor Septimius Severus\footnote{\emph{CIL VIII} 14370 (Avedda).}. Even though these statues represent the 
+statues of the emperor Septimius Severus. Even though these statues represent the 
 emperor, they must have been small ones because the base is only 54 cm high and 
-35 cm wide\footnote{Bergemann (n.
-\ref{note:18}) no.
-81.}.
-These small equestrian statues fitted probably into the context in which they stood.
-On the other hand, gigantic equestrian statues were erected of
+35 cm wide. These small equestrian statues fitted probably into the context in 
+which they stood. On the other hand, gigantic equestrian statues were erected of 
 all the emperors – and not only of them. An exact relation between the size of 
@@
 
-\begin{figure}[!bp]
-\centering
-\includegraphics[scale=0.5]{images/fig03}
-\caption{Brescia Basis with 2 pillars}
-\label{fig:3}
-\end{figure}
-
 Most \textit{statuae equestres }either rose from bricked bases or the base was 
 made of solid stone. The base for the statue of C. Minicius Italus in Aquileia 
-was built with bricks and then covered with marble slabs\footnote{\emph{CIL V} 875 = Dessau 1374 = Inscr.
-Aquileiae I 495; G.
-Alföldy, Römische Statuen in Venetia et Histria, Heidelberg 1984, 98 f.
-no.
-87.
-For a photo see EDCS-01600153.}.
-The one of the young
-senator L. Fabius Severus in Tergeste was made of a single solid stone\footnote{\emph{CIL V} 532 = Dessau 6680 = Inscr. Italiae X 31. Photo : EDCS-04200621.}. The latter 
+was built with bricks and then covered with marble slabs. The one of the young 
+senator L. Fabius Severus in Tergeste was made of a single solid stone. The latter 
 applies to the majority of these monuments. Even so, at least in the first century 
@@
 until recently. For there existed equestrian statues, which seemed much lighter 
-and did not stand on an apparently solid basis\footnote{\label{fn:30}For the following discussion see W.
-Eck - H.
-von Hesberg Tische als Statuenträger, MDAI (R) 111, 2004 [2006], 143 ff.}.
-Sometimes the base was simply made
+and did not stand on an apparently solid basis. Sometimes the base was simply made 
 of a foundation slab, two supporting pillars and a cover plate on top, on which 
-the equestrian statue stood (Fig. \ref{fig:1} and \ref{fig:2}). To the best of my knowledge, there 
+the equestrian statue stood. (Fig. 1 and 2) To the best of my knowledge, there 
 is only one fully surviving example of this type, which is today kept in the museum 
-of Brescia. (Fig. \ref{fig:3}) This example presents a posthumous honour for a 6-year old 
+of Brescia. (Fig. 3) This example presents a posthumous honour for a 6-year old 
 boy decreed by the \textit{ordo} \textit{decurionum }of Brixia. The setting up 
-of the equestrian statue was executed by the father of the deceased\footnote{\emph{CIL V} 4441 = Inscr. Italiae X 5, 232.}. In this case, 
+of the equestrian statue was executed by the father of the deceased. In this case, 
 we recognise such a particular type of equestrian statue only because the entire 
@@
 there are the so-called \textit{trapezophora }on which an inscription is not rarely 
-found\footnote{For more information see Eck - von Hesberg (n.
-\ref{fn:30}).}.
-Just to mention some: we know for the young senator P.
-Numicius Pica Caesianus\footnote{\emph{CIL VI} 3835 = D 911= VI 31742 = 31743.}
+found. Just to mention some: We know for the young senator P. Numicius Pica Caesianus 
 two \textit{trapezophora }from Rome, quite a number from Torino for Q. Glitius 
-Atilius Agricola\footnote{\emph{CIL V} 6974 – 6987; Eck - von Hesberg (n. \ref{fn:30}) 186 f.}, a single one for T. Flavius Cimber, a municipal magistrate from 
-Urvinum Mautaurense\footnote{\emph{CIL XI} 6062: \emph{T(ito) Flavio L(uci) f(ilio) Ste(llatina) Cimbro pont(ifici), aed(ili) bis, IIIIvir(o) i(ure) d(icundo), quinq(uennali), praef(ecto) fabr(um), d(ecreto) d(ecurionum)}.} and many others\footnote{See Eck - von Hesberg (n.
-\ref{fn:30}) 180 ff.: a list of the inscribed trapezophora which were known to us in 2003/4.}.
+Atilius Agricola, a single one for T. Flavius Cimber, a municipal magistrate from 
+Urvinum Mautaurense and many others.
 
 These epigraphic monuments can partly be found in the EAGLE database too. Here 
-one has to search for the term \textit{trapezophorus }as object type\footnote{\emph{Trapezophoron} is the normal \emph{terminus technicus}.}. Table feet 
+one has to search for the term \textit{trapezophorus }as object type. Table feet 
 of this kind, which contain inscriptions, are known in relatively big numbers because 
-of their special shape; they have always been categorised individually\footnote{For the literature see Eck - von Hesberg (n. \ref{fn:30}) notes 1-3.}. However, 
+of their special shape; they have always been categorised individually. However, 
 they were not hitherto considered as parts of statue bases but rather of tables, 
@@
 }with inscriptions were in reality parts of statue dedications, with an equestrian 
-statue which stood on this specially arranged base\footnote{Eck - von Hesberg (n. \ref{fn:30}) with the general discussion of this type of honorary monuments.}. As long as these inscriptions 
+statue which stood on this specially arranged base. As long as these inscriptions 
 are marked up with the keyword \textit{trapezophorus }(sic!) in the EAGLE database, 
@@
 keyword \textit{mensa}. If one is entering \textit{mensa }in the EDH, numerous 
-and extremely varied objects appear, which have by no means the same function\footnote{For an example of a mensa in the concrete sense: IRT 590: \emph{Ti(berius) Cl(audius) Amicus M(arcus) Heliodorius Apollonides aed(iles) mensas p(ecunia) s(ua) d(ono) d(ederunt)}; the inscription is engraved on the frame of the table; cf.
-also \emph{CIL III} 15184, 18 = AIJ 310.
-On the other side also a mensa ponderaria can be found by surging for a mensa: AE 1905, 37 = HD030144; \emph{CIL III} 15025 = HD005744.}.
+and extremely varied objects appear, which have by no means the same function. 
 If the keyword \textit{mensa }is entered in the EDH and connected with the search 
 parameter ``honorific inscription'', it is possible to find three \textit{trapezophora} 
-– in the category \textit{mensa}\footnote{HD002671 = EDR111713; HD006112 (Ruck) = EDR076783; HD025725 (Féraudi) = EDR073154 (also described as \emph{mensa}).}.
-But who would imagine that the combination
+– in the category \textit{mensa}. But who would imagine that the combination 
 of \textit{mensa} and honorific inscription is necessary to find this type of monument? 
@@
 \textit{mensa }can almost only be used as long as the word occurs in the inscribed 
-text, not as term for an archaeological object\footnote{For an exception see the preceding note.}. Again the harmonization of search 
-terms becomes extremely important in order to enable a quick and safe search\footnote{In EDR077116 an inscription with the text: \emph{L(ucius) Ansius Quintill[i]anus m\d{e}[nsam -{}-{}-]} is called trapezophorus, although in reality it is a \emph{mensa}, as the text itself tells us and as the photo in EDCS-10700899 clearly shows.
-The inscription is engraved on the frame of the tabula for a \emph{mensa}.}.
+text, not as term for an archaeological object. Again the harmonization of search 
+terms becomes extremely important in order to enable a quick and safe search.
 
@@
 The posthumous honours in Brescia for the 6-year old P. Matienus Proculus is, as 
-already mentioned, such an example\footnote{\emph{CIL V} 4441 = Inscr. Italiae X 5, 232.}. But in the databases, inscriptions for such 
+already mentioned, such an example. But in the databases, inscriptions for such 
 monuments are not described with their specific features\textbf{ }as\textbf{ }the 
@@
 the database. The text for Publius Matienus Proculus one would not even find with 
-the word \textit{titulus honorarius}, because it is categorised as sepulcralis\footnote{EDR090232. The precise terminus would be: \emph{titulus honorarius postumus}.}. 
+the word \textit{titulus honorarius}, because it is categorised as sepulcralis. 
 To single out the different types, one has to describe in the data bases the special 
@@
 
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.6]{images/fig04}
-\caption{Herennius Caecilianus -- frontside}
-\label{fig:4}
-\end{figure}
-
 In Sirmione (ancient Sirmio), at the Lago di Garda, an inscription was found in 
 1960. It once belonged to an honorific monument for the young senator C. Herennius 
-Caecilianus.
-The following text was published by Alberto Albertini in 1973 (Fig.
-\ref{fig:4})\footnote{\label{fn:46}A.
-Albertini, Un patrono di Verona del secondo secolo d.C.: G.
-Erennio Ceciliano, in: Il territorio Veronese in età romana Verona 1973, 439ff.
-= Alföldy, Römische Statuen (n.
-28) 253; R.
-Bertolazzi - V.
-Guidorizzi, Supplementa Italiae 28, nr.
-7.}:
+Caecilianus. The following text was published by Alberto Albertini in 1973 (fig. 
+4):
 
-\begin{quote}
-\textit{C(aio) Herennio} \\
-\textit{C(ai) f(ilio) Pob(lilia)} \\
-\textit{Caeciliano,} \\
-\textit{adlect(o) in senat(um)} \\
-\textit{ab imp(eratore) Hadriano} \\
-\textit{Aug(usto), q(uaestori) prov(inciae) Narb(onensis),} \\
-\textit{trib(uno) pleb(is), $\overline{IIII}$viro i(ure) d(icundo)} \\
-\textit{Veronae,} \\
-\textit{patrono} \\
-\textit{d(ecreto) d(ecurionum)}. \\
-\end{quote}
+\textit{C(aio) Herennio}
+
+\textit{C(ai) f(ilio) Pob(lilia)}
+
+\textit{Caeciliano,}
+
+\textit{adlect(o) in senat(um)}
+
+\textit{ab imp(eratore) Hadriano}
+
+\textit{Aug(usto), q(uaestori) prov(inciae) Narb(onensis),}
+
+\textit{trib(uno) pleb(is), I̅I̅I̅I̅viro i(ure) d(icundo)}
+
+\textit{Veronae,}
+
+\textit{patrono}
+
+\textit{d(ecreto) d(ecurionum)}.
 
@@
 with a statue. This was also assumed by Albertini, who suggested a bronze statue. 
-The text also appears in the EDR and EDH\footnote{EDR093835 and HD033596.}.
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.5]{images/fig05}
-\caption{Herennius Caecilianus -- backside}
-\label{fig:5}
-\end{figure}
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.6]{images/fig06}
-\caption{Herennius Caecilianus -- lateral}
-\label{fig:6}
-\end{figure}
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.46]{images/fig07}
-\caption{Herennius Caecilianus -- from above}
-\label{fig:7}
-\end{figure}
+The text also appears in the EDR and EDH.
 
@@
 The text is surrounded with a frame on the front side. The same frame surrounds 
-also both laterals and, above all, the backside, which is crucial (Fig. \ref{fig:5} and \ref{fig:6}). 
+also both laterals and, above all, the backside, which is crucial (fig. 5 and 6). 
 For this shows that the backside was elaborated with the intention to be exposed, 
@@
 the statue directly standing on it, in his view) was not \textit{adossata a una 
-parete, ma eretta in uno spazzio}\footnote{Albertini (n. \ref{fn:46}).}. It is correct that the inscription could not 
+parete, ma eretta in uno spazzio.} It is correct that the inscription could not 
 have been \textit{adossata a una parete}. However, both the first editor and all 
 the others who dealt with the inscription thereafter have simply not wondered how, 
-then, could the statue stand on a slab which is only 29 cm deep (Fig. \ref{fig:7}). Furthermore, 
+then, could the statue stand on a slab which is only 29 cm deep (fig. 7). Furthermore, 
 the three holes for the corresponding dowel on the top side of the slab show that 
@@
 on an cover plate, which was resting on two pillars, whose front side with the 
-inscription was 29 cm deep (the second pillar is lost). This ``lighter'' 
+inscription was 29 cm deep (the second pillar is lost). This \texttt{"}lighter\texttt{"} 
 version of an equestrian statue was perhaps chosen by Verona because the monument 
@@
 these too are elaborated on the backside with the intention to be exposed. One 
-of the pillars bore once an inscription that was later erased\footnote{A more detailed argumentation in Eck, Wie ehrt man Mitglieder der staatlich-städtischen Elite?
-(n.
-\ref{fn:12})}, which makes it impossible
-to know who was honoured in such a way (Fig. \ref{fig:8}, \ref{fig:9}, \ref{fig:10}).
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.15]{images/fig08}
-\caption{Brescia -- erased inscription on a pillar}
-\label{fig:8}
-\end{figure}
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.15]{images/fig09}
-\caption{Brescia -- pillar}
-\label{fig:9}
-\end{figure}
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.05]{images/fig10}
-\caption{Brescia -- pillar, reused later}
-\label{fig:10}
-\end{figure}
+of the pillars bore once an inscription that was later erased, which makes it impossible 
+to know who was honoured in such a way (fig. 8-10).
 
@@
 provide all those details which are necessary for the complete interpretation of 
-an epigraphic monument\footnote{For the acts of the Congress see n. \ref{fn:7} above.}. The high costs that the inclusion of many pictures in previous 
+an epigraphic monument. The high costs that the inclusion of many pictures in previous 
 publications entailed are not a crucial problem any longer.
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.06]{images/fig11}
-\caption{Base, Sarapis dedication -- Front}
-\label{fig:11}
-\end{figure}
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.06]{images/fig12}
-\caption{Base, Sarapis dedication -- Eagle}
-\label{fig:12}
-\end{figure}
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.06]{images/fig13}
-\caption{Base, Sarapis dedication -- Victoria}
-\label{fig:13}
-\end{figure}
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.06]{images/fig14}
-\caption{Base, Sarapis dedication -- Eagle}
-\label{fig:14}
-\end{figure}
 
@@
 now with an example that comes from the material of the \textit{Corpus Inscriptionum 
-Iudaeae/Palaestinae}\footnote{This monument will be included in volume V of the Corpus Inscriptionum Iudaeae/Palaestinae.}. A short time ago one could find the inscription only in the 
-database Clauss-Slaby and only the text\footnote{EDCS-15200169; now there are also photos connected to the text.
-Under the adresse: \url{http://www.antiquities.org.il/t/item_en.aspx?CurrentPageKe y=33&rock=0} the monument can be seen on the homepage of the Rockefellermuseum in Jerusalem; there is only mentioned that an inscription is written on the monument; but the text is not given}.
-Avi Yonah, who greatly contributed to
+Iudaeae/Palaestinae}. A short time ago on could find the inscription only in the 
+database Clauss-Slaby and only the text. Avi Yonah, who greatly contributed to 
 the collection and publication of epigraphic monuments during the time of the British 
 mandate in Palestine, published in 1946 an inscription found near the legionary 
-camp of the \textit{legio VI Ferrata}, near Caparcotna = Legio\footnote{M. Avi-Yonah, Newly discovered Latin and Greek inscriptions, QDAP 12, 1946, 89 = AE 1948, 145.}. It is a round monument, 
-1.05 m high, which he – like many epigraphists later – presented as an altar\footnote{For example.
-L.
-Vidman, Sylloge inscriptionum religionis Isiacae et Sarapiacae, Berlin 1969, 182f.
-no.
-361; F.
-Mora, Prosopographia Isiaca.
-I.
-Corpus Prosopographicum Religionis Isiacae, Leiden – New York – Kopenhagen – Köln 1990, 243 Nr.
-577; N.
-Belayche, Iudaea-Palaestina, The Pagan Cults in Roman Palestine (Second to Fourth Century), Tübingen 2001, 59 ff.; L.
-Bricault, Recueil des inscriptions concernant les cultes isiaques (RICIS), 3 vols., Paris 2005, 508f.
-no.
-403/0201; N.
-Belayche, Les dévotions à Isis et Sérapis dans la Judée-Palestine romaine, in: Nile into Tiber.
-Egypt in the Roman World, hg.
-L.
-Bricault – M.J.
-Versluys – P.G.P.
-Mayboom, Leiden 2007, 451f.; W.
-Eck, Rom und Judaea.
-Fünf Vorträge zur römischen Herrschaft in Palaestina, Tübingen 2007, 186f.; P.
-Figueras, The Pagan Image of Greco-Roman Palestine and Surrounding Lands, Oxford 2013, 78.
-96.}.
+camp of the \textit{legio VI Ferrata}, near Caparcotna = Legio. It is a round monument, 
+1.05 m high, which he – like many epigraphists later – presented as an altar. 
 The monument shows three perfectly elaborated relieves on three sides: a Victoria 
@@
 two eagles that carry a thunderbolt in the crawls and a crown in the beak. The 
-inscription on the front face reads (Fig.
-\ref{fig:11}, \ref{fig:12}, \ref{fig:13}, \ref{fig:14})\footnote{The text of the inscription is the one corrected by W.
-Eck, Sarapis und die legio VI Ferrata.
-Die Weihung einer Sarapisbüste für das Wohl des Kaisers, ZPE 198, 2016 (in print).}:
+inscription on the front face reads (fig. 11-14):
 
-\begin{quotation}
 \textit{Pro salute et incolumitate / domini nostri [[Imp(eratoris) Caes(aris) M(arci) 
@@
 p(ilus).}
-\end{quotation}
-
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.06]{images/fig15}
-\caption{Base, Sarapis dedication -- Upper side}
-\label{fig:15}
-\end{figure}
 
@@
 of the inscription, in which it is written: \textit{praesentissim}\textit{\textbf{um}}\textit{ 
-de}\textit{\textbf{um}}\textit{ Mag(num) Sarapid}\textit{\textbf{em}}\footnote{One exception: O.
-Stoll, Zwischen Integration und Abgrenzung.
-Die Religion des römischen Heeres im Nahen Osten.
-Studien zum Verhältnis von Armee und Zivilbevölkerung im römischen Syrien und den Nachbargebieten, St.
-Katharinen 2001, 280, who saw the consequences of the accusativ for the interpretation of the monument.}.
-This evidently
+de}\textit{\textbf{um}}\textit{ Mag(num) Sarapid}\textit{\textbf{em}}. This evidently 
 means that Sarapis is not mentioned here as the one to which something was dedicated 
@@
 peculiarity on the upper side of the base,\textbf{ }namely a completely rounded 
-hollow with a diameter of 29 cm and a depth of 9 cm (Fig. \ref{fig:15}). It is an almost 
+hollow with a diameter of 29 cm and a depth of 9 cm (fig. 15). It is an almost 
 half sphere that was perfectly chiselled and smoothed from the marble. Sometimes 
@@
 
-\begin{figure}[!bp]
-\centering
- \includegraphics[scale=0.5]{images/fig16}
-\caption{Base, Sarapis dedication -- Reconstruction of the bust of Sarapis}
-\label{fig:16}
-\end{figure}
-
 The representation of the god is not preserved, only the basis with the inscription 
-like in most other dedications. Nevertheless, if one checks the forms 
+like in most other dedications. Nevertheless,\textbf{ }if one checks the forms 
 in which Sarapis is figuratively represented, one immediately comes upon busts 
@@
 an explanation for the spherical form of the hollow carved on the upper part of 
-the base. Here the lower part of a globe would sink, on which the bust 
+the base.\textbf{ }Here the lower part of a globe would sink, on which the bust 
 of the god would likewise rest. This entire monument, the basis and the bust of 
-the god was set up in a shrine and probably took an important place there (Fig. \ref{fig:16})\footnote{Drawing of the reconstruction by Gisela Michel.}.
+the god was set up in a shrine and probably took an important place there (fig. 
+16).
 
@@
 
+Köln          Werner Eck
+
+Werner.Eck@uni-koeln.de
+
+\newpage
+
 \end{document}

### 2016-04-04 keynote/roueché/EAGLE_2016_Roueche.tex
Roueche's paper transformed into LaTeX2e, but refs left as footnotes

--- 2016-01-13 keynote/roueché/EAGLE_2016_Roueche.tex
@@
 \title{Where Eagles Dare}
-\author[]{Charlotte Roueché\corref{first}}
+%\author[]{Werner Eck\corref{first}}
+\author[kcl]{Charlotte Roueché\corref{first}}
+\address[kcl]{King's College London} 
+\cortext[first]{Corresponding author. Email: charlotte.roueche@kcl.ac.uk}
 \begin{document}
 \maketitle
-\begin{abstract}
-The publication of inscriptions in digital form has been evolving since the 1980s, and online publication since the 1990s.
-It is useful to see where EAGLE is positioned in this evolution – a huge step forward, but not the end of the road.
-What can we learn from our progress so far?
-What have the main challenges been?
-And what does this suggest about the next stages?
+%\begin{abstract}
+%\end{abstract}
 
-\end{abstract}
+I couldn't resist this title, taken from Shakespeare's \textit{Richard III} to 
+use for a Hollywood film, full of daring wartime adventures. As academics we may 
+underestimate the importance of courage in our undertakings. But the early epigraphers 
+required a good deal of \textbf{daring}; while recording inscriptions was straightforward 
+enough in Italy or north-western Europe, the eastern and southern regions of the 
+Greco-Roman world remained difficult and dangerous to visit until the late nineteenth 
+century or even later – as is again becoming true. The early travellers also 
+suffered from practical constraints; in the 18\textsuperscript{th} and 19\textsuperscript{th} centuries 
+they were limited by the quantity of paper that they had brought with them just 
+as in the 20\textsuperscript{th} century; Robert `Palmyra' Wood recorded inscriptions 
+on pages of his copy of Homer. He was travelling for several months, and was saving 
+his paper for sketches, plans and drawings of buildings\footnote{Robert Wood (1716/17–1771): doi:10.1093/ref:odnb/29891}. Similarly John Deering 
+recorded texts in a notebook with very small pages\footnote{John Peter Deering (1787–1850): doi:10.1093/ref:odnb/7420}. Twentieth century travellers 
+were to experience similar constraints on the amount of photographic film available 
+to them.
 
-Charlotte Roueché studied Classics at Cambridge, and Byzantine Studies in Paris and Washington; as a pupil of Joyce Reynolds, she became involved in publishing Greek inscriptions, with a particular focus on material from Late Antiquity.
-In 1984 she was appointed to King’s College London, where an important centre of Digital Humanities started to develop in the 1990s; this encouraged her to explore the possibilities for publishing inscribed texts online.
-In 1999 Tom Elliott published the first proposal for EpiDoc; working with him and Gabriel Bodard, Charlotte was able to use EpiDoc to publish several online corpora: \url{http://insaph.kcl.ac.uk/ala2004, http://insaph.kcl.ac.uk/iaph2007}, and \url{http://irt.kcl.ac.uk/irt2009}.
-At present she is preparing a digital corpus of inscriptions of Roman Cyrenaica, and developing plans for a shared portal, Inscriptions of Libya.
-She has also recently published an eleventh century Byzantine text online, in a project exploring the uses of RDF to analyse intertextuality: \url{http://ancientwisdoms.ac.uk}.
+At the same time, epigraphers have always been ingenious in their use of technical 
+solutions. The most dramatic of these is perhaps the development of \textbf{squeezes}, 
+which have turned out to provide records of enduring and continuing value. The 
+driving force here was the need to record inscriptions in languages – hieroglyphics, 
+cuneiform – which could not be interpreted, and even incised designs; copyists 
+who know Greek or Latin could record texts in those languages with relative ease, 
+but they too came to find squeezes useful in representing what they saw. Over time 
+the technology was improved; while the plaster casts of Egyptian paintings could 
+seriously damage the monument, paper squeezes improved in quality\footnote{\url{http://www.asia.si.edu/research/squeezeproject/sq_making.asp}}.
+
+During the 19th century, \textbf{travel} became steadily safer, and also easier 
+- most dramatically as railways began to open up new regions. Over the same period, 
+the creation of the first great corpora of inscriptions encouraged an increasing 
+standardisation of records. Early travellers recorded as much as they could, often 
+in haste: this normally took the form of a simple transcription, or drawing, of 
+the text, with sometimes a brief mention of its support.  Gradually, measurements 
+start to appear – often only of letters, then of the monument or fragment itself. 
+In 1890 the Austrian government set up the Kleinasiatische Kommission\footnote{\url{http://www.oeaw.ac.at/en/science-and-society/commissions/kleinasiatische-kommission/geschichte/}}, which provided 
+travellers with special notebooks: these were pre-printed with headings for Location 
+and Position, Material, Height, Width, Depth, Letter heights, Shape and Condition, 
+Number and location of squeeze, When copied and by whom\footnote{Fund- und Standort, Material, Höhe, Breite, Dicke, Buchstabenhöhe, Form und Erhaltung, Nummer und Ortsangabe des Abklatsches, Wann und von wem copirt}.
+On return to Vienna, the
+notebooks could be filed, and the squeezes stored, in bookshelves and drawers specially 
+designed for the purpose.
+
+At the same time, the publication of the corpora, and the great projects from Boeckh's 
+\textit{CIG} to  Mommsen's \textit{CIL} was revealing the \textbf{volume} of material. 
+ The Ottoman world was becoming increasingly accessible; in the western part of 
+the Roman Empire development and industrialisation – particularly the redevelopment 
+of Rome as a capital city -- increased the torrent of material. Publishing an abundance 
+of texts, accompanied by increasingly detailed information, required a systematised 
+response. Organisation could be thematic: Christian inscriptions, for example, 
+were identified as a separate category, requiring different expertise, although 
+this division has remained problematic. It could be geographic: the Berlin Academy 
+took responsibility for publishing the material from Italy and the west, while 
+the Austrian Academy was to deal with Asia Minor and the East; but national interests 
+also played a part, with Italian and French scholars publishing materials from 
+the epigraphically prolific north African regions which their governments controlled. 
+The situation also demanded finding aids – the \textit{PIR} can be seen as a 
+tool for accessing the material in \textit{CIL}. And the pressure for standardisation 
+continued, although it was not until the 1930s that use of the Leiden conventions 
+for publishing inscribed texts was agreed (and modifications continued)\footnote{B.A.
+van Groningen, ‘Projet d’unification des systèmes de signes critiques’, \emph{Chronique d’Égypte}, 7 (1932): 262–9.7 ; for further modifications, see S.Dow, \emph{Conventions in Editing: A Suggested Reformulation of the Leiden System} (Durham, NC: Duke University, 1969); H.
+Krummrey and S.
+Panciera, ‘Criteri di edizione e segni diacritici’, Tituli, 2 (1980.
+205-15; S.
+Panciera, ‘Struttura dei supplementi e segni diacritici dieci anni dopo’, \emph{Supplementa Italica} 8 (Rome: Quasar, 1991), pp.
+9–21.}.
+
+Much of this reflected a response to the increased volume of material becoming 
+available as the world changed. But the end of the nineteenth century saw the beginning 
+of a further technological revolution with the arrival of \textbf{photography}. 
+The use of cameras for archaeological records was at first limited to established 
+excavation sites, or cities such as Athens, with a reasonable amount of infrastructure 
+and protection for cumbersome equipment; but by the 1920s cameras were sufficiently 
+portable to be taken out into the countryside. More and more inscriptions were 
+photographed; but the traditions which had already developed meant that they were 
+not immediately seen as essential elements in publication. The \textit{Monumenta 
+Asiae Minoris Antiqua} represent an honourable exception, established with the 
+specific aim of taking and publishing photographs; but they had the support of 
+American funding\footnote{C. Roueche, `The history of an idea : tracing the origins of the MAMA project.' \emph{Roman Phrygia : Culture and Society}. ed. P. Thonemann (Cambridge, 2013),  249-264.}.
+
+Gradually, however, photographs began to effect transformations in scholarly practice. 
+They permitted a far better understanding of regional, cultural and chronological 
+distinctions, which could be communicated to readers who would have no opportunity 
+to see the stones themselves. And they provided an increasing understanding of 
+\textbf{context} – whose importance was emphasised by both Louis Robert and, 
+more recently, Werner Eck. The photograph can present the support, and, when applicable, 
+the setting of that support; Robert emphasised the importance of visualising the 
+landscape surrounding a particular community.
+
+Robert also demanded ever higher \textbf{standards} in the accompanying commentaries 
+on inscriptions. But all of this raised a huge logistical problem: a text accompanied 
+by a detailed description, a detailed commentary and one or more photographs requires 
+a good deal of space – and more and more texts were appearing. From the 1980s 
+onwards it was also becoming standard to provide a translation into a modern language. 
+Publication in book form was becoming increasingly expensive and burdensome.
+
+It was therefore changes which had been brought about by a series of technological 
+developments which led far-sighted scholars – in particular Silvio Panciera  
+and Geza Alföldy – to look to yet another technology. As early as the 1980s 
+they both saw the value of \textbf{computers} as tools for holding, organising 
+and searching large volumes of text; others quickly followed. Panciera also understood 
+early on that working in this medium required collaboration, and the use of agreed 
+standards, convening meetings to discuss such matters from 1989 onwards.  With 
+the arrival of the web, and the resultant possibilities for communication, these 
+requirements became ever more important; in the early 2000s the agreed conventions 
+of epigraphy were translated into a set of machine readable instructions by Tom 
+Elliott, when he developed EpiDoc\footnote{\url{http://epidoc.sourceforge.net/}}.
+
+At the same time, the steady improvements in technology were making  it possible, 
+by the 2000s, to exchange images as well as texts; and the arrival of the digital 
+camera was transforming the possibilities for photographing texts. The epigrapher 
+in the field no longer depends on a finite supply of film: the traditional shot 
+of several fragments photographed together for reasons of economy is disappearing. 
+Instead, the epigrapher should be expected to present images of every side of a 
+monument and its setting. All these developments both enabled and necessitated 
+the first large scale publications of inscriptions on line: \textit{Vindolanda 
+Tablets Online}\footnote{\url{http://vindolanda.csad.ox.ac.uk/}}, (2003), the \textit{U.S. Epigraphy project}\footnote{\url{http://usepigraphy.brown.edu}}, (2003–), \textit{Aphrodisias 
+in Late Antiquity} (2004)\footnote{\url{http://insaph.kcl.ac.uk/ala2004/}},\textsuperscript{ }  the \textit{Inscriptions of Aphrodisias} 
+(2007)\footnote{\url{http://insaph.kcl.ac.uk/iaph2007/}}, and others in preparation.
+
+While we were working on the materials from Aphrodisias, new possibilities were 
+opening up: connections were becoming faster and more ubiquitous. More and more 
+relevant material was being published online: what Tim Berners-Lee calls t\label{HGoBack}\textit{he 
+next Web of open}\textit{\textbf{, linked data. }}In 2008 we received a grant to 
+start exploring the use of geodata with inscriptions, in the \textit{Inscriptions 
+of Roman Tripolitania} (2009)\footnote{\url{http://inslib.kcl.ac.uk/irt2009/}}; this approach has been further developed in \textit{Monumenta 
+Asiae Minoris Antiqua} XI\footnote{\url{http://mama.csad.ox.ac.uk/}},  and other kinds of interconnection are being actively 
+explored. Linking is closely connected to \textbf{sharing}: it is becoming increasingly 
+clear that one way of ensuring the survival of our materials is by making them 
+openly available for others to reuse and share as widely as possible.
+
+One of the aims of both Alföldy and Panciera had been to develop collaborative 
+corpora, places where large quantities of texts could be shared to enable extensive 
+searching. After their projects moved onto the web it also became possible to include 
+more and more material – photographs and other images, and geodata. It was Silvia 
+Orlandi who realised that the next step was a portal, to offer access across these 
+and many other online epigraphic collections. Work over several years, by many 
+different scholars, had established EpiDoc as a robust system, particularly for 
+the exchange of information, and it was therefore available to build \textbf{EAGLE}. 
+The spirit of this enterprise was exactly the spirit behind Europeana –  a project 
+to present high quality records of heritage materials to a worldwide audience.
+
+This is an account, therefore, of an academic discipline which has evolved by engagement 
+with a series of technological developments over two centuries, and is continuing 
+to do so; it is also a story of developing steadily higher standards for the publication 
+of heritage materials. The current challenge is to confront the fact that such 
+materials will now be universally available, and must therefore be presented in 
+a way that helps and supports \textbf{new audiences}. For this the EAGLE project 
+has been developing valuable new resources – such as the mobile app –and, very 
+importantly, encouraging translations. The crucial thing to realise is that it 
+will not be possible to revert to earlier models: this project sets new, higher 
+standards for epigraphic publication. This is a project which will take the subject 
+into the future `on eagles' wings', as the Bible puts it – in another phrase 
+used by Hollywood\footnote{Exodus 19.4.}.
+
+They should be taking us all with them.
 

### 2016-04-14 keynote/Hughes/EAGLE_2016_Hughes.tex
Lorna Hughes' keynote paper partly converted to LaTeX (not finished!)

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Infrastructures for Digital Research: New opportunities and challenges}
\author[gla]{Lorna Hughes\corref{first}}
\address[uniba]{Humanities Advanced Technology and Information Institute. University of Glasgow} 
\cortext[first]{Corresponding author. Email: Lorna.Hughes@glasgow.ac.uk}

\begin{document}
\maketitle
\begin{abstract}
This article is based on a keynote presentation given at the EAGLE Conference, Sapeinza University, Rome, on January 29th 2016. It draws on the author’s experience in digital humanities and in developing a digital collections research programme based in the National Library of Wales as an evidence base for understanding ways that the use and impact of digital cultural heritage can be enhanced and increased: an understanding that can provide a critical framework for future development of digital content by Galleries, Libraries, Archives and Museums (the so-called GLAM) sector.  This research is currently informing approaches to developing and using digital content in research that will be delivered through Europeana Research, a programme of engagement with the research community to make the metadata and content of Europeana.eu more accessible for research. This programme will have wider applicability for the development of other Research Infrastructures in the Arts and Humanities, cultivating a better awareness of the conditions that encourage use and re-use of digital heritage for research, ultimately enabling further exploitation of the ‘digital turn’ through the use of content, methods and content for research.
\end{abstract}

\section{Introduction: understanding the use of digital content}

The past twenty years has seen a very visible expansion of the digitization of Europe’s Cultural Heritage. However, to put the scale of this investment in perspective, the ENUMERATE\footnote{\url{http://www.enumerate.eu/}} project has produced data about the volume of digital cultural heritage available across Europe. The results of their analysis published in 2014 \cite[6]{ENUMERATE:2014aa} shows that over 10\% of the collections of European museums, archives and libraries has been digitized, over 300 million items. At the present scale of progress, it will take over 30 years to digitize the rest: a task that will be more complex taking into consideration that a large amount of material remaining to be digitized is 20th and 21st century material, either moving image, audio, or born-digital. A growing number of organisations have developed digital strategies to address this: 36\% of the institutions surveyed by ENUMERATE in 2014 have a written digitisation strategy.

Of the content that has been digitized, though, only 34\% is available online, and of that only 3\% is available for open re-use, though Creative Commons and unrestricted licensing. However, more than half of the organisations surveyed do not implement measures to quantify how their digital content is used. This lack of analysis of use isn’t just an issue in cultural heritage digitization: in 2008, the UK’s Arts and Humanities Research Council carried out an analysis of sustainability planning for projects funded through its Resource Enhancement Scheme, 173 projects, an investment of approximately £40 million \citep{Robey:2008aa}. Of these projects, only just over half were collecting usage statistics \citep[153]{Robey:2011aa}. A lack of enquiry into the use and impact of digital content is not a new issue: indeed, there has been a disproportionate investment in creation, management and curation of digital resources versus use of digital content for scholarship (in the UK from 1999-2009, the AHRC invested approximately £1.5 million into research that addressed the use of digital content, tools, and methods for research, but during this period they invested significantly more in digital content creation and curation). The lack of investigation in this area raises some serious questions about the value of the significant international investment in digital cultural heritage.

\section{What are we doing with all this digital stuff?}

It is perplexing that more institutions don’t try to develop a better understanding of what their users do with digital content, because methodologies exist for gathering and analyzing this information. Using analytic tools including Google Analytics\footnote{\url{https://www.google.co.uk/analytics/}} as a baseline, it is possible to augment numeric data with other qualitative and quantitative measures fro assessing users, including structured surveys, interviews, and content analysis tools (looking at citations of digital content, and its embedding in blogs and other resources). These approaches have been refined in a tried and tested approach developed by the Oxford Internet Institute (OII), called Tools for the Impact of Digitised Scholarly Resources (TIDSR)\footnote{\url{http://microsites.oii.ox.ac.uk/tidsr/}}. This has been used successfully to measure the impact of a range of resources in the arts and humanities\footnote{For an overview of the use of this method on a collection of Northeren Irish Parlaimentary Papers, see \citet{Hughes:2015aa}.}.  Similarly, methods developed by the CIBER initiative\footnote{\url{http://ciber-research.eu/}}, originally based at University College London, enable an understanding of the `digital footprint' of users of heritage content\footnote{\url{https://www.ucl.ac.uk/news/news-articles/1001/10010802}, and see also \citet{Nicholas:2014aa}.}. These approaches are an interesting way to build a narrative around data: using statistics about use to build a picture of engagement with digital cultural heritage. Overall a striking finding is the change in information seeking behaviours: users will multitask, and do many things lightly rather than one thing deeply: they have become what David Nicholas of CIBER referred to as `web foxes' \citep{Nicholas:2015aa} as people bounce around the digital domain using fast and abbreviated searches. Visitors to sites will frequently only access one page, then move on. This is even more noticeable when users access heritage data using mobile phones devices on the go: searching is shorter, and less engaged.

In order to investigate these questions in more detail, from 2011-14 I undertook some analysis of the use of the collections of the National Library of Wales, specifically \emph{Welsh Journals Online} and \emph{Rhyfel Byd 1914-1918 a’r profiad Cymreig / Welsh experience of the First World War 1914-1918}\footnote{\emph{Welsh Journals Online}: \url{http://welshjournals.llgc.org.uk/}; \emph{Cymru 1914-1918a}: \url{http://www.cymru1914.org}. The study into the use of Welsh Journals Online was written about in \citet{Hughes:2012aa}; the research into the use of Cymru1914.org was presented at the Sheffield Digital Humanities Conference, 2014: \url{http://www.hrionline.ac.uk/dhc/paper/19}}. This analysis showed several interesting things. Google analytic data for both sites showed that users mostly come from search engines, not the library website or interface, which is interesting in terms of the investment in presenting content. Similarly, many users were referred through family history websites, or media stories about the resource, rather than academic or subject-specific resources that linked to it: the highest number of referrals to \emph{Rhyfel Byd 1914-1918 a’r profiad Cymreig / Welsh experience of the First World War 1914-1918} came from a story on the BBC Wales News Website about the launch of the resource\footnote{\url{http://www.bbc.co.uk/news/uk-wales-25126781}}. The most popular search results are location based: people are looking for historic information about places they know. Interviews with selected groups of users of both resources expanded on these findings, establishing that the search box enabled most searches: that the `googlization' of information seeking had not affected the quality of information they were able to find. In fact, the only limitation that users reported was that there was not enough digital content in each resource. This is a consistent finding when surveying users of digital cultural heritage: as fast as digital content is launched, the demand increases for more material: `we have too many digital collections' said no survey respondent, ever. The benefits of unlimited access to information are so great that they compensate for problems that arise from irrelevant data, poor metadata, or digital dead ends.

\section{`We are all digital humanists now'}

Humanities research also exemplifies this enthusiastic embrace of the digital: In many respects, we are all digital humanists now:  we all use digital data for research by default (at the very least, electronic catalogues that point to analogue resources). Scholars rely on a steady consumption of digital source materials for scholarship and pedagogy, mostly created through large-scale digitisation initiatives in universities, libraries, museums and archives and as well as by commercial entities \citep{Ell:2013aa}. And humanists do not just use data: many scholars now create and manage these resources; we communicate via blogs and twitter, and use digital dissemination methods for sharing and challenging research results.  For many, this represents a sea change in research practice within the recognizable lifetime of a career.

But digital humanities is more than just the use of digital content for searching and browsing. If I was to attempt a definition of \emph{`digital humanities'} as an emerging field of research, I would suggest that it is combination of using digital content; with digital methods for the analysis and interpretation of this content; tools for specific scholarly tasks; and communicating research to the widest possible audience using traditional and non-traditional publishing methods. At its most effective, digitally enabled research in the humanities can facilitate and enhance existing research, making research processes easier via the use of computational tools and methods. More interestingly, it can enable research that would be impossible to undertake without digital resources and methods, allowing the formulation of new research questions that are driven by insights only achievable through the use of new tools and methods.

At the core of this sort of approach is collaboration: with researchers from scientific and other humanities disciplines, computational and technical fields, as well as cultural heritage organizations. Another important collaboration is with the audiences for our work, as user-led design, and participation via mechanisms like crowdsourcing, informs the development of many initiatives. It also relies on underlying technical infrastructures.

I like this characterization of digital humanities as about \emph{content, tools and methods}, because it creates a working environment with raw materials, tools for working with the raw material, and expertise in digital methods. This may seem a reduction of humanities practice, one that confirms Gary Hall’s view of the digital humanities as a use of tools and technologies that does not rely on a theoretical framework \citep{Hall:2012aa}: the sort of model that one can see as the basis for a DH research infrastructure, putting the maker’s perspective at the centre.  However, another school of thought in the digital humanities, exemplified by Alan \citet{Liu:2012aa}, sees the emphasis on building and making as a space for exploring and critique of culture. I would argue that the maker and theorist perspectives are not incompatible: in fact, it is through developing and building digital projects in the humanities \emph{through practice} that we can conduct cultural and critical analysis, by questioning many of the assumptions on which digital resources are built and communicated, and developing a better framework for understanding the ways that working with digital content, tools and methods is transforming our consumption and production of knowledge. By subsequently analyzing use of the digital collections and content that have been built, we can also develop a better understanding of their role in the humanities research lifecycle, and start to address questions about how digital content in particular is not just helping us to do research more effectively, but a disruptive, transformative intervention in the research lifecycle.

To address some of these questions, from 2011-15 the European Science Foundation funded the Research Network in Digital Methods in the Arts and Humanities (NeDiMAH.eu), to look at the practice of digital humanities across Europe, and to understand what researchers need in order to do digitally enabled research in the future. From a detailed programme of methodologically focused activities, several conclusions can be drawn about what can make digital resources and collections more valuable for scholarship\footnote{Full reports of all NeDiMAH activities are available at the Network’s website, \url{nedimah.eu}. These activities were structured around Working Groups, each addressing a specific area of Digital Humanities methods: Space and Time; Information Visualization; Linked Data and Ontological Methods; Developing Digital Data; Large-Scale Text Collections; and Scholarly Digital Editions.}. These conclusions sit alongside the user analysis described above to suggest three core observations about what the research community needs from digital content, and the infrastructures that deliver this content: access to, and `knowability' of digital content; opening up content for use and re-use; and creating better environments for drawing together multimedia content from a variety of sources for analysis and publication. These are described in more detail in the sections that follow.

\section{Access to, and `knowability' of digital content}

Much of what is classified as `digital humanities' is consumption: scholarly use of digital resources and born-digital material, working in what Roy \citet[22]{Rosenzweig:2010aa} called a culture of abundance.

 While this presents new possibilities it also raises significant challenges, and there is a need for a deeper understanding of the digital resources scholars rely on. As Tim \citet{Hitchcock:2013aa} has described in his article \emph{`confronting the digital: or how academic history writing lost the plot'}, researchers often work with data that is what we would charitably call ‘limited’ in its potential for re-use, thanks to problems with OCR, markup, and description and provenance information. Similarly, limitations in metadata often mean that while keyword searching gives the satisfying `quick hit' of a result, the user frequently misses important contextual information that enables more mindful engagement with sources, especially archival content: the scale of the original sources, their condition and their environment are often hidden when using digital surrogates that often appear as disembodied objects, separated from their context. Ryan \citet{Cordell:2015aa}, who has theorized a `network-author' function in antebellum newspapers from research based on digital newspaper archives, has said that `most digital archives hide more than they reveal, as keyword searches require prior knowledge of the texts to be discovered and can lead to evidentiary excess'.

Another issue is the selection of content for digitization. Due to limitations of funding, availability of analogue sources, and copyright and licensing issues, many digital archives are not ‘complete’. The National Library of Wales received funding from the Welsh Government and European Regional Development Fund (ERDF) to digitize 1 million pages of \emph{Welsh Newspapers Online}\footnote{\url{http://newspapers.library.wales/}} from its holdings. These date from 1800-1919, in Welsh and English. As a selection of the newspaper coverage of the time, this is an important resource for research and teaching: however, it is by no means a full representation of all newspapers of the period.

%<Figure 1 Home page: Welsh Newspapers Online>
\begin{figure}[!htbp]
\centering
 \includegraphics[scale=.28]{img/fig1.jpg}
\caption{Home page: Welsh Newspapers Online}
\label{fig:1}
\end{figure}

Figure \ref{fig:1} shows a chart on the home page of the resource showing the materials selected for digitization: many are missing from the 1909-19 period as a smaller, separate funding source was used for this. From the earlier period, many of the newspaper runs held by the library are incomplete. However, for the user, it’s too easy to assume that the resource is a complete representation of \emph{all} Welsh Newspapers, ever: the chart only shows the proportion of the newspapers in the resource, not as a proportion of those published. As we know, users (especially students) frequently turn to digital content by default when looking for historical sources, and as a free resource, this newspaper archive is very likely to be used, leading to the likelihood of research being conducted from an incomplete dataset.

\section{Opening up content for use and re-use}

A second requirement for researchers is the need for better, and more open, solutions for analysis and linking of digital content, which is all too frequently locked in digital silos, unable to be easily re-used and linked with other data from other providers and collections. Evaluation of users of digital content, and assessment of a range of European Digital Humanities projects, shows that scholars frequently have very simple questions or ideas that they want to test with data at the desktop, and they do not want the technology to be a barrier. When data is locked in ‘silos’, it can be difficult for the end user to integrate the tools they need analysis and linking with content held remotely. There needs to be greater disassociation of text and data from platform and delivery mechanisms, liberating digital resources for purposes unanticipated by the creators of digital resources \citep{Robinson:2014aa}.

For example, users often want to be able to use simple, pattern-finding tools, like nGram, with a range of data sets to quickly test hypotheses – the example in Figure 2 shows a visualization, using nGram, of the term `Belgian Refugees' in newspapers in Welsh Newspapers Online from 1914-1919. The graph shows a spike of heightened interest in the almost fashionable cause of Belgian refugees in late 1914, which tailed off as the war continued. A similar pattern can be seen in newspapers from around the British Isles\footnote{\citet{Hughes:2016aa}. The data for this graph is accessible at \url{http://dx.doi.org/10.5525/gla.researchdata.301}}. For questions like this, requiring simple pattern finding, this sort of tool can be a compelling and timesaving aid for researchers. However, as the tool is not part of the resource, using it with this collection required liaison with the National Library of Wales developer team to extract and work with the core data in this way.

One of the huge advantages of digital access is making the invisible visible in archives. We lose this when we can’t easily re-use and re-purpose this content.

\begin{figure}[!htbp]
\centering
 \includegraphics[scale=0.08]{img/fig2.jpg}
\caption{nGram of the term `Belgian refugees' in Welsh Newspapers Online}
\label{fig:2}
\end{figure}
%<Figure 2: nGram of the term ‘Belgian refugees’ in Welsh Newspapers Online>

\section{Creating better environments for drawing together multimedia content from a variety of sources for analysis and publication}

The attraction of working in a digital environment is the ability to integrate sources in a variety of formats, and to use them for new and unforeseen purposes. An example of working with hybrid, multimedia content can be seen in an AHRC funded project, \emph{The Snows of Yesteryear: Narrating Extreme Weather}\footnote{\url{Eira.llgc.org.uk}. The partners in the project were the Centre for Advanced Welsh and Celtic Studies at the University of Wales, the ACRE Project at the UK Met Office, the department of performance at Aberystwyth University, and Welsh performance artist, Eddie Ladd.}. The project was a creative collaboration to uncover archives in the collections of the National Library of Wales that contain information about the impact of extreme weather in Wales during the period from about 1700 (the pre-weather instrument period) to the 1960s. This included manuscripts (literary works, diaries, letters); printed materials, including Welsh ballads; and other material, including art works and manuscripts. These were digitized and made accessible to a team of climate scientists from the ACRE Project at the UK Met Office who used the data to fill gaps in the global picture of extreme weather in this period, and to use digital tools and methods to `map' and visualize incidents of extreme weather: floods, storms, freezing conditions, and incidents including the `year without a summer', in 1816 (after the Tambora volcano eruption caused severe climate abnormalities). The project also carried out community engagement, gathering interviews with local communities to share their experiences of extreme weather events within living memory: floods, storms, and snowfall. Form these two strands of evidence, we were able to construct two types of interpretation and analysis: the scientific visualisations; and also a public performance piece, \emph{Ghost Dance}, by Eddie Ladd: this drew on disparate narratives describing events of the winter of 1963 in an allegorical, not didactic way.

The project required sustaining a complex, hybrid archive of materials related to extreme weather impact in Wales, their analysis through performance and scientific visualization, and the communities that contribute this data. As such, it exemplifies the possibilities of digitally integrated research as set out in the Digital Humanities Manifesto: the ability to combine research, curation, and archive management to re-imagine the creation of digital content as a process of creative making, bringing together scholars and curators in ways that recast the role of each, driven by a shared goal of using historical sources for new purposes These approaches effect collaboration of disciplines and data types as an act of curation as much as a piece of scholarship, one in which scholars will not just create not just a collection of sources, but also effect a convergence of practices: artistic, scientific, and humanistic, and the ability to work with connected communities around the content\footnote{\url{http://manifesto.humanities.ucla.edu/}}. Documenting and publishing this sort of practice over the long term will also create a range of interesting issues -- how do we replicate the relationships between archives, scientific visualization, and performance? How is provenance of archives retained when embedded in a scientific visualization, or a performance? How will the convergence of digital and born digital be preserved and replicated over the long term? This form of knowledge production is an exciting means of determining the interface to information, data, and knowledge. It also raises significant information management research challenges associated with managing and sustaining complex collections over time, pushing the boundaries of what is currently known about stewardship and curation of humanities and cultural heritage content and data.

\bibliographystyle{sapauth-eng}
\bibliography{../../EAGLE}
\end{document}

### 2016-04-14 keynote/Hughes/bio/EAGLE_2016_Hughes.tex
Lorna Hughes' keynote paper partly converted to LaTeX (not finished!)

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Infrastructures for Digital Research: New opportunities and challenges}
\author[]{Lorna Hughes\corref{first}}
\begin{document}
\maketitle
\begin{abstract}
The expansion of digital cultural heritage from library, archive, museum and university collections has created tremendous opportunities for research, not least through new methods for collaboration and communication. The ‘digital turn’ has also enabled innovative approaches to representing spatial and temporal aspects of primary source materials. However, challenges still exist in accessing and using digital heritage: access to the content; availability and use of digital tools and methods; and the publication of research outputs in ways that are recognized by our peers.
 This presentation will discuss some approaches to addressing these issues, based on the theory and practice informing the development of digital research infrastructures. It will also discuss the role of the Europeana in the digital research space, and some important new developments internationally to support the use and description of digital content, tools and methods for research. 
\end{abstract}

\vspace{1cm}
Lorna M. Hughes is Professor of Digital Humanities at the University of Glasgow.Her research addresses the creation and use of digital cultural heritage for research, with a focus on collaborations between the humanities and scientific disciplines. A specialist in digital humanities methods, Hughes is the author of Digitizing Collections: Strategic Issues for the Information Manager (London: Facet, 2004), the editor of Evaluating \& Measuring the Value, Use and Impact of Digital Collections (London: Facet, 2011), and the co-editor of The Virtual Representation of the Past (London: Ashgate, 2007). She was the Chair of the European Science Foundation (ESF) Network for Digital Methods in the Arts and Humanities (\url{www.nedimah.eu}) from 2011-15, which developed the NeDiMAH Methods Ontology for the Digital Humanities (NeMO: \url{nemo.dcu.gr/}). Other notable digital projects include the AHRC-funded The Snows of Yesteryear: Narrating Extreme Weather (\url{eira.llgc.org.uk}) and the Jisc-funded digital archive, The Welsh Experience of the First World War (\url{cymruww1.llgc.org.uk}).

\end{document}

### 2016-04-14 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
Lorna Hughes' keynote paper partly converted to LaTeX (not finished!)

@@
  \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
-\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from \emph{Iscrizioni} 1997, sch. 3.12.14) and
+\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from \citealp{Manzella1997Felle}, sch. 3.12.14) and
 edition in ICVR, VII 19820.}
@@
 
-\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from \emph{Iscrizioni} 1997, sch. 3.8.3) and edition in
+\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from \citealp{Manzella1997Felle}, sch. 3.8.3) and edition in
 ICVR, VIII 22407.}
@@
 this aim: I think for instance to the solutions presented during the VIth EAGLE International Event in Bari by the
-lecture offered by Rebecca Benefiel and Holly Sypniewski about the \emph{Ancient Graffiti of Herculaneum project}.}, that can
+lecture offered by Rebecca Benefiel and Holly Sypniewski about the \emph{Ancient Graffiti of Herculaneum project}; see also in this volume the paper by Rebecca Benefiel}, that can
 be a suitable tool also to trail the activity of different stonecutters’ workshops that served – in the case of EDB
@@
 \caption{Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
-``useless'' caption  \emph{s(an)c(t)a Agnes}, age of pope Honorius I, 625/638. (photo: A. E. Felle).}
+``useless'' caption  \emph{s(an)c(t)a Agnes}, age of pope Honorius I. (photo: A. E. Felle).}
 \label{fig:8}
@@
 John 10.7: \begin{otherlanguage*}{greek}
-Εἶπεν οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι ἐγώ εἰμι ἡ θύρα τῶν προβάτων.
+\underline{Εἶπεν} οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι \underline{ἐγώ εἰμι ἡ θύρα τῶν προβάτων}.
 \end{otherlanguage*} (\emph{Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep});
@@
 John 10, 9: \begin{otherlanguage*}{greek}
-ἐγώ εἰμι ἡ θύρα· δι' ἐμοῦ ἐάν τις εἰσέλθῃ σωθήσεται καὶ εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει
+ἐγώ εἰμι ἡ θύρα· \underline{δι' ἐμοῦ ἐάν τις εἰσέλθῃ} σωθήσεται καὶ \underline{εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει}
 \end{otherlanguage*}  ( $^\textit{9}$ \emph{I am the gate;
@@
 building. Then, we have to face the issue of the description of the writing not only from the necessary point of
-view of palaeography ( we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
+view of palaeography (we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
 also in order to perceive and to understand its non-verbal significance: by its disposition, direction, shape. The

### 2016-05-04 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
FINAL folder added to hold all the edited files and final versions by La Sapienza; paper 40 by Felle added

@@
 \usepackage[authoryear]{natbib}
+\bibpunct{(}{)}{;}{a}{}{,}
 % custom commands
@@
 
-\newpage
+%\newpage
 On the open \emph{codex} is inscribed a focused - but barely readable - quotation from John 10, verses 7 and 9, where Jesus
@@
 \begin{quotation}
-John 10.7: \begin{otherlanguage*}{greek}
+John 10.7: \selectlanguage{greek}
 \underline{Εἶπεν} οὖν πάλιν ὁ ᾿Ιησοῦς, ᾿Αμὴν ἀμὴν λέγω ὑμῖν ὅτι \underline{ἐγώ εἰμι ἡ θύρα τῶν προβάτων}.
-\end{otherlanguage*} (\emph{Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep});
+\selectlanguage{english} (\emph{Therefore Jesus said again, ‘Very truly I tell you, I am the gate for the sheep});
 
-John 10, 9: \begin{otherlanguage*}{greek}
+John 10, 9: \selectlanguage{greek}
 ἐγώ εἰμι ἡ θύρα· \underline{δι' ἐμοῦ ἐάν τις εἰσέλθῃ} σωθήσεται καὶ \underline{εἰσελεύσεται καὶ ἐξελεύσεται καὶ νομὴν εὑρήσει}
-\end{otherlanguage*}  ( $^\textit{9}$ \emph{I am the gate;
+\selectlanguage{english}  ( $^\textit{9}$ \emph{I am the gate;
 whoever enters through me will be saved.} $^{[a]}$ \emph{They will come in and go out, and find pasture}).
@@
 \begin{quotation}
-((crux)) \begin{otherlanguage*}{greek}
+((crux)) \selectlanguage{greek}
 εἶπεν ὁ κ(ύριο)ς \textlatin{|} ἐγώ εἰμι \textlatin{|} ἡ θύρα τῶν \textlatin{|} προβάτων· \textlatin{|} δι᾿ ἐμοῦ \textlatin{||} ἐάν τις \textlatin{|} εἰσέλθῃ \textlatin{|} εἰσελεύσετ(αι) \textlatin{|} κ(αὶ) ἐξελεύσετ(αι) \textlatin{|} κ(αὶ) νομὴν \textlatin{|} εὑρήσει.
-\end{otherlanguage*}
+\selectlanguage{english}
 
@@
 
-\bibliographystyle{sapauth-eng}
+\bibliographystyle{sapienza-chicago}
 \bibliography{../../EAGLE}
